---
params:
  reportTitle: "CountQC"
title: "`r params$reportTitle`"
metadata-files: ["_fgcz-report.yml"]
---

```{r setup}
#| include: false
library(ezRun)
library(SummarizedExperiment)
library(DT)
library(kableExtra)
library(WGCNA)
library(plotly)
library(matrixStats)
library(reshape2)
library(tidyverse)
library(writexl)
library(gridExtra)
library(cowplot)
library(ggpubr)
library(cluster)
library(fpc)
library(clusterSim)
library(ggrepel)
library(htmlwidgets)
library(jsonlite)

## ── Data loading (qs2 with rds fallback, checks data/ first) ──
.fgcz_load <- function(name, nthreads = 4L) {
  paths <- c(
    file.path("data", paste0(name, ".qs2")),
    paste0(name, ".qs2"),
    file.path("data", paste0(name, ".rds")),
    paste0(name, ".rds")
  )
  for (p in paths) {
    if (file.exists(p)) {
      if (grepl("\\.qs2$", p)) return(qs2::qs_read(p, nthreads = nthreads))
      return(readRDS(p))
    }
  }
  stop("No .qs2 or .rds file found for: ", name)
}

## ── DT helper ─────────────────────────────────────────────────
.dt <- function(df, caption = NULL, pageLength = 10L,
                filter = "none", rownames = FALSE, escape = TRUE, ...) {
  DT::datatable(
    data = df, caption = caption, filter = filter,
    rownames = rownames, escape = escape,
    class = "compact stripe",
    options = list(
      pageLength = pageLength,
      scrollX = TRUE,
      dom = "tip",
      columnDefs = list(
        list(targets = "_all", className = "dt-left")
      )
    ),
    ...
  ) |> DT::formatStyle(columns = seq_len(ncol(df)),
                        `text-align` = "left",
                        `white-space` = "normal",
                        `word-wrap` = "break-word",
                        `max-width` = "300px")
}

## Emit a DT widget as raw HTML inside results='asis' chunks
.dt_asis <- function(...) {
  widget <- .dt(...)
  cat(as.character(htmltools::as.tags(widget)))
}

## ── Output directories ────────────────────────────────────────
for (d in c("plots", "excel", "html")) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}

## ── Plot saving (PNG + PDF to plots/) ─────────────────────────
.save_plot <- function(plot, name, width = 6, height = 5, dpi = 300) {
  if (!dir.exists("plots")) dir.create("plots", recursive = TRUE)
  png_path <- file.path("plots", paste0(name, ".png"))
  pdf_path <- file.path("plots", paste0(name, ".pdf"))
  ggplot2::ggsave(png_path, plot = plot, width = width, height = height,
                  dpi = dpi, bg = "white")
  ggplot2::ggsave(pdf_path, plot = plot, width = width, height = height,
                  device = "pdf", bg = "white")
  invisible(png_path)
}

## ── Agent metadata writer ─────────────────────────────────────
.write_metadata <- function(...) {
  args <- list(...)
  args$generatedAt <- format(Sys.time(), "%Y-%m-%dT%H:%M:%S%z")
  args$softwareVersions <- list(
    R = paste0(R.version$major, ".", R.version$minor),
    quarto = tryCatch(system("quarto --version", intern = TRUE)[1],
                      error = function(e) "unknown"))
  jsonlite::write_json(args, path = "report-metadata.json",
                       pretty = TRUE, auto_unbox = TRUE, null = "null")
}

rawData <- .fgcz_load("rawData")
output  <- metadata(rawData)$output
param   <- metadata(rawData)$param
```

```{r prepare-data}
#| include: false
debug <- FALSE

seqAnno <- data.frame(rowData(rawData), row.names = rownames(rawData),
                      check.names = FALSE)
dataset <- data.frame(colData(rawData), row.names = colnames(rawData),
                      check.names = FALSE)

types <- data.frame(row.names = rownames(seqAnno))
for (nm in setdiff(na.omit(unique(seqAnno$type)), "")) {
  types[[nm]] <- seqAnno$type == nm
}

metadata(rawData)$design <- ezDesignFromDataset(dataset, param)
colData(rawData)$conds   <- ezConditionsFromDesign(metadata(rawData)$design,
                                                   maxFactors = 2)
metadata(rawData)$condColors <- lapply(metadata(rawData)$design, getCondsColors)
metadata(rawData)$sampleColors <- list()
for (cond in colnames(metadata(rawData)$design)) {
  metadata(rawData)$sampleColors[[cond]] <-
    set_names(metadata(rawData)$condColors[[cond]][metadata(rawData)$design[[cond]]],
              rownames(metadata(rawData)$design))
}
```

::: {.panel-tabset}

# Settings

`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`

```{r check-enough-samples}
#| results: asis
if (ncol(rawData) < 2) {
  cat("Note: Statistics and Plots are not available for single sample experiments.", "\n")
  cat("Run the report again with multiple samples selected.", "\n")
  knitr::knit_exit()
}
```

```{r settings-table}
settings <- character()
settings["Reference:"]          <- param$refBuild
settings["Normalization method:"] <- param$normMethod
if (param$useSigThresh) {
  settings["Log2 signal threshold:"]    <- signif(log2(param$sigThresh), digits = 4)
  settings["Linear signal threshold:"]  <- signif(param$sigThresh, digits = 4)
}
settings["Feature level:"]        <- metadata(rawData)$featureLevel
settings["Number of features:"]   <- nrow(rawData)
settings["Data Column Used:"]     <- metadata(rawData)$countName

settings_df <- data.frame(Setting = settings, row.names = names(settings))
.dt(settings_df, caption = "Analysis settings", filter = "none",
    pageLength = 20L, rownames = TRUE)
```

```{r live-report}
#| echo: false
#| results: asis
if (exists("output") && !is.null(output)) {
  liveReportLink <- output$getColumn("Live Report")
  result <- EzResult(se = rawData)
  result$saveToFile(basename(output$getColumn("Live Report")))
  cat(paste0("[Live Report and Visualizations](", liveReportLink, ")", "{target=\"_blank\"}"), "\n")
}
```

# Data Files

The data files are in tabular text format and can also be opened with a spreadsheet program (e.g. Excel).

When opening with Excel, do make sure that the Gene symbols are loaded into a column formatted as 'text' that prevents
conversion of the symbols to dates). See

(https://www.genenames.org/help/importing-gene-symbol-data-into-excel-correctly)

```{r prepare-signal}
#| include: false
assays(rawData)$signal <- ezNorm(assays(rawData)$counts,
                                 presentFlag = assays(rawData)$presentFlag,
                                 method = param$normMethod) + param$minSignal
signalRange <- range(assays(rawData)$signal, na.rm = TRUE)

signalCond <- rowsum(t(log2(assays(rawData)$signal)), group = colData(rawData)$conds)
signalCond <- 2^t(sweep(signalCond, 1,
                        table(colData(rawData)$conds)[rownames(signalCond)], FUN = "/"))

isPresentCond <- rowsum(t(assays(rawData)$presentFlag * 1), group = colData(rawData)$conds)
isPresentCond <- t(sweep(isPresentCond, 1,
                         table(colData(rawData)$conds)[rownames(isPresentCond)], FUN = "/")) >= 0.5
rowData(rawData)$isValid <- rowMeans(isPresentCond) >= 0.5
```

```{r write-data-files}
#| include: false
if (!is.null(assays(rawData)$presentFlag)) {
  combined <- interleaveMatricesByColumn(assays(rawData)$signal,
                                        assays(rawData)$presentFlag)
} else {
  combined <- assays(rawData)$signal
}
if (!is.null(seqAnno)) {
  combined <- cbind(seqAnno[rownames(combined), , drop = FALSE], combined)
}
if (!is.null(combined$featWidth)) {
  combined$featWidth <- as.integer(combined$featWidth)
}

combined$isPresent_50P <- rowMeans(assays(rawData)$presentFlag) >= 0.5

# Raw counts
countFile <- file.path("excel", paste0(ezValidFilename(param$name), "-raw-count.xlsx"))
counts <- as_tibble(assays(rawData)$counts, rownames = "Feature ID")
if (!is.null(rowData(rawData)$gene_name)) {
  counts <- left_join(counts,
                      as_tibble(rowData(rawData)[, "gene_name", drop = FALSE],
                                rownames = "Feature ID"))
  counts <- dplyr::select(counts, "Feature ID", gene_name, everything())
}
write_xlsx(counts, path = countFile)

# Normalised signal
signalFile <- file.path("excel", paste0(ezValidFilename(param$name), "-normalized-signal.xlsx"))
write_xlsx(combined, path = signalFile)

selectSignals <- grepl("Signal", colnames(combined))
combined$"Mean signal"    <- rowMeans(combined[, selectSignals], na.rm = TRUE)
combined$"Maximum signal" <- rowMaxs(as.matrix(combined[, selectSignals]), na.rm = TRUE)
topGenesPerSample <- apply(combined[, selectSignals], 2, function(col) {
  col <- sort(col, decreasing = TRUE)
  if (length(col) > 100) col <- col[1:100]
  return(names(col))
})
topGenes <- unique(as.character(topGenesPerSample))

combined <- combined[order(combined$"Maximum signal", decreasing = TRUE), , drop = FALSE]
useInInteractiveTable <- c("seqid", "gene_name", "Maximum signal",
                           "Mean signal", "description", "featWidth", "gc")
useInInteractiveTable <- intersect(useInInteractiveTable, colnames(combined))
tableLink <- file.path("html", sub(".xlsx", "-viewHighExpressionGenes.html",
                                   basename(signalFile)))
## Select top genes
combinedTopGenes <- combined[which(rownames(combined) %in% topGenes), ]
## Restrict number of table rows if necessary
topGenesDf <- head(combinedTopGenes[, useInInteractiveTable, drop = FALSE],
                   param$maxTableRows)
ezInteractiveTableRmd(topGenesDf, digits = 3,
                      colNames = c("ID", colnames(topGenesDf)),
                      title = paste("Showing the top", nrow(topGenesDf),
                                    "genes with the highest expression")) |>
  DT::saveWidget(tableLink)

rpkmFile <- file.path("excel", paste0(ezValidFilename(param$name), "-rpkm.xlsx"))
rpkm <- as_tibble(getRpkm(rawData), rownames = "Feature ID")
if (!is.null(rowData(rawData)$gene_name)) {
  rpkm <- left_join(rpkm,
                    as_tibble(rowData(rawData)[, "gene_name", drop = FALSE],
                              rownames = "Feature ID"))
  rpkm <- dplyr::select(rpkm, "Feature ID", gene_name, everything())
}
write_xlsx(rpkm, path = rpkmFile)

tpmFile <- file.path("excel", paste0(ezValidFilename(param$name), "-tpm.xlsx"))
tpm <- as_tibble(getTpm(rawData), rownames = "Feature ID")
if (!is.null(rowData(rawData)$gene_name)) {
  tpm <- left_join(tpm,
                   as_tibble(rowData(rawData)[, "gene_name", drop = FALSE],
                             rownames = "Feature ID"))
  tpm <- dplyr::select(tpm, "Feature ID", gene_name, everything())
}
write_xlsx(tpm, path = tpmFile)
```

```{r data-file-links}
#| echo: false
#| results: asis
#| message: false
for (each in c(countFile, signalFile, rpkmFile, tpmFile, tableLink)) {
  cat("\n")
  cat(paste0("[", each, "](", each, ")"))
  cat("\n")
}
```

# Count Statistics

```{r count-statistics}
#| echo: false
#| message: false
#| warning: false
#| fig-width: !expr "min(7 + (ncol(rawData) - 10) * 0.3, 30)"
toPlot <- tibble(samples      = colnames(rawData),
                 totalCounts  = signif(colSums(assays(rawData)$counts) / 1e6, digits = 3),
                 presentCounts = colSums(assays(rawData)$presentFlag),
                 percentages  = paste(signif(100 * presentCounts / nrow(rawData), digits = 2), "%"))
m <- list(l = 80, r = 80, b = 200, t = 100, pad = 0)

## Total reads
plot_ly(toPlot, x = ~samples,
        y = ~totalCounts, type = "bar") |>
  layout(title  = "Total reads",
         yaxis  = list(title = "Counts [Mio]"),
         margin = m)

plot_ly(toPlot, x = ~samples, y = ~presentCounts, type = "bar",
        text = ~percentages, textposition = "auto") |>
  layout(title  = "Genomic features with reads above threshold",
         yaxis  = list(title = "Counts"),
         margin = m)

## Check for dominant genes in samples
countFracPerSample <- sweep(counts[, 3:ncol(counts)], 2,
                            colSums(counts[, 3:ncol(counts)]), FUN = "/")
rownames(countFracPerSample) <- make.unique(counts$gene_name)

df <- countFracPerSample |>
  as.data.frame() |>
  tibble::rownames_to_column("Gene") |>
  pivot_longer(-Gene, names_to = "Sample", values_to = "Fraction")

top3 <- df |>
  group_by(Sample) |>
  slice_max(Fraction, n = 3, with_ties = FALSE) |>
  mutate(rank = factor(row_number(), levels = 1:3)) |>
  ungroup()

p <- ggplot(df, aes(x = Sample, y = Fraction)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90")
p <- p + geom_point(data = top3, aes(colour = rank), size = 2,
                    position = position_dodge(width = 0.4))
p <- p + geom_text_repel(data = top3,
                         aes(label = Gene, colour = rank),
                         size          = 2.5,
                         nudge_y       = 0.02,
                         direction     = "y",
                         segment.size  = 0.2,
                         segment.alpha = 0.4,
                         max.overlaps  = Inf) +
  scale_colour_manual(values = c("red", "orange", "gold"), name = "Top-gene rank")
p <- p + theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        plot.margin = margin(5.5, 20, 5.5, 5.5, "pt"))
p <- p + labs(y = "Fraction of total expression",
              title = "Top 3 over-represented genes per sample")
p
```

# Correlation/Clustering Plot

```{r prepare-data-correlation}
#| include: false
if (!is.null(seqAnno$IsControl)) {
  rowData(rawData)$isValid <- rowData(rawData)$isValid & !seqAnno$IsControl
}
```

```{r check-enough-features}
#| results: asis
if (sum(rowData(rawData)$isValid) < 10) {
  cat("Not enough valid features for further plots", "\n")
  knitr::knit_exit()
}
```

```{r filter-valid-features}
#| include: false
rawDataAll <- rawData
rawData    <- rawData[rowData(rawData)$isValid, ]
assays(rawData)$log2signal <- log2(assays(rawData)$signal +
                                     param$backgroundExpression)
assays(rawData)$log2signalCentered <- sweep(assays(rawData)$log2signal, 1,
                                            rowMeans(assays(rawData)$log2signal))

metadata(rawData)$design <- ezDesignFromDataset(dataset, param)
colData(rawData)$conds   <- ezConditionsFromDesign(metadata(rawData)$design,
                                                   maxFactors = 2)

if (param$selectByFtest) {
  f_test_results <- apply(assays(rawData)$log2signalCentered, 1, function(row) {
    data <- data.frame(Expression = row, Group = colData(rawData)$conds)
    fit  <- lm(Expression ~ Group, data = data)
    anova(fit)["Group", "Pr(>F)"]
  })
  metadata(rawData)$topGenes <- names(sort(f_test_results)[1:param$topGeneSize])
} else {
  metadata(rawData)$topGenes <- rownames(rawData)[
    head(order(rowSds(assays(rawData)$log2signal, na.rm = TRUE),
               decreasing = TRUE), param$topGeneSize)]
}
```

::: {.panel-tabset}

## Sample correlation

```{r sample-correlation-setup}
#| include: false
figure.height <- min(max(7, ncol(rawData) * 0.3), 30)
```

```{r plot-sample-correlation}
#| echo: false
#| fig-height: !expr "figure.height"
#| fig-width: !expr "figure.height * 1.25"
## All genes
ezCorrelationPlot(cor(assays(rawData)$log2signal, use = "complete.obs"),
                  cond = colData(rawData)$conds,
                  condLabels = colData(rawData)$conds,
                  main = paste0("all present genes (", nrow(rawData), ")"),
                  colors = metadata(rawData)$sampleColors[[1]])

## Top genes
ezCorrelationPlot(cor(assays(rawData)$log2signal[metadata(rawData)$topGenes, ],
                      use = "complete.obs"),
                  cond = colData(rawData)$conds,
                  condLabels = colData(rawData)$conds,
                  main = paste0("top genes (", length(metadata(rawData)$topGenes), ")"),
                  colors = metadata(rawData)$sampleColors[[1]])
```

```{r sample-clustering}
#| echo: false
#| fig-height: 7
#| fig-width: !expr "min(7 + (ncol(rawData) - 10) * 0.3, 30)"
#| results: asis
if (ncol(rawData) > 3) {
  ## All genes
  d  <- as.dist(1 - cor(assays(rawData)$log2signal, use = "complete.obs"))
  hc <- hclust(d, method = "ward.D2")
  plotDendroAndColors(hc,
                      colors = as.data.frame(bind_cols(metadata(rawData)$sampleColors)),
                      autoColorHeight = TRUE, hang = -0.1,
                      main = "all present genes")

  ## Top genes
  d  <- as.dist(1 - cor(assays(rawData)$log2signal[metadata(rawData)$topGenes, ],
                        use = "complete.obs"))
  hc <- hclust(d, method = "ward.D2")
  plotDendroAndColors(hc,
                      colors = as.data.frame(bind_cols(metadata(rawData)$sampleColors)),
                      autoColorHeight = TRUE, hang = -0.1,
                      main = paste("top", length(metadata(rawData)$topGenes), " genes"))
}
```

:::

# Clustering of High Variance Features

```{r clustering-high-variance}
#| echo: false
#| message: false
#| fig-width: !expr "max(8, 4 + 0.15 * ncol(rawData))"
#| fig-height: 10
#| results: asis
#| eval: !expr "!debug"
if (ncol(rawData) > 3) {
  varFeatures <- rownames(rawData)[
    head(order(rowSds(assays(rawData)$log2signal), decreasing = TRUE),
         param$maxGenesForClustering)]
  notVarFeatures <- rownames(rawData)[
    which(rowSds(assays(rawData)$log2signal) <= param$highVarThreshold)]
  varFeatures <- setdiff(varFeatures, notVarFeatures)
  param$highVarThreshold <- signif(
    min(rowSds(assays(rawData)$log2signal[varFeatures, ])), digits = 3)

  if (length(varFeatures) > param$minGenesForClustering) {
    cat(paste("Threshold for std. dev. of log2 signal across samples:",
              param$highVarThreshold), "\n")
    cat("\n")
    cat(paste("Number of features with high std. dev.:", length(varFeatures)), "\n")
    cat("\n")
    clusterColors <- hcl.colors(param$nSampleClusters, palette = "Spectral")

    clusterResult <- clusterPheatmap(
      x      = assays(rawData)$log2signalCentered[varFeatures, ],
      design = metadata(rawData)$design,
      param  = param, clusterColors = clusterColors,
      lim    = c(-param$logColorRange, param$logColorRange),
      doClusterColumns = TRUE,
      condColors = metadata(rawData)$condColors)

    xTmp <- enframe(clusterResult$clusterNumbers, "feature_id", value = "Cluster")
    xTmp <- left_join(xTmp,
                      as_tibble(rowData(rawData)[, "gene_name", drop = FALSE],
                                rownames = "feature_id"))
    xTmp <- arrange(xTmp, Cluster)
    write_tsv(xTmp, file = "cluster-heatmap-clusterMembers.txt")

    plot(clusterResult$pheatmap$gtable)

    if (doGo(param, seqAnno)) {
      clusterResult <- goClusterResults(
        assays(rawData)$log2signalCentered[varFeatures, ],
        param, clusterResult,
        seqAnno          = seqAnno,
        universeProbeIds = rownames(seqAnno))
    }

    if (!is.null(clusterResult$GO)) {
      goTables <- goClusterTableRmd(param, clusterResult, seqAnno)
      if (doEnrichr(param)) {
        goAndEnrichr <- cbind(goTables$linkTable, goTables$enrichrTable)
      } else {
        goAndEnrichr <- goTables$linkTable
      }
      bgColors <- rep(gsub("FF$", "", clusterResult$clusterColors))
      rownames(goAndEnrichr) <- paste0('<font color="', bgColors,
                                       '">Cluster ', rownames(goAndEnrichr),
                                       '</font>')
      kable(goAndEnrichr, escape = FALSE, row.names = TRUE, format = "html",
            caption = "GO categories of feature clusters") |>
        kable_styling(bootstrap_options = "striped",
                      full_width = F, position = "float_right") |>
        footnote(general = "Cluster font color corresponds to the row colors in the heatmap plot.")
    } else {
      cat("No information available", "\n")
    }
  }
} else {
  varFeatures <- c()
}
```

```{r render-go-cluster}
#| include: false
#| eval: !expr "!debug"
if (ncol(rawData) > 3) {
  if (length(varFeatures) > param$minGenesForClustering) {
    if (!is.null(clusterResult$GO)) {
      file <- file.path(system.file("templates", package = "ezRun"),
                        "CountQC_goClusterTable.Rmd")
      file.copy(from = file, to = basename(file), overwrite = TRUE)
      rmarkdown::render(input = basename(file), envir = new.env(),
                        output_dir = "html", output_file = "goClusterTable.html",
                        quiet = TRUE)
    }
  }
}
```

```{r go-cluster-link}
#| echo: false
#| results: asis
#| eval: !expr "!debug"
if (file.exists("html/goClusterTable.html")) {
  cat(paste0("[GO cluster tables](html/goClusterTable.html)"), "\n")
  cat("\n")
}
```

```{r cluster-heatmap-members}
#| echo: false
#| results: asis
#| eval: !expr "!debug"
if (file.exists("cluster-heatmap-clusterMembers.txt")) {
  cat(paste0("[Heatmap cluster members](cluster-heatmap-clusterMembers.txt)"), "\n")
  cat("\n")
}
```

<details>
<summary>**Evaluate number of gene clusters**</summary>
Multiple metrics were computed for a comprehensive assessment.

* Compactness: Dunn Index, CH Index
* Separation: Silhouette Width, Davies-Bouldin Index
* Cluster Stability: Gap Statistic, Entropy

```{r cluster-scores}
#| echo: false
#| message: false
#| results: asis
#| eval: !expr "!debug"
if (length(varFeatures) > param$minGenesForClustering) {

  dist_matrix <- dist(assays(rawData)$log2signalCentered[varFeatures, ])
  hc          <- hclust(dist_matrix, method = "ward.D2")

  hclust_cluster <- function(X, k) {
    res          <- data.frame(clusters = rep(0, nrow(X)))
    dist_matrix  <- dist(X)
    hc           <- hclust(dist_matrix, method = "ward.D2")
    res$clusters <- cutree(hc, k = k)
    return(res)
  }
  gap_stat <- clusGap(assays(rawData)$log2signalCentered[varFeatures, ],
                      FUN = hclust_cluster, K.max = 10, B = 50)

  myClusterStats <- data.frame(nClusters = c(2:10), dunn = NA, ch = NA,
                               db = NA, silhouette = NA, entropy = NA,
                               gap = round(gap_stat$Tab[2:10, 3], 4))
  for (i in 2:10) {
    clusters       <- cutree(hc, k = i)
    cluster_scores <- cluster.stats(dist_matrix, clusters)
    myClusterStats$dunn[i - 1]      <- round(cluster_scores$dunn, 4)
    myClusterStats$ch[i - 1]        <- round(cluster_scores$ch, 4)
    myClusterStats$db[i - 1]        <- round(
      index.DB(assays(rawData)$log2signalCentered[varFeatures, ], clusters)$DB, 4)
    myClusterStats$silhouette[i - 1] <- round(cluster_scores$avg.silwidth, 4)
    myClusterStats$entropy[i - 1]   <- round(cluster_scores$entropy, 4)
  }

  datatable(myClusterStats,
            rownames = FALSE,
            options  = list(dom = "t")) |>
    formatStyle("dunn",
                background         = styleColorBar(range(myClusterStats$dunn), "lightblue"),
                backgroundSize     = "98% 15px",
                backgroundRepeat   = "no-repeat",
                backgroundPosition = "center") |>
    formatStyle("ch",
                background         = styleColorBar(range(myClusterStats$ch), "lightblue"),
                backgroundSize     = "98% 15px",
                backgroundRepeat   = "no-repeat",
                backgroundPosition = "center") |>
    formatStyle("db",
                background         = styleColorBar(range(myClusterStats$db), "lightgreen"),
                backgroundSize     = "98% 15px",
                backgroundRepeat   = "no-repeat",
                backgroundPosition = "center") |>
    formatStyle("silhouette",
                background         = styleColorBar(range(myClusterStats$silhouette), "lightgreen"),
                backgroundSize     = "98% 15px",
                backgroundRepeat   = "no-repeat",
                backgroundPosition = "center") |>
    formatStyle("entropy",
                background         = styleColorBar(range(myClusterStats$entropy), "lightsalmon"),
                backgroundSize     = "98% 15px",
                backgroundRepeat   = "no-repeat",
                backgroundPosition = "center") |>
    formatStyle("gap",
                background         = styleColorBar(range(myClusterStats$gap), "lightsalmon"),
                backgroundSize     = "98% 15px",
                backgroundRepeat   = "no-repeat",
                backgroundPosition = "center")
}
```
</details>

# MDS Plot

```{r mds-message}
#| echo: false
#| results: asis
#| message: false
if (ncol(rawData) <= 3) {
  cat("MDS plot is not possible with fewer than 4 samples. \n")
}
```

```{r mds-plots}
#| echo: false
#| message: false
#| warning: false
#| fig-height: 7
#| fig-width: 9
if (ncol(rawData) > 3) {
  ezMdsPlotly(assays(rawData)$log2signal,
              design     = metadata(rawData)$design,
              ndim       = 3, main = "mdsPlot_PresentGenes 3D",
              condColors = metadata(rawData)$condColors[[1]])
}
if (ncol(rawData) > 3) {
  ezMdsPlotly(assays(rawData)$log2signal[metadata(rawData)$topGenes, ],
              design     = metadata(rawData)$design, ndim = 3,
              main       = "mdsPlot_TopGenes 3D",
              condColors = metadata(rawData)$condColors[[1]])
}
if (ncol(rawData) > 3) {
  ezMdsPlotly(assays(rawData)$log2signal,
              design     = metadata(rawData)$design, ndim = 2,
              main       = "mdsPlot_PresentGenes 2D",
              condColors = metadata(rawData)$condColors[[1]])
}
if (ncol(rawData) > 3) {
  ezMdsPlotly(assays(rawData)$log2signal[metadata(rawData)$topGenes, ],
              design     = metadata(rawData)$design, ndim = 2,
              main       = "mdsPlot_TopGenes 2D",
              condColors = metadata(rawData)$condColors[[1]])
}
```

# Scatter Plots by Conditions

```{r scatter-plot-setup}
#| echo: false
#| message: false
#| results: asis
scatterPlotData <- countQcScatterPlots(param, metadata(rawData)$design,
                                       colData(rawData)$conds, rawDataAll,
                                       signalCond, isPresentCond, types = types)
nPlots <- max(scatterPlotData$nPlots, 1)
```

```{r scatter-plot-all-pairs}
#| echo: false
#| message: false
#| results: asis
#| fig-width: !expr "min(max(ncol(signalCond) * 2, 4), 20)"
#| fig-height: !expr "min(max(ncol(signalCond) * 2, 4), 20) * (16 / 15)"
#| eval: !expr "ncol(rawData) > 3 & param$writeScatterPlots"
qcScatterFiles <- scatterPlotData$allPairs

cat("\n")
cat("#### allPairs \n")
for (gridOfPlots in qcScatterFiles) {
  scatterPlots <- gridOfPlots$scatter
  axisLabels   <- gridOfPlots$axisLabels
  nItems       <- gridOfPlots$nItems

  grobList        <- list()
  idxLabel        <- 1
  idxScatterPlot  <- 1
  fontSize        <- max(4, 10 - nItems)

  for (idxRow in 1:(nItems + 1)) {
    if (idxRow == nItems + 1) {
      axisLabel <- ""
    } else {
      axisLabel <- axisLabels[idxLabel]
      idxLabel  <- idxLabel + 1
    }
    colLabel  <- text_grob(axisLabel, rot = 90, size = fontSize)
    grobList  <- c(grobList, list(colLabel))
  }

  for (rowIndex in 1:nItems) {
    for (colIndex in 1:nItems) {
      if (colIndex == nItems) {
        colLabel  <- text_grob(axisLabels[idxLabel], size = fontSize)
        grobList  <- c(grobList, list(scatterPlots[[idxScatterPlot]]), list(colLabel))
        idxLabel  <- idxLabel + 1
      } else {
        grobList  <- c(grobList, list(scatterPlots[[idxScatterPlot]]))
      }
      idxScatterPlot <- idxScatterPlot + 1
    }
  }

  plotGrob <- plot_grid(
    plotlist    = grobList,
    nrow        = nItems + 1,
    ncol        = nItems + 1,
    rel_heights = c(rep(1, nItems), 0.1),
    rel_widths  = c(0.1, rep(1, nItems)),
    byrow       = FALSE
  )

  title <- ggdraw() +
    draw_label(gridOfPlots$main, x = 0, hjust = 0) +
    theme(plot.margin = margin(0, 0, 0, 7))

  print(plot_grid(title, plotGrob, nrow = 2, ncol = 1, rel_heights = c(1, 15)))
  cat("\n")
}
cat("\n")
```

```{r scatter-plot-narrow}
#| echo: false
#| message: false
#| results: asis
#| fig-width: !expr "min(nPlots, 6) * 5"
#| fig-height: !expr "max(ceiling(nPlots / 6), 1) * 5"
#| out-width: !expr "paste0(min(nPlots + 1, 6) * 167, 'px')"
#| eval: !expr "ncol(rawData) > 3 & param$writeScatterPlots"
qcScatterFiles <- scatterPlotData$narrowPlots

if (length(qcScatterFiles) > 0) {
  for (i in 1:length(qcScatterFiles)) {
    cat("\n")
    cat("####", names(qcScatterFiles)[i], "\n")
    rowsOfPlots <- qcScatterFiles[[i]]
    for (rowOfPlots in rowsOfPlots) {
      scatterPlots <- rowOfPlots$scatters

      legend_temp <- get_legend(
        scatterPlots[[1]] +
          theme(legend.text = element_text(size = 10)) +
          guides(color = guide_legend(override.aes = list(size = 5)))
      )

      scatterPlots <- lapply(scatterPlots, function(x) {
        x +
          coord_fixed(xlim = c(1e1, NA), ylim = c(1e1, NA), expand = TRUE) +
          theme(legend.position = "none",
                text             = element_text(size = 10),
                aspect.ratio     = 1)
      })

      scatterPlots[["legend"]] <- legend_temp

      numRowsFinal <- rowOfPlots$nrow
      if (rowOfPlots$ncol * rowOfPlots$nrow > length(rowOfPlots$scatters)) {
        numColsFinal <- rowOfPlots$ncol
      } else {
        numColsFinal <- rowOfPlots$ncol + 1
        if (length(rowOfPlots$scatters) < numColsFinal * (numRowsFinal - 1)) {
          numRowsFinal <- numRowsFinal - 1
        }
      }

      grid.arrange(grobs   = scatterPlots,
                   nrow    = numRowsFinal,
                   ncol    = numColsFinal,
                   widths  = rep(6, numColsFinal),
                   heights = rep(6, numRowsFinal))
      cat("\n")
    }
    cat("\n")
  }
}
```

# EnrichR for topMarkers

Based on top 500 genes per sample selected by TPM and filtered for known housekeeping genes reported in the [HRT atlas](https://housekeeping.unicamp.br/homePageGlobal.html).

```{r enrichr-markers}
#| echo: false
#| results: asis
#| warning: false
#| eval: !expr "unique(dataset$featureLevel) != 'smRNA'"
maxGenes <- 500
dat      <- tpm[, 3:ncol(tpm)]
rownames(dat) <- make.unique(tpm$gene_name)
topMarkersPerSample <- apply(dat, 2, function(col) {
  col <- sort(col, decreasing = TRUE)
  if (length(col) > maxGenes) col <- col[1:maxGenes]
  return(names(col))
})

hrtGenes <- ezRead.table("/srv/GT/databases/HRT/Human_Mouse_Common.csv",
                         sep = ";", row.names = NULL)
hrtGenes <- c(hrtGenes$Human, hrtGenes$Mouse)
topMarkers <- list()
for (j in 1:ncol(topMarkersPerSample)) {
  topMarkers[[colnames(topMarkersPerSample)[j]]] <-
    topMarkersPerSample[!topMarkersPerSample[, j] %in% hrtGenes, j]
}
topMarkers[["AllSamples"]] <- unique(unlist(lapply(topMarkers, head, n = maxGenes / 2),
                                            use.names = FALSE))
topMarkers <- topMarkers[c("AllSamples", colnames(topMarkersPerSample))]

jsCall <- paste0('enrich({list: "', sapply(topMarkers, paste, collapse = "\\n"),
                 '", popup: true});')
enrichrCalls <- paste0("<a href='javascript:void(0)' onClick='", jsCall,
                       "'>Analyse at Enrichr website</a>")
enrichrTable <- tibble(Sample              = names(topMarkers),
                       "# of posMarkers"   = lengths(topMarkers),
                       "Enrichr link"      = enrichrCalls)
kable(enrichrTable, format = "html", escape = FALSE,
      caption = paste0("GeneSet enrichment")) |>
  kable_styling("striped", full_width = F, position = "left")
```

# Input Dataset

```{r input-dataset}
#| echo: false
ezInteractiveTableRmd(values = dataset, digits = 4)
```

```{r save-rds}
#| include: false
saveRDS(rawData, "rawData_final.rds")
```

# Session Info

```{r session-info}
sessionInfo()
```

:::

```{r agent-metadata}
#| include: false
.write_metadata(
  `@context`  = "https://fgcz.ch/report-schema/v1",
  reportType  = "count-qc",
  omicsType   = "rnaseq",
  organism    = if (!is.null(param$refBuild)) param$refBuild else NA,
  nSamples    = ncol(rawData),
  nFeatures   = nrow(rawData),
  normMethod  = param$normMethod,
  files       = list(
    countFile  = countFile,
    signalFile = signalFile,
    plots      = "plots/"
  )
)
```
