---
params:
  reportTitle: "SUSHI Report"
title: "`r params$reportTitle`"
metadata-files: ["_fgcz-report.yml"]
---

```{r setup}
#| include: false
library(ezRun)
library(SummarizedExperiment)
library(DT)
library(kableExtra)
library(plotly)
library(htmlwidgets)
library(tidyverse)
library(readxl)
library(writexl)
library(BiocParallel)
library(DOSE)
library(clusterProfiler)
library(jsonlite)

## ── Data loading (qs2 with rds fallback, checks data/ first) ──
.fgcz_load <- function(name, nthreads = 4L) {
  paths <- c(
    file.path("data", paste0(name, ".qs2")),
    paste0(name, ".qs2"),
    file.path("data", paste0(name, ".rds")),
    paste0(name, ".rds")
  )
  for (p in paths) {
    if (file.exists(p)) {
      if (grepl("\\.qs2$", p)) return(qs2::qs_read(p, nthreads = nthreads))
      return(readRDS(p))
    }
  }
  stop("No .qs2 or .rds file found for: ", name)
}

output <- .fgcz_load("output")
se     <- .fgcz_load("deResult")
param  <- .fgcz_load("param")

if (isTRUE(param$runGO)) {
  enrichInput      <- metadata(se)$enrichInput
  enrichResult     <- metadata(se)$enrichResult
  enrichResultGSEA <- metadata(se)$enrichResultGSEA
}

debug <- FALSE

BPPARAM <- BiocParallel::MulticoreParam(workers = param$cores)
register(BPPARAM)

## ── DT helper ─────────────────────────────────────────────────
.dt <- function(df, caption = NULL, pageLength = 10L,
                filter = "none", rownames = FALSE, escape = TRUE, ...) {
  DT::datatable(
    data = df, caption = caption, filter = filter,
    rownames = rownames, escape = escape,
    class = "compact stripe",
    options = list(
      pageLength = pageLength,
      scrollX = TRUE,
      dom = "tip",
      columnDefs = list(
        list(targets = "_all", className = "dt-left")
      )
    ),
    ...
  ) |> DT::formatStyle(columns = seq_len(ncol(df)),
                        `text-align` = "left",
                        `white-space` = "normal",
                        `word-wrap` = "break-word",
                        `max-width` = "300px")
}

## Emit a DT widget as raw HTML inside results='asis' chunks
.dt_asis <- function(...) {
  widget <- .dt(...)
  cat(as.character(htmltools::as.tags(widget)))
}

## ── Output directories ────────────────────────────────────────
for (d in c("plots", "excel", "html")) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}

## ── Plot saving (PNG + PDF to plots/) ─────────────────────────
.save_plot <- function(plot, name, width = 6, height = 5, dpi = 300) {
  if (!dir.exists("plots")) dir.create("plots", recursive = TRUE)
  png_path <- file.path("plots", paste0(name, ".png"))
  pdf_path <- file.path("plots", paste0(name, ".pdf"))
  ggplot2::ggsave(png_path, plot = plot, width = width, height = height,
                  dpi = dpi, bg = "white")
  ggplot2::ggsave(pdf_path, plot = plot, width = width, height = height,
                  device = "pdf", bg = "white")
  invisible(png_path)
}

## ── Agent metadata writer ─────────────────────────────────────
.write_metadata <- function(...) {
  args <- list(...)
  args$generatedAt <- format(Sys.time(), "%Y-%m-%dT%H:%M:%S%z")
  args$softwareVersions <- list(
    R = paste0(R.version$major, ".", R.version$minor),
    quarto = tryCatch(system("quarto --version", intern = TRUE)[1],
                      error = function(e) "unknown"))
  jsonlite::write_json(args, path = "report-metadata.json",
                       pretty = TRUE, auto_unbox = TRUE, null = "null")
}
```

```{r prepare-data}
#| include: false
param   <- metadata(se)$param
seqAnno <- data.frame(rowData(se), row.names = rownames(se), check.names = FALSE)
dataset <- data.frame(colData(se), check.names = FALSE)

if (length(unique(dataset$featureLevel)) == 1L &&
    unique(dataset$featureLevel) == "smRNA") {
  param$runGO <- FALSE
}

design <- ezDesignFromDataset(dataset, param)
hasMassiveFeats <- nrow(seqAnno) >= 200e3
```

::: {.panel-tabset}

# Settings

`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`

::: {.callout-note collapse="true"}
## Analysis Parameters

```{r settings-table}
## Build a clean row-per-parameter table
raw <- makeCountResultSummary(param, se)
settings_df <- data.frame(
  Parameter = names(raw),
  Value = as.character(raw),
  row.names = NULL
)
.dt(settings_df, caption = "Analysis settings", filter = "none", pageLength = 30)
```
:::

# Results Summary

::: {.callout-tip collapse="true"}
## Overview

```{r summary}
settings <- data.frame(
  Metric = character(), Value = character(),
  stringsAsFactors = FALSE)
settings <- rbind(settings, data.frame(Metric = "Reference", Value = param$refBuild))
settings <- rbind(settings, data.frame(Metric = "Number of features", Value = as.character(nrow(se))))
if (!is.null(rowData(se)$isPresentProbe)) {
  settings <- rbind(settings, data.frame(
    Metric = "Features above threshold",
    Value = as.character(sum(rowData(se)$isPresentProbe))))
}
.dt(settings, caption = "Result summary", filter = "none")
```
:::

### Number of significants (log2 fold-change)

```{r sig-counts}
## Build significance table in log2 space with linear FC values
sc <- as.data.frame(makeSignificantCounts(se))

## Add a column showing the linear fold-change equivalent for each row name
## Row names are typically thresholds like "0.5", "1", "2" etc.
if (all(grepl("^[0-9.]+$", rownames(sc)))) {
  log2_thresh <- as.numeric(rownames(sc))
  sc <- cbind(
    data.frame(
      `log2 FC threshold` = rownames(sc),
      `linear FC` = paste0(round(2^log2_thresh, 2), "x"),
      check.names = FALSE),
    sc)
  rownames(sc) <- NULL
}

.dt(sc, caption = "Significant feature counts by p-value and log2 fold-change")
```

```{r result-file}
#| include: false
resultFile <- makeResultFile(param, se)

## Move result files into tidy directories
if (file.exists(resultFile$resultFile)) {
  new_xlsx <- file.path("excel", basename(resultFile$resultFile))
  file.rename(resultFile$resultFile, new_xlsx)
  resultFile$resultFile <- new_xlsx
}
if (file.exists(resultFile$resultHtml)) {
  new_html <- file.path("html", basename(resultFile$resultHtml))
  file.rename(resultFile$resultHtml, new_html)
  resultFile$resultHtml <- new_html
}

ezWrite.table(
  as.data.frame(colData(se))[, "sf", drop = FALSE],
  file = file.path("excel", "scalingfactors.txt"), head = "Name"
)
liveReportLink <- output$getColumn("Live Report")
```

Full result table: [`r resultFile$resultFile`](`r resultFile$resultFile`)

[Live Report and Visualisations](`r liveReportLink`){target="_blank"}

### All genes

```{r all-genes-table}
rd <- as.data.frame(rowData(se))
gene_table <- data.frame(
  gene_id    = rownames(rd),
  gene_name  = rd$gene_name,
  type       = rd$type,
  description = rd$description,
  log2FC     = round(rd$log2Ratio, 3),
  linear_FC  = round(2^rd$log2Ratio, 3),
  pValue     = rd$pValue,
  FDR        = rd$fdr,
  tested     = ifelse(rd$usedInTest, "yes", "no"),
  present    = if (!is.null(rd$isPresentProbe)) ifelse(rd$isPresentProbe, "yes", "no") else NA,
  row.names  = NULL,
  stringsAsFactors = FALSE
)
## Sort by p-value
gene_table <- gene_table[order(gene_table$pValue), ]

DT::datatable(
  gene_table,
  caption = "All genes — filter and search",
  filter = "top",
  rownames = FALSE,
  class = "compact stripe",
  options = list(
    pageLength = 15,
    scrollX = TRUE,
    dom = "ftip",
    order = list(list(6, "asc")),
    columnDefs = list(
      list(targets = "_all", className = "dt-left")
    )
  )
) |>
  DT::formatStyle(columns = seq_len(ncol(gene_table)),
                   `text-align` = "left",
                   `white-space` = "normal",
                   `word-wrap` = "break-word",
                   `max-width` = "300px") |>
  DT::formatSignif(c("pValue", "FDR"), digits = 3)
```

# Significant Genes

```{r scatter-data}
#| include: false
scatterData <- makeTestScatterData(param, se)
logSignal   <- scatterData$logSignal
groupMeans  <- scatterData$groupMeans
types       <- scatterData$types
isTwoGroup  <- ncol(groupMeans) == 2 & !is.null(param$sampleGroup) &
  !is.null(param$refGroup)
```

::: {.callout-note collapse="true"}
## Thresholds

```{r sig-settings}
thresh_df <- data.frame(Metric = character(), Value = character(),
                        stringsAsFactors = FALSE)
if (!is.null(param$pValueHighlightThresh))
  thresh_df <- rbind(thresh_df, data.frame(
    Metric = "P-value threshold", Value = paste("p <=", param$pValueHighlightThresh)))
if (!is.null(param$log2RatioHighlightThresh))
  thresh_df <- rbind(thresh_df, data.frame(
    Metric = "Log2 ratio threshold", Value = paste(">=", param$log2RatioHighlightThresh)))
thresh_df <- rbind(thresh_df, data.frame(
  Metric = "Significant genes", Value = as.character(sum(types$Significants))))
.dt(thresh_df, filter = "none")
```
:::

Subsequent plots highlight significant genes in [blue]{style="color:#A6CEE3FF"}.

[Interactive table of significant genes](`r resultFile$resultHtml`)

::: {.panel-tabset}

## Between-group Comparison

```{r scatter-plot}
#| eval: !expr "isTRUE(isTwoGroup) && !debug"
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Scatter plot comparing average expression between sample and reference groups. Significant genes highlighted in blue."
sampleValues <- 2^groupMeans[, param$sampleGroup]
refValues    <- 2^groupMeans[, param$refGroup]

p_scatter <- ezXYScatter.2(
  xVec = refValues, yVec = sampleValues,
  isPresent = rowData(se)$usedInTest,
  types = types, names = rowData(se)$gene_name,
  xlab = param$refGroup, ylab = param$sampleGroup,
  main = "Comparison of average expression",
  mode = "ggplot2"
)
print(p_scatter)
.save_plot(p_scatter, paste0(param$comparison, "-scatter"))

if (!isTRUE(hasMassiveFeats)) {
  p_scatter_ly <- ezXYScatter.2(
    xVec = refValues, yVec = sampleValues,
    isPresent = rowData(se)$usedInTest,
    types = types, names = rowData(se)$gene_name,
    xlab = param$refGroup, ylab = param$sampleGroup,
    main = "Comparison of average expression",
    mode = "plotly"
  )
  scatterHtml <- file.path("html", paste0(param$comparison, "-scatter.html"))
  DT::saveWidget(as_widget(p_scatter_ly), scatterHtml)
}
```

```{r scatter-link}
#| echo: false
#| results: asis
#| eval: !expr "isTRUE(isTwoGroup) && isFALSE(hasMassiveFeats) && !debug"
cat(paste0("[Interactive comparison plot](", scatterHtml, ")"), "\n")
```

## Volcano (p-value)

```{r volcano-pvalue}
#| eval: !expr "isTRUE(isTwoGroup) && !debug"
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Volcano plot showing log2 fold-change vs -log10(p-value). Significant genes highlighted in blue."
p_volcano <- ezVolcano(
  log2Ratio = rowData(se)$log2Ratio,
  pValue = rowData(se)$pValue, yType = "p-value",
  isPresent = rowData(se)$usedInTest, types = types,
  names = rowData(se)$gene_name,
  main = param$comparison,
  mode = "ggplot2"
)
print(p_volcano)
.save_plot(p_volcano, paste0(param$comparison, "-volcano-pvalue"))

if (isFALSE(hasMassiveFeats)) {
  p_volcano_ly <- ezVolcano(
    log2Ratio = rowData(se)$log2Ratio,
    pValue = rowData(se)$pValue, yType = "p-value",
    isPresent = rowData(se)$usedInTest, types = types,
    names = rowData(se)$gene_name,
    main = param$comparison,
    mode = "plotly"
  )
  volcanoHtml <- file.path("html", paste0(param$comparison, "-volcano.html"))
  DT::saveWidget(as_widget(p_volcano_ly), volcanoHtml)
}
```

```{r volcano-link}
#| echo: false
#| results: asis
#| eval: !expr "isTRUE(isTwoGroup) && isFALSE(hasMassiveFeats) && !debug"
cat(paste0("[Interactive volcano plot](", volcanoHtml, ")"), "\n")
```

## Volcano (FDR)

```{r volcano-fdr}
#| eval: !expr "isTRUE(isTwoGroup) && !debug"
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Volcano plot showing log2 fold-change vs -log10(FDR). Significant genes highlighted in blue."
p_volcano_fdr <- ezVolcano(
  log2Ratio = rowData(se)$log2Ratio,
  pValue = rowData(se)$fdr, yType = "FDR",
  isPresent = rowData(se)$usedInTest, types = types,
  names = rowData(se)$gene_name,
  main = param$comparison,
  mode = "ggplot2"
)
print(p_volcano_fdr)
.save_plot(p_volcano_fdr, paste0(param$comparison, "-volcano-FDR"))
```

## P-value Histogram

```{r pvalue-hist}
#| eval: !expr "!debug"
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "Distribution of p-values for tested (blue) and absent (orange) features. Horizontal line shows the expected uniform density."
myBreaks <- seq(0, 1, by = 0.002)
histUsed <- hist(rowData(se)$pValue[rowData(se)$usedInTest],
                 breaks = myBreaks, plot = FALSE)
histAbs  <- hist(rowData(se)$pValue[!rowData(se)$usedInTest],
                 breaks = myBreaks, plot = FALSE)
xx <- rbind(used = histUsed$counts, absent = histAbs$counts)
xx <- shrinkToRange(xx, c(0, max(xx["used", ])))
barplot(xx,
        space = 0, border = NA, col = c("blue", "darkorange"),
        xlab = "p-value", ylab = "counts",
        ylim = c(0, max(xx["used", ])),
        main = "p-value histogram")
abline(h = sum(rowData(se)$usedInTest) / ncol(xx))
at <- c(0.01, 0.1, 0.25, 0.5, 0.75, 1)
axis(1, at = at * ncol(xx), labels = at)
legend("top", c("used", "not expressed"),
       col = c("blue", "darkorange"), pch = 20, cex = 1)
```

## Intra-group

```{r intra-group}
#| echo: false
#| results: asis
#| eval: !expr "!debug"
theRange <- 2^(range(logSignal, na.rm = TRUE))
x <- logSignal
if (ncol(x) <= 100) {
  if (!ezIsSpecified(param$grouping2)) {
    for (group in unique(c(param$refGroup, colnames(groupMeans)))) {
      idx <- which(group == param$grouping)
      if (length(idx) > 1) {
        cat("\n")
        cat(paste("#### Intra-group Comparison:", group), "\n")
        pngName <- file.path("plots", paste0(group, "-scatter.png"))
        xlab <- paste("Avg of", group)
        refValues <- groupMeans[, group]
        plotCmd <- expression({
          ezScatter(
            x = 2^refValues, y = 2^x[, idx, drop = FALSE],
            isPresent = assays(se)$isPresent[, idx, drop = FALSE],
            types = types, lim = theRange, xlab = xlab)
        })
        includeHtml <- ezImageFileLink(plotCmd, file = pngName,
          width = min(ncol(as.matrix(x[, idx, drop = FALSE])), 6) * 400,
          height = ceiling(ncol(as.matrix(x[, idx, drop = FALSE])) / 6) * 400)
        cat(includeHtml)

        if (ncol(groupMeans) == 2) {
          otherGroup <- setdiff(colnames(groupMeans), group)
          pngName <- file.path("plots", paste0(group, "-over-", otherGroup, "-scatter.png"))
          xlab <- paste("Avg of", otherGroup)
          refValues <- groupMeans[, otherGroup]
          plotCmd <- expression({
            ezScatter(
              x = 2^refValues, y = 2^x[, idx, drop = FALSE],
              isPresent = assays(se)$isPresent[, idx, drop = FALSE],
              types = types, lim = theRange, xlab = xlab)
          })
          includeHtml <- ezImageFileLink(plotCmd, file = pngName,
            width = min(ncol(as.matrix(x[, idx, drop = FALSE])), 6) * 400,
            height = ceiling(ncol(as.matrix(x[, idx, drop = FALSE])) / 6) * 400)
          cat(includeHtml)
        }
      }
      cat("\n")
    }
  } else {
    cat(paste("Pairs:", param$sampleGroup, "over", param$refGroup), "\n")
    use <- param$grouping %in% c(param$sampleGroup, param$refGroup)
    if (all(table(param$grouping2[use], param$grouping[use]) == 1)) {
      groups <- paste(param$grouping, param$grouping2, sep = "--")
      sampleGroups <- sort(unique(groups[param$grouping == param$sampleGroup]))
      refGroups <- sort(unique(groups[param$grouping == param$refGroup]))
      avgValues <- averageAcrossColumns(x[, use], groups[use], mean)
      avgPresent <- averageAcrossColumns(x[, use], groups[use], function(x) mean(x) > 0.5)
      sampleValues <- avgValues[, sampleGroups, drop = FALSE]
      refValues <- avgValues[, refGroups, drop = FALSE]
      samplePresent <- avgPresent[, sampleGroups, drop = FALSE]
      refPresent <- avgPresent[, refGroups, drop = FALSE]
      pngName <- file.path("plots", paste0(param$sampleGroup, "-over-", param$refGroup, "-pairs.png"))
      plotCmd <- expression({
        ezScatter(x = 2^refValues, y = 2^sampleValues,
                  isPresent = samplePresent | refPresent,
                  types = types, lim = theRange, xlab = colnames(refValues))
      })
      includeHtml <- ezImageFileLink(plotCmd, file = pngName,
        width = min(ncol(as.matrix(sampleValues)), 6) * 400,
        height = ceiling(ncol(as.matrix(sampleValues)) / 6) * 400)
      cat(includeHtml)
    }
  }
}
```

:::

# Clustering

```{r cluster-setup}
#| eval: !expr "!debug"
use <- rowData(se)$pValue < param$pValueHighlightThresh &
  abs(rowData(se)$log2Ratio) > param$log2RatioHighlightThresh &
  rowData(se)$usedInTest
use[is.na(use)] <- FALSE
if (sum(use) > param$maxGenesForClustering) {
  use[use] <- rank(rowData(se)$pValue[use], ties.method = "max") <=
    param$maxGenesForClustering
}
```

::: {.callout-note collapse="true"}
## Clustering Thresholds

```{r cluster-thresh}
#| eval: !expr "!debug"
thresh_df <- data.frame(
  Metric = character(), Value = character(), stringsAsFactors = FALSE)
thresh_df <- rbind(thresh_df, data.frame(
  Metric = "Significance threshold", Value = as.character(param$pValueHighlightThresh)))
if (param$log2RatioHighlightThresh > 0) {
  thresh_df <- rbind(thresh_df, data.frame(
    Metric = "log2 Ratio threshold", Value = as.character(param$log2RatioHighlightThresh)))
}
thresh_df <- rbind(thresh_df, data.frame(
  Metric = "Significant features", Value = as.character(sum(use))))
.dt(thresh_df, filter = "none")
```
:::

### Cluster Heatmap

```{r cluster-heatmap}
#| eval: !expr "!debug && !hasMassiveFeats"
#| fig-width: !expr "max(8, 4 + 0.15 * ncol(se))"
#| fig-height: 10
#| fig-retina: 2
#| fig-cap: "Heatmap of significant features clustered by expression pattern. Rows are centred by gene; columns ordered by group."
logSignal <- log2(assays(se)$xNorm + param$backgroundExpression)
if (sum(use) > param$minGenesForClustering) {
  xCentered <- logSignal[use, ]
  if (param$onlyCompGroupsHeatmap) {
    columnsToSubset <- param$grouping %in% c(param$refGroup, param$sampleGroup)
    param$grouping <- param$grouping[columnsToSubset]
    xCentered <- xCentered[, columnsToSubset]
    design <- design[columnsToSubset, , drop = FALSE]
  }
  if (!is.null(param$useRefGroupAsBaseline) && param$useRefGroupAsBaseline) {
    xCentered <- sweep(xCentered, MARGIN = 1,
                       rowMeans(xCentered[, param$grouping == param$refGroup]),
                       FUN = "-")
  } else {
    xCentered <- sweep(xCentered, MARGIN = 1, rowMeans(xCentered), FUN = "-")
  }
  xCentered <- xCentered[, order(param$grouping)]
  sampleColors <- getSampleColors(param$grouping)[order(param$grouping)]

  nClusters <- 6
  clusterColors <- hcl.colors(nClusters, palette = "Spectral")
  clusterResult <- clusterResults(xCentered, nClusters = nClusters,
                                  clusterColors = clusterColors)

  design <- design[, c(param$groupingName,
                        setdiff(colnames(design), param$groupingName)),
                   drop = FALSE]

  clusterResult <- clusterPheatmap(xCentered, design, param, clusterColors,
    lim = c(-param$logColorRange, param$logColorRange),
    doClusterColumns = FALSE, sampleColors = sampleColors)
  plot(clusterResult$pheatmap$gtable)

  if (doGo(param, seqAnno)) {
    clusterResult <- goClusterResults(xCentered, param, clusterResult,
      seqAnno = seqAnno,
      universeProbeIds = rownames(seqAnno)[rowData(se)$isPresentProbe])
  }

  ## Append cluster info to result file
  resultLoaded <- read_excel(resultFile$resultFile)
  resultLoaded$Cluster <- clusterResult$clusterNumbers[resultLoaded[[1]]]
  write_xlsx(resultLoaded, path = resultFile$resultFile)

  if (!is.null(clusterResult$GO)) {
    goTables <- goClusterTableRmd(param, clusterResult, seqAnno)
    if (doEnrichr(param)) {
      goAndEnrichr <- cbind(goTables$linkTable, goTables$enrichrTable)
    } else {
      goAndEnrichr <- goTables$linkTable
    }
    bgColors <- gsub("FF$", "", clusterResult$clusterColors)
    rownames(goAndEnrichr) <- paste0(
      '<font color="', bgColors, '">Cluster ',
      rownames(goAndEnrichr), "</font>")
    .dt(goAndEnrichr, caption = "GO categories of feature clusters",
        rownames = TRUE, escape = FALSE)
  }
}
```

```{r render-go-cluster}
#| include: false
#| eval: !expr "!debug && !hasMassiveFeats"
if (sum(use) > param$minGenesForClustering) {
  if (!is.null(clusterResult$GO)) {
    file <- file.path(
      system.file("templates", package = "ezRun"),
      "twoGroups_goClusterTable.Rmd")
    file.copy(from = file, to = basename(file), overwrite = TRUE)
    rmarkdown::render(
      input = basename(file), envir = new.env(),
      output_dir = "html", output_file = "goClusterTable.html",
      quiet = TRUE)
  }
}
```

```{r go-cluster-link}
#| echo: false
#| results: asis
#| eval: !expr "!debug && !hasMassiveFeats"
if (file.exists("html/goClusterTable.html")) {
  cat(paste0("[GO cluster tables](html/goClusterTable.html)"), "\n")
}
```

# Enrichment

::: {.panel-tabset}

## Enrichr

::: {.callout-note collapse="true"}
## About Enrichr
Enrichr is a web server collecting many databases that can be interrogated using gene lists. It takes as input differentially expressed genes. **It is recommended to search various databases such as cell type, transcription factors, mutations, diseases.** [Visit the Enrichr Bioconductor page](https://bioconductor.org/packages/release/bioc/html/goseq.html)
:::

```{r enrichr}
#| eval: !expr isTRUE(param$runGO)
if (doEnrichr(param)) {
  selectionNames <- names(enrichInput$selections)

  enrichrLinks <- ezMatrix("", rows = selectionNames,
    cols = c("Number of Genes", "External", "Precomputed", "CutOffs"))
  maxResultsPerLibrary <- 5

  for (mySel in selectionNames) {
    genesToUse <- enrichInput$selections[[mySel]]
    genesToUse <- rowData(se)[genesToUse, "gene_name"]
    enrichrLinks[mySel, "Number of Genes"] <- length(genesToUse)
    enrichrLinks[mySel, "CutOffs"] <- paste("p <=", param$pValThreshGO)
    if (mySel == "upGenes")
      enrichrLinks[mySel, "CutOffs"] <- paste(enrichrLinks[mySel, "CutOffs"],
        ", log2 ratio >", param$log2RatioThreshGO)
    if (mySel == "downGenes")
      enrichrLinks[mySel, "CutOffs"] <- paste(enrichrLinks[mySel, "CutOffs"],
        ", log2 ratio <", -param$log2RatioThreshGO)
    jsCall <- paste0('enrich({list: "',
      paste(genesToUse, collapse = "\\n"), '", popup: true});')
    enrichrLinks[mySel, "External"] <- paste0(
      "<a href='javascript:void(0)' onClick='", jsCall,
      "'>Analyse at Enrichr website</a>")
    resMerged <- NA
    if (!is.null(genesToUse) && length(genesToUse) > 3 &&
        param$doPrecomputeEnrichr) {
      resList <- tryCatch(
        runEnrichr(genesToUse, maxResult = maxResultsPerLibrary),
        error = function(e) { message("enrichr failed"); NULL })
      resList <- lapply(names(resList), function(nm) {
        cbind("Gene-set library" = nm, resList[[nm]][, c(2:5, 7:10)])
      })
      if (length(resList) > 0) {
        resMerged <- do.call("rbind", resList)
        resMerged <- resMerged[order(-resMerged[, 5]), ]
        resMerged[, c(3, 6:8)] <- apply(resMerged[, c(3, 6:8)], 2, sprintf, fmt = "%0.2e")
        resMerged[, c(4, 5)] <- apply(resMerged[, c(4, 5)], 2, sprintf, fmt = "%0.3f")
        enrichrTablePath <- file.path("html", paste0("enrichrTable_", mySel, ".html"))
        ezInteractiveTable(resMerged, title = paste("Enrichr report for", mySel)) |>
          DT::saveWidget(enrichrTablePath)
        enrichrLinks[mySel, "Precomputed"] <- ezLink(
          link = enrichrTablePath, label = "Report", target = "_blank")
      } else {
        enrichrLinks[mySel, "Precomputed"] <- "No significant results"
      }
    } else {
      enrichrLinks[mySel, "Precomputed"] <- "Not run"
    }
  }
  .dt(as.data.frame(enrichrLinks), caption = "Enrichr results",
      rownames = TRUE, escape = FALSE)
}
if (!doEnrichr(param)) {
  cat("\n", getOrganism(param$ezRef), "is not supported by Enrichr.\n")
}
```

## ORA

::: {.callout-note collapse="true"}
## About ORA
The Over Representation Analysis (ORA), also known as the hypergeometric test, estimates whether selected genes are enriched for specific Gene Ontology categories. **Recommended when the difference between groups is large (500+ genes above thresholds).** Uses [clusterProfiler](https://bioconductor.org/packages/release/bioc/html/clusterProfiler.html).

- Gene Selection: `r if(isTRUE(param$runGO)) paste0("p <= ", param$pValThreshGO, ", Up: log2 ratio >", param$log2RatioThreshGO, " / Down: log2 ratio < ", -param$log2RatioThreshGO)`
- Candidate Terms: `r if(isTRUE(param$runGO)) paste("fdr <=", param$fdrThreshORA)`
:::

```{r ora-prepare}
#| include: false
#| eval: !expr "!debug && isTRUE(param$runGO)"
enrichResultPlots <- list()
for (onto in names(enrichResult)) {
  enrichResultPlots[[onto]] <- list()
  for (sig in names(enrichResult[[onto]])) {
    if (is.na(enrichResult[[onto]][[sig]][1][[1]])) next
    erp <- list()
    er <- enrichResult[[onto]][[sig]]
    er@result$Label <- ""
    for (i in seq_along(er@result$Description)) {
      desc <- er@result$Description[i]
      if (!is.na(desc) && nchar(desc) < 35) {
        syn <- desc
      } else {
        require(AnnotationDbi); require(GO.db)
        syn <- Synonym(GOTERM[[er@result$ID[i]]])
        syn <- grep("GO:", syn, invert = TRUE, value = TRUE)
        syn <- head(syn[order(nchar(syn))], 1)
        if (length(syn) == 0 || is.na(syn)) syn <- substr(desc, 1, 35)
      }
      er@result$Label[i] <- paste0(er@result$ID[i], "\n", syn)
    }
    er@result$geneID <- er@result$geneName
    erp[["df"]] <- as.data.frame(
      er@result[, c("ID", "Description", "GeneRatio", "BgRatio",
                     "pvalue", "p.adjust", "geneName")])
    erp[["df"]] <- erp[["df"]][erp[["df"]]$p.adjust <= param$fdrThreshORA, ]
    erp[["barplot"]] <- enrichplot:::barplot.enrichResult(
      er, showCategory = 20, title = paste(onto, sig))
    erp[["barplot"]] <- erp[["barplot"]] +
      scale_y_discrete(labels = rev(er@result$Label[
        match(erp$barplot$plot_env$df$ID, er@result$ID)])) +
      theme(axis.text.y = element_text(vjust = -0.01, size = 8))
    sigGeneIds <- enrichInput$selections[[sig]]
    log2Ratio <- enrichInput$log2Ratio[sigGeneIds]
    names(log2Ratio) <- enrichInput$seqAnno[sigGeneIds, "gene_name"]
    log2Ratio <- sort(log2Ratio, decreasing = TRUE)
    erp[["cnet"]] <- enrichplot::cnetplot(
      x = er, color.params = list(foldChange = log2Ratio),
      cex.params = list(category_label = 0.7, gene_label = 0.6))
    enrichResultPlots[[onto]][[sig]] <- erp
  }
}
## Combined ORA xlsx
oraFile <- file.path("excel", paste0(param$comparison, "--ORA_results.xlsx"))
oraList <- list()
for (onto in names(enrichResult)) {
  for (sig in names(enrichResult[[onto]])) {
    if (is.na(enrichResult[[onto]][[sig]][1][[1]])) next
    oraList[[paste0(onto, "_", sig)]] <- as.data.frame(enrichResult[[onto]][[sig]]@result)
  }
}
if (length(oraList) >= 1) openxlsx::write.xlsx(oraList, file = oraFile)
```

```{r ora-tabs}
#| echo: false
#| results: asis
#| eval: !expr isTRUE(param$runGO)
#| fig-height: 6
#| fig-width: 6

## ── Overview tab ──
cat("\n\n::: {.panel-tabset}\n\n")
cat("### Overview\n\n")

ora.overview <- ezFrame("Number of Significant Terms" = numeric())
for (goDomain in names(enrichResult)) {
  for (resName in names(enrichResult[[goDomain]])) {
    testName <- paste0(goDomain, ": ", resName)
    ora.overview[testName, "Number of Significant Terms"] <-
      nrow(enrichResult[[goDomain]][[resName]]@result[
        enrichResult[[goDomain]][[resName]]@result$p.adjust <= param$fdrThreshORA, ])
  }
}
.dt_asis(format(ora.overview, digits = 3, scientific = -2),
    caption = "Hypergeometric Over-representation Test",
    rownames = TRUE)
cat("\n\n")
if (file.exists(oraFile))
  cat(paste0("[Download ORA results (Excel)](", oraFile, ")"), "\n\n")

## ── Per-ontology tabs ──
for (onto in names(enrichResult)) {
  cat("###", onto, "\n\n")
  cat("::: {.panel-tabset}\n\n")
  for (sig in names(enrichResult[[onto]])) {
    cat("####", sig, "\n\n")
    if (is.na(enrichResult[[onto]][[sig]][1][[1]])) {
      cat("No significant GO terms detected.\n\n")
      next
    }
    erp <- enrichResultPlots[[onto]][[sig]]

    cat("::: {.panel-tabset}\n\n")
    cat("##### Barplot\n\n")
    print(erp$barplot)
    cat("\n\n##### Network\n\n")
    print(erp$cnet)
    cat("\n\n##### Table\n\n")
    .dt_asis(erp$df, caption = paste(onto, sig, "— significant terms"))
    cat("\n\n:::\n\n")
  }
  cat(":::\n\n")
}
cat(":::\n\n")
```

## GSEA

::: {.callout-note collapse="true"}
## About GSEA
Gene Set Enrichment Analysis (GSEA) calculates an enrichment score by screening all genes and their fold-changes. **Recommended when the difference between groups is small or when combining results.** Uses [clusterProfiler](https://bioconductor.org/packages/release/bioc/html/clusterProfiler.html).

- Candidate Terms: `r if(isTRUE(param$runGO)) paste("fdr <=", param$fdrThreshGSEA)`
- Gene Ranking: all present genes sorted by log2 ratio
:::

```{r gsea-prepare}
#| include: false
#| eval: !expr "!debug && isTRUE(param$runGO)"
enrichResultGSEAPlots <- list()
for (onto in names(enrichResultGSEA)) {
  enrichResultGSEAPlots[[onto]] <- list()
  if (is.na(enrichResultGSEA[[onto]][1][[1]])) next
  erp <- list()
  er <- enrichResultGSEA[[onto]]
  er@result$Label <- ""
  for (i in seq_along(er@result$Description)) {
    desc <- er@result$Description[i]
    if (!is.na(desc) && nchar(desc) < 35) {
      syn <- desc
    } else {
      require(AnnotationDbi); require(GO.db)
      syn <- Synonym(GOTERM[[er@result$ID[i]]])
      syn <- grep("GO:", syn, invert = TRUE, value = TRUE)
      syn <- head(syn[order(nchar(syn))], 1)
      if (length(syn) == 0 || is.na(syn)) syn <- substr(desc, 1, 35)
    }
    er@result$Label[i] <- paste0(er@result$ID[i], "\n", syn)
  }
  erp[["df"]] <- as.data.frame(er@result[, c(
    "ID", "Description", "setSize", "enrichmentScore", "NES",
    "pvalue", "p.adjust", "geneName")])
  erp[["ridgeplot"]] <- enrichplot::ridgeplot(x = er, showCategory = 20)
  erp[["ridgeplot"]] <- erp[["ridgeplot"]] +
    scale_y_discrete(labels = er@result$Label[
      match(levels(as.factor(erp$ridgeplot$plot_env$gs2val.df$category)),
            er@result$Description)]) +
    theme(axis.text.y = element_text(vjust = -0.01, size = 8))
  erp[["gseaplot"]] <- enrichplot::gseaplot2(x = er, geneSetID = 1)

  ce_list <- NA
  for (i in seq_along(er@result$core_enrichment)) {
    core_enrich <- unlist(strsplit(er@result$core_enrichment[i], split = "/"))
    ce_list <- c(ce_list, core_enrich)
    core_enrich_symbol <- enrichInput$seqAnno$gene_name[
      enrichInput$seqAnno$gene_id %in% core_enrich]
    er@result$core_enrichment[i] <- paste(core_enrich_symbol, collapse = "/")
  }
  sigGeneIds <- ce_list[!is.na(ce_list)]
  log2Ratio <- enrichInput$log2Ratio[sigGeneIds]
  names(log2Ratio) <- enrichInput$seqAnno[sigGeneIds, "gene_name"]
  log2Ratio <- sort(log2Ratio, decreasing = TRUE)
  erp[["cnet"]] <- enrichplot::cnetplot(
    x = er, color.params = list(foldChange = log2Ratio),
    cex.params = list(category_label = 0.7, gene_label = 0.6))
  enrichResultGSEAPlots[[onto]] <- erp
}
## Combined GSEA xlsx
gseaFile <- file.path("excel", paste0(param$comparison, "--GSEA_results.xlsx"))
gseaList <- list()
for (onto in names(enrichResultGSEA)) {
  if (is.na(enrichResultGSEA[[onto]][1][[1]])) next
  gseaList[[onto]] <- as.data.frame(enrichResultGSEA[[onto]]@result)
}
if (length(gseaList) >= 1) openxlsx::write.xlsx(gseaList, file = gseaFile)
```

```{r gsea-tabs}
#| echo: false
#| results: asis
#| eval: !expr isTRUE(param$runGO)
#| fig-height: 6
#| fig-width: 6

## ── Overview tab ──
cat("\n\n::: {.panel-tabset}\n\n")
cat("### Overview\n\n")

gsea.overview <- ezFrame("Number of Significant Pathways" = numeric())
for (i in seq_along(names(enrichResultGSEA))) {
  testName <- names(enrichResultGSEA)[i]
  gsea.overview[testName, "Number of Significant Pathways"] <-
    nrow(enrichResultGSEA[[i]]@result)
}
.dt_asis(format(gsea.overview, digits = 3, scientific = -2),
    caption = "GSEA overview", rownames = TRUE)
cat("\n\n")
if (file.exists(gseaFile))
  cat(paste0("[Download GSEA results (Excel)](", gseaFile, ")"), "\n\n")

## ── Per-ontology tabs ──
for (onto in names(enrichResultGSEA)) {
  cat("###", onto, "\n\n")
  if (is.na(enrichResultGSEA[[onto]][1][[1]])) {
    cat("No significant GO terms detected.\n\n")
    next
  }
  erp <- enrichResultGSEAPlots[[onto]]

  cat("::: {.panel-tabset}\n\n")
  cat("#### Ridge Plot\n\n")
  print(erp$ridgeplot)
  cat("\n\n#### Network\n\n")
  print(erp$cnet)
  cat("\n\n#### Table\n\n")
  .dt_asis(erp$df, caption = paste(onto, "— GSEA results"))
  cat("\n\n:::\n\n")
}
cat(":::\n\n")
```

:::

# Technical Bias

::: {.callout-warning collapse="true"}
## About this test
We define 4 gene sets: high/low GC content (5th/95th percentile) and long/short genes (5th/95th percentile). We test if up- or down-regulated genes are associated with these sets. Significant associations (p < 0.001) may indicate false positives due to technical bias.
:::

```{r bias}
x <- rowData(se)[rowData(se)$usedInTest, ]

if (all(c("gc", "featWidth") %in% colnames(x))) {
  if (!all(is.na(x$gc)) & !all(is.na(x$featWidth))) {
    gcThresh <- quantile(x$gc, c(0.05, 0.95))
    widthThresh <- quantile(x$featWidth, c(0.05, 0.95))
    biasTable <- data.frame(
      "low GC" = x$gc < gcThresh[1],
      "high GC" = x$gc > gcThresh[2],
      "short genes" = x$featWidth < widthThresh[1],
      "long genes" = x$featWidth > widthThresh[2],
      check.names = FALSE)
    isUp <- x$pValue < param$pValueHighlightThresh &
      x$log2Ratio > param$log2RatioHighlightThresh
    isDown <- x$pValue < param$pValueHighlightThresh &
      x$log2Ratio < -param$log2RatioHighlightThresh
    sigTable <- data.frame(
      "Up-regulation" = isUp,
      "Down-regulation" = isDown,
      check.names = FALSE)

    tests <- expand.grid(sig = names(sigTable), bias = names(biasTable))
    testTable <- ezFrame(
      "overlapping/total genes" = character(),
      "odds ratio" = numeric(),
      "p-value" = numeric())
    for (i in seq_len(nrow(tests))) {
      testName <- paste(tests$bias[i], " -- ", tests$sig[i])
      myBias <- biasTable[[tests$bias[i]]]
      mySig <- sigTable[[tests$sig[i]]]
      if (length(table(mySig)) == 2 && length(table(myBias)) == 2) {
        res.fisher <- fisher.test(myBias, mySig, alternative = "greater")
      } else {
        res.fisher <- list(estimate = NA, p.value = NA)
      }
      testTable[testName, "overlapping/total genes"] <-
        paste(sum(myBias & mySig), sum(myBias), sep = "/")
      testTable[testName, "odds ratio"] <- res.fisher$estimate
      testTable[testName, "p-value"] <- res.fisher$p.value
    }
    .dt(format(testTable, digits = 3, scientific = -2),
        caption = paste("Association test:",
          sum(isUp), "up-regulated and",
          sum(isDown), "down-regulated genes"),
        rownames = TRUE)
  }
}
```

# Input Dataset

```{r input-dataset}
ezInteractiveTableRmd(values = dataset, digits = 4)
```

# Session Info

```{r session-info}
sessionInfo()
```

:::

```{r agent-metadata}
#| include: false
## Write agent metadata sidecar
topN <- min(20, nrow(se))
rd <- as.data.frame(rowData(se))
rd_sig <- rd[!is.na(rd$fdr) & rd$fdr < param$pValueHighlightThresh, ]
rd_sig <- rd_sig[order(rd_sig$fdr), ]
rd_sig <- head(rd_sig, topN)

## Clean up stray files from ezRun sub-renders
unlink(list.files(pattern = "^enrichr-error-.*\\.rds$"))
unlink(list.files(pattern = "^Cluster-.*\\.html$"))
unlink("twoGroups_goClusterTable.Rmd")

.write_metadata(
  `@context` = "https://fgcz.ch/report-schema/v1",
  reportType = "differential-expression",
  omicsType = "rnaseq",
  organism = if (!is.null(param$refBuild)) param$refBuild else NA,
  comparisons = list(list(
    name = param$comparison,
    sample = param$sampleGroup,
    reference = param$refGroup,
    method = if (!is.null(metadata(se)$method)) metadata(se)$method else "unknown"
  )),
  thresholds = list(
    pValue = param$pValueHighlightThresh,
    log2FC = param$log2RatioHighlightThresh,
    fdr = param$pValueHighlightThresh
  ),
  counts = list(
    totalTested = sum(rd$usedInTest, na.rm = TRUE),
    significantUp = sum(rd$usedInTest & rd$pValue < param$pValueHighlightThresh &
      rd$log2Ratio > param$log2RatioHighlightThresh, na.rm = TRUE),
    significantDown = sum(rd$usedInTest & rd$pValue < param$pValueHighlightThresh &
      rd$log2Ratio < -param$log2RatioHighlightThresh, na.rm = TRUE)
  ),
  topHits = if (nrow(rd_sig) > 0) {
    lapply(seq_len(nrow(rd_sig)), function(i) list(
      id = rownames(rd_sig)[i], name = rd_sig$gene_name[i],
      log2FC = round(rd_sig$log2Ratio[i], 3),
      fdr = signif(rd_sig$fdr[i], 3)))
  } else list(),
  files = list(
    resultTable = resultFile$resultFile,
    resultHtml = resultFile$resultHtml,
    plots = "plots/"
  )
)
```
