---
title: "`r if (exists('reportTitle')) reportTitle else 'FGCZ Proteomics Report'`"
metadata-files: ["_fgcz-report.yml"]
---

```{r setup}
#| include: false
source("_helpers.R")

library(SummarizedExperiment)
library(DT)
library(ggplot2)
library(tidyverse)
library(pheatmap)
library(writexl)
library(jsonlite)

## Load data (qs2 with rds fallback)
se <- fgczLoadData("deResult")
```

```{r prepare-data}
#| include: false
## Extract metadata
bfabric_urls <- metadata(se)$bfabric_urls
contrasts    <- metadata(se)$contrasts
formula_info <- metadata(se)$formula

## Parse contrast information
contrast_names <- rownames(contrasts)

## Build per-contrast result tables from rowData
rd <- as.data.frame(rowData(se))

## Identify contrast column prefixes (e.g. "constrast_OT_vs_Manual")
contrast_prefixes <- paste0("constrast_", contrast_names)

## Sample info
sample_info <- as.data.frame(colData(se))

## Group column
group_col <- grep("^G_|^group|^Group", colnames(sample_info), value = TRUE)[1]
if (is.na(group_col)) group_col <- colnames(sample_info)[3]  # fallback

## Thresholds
fdr_threshold  <- 0.05
diff_threshold <- 1.0
```

::: {.panel-tabset}

# Settings

`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`

::: {.callout-note collapse="true"}
## Analysis Parameters

```{r settings-table}
settings_df <- data.frame(
  Parameter = c("Formula", "FDR threshold", "Difference threshold",
                "Number of proteins", "Number of samples",
                "Number of contrasts"),
  Value = c(
    formula_info$formula[1],
    fdr_threshold,
    diff_threshold,
    nrow(se),
    ncol(se),
    length(contrast_names)
  ),
  row.names = NULL
)
fgczDatatable(settings_df, caption = "Analysis settings", filter = "none")
```
:::

::: {.callout-note collapse="true"}
## Contrasts

```{r contrasts-table}
fgczDatatable(
  data.frame(
    Name = rownames(contrasts),
    Formula = contrasts$contrast,
    row.names = NULL
  ),
  caption = "Contrasts tested",
  filter = "none"
)
```
:::

::: {.callout-tip collapse="true"}
## B-Fabric Links

```{r bfabric-links}
#| results: asis
if (!is.null(bfabric_urls)) {
  for (nm in names(bfabric_urls)) {
    cat(paste0("- [", nm, "](", bfabric_urls[[nm]], "){target='_blank'}"), "\n")
  }
}
```
:::

# Results Summary

```{r results-summary}
summary_rows <- list()
for (cn in contrast_names) {
  prefix <- paste0("constrast_", cn)
  fdr_col  <- paste0(prefix, ".FDR")
  diff_col <- paste0(prefix, ".diff")
  if (all(c(fdr_col, diff_col) %in% colnames(rd))) {
    n_sig   <- sum(rd[[fdr_col]] < fdr_threshold & abs(rd[[diff_col]]) > diff_threshold,
                   na.rm = TRUE)
    n_up    <- sum(rd[[fdr_col]] < fdr_threshold & rd[[diff_col]] > diff_threshold,
                   na.rm = TRUE)
    n_down  <- sum(rd[[fdr_col]] < fdr_threshold & rd[[diff_col]] < -diff_threshold,
                   na.rm = TRUE)
    summary_rows[[cn]] <- data.frame(
      Contrast = cn, Total = nrow(rd),
      Significant = n_sig, Up = n_up, Down = n_down,
      row.names = NULL)
  }
}
summary_df <- do.call(rbind, summary_rows)
fgczDatatable(summary_df, caption = "Significant protein counts per contrast",
              filter = "none")
```

```{r write-results-xlsx}
#| include: false
## Write combined results Excel
result_sheets <- list()
for (cn in contrast_names) {
  prefix <- paste0("constrast_", cn)
  cols <- grep(paste0("^", prefix), colnames(rd), value = TRUE)
  short_cols <- gsub(paste0("^", prefix, "\\."), "", cols)
  df_out <- rd[, cols, drop = FALSE]
  colnames(df_out) <- short_cols
  df_out$protein_Id <- rownames(rd)
  df_out <- df_out[, c("protein_Id", setdiff(colnames(df_out), "protein_Id"))]
  result_sheets[[cn]] <- df_out
}
result_file <- "result--proteomics.xlsx"
write_xlsx(result_sheets, path = result_file)
```

Full result table: [`r result_file`](`r result_file`)

# Quality Control

::: {.panel-tabset}

## Missing Values

::: {.callout-warning collapse="true"}
## About missing values
Absent protein measurements may indicate biological relevance or technical problems. The heatmap below shows present (white) vs absent (black) patterns across samples.
:::

```{r missing-heatmap}
#| fig-width: 8
#| fig-height: 7
mat <- assay(se, "rawData")
na_mat <- ifelse(is.na(mat), 1, 0)
## Only show proteins with at least one NA
has_na <- rowSums(na_mat) > 0
n_with_na <- sum(has_na)

if (n_with_na > 0 && n_with_na < 5000) {
  anno_col <- data.frame(Group = sample_info[[group_col]],
                         row.names = colnames(se))
  pheatmap(na_mat[has_na, , drop = FALSE],
           color = c("white", "black"),
           cluster_rows = TRUE, cluster_cols = TRUE,
           show_rownames = FALSE,
           annotation_col = anno_col,
           main = paste0("Missing value pattern (", n_with_na,
                         " proteins with >= 1 NA)"),
           legend = FALSE)
} else if (n_with_na >= 5000) {
  cat("Too many proteins with missing values to display heatmap:",
      n_with_na, "\n")
} else {
  cat("No missing values detected.\n")
}
```

## Abundance Distribution

```{r density-plots}
#| fig-width: 10
#| fig-height: 5
raw_long <- as.data.frame(assay(se, "rawData")) |>
  pivot_longer(everything(), names_to = "Sample", values_to = "Abundance") |>
  mutate(Type = "Raw")

trans_long <- as.data.frame(assay(se, "transformedData")) |>
  pivot_longer(everything(), names_to = "Sample", values_to = "Abundance") |>
  mutate(Type = "Normalised")

combined <- bind_rows(raw_long, trans_long)

ggplot(combined, aes(x = Abundance, colour = Sample)) +
  geom_density(na.rm = TRUE, linewidth = 0.5) +
  facet_wrap(~Type, scales = "free") +
  theme_minimal() +
  theme(legend.position = if (ncol(se) > 12) "none" else "right") +
  labs(x = "log2 Abundance", y = "Density",
       title = "Protein abundance distribution")
```

## CV Analysis

```{r cv-analysis}
#| fig-width: 7
#| fig-height: 4
mat_norm <- assay(se, "transformedData")
groups <- sample_info[[group_col]]

cv_list <- list()
for (grp in unique(groups)) {
  idx <- which(groups == grp)
  if (length(idx) >= 2) {
    row_means <- rowMeans(mat_norm[, idx], na.rm = TRUE)
    row_sds   <- matrixStats::rowSds(mat_norm[, idx], na.rm = TRUE)
    cv <- (row_sds / abs(row_means)) * 100
    cv_list[[grp]] <- data.frame(Group = grp, CV = cv[is.finite(cv)])
  }
}
## Add "All"
all_means <- rowMeans(mat_norm, na.rm = TRUE)
all_sds   <- matrixStats::rowSds(mat_norm, na.rm = TRUE)
cv_all <- (all_sds / abs(all_means)) * 100
cv_list[["All"]] <- data.frame(Group = "All", CV = cv_all[is.finite(cv_all)])

cv_df <- do.call(rbind, cv_list)

p_cv <- ggplot(cv_df, aes(x = Group, y = CV, fill = Group)) +
  geom_violin(draw_quantiles = 0.5, alpha = 0.7) +
  theme_minimal() +
  labs(x = NULL, y = "CV (%)", title = "Coefficient of Variation") +
  theme(legend.position = "none") +
  coord_cartesian(ylim = c(0, quantile(cv_df$CV, 0.99)))
print(p_cv)
fgczSavePlot(p_cv, "cv-violin", width = 7, height = 4)
```

## PCA

```{r pca}
#| fig-width: 7
#| fig-height: 6
mat_pca <- assay(se, "transformedData")
mat_pca <- mat_pca[complete.cases(mat_pca), ]

if (nrow(mat_pca) >= 10) {
  pca_res <- prcomp(t(mat_pca), center = TRUE, scale. = TRUE)
  pca_df <- data.frame(
    PC1 = pca_res$x[, 1], PC2 = pca_res$x[, 2],
    Sample = colnames(mat_pca),
    Group = sample_info[[group_col]])
  var_expl <- round(summary(pca_res)$importance[2, 1:2] * 100, 1)

  p_pca <- ggplot(pca_df, aes(x = PC1, y = PC2, colour = Group, label = Sample)) +
    geom_point(size = 3) +
    ggrepel::geom_text_repel(size = 3, max.overlaps = 15) +
    theme_minimal() +
    labs(x = paste0("PC1 (", var_expl[1], "%)"),
         y = paste0("PC2 (", var_expl[2], "%)"),
         title = "PCA of normalised abundances") +
    coord_fixed()
  print(p_pca)
  fgczSavePlot(p_pca, "pca", width = 7, height = 6)
}
```

:::

# Differential Expression

::: {.panel-tabset}

## Volcano

```{r volcano-plots}
#| fig-width: 9
#| fig-height: 7
#| results: asis
for (cn in contrast_names) {
  prefix   <- paste0("constrast_", cn)
  diff_col <- paste0(prefix, ".diff")
  fdr_col  <- paste0(prefix, ".FDR")
  name_col <- paste0(prefix, ".gene_name")
  model_col <- paste0(prefix, ".modelName")

  if (!all(c(diff_col, fdr_col) %in% colnames(rd))) next

  plot_df <- data.frame(
    diff = rd[[diff_col]],
    FDR  = rd[[fdr_col]],
    name = if (name_col %in% colnames(rd)) rd[[name_col]] else rownames(rd),
    model = if (model_col %in% colnames(rd)) rd[[model_col]] else "unknown"
  )
  plot_df <- plot_df[!is.na(plot_df$FDR), ]

  p <- ggplot(plot_df, aes(x = diff, y = -log10(FDR), colour = model)) +
    geom_point(alpha = 0.5, size = 1) +
    geom_hline(yintercept = -log10(fdr_threshold), linetype = "dashed",
               colour = "red") +
    geom_vline(xintercept = c(-diff_threshold, diff_threshold),
               linetype = "dashed", colour = "darkgreen") +
    theme_minimal() +
    labs(x = "Difference (log2)", y = "-log10(FDR)",
         title = cn, colour = "Model") +
    scale_colour_manual(values = c("Linear_Model_moderated" = "black",
                                   "Imputed_Mean_moderated" = "orange"))
  print(p)
  fgczSavePlot(p, paste0("volcano-", cn))
  cat("\n\n")
}
```

## Results Table

```{r results-table}
#| results: asis
for (cn in contrast_names) {
  prefix <- paste0("constrast_", cn)
  cols_needed <- c(
    paste0(prefix, c(".protein_Id", ".gene_name", ".description",
                     ".diff", ".FDR", ".modelName")))
  cols_avail <- intersect(cols_needed, colnames(rd))

  if (length(cols_avail) >= 3) {
    df_show <- rd[, cols_avail, drop = FALSE]
    colnames(df_show) <- gsub(paste0("^", prefix, "\\."), "", colnames(df_show))
    if ("diff" %in% colnames(df_show))
      df_show$diff <- round(df_show$diff, 3)
    if ("FDR" %in% colnames(df_show))
      df_show$FDR <- signif(df_show$FDR, 3)

    cat("###", cn, "\n\n")
    print(fgczDatatable(df_show, caption = paste("All proteins —", cn)))
    cat("\n\n")
  }
}
```

:::

# Significant Proteins

```{r significant-proteins}
#| results: asis
for (cn in contrast_names) {
  prefix   <- paste0("constrast_", cn)
  diff_col <- paste0(prefix, ".diff")
  fdr_col  <- paste0(prefix, ".FDR")
  name_col <- paste0(prefix, ".gene_name")
  desc_col <- paste0(prefix, ".description")

  if (!all(c(diff_col, fdr_col) %in% colnames(rd))) next

  is_sig <- !is.na(rd[[fdr_col]]) &
    rd[[fdr_col]] < fdr_threshold &
    abs(rd[[diff_col]]) > diff_threshold

  sig_df <- data.frame(
    protein_Id = rownames(rd)[is_sig],
    gene_name  = if (name_col %in% colnames(rd)) rd[is_sig, name_col] else NA,
    description = if (desc_col %in% colnames(rd)) rd[is_sig, desc_col] else NA,
    diff = round(rd[is_sig, diff_col], 3),
    FDR  = signif(rd[is_sig, fdr_col], 3),
    row.names = NULL
  )
  sig_df <- sig_df[order(sig_df$FDR), ]

  cat("###", cn, paste0("(", nrow(sig_df), " significant)"), "\n\n")
  print(fgczDatatable(sig_df, caption = paste("Significant proteins —", cn)))
  cat("\n\n")
}
```

### Heatmap of Significant Proteins

```{r sig-heatmap}
#| fig-width: 8
#| fig-height: 8
## Combine significant proteins across all contrasts
all_sig_ids <- c()
for (cn in contrast_names) {
  prefix   <- paste0("constrast_", cn)
  diff_col <- paste0(prefix, ".diff")
  fdr_col  <- paste0(prefix, ".FDR")
  if (all(c(diff_col, fdr_col) %in% colnames(rd))) {
    is_sig <- !is.na(rd[[fdr_col]]) &
      rd[[fdr_col]] < fdr_threshold &
      abs(rd[[diff_col]]) > diff_threshold
    all_sig_ids <- union(all_sig_ids, rownames(rd)[is_sig])
  }
}

mat_sig <- assay(se, "transformedData")[all_sig_ids, , drop = FALSE]
mat_sig <- mat_sig[complete.cases(mat_sig), , drop = FALSE]

if (nrow(mat_sig) >= 2 && nrow(mat_sig) <= 500) {
  ## Row-centre
  mat_sig_scaled <- t(scale(t(mat_sig)))
  anno_col <- data.frame(Group = sample_info[[group_col]],
                         row.names = colnames(se))
  show_rownames <- nrow(mat_sig_scaled) <= 50

  pheatmap(mat_sig_scaled,
           cluster_rows = TRUE, cluster_cols = TRUE,
           show_rownames = show_rownames,
           annotation_col = anno_col,
           main = paste0("Significant proteins (n=", nrow(mat_sig_scaled), ")"),
           color = colorRampPalette(c("blue", "white", "red"))(100))
} else if (nrow(mat_sig) > 500) {
  cat("Too many significant proteins for heatmap display:", nrow(mat_sig), "\n")
} else {
  cat("Fewer than 2 significant proteins with complete data.\n")
}
```

# Technical QC

::: {.callout-note collapse="true"}
## Model Types
Proteins with sufficient observations use `Linear_Model_moderated`. Those with missing group means use `Imputed_Mean_moderated` — these are less reliable.
:::

```{r model-summary}
model_summary <- list()
for (cn in contrast_names) {
  model_col <- paste0("constrast_", cn, ".modelName")
  if (model_col %in% colnames(rd)) {
    tbl <- table(rd[[model_col]])
    model_summary[[cn]] <- data.frame(
      Contrast = cn,
      Model = names(tbl),
      Count = as.integer(tbl),
      row.names = NULL)
  }
}
if (length(model_summary) > 0) {
  model_df <- do.call(rbind, model_summary)
  fgczDatatable(model_df, caption = "Modelling approach per contrast",
                filter = "none")
}
```

# Input Dataset

```{r input-dataset}
fgczDatatable(sample_info, caption = "Sample annotation")
```

# Session Info

```{r session-info}
sessionInfo()
```

:::

```{r agent-metadata}
#| include: false
## Write agent metadata sidecar
comparison_list <- lapply(contrast_names, function(cn) {
  prefix <- paste0("constrast_", cn)
  fdr_col  <- paste0(prefix, ".FDR")
  diff_col <- paste0(prefix, ".diff")

  n_sig_up   <- sum(rd[[fdr_col]] < fdr_threshold &
                       rd[[diff_col]] > diff_threshold, na.rm = TRUE)
  n_sig_down <- sum(rd[[fdr_col]] < fdr_threshold &
                       rd[[diff_col]] < -diff_threshold, na.rm = TRUE)
  list(
    name = cn,
    formula = contrasts[cn, "contrast"],
    method = "prolfqua",
    significantUp = n_sig_up,
    significantDown = n_sig_down
  )
})

## Top hits from first contrast
first_cn <- contrast_names[1]
prefix   <- paste0("constrast_", first_cn)
fdr_col  <- paste0(prefix, ".FDR")
diff_col <- paste0(prefix, ".diff")
name_col <- paste0(prefix, ".gene_name")

top_idx <- order(rd[[fdr_col]])[1:min(20, nrow(rd))]
top_hits <- data.frame(
  id = rownames(rd)[top_idx],
  name = if (name_col %in% colnames(rd)) rd[top_idx, name_col] else rownames(rd)[top_idx],
  log2FC = round(rd[top_idx, diff_col], 3),
  fdr = signif(rd[top_idx, fdr_col], 3)
)

fgczAgentMetadata(
  reportType = "differential-expression",
  omicsType = "proteomics",
  comparisons = comparison_list,
  thresholds = list(fdr = fdr_threshold, diff = diff_threshold),
  counts = list(
    totalTested = nrow(rd),
    significantUp = sum(summary_df$Up),
    significantDown = sum(summary_df$Down)
  ),
  topHits = top_hits,
  files = list(
    resultTable = result_file,
    plots = "plots/"
  )
)
```
