---
params:
  reportTitle: "ScSeurat"
title: "`r params$reportTitle`"
metadata-files: ["_fgcz-report.yml"]
---

```{r setup}
#| include: false
library(clustree)
library(kableExtra)
library(htmltools)
library(DT)
library(NMF)
library(pheatmap)
library(viridis)
library(cowplot)
library(scran)
library(RColorBrewer)
library(plotly)
library(tidyverse)
library(SingleR)
library(scater)
library(Seurat)
library(AUCell)
library(ezRun)
library(HDF5Array)
library(enrichR)
library(patchwork)
library(SCpubr)
library(dplyr)
library(cellxgenedp)
library(UpSetR)
library(jsonlite)

## ── Data loading (qs2 with rds fallback, checks data/ first) ──
.fgcz_load <- function(name, nthreads = 4L) {
  paths <- c(
    file.path("data", paste0(name, ".qs2")),
    paste0(name, ".qs2"),
    file.path("data", paste0(name, ".rds")),
    paste0(name, ".rds")
  )
  for (p in paths) {
    if (file.exists(p)) {
      if (grepl("\\.qs2$", p)) return(qs2::qs_read(p, nthreads = nthreads))
      return(readRDS(p))
    }
  }
  stop("No .qs2 or .rds file found for: ", name)
}

## ── DT helper ─────────────────────────────────────────────────
.dt <- function(df, caption = NULL, pageLength = 10L,
                filter = "none", rownames = FALSE, escape = TRUE, ...) {
  DT::datatable(
    data = df, caption = caption, filter = filter,
    rownames = rownames, escape = escape,
    class = "compact stripe",
    options = list(
      pageLength = pageLength,
      scrollX = TRUE,
      dom = "tip",
      columnDefs = list(
        list(targets = "_all", className = "dt-left")
      )
    ),
    ...
  ) |> DT::formatStyle(columns = seq_len(ncol(df)),
                        `text-align` = "left",
                        `white-space` = "normal",
                        `word-wrap` = "break-word",
                        `max-width` = "300px")
}

## Emit a DT widget as raw HTML inside results='asis' chunks
.dt_asis <- function(...) {
  widget <- .dt(...)
  cat(as.character(htmltools::as.tags(widget)))
}

## ── Output directories ────────────────────────────────────────
for (d in c("plots", "excel", "html")) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}

## ── Plot saving (PNG + PDF to plots/) ─────────────────────────
.save_plot <- function(plot, name, width = 6, height = 5, dpi = 300) {
  if (!dir.exists("plots")) dir.create("plots", recursive = TRUE)
  png_path <- file.path("plots", paste0(name, ".png"))
  pdf_path <- file.path("plots", paste0(name, ".pdf"))
  ggplot2::ggsave(png_path, plot = plot, width = width, height = height,
                  dpi = dpi, bg = "white")
  ggplot2::ggsave(pdf_path, plot = plot, width = width, height = height,
                  device = "pdf", bg = "white")
  invisible(png_path)
}

## ── Agent metadata writer ─────────────────────────────────────
.write_metadata <- function(...) {
  args <- list(...)
  args$generatedAt <- format(Sys.time(), "%Y-%m-%dT%H:%M:%S%z")
  args$softwareVersions <- list(
    R = paste0(R.version$major, ".", R.version$minor),
    quarto = tryCatch(system("quarto --version", intern = TRUE)[1],
                      error = function(e) "unknown"))
  jsonlite::write_json(args, path = "report-metadata.json",
                       pretty = TRUE, auto_unbox = TRUE, null = "null")
}

## ── Load required objects ─────────────────────────────────────
param             <- .fgcz_load("param")
output            <- .fgcz_load("output")
scData            <- .fgcz_load("scData", nthreads = param$cores)
allCellsMeta      <- .fgcz_load("allCellsMeta")
cellsPerGeneFraction <- .fgcz_load("cellsPerGeneFraction")

scData$qc.doublet <- scData$doubletClass %in% "doublet"
species <- getSpecies(param$refBuild)
```

```{r load-optional}
#| include: false

## ── Optional annotation objects ──────────────────────────────
aziResults <- if (file.exists("aziResults.qs2")) {
  az <- .fgcz_load("aziResults")
  scData <- AddMetaData(scData, az)
  az
} else if (ezRobjExists("aziResults")) {
  az <- ezLoadRobj("aziResults")
  scData <- AddMetaData(scData, az)
  az
} else {
  NULL
}

cellxgeneResults <- if (file.exists("cellxgeneResults.qs2")) {
  cxg <- .fgcz_load("cellxgeneResults")
  scData <- AddMetaData(scData, cxg)
  cxg
} else if (ezRobjExists("cellxgeneResults")) {
  cxg <- ezLoadRobj("cellxgeneResults")
  scData <- AddMetaData(scData, cxg)
  cxg
} else {
  NULL
}

singler.results <- if (file.exists("singler.results.qs2")) {
  .fgcz_load("singler.results")
} else if (ezRobjExists("singler.results")) {
  ezLoadRobj("singler.results")
} else {
  NULL
}

cells.AUC <- if (file.exists("cells.AUC.qs2")) {
  .fgcz_load("cells.AUC")
} else if (ezRobjExists("cells.AUC")) {
  ezLoadRobj("cells.AUC")
} else {
  NULL
}

enrichRout <- if (file.exists("enrichRout.qs2")) {
  er <- .fgcz_load("enrichRout")
  if (length(er) == 1) NULL else er
} else if (ezRobjExists("enrichRout")) {
  er <- ezLoadRobj("enrichRout")
  if (length(er) == 1) NULL else er
} else {
  NULL
}

pathwayActivity <- if (file.exists("pathwayActivity.qs2")) {
  .fgcz_load("pathwayActivity")
} else if (ezRobjExists("pathwayActivity")) {
  ezLoadRobj("pathwayActivity")
} else {
  NULL
}

TFActivity <- if (file.exists("TFActivity.qs2")) {
  .fgcz_load("TFActivity")
} else if (ezRobjExists("TFActivity")) {
  ezLoadRobj("TFActivity")
} else {
  NULL
}

## scType results
sctype_results <- if (file.exists("sctype_results.qs2")) {
  res <- .fgcz_load("sctype_results")
  if ("scData" %in% names(res)) res$scData <- NULL
  res
} else if (ezRobjExists("sctype_results")) {
  res <- ezLoadRobj("sctype_results")
  if ("scData" %in% names(res)) res$scData <- NULL
  res
} else {
  NULL
}
sctype_available <- !is.null(sctype_results) &&
  "sctype_classification" %in% colnames(scData@meta.data)

## Azimuth Pan-Human availability
azimuth_available <- "final_level_labels" %in% colnames(scData@meta.data)

## ── Shared color palette ──────────────────────────────────────
shared_colors <- c(
  scales::hue_pal()(12),
  DiscretePalette(n = 26, palette = "alphabet2"),
  DiscretePalette(n = 36, palette = "polychrome"),
  DiscretePalette(n = 24, palette = "stepped"),
  DiscretePalette(n = 26, palette = "alphabet2"),
  DiscretePalette(n = 36, palette = "polychrome")
)
names(shared_colors) <- NULL

## ── posMarkers ────────────────────────────────────────────────
posMarkers <- readxl::read_xlsx("posMarkers.xlsx")
posMarkers$gene <- as.factor(posMarkers$gene)
clusterSet <- posMarkers$cluster |> as.integer() |> unique() |> sort(na.last = TRUE)
if (any(is.na(clusterSet))) {
  posMarkers$cluster <- as.factor(posMarkers$cluster)
} else {
  posMarkers$cluster <- factor(posMarkers$cluster, levels = clusterSet)
}

var_height <- 1
clusterNameNChar <- max(nchar(levels(Idents(scData))))
```

::: {.panel-tabset}

# Quality control

`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`

```{r prepare-qc}
#| include: false
xxAll <- SingleCellExperiment(colData = allCellsMeta)
xxAll$discard <- !xxAll$useCell
```

### QC diagnostic plots

```{r qc-plots-line1}
#| fig-height: 4
#| fig-width: 8

xxAll$genePerCount <- xxAll$nFeature_RNA / xxAll$nCount_RNA

plot1 <- plotColData(
  xxAll, x = "Sample",
  y = "nCount_RNA",
  colour_by = "discard"
) +
  scale_y_log10() +
  labs(title = "UMI counts",
       x = "",
       y = "UMI counts per spot") +
  theme(
    legend.position = "top",
  )

plot1 + plot_layout(ncol = 2)
```

```{r qc-plots-line2}
#| fig-width: 8
#| fig-height: 4

plot2 <- plotColData(
  xxAll,
  x = "Sample",
  y = "genePerCount",
  colour_by = "discard"
) +
  labs(
    title = "Detected genes",
    x = "",
    y = "Detected genes"
  ) +
  theme(
    legend.position = "none",
  )

plot3 <- plotColData(
  xxAll,
  x = "nCount_RNA",
  y = "genePerCount",
  colour_by = "discard",
) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = "",
    x = "UMI counts",
    y = "Detected genes"
  ) +
  theme(
    legend.position = "none",
  )

plot2 + plot3
```

```{r qc-plots-line3}
#| fig-width: 8
#| fig-height: 4

plot4 <- plotColData(
  xxAll,
  x = "Sample",
  y = "percent_mito",
  colour_by = "discard"
) +
  labs(
    title = "Mito percent",
    x = "",
    y = "Mitochondrial reads %"
  ) +
  theme(
    legend.position = "none"
    )

plot5 <- plotColData(
  xxAll,
  x = "nCount_RNA",
  y = "percent_mito",
  colour_by = "discard"
) +
  scale_x_log10() +
  labs(
    x = "UMI counts",
    y = "Mitochondrial reads %"
  ) +
  theme(
    legend.position = "none"
  )

plot4 + plot5
```

```{r qc-plots-line4}
#| fig-width: 8
#| fig-height: 4

plot6 <- plotColData(
  xxAll,
  x = "Sample",
  y = "percent_riboprot",
  colour_by = "discard"
) +
  labs(
    x = "",
    title = "Ribosomal percent",
    y = "Ribosomal protein reads %"
  ) +
  theme(
    legend.position = "none"
  )

plot7 <- plotColData(
  xxAll,
  x = "nCount_RNA",
  y = "percent_riboprot",
  colour_by = "discard"
) +
  scale_x_log10() +
  labs(
    x = "UMI counts",
    y = "Ribosomal protein reads %"
  ) +
  theme(
    legend.position = "none"
  )

plot6 + plot7
```

```{r qc-plots-hemoglobin}
#| fig-width: 8
#| fig-height: 4

plot_hb1 <- plotColData(
  xxAll,
  x = "Sample",
  y = "percent_hb",
  colour_by = "discard"
) +
  labs(
    title = "Hemoglobin percent",
    x = "",
    y = "Hemoglobin reads %"
  ) +
  theme(legend.position = "none")

plot_hb2 <- plotColData(
  xxAll,
  x = "nCount_RNA",
  y = "percent_hb",
  colour_by = "discard"
) +
  scale_x_log10() +
  labs(
    x = "UMI counts",
    y = "Hemoglobin reads %"
  ) +
  theme(legend.position = "none")

plot_hb1 + plot_hb2
```

```{r qc-plots-line5}
#| fig-width: 8
#| fig-height: 4

plot8 <- plotColData(
  xxAll,
  x = "Sample",
  y = "doubletScore",
  colour_by = "discard"
) +
  labs(
    title = "Doublet score",
    y = "Doublet score"
  ) +
  theme(
    legend.position = "none",
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank()
  )

plot9 <- plotColData(
  xxAll,
  x = "nCount_RNA",
  y = "doubletScore",
  colour_by = "discard"
) +
  labs(
    x = "UMI counts",
    y = "Percent ribosomal protein reads"
  ) +
  theme(
    legend.position = "none"
  )

plot8 + plot9
```

<br/>

### Cell filtering

::: {.row}
::: {.col-md-6}

```{r low-quality-cells}
#| fig-align: center
qcList <- lapply(allCellsMeta[, grep("^qc", colnames(allCellsMeta))], function(x) {
  rownames(allCellsMeta)[x]
})
qcList <- qcList[sapply(qcList, length) > 0]
if (length(qcList) > 1) {
  UpSetR::upset(fromList(qcList), nsets = 5)
}
```

:::

::: {.col-md-6}

```{r qc-table}
#| fig-align: center
qcStats <- ezFrame("Number of Cells" = integer(0),
                   "Min Pass Value"  = integer(0),
                   "Max Pass Value"  = integer(0),
                   Fraction          = integer(0))
qcStats["Library Size", ] <- c(
  c(sum(allCellsMeta$qc.lib), range(allCellsMeta$nCount_RNA[allCellsMeta$useCell])) |> as.integer(),
  round(sum(allCellsMeta$qc.lib) / nrow(allCellsMeta), 2)
)
qcStats["Expressed genes", ] <- c(
  c(sum(allCellsMeta$qc.nexprs), range(allCellsMeta$nFeature_RNA[allCellsMeta$useCell])) |> as.integer(),
  round(sum(allCellsMeta$qc.nexprs) / nrow(allCellsMeta), 2)
)
qcStats["Mitochondrial Percent", ] <- c(
  c(sum(allCellsMeta$qc.mito), range(allCellsMeta$percent_mito[allCellsMeta$useCell])) |> as.integer(),
  round(sum(allCellsMeta$qc.mito) / nrow(allCellsMeta), 2)
)
qcStats["Ribosomal protein Percent", ] <- c(
  c(sum(allCellsMeta$qc.riboprot), range(allCellsMeta$percent_riboprot[allCellsMeta$useCell])) |> as.integer(),
  round(sum(allCellsMeta$qc.riboprot) / nrow(allCellsMeta), 2)
)
qcStats["Doublets", ] <- c(
  sum(allCellsMeta$qc.doublet),
  NA,
  NA,
  round(sum(allCellsMeta$qc.doublet) / nrow(allCellsMeta), 2)
)
qcStats["Empty Droplets", ] <- c(
  sum(allCellsMeta$qc.empty),
  NA,
  NA,
  round(sum(allCellsMeta$qc.empty) / nrow(allCellsMeta), 2)
)
qcStats["Total removed (thresholding)", ] <- c(
  sum(!allCellsMeta$useCell),
  NA,
  NA,
  round(sum(!allCellsMeta$useCell) / nrow(allCellsMeta), 2)
)
qcStats["Total removed (manual)", ] <- c(
  sum(allCellsMeta$useCell) - ncol(scData),
  NA,
  NA,
  round((sum(allCellsMeta$useCell) - ncol(scData)) / nrow(allCellsMeta), 2)
)
qcStats["Cells remaining", ] <- c(
  sum(ncol(scData)),
  NA,
  NA,
  round(sum(ncol(scData)) / nrow(allCellsMeta), 2)
)

qc_dt <- as.data.frame(qcStats) |>
  tibble::rownames_to_column(var = "Metric")
qc_dt <- qc_dt[, c("Metric", setdiff(names(qc_dt), "Metric"))]

dt <- DT::datatable(
  qc_dt,
  rownames = FALSE,
  options  = list(
    autoWidth = TRUE,
    dom       = "t"
  )
)

htmlwidgets::prependContent(
  dt,
  htmltools::tags$style(htmltools::HTML(
    "table.dataTable tbody td:empty::before{content:'-'; color:#6c757d;}"
  ))
)
```

:::
:::

<br/>

### Dimensionality reduction

```{r scree-plot}
#| fig-align: center
#| fig-width: 14

pca_name <- if ("pca.sketch" %in% names(scData@reductions)) "pca.sketch" else "pca"

k      <- param$npcs
top_n  <- max(80, k)

stdev  <- scData[[pca_name]]@stdev |> head(top_n)
pct    <- ((stdev^2) / sum(stdev^2)) * 100
cumu   <- cumsum(pct)

plot_df <- data.frame(
  PC   = seq_along(pct),
  pct  = pct,
  cumu = cumu,
  used = seq_along(pct) <= k)

ggplot(plot_df, aes(x = PC)) +
  geom_col(aes(y = pct, fill = used),
           width = 0.7,
           show.legend = FALSE) +
  scale_fill_manual(values = c(`TRUE` = "grey20",
                               `FALSE` = "grey75")) +
  geom_line(aes(y = cumu / max(cumu) * max(pct), group = 1),
            linewidth = 0.5,
            color = "orange") +
  geom_point(aes(y = cumu / max(cumu) * max(pct)),
             size = 1,
             color = "orange") +
  geom_vline(xintercept = k,
             linetype = "dashed",
             color = "grey40") +
  scale_y_continuous(
    name = "Variance explained (%)",
    sec.axis = sec_axis(~ . / max(plot_df$pct) * 100,
                        name = "Cumulative variance (%)")) +
  labs(
    title    = "PCA variance explained",
    subtitle = paste0("PCs used: ", k, "   |   Cumulative: ", sprintf('%.2f%%', cumu[k])),
    x = NULL) +
  theme_bw() +
  theme(
    axis.title.y.left  = element_text(color = "grey20"),
    axis.title.y.right = element_text(color = "orange"),
    plot.title         = element_text(hjust = 0.5),
    panel.grid.minor   = element_blank())
```

<br/>

### GeneSet enrichment

```{r explore-hvgs}
#| results: asis
HVGs <- VariableFeatures(object = scData)

jsCall <- paste0('enrich({list: "', paste(HVGs, collapse = "\\n"), '", popup: true});')
enrichrCall <- paste0("<a href='javascript:void(0)' onClick='",
                      jsCall,
                      "'>Enrichr website</a>")

enrichrTable <- tibble(GeneSet = 'HVGs',
                       "N HVGs" = length(HVGs),
                       "Link" = enrichrCall)

kable(enrichrTable,
      format = "html",
      escape = FALSE
      ) |>
  kable_styling("striped",
                full_width = F,
                position = "left")
```

The dimension reduction is based on the top `r param$nfeatures` highly variable genes (HVGs): `r enrichrCall`

#### Selected QC metrics

We use several common QC metrics to identify low-quality cells based on their expression profiles. The metrics that were chosen are described below.

1. The library size is defined as the total sum of counts across all relevant features for each cell. Cells with small library sizes are of low quality as the RNA has been lost at some point during library preparation.
2. The number of expressed features in each cell is defined as the number of genes with non-zero counts for that cell. Any cell with very few expressed genes is likely to be of poor quality as the diverse transcript population has not been successfully captured.
3. The proportions of mitochondrial and ribosomal genes per cell. High proportions are indicative of poor-quality cells (Islam et al. 2014; Ilicic et al. 2016), presumably because of the loss of cytoplasmic RNA from perforated cells.

<br/>

#### Diagnostic plots

A key assumption here is that the QC metrics are independent of the biological state of each cell. Poor values (e.g., low library sizes, high mitochondrial or ribosomal proportions) are presumed to be driven by technical factors rather than biological processes, meaning that the subsequent removal of cells will not misrepresent the biology in downstream analyses. Major violations of this assumption would potentially result in the loss of cell types that have, say, systematically low RNA content or high numbers of mitochondria. We can check for such violations using some diagnostics plots. In the most ideal case, we would see normal distributions that would justify the thresholds used in outlier detection. A large proportion of cells in another mode suggests that the QC metrics might be correlated with some biological state, potentially leading to the loss of distinct cell types during filtering.

<br>

#### Dimensionality reduction

Dimensionality reduction aims to reduce the number of separate dimensions in the data. This is possible because different genes are correlated if they are affected by the same biological process. Thus, we do not need to store separate information for individual genes, but can instead compress multiple features into a single dimension. This reduces computational work in downstream analyses, as calculations only need to be performed for a few dimensions rather than thousands of genes; reduces noise by averaging across multiple genes to obtain a more precise representation of the patterns in the data, and enables effective plotting of the data.

The numbers of PCs that should be retained for downstream analyses typically range from 10 to 50. However, identifying the true dimensionality of a dataset can be challenging, that's why we recommend considering the 'Elbow plot' approach. a ranking of principal components based on the percentage of variance explained by each one. The assumption is that each of the top PCs capturing biological signal should explain much more variance than the remaining PCs. Thus, there should be a sharp drop in the percentage of variance explained when we move past the last "biological" PC. This manifests as an elbow in the scree plot, the location of which serves as a natural choice for a number of PCs.

# Clustering

`r paste("Resolution:", param$resolution)`

```{r clustering-plot1}
#| fig-width: 9

n_clusters <- nlevels(Idents(scData))
show_leg   <- n_clusters <= 20

p1 <- DimPlot(scData,
              label = FALSE,
              pt.size = 1.5) +
  labs(color = "Clusters") +
  guides(color = if (show_leg) guide_legend(title = "Clusters") else "none") +
  ggtitle("UMAP") +
  theme_void() +
  theme(legend.position = "none")

p1 <- Seurat::LabelClusters(
  plot     = p1,
  id       = "ident",
  fontface = "bold",
  color    = "#383838",
  size     = 4
)

p2 <- DimPlot(scData,
              label = FALSE,
              reduction = "tsne",
              pt.size = 1.5) +
  labs(color = "Clusters") +
  guides(color = if (show_leg) guide_legend(title = "Clusters") else "none") +
  ggtitle("TSNE") +
  theme_void()

p2 <- Seurat::LabelClusters(
  plot     = p2,
  id       = "ident",
  fontface = "bold",
  color    = "#383838",
  size     = 4
)

(p1 + p2) +
  plot_layout(nrow = 1, guides = if (show_leg) "keep" else "auto")
```

<br/>

::: {.row}
::: {.col-md-6}

```{r cells-per-cluster}
cellIdents_perSample <- data.frame(ident = Idents(scData), Sample = scData$Sample)

if (!('Batch' %in% colnames(scData@meta.data))) {
  scData@meta.data$Batch <- scData@meta.data$Sample
}

cellIdents_perSample <- scData@meta.data[, c('seurat_clusters', 'Batch')]
cellIdents_perSample$Batch <- factor(cellIdents_perSample$Batch)

ggplot(cellIdents_perSample, aes(x = seurat_clusters,
                    fill = Batch)) +
  geom_bar() +
  labs(title = "Cells per cluster and sample (counts)",
       x = NULL,
       y = "N cells"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        axis.text.x = element_text(angle = 45)
  )
```

:::

::: {.col-md-6}

```{r cells-cluster-table}
df <- cellsProportion(scData,
                      groupVar1 = "seurat_clusters",
                      groupVar2 = "Sample") |>
  as.data.frame() |>
  dplyr::rename(Cluster = seurat_clusters)

is_pct_col <- grepl("_fraction$", colnames(df))

new_names <- colnames(df)
new_names[is_pct_col] <- sub("_fraction$", " (%)", new_names[is_pct_col])

is_count_col <- !is_pct_col & new_names != "Cluster"
new_names[is_count_col] <- paste0(new_names[is_count_col], " (n)")

colnames(df) <- new_names

datatable(
  df,
  caption   = htmltools::tags$caption("Cells per cluster and sample"),
  extensions = c("Buttons"),
  rownames  = FALSE,
  escape    = TRUE,
  options   = list(
    paging    = TRUE,
    dom       = "Bfrtip",
    autoWidth = TRUE,
    buttons   = list(list(extend = "excel",
                          text = "Excel",
                          filename = "clusters")
                     )
  )
) |> formatPercentage(columns = which(is_pct_col), digits = 1)
```

:::
:::

### Clustering

In order to find clusters of cells we first built a graph called K-nearest neighbor (KNN), where each node is a cell that is connected to its nearest neighbors in the high-dimensional space. Edges are weighted based on the similarity between the cells involved, with higher weight given to cells that are more closely related. This step takes as input the previously defined dimensionality of the dataset (first `r param$npcs` PCs). We then applied algorithms to identify "communities" of cells that are more connected to cells in the same community than they are to cells of different communities. Each community represents a cluster that we can use for downstream interpretation.

We can visualize the distribution of clusters in the TSNE and UMAP plots. However, we should not perform downstream analyses directly on their coordinates. These plots are most useful for checking whether two clusters are actually neighboring subclusters or whether a cluster can be split into further subclusters.

# Cluster assessment

### Segregation of clusters by various sources of uninteresting variation.

Once we have created the clusters we need to assess if the clustering was driven by technical artifacts or uninteresting biological variability, such as cell cycle, mitochondrial gene expression.
We can explore whether the cells cluster by the different cell cycle phases. In such a case, we would have clusters where most of the cells would be in one specific phase. This bias could be taken into account when normalizing and transforming the data prior to clustering. We can also look at the total number of reads, genes detected and mitochondrial gene expression. The clusters should be more or less even but if we observe big differences among some of them for these metrics, we will keep an eye on them and see if the cell types we identify later can explain the differences.

```{r cell-cycle}
#| fig-width: !expr min(6 + clusterNameNChar*0.2, 10)
#| fig-height: !expr min(6 + clusterNameNChar*0.2, 10)
#| eval: !expr !is.null(scData@meta.data$CellCycle)
if (any(!is.na(scData$CellCycle))) {
  DimPlot(scData, label = FALSE, group.by = "CellCycle")
}
p2 <- DimPlot(scData, reduction = "umap", label = FALSE, pt.size = 1.5) + labs(color = "seurat_clusters") + NoLegend()
p2 <- LabelClusters(p2, id = "ident", fontface = "bold", color = "black", size = 6)
p2
```

```{r uninteresting-variation}
#| fig-width: !expr min(6 + clusterNameNChar*0.2, 10)
#| fig-height: !expr min(6 + clusterNameNChar*0.2, 10)
VlnPlot(scData, features = "nCount_RNA", log = TRUE) + ggtitle("Number of UMIs vs ident") + NoLegend()
FeaturePlot(scData, features = "nCount_RNA", label = TRUE)

VlnPlot(scData, features = "nFeature_RNA") + ggtitle("Number of genes vs ident") + NoLegend()
FeaturePlot(scData, features = "nFeature_RNA", label = TRUE)

VlnPlot(scData, features = "percent_mito") + ggtitle("Mitochondrial percentage vs ident") + NoLegend()
FeaturePlot(scData, features = "percent_mito", label = TRUE)

VlnPlot(scData, features = "percent_riboprot") + ggtitle("Ribosomal protein percentage vs ident") + NoLegend()
FeaturePlot(scData, features = "percent_riboprot", label = TRUE)

VlnPlot(scData, features = "percent_hb") + ggtitle("Hemoglobin percentage vs ident") + NoLegend()
FeaturePlot(scData, features = "percent_hb", label = TRUE)

VlnPlot(scData, features = "doubletScore") + ggtitle("DoubletScore vs ident") + NoLegend()
FeaturePlot(scData, features = "doubletScore", label = TRUE)
```

```{r decontx}
#| fig-width: !expr min(6 + clusterNameNChar*0.2, 10)
#| fig-height: !expr min(6 + clusterNameNChar*0.2, 10)
if (!is.null(scData@meta.data$DecontX_contFrac)) {
  VlnPlot(scData, features = "DecontX_contFrac", y.max = 1) + ggtitle("DecontX ambient fraction vs ident") + NoLegend()
  FeaturePlot(scData, features = "DecontX_contFrac", label = TRUE)
}
```

```{r soupx}
#| fig-width: !expr min(6 + clusterNameNChar*0.2, 10)
#| fig-height: !expr min(6 + clusterNameNChar*0.2, 10)
if (!is.null(scData@meta.data$SoupX_contFrac)) {
  VlnPlot(scData, features = "SoupX_contFrac", y.max = 1) + ggtitle("SoupX ambient fraction vs ident") + NoLegend()
  FeaturePlot(scData, features = "SoupX_contFrac", label = TRUE)
}
```

```{r cell-probability}
#| fig-width: !expr min(6 + clusterNameNChar*0.2, 10)
#| fig-height: !expr min(6 + clusterNameNChar*0.2, 10)
if (!is.null(scData@meta.data$negLog10CellPValue)) {
  VlnPlot(scData, features = "negLog10CellPValue") + ggtitle("EmptyDrops Cell Confidence") + NoLegend()
  FeaturePlot(scData, features = "negLog10CellPValue", label = TRUE) + labs(color = "EmptyDrops\nCell Confidence")
}
```

```{r control-seq-plot}
#| fig-width: 12
#| fig-height: !expr shrinkToRange(length(param$controlSeqs)*3/4, c(6, 18))
#| eval: !expr ezIsSpecified(param$controlSeqs)
genesToPlot <- gsub("_", "-", param$controlSeqs)
genesToPlot <- intersect(genesToPlot, rownames(scData))
if (length(genesToPlot) > 0) {
  if (length(genesToPlot) < 20) {
    FeaturePlot(scData, genesToPlot) + NoLegend()
  } else {
    DoHeatmap(scData, features = genesToPlot)
    DotPlot(scData, features = genesToPlot) + coord_flip() +
      theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
  }
}
```

#### Cluster resolution

One of the most important parameters when clustering is k, the number of nearest neighbors used to construct the graph. This controls the resolution of the clustering where higher k yields a more inter-connected graph and broader clusters. Users can experiment with different values of k to obtain a satisfactory resolution. We recommend increasing the resolution when a rare population is expected.
Below, it is shown a clustering tree that helps us to visualize the relationships between clusters at a range of resolutions. Each cluster forms a node in the tree and edges are constructed by considering the cells in a cluster at a lower resolution that end up in a cluster at the next highest resolution. By connecting clusters in this way, we can see how clusters are related to each other, which are clearly distinct and which are unstable. The size of each node is related to the number of cells in each cluster and the color indicates the clustering resolution. Edges are colored according to the number of cells they represent and the transparency shows the incoming node proportion, the number of cells in the edge divided by the number of samples in the node it points to.

```{r clustree}
#| fig-align: center
#| fig-width: 8
#| fig-height: 8
clustree::clustree(scData, prefix = paste0(DefaultAssay(scData), "_snn_res."))
```

# Cluster markers

```{r wilcoxon-test}
#| results: asis
#| eval: !expr param$DE.method == "wilcox"
cat("We found positive markers that defined clusters compared to all other cells via differential expression. The test we used was the Wilcoxon Rank Sum test. Genes with an average, at least 0.25-fold difference (log-scale) between the cells in the tested cluster and the rest of the cells and an adjusted p-value < 0.05 were declared as significant.")
```

```{r lr-test}
#| results: asis
#| eval: !expr param$DE.method == "LR"
cat("We found positive markers that defined clusters compared to all other cells via differential expression using a logistic regression test and including in the model the cell cycle as the batch effect. Genes with an average, at least 0.25-fold difference (log-scale) between the cells in the tested cluster and the rest of the cells and an adjusted p-value < 0.05 were declared as significant.")
```

```{r pos-markers}
ezInteractiveTableRmd(posMarkers, digits = 3, rowNames = FALSE)
```

# Marker plots

Here, we use a heatmap and a dotplot to visualize simultaneously the top 5 markers in each cluster. Be aware that some genes may be in the top markers for different clusters.

<br/>

```{r marker-plot-setup}
#| include: false
top5 <- posMarkers |>
  group_by(cluster) |>
  slice_max(n = 5, order_by = tibble(diff_pct, avg_log2FC))
genesToPlot <- c(gsub("_", "-", param$controlSeqs), unique(as.character(top5$gene)))
genesToPlot <- intersect(genesToPlot, rownames(scData))
```

#### Heatmap & DotPlot

```{r marker-heatmap}
#| fig-width: 12
#| fig-height: !expr length(levels(Idents(scData)))
DoHeatmap(scData, features = unique(genesToPlot))
DotPlot(scData, features = genesToPlot) + coord_flip() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

#### Group-wise DE Plot

```{r marker-groupwise-plot}
#| fig-width: 13
#| fig-height: !expr length(levels(Idents(scData)))
do_GroupwiseDEHeatmap(sample = scData,
                   de_genes = posMarkers |>
                     mutate(gene = as.character(gene),
                            p_val_adj = ifelse(p_val_adj < 1e-100, 1e-100, p_val_adj)),
                   min.cutoff = NA)
```

```{r plot-user-genes}
#| fig-width: 12
#| fig-height: !expr shrinkToRange(length(param$controlSeqs)/5, c(6, 12))
#| eval: !expr ezIsSpecified(param$controlSeqs)
genesToPlot <- intersect(param$controlSeqs, rownames(scData))

if (length(genesToPlot) > 0) {
  DoHeatmap(scData, features = unique(genesToPlot))
  DotPlot(scData, features = genesToPlot) + coord_flip() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
} else {
  cat("controlseqs not found")
}
```

# Cell type annotation

::: {.panel-tabset}

## EnrichR

Cell type scoring using the EnrichR tool. This approach consists in performing a gene set enrichment analysis on the marker genes defining each cluster. This identifies the pathways and processes that are (relatively) active in each cluster based on the upregulation of the associated genes compared to other clusters.

```{r enrichr-per-cluster-prep}
#| results: asis
markersPerClusterTable <- c()
eachCluster <- 0
for (eachCluster in levels(posMarkers$cluster)) {
  markersPerCluster <- dplyr::filter(posMarkers, cluster == eachCluster) |>
    dplyr::arrange(desc(avg_log2FC))
  markersPerCluster <- head(markersPerCluster, min(nrow(markersPerCluster), 500))
  markersPerClusterTable <- rbind(markersPerClusterTable, markersPerCluster)
}

genesPerCluster <- split(markersPerClusterTable$gene, markersPerClusterTable$cluster)
jsCall <- paste0('enrich({list: "', sapply(genesPerCluster, paste, collapse = "\\n"), '", popup: true});')
enrichrCalls <- paste0("<a href='javascript:void(0)' onClick='", jsCall,
                       "'>Analyse at Enrichr website</a>")
```

```{r enrichr-markers}
#| results: asis
enrichrTable <- tibble(Cluster = names(genesPerCluster),
                       "# of posMarkers" = lengths(genesPerCluster),
                       "Enrichr link" = enrichrCalls)
if (!is.null(enrichRout) && length(enrichRout) > 0) {
  enrichRTerm <- as.data.frame(do.call(rbind, lapply(enrichRout, as.vector)))
  enrichRTerm <- map_df(enrichRTerm, ~ map_df(.x, ~ replace(.x, is.null(.x), NA)), .id = "database")
  enrichRTerm <- enrichRTerm |>
    group_by(Cluster, database) |>
    summarise(topTerms = paste(Term, collapse = "; ")) |>
    as.data.frame()
  if (nrow(enrichRTerm) > 0) {
    enrichRTerm <- reshape2::dcast(enrichRTerm, ... ~ database)
    enrichrTable <- merge(enrichrTable, enrichRTerm, by = "Cluster")
  }
}

kable(enrichrTable, format = "html", escape = FALSE,
      caption = paste0("GeneSet enrichment")) |>
  kable_styling("striped", full_width = F, position = "left")
```

```{r plot-enrichr-per-cluster}
#| fig-height: 5
#| fig-width: 10
#| eval: !expr !is.null(enrichRout)
for (cluster in names(enrichRout)) {
  enrichRDataFrame <- as.data.frame(do.call(rbind, enrichRout[[cluster]]))
  enrichRDataFrame$database <- sapply(rownames(enrichRDataFrame), function(x) str_split(x, "[.]")[[1]][1])
  print(ggplot(as.data.table(enrichRDataFrame), aes(database, Term))
        +
          geom_point(aes(color = OverlapGenesN, size = Combined.Score))
        +
          scale_color_viridis()
        +
          scale_fill_viridis()
        +
          labs(y = "Term", x = "Cluster", title = paste0("Cluster: ", cluster))
        +
          theme_bw()
        +
          theme(
            axis.title.x = element_blank(),
            axis.text.x = element_text(angle = 45, hjust = 1)
          ))
}
```

<br/>

## SingleR

Matching of clusters to annotated reference datasets using SingleR.

```{r singler}
#| fig-height: 12
#| fig-width: 12
#| results: asis
#| eval: !expr !is.null(singler.results)
clusterInfosList <- list()
plotlyPlots <- list()
nTopMarkers <- 10
for (r in names(singler.results)) {
  cat("##### Heatmap: ", r, "\n\n")
  print(plotScoreHeatmap(singler.results[[r]]$single.fine))
  print(plotScoreHeatmap(singler.results[[r]]$cluster.fine, show_colnames = TRUE))
  cat("\n\n")
  singler.single.labels  <- singler.results[[r]]$single.fine$labels
  singler.cluster.labels <- singler.results[[r]]$cluster.fine$labels[match(Idents(scData), rownames(singler.results[[r]]$cluster.fine))]

  cellInfo <- tibble(
    Cells = colnames(scData),
    Cluster = Idents(scData),
    SingleR.labels.cluster = singler.cluster.labels,
    SingleR.labels.single  = singler.single.labels) |>
    left_join(as_tibble(scData@reductions$umap@cell.embeddings, rownames = "Cells"))

  nrOfLabels_cluster <- length(unique(cellInfo$SingleR.labels.cluster))
  nrOfLabels_single  <- length(unique(cellInfo$SingleR.labels.single))

  if (nrOfLabels_single <= 9) {
    colsLabels <- brewer.pal(nrOfLabels_single, "Set1")
  } else {
    colsLabels <- colorRampPalette(brewer.pal(9, "Set1"))(nrOfLabels_single)
  }

  x <- list(title = "umap_1", zeroline = FALSE)
  y <- list(title = "umap_2", zeroline = FALSE)

  p1 <- plot_ly(cellInfo, x = ~umap_1, y = ~umap_2, color = ~SingleR.labels.single,
                text = ~paste("Cluster: ", Cluster,
                              "\nCell: ", Cells,
                              "\nSingleR.labels.cluster: ",
                              SingleR.labels.single),
                type = 'scatter', mode = "markers", marker = list(size = 5, opacity = 0.5),
                colors = colsLabels, width = 800, height = 500) |>
    layout(xaxis = x, yaxis = y, title = paste0(r, ": Single"))
  p2 <- plot_ly(cellInfo, x = ~umap_1, y = ~umap_2, color = ~SingleR.labels.cluster,
                text = ~paste("Cluster: ", Cluster,
                              "\nCell: ", Cells,
                              "\nSingleR.labels.cluster: ", SingleR.labels.cluster),
                type = 'scatter', mode = "markers", marker = list(size = 5, opacity = 0.5),
                colors = colsLabels, width = 800, height = 500) |>
    layout(xaxis = x, yaxis = y, title = paste0(r, ": Cluster"))
  plotlyPlots <- append(plotlyPlots, list(p1, p2))
  cat("\n\n")
  clusterInfos <- ezFrame(Sample  = as.character(unique(scData$Sample)),
                          Reference = r,
                          Cluster = levels(Idents(scData)))
  if (!is.null(singler.results)) {
    clusterInfos$SinglerCellType <- singler.results[[r]]$cluster.fine[clusterInfos$Cluster, "pruned.labels"]
  }
  topMarkers <- posMarkers |> group_by(cluster) |>
    slice_max(n = nTopMarkers, order_by = avg_log2FC)
  topMarkerString <- sapply(split(topMarkers$gene, topMarkers$cluster), paste, collapse = ", ")
  clusterInfos[["TopMarkers"]] <- topMarkerString[clusterInfos$Cluster]
  clusterInfos[["Percentage of Cells"]] <- table(Idents(scData)) / nrow(scData)
  clusterInfosList <- append(clusterInfosList, list(clusterInfos))
  cat("\n\n")
}
```

<br>

```{r singler-labels-umap}
#| results: asis
#| eval: !expr !is.null(singler.results)
cat("##### UMAPs with SingleR Labels\n\n")
cat("The single cell (top) and cluster (bottom) annotations are also shown on the UMAPs. Place the mouse over the cells to get information such as their UMAP coordinates, original cluster, the cells name and the label assigned by SingleR. You can also zoom in specific areas of the UMAP by dragging and drop with the mouse.\n")
tl <- htmltools::tagList(plotlyPlots)
tl
cat("\n\n")
```

<br/>

```{r singler-labels-table}
#| results: asis
#| eval: !expr !is.null(singler.results)
cat("##### Table of SingleR Labels\n\n")
cat(sprintf("The following table summarises the cluster-level labels for each reference (%s) and displays the top %s genes for that cluster in terms of their average log2-fold change. Note the top markers for every cluster will be identical regardless of reference used since this information is specific to the cluster and not the annotation reference.", paste(names(singler.results), collapse = ", "), nTopMarkers))
ezInteractiveTableRmd(bind_rows(clusterInfosList))
cat("\n\n")
```

## AUCell

Matching the cells to marker gene sets that are known to be characteristic for cell types. We use marker gene sets from the CellMarker database (Zhang X., Lan Y., Xu J., Quan F., Zhao E., Deng C., et al. (2019). CellMarker: a manually curated resource of cell markers in human and mouse. Nucleic Acids Res. 47, D721–d728. 10.1093/nar/gky900).

```{r auc-assignments}
#| eval: !expr !is.null(cells.AUC)
tab <- NULL
try({
  cells_assignment <- AUCell_exploreThresholds(cells.AUC, plotHist = FALSE, assign = TRUE, nCores = param$cores)
  cellsAssigned <- lapply(cells_assignment, function(x) x$assignment)
  assignmentTable <- reshape2::melt(cellsAssigned, value.name = "cell")
  colnames(assignmentTable)[2] <- "geneSet"
  tras_cells.AUC <- t(assay(cells.AUC))
  full.labels <- colnames(tras_cells.AUC)[max.col(tras_cells.AUC)]
  tab <- table(full.labels, Idents(scData))
  var_height <- shrinkToRange(nrow(tab) * 0.5, c(6, 16))
})
```

```{r auc-assignments-text}
#| results: asis
#| eval: !expr !is.null(cells.AUC)
cat("We can explore the cell assignment results using different plots. Below, we show a heatmap that represents the number of cells (in log scale) from each cluster that were assigned to the different cell types. After calculating an AUC score for each cell and cell type, we assign cell type identity by taking the cell type with the top AUC as the label for that cell. Some cell types may be missing because no cells obtained their top AUC score for it.")
cat("\n\n")
```

```{r auc-assignments-heatmap}
#| results: asis
#| fig-align: center
#| fig-width: 10
#| fig-height: !expr var_height
#| eval: !expr !is.null(cells.AUC)
if (!is.null(tab) && nrow(tab) > 2) {
  pheatmap(log10(tab + 10), color = viridis::viridis(100))
}
```

```{r auc-tsne-text}
#| results: asis
#| eval: !expr !is.null(cells.AUC)
cat("The plots below show for every cell type:\n")
cat('\n')
cat("1) The distribution of the AUC values in the cells. The ideal situation will be a bi-modal distribution, in which most cells in the dataset have a low \"AUC\" compared to a population of cells with a higher value.  The size of the gene-set will also affect the results. With smaller gene-genes (fewer genes), it is more likely to get cells with AUC = 0. While this is the case of the \"perfect markers\" it is also easier to get it by chance with small datasets. The vertical bars correspond to several thresholds that could be used to consider a gene-set 'active'. The thickest vertical line indicates the threshold selected by default: the highest value to reduce the false positives.\n")
cat('\n')
cat("2) The t-SNE can be colored based on the AUC scores. To highlight the cluster of cells that are more likely of the cell type according to the signatures, we split the cells into cells that passed the assignment threshold (colored in blue), and cells that didn't (colored in gray).\n")
cat('\n')
cat("3) The last TSNE represents the AUC scores values. The darker a cell is the higher AUC score it obtained, i.e. the cell is more enriched in that cell type.")
cat('\n')
cellsUmap    <- scData@reductions$umap@cell.embeddings
minAucThresh <- 0.2
filtered_cells.AUC <- cells.AUC[rowSums(assay(cells.AUC) >= minAucThresh) > 0, ]
maxAucScore  <- rowMax(assay(filtered_cells.AUC))
scoreOrder   <- order(maxAucScore, decreasing = TRUE)
nPlots       <- 50
```

```{r auc-tsne-hist}
#| eval: !expr !is.null(cells.AUC)
#| fig-height: 5
#| fig-width: 5
try({
  AUCell_plotTSNE(tSNE     = cellsUmap,
                  cellsAUC = filtered_cells.AUC[head(scoreOrder, nPlots), ],
                  plots    = c("histogram", "binaryAUC"))
})
```

## Azimuth

```{r azimuth}
#| eval: !expr !is.null(scData@meta.data$Azimuth.celltype.l1)
#| fig-width: 15
#| fig-height: 8
options(ggrepel.max.overlaps = Inf)

cat(sprintf("Analysis was performed using the Azimuth reference map: %s", param$Azimuth))

DimPlot(scData, label = TRUE, group.by = "Azimuth.celltype.l1", repel = TRUE, reduction = "umap", cols = shared_colors) +
  theme(aspect.ratio = 1)
if ('Azimuth.celltype.l2' %in% colnames(scData@meta.data))
  DimPlot(scData, label = TRUE, group.by = "Azimuth.celltype.l2", repel = TRUE, reduction = "umap", cols = shared_colors) +
  theme(aspect.ratio = 1)
if ('Azimuth.celltype.l3' %in% colnames(scData@meta.data))
  DimPlot(scData, label = TRUE, group.by = "Azimuth.celltype.l3", repel = TRUE, reduction = "umap", cols = shared_colors) +
  theme(aspect.ratio = 1)
if ('Azimuth.celltype.l4' %in% colnames(scData@meta.data))
  DimPlot(scData, label = TRUE, group.by = "Azimuth.celltype.l4", repel = TRUE, reduction = "umap", cols = shared_colors) +
  theme(aspect.ratio = 1)
```

## Azimuth Pan-Human

`r if(azimuth_available) "Azimuth Pan-Human provides neural network-based cell type annotation using the Satija lab reference (HUMAN DATASETS ONLY)."`

```{r azimuth-panhuman}
#| eval: !expr azimuth_available
#| results: asis
#| fig-width: 15
#| fig-height: 11
if (azimuth_available) {
  cat("**Reference:** Pan-Human Azimuth neural network model (CloudAzimuth)\n\n")

  scData_az <- scData

  if ("final_level_labels" %in% colnames(scData_az@meta.data)) {
    n_annotated <- sum(!is.na(scData_az@meta.data$final_level_labels))
    cat("**Cells annotated:** ", n_annotated, "\n\n")
  }

  if ("final_level_labels" %in% colnames(scData_az@meta.data)) {
    cat("#### Final Level Cell Types\n\n")
    reduction_to_use <- if ("azimuth_umap" %in% names(scData_az@reductions)) "azimuth_umap" else "umap"
    print(DimPlot(scData_az, group.by = "final_level_labels", label = TRUE, repel = TRUE,
                  reduction = reduction_to_use, cols = shared_colors) +
            ggtitle("Azimuth Pan-Human: Final Level Labels") +
            theme(legend.position = "right") +
            guides(color = guide_legend(override.aes = list(size = 3), ncol = 2)) +
            coord_fixed())
    cat("\n\n")
  }

  if ("azimuth_broad" %in% colnames(scData_az@meta.data)) {
    cat("#### Broad Level Cell Types\n\n")
    reduction_to_use <- if ("azimuth_umap" %in% names(scData_az@reductions)) "azimuth_umap" else "umap"
    print(DimPlot(scData_az, group.by = "azimuth_broad", label = TRUE, repel = TRUE,
                  reduction = reduction_to_use, cols = shared_colors) +
            ggtitle("Azimuth Pan-Human: Broad Level Labels") +
            theme(legend.position = "right") +
            guides(color = guide_legend(override.aes = list(size = 3), ncol = 2)) +
            coord_fixed())
    cat("\n\n")
  }

  if ("azimuth_medium" %in% colnames(scData_az@meta.data)) {
    cat("#### Medium Level Cell Types\n\n")
    reduction_to_use <- if ("azimuth_umap" %in% names(scData_az@reductions)) "azimuth_umap" else "umap"
    print(DimPlot(scData_az, group.by = "azimuth_medium", label = TRUE, repel = TRUE,
                  reduction = reduction_to_use, cols = shared_colors) +
            ggtitle("Azimuth Pan-Human: Medium Level Labels") +
            theme(legend.position = "right") +
            guides(color = guide_legend(override.aes = list(size = 3), ncol = 2)) +
            coord_fixed())
    cat("\n\n")
  }

  if ("azimuth_fine" %in% colnames(scData_az@meta.data)) {
    cat("#### Fine Level Cell Types\n\n")
    reduction_to_use <- if ("azimuth_umap" %in% names(scData_az@reductions)) "azimuth_umap" else "umap"
    print(DimPlot(scData_az, group.by = "azimuth_fine", label = TRUE, repel = TRUE,
                  reduction = reduction_to_use, cols = shared_colors) +
            ggtitle("Azimuth Pan-Human: Fine Level Labels") +
            theme(legend.position = "right") +
            guides(color = guide_legend(override.aes = list(size = 3), ncol = 2)) +
            coord_fixed())
    cat("\n\n")
  }

  if ("final_level_confidence" %in% colnames(scData_az@meta.data)) {
    cat("#### Confidence Scores by Cell\n\n")
    reduction_to_use <- if ("azimuth_umap" %in% names(scData_az@reductions)) "azimuth_umap" else "umap"
    print(FeaturePlot(scData_az, features = "final_level_confidence",
                      reduction = reduction_to_use) +
            ggtitle("Azimuth Pan-Human: Confidence Scores") +
            scale_colour_viridis_c() +
            coord_fixed())
    cat("\n\n")
  }

  if ("final_level_confidence" %in% colnames(scData_az@meta.data)) {
    cat("#### Confidence Score Distribution\n\n")
    p <- ggplot(scData_az@meta.data, aes(x = final_level_confidence)) +
      geom_histogram(bins = 30, fill = "skyblue", color = "white", alpha = 0.8) +
      theme_minimal() +
      labs(title = "Azimuth Pan-Human: Softmax Probabilities",
           x = "Softmax Probability",
           y = "Number of Cells")
    print(p)
    cat("\n\n")
  }

  if ("final_level_confidence" %in% colnames(scData_az@meta.data) &&
      "final_level_labels" %in% colnames(scData_az@meta.data)) {
    cat("#### Confidence Score by Cell Type\n\n")
    p <- ggplot(scData_az@meta.data,
                aes(x = final_level_labels, y = final_level_confidence, fill = final_level_labels)) +
      geom_violin(alpha = 0.8, width = 0.8, scale = "width") +
      geom_boxplot(width = 0.3, outlier.size = 0.5, fill = "white") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            legend.position = "none") +
      scale_fill_manual(values = shared_colors) +
      labs(title = "Azimuth Pan-Human: Confidence by Cell Type",
           x = "Cell Type",
           y = "Confidence Score")
    print(p)
    cat("\n\n")
  }
}
```

## Cell x gene

```{r cellxgene}
#| eval: !expr !is.null(cellxgeneResults)
#| fig-width: 7
#| fig-height: 7
options(ggrepel.max.overlaps = Inf)
n_celltype  <- length(unique(scData$predicted.id.cellxgene.authorlabel))
sample_name <- names(scData)

scData$cellxgene_label_short <- scData$predicted.id.cellxgene.authorlabel |>
  as.character() |>
  gsub(",", "", x = _) |>
  gsub("-positive", "-pos", x = _) |>
  gsub("-negative", "-neg", x = _) |>
  gsub("regulatory", "reg", x = _) |>
  {\(x) ifelse(nchar(x) > 30, paste0(substr(x, 1, 8), "...", substr(x, nchar(x) - 14, nchar(x))), x)}()

p1 <- DimPlot(scData, label = TRUE, repel = TRUE, group.by = 'cellxgene_label_short', cols = shared_colors) +
  ggtitle(paste(sample_name, "- CellxGene - Predicted ID")) +
  guides(color = guide_legend(ncol = 4, byrow = TRUE, override.aes = list(size = 4, alpha = 1)))

legend_plot_p1 <- ggdraw(get_legend(p1)) + theme(plot.margin = margin(0, 0, 0, 0))
main_plot_p1   <- p1 + theme(legend.position = "none", plot.margin = margin(0, 0, 0, 0))
big_plot1 <- (main_plot_p1 / legend_plot_p1) +
  plot_layout(heights = c(3, 1))

p2 <- FeaturePlot(scData, features = "prediction.score.max", cols = c("lightgrey", "dodgerblue"), pt.size = 1, max.cutoff = 'q97', order = TRUE) +
  ggtitle(paste(sample_name, "- mapping score"))

p3 <- VlnPlot(scData, features = 'prediction.score.max', group.by = 'cellxgene_label_short', cols = shared_colors, alpha = 0.3) +
  ggtitle(paste(sample_name, "- prediction.score.max grouped by author's cell label")) +
  theme(legend.position = "none")
```

```{r cellxgene-info}
#| eval: !expr !is.null(cellxgeneResults)
#| results: asis
cat("\n\n* The URL of the reference data is:", param$cellxgeneUrl)

doi_info <- attr(cellxgeneResults, "doi")
if (!is.null(doi_info) && !is.na(doi_info$doi)) {
  cat("\n\n* The DOI of the reference data is:", doi_info$doi)
} else if (!is.null(doi_info)) {
  cat("\n\n* DOI:", doi_info$message)
}
```

```{r cellxgene-plots}
#| eval: !expr !is.null(cellxgeneResults)
#| fig-width: 12
#| fig-height: 10
tryCatch({
  print(big_plot1)
  print(p2 + coord_fixed())
  print(p3)
}, error = function(e) {
  print(big_plot1)
  print(p2)
  print(p3)
})
```

## scType

`r if(sctype_available) "scType provides fully-automated cell-type identification using specific marker combinations (human and mouse supported)."`

```{r sctype-results}
#| eval: !expr sctype_available
#| results: asis
#| fig-width: 14
#| fig-height: 10
if (sctype_available && !is.null(sctype_results)) {
  cat("**Detected tissue:** ", sctype_results$tissue_type, "\n\n")

  scData_sc <- scData

  if ("sctype_classification" %in% colnames(scData_sc@meta.data)) {
    cat("#### scType Cell Type Classification\n\n")
    print(DimPlot(scData_sc, reduction = "umap", label = TRUE, repel = TRUE,
                  group.by = 'sctype_classification', cols = shared_colors) +
            ggtitle(paste("scType Cell Type Annotation -", sctype_results$tissue_type)) +
            theme(legend.position = "right") +
            coord_fixed())
    cat("\n\n")
  }
}
```

:::

# Links

### Interactive Single Cell Explorer

The link below will take you to an interactive Shiny app which can be used to make
production-ready plots and figures. You can also use this app to explore your own markers and create corresponding feature plots.

[Single Cell Explorer](`r {paste0("https://fgcz-shiny.uzh.ch/exploreSC", "/?data=", output$getColumn("SC Cluster Report"), "/scData.qs2")}`){target="_blank"}

### Data files

[clusterInfos.xlsx](clusterInfos.xlsx) -- use this xls file to manually relabel the clusters

[posMarkers](posMarkers.xlsx) -- Excel table with all marker genes

[scData.qs2](scData.qs2) -- final seurat object in [qs2 format](https://github.com/qsbase/qs2)

`r if(sctype_available) "[scType_results.xlsx](scType_results.xlsx) -- scType cell type annotation results"`

# Parameters

### Parameters used

```{r report-parameters}
#| echo: false
```

# Session Info

```{r session-info}
sessionInfo()
```

:::

```{r write-metadata}
#| include: false
annotation_methods <- c(
  if (!is.null(singler.results)) "singler" else NULL,
  if (!is.null(aziResults)) "azimuth" else NULL,
  if (!is.null(cells.AUC)) "AUCell" else NULL,
  if (sctype_available) "scType" else NULL,
  if (!is.null(cellxgeneResults)) "cellxgene" else NULL
)

.write_metadata(
  reportType        = "single-cell-seurat",
  omicsType         = "scrna",
  organism          = if (!is.null(param$refBuild)) param$refBuild else NA,
  nCellsFiltered    = ncol(scData),
  nClusters         = length(unique(scData$seurat_clusters)),
  annotationMethods = annotation_methods,
  files             = list(plots = "plots/", html = "html/")
)
```
