---
params:
  reportTitle: "Analysis"
title: "`r params$reportTitle`"
metadata-files: ["_fgcz_report.yml"]
---

```{r setup}
#| include: false
library(ezRun)
library(DT)
library(tidyverse)

.dt <- function(df, caption = NULL, pageLength = 10L,
                filter = "none", rownames = FALSE, escape = TRUE, ...) {
  DT::datatable(
    data = df, caption = caption, filter = filter,
    rownames = rownames, escape = escape,
    class = "compact stripe",
    options = list(
      pageLength = pageLength,
      scrollX = TRUE,
      dom = "tip",
      columnDefs = list(list(targets = "_all", className = "dt-left"))
    ),
    ...
  ) |> DT::formatStyle(columns = seq_len(ncol(df)),
                        `text-align` = "left",
                        `white-space` = "normal",
                        `word-wrap` = "break-word",
                        `max-width` = "300px")
}

.save_plot <- function(plot, name, width = 6, height = 5, dpi = 300) {
  if (!dir.exists("plots")) dir.create("plots", recursive = TRUE)
  ggplot2::ggsave(file.path("plots", paste0(name, ".png")),
                  plot = plot, width = width, height = height, dpi = dpi, bg = "white")
  ggplot2::ggsave(file.path("plots", paste0(name, ".pdf")),
                  plot = plot, width = width, height = height, device = "pdf", bg = "white")
}
```

::: {.panel-tabset}

# First

<!-- PATTERN 1 — section-level callout: a full-width note that frames or
     caveats the WHOLE sub-tab (methods summary, data source, scope). Just a
     plain ::: callout (3 colons), no columns. -->

::: {.callout-note}
## About this section

Use a full-width callout when the note applies to *everything* in the sub-tab
rather than one figure. Swap `note` for `tip` / `warning` / `important` to
change the colour and icon; add `collapse="true"` to fold it shut by default.
:::

Main content of the tab then runs full-width underneath.

# Second

<!-- PATTERN 2 — figure + side callout, with numbers pulled LIVE from the qmd.
     Compute the values in a hidden chunk, then reference them in the callout
     with inline R expressions so the prose can never drift from the figure.
     Fence colon-counts MUST nest (outer > inner) or the blocks won't close:
       11 colons = .columns wrapper · 10 = each .column · 3 = the .callout -->

```{r second-demo}
#| include: false
demo_df   <- mtcars
demo_n    <- nrow(demo_df)
demo_mean <- round(mean(demo_df$mpg), 1)
demo_plot <- ggplot(demo_df, aes(wt, mpg)) +
  geom_point(size = 2) +
  labs(x = "Weight (1000 lbs)", y = "Fuel economy (mpg)") +
  theme_bw()
```

::::::::::: {.columns}

:::::::::: {.column width="73%"}

```{r second-plot}
demo_plot
```

::::::::::

:::::::::: {.column width="2%"}
::::::::::

:::::::::: {.column width="25%"}

::: {.callout-tip collapse="true"}
## What this shows

Across the **`r demo_n`** cars, mean fuel economy is **`r demo_mean` mpg**.
Both figures are inline R expressions reading the chunk on the left, so the
caption stays in lock-step with the data instead of being hard-typed.
:::

::::::::::

:::::::::::

# Session Info

```{r session-info}
sessionInfo()
```

:::
