---
params:
  reportTitle: "Analysis"
title: "`r params$reportTitle`"
metadata-files: ["_fgcz-report.yml"]
---

```{r setup}
#| include: false
library(ezRun)
library(DT)
library(tidyverse)
```

::: {.panel-tabset}

# First

<!-- PATTERN 1 — section-level callout: a full-width note that frames or
     caveats the WHOLE sub-tab (methods summary, data source, scope). Just a
     plain ::: callout (3 colons), no columns. -->

::: {.callout-note}
## About this section

Use a full-width callout when the note applies to *everything* in the sub-tab
rather than one figure. Swap `note` for `tip` / `warning` / `important` to
change the colour and icon; add `collapse="true"` to fold it shut by default.
:::

Main content of the tab then runs full-width underneath.

# Second

<!-- PATTERN 2 — OPTIONAL plot + side callout (flush, same height).
     RULE: every figure carries a native #| fig-cap caption (below the plot) —
     that alone is the default figure style and obeys the yaml out.width (see
     the Nested tab for a plain example).
     EXTRA (opt-in): a green callout beside the plot for annotation, stretched
     to the plot's height via .fig-row in fgcz.scss. In THIS side-by-side
     layout the plot fills its column (size set by .fig-main in fgcz.scss, not
     out.width). Keep numbers as inline R so the note can't drift from the
     figure. The callout keeps its natural height so collapse="true" works
     (it is NOT locked to the plot height); on narrow screens it drops below
     the plot. Fence colons nest: 5 = .fig-row · 4 = each side · 3 = .callout. -->

```{r second-demo}
#| include: false
demo_df   <- mtcars
demo_n    <- nrow(demo_df)
demo_mean <- round(mean(demo_df$mpg), 1)
demo_plot <- ggplot(demo_df, aes(wt, mpg)) +
  geom_point(size = 2) +
  labs(x = "Weight (1000 lbs)", y = "Fuel economy (mpg)") +
  theme_bw()
```

::::: {.fig-row}

:::: {.fig-main}

```{r second-plot}
#| fig-cap: "Fuel economy versus weight across the mtcars dataset."
demo_plot
```

::::

:::: {.fig-side}

::: {.callout-tip collapse="true"}
## What this shows

Each point is one of the **`r demo_n`** cars; mean fuel economy is
**`r demo_mean` mpg**, and heavier cars are clearly less efficient. This green
box is an optional annotation beside the plot — the figure's own caption sits
underneath as usual.
:::

::::

:::::

# Nested

<!-- PATTERN 3 — nested tabset (card-in-card). The inner ::: {.panel-tabset}
     uses ## sub-headings for its tabs, one level deeper than the # tabs of
     the outer set. With fgcz.scss the inner set renders as a card inside the
     outer card. Chunks call existing functions only — no template-local
     helper defs (DT::datatable for the table, ggplot2 for the figure). -->

::: {.panel-tabset}

## Table

```{r nested-table}
ezRun::ezInteractiveTableRmd(
  head(mtcars[, 1:6], 8),
  title = "Demo table — first rows of mtcars"
)
```

## Figure

```{r nested-figure}
#| fig-cap: "Demo figure — fuel economy by cylinder count."
ggplot(mtcars, aes(factor(cyl), mpg, fill = factor(cyl))) +
  geom_boxplot(show.legend = FALSE) +
  labs(x = "Cylinders", y = "Fuel economy (mpg)") +
  theme_bw()
```

:::

# Three-level

<!-- PATTERN 4 — three levels of nesting (tabset → tabset → tabset). Every
     depth renders as the IDENTICAL folder-tab card (no per-level colour; see
     fgcz.scss) — depth reads from the nested card borders alone.
     Fence colons step DOWN per level so nesting is unambiguous:
     4 colons = level-2 tabset · 3 colons = each level-3 tabset. Tabs at each
     level are just headings: ## for level 2, ### for level 3. -->

:::: {.panel-tabset}

## By dimension

::: {.panel-tabset}

### Sepal

```{r tl-sepal}
#| fig-cap: "Sepal width versus length in the iris dataset, by species."
ggplot(iris, aes(Sepal.Length, Sepal.Width, colour = Species)) +
  geom_point(size = 2) +
  labs(x = "Sepal length (cm)", y = "Sepal width (cm)", colour = "Species") +
  theme_bw()
```

### Petal

```{r tl-petal}
#| fig-cap: "Petal width versus length in the iris dataset, by species."
ggplot(iris, aes(Petal.Length, Petal.Width, colour = Species)) +
  geom_point(size = 2) +
  labs(x = "Petal length (cm)", y = "Petal width (cm)", colour = "Species") +
  theme_bw()
```

:::

## Distributions

::: {.panel-tabset}

### Sepal length

```{r tl-sepal-hist}
#| fig-cap: "Distribution of sepal length across species."
ggplot(iris, aes(Sepal.Length, fill = Species)) +
  geom_histogram(bins = 20, alpha = 0.7, position = "identity") +
  labs(x = "Sepal length (cm)", y = "Count", fill = "Species") +
  theme_bw()
```

### Petal length

```{r tl-petal-hist}
#| fig-cap: "Distribution of petal length across species."
ggplot(iris, aes(Petal.Length, fill = Species)) +
  geom_histogram(bins = 20, alpha = 0.7, position = "identity") +
  labs(x = "Petal length (cm)", y = "Count", fill = "Species") +
  theme_bw()
```

:::

::::

# Five-level

<!-- PATTERN 5 — five levels of nesting (tabset → … → tabset), the deepest
     nesting the template demonstrates. Every depth renders as the IDENTICAL
     folder-tab card (no per-level colour; see fgcz.scss) — depth reads from the
     nested card borders alone. This top tab is level 1; each nested
     ::: {.panel-tabset} steps one level deeper.
     Fence colons step DOWN per level so nesting stays unambiguous:
     6 = level-2 tabset · 5 = level-3 · 4 = level-4 · 3 = level-5. Tab headings
     are ## (L2) · ### (L3) · #### (L4) · ##### (L5). -->

:::::: {.panel-tabset}

## Level 2 · A

```{r fl-l2a}
#| fig-cap: "Level 2 — fuel economy versus weight (mtcars)."
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(size = 2, colour = "#1B4079") +
  labs(x = "Weight (1000 lbs)", y = "Fuel economy (mpg)") +
  theme_bw()
```

## Level 2 · B (go deeper)

::::: {.panel-tabset}

### Level 3 · A

```{r fl-l3a}
#| fig-cap: "Level 3 — fuel economy by cylinder count (mtcars)."
ggplot(mtcars, aes(factor(cyl), mpg, fill = factor(cyl))) +
  geom_boxplot(show.legend = FALSE) +
  labs(x = "Cylinders", y = "Fuel economy (mpg)") +
  theme_bw()
```

### Level 3 · B (go deeper)

:::: {.panel-tabset}

#### Level 4 · A

```{r fl-l4a}
#| fig-cap: "Level 4 — sepal dimensions by species (iris)."
ggplot(iris, aes(Sepal.Length, Sepal.Width, colour = Species)) +
  geom_point(size = 2) +
  labs(x = "Sepal length (cm)", y = "Sepal width (cm)", colour = "Species") +
  theme_bw()
```

#### Level 4 · B (go deeper)

::: {.panel-tabset}

##### Level 5 · A

```{r fl-l5a}
#| fig-cap: "Level 5 — distribution of petal length across species (iris)."
ggplot(iris, aes(Petal.Length, fill = Species)) +
  geom_histogram(bins = 20, alpha = 0.7, position = "identity") +
  labs(x = "Petal length (cm)", y = "Count", fill = "Species") +
  theme_bw()
```

##### Level 5 · B

```{r fl-l5b}
#| fig-cap: "Level 5 — petal dimensions by species (iris)."
ggplot(iris, aes(Petal.Length, Petal.Width, colour = Species)) +
  geom_point(size = 2) +
  labs(x = "Petal length (cm)", y = "Petal width (cm)", colour = "Species") +
  theme_bw()
```

:::

::::

:::::

::::::

# Session Info

```{r session-info}
sessionInfo()
```

:::
