---
title: "Reference Genome Folders"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reference Genome Folders}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r, warning=FALSE,message=FALSE,split=TRUE}
library(ezRun)
```

<!-- see: http://stackoverflow.com/questions/19372260/how-to-get-rstudio-to-automatically-compile-r-markdown-vignettes -->
<!-- use knit button or devtools::build_vignettes() -->
<!-- do not use: devtools::install(build_vignettes = TRUE) -->

# Introduction

`ezRun` makes use of reference genomes stored in a canonical directory structure. Reference genomes can be built from
a gtf file containing the gene annotations and a fasta file holding the genome sequence. With the file-based representation
the reference genomes are available to external tools like read aligners and others.

The commands to generate a reference are:
```{r,split=TRUE}
refBuild = "Saccharomyces_cerevisiae/Ensembl/EF4/Annotation/Version-2013-03-18"
genomesRoot = "./refExample"
system("rm -rf refExample")
param = ezParam(list(refBuild=refBuild, genomesRoot=genomesRoot))
gtf = system.file("extdata/genes.gtf", package="ezRun", mustWork = TRUE)
fp = system.file("extdata/genome.fa", package="ezRun", mustWork = TRUE)
buildRefDir(param$ezRef, fp, gtf)
buildIgvGenome(param$ezRef)
seqAnno = writeAnnotationFromGtf(param=param)# featureFile=param$ezRef["refFeatureFile"], featAnnoFile=myRef["refAnnotationFile"])
```


# Structure of a reference genome folder

The reference genome folder is inspired by illumina's iGenome folders but has differences with respect to handling
different annotation versions.

A reference genome build must be in a folder with the path

```
<species> / <provider> / <build name>
```

Build names should be unique within an installation. Within each build the sub-directories are:
```
└── <provider>
    └── <genome_build>
        ├── Annotation
        │   └── Genes (is a link to the latest version)
        |     └──  Version-<date>
        |           └── Genes
        |                └── genes.gtf
        |                └── features.gtf
        └── Sequence
            └── WholeGenomeFasta
                 └── genome.fa
                 └── genome.fa.fai
                 └── genome.dict
```

<!-- currently not implemented
├── Chromosomes
                └── <chromosome/contig names>.fa
-->

Indices of the various aligners will be built on demand and also placed in the folder structure.



# Genome annotation files

Genome annotation files that define features (e.g. gene loci) on the genome must be in gtf file format. If the input file is in gff format, it can be converted to gtf with the `gffread` utility from the cufflinks suite
```
gffread -E -T -o genes.gtf genes.gff 
```

By convention the content of the feature files in the reference folders is:  

* genes.gtf -- covers all protein coding loci in the reference (i.e. all loci where transcripts with a poly-A are generated); designed for poly-A RNA-seq  
* transcribed.gtf -- covers all genome features that are transcribed, including non-coding RNAs like lincRNAs, micro-RNAs, ...;
designed for whole transcriptome appraoches (e.g. ribo-depletion)  
* features.gtf -- all annotated features in the genome, including pseudogenes, etc.; meant for whole genome exploratory analyses. Note: the included pseudogenes may give rise to multi-mapping reads in RNA-seq approaches because of the inclusion of loci that are never transcribed.  

# Feature annotation file
Every file `<name>.gtf` must be accompanied by a file `<name>_annotation.txt`. This annotation file must have rownames that match the attribute `transcript_id`
in the .gtf file. This annotation file will probably need additional enhancements. It may can contain annotations like  

* gene_name  
* description  
* GO categories  


# Processing and checks of genome assemblies and gene annotation files

The buildRefDir function does:  

* from the fasta file remove anything everything after the first space on the sequence identifier lines  
* remove contigs where the contig-name includeds the word PATCH  
* verify that the gene annotation file uses the same chromosome names as the fasta file  
* remove chromosomes from the annotation files that are not present in the fasta file  
* create a fasta index file with  
```
samtools faidx genome.fa
```
* create the genome dict file with
```
java -jar picard.jar CreateSequenceDictionary R=genome.fa O=genome.dict
```
