## ATAC (chromatin accessibility) {.tabset}

```{r atac-prep, include=FALSE}
library(Signac)
n_peaks <- nrow(scMultiData[["ATAC"]])
n_cells_atac <- ncol(scMultiData)
DefaultAssay(scMultiData) <- "ATAC"
if (!"nucleosome_signal" %in% colnames(scMultiData@meta.data)) {
  scMultiData <- tryCatch(
    NucleosomeSignal(scMultiData, assay = "ATAC"),
    error = function(e) scMultiData
  )
}
if (!"TSS.enrichment" %in% colnames(scMultiData@meta.data)) {
  scMultiData <- tryCatch(
    TSSEnrichment(scMultiData, assay = "ATAC", fast = TRUE),
    error = function(e) scMultiData
  )
}
DefaultAssay(scMultiData) <- "RNA"
has_ga <- "GeneActivity" %in% Assays(scMultiData)
```

### UMAP

```{r atac-umaps, echo=FALSE, fig.width=14, fig.height=7}
plots <- list()
if ("atac.umap" %in% names(scMultiData@reductions) && !is.null(cluster_col)) {
  plots[["clusters"]] <- DimPlot(scMultiData, reduction = "atac.umap",
                                 group.by = cluster_col, raster = FALSE,
                                 label = TRUE, repel = TRUE, order = TRUE,
                                 cols = cluster_pal) +
    theme(aspect.ratio = 1) + ggtitle(paste0("ATAC UMAP - clusters (", cluster_col, ")"))
}
if ("atac.umap" %in% names(scMultiData@reductions) && !is.null(celltype_col)) {
  plots[["celltype"]] <- DimPlot(scMultiData, reduction = "atac.umap",
                                 group.by = celltype_col, raster = FALSE,
                                 label = TRUE, repel = TRUE, order = TRUE) +
    theme(aspect.ratio = 1) + ggtitle(paste0("ATAC UMAP - cell type (", celltype_col, ")"))
}
if (length(plots) >= 2) {
  print(patchwork::wrap_plots(plots, ncol = 2))
} else if (length(plots) == 1) {
  print(plots[[1]])
}
```

### Gene activity FeaturePlots

```{r atac-gene-activity, echo=FALSE, fig.width=12, fig.height=10}
if (has_ga) {
  scMultiData_ga <- scMultiData
  DefaultAssay(scMultiData_ga) <- "GeneActivity"
  ga_mat <- tryCatch(
    as.matrix(GetAssayData(scMultiData_ga, assay = "GeneActivity", layer = "data")),
    error = function(e) NULL
  )
  if (!is.null(ga_mat)) {
    ga_means <- rowMeans(ga_mat)
    top_ga <- names(sort(ga_means, decreasing = TRUE))[1:min(8L, length(ga_means))]
    red <- if ("atac.umap" %in% names(scMultiData_ga@reductions)) "atac.umap" else rna_umap
    p <- FeaturePlot(scMultiData_ga, features = top_ga, reduction = red,
                     raster = FALSE, ncol = 4) &
      theme(aspect.ratio = 1) &
      scale_color_viridis_c(option = "D")
    print(p)
  }
} else {
  cat("GeneActivity assay not available - genome annotation could not be resolved.\n")
}
```

### Fragment counts

```{r atac-frag, echo=FALSE, fig.width=8, fig.height=5}
print(VlnPlot(scMultiData, features = "nCount_ATAC", group.by = cluster_col %||% "Sample",
              pt.size = 0, cols = cluster_pal) + scale_y_log10() + NoLegend())
```

### TSS enrichment

```{r atac-tss, echo=FALSE, fig.width=8, fig.height=5}
if ("TSS.enrichment" %in% colnames(scMultiData@meta.data)) {
  print(VlnPlot(scMultiData, features = "TSS.enrichment", group.by = cluster_col %||% "Sample",
                pt.size = 0, cols = cluster_pal) + NoLegend())
} else {
  cat("TSS enrichment unavailable - genome annotation could not be resolved.\n")
}
```

### Nucleosome signal

```{r atac-nuc, echo=FALSE, fig.width=8, fig.height=5}
if ("nucleosome_signal" %in% colnames(scMultiData@meta.data)) {
  print(VlnPlot(scMultiData, features = "nucleosome_signal", group.by = cluster_col %||% "Sample",
                pt.size = 0, cols = cluster_pal) + NoLegend())
} else {
  cat("Nucleosome signal unavailable - genome annotation could not be resolved.\n")
}
```

### Summary

**What this tab shows.** Per-cell **chromatin accessibility (ATAC)**
quantification from a CellRanger ARC multiome sample. ATAC measures
which regions of the genome are open and accessible to transcription
machinery — peaks are called by CellRanger ARC across the dataset, and
each cell gets a count for how many ATAC fragments fell into each peak.
We then build a **Signac ChromatinAssay**, run TF-IDF normalization +
SVD (LSI) for dimensionality reduction, and compute an ATAC UMAP. A
**GeneActivity** assay aggregates accessibility within ±2 kb of each
gene's TSS as a proxy for "this gene is regulatorily active in this
cell" (orthogonal to RNA, which measures transcript abundance after the
fact).

**Why look at this tab.** Open chromatin marks cells that are *poised*
to express a program, not just cells already expressing it — useful for
catching transitional states (e.g. progenitors priming a lineage),
finding regulatory elements (enhancers / promoters) that distinguish
clusters, and validating cell-type calls with an independent modality.
The Gene activity FeaturePlots overlay key marker genes' accessibility
on the RNA UMAP so you can compare "is the gene's locus open?" vs "is
the gene transcribed?" side by side. The Fragment counts / TSS
enrichment / Nucleosome signal subtabs are standard ATAC QC: low TSS
enrichment or high nucleosome signal both flag low-quality cells whose
ATAC libraries are dominated by noise.

**Panel:** `r format(n_peaks, big.mark = ",")` peaks called over `r format(n_cells_atac, big.mark = ",")` cells.
**Gene activity assay:** `r if (has_ga) paste0("**", format(nrow(scMultiData[["GeneActivity"]]), big.mark = ","), "** genes (Signac::GeneActivity over peak-to-gene mapping using the EnsDb annotation matched to the refBuild)") else "_not built — typically because the EnsDb package matching this refBuild was not available at runtime, so peak-to-gene mapping was skipped. The Gene activity FeaturePlots subtab will fall back to a placeholder._"`.

**UMAP layout in plots:** `atac.umap` — computed on the ATAC LSI reduction (TF-IDF + SVD on the peak matrix). This is independent of the RNA UMAP; the same cells will sit in different positions in the two layouts.

**Clustering column used in plots:** `r if (!is.null(cluster_col)) paste0("**", cluster_col, "** — inherited from the RNA modality (no ATAC-specific clustering is computed). For multi-modal cluster IDs that fold ATAC into the integration, see the Multi-modal integration (WNN) tab.") else "_no cluster column found on the object._"`.
