## WNN {.tabset}

```{r wnn-prep, include=FALSE}
red_used <- intersect(c("pca", "adt.pca", "lsi"), names(scMultiData@reductions))
weight_cols <- grep("\\.weight$", colnames(scMultiData@meta.data), value = TRUE)
wnn_resolution <- modalities$wnnResolution %||% scMultiData@misc$wnn_resolution %||% 0.5
wnn_cluster_col <- scMultiData@misc$wnn_cluster_col %||% paste0("wsnn_res.", wnn_resolution)
if (!wnn_cluster_col %in% colnames(scMultiData@meta.data)) {
  wsnn_hits <- grep("^wsnn_res\\.", colnames(scMultiData@meta.data), value = TRUE)
  wnn_cluster_col <- if (length(wsnn_hits) > 0) wsnn_hits[1] else cluster_col
}
```

*Clustering used:* **`r wnn_cluster_col`** — Leiden on the weighted SNN graph at resolution `r wnn_resolution`. Integrated reductions: **`r paste(red_used, collapse = ", ")`**.

### UMAP

```{r wnn-umaps, echo=FALSE, fig.width=14, fig.height=7}
plots <- list()
if ("wnn.umap" %in% names(scMultiData@reductions) && !is.null(wnn_cluster_col)) {
  plots[["clusters"]] <- DimPlot(scMultiData, reduction = "wnn.umap",
                                 group.by = wnn_cluster_col, raster = FALSE,
                                 label = TRUE, repel = TRUE, order = TRUE,
                                 cols = cluster_pal) +
    theme(aspect.ratio = 1) + ggtitle(paste0("WNN UMAP - clusters (", wnn_cluster_col, ")"))
}
if ("wnn.umap" %in% names(scMultiData@reductions) && !is.null(celltype_col)) {
  plots[["celltype"]] <- DimPlot(scMultiData, reduction = "wnn.umap",
                                 group.by = celltype_col, raster = FALSE,
                                 label = TRUE, repel = TRUE, order = TRUE) +
    theme(aspect.ratio = 1) + ggtitle(paste0("WNN UMAP - cell type (", celltype_col, ")"))
}
if (length(plots) >= 2) {
  print(patchwork::wrap_plots(plots, ncol = 2))
} else if (length(plots) == 1) {
  print(plots[[1]])
}
```

### Cluster DEGs {.tabset}

Wilcoxon `FindAllMarkers` per modality, grouped by **`r wnn_cluster_col`** (`min.pct = 0.25`, `logfc.threshold = 0.25`, only positive markers). One subtab per modality available on the object: top markers as DotPlot + interactive sortable / filterable / Excel-exportable table. RNA uses the `RNA` assay; ADT uses the `ADT` assay; ATAC uses the `GeneActivity` assay (gene-level chromatin accessibility derived from the fragments).

```{r wnn-deg-compute, include=FALSE}
deg_assays <- c()
if ("RNA" %in% Assays(scMultiData)) deg_assays <- c(deg_assays, RNA = "RNA")
if ("ADT" %in% Assays(scMultiData)) deg_assays <- c(deg_assays, ADT = "ADT")
if ("GeneActivity" %in% Assays(scMultiData)) deg_assays <- c(deg_assays, ATAC = "GeneActivity")

deg_results <- list()
top_features  <- list()
if (!is.null(wnn_cluster_col) && wnn_cluster_col %in% colnames(scMultiData@meta.data)) {
  for (label in names(deg_assays)) {
    assay_name <- deg_assays[[label]]
    scMultiData_wm <- scMultiData
    DefaultAssay(scMultiData_wm) <- assay_name
    Idents(scMultiData_wm) <- wnn_cluster_col
    n_features <- nrow(scMultiData_wm[[assay_name]])
    if (n_features < 2L) next
    mk <- tryCatch(
      Seurat::FindAllMarkers(scMultiData_wm, assay = assay_name, only.pos = TRUE,
                             min.pct = 0.25, logfc.threshold = 0.25,
                             verbose = FALSE),
      error = function(e) { message("FindAllMarkers ", label, " failed: ",
                                    conditionMessage(e)); NULL }
    )
    if (!is.null(mk) && nrow(mk) > 0) {
      mk <- mk[, c("gene", "cluster", "avg_log2FC", "p_val_adj", "pct.1", "pct.2")]
      mk$avg_log2FC <- round(mk$avg_log2FC, 3)
      mk$pct.1 <- round(mk$pct.1, 3); mk$pct.2 <- round(mk$pct.2, 3)
      deg_results[[label]] <- mk
      # Reverse order so coord_flip in the subtab places cluster 0 at the TOP
      # (DotPlot + coord_flip puts features[1] at the bottom of the y-axis).
      top_features[[label]] <- mk |>
        dplyr::group_by(cluster) |>
        dplyr::slice_max(order_by = avg_log2FC, n = 5) |>
        dplyr::pull(gene) |>
        unique() |>
        rev()
    }
  }
}
```

```{r wnn-deg-children, echo=FALSE, results='asis'}
# Render each modality as a knit_child so the embedded ezInteractiveTableRmd
# DT widget loads its CSS/JS deps correctly (asis + print() loses them).
if (length(deg_results) == 0) {
  cat("No WNN cluster markers detected.\n\n")
} else {
  subtab_path <- system.file("templates/_scMultiOmics_wnn_deg_subtab.Rmd",
                             package = "ezRun")
  for (lbl in names(deg_results)) {
    label      <- lbl
    assay_name <- deg_assays[[lbl]]
    mk         <- deg_results[[lbl]]
    top_genes  <- top_features[[lbl]]
    cat(sprintf("\n#### %s\n\n", label))
    cat(knitr::knit_child(subtab_path, envir = environment(), quiet = TRUE),
        sep = "\n")
  }
}
```

### Per-modality weights (Violin)

```{r wnn-weights-vln, echo=FALSE, fig.width=10, fig.height=5}
if (length(weight_cols) > 0 && !is.null(wnn_cluster_col)) {
  print(VlnPlot(scMultiData, features = weight_cols,
                group.by = wnn_cluster_col, pt.size = 0,
                cols = cluster_pal,
                ncol = length(weight_cols)))
} else {
  cat("No modality weight columns - WNN may not have run.\n")
}
```

### Per-modality weights (FeaturePlot)

```{r wnn-weights-feat, echo=FALSE, fig.width=12, fig.height=4 * ceiling(max(1, length(weight_cols))/3)}
if (length(weight_cols) > 0 && "wnn.umap" %in% names(scMultiData@reductions)) {
  p <- FeaturePlot(scMultiData, features = weight_cols,
                   reduction = "wnn.umap", raster = FALSE,
                   ncol = min(3L, length(weight_cols))) &
    theme(aspect.ratio = 1) &
    scale_color_viridis_c(option = "C")
  print(p)
} else {
  cat("FeaturePlot unavailable - WNN UMAP missing.\n")
}
```


### Modality correlation {.tabset}

```{r wnn-corr-prep, include=FALSE}
.wnn_make_group_palette <- function(groups) {
  lvls <- unique(groups)
  if (all(!is.na(suppressWarnings(as.integer(lvls))))) lvls <- as.character(sort(as.integer(lvls)))
  else lvls <- sort(lvls)
  setNames(unname(pals::polychrome())[seq_along(lvls)], lvls)
}

.wnn_format_p <- function(p) {
  if (is.na(p)) "NA" else if (p < 0.001) "< 0.001" else sprintf("= %.3f", p)
}

.wnn_plot_stats <- function(data) {
  dplyr::bind_rows(lapply(split(data, data$plotFeature), function(df) {
    x <- df$rnaMean; y <- df$sourceMean
    keep <- is.finite(x) & is.finite(y); x <- x[keep]; y <- y[keep]
    if (length(x) < 3 || length(unique(x)) < 2 || length(unique(y)) < 2) {
      rho <- pv <- r2 <- NA_real_
    } else {
      ct <- suppressWarnings(cor.test(x, y, method = "spearman", exact = FALSE))
      rho <- unname(ct$estimate); pv <- ct$p.value; r2 <- summary(lm(y ~ x))$r.squared
    }
    data.frame(plotFeature = df$plotFeature[1],
               statLabel = paste0("rho = ", ifelse(is.na(rho), "NA", sprintf("%.2f", rho)),
                                  "\np ", .wnn_format_p(pv),
                                  "\nR^2 = ", ifelse(is.na(r2), "NA", sprintf("%.2f", r2))),
               stringsAsFactors = FALSE)
  }))
}

.wnn_pair_scatter <- function(plotData, yLabel, ncol = 5, statSide = "left", aspectRatio = NULL) {
  pal <- .wnn_make_group_palette(plotData$group)
  plotData$group <- factor(plotData$group, levels = names(pal))
  facetLevels <- if (is.factor(plotData$plotFeature)) levels(plotData$plotFeature) else unique(plotData$plotFeature)
  plotData$plotFeature <- factor(plotData$plotFeature, levels = facetLevels)
  stats <- .wnn_plot_stats(plotData)
  stats$plotFeature <- factor(stats$plotFeature, levels = facetLevels)
  statX <- if (identical(statSide, "right")) Inf else -Inf
  statH <- if (identical(statSide, "right")) 1.05 else -0.05
  p <- ggplot(plotData, aes(x = rnaMean, y = sourceMean)) +
    geom_smooth(method = "lm", se = FALSE, color = "grey35", linewidth = 0.4) +
    geom_point(aes(color = group), size = 1.8) +
    geom_text(data = stats, aes(label = statLabel),
              x = statX, y = Inf, hjust = statH, vjust = 1.1, size = 2.7, inherit.aes = FALSE) +
    facet_wrap(~ plotFeature, scales = "free", ncol = ncol) +
    scale_color_manual(values = pal, drop = FALSE) +
    guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
    labs(x = "Mean RNA expression", y = yLabel, color = "Cluster") +
    theme_bw() + theme(legend.position = "top")
  if (!is.null(aspectRatio)) p <- p + theme(aspect.ratio = aspectRatio)
  p
}

wnnCorrGroupCol <- if (!is.null(wnn_cluster_col) && wnn_cluster_col %in% colnames(scMultiData@meta.data)) wnn_cluster_col else NULL
wnnCorrAvail <- "GeneActivity" %in% Seurat::Assays(scMultiData) &&
                "RNA" %in% Seurat::Assays(scMultiData) &&
                !is.null(wnnCorrGroupCol)
wnnCorrPseudobulk <- data.frame()
wnnCorrCorr <- data.frame()
wnnGlobalPseudobulk <- data.frame()

if (wnnCorrAvail) {
  gaFeats <- rownames(scMultiData[["GeneActivity"]])
  rnaFeats <- rownames(scMultiData[["RNA"]])
  matchIdx <- match(gaFeats, rnaFeats)
  fmap <- data.frame(
    sourceFeature = gaFeats,
    sourceDisplay = gaFeats,
    rnaFeature    = rnaFeats[matchIdx],
    stringsAsFactors = FALSE
  )
  fmap <- fmap[!is.na(fmap$rnaFeature), , drop = FALSE]

  if (nrow(fmap) > 0) {
    avgGa <- as.matrix(Seurat::AverageExpression(scMultiData, assays = "GeneActivity",
                          group.by = wnnCorrGroupCol, layer = "data", verbose = FALSE)$GeneActivity)
    avgRna <- as.matrix(Seurat::AverageExpression(scMultiData, assays = "RNA",
                          group.by = wnnCorrGroupCol, layer = "data", verbose = FALSE)$RNA)
    grps <- colnames(avgGa)
    wnnCorrPseudobulk <- dplyr::bind_rows(lapply(seq_len(nrow(fmap)), function(i) {
      sf <- fmap$sourceFeature[i]; sd <- fmap$sourceDisplay[i]; rf <- fmap$rnaFeature[i]
      data.frame(sourceFeature = sf, sourceDisplay = sd, rnaFeature = rf, group = grps,
                 sourceMean = if (sf %in% rownames(avgGa)) avgGa[sf, ] else NA_real_,
                 rnaMean    = if (rf %in% rownames(avgRna)) avgRna[rf, ] else NA_real_,
                 stringsAsFactors = FALSE)
    }))
    wnnGlobalPseudobulk <- wnnCorrPseudobulk
    wnnCorrCorr <- dplyr::bind_rows(lapply(
      split(wnnCorrPseudobulk, paste(wnnCorrPseudobulk$sourceFeature, wnnCorrPseudobulk$rnaFeature)),
      function(df) {
        x <- df$rnaMean; y <- df$sourceMean
        keep <- is.finite(x) & is.finite(y); x <- x[keep]; y <- y[keep]
        rho <- if (length(x) >= 3 && length(unique(x)) > 1 && length(unique(y)) > 1)
          suppressWarnings(stats::cor(x, y, method = "spearman")) else NA_real_
        data.frame(sourceFeature = df$sourceFeature[1], sourceDisplay = df$sourceDisplay[1],
                   rnaFeature = df$rnaFeature[1], spearmanPseudobulk = rho,
                   stringsAsFactors = FALSE)
      }
    ))
  } else {
    wnnCorrAvail <- FALSE
  }
}
```

#### Multimodal Correlation overview

```{r wnn-corr-overview-barplot, echo=FALSE, fig.width=10, fig.height=6}
if (!wnnCorrAvail || nrow(wnnCorrCorr) == 0) {
  cat("No ATAC/RNA correlations available.\n")
} else {
  pd <- wnnCorrCorr[!is.na(wnnCorrCorr$spearmanPseudobulk), ]
  topC <- pd[order(pd$spearmanPseudobulk, decreasing = TRUE)[seq_len(min(20L, nrow(pd)))], ]
  topD <- pd[order(pd$spearmanPseudobulk)[seq_len(min(20L, nrow(pd)))], ]
  topData <- unique(rbind(topC, topD))
  print(ggplot(topData, aes(x = reorder(sourceDisplay, spearmanPseudobulk), y = spearmanPseudobulk)) +
    geom_col(fill = "#2C7FB8") +
    coord_flip() +
    labs(x = NULL, y = "Pseudobulk Spearman ρ") +
    theme_bw())
}
```

```{r wnn-corr-overview-density, echo=FALSE, fig.width=7.2, fig.height=4.8}
if (wnnCorrAvail && nrow(wnnGlobalPseudobulk) > 0) {
  pd <- wnnGlobalPseudobulk
  pd$logRna <- log1p(pd$rnaMean)
  pd$logSrc <- log1p(pd$sourceMean)
  pd <- pd[!(pd$logRna == 0 & pd$logSrc == 0) & is.finite(pd$logRna) & is.finite(pd$logSrc), ]
  if (nrow(pd) > 0) {
    x <- pd$logRna; y <- pd$logSrc
    ct <- suppressWarnings(cor.test(x, y, method = "spearman", exact = FALSE))
    r2 <- summary(lm(y ~ x))$r.squared
    statLbl <- paste0("rho = ", sprintf("%.2f", unname(ct$estimate)),
                      "\np ", .wnn_format_p(ct$p.value),
                      "\nR^2 = ", sprintf("%.2f", r2))
    print(ggplot(pd, aes(x = logRna, y = logSrc)) +
      geom_bin2d(bins = 85) +
      scale_fill_viridis_c(option = "C", trans = "log10") +
      geom_smooth(color = "black", method = "lm", se = FALSE, linewidth = 0.5) +
      annotate("text", x = Inf, y = -Inf, label = statLbl,
               hjust = 1.05, vjust = -0.1, size = 3) +
      labs(x = "log1p mean RNA expression", y = "log1p mean GeneActivity signal",
           fill = "Point density") +
      theme_bw() + theme(legend.position = "bottom"))
  }
}
```

```{r wnn-corr-overview-table, echo=FALSE}
if (wnnCorrAvail && nrow(wnnCorrCorr) > 0) {
  td <- wnnCorrCorr[order(abs(wnnCorrCorr$spearmanPseudobulk), decreasing = TRUE, na.last = TRUE), ]
  td$spearmanPseudobulk <- round(td$spearmanPseudobulk, 3)
  DT::datatable(td, rownames = FALSE, filter = "top", extensions = "Buttons",
                options = list(dom = "Bfrtip", buttons = c("copy", "csv", "excel"), pageLength = 25))
}
```

#### Top concordant markers

```{r wnn-corr-concordant-scatter, echo=FALSE, fig.width=12, fig.height=8}
if (!wnnCorrAvail || nrow(wnnCorrCorr) == 0) {
  cat("No ATAC/RNA correlations available.\n")
} else {
  topPairs <- wnnCorrCorr[!is.na(wnnCorrCorr$spearmanPseudobulk), ]
  topPairs <- head(topPairs[order(topPairs$spearmanPseudobulk, decreasing = TRUE), ], 10L)
  pd <- wnnCorrPseudobulk[
    paste(wnnCorrPseudobulk$sourceFeature, wnnCorrPseudobulk$rnaFeature) %in%
    paste(topPairs$sourceFeature, topPairs$rnaFeature), ]
  pd$plotFeature <- factor(pd$sourceDisplay, levels = unique(topPairs$sourceDisplay))
  if (nrow(pd) > 0) print(.wnn_pair_scatter(pd, "Mean GeneActivity signal", ncol = 5, statSide = "left", aspectRatio = 1))
}
```

```{r wnn-corr-concordant-table, echo=FALSE}
if (wnnCorrAvail && nrow(wnnCorrCorr) > 0) {
  td <- wnnCorrCorr[!is.na(wnnCorrCorr$spearmanPseudobulk), ]
  td <- head(td[order(td$spearmanPseudobulk, decreasing = TRUE), ], 10L)
  td$spearmanPseudobulk <- round(td$spearmanPseudobulk, 3)
  DT::datatable(td, rownames = FALSE, filter = "top", extensions = "Buttons",
                options = list(dom = "Bfrtip", buttons = c("copy", "csv", "excel"), pageLength = 10))
}
```

#### Top discordant markers

```{r wnn-corr-discordant-scatter, echo=FALSE, fig.width=12, fig.height=8}
if (!wnnCorrAvail || nrow(wnnCorrCorr) == 0) {
  cat("No ATAC/RNA correlations available.\n")
} else {
  topPairs <- wnnCorrCorr[!is.na(wnnCorrCorr$spearmanPseudobulk), ]
  topPairs <- head(topPairs[order(topPairs$spearmanPseudobulk), ], 10L)
  pd <- wnnCorrPseudobulk[
    paste(wnnCorrPseudobulk$sourceFeature, wnnCorrPseudobulk$rnaFeature) %in%
    paste(topPairs$sourceFeature, topPairs$rnaFeature), ]
  pd$plotFeature <- factor(pd$sourceDisplay, levels = unique(topPairs$sourceDisplay))
  if (nrow(pd) > 0) print(.wnn_pair_scatter(pd, "Mean GeneActivity signal", ncol = 5, statSide = "right", aspectRatio = 1))
}
```

```{r wnn-corr-discordant-table, echo=FALSE}
if (wnnCorrAvail && nrow(wnnCorrCorr) > 0) {
  td <- wnnCorrCorr[!is.na(wnnCorrCorr$spearmanPseudobulk), ]
  td <- head(td[order(td$spearmanPseudobulk), ], 10L)
  td$spearmanPseudobulk <- round(td$spearmanPseudobulk, 3)
  DT::datatable(td, rownames = FALSE, filter = "top", extensions = "Buttons",
                options = list(dom = "Bfrtip", buttons = c("copy", "csv", "excel"),
                               pageLength = 10, ordering = FALSE))
}
```


### Summary

Integrated reductions: **`r paste(red_used, collapse = ", ")`**. Resolution: **`r wnn_resolution`**. Cluster column: **`r wnn_cluster_col`**.
