## ADT {.tabset}

```{r adt-prep, include=FALSE}
n_adt   <- nrow(scMultiData[["ADT"]])
adt_features <- rownames(scMultiData[["ADT"]])
adt_counts_mat <- tryCatch(
  as.matrix(GetAssayData(scMultiData, assay = "ADT", layer = "counts")),
  error = function(e) NULL
)
mean_umi   <- if (is.matrix(adt_counts_mat)) round(mean(colSums(adt_counts_mat)), 1) else NA
median_umi <- if (is.matrix(adt_counts_mat)) round(median(colSums(adt_counts_mat)), 1) else NA
adt_norm <- if (!is.null(param$adtNorm)) param$adtNorm else "ADTnorm"

# Top-N ADT per group, ordered for a clean diagonal after coord_flip.
# (1) drop groups with < min_cells (rare Azimuth labels add noise + bloat)
# (2) z-score mean expression per row → group-specific markers
# (3) per group take top n
# (4) order final union by each marker's BEST group → diagonal pattern even when
#     groups share markers (e.g. CD3 across all T-cell subsets)
# Isotype controls excluded.
adt_top_per_group <- function(obj, group_col, n = 3L, min_cells = 30L) {
  groups <- obj[[group_col, drop = TRUE]]
  group_counts <- table(groups, useNA = "no")
  keep <- names(group_counts)[group_counts >= min_cells]
  if (length(keep) < 2L) keep <- names(group_counts)  # fallback
  avg <- Seurat::AverageExpression(obj, assays = "ADT", group.by = group_col,
                                   layer = "data", verbose = FALSE)$ADT
  # Seurat v5 may return a data.frame; coerce so t() / scale() work.
  avg <- as.matrix(avg)
  # Seurat replaces underscores in idents with dashes for column names → match either.
  keep_seurat <- gsub("_", "-", keep)
  col_keep <- colnames(avg) %in% keep | colnames(avg) %in% keep_seurat
  if (any(col_keep)) avg <- avg[, col_keep, drop = FALSE]
  avg <- avg[!grepl("(?i)isotype", rownames(avg)), , drop = FALSE]
  if (ncol(avg) < 2L || nrow(avg) == 0L) {
    return(rownames(avg)[seq_len(min(n, nrow(avg)))])
  }
  z <- t(scale(t(avg)))
  z[!is.finite(z)] <- 0
  picked <- unique(unlist(lapply(seq_len(ncol(z)), function(j) {
    names(sort(z[, j], decreasing = TRUE))[seq_len(min(n, nrow(z)))]
  })))
  z_pick <- z[picked, , drop = FALSE]
  best_group <- apply(z_pick, 1, which.max)
  best_z     <- apply(z_pick, 1, max)
  ord <- order(best_group, -best_z)
  rev(picked[ord])  # coord_flip reverses y-axis
}
```

*Clustering used:* **`r if (!is.null(cluster_col)) cluster_col else "_none_"`** — RNA-derived (no ADT-specific clustering computed). UMAP layout is from the ADT-only PCA (`adt.pca`).

### UMAP

```{r adt-umaps, echo=FALSE, fig.width=14, fig.height=7}
plots <- list()
if ("adt.umap" %in% names(scMultiData@reductions) && !is.null(cluster_col)) {
  plots[["clusters"]] <- DimPlot(scMultiData, reduction = "adt.umap",
                                 group.by = cluster_col, raster = FALSE,
                                 label = TRUE, repel = TRUE, order = TRUE,
                                 cols = unname(pals::polychrome())) +
    theme(aspect.ratio = 1) + ggtitle(paste0("ADT UMAP - clusters (", cluster_col, ")"))
}
if ("adt.umap" %in% names(scMultiData@reductions) && !is.null(celltype_col)) {
  plots[["celltype"]] <- DimPlot(scMultiData, reduction = "adt.umap",
                                 group.by = celltype_col, raster = FALSE,
                                 label = TRUE, repel = TRUE, order = TRUE) +
    theme(aspect.ratio = 1) + ggtitle(paste0("ADT UMAP - cell type (", celltype_col, ")"))
}
if (length(plots) >= 2) {
  print(patchwork::wrap_plots(plots, ncol = 2))
} else if (length(plots) == 1) {
  print(plots[[1]])
}
```

### DotPlot - by cluster

DotPlot of normalized ADT signal (`r adt_norm`) — top 3 markers per cluster (z-scored across clusters; isotypes excluded). Colour = viridis. Grouped by `r if (!is.null(cluster_col)) sprintf("**%s**", cluster_col) else "_no clustering column_"`.

```{r adt-dotplot-cluster, echo=FALSE, fig.width=12, fig.height=14}
if (!is.null(cluster_col)) {
  scMultiData_dp <- scMultiData
  Idents(scMultiData_dp) <- cluster_col
  DefaultAssay(scMultiData_dp) <- "ADT"
  feats <- adt_top_per_group(scMultiData_dp, cluster_col, n = 3L)
  p <- DotPlot(scMultiData_dp, features = feats, assay = "ADT") +
    coord_flip() +
    scale_color_viridis_c(option = "D") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(p)
}
```

### DotPlot - by cell type

DotPlot of normalized ADT signal (`r adt_norm`) — top 3 markers per cell type (z-scored; isotypes excluded; cell types with < 30 cells dropped to keep the panel readable). Grouped by `r if (!is.null(celltype_col)) sprintf("**%s**", celltype_col) else "_no annotation column_"`.

```{r adt-dotplot-celltype, echo=FALSE, fig.width=14, fig.height=16}
if (!is.null(celltype_col)) {
  scMultiData_dp <- scMultiData
  Idents(scMultiData_dp) <- celltype_col
  DefaultAssay(scMultiData_dp) <- "ADT"
  feats <- adt_top_per_group(scMultiData_dp, celltype_col, n = 3L, min_cells = 30L)
  message("ADT DotPlot by cell type: ", length(feats), " features (top 3 per type, dedup; ",
          "of ", nrow(scMultiData_dp[["ADT"]]), " ADT features total)")
  p <- DotPlot(scMultiData_dp, features = feats, assay = "ADT") +
    coord_flip() +
    scale_color_viridis_c(option = "D") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(p)
} else {
  cat("No cell-type annotation column on the object - cluster panel only.\n")
}
```

### Top markers (FeaturePlot)

```{r adt-feature-plots, echo=FALSE, fig.width=14, fig.height=7}
if (!is.null(adt_counts_mat)) {
  data_mat <- tryCatch(
    as.matrix(GetAssayData(scMultiData, assay = "ADT", layer = "data")),
    error = function(e) adt_counts_mat
  )
  # Skip isotype controls when ranking top ADT markers
  rm_iso <- !grepl("(?i)isotype", rownames(data_mat))
  ranked <- names(sort(rowMeans(data_mat[rm_iso, , drop = FALSE]), decreasing = TRUE))
  top_features <- ranked[seq_len(min(8L, length(ranked)))]
  scMultiData_fp <- scMultiData
  DefaultAssay(scMultiData_fp) <- "ADT"
  p <- FeaturePlot(scMultiData_fp, features = top_features,
                   reduction = "adt.umap", raster = FALSE, ncol = 4) &
    scale_color_viridis_c(option = "D") &
    theme(plot.margin = ggplot2::margin(2, 4, 2, 4, "pt"),
          plot.title  = ggplot2::element_text(size = 11, margin = ggplot2::margin(b = 2)))
  print(p)
}
```

### QC

`r format(n_adt, big.mark = ",")` ADT features, normalization: **`r adt_norm`**, mean `r mean_umi` UMI/cell, median `r median_umi` UMI/cell. Cell-type column: `r if (!is.null(celltype_col)) sprintf("**%s**", celltype_col) else "_none available_"`.
