## ATAC {.tabset}

```{r atac-prep, include=FALSE}
library(Signac)
n_peaks <- nrow(scMultiData[["ATAC"]])
n_cells_atac <- ncol(scMultiData)
DefaultAssay(scMultiData) <- "ATAC"
if (!"nucleosome_signal" %in% colnames(scMultiData@meta.data)) {
  scMultiData <- tryCatch(
    NucleosomeSignal(scMultiData, assay = "ATAC"),
    error = function(e) scMultiData
  )
}
if (!"TSS.enrichment" %in% colnames(scMultiData@meta.data)) {
  scMultiData <- tryCatch(
    TSSEnrichment(scMultiData, assay = "ATAC", fast = TRUE),
    error = function(e) scMultiData
  )
}
DefaultAssay(scMultiData) <- "RNA"
has_ga <- "GeneActivity" %in% Assays(scMultiData)
```

*Clustering used:* **`r if (!is.null(cluster_col)) cluster_col else "_none_"`** — RNA-derived (no ATAC-specific clustering computed). UMAP layout is from the ATAC LSI reduction (`atac.umap`).

### UMAP

```{r atac-umaps, echo=FALSE, fig.width=14, fig.height=7}
plots <- list()
if ("atac.umap" %in% names(scMultiData@reductions) && !is.null(cluster_col)) {
  plots[["clusters"]] <- DimPlot(scMultiData, reduction = "atac.umap",
                                 group.by = cluster_col, raster = FALSE,
                                 label = TRUE, repel = TRUE, order = TRUE,
                                 cols = unname(pals::polychrome())) +
    theme(aspect.ratio = 1) + ggtitle(paste0("ATAC UMAP - clusters (", cluster_col, ")"))
}
if ("atac.umap" %in% names(scMultiData@reductions) && !is.null(celltype_col)) {
  plots[["celltype"]] <- DimPlot(scMultiData, reduction = "atac.umap",
                                 group.by = celltype_col, raster = FALSE,
                                 label = TRUE, repel = TRUE, order = TRUE) +
    theme(aspect.ratio = 1) + ggtitle(paste0("ATAC UMAP - cell type (", celltype_col, ")"))
}
if (length(plots) >= 2) {
  print(patchwork::wrap_plots(plots, ncol = 2))
} else if (length(plots) == 1) {
  print(plots[[1]])
}
```

### Gene activity FeaturePlots

```{r atac-gene-activity, echo=FALSE, fig.width=12, fig.height=10}
if (has_ga) {
  scMultiData_ga <- scMultiData
  DefaultAssay(scMultiData_ga) <- "GeneActivity"
  ga_mat <- tryCatch(
    as.matrix(GetAssayData(scMultiData_ga, assay = "GeneActivity", layer = "data")),
    error = function(e) NULL
  )
  if (!is.null(ga_mat)) {
    ga_means <- rowMeans(ga_mat)
    top_ga <- names(sort(ga_means, decreasing = TRUE))[1:min(8L, length(ga_means))]
    red <- if ("atac.umap" %in% names(scMultiData_ga@reductions)) "atac.umap" else rna_umap
    p <- FeaturePlot(scMultiData_ga, features = top_ga, reduction = red,
                     raster = FALSE, ncol = 4) &
      theme(aspect.ratio = 1) &
      scale_color_viridis_c(option = "D")
    print(p)
  }
} else {
  cat("GeneActivity assay not available - genome annotation could not be resolved.\n")
}
```

### Fragment counts

```{r atac-frag, echo=FALSE, fig.width=8, fig.height=5}
print(VlnPlot(scMultiData, features = "nCount_ATAC", group.by = cluster_col %||% "Sample",
              pt.size = 0) + scale_y_log10() + NoLegend())
```

### TSS enrichment

```{r atac-tss, echo=FALSE, fig.width=8, fig.height=5}
if ("TSS.enrichment" %in% colnames(scMultiData@meta.data)) {
  print(VlnPlot(scMultiData, features = "TSS.enrichment", group.by = cluster_col %||% "Sample",
                pt.size = 0) + NoLegend())
} else {
  cat("TSS enrichment unavailable - genome annotation could not be resolved.\n")
}
```

### Nucleosome signal

```{r atac-nuc, echo=FALSE, fig.width=8, fig.height=5}
if ("nucleosome_signal" %in% colnames(scMultiData@meta.data)) {
  print(VlnPlot(scMultiData, features = "nucleosome_signal", group.by = cluster_col %||% "Sample",
                pt.size = 0) + NoLegend())
} else {
  cat("Nucleosome signal unavailable - genome annotation could not be resolved.\n")
}
```

### Summary

`r format(n_peaks, big.mark = ",")` peaks, `r format(n_cells_atac, big.mark = ",")` cells. `r if (has_ga) paste0("GeneActivity assay: ", format(nrow(scMultiData[["GeneActivity"]]), big.mark = ","), " genes.") else "GeneActivity assay not available."`
