## ADT

```{r adt-umaps, echo=FALSE, fig.width=14, fig.height=7}
plots <- list()
if ("adt.umap" %in% names(scMultiData@reductions) && !is.null(cluster_col)) {
  plots[["clusters"]] <- DimPlot(scMultiData, reduction = "adt.umap",
                                 group.by = cluster_col, raster = FALSE,
                                 label = TRUE, repel = TRUE, order = TRUE,
                                 cols = unname(pals::polychrome())) +
    theme(aspect.ratio = 1) + ggtitle("ADT UMAP - clusters")
}
if ("adt.umap" %in% names(scMultiData@reductions) && !is.null(celltype_col)) {
  plots[["celltype"]] <- DimPlot(scMultiData, reduction = "adt.umap",
                                 group.by = celltype_col, raster = FALSE,
                                 label = TRUE, repel = TRUE, order = TRUE) +
    theme(aspect.ratio = 1) + ggtitle("ADT UMAP - cell type")
}
if (length(plots) >= 2) {
  print(patchwork::wrap_plots(plots, ncol = 2))
} else if (length(plots) == 1) {
  print(plots[[1]])
}
```

```{r adt-prep, include=FALSE}
n_adt   <- nrow(scMultiData[["ADT"]])
adt_features <- rownames(scMultiData[["ADT"]])
adt_counts_mat <- tryCatch(
  as.matrix(GetAssayData(scMultiData, assay = "ADT", layer = "counts")),
  error = function(e) NULL
)
mean_umi   <- if (is.matrix(adt_counts_mat)) round(mean(colSums(adt_counts_mat)), 1) else NA
median_umi <- if (is.matrix(adt_counts_mat)) round(median(colSums(adt_counts_mat)), 1) else NA
adt_norm <- if (!is.null(param$adtNorm)) param$adtNorm else "ADTnorm"
```

The DotPlot below shows the **normalized ADT signal** (`r adt_norm` normalization, then z-scored across cells via Seurat's `DotPlot` default). Colour ramp uses **viridis** so the same scale applies across panels.

### DotPlot {.tabset}

#### By cluster

```{r adt-dotplot-cluster, echo=FALSE, fig.width=10, fig.height=max(6, length(adt_features) * 0.18)}
if (!is.null(cluster_col)) {
  scMultiData_dp <- scMultiData
  Idents(scMultiData_dp) <- cluster_col
  DefaultAssay(scMultiData_dp) <- "ADT"
  p <- DotPlot(scMultiData_dp, features = unique(adt_features), assay = "ADT") +
    coord_flip() +
    scale_color_viridis_c(option = "D") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = paste("ADT signal per cluster (", adt_norm, ")"))
  print(p)
}
```

#### By cell type

```{r adt-dotplot-celltype, echo=FALSE, fig.width=10, fig.height=max(6, length(adt_features) * 0.18)}
if (!is.null(celltype_col)) {
  scMultiData_dp <- scMultiData
  Idents(scMultiData_dp) <- celltype_col
  DefaultAssay(scMultiData_dp) <- "ADT"
  p <- DotPlot(scMultiData_dp, features = unique(adt_features), assay = "ADT") +
    coord_flip() +
    scale_color_viridis_c(option = "D") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = paste0("ADT signal per cell type (", celltype_col,
                        " annotation, ", adt_norm, ")"))
  print(p)
} else {
  cat("No cell-type annotation column on the object - cluster panel only.\n")
}
```

### Top markers (FeaturePlot)

```{r adt-feature-plots, echo=FALSE, fig.width=12, fig.height=10}
if (!is.null(adt_counts_mat)) {
  data_mat <- tryCatch(
    as.matrix(GetAssayData(scMultiData, assay = "ADT", layer = "data")),
    error = function(e) adt_counts_mat
  )
  top_features <- names(sort(rowMeans(data_mat), decreasing = TRUE))[seq_len(min(8L, nrow(data_mat)))]
  scMultiData_fp <- scMultiData
  DefaultAssay(scMultiData_fp) <- "ADT"
  p <- FeaturePlot(scMultiData_fp, features = top_features,
                   reduction = "adt.umap", raster = FALSE, ncol = 4) &
    theme(aspect.ratio = 1) &
    scale_color_viridis_c(option = "D")
  print(p)
}
```

### QC

`r format(n_adt, big.mark = ",")` ADT features, normalization: **`r adt_norm`**, mean `r mean_umi` UMI/cell, median `r median_umi` UMI/cell. Cell-type column used for annotation panels: `r if (!is.null(celltype_col)) sprintf("**%s**", celltype_col) else "_none available_"`.
