## WNN {.tabset}

```{r wnn-prep, include=FALSE}
red_used <- intersect(c("pca", "adt.pca", "lsi"), names(scMultiData@reductions))
weight_cols <- grep("\\.weight$", colnames(scMultiData@meta.data), value = TRUE)
wnn_resolution <- modalities$wnnResolution %||% scMultiData@misc$wnn_resolution %||% 0.5
wnn_cluster_col <- scMultiData@misc$wnn_cluster_col %||% paste0("wsnn_res.", wnn_resolution)
if (!wnn_cluster_col %in% colnames(scMultiData@meta.data)) {
  wsnn_hits <- grep("^wsnn_res\\.", colnames(scMultiData@meta.data), value = TRUE)
  wnn_cluster_col <- if (length(wsnn_hits) > 0) wsnn_hits[1] else cluster_col
}
```

*Clustering used:* **`r wnn_cluster_col`** — Leiden on the weighted SNN graph at resolution `r wnn_resolution`. Integrated reductions: **`r paste(red_used, collapse = ", ")`**.

### UMAP

```{r wnn-umaps, echo=FALSE, fig.width=14, fig.height=7}
plots <- list()
if ("wnn.umap" %in% names(scMultiData@reductions) && !is.null(wnn_cluster_col)) {
  plots[["clusters"]] <- DimPlot(scMultiData, reduction = "wnn.umap",
                                 group.by = wnn_cluster_col, raster = FALSE,
                                 label = TRUE, repel = TRUE, order = TRUE,
                                 cols = unname(pals::polychrome())) +
    theme(aspect.ratio = 1) + ggtitle(paste0("WNN UMAP - clusters (", wnn_cluster_col, ")"))
}
if ("wnn.umap" %in% names(scMultiData@reductions) && !is.null(celltype_col)) {
  plots[["celltype"]] <- DimPlot(scMultiData, reduction = "wnn.umap",
                                 group.by = celltype_col, raster = FALSE,
                                 label = TRUE, repel = TRUE, order = TRUE) +
    theme(aspect.ratio = 1) + ggtitle(paste0("WNN UMAP - cell type (", celltype_col, ")"))
}
if (length(plots) >= 2) {
  print(patchwork::wrap_plots(plots, ncol = 2))
} else if (length(plots) == 1) {
  print(plots[[1]])
}
```

### Cluster DEGs {.tabset}

Wilcoxon `FindAllMarkers` per modality, grouped by **`r wnn_cluster_col`** (`min.pct = 0.25`, `logfc.threshold = 0.25`, only positive markers). One subtab per modality available on the object: top markers as DotPlot + interactive sortable / filterable / Excel-exportable table. RNA uses the `RNA` assay; ADT uses the `ADT` assay; ATAC uses the `GeneActivity` assay (gene-level chromatin accessibility derived from the fragments).

```{r wnn-deg-compute, include=FALSE}
deg_assays <- c()
if ("RNA" %in% Assays(scMultiData)) deg_assays <- c(deg_assays, RNA = "RNA")
if ("ADT" %in% Assays(scMultiData)) deg_assays <- c(deg_assays, ADT = "ADT")
if ("GeneActivity" %in% Assays(scMultiData)) deg_assays <- c(deg_assays, ATAC = "GeneActivity")

deg_results <- list()
top_features  <- list()
if (!is.null(wnn_cluster_col) && wnn_cluster_col %in% colnames(scMultiData@meta.data)) {
  for (label in names(deg_assays)) {
    assay_name <- deg_assays[[label]]
    scMultiData_wm <- scMultiData
    DefaultAssay(scMultiData_wm) <- assay_name
    Idents(scMultiData_wm) <- wnn_cluster_col
    n_features <- nrow(scMultiData_wm[[assay_name]])
    if (n_features < 2L) next
    mk <- tryCatch(
      Seurat::FindAllMarkers(scMultiData_wm, assay = assay_name, only.pos = TRUE,
                             min.pct = 0.25, logfc.threshold = 0.25,
                             verbose = FALSE),
      error = function(e) { message("FindAllMarkers ", label, " failed: ",
                                    conditionMessage(e)); NULL }
    )
    if (!is.null(mk) && nrow(mk) > 0) {
      mk <- mk[, c("gene", "cluster", "avg_log2FC", "p_val_adj", "pct.1", "pct.2")]
      mk$avg_log2FC <- round(mk$avg_log2FC, 3)
      mk$pct.1 <- round(mk$pct.1, 3); mk$pct.2 <- round(mk$pct.2, 3)
      deg_results[[label]] <- mk
      top_features[[label]] <- mk |>
        dplyr::group_by(cluster) |>
        dplyr::slice_max(order_by = avg_log2FC, n = 5) |>
        dplyr::pull(gene) |>
        unique()
    }
  }
}
```

```{r wnn-deg-children, echo=FALSE, results='asis'}
# Render each modality as a knit_child so the embedded ezInteractiveTableRmd
# DT widget loads its CSS/JS deps correctly (asis + print() loses them).
if (length(deg_results) == 0) {
  cat("No WNN cluster markers detected.\n\n")
} else {
  subtab_path <- system.file("templates/_scMultiOmics_wnn_deg_subtab.Rmd",
                             package = "ezRun")
  for (lbl in names(deg_results)) {
    label      <- lbl
    assay_name <- deg_assays[[lbl]]
    mk         <- deg_results[[lbl]]
    top_genes  <- top_features[[lbl]]
    cat(sprintf("\n#### %s\n\n", label))
    cat(knitr::knit_child(subtab_path, envir = environment(), quiet = TRUE),
        sep = "\n")
  }
}
```

### Per-modality weights (Violin)

```{r wnn-weights-vln, echo=FALSE, fig.width=10, fig.height=5}
if (length(weight_cols) > 0 && !is.null(wnn_cluster_col)) {
  print(VlnPlot(scMultiData, features = weight_cols,
                group.by = wnn_cluster_col, pt.size = 0,
                ncol = length(weight_cols)))
} else {
  cat("No modality weight columns - WNN may not have run.\n")
}
```

### Per-modality weights (FeaturePlot)

```{r wnn-weights-feat, echo=FALSE, fig.width=12, fig.height=4 * ceiling(max(1, length(weight_cols))/3)}
if (length(weight_cols) > 0 && "wnn.umap" %in% names(scMultiData@reductions)) {
  p <- FeaturePlot(scMultiData, features = weight_cols,
                   reduction = "wnn.umap", raster = FALSE,
                   ncol = min(3L, length(weight_cols))) &
    theme(aspect.ratio = 1) &
    scale_color_viridis_c(option = "C")
  print(p)
} else {
  cat("FeaturePlot unavailable - WNN UMAP missing.\n")
}
```

### Summary

Integrated reductions: **`r paste(red_used, collapse = ", ")`**. Resolution: **`r wnn_resolution`**. Cluster column: **`r wnn_cluster_col`**.
