## VDJ {.tabset}

```{r vdj-prep, include=FALSE}
library(scRepertoire)
has_vdjT <- isTRUE(modalities$hasVDJ_T)
has_vdjB <- isTRUE(modalities$hasVDJ_B)
chain_label <- paste(c(if (has_vdjT) "TCR" else NULL,
                       if (has_vdjB) "BCR" else NULL), collapse = " + ")
chain_prefix <- if (has_vdjT) "TR" else "IG"

clone_pal <- ezRun::cloneSizePalette()
n_with_clone <- sum(!is.na(scMultiData@meta.data[["CTaa"]]))
n_total      <- ncol(scMultiData)
pct_recovery <- round(100 * n_with_clone / n_total, 1)

group_for_clonal <- cluster_col %||% "Sample"

# Reuse the canonical cluster palette / numeric ordering from the parent Rmd.
# When VDJ is grouped by something other than cluster_col (e.g. "Sample"),
# fall back to a fresh polychrome and the alphabetical level order.
if (identical(group_for_clonal, cluster_col) && !is.null(cluster_levels)) {
  group_levels <- cluster_levels
} else {
  lvls <- levels(as.factor(scMultiData@meta.data[[group_for_clonal]]))
  num  <- suppressWarnings(as.integer(lvls))
  group_levels <- if (all(!is.na(num))) as.character(sort(num)) else lvls
}
group_pal <- if (identical(group_for_clonal, cluster_col) && !is.null(cluster_pal)) {
  cluster_pal
} else if (!is.null(group_levels)) {
  setNames(unname(pals::polychrome())[seq_along(group_levels)], group_levels)
} else NULL

safe_print <- function(..., recolor = TRUE) {
  args <- list(...)
  expr_call <- args[[1]]
  p <- tryCatch(eval(expr_call), error = function(e) {
    message("Plot skipped: ", conditionMessage(e)); NULL
  })
  if (!is.null(p)) {
    if (isTRUE(recolor) && !is.null(group_pal) && inherits(p, "ggplot")) {
      p <- p + scale_fill_manual(values = group_pal, drop = FALSE,
                                 limits = names(group_pal))
    }
    print(p)
  }
}
```

*VDJ data is metadata-only* — it overlays cloneSize on the RNA / ADT / WNN UMAPs but produces no clustering of its own. Bar plots below are grouped by **`r group_for_clonal`** (the active clustering column).

### cloneSize on UMAPs

```{r vdj-umap, echo=FALSE, fig.width=20, fig.height=10}
plots <- list()
p_rna <- ezRun::cloneDimPlot(scMultiData, rna_umap, palette = clone_pal,
                             title = paste(chain_label, "clone size on RNA UMAP"))
if (!is.null(p_rna)) plots[["RNA"]] <- p_rna
if ("adt.umap" %in% names(scMultiData@reductions)) {
  p_adt <- ezRun::cloneDimPlot(scMultiData, "adt.umap", palette = clone_pal,
                               title = paste(chain_label, "on ADT UMAP"))
  if (!is.null(p_adt)) plots[["ADT"]] <- p_adt
}
if ("wnn.umap" %in% names(scMultiData@reductions)) {
  p_wnn <- ezRun::cloneDimPlot(scMultiData, "wnn.umap", palette = clone_pal,
                               title = paste(chain_label, "on WNN UMAP"))
  if (!is.null(p_wnn)) plots[["WNN"]] <- p_wnn
}
if (length(plots) >= 2) {
  print(patchwork::wrap_plots(plots, ncol = min(length(plots), 2L), guides = "collect"))
} else if (length(plots) == 1) {
  print(plots[[1]])
}
```

### Per cluster / cell type {.tabset}

#### clonalOccupy by cluster

```{r vdj-occ-cluster, echo=FALSE, fig.width=10, fig.height=10}
if (!is.null(cluster_col)) {
  p_n   <- ezRun::cloneOccupyFull(scMultiData, x.axis = cluster_col, palette = clone_pal)
  p_pct <- ezRun::cloneOccupyFull(scMultiData, x.axis = cluster_col,
                                  palette = clone_pal, proportion = TRUE)
  print(patchwork::wrap_plots(p_n, p_pct, ncol = 1, guides = "collect"))
}
```

#### clonalOccupy by cell type

```{r vdj-occ-celltype, echo=FALSE, fig.width=10, fig.height=10}
if (!is.null(celltype_col)) {
  p_n   <- ezRun::cloneOccupyFull(scMultiData, x.axis = celltype_col, palette = clone_pal)
  p_pct <- ezRun::cloneOccupyFull(scMultiData, x.axis = celltype_col,
                                  palette = clone_pal, proportion = TRUE)
  print(patchwork::wrap_plots(p_n, p_pct, ncol = 1, guides = "collect"))
} else {
  cat("No cell-type annotation column on the object.\n")
}
```

### Repertoire statistics {.tabset}

#### clonalQuant

```{r vdj-quant, echo=FALSE, fig.width=8, fig.height=5}
safe_print(bquote(clonalQuant(scMultiData, cloneCall = "strict", chain = "both",
                              scale = TRUE, group.by = .(group_for_clonal),
                              order.by = .(group_levels))))
```

#### clonalHomeostasis

```{r vdj-homeostasis, echo=FALSE, fig.width=8, fig.height=5}
# scRepertoire fills by Var2 (homeostasis bins: Rare/Small/Medium/Large/Hyperexpanded),
# NOT by cluster - skip the polychrome recolor.
safe_print(recolor = FALSE,
           bquote(clonalHomeostasis(scMultiData, cloneCall = "strict",
                                    group.by = .(group_for_clonal),
                                    order.by = .(group_levels))))
```

#### clonalProportion

```{r vdj-proportion, echo=FALSE, fig.width=8, fig.height=5}
# scRepertoire fills by Var2 (clonalSplit bins like [1:10], [11:100]),
# NOT by cluster - skip the polychrome recolor.
safe_print(recolor = FALSE,
           bquote(clonalProportion(scMultiData, cloneCall = "strict",
                                   group.by = .(group_for_clonal),
                                   order.by = .(group_levels))))
```

#### clonalDiversity

```{r vdj-diversity, echo=FALSE, fig.width=10, fig.height=5}
safe_print(bquote(clonalDiversity(scMultiData, cloneCall = "strict",
                                  group.by = .(group_for_clonal),
                                  order.by = .(group_levels),
                                  metrics = c("shannon", "inv.simpson", "norm.entropy"))))
```

#### CDR3aa length

```{r vdj-aa-length, echo=FALSE, fig.width=9, fig.height=5}
safe_print(bquote(clonalLength(scMultiData, cloneCall = "aa", chain = "both",
                               scale = TRUE, group.by = .(group_for_clonal),
                               order.by = .(group_levels))))
```

### Top 3 clones on UMAP

```{r vdj-top3-umap, echo=FALSE, fig.width=15, fig.height=5}
if ("CTstrict" %in% colnames(scMultiData@meta.data)) {
  ct_counts <- sort(table(scMultiData$CTstrict), decreasing = TRUE)
  top3 <- names(ct_counts)[seq_len(min(3L, length(ct_counts)))]
  if (length(top3) > 0) {
    emb <- as.data.frame(Embeddings(scMultiData, rna_umap))
    xcol <- colnames(emb)[1]; ycol <- colnames(emb)[2]
    plots <- lapply(seq_along(top3), function(i) {
      key <- top3[i]
      cdr3 <- scMultiData$CTaa[scMultiData$CTstrict == key &
                               !is.na(scMultiData$CTstrict)][1]
      cdr3_short <- if (is.na(cdr3)) "" else
                    paste0("\n", substr(cdr3, 1, 35),
                           if (nchar(cdr3) > 35) "..." else "")
      in_clone <- !is.na(scMultiData$CTstrict) & scMultiData$CTstrict == key
      df_bg <- emb[!in_clone, ]
      df_fg <- emb[in_clone, ]
      ggplot() +
        geom_point(data = df_bg, aes(.data[[xcol]], .data[[ycol]]),
                   color = "grey85", size = 0.4, alpha = 0.6) +
        geom_point(data = df_fg, aes(.data[[xcol]], .data[[ycol]]),
                   color = "#E45358", size = 1.6) +
        theme_classic() +
        theme(aspect.ratio = 1, panel.grid = element_blank()) +
        ggtitle(sprintf("Clone %d (%d cells)%s", i, ct_counts[key], cdr3_short))
    })
    print(patchwork::wrap_plots(plots, ncol = 3))
  }
}
```

### Top clones

```{r vdj-top-clones, echo=FALSE}
if ("CTaa" %in% colnames(scMultiData@meta.data)) {
  ct <- scMultiData@meta.data[["CTaa"]]; ct <- ct[!is.na(ct)]
  if (length(ct) > 0) {
    top_clones <- as.data.frame(sort(table(ct), decreasing = TRUE)[1:min(10, length(unique(ct)))])
    colnames(top_clones) <- c("CDR3 amino acids (paired)", "Cells")
    knitr::kable(top_clones)
  }
}
```

### Summary

`r format(n_with_clone, big.mark = ",")` / `r format(n_total, big.mark = ",")` (`r pct_recovery`%) cells with paired `r chain_label` clone; `r if ("CTaa" %in% colnames(scMultiData@meta.data)) format(length(unique(na.omit(scMultiData@meta.data[["CTaa"]]))), big.mark = ",") else "NA"` unique CDR3aa clonotypes.
