## VDJ

```{r vdj-prep, include=FALSE}
library(scRepertoire)

has_vdjT <- isTRUE(modalities$hasVDJ_T)
has_vdjB <- isTRUE(modalities$hasVDJ_B)
chain_label <- paste(c(if (has_vdjT) "TCR" else NULL,
                       if (has_vdjB) "BCR" else NULL), collapse = " + ")
chain_prefix <- if (has_vdjT) "TR" else "IG"

ct_cols   <- intersect(c("CTaa", "CTgene", "cloneSize", "Frequency", "clonalFrequency"),
                       colnames(scMultiData@meta.data))
n_with_clone <- sum(!is.na(scMultiData@meta.data[["CTaa"]]))
n_total      <- ncol(scMultiData)
pct_recovery <- round(100 * n_with_clone / n_total, 1)

# Top clones table
top_clones <- NA
if ("CTaa" %in% colnames(scMultiData@meta.data)) {
  ct <- scMultiData@meta.data[["CTaa"]]; ct <- ct[!is.na(ct)]
  if (length(ct) > 0) {
    top_clones <- as.data.frame(sort(table(ct), decreasing = TRUE)[1:min(10, length(unique(ct)))])
    colnames(top_clones) <- c("CDR3 amino acids (paired)", "Cells")
  }
}

cluster_col <- if ("seurat_clusters" %in% colnames(scMultiData@meta.data)) {
  "seurat_clusters"
} else NULL

safe_print <- function(expr_call, ...) {
  p <- tryCatch(eval(expr_call), error = function(e) {
    message("Plot skipped: ", conditionMessage(e)); NULL
  })
  if (!is.null(p)) print(p)
}
```

**`r chain_label` repertoire summary**

- **Cells with paired clone:** `r format(n_with_clone, big.mark = ",")` / `r format(n_total, big.mark = ",")` (`r pct_recovery`%)
- **Unique CDR3aa clonotypes:** `r if ("CTaa" %in% colnames(scMultiData@meta.data)) format(length(unique(na.omit(scMultiData@meta.data[["CTaa"]]))), big.mark = ",") else "NA"`
- **Sample(s):** `r paste(unique(scMultiData$Sample), collapse = ", ")`
- **Clone-size bins (cloneSize):**

```{r vdj-clone-size, echo=FALSE}
if ("cloneSize" %in% colnames(scMultiData@meta.data)) {
  cs <- as.data.frame(table(scMultiData$cloneSize, useNA = "ifany"))
  colnames(cs) <- c("cloneSize bin", "Cells")
  knitr::kable(cs)
}
```

**Top 10 paired clones**

```{r vdj-top-clones, echo=FALSE}
if (is.data.frame(top_clones)) knitr::kable(top_clones)
```

### Clone size landscape {.tabset}

#### clonalQuant

```{r vdj-quant, echo=FALSE, fig.width=8, fig.height=5}
safe_print(quote(clonalQuant(scMultiData, cloneCall = "strict", chain = "both",
                             scale = TRUE, group.by = "Sample")))
```

#### clonalAbundance

```{r vdj-abundance, echo=FALSE, fig.width=8, fig.height=5}
safe_print(quote(clonalAbundance(scMultiData, cloneCall = "strict",
                                 scale = FALSE, group.by = "Sample")))
```

#### clonalHomeostasis

```{r vdj-homeostasis, echo=FALSE, fig.width=8, fig.height=5}
safe_print(quote(clonalHomeostasis(scMultiData, cloneCall = "strict", group.by = "Sample")))
```

#### clonalProportion

```{r vdj-proportion, echo=FALSE, fig.width=8, fig.height=5}
safe_print(quote(clonalProportion(scMultiData, cloneCall = "strict", group.by = "Sample")))
```

#### clonalDiversity

```{r vdj-diversity, echo=FALSE, fig.width=10, fig.height=5}
safe_print(quote(clonalDiversity(scMultiData, cloneCall = "strict", group.by = "Sample",
                                 metrics = c("shannon", "inv.simpson", "norm.entropy"))))
```

### Clone size on embeddings {.tabset}

#### RNA UMAP

```{r vdj-overlay-rna, echo=FALSE, fig.width=8, fig.height=8}
if ("umap" %in% names(scMultiData@reductions) && "cloneSize" %in% colnames(scMultiData@meta.data)) {
  p <- DimPlot(scMultiData, reduction = "umap", group.by = "cloneSize",
               raster = FALSE, order = TRUE) +
    theme(aspect.ratio = 1) +
    ggtitle(paste(chain_label, "clone size on RNA UMAP"))
  print(p)
}
```

#### ADT UMAP

```{r vdj-overlay-adt, echo=FALSE, fig.width=8, fig.height=8}
if ("adt.umap" %in% names(scMultiData@reductions) && "cloneSize" %in% colnames(scMultiData@meta.data)) {
  p <- DimPlot(scMultiData, reduction = "adt.umap", group.by = "cloneSize",
               raster = FALSE, order = TRUE) +
    theme(aspect.ratio = 1) +
    ggtitle(paste(chain_label, "clone size on ADT UMAP"))
  print(p)
} else {
  cat("ADT UMAP not available - skipped.\n")
}
```

#### Top clones highlighted

```{r vdj-highlight, echo=FALSE, fig.width=8, fig.height=8}
if (is.data.frame(top_clones) && "umap" %in% names(scMultiData@reductions)) {
  top_seqs <- as.character(top_clones[seq_len(min(5L, nrow(top_clones))), 1])
  obj_h <- tryCatch(
    highlightClones(scMultiData, cloneCall = "aa", sequence = top_seqs),
    error = function(e) NULL
  )
  if (!is.null(obj_h)) {
    p <- DimPlot(obj_h, reduction = "umap", group.by = "highlight",
                 raster = FALSE, order = TRUE, cols = c("grey85", scales::hue_pal()(length(top_seqs)))) +
      theme(aspect.ratio = 1) +
      ggtitle("Top 5 clones highlighted (RNA UMAP)")
    print(p)
  }
}
```

### V / J gene usage {.tabset}

#### V gene barplot

```{r vdj-vgene, echo=FALSE, fig.width=12, fig.height=5}
v_axis <- paste0(chain_prefix, "BV")
safe_print(bquote(vizGenes(scMultiData, x.axis = .(v_axis), y.axis = NULL,
                           plot = "barplot", scale = TRUE)))
```

#### V x J heatmap

```{r vdj-vj-heatmap, echo=FALSE, fig.width=10, fig.height=8}
v_axis <- paste0(chain_prefix, "BV")
j_axis <- paste0(chain_prefix, "BJ")
safe_print(bquote(vizGenes(scMultiData, x.axis = .(v_axis), y.axis = .(j_axis),
                           plot = "heatmap", scale = TRUE)))
```

#### CDR3aa length distribution

```{r vdj-aa-length, echo=FALSE, fig.width=9, fig.height=5}
safe_print(quote(clonalLength(scMultiData, cloneCall = "aa", chain = "both",
                              scale = TRUE, group.by = "Sample")))
```

### Per-cluster clonal occupancy

```{r vdj-bycluster, echo=FALSE, fig.width=10, fig.height=5}
if (!is.null(cluster_col)) {
  safe_print(bquote(clonalOccupy(scMultiData, x.axis = .(cluster_col))))
}
```
