## VDJ

```{r vdj-prep, include=FALSE}
library(scRepertoire)
has_vdjT <- isTRUE(modalities$hasVDJ_T)
has_vdjB <- isTRUE(modalities$hasVDJ_B)
chain_label <- paste(c(if (has_vdjT) "TCR" else NULL,
                       if (has_vdjB) "BCR" else NULL), collapse = " + ")
chain_prefix <- if (has_vdjT) "TR" else "IG"

clone_pal <- ezRun::cloneSizePalette()
n_with_clone <- sum(!is.na(scMultiData@meta.data[["CTaa"]]))
n_total      <- ncol(scMultiData)
pct_recovery <- round(100 * n_with_clone / n_total, 1)

# Force a colour scale layer for DimPlot(group.by = "cloneSize") so it matches
# scRepertoire bar plots. We also set the order so the densest categories
# render on top.
clone_dim <- function(obj, reduction) {
  if (!"cloneSize" %in% colnames(obj@meta.data)) return(NULL)
  # Force the cloneSize column to a factor with all 6 levels so the legend
  # shows every bin (Hyperexpanded -> None) even when only a few are populated.
  obj$cloneSize <- factor(obj$cloneSize, levels = names(clone_pal))
  DimPlot(obj, reduction = reduction, group.by = "cloneSize",
          raster = FALSE, order = TRUE) +
    scale_color_manual(values = clone_pal, na.value = "grey85",
                       limits = names(clone_pal), breaks = names(clone_pal),
                       drop = FALSE) +
    theme(aspect.ratio = 1)
}

safe_print <- function(expr_call) {
  p <- tryCatch(eval(expr_call), error = function(e) {
    message("Plot skipped: ", conditionMessage(e)); NULL
  })
  if (!is.null(p)) print(p)
}
```

```{r vdj-umap, echo=FALSE, fig.width=14, fig.height=7}
plots <- list()
p_rna <- clone_dim(scMultiData, rna_umap)
if (!is.null(p_rna)) plots[["RNA"]] <- p_rna + ggtitle(paste(chain_label, "clone size on RNA UMAP"))
if ("adt.umap" %in% names(scMultiData@reductions)) {
  p_adt <- clone_dim(scMultiData, "adt.umap")
  if (!is.null(p_adt)) plots[["ADT"]] <- p_adt + ggtitle(paste(chain_label, "on ADT UMAP"))
}
if ("wnn.umap" %in% names(scMultiData@reductions)) {
  p_wnn <- clone_dim(scMultiData, "wnn.umap")
  if (!is.null(p_wnn)) plots[["WNN"]] <- p_wnn + ggtitle(paste(chain_label, "on WNN UMAP"))
}
if (length(plots) >= 2) {
  print(patchwork::wrap_plots(plots, ncol = min(length(plots), 2L), guides = "collect"))
} else if (length(plots) == 1) {
  print(plots[[1]])
}
```

### Repertoire statistics {.tabset}

#### clonalQuant

```{r vdj-quant, echo=FALSE, fig.width=8, fig.height=5}
group_for_clonal <- cluster_col %||% "Sample"
safe_print(bquote(clonalQuant(scMultiData, cloneCall = "strict", chain = "both",
                              scale = TRUE, group.by = .(group_for_clonal))))
```

#### clonalAbundance

```{r vdj-abundance, echo=FALSE, fig.width=8, fig.height=5}
safe_print(bquote(clonalAbundance(scMultiData, cloneCall = "strict",
                                  scale = FALSE, group.by = .(group_for_clonal))))
```

#### clonalHomeostasis

```{r vdj-homeostasis, echo=FALSE, fig.width=8, fig.height=5}
safe_print(bquote(clonalHomeostasis(scMultiData, cloneCall = "strict",
                                    group.by = .(group_for_clonal))))
```

#### clonalProportion

```{r vdj-proportion, echo=FALSE, fig.width=8, fig.height=5}
safe_print(bquote(clonalProportion(scMultiData, cloneCall = "strict",
                                   group.by = .(group_for_clonal))))
```

#### clonalDiversity

```{r vdj-diversity, echo=FALSE, fig.width=10, fig.height=5}
safe_print(bquote(clonalDiversity(scMultiData, cloneCall = "strict",
                                  group.by = .(group_for_clonal),
                                  metrics = c("shannon", "inv.simpson", "norm.entropy"))))
```

#### CDR3aa length

```{r vdj-aa-length, echo=FALSE, fig.width=9, fig.height=5}
safe_print(bquote(clonalLength(scMultiData, cloneCall = "aa", chain = "both",
                               scale = TRUE, group.by = .(group_for_clonal))))
```

### Per cluster / cell type {.tabset}

#### clonalOccupy by cluster

```{r vdj-occ-cluster, echo=FALSE, fig.width=10, fig.height=5}
if (!is.null(cluster_col)) {
  safe_print(bquote(clonalOccupy(scMultiData, x.axis = .(cluster_col)) +
                      scale_fill_manual(values = clone_pal, na.value = "grey85",
                                              limits = names(clone_pal),
                                              breaks = names(clone_pal),
                                              drop = FALSE)))
}
```

#### clonalOccupy by cell type

```{r vdj-occ-celltype, echo=FALSE, fig.width=10, fig.height=5}
if (!is.null(celltype_col)) {
  safe_print(bquote(clonalOccupy(scMultiData, x.axis = .(celltype_col)) +
                      scale_fill_manual(values = clone_pal, na.value = "grey85",
                                              limits = names(clone_pal),
                                              breaks = names(clone_pal),
                                              drop = FALSE)))
} else {
  cat("No cell-type annotation column on the object.\n")
}
```

### V / J gene usage {.tabset}

#### V gene barplot

```{r vdj-vgene, echo=FALSE, fig.width=12, fig.height=5}
v_axis <- paste0(chain_prefix, "BV")
safe_print(bquote(vizGenes(scMultiData, x.axis = .(v_axis), y.axis = NULL,
                           plot = "barplot", scale = TRUE)))
```

#### V x J heatmap

```{r vdj-vj-heatmap, echo=FALSE, fig.width=10, fig.height=8}
v_axis <- paste0(chain_prefix, "BV")
j_axis <- paste0(chain_prefix, "BJ")
safe_print(bquote(vizGenes(scMultiData, x.axis = .(v_axis), y.axis = .(j_axis),
                           plot = "heatmap", scale = TRUE)))
```

### Top clones

```{r vdj-top-clones, echo=FALSE}
if ("CTaa" %in% colnames(scMultiData@meta.data)) {
  ct <- scMultiData@meta.data[["CTaa"]]; ct <- ct[!is.na(ct)]
  if (length(ct) > 0) {
    top_clones <- as.data.frame(sort(table(ct), decreasing = TRUE)[1:min(10, length(unique(ct)))])
    colnames(top_clones) <- c("CDR3 amino acids (paired)", "Cells")
    knitr::kable(top_clones)
  }
}
```

### Summary

`r format(n_with_clone, big.mark = ",")` / `r format(n_total, big.mark = ",")` (`r pct_recovery`%) cells with paired `r chain_label` clone; `r if ("CTaa" %in% colnames(scMultiData@meta.data)) format(length(unique(na.omit(scMultiData@meta.data[["CTaa"]]))), big.mark = ",") else "NA"` unique CDR3aa clonotypes.
