---
params:
  reportTitle: "ChIP-seq Peak Comparison"
title: "`r params$reportTitle`"
---

<!-- Styling comes from the fgczQuartoTemplate assets staged next to this file by
     ezRun::makeQuartoReport(), which calls fgczQuartoTemplate::fgcz_render().
     There is deliberately no format: block here. -->

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
suppressPackageStartupMessages({
  library(GenomicRanges); library(ggplot2); library(data.table)
})
## allow non-UCSC chromosome names (e.g. Ensembl "1", bacterial RefSeq "NC_*")
## in Gviz tracks, otherwise DataTrack/AnnotationTrack reject them.
options(ucscChromosomeNames = FALSE)

.fgcz_load <- function(name, nthreads = 4L) {
  paths <- c(file.path("data", paste0(name, ".qs2")), paste0(name, ".qs2"),
             file.path("data", paste0(name, ".rds")), paste0(name, ".rds"))
  for (p in paths) if (file.exists(p)) {
    if (grepl("\\.qs2$", p)) return(qs2::qs_read(p, nthreads = nthreads))
    return(readRDS(p))
  }
  stop("No .qs2/.rds file found for: ", name)
}
.tbl <- function(df, digits = 3, ...) {
  df <- as.data.frame(df)
  num <- vapply(df, is.numeric, logical(1)); df[num] <- lapply(df[num], round, digits)
  if (requireNamespace("DT", quietly = TRUE))
    return(DT::datatable(df, class = "compact stripe", rownames = FALSE,
                         options = list(pageLength = 10, scrollX = TRUE, dom = "tip"), ...))
  knitr::kable(df)
}
.noteHTML <- function(txt) htmltools::HTML(sprintf(
  '<div class="callout callout-style-default callout-note"><div class="callout-body-container callout-body"><p>%s</p></div></div>', txt))
.have <- function(x) !is.null(x) && (is.data.frame(x) || length(x) > 0)

chipData    <- .fgcz_load("chipData")
param       <- chipData$param
sampleNames <- chipData$sampleNames
condition   <- chipData$condition
nSamples    <- chipData$nSamples
peaks       <- chipData$peaks
cvdPalette  <- grDevices::palette.colors(palette = "Okabe-Ito")   # colour-blind safe
sampleCol   <- setNames(rep(cvdPalette, length.out = nSamples), sampleNames)
```

::: {.panel-tabset}

# Overview

```{r overview-text}
#| output: asis
cat(sprintf("Reference build: **`%s`** &nbsp;|&nbsp; peak column: **%s** &nbsp;|&nbsp; mark type: **%s**\n\n",
            param$refBuild, chipData$peakCol, param$markType))
cat(sprintf("top peaks (n) / rankBy: **%s / %s** &nbsp;|&nbsp; quantify from: **%s** &nbsp;|&nbsp; normalization: **%s** &nbsp;|&nbsp; consensus min samples: **%s**\n\n",
            param$topN, param$rankBy, param$quantifyFrom, param$normalization,
            param$minSamplesForConsensus))
```

```{r overview-table}
info <- data.frame(
  Sample = sampleNames,
  Condition = if (!is.null(condition)) as.character(condition) else "NA",
  Peaks = chipData$peakStats$nPeaks[match(sampleNames, chipData$peakStats$sample)],
  BAM = ifelse(chipData$hasBam, "yes", "no"),
  BigWig = ifelse(chipData$hasBw, "yes", "no"),
  stringsAsFactors = FALSE)
.tbl(info)
```

### Peaks dropped during harmonisation

```{r droplog}
if (.have(chipData$dropLog) && nrow(chipData$dropLog) > 0) .tbl(chipData$dropLog) else
  .noteHTML("No peaks were dropped (all on-reference, in-bounds, not blacklisted).")
```

# QC

::: {.panel-tabset}

## Summary

```{r qcflags-table}
if (.have(chipData$qcFlags) && nrow(chipData$qcFlags) > 0) {
  fl <- chipData$qcFlags
  wide <- reshape(fl[, c("sample", "metric", "value")], idvar = "sample",
                  timevar = "metric", direction = "wide")
  colnames(wide) <- sub("^value\\.", "", colnames(wide))
  .tbl(wide)
} else .noteHTML("No flaggable QC metrics available for this dataset.")
```

::: {.callout-tip}
## Flags
**good** ≥ good threshold &nbsp; **acceptable** ≥ min &nbsp; **low** < min. ENCODE NRF/PBC/NSC/RSC thresholds are calibrated for large eukaryotic genomes; on compact genomes read positions saturate and these metrics read low even for good libraries.
:::

## Alignment QC

::: {.callout-note collapse="true"}
## Metric definitions
- **NRF** (Non-Redundant Fraction) — distinct read start positions ÷ total mapped reads. Low = PCR duplication (or, on compact genomes, positional saturation).
- **PBC1** (PCR Bottleneck Coefficient 1) — positions with exactly one read ÷ distinct positions.
- **PBC2** (PCR Bottleneck Coefficient 2) — positions with one read ÷ positions with two reads.
- **duplicationRate** — fraction of reads flagged as duplicates. **mapq30Fraction** — fraction with MAPQ ≥ 30.
:::

```{r aln}
if (!chipData$hasBam) {
  .noteHTML("No BAM files in the dataset — alignment QC is unavailable.")
} else {
  tab <- chipData$alnStats
  if (.have(chipData$complexity)) tab <- merge(tab, chipData$complexity, by = "sample", all = TRUE)
  .tbl(tab)
}
```

## Signal QC

::: {.callout-note collapse="true"}
## Metric definitions
- **FRiP** (Fraction of Reads in Peaks) — aligned reads inside called peaks ÷ total; higher = better enrichment.
- **NSC** (Normalized Strand Cross-Correlation) — fragment-length cross-correlation peak relative to background; > 1 = enrichment.
- **RSC** (Relative Strand Cross-Correlation) — fragment-length peak height (above background) relative to the read-length phantom peak; ≥ 0.8 = good signal.
:::

```{r signalqc-table}
if (!chipData$hasBam) {
  .noteHTML("No BAM files — FRiP and NSC/RSC are unavailable.")
} else {
  parts <- list()
  if (.have(chipData$frip)) parts$FRiP <- chipData$frip
  if (.have(chipData$ccMetrics)) parts$CC <- chipData$ccMetrics
  .tbl(Reduce(function(a, b) merge(a, b, by = "sample", all = TRUE), parts))
}
```

```{r frip-plot}
#| fig-height: 3.5
if (chipData$hasBam && .have(chipData$frip)) {
  d <- chipData$frip; d$sample <- factor(d$sample, levels = sampleNames)
  ggplot(d, aes(sample, frip, fill = sample)) + geom_col() +
    scale_fill_manual(values = sampleCol, guide = "none") +
    labs(y = "FRiP", x = NULL, title = "Fraction of reads in peaks") +
    theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

```{r bwcor}
#| fig-height: 5
if (chipData$hasBw && !is.null(chipData$bwCor)) {
  m <- chipData$bwCor$spearman
  if (requireNamespace("ComplexHeatmap", quietly = TRUE))
    ComplexHeatmap::Heatmap(m, name = "Spearman", column_title = "BigWig genome-wide correlation")
  else print(m)
}
```

## Peak QC

```{r peakqc-table}
.tbl(chipData$peakStats)
```

```{r peakqc-width}
#| fig-height: 4
if (length(peaks)) {
  wd <- do.call(rbind, lapply(sampleNames, function(s)
    data.frame(sample = s, width = BiocGenerics::width(peaks[[s]]))))
  ggplot(wd, aes(sample, width, fill = sample)) + geom_violin(scale = "width") +
    scale_y_log10() + scale_fill_manual(values = sampleCol, guide = "none") +
    labs(title = "Peak width distribution", y = "width (bp, log10)", x = NULL) +
    theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

```{r peakqc-anno}
#| fig-height: 4
if (!is.null(chipData$consAnno)) {
  tryCatch(ChIPseeker::plotAnnoBar(chipData$consAnno),
           error = function(e) .noteHTML("Annotation bar plot unavailable."))
} else .noteHTML("Consensus annotation not available (no TxDb built from the GTF).")
```

### Signal concentration

Peaks ranked strongest-first (x); cumulative fraction of total signal (y). A steep rise = signal concentrated in a few strong peaks; a near-diagonal (dashed line) = signal spread evenly, itself a QC warning.

```{r peakqc-cumsignal}
#| fig-height: 4
cs <- chipData$cumSignal
if (length(cs)) {
  d <- do.call(rbind, lapply(names(cs), function(s) if (nrow(cs[[s]])) cbind(cs[[s]], sample = s)))
  if (!is.null(d))
    ggplot(d, aes(fracPeaks, cumFracSignal, colour = sample)) + geom_line(linewidth = 0.8) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
      scale_colour_manual(values = sampleCol) +
      labs(title = "Signal concentration", x = "fraction of peaks (ranked strongest first)",
           y = "cumulative fraction of signal") + theme_bw()
}
```

### Genomic footprint

```{r peakqc-footprint-table}
if (.have(chipData$footprint$summary)) .tbl(chipData$footprint$summary)
```

```{r peakqc-footprint-plot}
#| fig-height: 4
fp <- chipData$footprint
if (.have(fp$perChrom))
  ggplot(fp$perChrom, aes(chrom, bpPerMb, fill = sample)) +
    geom_col(position = "dodge") + scale_fill_manual(values = sampleCol) +
    labs(title = "Peak footprint per chromosome (bp per Mb)", x = NULL) +
    theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r peakqc-ideogram}
#| fig-height: 6
if (requireNamespace("karyoploteR", quietly = TRUE) && length(peaks)) {
  tryCatch({
    si <- chipData$seqinfo
    custom <- GRanges(GenomeInfoDb::seqnames(si), IRanges::IRanges(1, GenomeInfoDb::seqlengths(si)))
    kp <- karyoploteR::plotKaryotype(genome = custom)
    for (i in seq_along(peaks))
      karyoploteR::kpPlotRegions(kp, data = peaks[[i]], col = cvdPalette[i],
        r0 = (i - 1) / nSamples, r1 = i / nSamples)
  }, error = function(e) NULL)
}
```

```{r peakqc-ideogram-note}
#| output: asis
if (grepl("^Salmonella|bacter", param$refBuild, ignore.case = TRUE))
  cat("\n::: {.callout-note}\nBacterial genome: the ideogram is a linear contig view, not a cytoband karyotype. Each horizontal track is one sample's peaks.\n:::\n")
```

:::

# Comparison

::: {.panel-tabset}

## Overlap

```{r overlap-all}
#| fig-height: 5
if (nSamples < 2) {
  .noteHTML("Only one sample — pairwise overlap is not applicable.")
} else {
  m <- chipData$jaccard
  if (requireNamespace("ComplexHeatmap", quietly = TRUE))
    ComplexHeatmap::Heatmap(m, name = "Jaccard", column_title = "Pairwise base-pair Jaccard (all peaks)")
  else print(m)
}
```

```{r overlap-sets}
#| fig-height: 5
if (nSamples >= 2) {
  occ <- chipData$occupancy
  if (nSamples <= 20 && requireNamespace("UpSetR", quietly = TRUE)) {
    df <- as.data.frame(occ); df[] <- lapply(df, as.integer)
    tryCatch(UpSetR::upset(df, nsets = ncol(df), order.by = "freq"), error = function(e) NULL)
  } else {
    spec <- table(rowSums(occ))
    barplot(spec, xlab = "number of samples sharing a consensus region",
            ylab = "consensus regions", main = "Occupancy spectrum", col = cvdPalette[3])
  }
}
```

## Consensus

```{r consensus-text}
#| output: asis
cons <- chipData$consensus
cat(sprintf("Consensus = genomic regions occupied by **at least %s sample(s)** (minSamplesForConsensus). ",
            param$minSamplesForConsensus))
if (param$minSamplesForConsensus <= 1) {
  cat("At the current setting of 1 this is the **union of all peaks**; the `occupancy` column (and `occupancy_matrix.tsv`) lets you filter to shared regions.\n\n")
} else {
  cat("\n\n")
}
if (length(cons) > 0) {
  cat(sprintf("**%d consensus regions**, total %s bp. Occupancy spectrum (how many samples share each region):\n\n",
              length(cons), format(sum(BiocGenerics::width(cons)), big.mark = ",")))
}
```

```{r consensus}
if (length(cons) == 0) .noteHTML("No consensus regions at the chosen minSamplesForConsensus.") else
  .tbl(as.data.frame(table(occupancy = S4Vectors::mcols(cons)$occupancy)))
```

```{r consensus-width}
#| fig-height: 4
if (length(cons) > 0)
  ggplot(data.frame(width = BiocGenerics::width(cons)), aes(width)) +
    geom_histogram(bins = 40, fill = cvdPalette[2]) + scale_x_log10() +
    labs(title = "Consensus peak width", x = "width (bp, log10)") + theme_bw()
```

## Quantitative

```{r quant-cor}
#| fig-height: 5
mat <- chipData$normMat
if (is.null(mat) || nrow(mat) < 2 || nSamples < 2) {
  .noteHTML("No quantitative signal matrix available (needs >= 2 samples).")
} else {
  cc <- suppressWarnings(cor(mat, method = "spearman", use = "pairwise.complete.obs"))
  if (requireNamespace("ComplexHeatmap", quietly = TRUE))
    ComplexHeatmap::Heatmap(cc, name = "Spearman", column_title = "Consensus-signal sample correlation")
  else print(cc)
}
```

```{r quant-pca}
#| fig-height: 5
if (!is.null(mat) && nSamples >= 3) {
  pc <- tryCatch(prcomp(t(log1p(mat)), scale. = FALSE), error = function(e) NULL)
  if (!is.null(pc)) {
    d <- data.frame(PC1 = pc$x[, 1], PC2 = pc$x[, 2], sample = colnames(mat))
    lab <- if (requireNamespace("ggrepel", quietly = TRUE))
      ggrepel::geom_text_repel(show.legend = FALSE) else geom_text(vjust = -0.8, show.legend = FALSE)
    ggplot(d, aes(PC1, PC2, colour = sample, label = sample)) + geom_point(size = 3) + lab +
      scale_colour_manual(values = sampleCol, guide = "none") +
      labs(title = "PCA of consensus signal") + theme_bw()
  }
} else if (!is.null(mat)) .noteHTML("PCA needs at least 3 samples.")
```

:::

# Top peaks coverage

The `r param$topN` most-significant peaks (highest `r param$rankBy`, i.e. largest fold-enrichment / log2 ratio over background) across the whole dataset. For each locus, one coverage track per sample (BigWig signal ± `r param$profileExtend` bp), so you can read directly whether a strong site in one sample is also occupied in the others. The grey **peak** track marks the called peak; the **gene** track at the bottom shows any genes in the window, coloured by strand (**blue = + / forward**, **orange = − / reverse**) with arrowheads pointing in the direction of transcription.

```{r topcov}
#| fig-height: 6
#| results: asis
lc <- chipData$locusCov
if (length(lc) == 0) {
  cat('<div class="callout callout-note"><p>No BigWig tracks — per-locus coverage tracks are unavailable.</p></div>\n')
} else if (!requireNamespace("Gviz", quietly = TRUE)) {
  cat('<div class="callout callout-note"><p>Gviz is not installed — cannot draw coverage tracks.</p></div>\n')
} else {
  ## Gviz's .onLoad resets this to TRUE, so it MUST be set after the namespace
  ## is loaded (above) to allow non-UCSC contig names (Ensembl / bacterial RefSeq).
  options(ucscChromosomeNames = FALSE)
  for (i in seq_along(lc)) {
    L <- lc[[i]]
    chr <- as.character(GenomeInfoDb::seqnames(L$region))
    ymax <- max(L$cov, na.rm = TRUE); if (!is.finite(ymax) || ymax <= 0) ymax <- 1
    ## axis, then one coverage track per sample, then the peak call ...
    tracks <- list(Gviz::GenomeAxisTrack()); sizes <- 1
    for (s in sampleNames) {
      tracks <- c(tracks, Gviz::DataTrack(
        range = L$binGR, data = L$cov[, s], name = s, type = "h",
        col = sampleCol[[s]], col.histogram = sampleCol[[s]],
        fill.histogram = sampleCol[[s]], ylim = c(0, ymax)))
      sizes <- c(sizes, 2)
    }
    tracks <- c(tracks, Gviz::AnnotationTrack(L$peak, name = "peak", fill = "grey40", col = NA))
    sizes <- c(sizes, 0.6)
    ## ... and the gene track at the BOTTOM, coloured/arrowed by strand
    ## (blue = + / forward, orange = - / reverse).
    if (!is.null(chipData$genes)) {
      g <- IRanges::subsetByOverlaps(chipData$genes, L$region)
      if (length(g) > 0L) {
        g <- GenomeInfoDb::keepSeqlevels(g, chr, pruning.mode = "coarse")
        g$feature <- ifelse(as.character(BiocGenerics::strand(g)) == "-", "reverse", "forward")
        gtrack <- tryCatch(Gviz::GeneRegionTrack(
          g, name = "genes", transcriptAnnotation = "symbol", showId = TRUE,
          col = NA, fontcolor.group = "black",
          forward = "#0072B2", reverse = "#D55E00",
          arrowHeadWidth = 20, arrowHeadMaxWidth = 30),
          error = function(e) NULL)
        if (!is.null(gtrack)) { tracks <- c(tracks, gtrack); sizes <- c(sizes, 1.8) }
      }
    }
    Gviz::plotTracks(tracks, chromosome = chr,
                     from = BiocGenerics::start(L$region), to = BiocGenerics::end(L$region),
                     main = L$label, cex.main = 0.75, sizes = sizes)
  }
}
```

# Downloads

```{r downloads}
#| output: asis
cat(sprintf("The consensus files below contain regions occupied by **≥ %s sample(s)** (the `occupancy` column gives the exact count per region).\n\n",
            param$minSamplesForConsensus))
files <- c("consensus_peaks.bed", "occupancy_matrix.tsv", "consensus_signal_counts.tsv",
           "consensus_signal_normalized.tsv", "top_significant_peaks.bed", "qc_metrics.xlsx")
present <- files[file.exists(files)]
if (length(present)) {
  cat("Result files in the report folder:\n\n")
  for (f in present) cat(sprintf("- [%s](%s)\n", f, f))
} else cat("::: {.callout-note}\nResult files are written alongside this report in the job output folder.\n:::\n")
```

# SessionInfo

```{r session}
if (exists("ezSessionInfo", mode = "function")) ezSessionInfo() else sessionInfo()
```

:::
