---
params:
  reportTitle: "CountQC"
  # B-Fabric identifiers, shown in Report provenance and used to name the ZIP the
  # Download panel builds. Leave empty to auto-detect from the render path (a
  # gstore-style .../pXXXXX/oXXXXX_App_date/... yields both); pass explicitly when
  # rendering somewhere that has no such path.
  project: ""
  order: ""
title: "`r params$reportTitle`"
---

<!-- Styling comes from the fgczQuartoTemplate assets staged next to this file
     (_metadata.yml is applied by Quarto automatically, so there is nothing to
     declare here). ezRun stages them via makeQuartoReport(), which calls
     fgczQuartoTemplate::fgcz_render(). The optional tab colour/numbering and the
     Find/Download toolbar are render-time arguments there, not params. -->


```{r setup}
#| include: false
library(ezRun)
library(SummarizedExperiment)
library(DT)
library(kableExtra)
library(WGCNA)
library(plotly)
library(matrixStats)
library(reshape2)
library(tidyverse)
library(writexl)
library(gridExtra)
library(cowplot)
library(ggpubr)
library(cluster)
library(fpc)
library(clusterSim)
library(ggrepel)
library(htmlwidgets)
library(jsonlite)

## ── Data loading (qs2 with rds fallback, checks data/ first) ──
.fgcz_load <- function(name, nthreads = 4L) {
  paths <- c(
    file.path("data", paste0(name, ".qs2")),
    paste0(name, ".qs2"),
    file.path("data", paste0(name, ".rds")),
    paste0(name, ".rds")
  )
  for (p in paths) {
    if (file.exists(p)) {
      if (grepl("\\.qs2$", p)) return(qs2::qs_read(p, nthreads = nthreads))
      return(readRDS(p))
    }
  }
  stop("No .qs2 or .rds file found for: ", name)
}

## ── DT helper ─────────────────────────────────────────────────
.dt <- function(df, caption = NULL, pageLength = 10L,
                filter = "none", rownames = FALSE, escape = TRUE, ...) {
  DT::datatable(
    data = df, caption = caption, filter = filter,
    rownames = rownames, escape = escape,
    class = "compact stripe",
    options = list(
      pageLength = pageLength,
      scrollX = TRUE,
      dom = "tip",
      columnDefs = list(
        list(targets = "_all", className = "dt-left")
      )
    ),
    ...
  ) |> DT::formatStyle(columns = seq_len(ncol(df)),
                        `text-align` = "left",
                        `white-space` = "normal",
                        `word-wrap` = "break-word",
                        `max-width` = "300px")
}

## ── Output directories ────────────────────────────────────────
for (d in c("plots", "excel", "html")) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}

## ── Agent metadata writer ─────────────────────────────────────
.write_metadata <- function(...) {
  args <- list(...)
  args$generatedAt <- format(Sys.time(), "%Y-%m-%dT%H:%M:%S%z")
  args$softwareVersions <- list(
    R = paste0(R.version$major, ".", R.version$minor),
    quarto = tryCatch(system("quarto --version", intern = TRUE)[1],
                      error = function(e) "unknown"))
  jsonlite::write_json(args, path = "report-metadata.json",
                       pretty = TRUE, auto_unbox = TRUE, null = "null")
}

rawData <- .fgcz_load("rawData")
output  <- metadata(rawData)$output
param   <- metadata(rawData)$param
```

```{r prepare-data}
#| include: false
debug <- FALSE

seqAnno <- data.frame(rowData(rawData), row.names = rownames(rawData),
                      check.names = FALSE)
dataset <- data.frame(colData(rawData), row.names = colnames(rawData),
                      check.names = FALSE)

types <- data.frame(row.names = rownames(seqAnno))
for (nm in setdiff(na.omit(unique(seqAnno$type)), "")) {
  types[[nm]] <- seqAnno$type == nm
}

metadata(rawData)$design <- ezDesignFromDataset(dataset, param)
colData(rawData)$conds   <- ezConditionsFromDesign(metadata(rawData)$design,
                                                   maxFactors = 2)
metadata(rawData)$condColors <- lapply(metadata(rawData)$design, getCondsColors)
metadata(rawData)$sampleColors <- list()
for (cond in colnames(metadata(rawData)$design)) {
  metadata(rawData)$sampleColors[[cond]] <-
    set_names(metadata(rawData)$condColors[[cond]][metadata(rawData)$design[[cond]]],
              rownames(metadata(rawData)$design))
}
```

```{r fgcz-bfabric}
#| echo: false
## B-Fabric identity + provenance, resolved once and reused by the Download
## panel's ZIP name and the Report provenance tab.
##
## The ids are read from the SUSHI job description, NOT from getwd(). Under
## SUSHI the report is rendered in a scratch directory
## (/scratch/CountQC_<workunit>_<date>_temp<pid>/<name>) which contains neither
## the project nor the order, so sniffing the working directory yields nothing
## on exactly the runs that matter. param$resultDir carries both, e.g.
## "p36207/o41819_CountQC_2026-06-18--17-16-12".
##
## Sources are tried in order of trustworthiness and the first hit wins:
##   1. an explicit -P project/-P order passed to this render
##   2. a structured B-Fabric column on the output dataset, where one exists
##      (the order has "Order Id [B-Fabric]"; there is no project equivalent)
##   3. param$resultDir           (set by SUSHI for every real job)
##   4. output$getColumn("Report") (same string plus the report subdirectory)
##   5. getwd()                    (only useful when rendering by hand inside a
##                                  gstore-style path)
## Anything still unresolved stays "" and is simply not displayed.
##
## NOTE there is deliberately no Workunit row. SUSHI registers the workunit in
## B-Fabric only AFTER the job has produced its output, so at render time the id
## does not exist yet — it cannot be recovered here by any means.

## First non-empty value of a column on the output dataset, or "" if the column
## is absent (it varies by app).
.fgcz_meta <- function(col) {
  v <- tryCatch(unique(output$meta[[col]]), error = function(e) NULL)
  v <- as.character(v[!is.na(v)])
  v <- v[nzchar(v)]
  if (length(v)) v[1] else ""
}

.fgcz_id <- function(explicit, prefix) {
  x <- as.character(explicit %||% "")
  if (!nzchar(x)) {
    sources <- c(
      as.character(param$resultDir %||% ""),
      tryCatch(as.character(output$getColumn("Report")), error = function(e) ""),
      getwd()
    )
    pat <- sprintf("(?<![A-Za-z0-9])%s[0-9]{3,6}(?![0-9])", prefix)
    for (s in sources) {
      if (!nzchar(s)) next
      hits <- regmatches(s, gregexpr(pat, s, perl = TRUE))[[1]]
      ## Last match wins: a nested working copy under another project's tree
      ## still resolves to the directory actually being rendered.
      if (length(hits)) { x <- hits[length(hits)]; break }
    }
  }
  if (!nzchar(x)) return("")
  ## Accept a bare number ("36207") as well as a prefixed one ("p36207").
  if (!grepl(sprintf("^%s", prefix), x)) x <- paste0(prefix, x)
  x
}
## The order has a structured source, so prefer it over parsing a path; .fgcz_id
## prefixes a bare number, so "41819" becomes "o41819". The project has no such
## column and still falls back to the path sources.
.fgcz_order_explicit <- as.character(params$order %||% "")
if (!nzchar(.fgcz_order_explicit)) {
  .fgcz_order_explicit <- .fgcz_meta("Order Id [B-Fabric]")
}

.fgcz_project <- .fgcz_id(params$project, "p")
.fgcz_order   <- .fgcz_id(.fgcz_order_explicit, "o")

## Who rendered this, and when. Sys.info() is the OS account the render ran as —
## under SUSHI that is the service account, not the submitting scientist.
.fgcz_user <- unname(Sys.info()[["user"]])
if (!nzchar(.fgcz_user %||% "")) .fgcz_user <- Sys.getenv("USER", "unknown")
.fgcz_rendered <- format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")
```

::: {.panel-tabset}

# Overview

::: {.callout-note}
## What this report is

**CountQC** is a quality-control report for a *count matrix*: one row per
feature (gene or transcript), one column per sample. It does not test any
hypothesis — it answers "is this dataset sound enough to analyse?" before any
differential-expression work begins.

Read it in three passes:

1. **Count Statistics** — did every sample get enough reads, and is any single
   feature swamping a library?
2. **Correlation/Clustering** — do samples group the way the experimental
   design says they should? Unexpected grouping here usually means a batch or
   labelling problem, not biology.
3. **Clustering of High Variance Features** and **MDS** — what structure is
   actually driving the variance?
:::

<!-- TODO(visual abstract): the FGCZ report layout asks for a static diagram here
     showing this report type's purpose, input, analysis path and output — the
     equivalent of fgcz-report-overview.svg in the template. One needs authoring
     for CountQC; it is a per-report-type asset, not a run-specific figure. -->

```{r overview-input-summary}
#| echo: false
## Plain, human-readable summary of what went in. Identifiers, timestamps and
## software versions deliberately live in Session Info, not here.
.ov <- c(
  "Samples"          = ncol(rawData),
  "Features"         = nrow(rawData),
  "Feature level"    = if (!is.null(param$featureLevel)) param$featureLevel else NA,
  "Normalisation"    = if (!is.null(param$normMethod)) param$normMethod else NA,
  "Reference build"  = if (!is.null(param$refBuild)) param$refBuild else NA
)
.ov <- .ov[!is.na(.ov) & nzchar(as.character(.ov))]
knitr::kable(
  data.frame(Field = names(.ov), Value = as.character(.ov), check.names = FALSE),
  caption = "Input summary for this run."
)
```

# Settings

`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`

```{r check-enough-samples}
#| results: asis
if (ncol(rawData) < 2) {
  cat("Note: Statistics and Plots are not available for single sample experiments.", "\n")
  cat("Run the report again with multiple samples selected.", "\n")
  knitr::knit_exit()
}
```

```{r settings-table}
settings <- character()
settings["Reference:"]          <- param$refBuild
settings["Normalization method:"] <- param$normMethod
if (param$useSigThresh) {
  settings["Log2 signal threshold:"]    <- signif(log2(param$sigThresh), digits = 4)
  settings["Linear signal threshold:"]  <- signif(param$sigThresh, digits = 4)
}
settings["Feature level:"]        <- metadata(rawData)$featureLevel
settings["Number of features:"]   <- nrow(rawData)
settings["Data Column Used:"]     <- metadata(rawData)$countName

settings_df <- data.frame(Setting = settings, row.names = names(settings))
.dt(settings_df, caption = "Analysis settings", filter = "none",
    pageLength = 20L, rownames = TRUE)
```

```{r live-report}
#| echo: false
#| results: asis
if (exists("output") && !is.null(output)) {
  liveReportLink <- output$getColumn("Live Report")
  result <- EzResult(se = rawData)
  result$saveToFile(basename(output$getColumn("Live Report")))
  cat(paste0("[Live Report and Visualizations](", liveReportLink, ")", "{target=\"_blank\"}"), "\n")
}
```

# Data Files

The data files are in tabular text format and can also be opened with a spreadsheet program (*e.g.*, Excel).

When opening with Excel, do make sure that the gene symbols are loaded into a column formatted as 'text', this prevents
conversion of the symbols to dates. See [here](https://www.genenames.org/help/importing-gene-symbol-data-into-excel-correctly)

```{r prepare-signal}
#| include: false
assays(rawData)$signal <- ezNorm(assays(rawData)$counts,
                                 presentFlag = assays(rawData)$presentFlag,
                                 method = param$normMethod) + param$minSignal
signalRange <- range(assays(rawData)$signal, na.rm = TRUE)

signalCond <- rowsum(t(log2(assays(rawData)$signal)), group = colData(rawData)$conds)
signalCond <- 2^t(sweep(signalCond, 1,
                        table(colData(rawData)$conds)[rownames(signalCond)], FUN = "/"))

isPresentCond <- rowsum(t(assays(rawData)$presentFlag * 1), group = colData(rawData)$conds)
isPresentCond <- t(sweep(isPresentCond, 1,
                         table(colData(rawData)$conds)[rownames(isPresentCond)], FUN = "/")) >= 0.5
rowData(rawData)$isValid <- rowMeans(isPresentCond) >= 0.5
```

```{r write-data-files}
#| include: false
if (!is.null(assays(rawData)$presentFlag)) {
  combined <- interleaveMatricesByColumn(assays(rawData)$signal,
                                        assays(rawData)$presentFlag)
} else {
  combined <- assays(rawData)$signal
}
if (!is.null(seqAnno)) {
  combined <- cbind(seqAnno[rownames(combined), , drop = FALSE], combined)
}
if (!is.null(combined$featWidth)) {
  combined$featWidth <- as.integer(combined$featWidth)
}

combined$isPresent_50P <- rowMeans(assays(rawData)$presentFlag) >= 0.5

# Raw counts
countFile <- file.path("excel", paste0(ezValidFilename(param$name), "-raw-count.xlsx"))
counts <- as_tibble(assays(rawData)$counts, rownames = "Feature ID")
if (!is.null(rowData(rawData)$gene_name)) {
  counts <- left_join(counts,
                      as_tibble(rowData(rawData)[, "gene_name", drop = FALSE],
                                rownames = "Feature ID"))
  counts <- dplyr::select(counts, "Feature ID", gene_name, everything())
}
write_xlsx(counts, path = countFile)

# Normalised signal
signalFile <- file.path("excel", paste0(ezValidFilename(param$name), "-normalized-signal.xlsx"))
write_xlsx(combined, path = signalFile)

selectSignals <- grepl("Signal", colnames(combined))
combined$"Mean signal"    <- rowMeans(combined[, selectSignals], na.rm = TRUE)
combined$"Maximum signal" <- rowMaxs(as.matrix(combined[, selectSignals]), na.rm = TRUE)
topGenesPerSample <- apply(combined[, selectSignals], 2, function(col) {
  col <- sort(col, decreasing = TRUE)
  if (length(col) > 100) col <- col[1:100]
  return(names(col))
})
topGenes <- unique(as.character(topGenesPerSample))

combined <- combined[order(combined$"Maximum signal", decreasing = TRUE), , drop = FALSE]
useInInteractiveTable <- c("seqid", "gene_name", "Maximum signal",
                           "Mean signal", "description", "featWidth", "gc")
useInInteractiveTable <- intersect(useInInteractiveTable, colnames(combined))
tableLink <- file.path("html", sub(".xlsx", "-viewHighExpressionGenes.html",
                                   basename(signalFile)))
## Select top genes
combinedTopGenes <- combined[which(rownames(combined) %in% topGenes), ]
## Restrict number of table rows if necessary
topGenesDf <- head(combinedTopGenes[, useInInteractiveTable, drop = FALSE],
                   param$maxTableRows)
ezInteractiveTableRmd(topGenesDf, digits = 3,
                      colNames = c("ID", colnames(topGenesDf)),
                      title = paste("Showing the top", nrow(topGenesDf),
                                    "genes with the highest expression")) |>
  DT::saveWidget(tableLink)

rpkmFile <- file.path("excel", paste0(ezValidFilename(param$name), "-rpkm.xlsx"))
rpkm <- as_tibble(getRpkm(rawData), rownames = "Feature ID")
if (!is.null(rowData(rawData)$gene_name)) {
  rpkm <- left_join(rpkm,
                    as_tibble(rowData(rawData)[, "gene_name", drop = FALSE],
                              rownames = "Feature ID"))
  rpkm <- dplyr::select(rpkm, "Feature ID", gene_name, everything())
}
write_xlsx(rpkm, path = rpkmFile)

tpmFile <- file.path("excel", paste0(ezValidFilename(param$name), "-tpm.xlsx"))
tpm <- as_tibble(getTpm(rawData), rownames = "Feature ID")
if (!is.null(rowData(rawData)$gene_name)) {
  tpm <- left_join(tpm,
                   as_tibble(rowData(rawData)[, "gene_name", drop = FALSE],
                             rownames = "Feature ID"))
  tpm <- dplyr::select(tpm, "Feature ID", gene_name, everything())
}
write_xlsx(tpm, path = tpmFile)
```

```{r data-file-links}
#| echo: false
#| results: asis
#| message: false
for (each in c(countFile, signalFile, rpkmFile, tpmFile, tableLink)) {
  cat("\n")
  cat(paste0("[", each, "](", each, ")"))
  cat("\n")
}
```

# Count Statistics

::: {.panel-tabset}

## Read counts

```{r count-statistics}
#| layout-ncol: 2
toPlot <- tibble(samples      = colnames(rawData),
                 totalCounts  = signif(colSums(assays(rawData)$counts) / 1e6, digits = 3),
                 presentCounts = colSums(assays(rawData)$presentFlag),
                 percentages  = paste(signif(100 * presentCounts / nrow(rawData), digits = 2), "%"))
m <- list(l = 80, r = 80, b = 120, t = 60, pad = 0)   # b: room for rotated sample labels — raise if names are long

## Total reads
plot_ly(toPlot, x = ~samples,
        y = ~totalCounts, type = "bar") |>
  layout(title  = "Total reads",
         yaxis  = list(title = "Counts [Mio]"),
         height = 420,                 # plotly sizes in px; knitr fig-height is ignored for widgets
         width = min(7 + (ncol(rawData) - 10) * 0.3, 30) * 96,   # inches → px
         margin = m)

## Above genomic threshold 
plot_ly(toPlot, x = ~samples, y = ~presentCounts, type = "bar",
        text = ~percentages, textposition = "auto") |>
  layout(title  = "Genomic features with reads above threshold",
         yaxis  = list(title = "Counts"),
         height = 420,
         width = min(7 + (ncol(rawData) - 10) * 0.3, 30) * 96,
         margin = m)
```

::: {.callout-note}
Total reads per sample (left) and the number of genomic features with reads above the detection threshold (right). Interactive (plotly) plots ignore Quarto's `fig-cap`, so this is a manual caption.
:::

## Dominant genes

```{r dominant-genes}
#| fig-height: 5
#| fig-width: 7
#| fig-cap: "Top 3 over-represented genes per sample, drawn over each sample's distribution of expression fractions."
## Check for dominant genes in samples

countFracPerSample <- sweep(counts[, 3:ncol(counts)], 2,
                            colSums(counts[, 3:ncol(counts)]), FUN = "/")
rownames(countFracPerSample) <- make.unique(counts$gene_name)

df <- countFracPerSample |>
  as.data.frame() |>
  tibble::rownames_to_column("Gene") |>
  pivot_longer(-Gene, names_to = "Sample", values_to = "Fraction")

top3 <- df |>
  group_by(Sample) |>
  slice_max(Fraction, n = 3, with_ties = FALSE) |>
  mutate(rank = factor(row_number(), levels = 1:3)) |>
  ungroup()

p <- ggplot(df, aes(x = Sample, y = Fraction)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90")
p <- p + geom_point(data = top3, aes(colour = rank), size = 2,
                    position = position_dodge(width = 0.4))
p <- p + geom_text_repel(data = top3,
                         aes(label = Gene, colour = rank),
                         size          = 2.5,
                         nudge_y       = 0.02,
                         direction     = "y",
                         segment.size  = 0.2,
                         segment.alpha = 0.4,
                         max.overlaps  = Inf) +
  scale_colour_manual(values = c("red", "orange", "gold"), name = "Top-gene rank")
p <- p + theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        plot.margin = margin(5.5, 20, 5.5, 5.5, "pt"))
p <- p + labs(y = "Fraction of total expression",
              title = "Top 3 over-represented genes per sample")
p
```

:::

# Correlation/Clustering Plot

```{r prepare-data-correlation}
if (!is.null(seqAnno$IsControl)) {
  rowData(rawData)$isValid <- rowData(rawData)$isValid & !seqAnno$IsControl
}
```

```{r check-enough-features}
#| results: asis
if (sum(rowData(rawData)$isValid) < 10) {
  cat("Not enough valid features for further plots", "\n")
  knitr::knit_exit()
}
```

```{r filter-valid-features}
rawDataAll <- rawData
rawData    <- rawData[rowData(rawData)$isValid, ]
assays(rawData)$log2signal <- log2(assays(rawData)$signal +
                                     param$backgroundExpression)
assays(rawData)$log2signalCentered <- sweep(assays(rawData)$log2signal, 1,
                                            rowMeans(assays(rawData)$log2signal))

metadata(rawData)$design <- ezDesignFromDataset(dataset, param)
colData(rawData)$conds   <- ezConditionsFromDesign(metadata(rawData)$design,
                                                   maxFactors = 2)

if (param$selectByFtest) {
  f_test_results <- apply(assays(rawData)$log2signalCentered, 1, function(row) {
    data <- data.frame(Expression = row, Group = colData(rawData)$conds)
    fit  <- lm(Expression ~ Group, data = data)
    anova(fit)["Group", "Pr(>F)"]
  })
  metadata(rawData)$topGenes <- names(sort(f_test_results)[1:param$topGeneSize])
} else {
  metadata(rawData)$topGenes <- rownames(rawData)[
    head(order(rowSds(assays(rawData)$log2signal, na.rm = TRUE),
               decreasing = TRUE), param$topGeneSize)]
}
```

::: {.panel-tabset}

## Sample correlation

```{r sample-correlation-setup}
# Width grows with sample count (more samples → wider matrix); height stays fixed.
figure.width <- min(6 + (ncol(rawData) - 10), 30)
figure.height <- min(6 + (ncol(rawData) - 10) * 0.3, 30)
```

```{r plot-sample-correlation}
#| layout-ncol: 2
#| out-width: "100%"
#| fig-width: !expr "figure.width"
#| fig-height: !expr "figure.height"

## All genes
ezCorrelationPlot(cor(assays(rawData)$log2signal, use = "complete.obs"),
                  cond = colData(rawData)$conds,
                  condLabels = colData(rawData)$conds,
                  main = paste0("all present genes (", nrow(rawData), ")"),
                  colors = metadata(rawData)$sampleColors[[1]])

## Top genes
ezCorrelationPlot(cor(assays(rawData)$log2signal[metadata(rawData)$topGenes, ],
                      use = "complete.obs"),
                  cond = colData(rawData)$conds,
                  condLabels = colData(rawData)$conds,
                  main = paste0("top genes (", length(metadata(rawData)$topGenes), ")"),
                  colors = metadata(rawData)$sampleColors[[1]])
```

Pearson correlation of samples based on all detected (left) and top 100 most abundant (right) features.

## Sample clustering

```{r sample-clustering}
#| layout-ncol: 2
#| fig-height: 6
#| fig-width: !expr "min(7 + (ncol(rawData) - 10) * 0.3, 30)"
#| results: asis
#| fig-cap: "Hierarchical clustering (Ward D2) of samples based on all detected (left) and top 100 most abundant (right) features."

if (ncol(rawData) > 3) {
  ## All genes
  d  <- as.dist(1 - cor(assays(rawData)$log2signal, use = "complete.obs"))
  hc <- hclust(d, method = "ward.D2")
  plotDendroAndColors(hc,
                      colors = as.data.frame(bind_cols(metadata(rawData)$sampleColors)),
                      autoColorHeight = TRUE, hang = -0.1,
                      main = "all present genes")

  ## Top genes
  d  <- as.dist(1 - cor(assays(rawData)$log2signal[metadata(rawData)$topGenes, ],
                        use = "complete.obs"))
  hc <- hclust(d, method = "ward.D2")
  plotDendroAndColors(hc,
                      colors = as.data.frame(bind_cols(metadata(rawData)$sampleColors)),
                      autoColorHeight = TRUE, hang = -0.1,
                      main = paste("top", length(metadata(rawData)$topGenes), " genes"))
}
```

:::

# Clustering of High Variance Features

```{r clustering-high-variance}
#| fig-width: !expr "max(8, 4 + 0.15 * ncol(rawData))"
#| fig-height: 10
#| fig-cap: "Samples as columns, features as rows. Feaures have been row-centred (but not scaled.)"
#| results: asis
#| eval: !expr "!debug"
if (ncol(rawData) > 3) {
  varFeatures <- rownames(rawData)[
    head(order(rowSds(assays(rawData)$log2signal), decreasing = TRUE),
         param$maxGenesForClustering)]
  notVarFeatures <- rownames(rawData)[
    which(rowSds(assays(rawData)$log2signal) <= param$highVarThreshold)]
  varFeatures <- setdiff(varFeatures, notVarFeatures)
  param$highVarThreshold <- signif(
    min(rowSds(assays(rawData)$log2signal[varFeatures, ])), digits = 3)

  if (length(varFeatures) > param$minGenesForClustering) {
    cat("\n::: {.callout-note}\n")
    cat(paste0("Threshold for std. dev. of log2 signal across samples: **",
               param$highVarThreshold, "**\n\n"))
    cat(paste0("Number of features with high std. dev.: **",
               length(varFeatures), "**\n"))
    cat(":::\n\n")
    clusterColors <- hcl.colors(param$nSampleClusters, palette = "Spectral")

    clusterResult <- clusterPheatmap(
      x      = assays(rawData)$log2signalCentered[varFeatures, ],
      design = metadata(rawData)$design,
      param  = param, clusterColors = clusterColors,
      lim    = c(-param$logColorRange, param$logColorRange),
      doClusterColumns = TRUE,
      condColors = metadata(rawData)$condColors)

    xTmp <- enframe(clusterResult$clusterNumbers, "feature_id", value = "Cluster")
    xTmp <- left_join(xTmp,
                      as_tibble(rowData(rawData)[, "gene_name", drop = FALSE],
                                rownames = "feature_id"))
    xTmp <- arrange(xTmp, Cluster)
    write_tsv(xTmp, file = "cluster-heatmap-clusterMembers.txt")

    if (doGo(param, seqAnno)) {
      clusterResult <- goClusterResults(
        assays(rawData)$log2signalCentered[varFeatures, ],
        param, clusterResult,
        seqAnno          = seqAnno,
        universeProbeIds = rownames(seqAnno))
    }

    if (!is.null(clusterResult$GO)) {
      goTables <- goClusterTableRmd(param, clusterResult, seqAnno)
      if (doEnrichr(param)) {
        goAndEnrichr <- cbind(goTables$linkTable, goTables$enrichrTable)
      } else {
        goAndEnrichr <- goTables$linkTable
      }
      bgColors <- rep(gsub("FF$", "", clusterResult$clusterColors))
      rownames(goAndEnrichr) <- paste0('<font color="', bgColors,
                                       '">Cluster ', rownames(goAndEnrichr),
                                       '</font>')
      goKable <- kable(goAndEnrichr, escape = FALSE, row.names = TRUE, format = "html",
            caption = "GO categories of feature clusters") |>
        kable_styling(bootstrap_options = "striped",
                      full_width = F, position = "float_right") |>
        footnote(general = "Cluster font color corresponds to the row colors in the heatmap plot.")
      # Emit the float_right table BEFORE the plot so it sits beside the heatmap
      # at the top (a right-floated block can only sit alongside content that
      # follows it in source order; if drawn after the plot it drops below).
      # Blank-line-delimited raw HTML: auto-printing a kableExtra table as the
      # tail of a results:asis chunk leaves Pandoc in a verbatim state that
      # swallows ALL following content into an unclosed <pre><code>.
      cat("\n\n", as.character(goKable), "\n\n", sep = "")
    } else {
      cat("No information available", "\n")
    }

    plot(clusterResult$pheatmap$gtable)
  }
} else {
  varFeatures <- c()
}
```

```{r render-go-cluster}
#| eval: !expr "!debug"
if (ncol(rawData) > 3) {
  if (length(varFeatures) > param$minGenesForClustering) {
    if (!is.null(clusterResult$GO)) {
      file <- file.path(system.file("templates", package = "ezRun"),
                        "CountQC_goClusterTable.Rmd")
      file.copy(from = file, to = basename(file), overwrite = TRUE)
      rmarkdown::render(input = basename(file), envir = new.env(),
                        output_dir = "html", output_file = "goClusterTable.html",
                        quiet = TRUE)
    }
  }
}
```

```{r go-cluster-link}
#| results: asis
#| eval: !expr "!debug"
if (file.exists("html/goClusterTable.html")) {
  cat(paste0("[GO cluster tables](html/goClusterTable.html)"), "\n")
  cat("\n")
}
```

```{r cluster-heatmap-members}
#| results: asis
#| eval: !expr "!debug"
if (file.exists("cluster-heatmap-clusterMembers.txt")) {
  cat(paste0("[Heatmap cluster members](cluster-heatmap-clusterMembers.txt)"), "\n")
  cat("\n")
}
```

::: {.callout-note collapse="true" title="Evaluate number of gene clusters"}
Multiple metrics were computed for a comprehensive assessment.

* Compactness: Dunn Index, CH Index
* Separation: Silhouette Width, Davies-Bouldin Index
* Cluster Stability: Gap Statistic, Entropy

```{r cluster-scores}
#| eval: !expr "!debug"
if (length(varFeatures) > param$minGenesForClustering) {

  dist_matrix <- dist(assays(rawData)$log2signalCentered[varFeatures, ])
  hc          <- hclust(dist_matrix, method = "ward.D2")

  hclust_cluster <- function(X, k) {
    res          <- data.frame(clusters = rep(0, nrow(X)))
    dist_matrix  <- dist(X)
    hc           <- hclust(dist_matrix, method = "ward.D2")
    res$clusters <- cutree(hc, k = k)
    return(res)
  }
  gap_stat <- clusGap(assays(rawData)$log2signalCentered[varFeatures, ],
                      FUN = hclust_cluster, K.max = 10, B = 50)

  myClusterStats <- data.frame(nClusters = c(2:10), dunn = NA, ch = NA,
                               db = NA, silhouette = NA, entropy = NA,
                               gap = round(gap_stat$Tab[2:10, 3], 4))
  for (i in 2:10) {
    clusters       <- cutree(hc, k = i)
    cluster_scores <- cluster.stats(dist_matrix, clusters)
    myClusterStats$dunn[i - 1]      <- round(cluster_scores$dunn, 4)
    myClusterStats$ch[i - 1]        <- round(cluster_scores$ch, 4)
    myClusterStats$db[i - 1]        <- round(
      index.DB(assays(rawData)$log2signalCentered[varFeatures, ], clusters)$DB, 4)
    myClusterStats$silhouette[i - 1] <- round(cluster_scores$avg.silwidth, 4)
    myClusterStats$entropy[i - 1]   <- round(cluster_scores$entropy, 4)
  }

  datatable(myClusterStats,
            rownames = FALSE,
            options  = list(dom = "t")) |>
    formatStyle("dunn",
                background         = styleColorBar(range(myClusterStats$dunn), "lightblue"),
                backgroundSize     = "98% 15px",
                backgroundRepeat   = "no-repeat",
                backgroundPosition = "center") |>
    formatStyle("ch",
                background         = styleColorBar(range(myClusterStats$ch), "lightblue"),
                backgroundSize     = "98% 15px",
                backgroundRepeat   = "no-repeat",
                backgroundPosition = "center") |>
    formatStyle("db",
                background         = styleColorBar(range(myClusterStats$db), "lightgreen"),
                backgroundSize     = "98% 15px",
                backgroundRepeat   = "no-repeat",
                backgroundPosition = "center") |>
    formatStyle("silhouette",
                background         = styleColorBar(range(myClusterStats$silhouette), "lightgreen"),
                backgroundSize     = "98% 15px",
                backgroundRepeat   = "no-repeat",
                backgroundPosition = "center") |>
    formatStyle("entropy",
                background         = styleColorBar(range(myClusterStats$entropy), "lightsalmon"),
                backgroundSize     = "98% 15px",
                backgroundRepeat   = "no-repeat",
                backgroundPosition = "center") |>
    formatStyle("gap",
                background         = styleColorBar(range(myClusterStats$gap), "lightsalmon"),
                backgroundSize     = "98% 15px",
                backgroundRepeat   = "no-repeat",
                backgroundPosition = "center")
}
```
:::

# MDS Plot

```{r mds-message}
#| results: asis
if (ncol(rawData) <= 3) {
  cat("MDS plot is not possible with fewer than 4 samples. \n")
}
```

::::: {.fig-row}
:::: {.fig-main}
::: {.fig-grid}
```{r mds-plots}
# plotly in a (hidden) tab can't measure its container, so it needs an EXPLICIT
# size. layout(height/width) sizes the DRAWN plot; setting p$width/p$height sizes
# the htmlwidget CONTAINER too — without that, the container stays at knitr's
# fig-height (~480px) and leaves a tall empty band under each 380px plot (the
# "white space between 3D and 2D"). .fig-grid (fgcz.scss) flex-WRAPS the four
# widgets so the pair wraps to one-per-row when the column is narrow instead of
# overflowing layout-ncol's fixed 2-up grid into the callout.
mdsHeight <- 380   # px per plot — tweak to taste
mdsWidth  <- 460   # px per plot — two + gap must fit the .fig-main column
mds <- function(p) {                       # lock BOTH the plot and its container
  p$width <- mdsWidth; p$height <- mdsHeight
  plotly::layout(p, height = mdsHeight, width = mdsWidth)
}
if (ncol(rawData) > 3) {
  ezMdsPlotly(assays(rawData)$log2signal,
              design     = metadata(rawData)$design,
              ndim       = 3, main = "mdsPlot_PresentGenes 3D",
              condColors = metadata(rawData)$condColors[[1]]) |> mds()
}
if (ncol(rawData) > 3) {
  ezMdsPlotly(assays(rawData)$log2signal[metadata(rawData)$topGenes, ],
              design     = metadata(rawData)$design, ndim = 3,
              main       = "mdsPlot_TopGenes 3D",
              condColors = metadata(rawData)$condColors[[1]]) |> mds()
}
if (ncol(rawData) > 3) {
  ezMdsPlotly(assays(rawData)$log2signal,
              design     = metadata(rawData)$design, ndim = 2,
              main       = "mdsPlot_PresentGenes 2D",
              condColors = metadata(rawData)$condColors[[1]]) |> mds()
}
if (ncol(rawData) > 3) {
  ezMdsPlotly(assays(rawData)$log2signal[metadata(rawData)$topGenes, ],
              design     = metadata(rawData)$design, ndim = 2,
              main       = "mdsPlot_TopGenes 2D",
              condColors = metadata(rawData)$condColors[[1]]) |> mds()
}
```
:::
::::
:::: {.fig-side}
::: {.callout-note}
## Notes

MDS plots are like PCA plots, each dot is one of the samples in the dataset, and the closer the dots are to each other, the more similar they are in terms of feature expression/abundance, relative to the other samples.
:::
::::
:::::

# Scatter Plots by Conditions

```{r scatter-plot-setup}
#| results: asis
#| fig-height: 6
#| fig-width: 6
#| fig-cap: ""
scatterPlotData <- countQcScatterPlots(param, metadata(rawData)$design,
                                       colData(rawData)$conds, rawDataAll,
                                       signalCond, isPresentCond, types = types)
nPlots <- max(scatterPlotData$nPlots, 1)
```

::: {.panel-tabset}

```{r scatter-plot-all-pairs}
#| echo: false
#| message: false
#| results: asis
#| fig-width: !expr "min(max(ncol(signalCond) * 2, 4), 20)"
#| fig-height: !expr "min(max(ncol(signalCond) * 2, 4), 20) * (16 / 15)"
#| eval: !expr "ncol(rawData) > 3 & param$writeScatterPlots"
qcScatterFiles <- scatterPlotData$allPairs

cat("\n#### All pairs\n\n")
for (gridOfPlots in qcScatterFiles) {
  scatterPlots <- gridOfPlots$scatter
  axisLabels   <- gridOfPlots$axisLabels
  nItems       <- gridOfPlots$nItems

  grobList        <- list()
  idxLabel        <- 1
  idxScatterPlot  <- 1
  fontSize        <- max(4, 10 - nItems)

  for (idxRow in 1:(nItems + 1)) {
    if (idxRow == nItems + 1) {
      axisLabel <- ""
    } else {
      axisLabel <- axisLabels[idxLabel]
      idxLabel  <- idxLabel + 1
    }
    colLabel  <- text_grob(axisLabel, rot = 90, size = fontSize)
    grobList  <- c(grobList, list(colLabel))
  }

  for (rowIndex in 1:nItems) {
    for (colIndex in 1:nItems) {
      if (colIndex == nItems) {
        colLabel  <- text_grob(axisLabels[idxLabel], size = fontSize)
        grobList  <- c(grobList, list(scatterPlots[[idxScatterPlot]]), list(colLabel))
        idxLabel  <- idxLabel + 1
      } else {
        grobList  <- c(grobList, list(scatterPlots[[idxScatterPlot]]))
      }
      idxScatterPlot <- idxScatterPlot + 1
    }
  }

  plotGrob <- plot_grid(
    plotlist    = grobList,
    nrow        = nItems + 1,
    ncol        = nItems + 1,
    rel_heights = c(rep(1, nItems), 0.1),
    rel_widths  = c(0.1, rep(1, nItems)),
    byrow       = FALSE
  )

  title <- ggdraw() +
    draw_label(gridOfPlots$main, x = 0, hjust = 0) +
    theme(plot.margin = margin(0, 0, 0, 7))

  print(plot_grid(title, plotGrob, nrow = 2, ncol = 1, rel_heights = c(1, 15)))
  cat("\n\n*All pairwise scatter plots of log2 signal between samples; each point is a feature. Tight scatter along the diagonal indicates highly correlated samples.*\n\n")
}
cat("\n")
```

```{r scatter-plot-narrow}
#| echo: false
#| message: false
#| results: asis
#| fig-width: !expr "min(nPlots, 6) * 5"
#| fig-height: !expr "max(ceiling(nPlots / 6), 1) * 5"
#| out-width: !expr "paste0(min(nPlots + 1, 6) * 167, 'px')"
#| eval: !expr "ncol(rawData) > 3 & param$writeScatterPlots"
qcScatterFiles <- scatterPlotData$narrowPlots

if (length(qcScatterFiles) > 0) {
  for (i in 1:length(qcScatterFiles)) {
    cat("\n####", names(qcScatterFiles)[i], "\n\n")
    rowsOfPlots <- qcScatterFiles[[i]]
    for (rowOfPlots in rowsOfPlots) {
      scatterPlots <- rowOfPlots$scatters

      legend_temp <- get_legend(
        scatterPlots[[1]] +
          theme(legend.text = element_text(size = 10)) +
          guides(color = guide_legend(override.aes = list(size = 5)))
      )

      scatterPlots <- lapply(scatterPlots, function(x) {
        x +
          coord_fixed(xlim = c(1e1, NA), ylim = c(1e1, NA), expand = TRUE) +
          theme(legend.position = "none",
                text             = element_text(size = 10),
                aspect.ratio     = 1)
      })

      scatterPlots[["legend"]] <- legend_temp

      numRowsFinal <- rowOfPlots$nrow
      if (rowOfPlots$ncol * rowOfPlots$nrow > length(rowOfPlots$scatters)) {
        numColsFinal <- rowOfPlots$ncol
      } else {
        numColsFinal <- rowOfPlots$ncol + 1
        if (length(rowOfPlots$scatters) < numColsFinal * (numRowsFinal - 1)) {
          numRowsFinal <- numRowsFinal - 1
        }
      }

      grid.arrange(grobs   = scatterPlots,
                   nrow    = numRowsFinal,
                   ncol    = numColsFinal,
                   widths  = rep(6, numColsFinal),
                   heights = rep(6, numRowsFinal))
      cat("\n")
    }
    cat(paste0("\n*Scatter plots for **", names(qcScatterFiles)[i],
               "**: log2 signal per sample, points coloured by feature type. ",
               "Points off the diagonal are features differing between the conditions.*\n\n"))
  }
}
```

:::

# EnrichR for topMarkers

Based on top 500 genes per sample selected by TPM and filtered for known housekeeping genes reported in the [HRT atlas](https://housekeeping.unicamp.br/homePageGlobal.html).

```{r enrichr-markers}
#| echo: false
#| results: asis
#| warning: false
#| eval: !expr "unique(dataset$featureLevel) != 'smRNA'"
maxGenes <- 500
dat      <- tpm[, 3:ncol(tpm)]
rownames(dat) <- make.unique(tpm$gene_name)
topMarkersPerSample <- apply(dat, 2, function(col) {
  col <- sort(col, decreasing = TRUE)
  if (length(col) > maxGenes) col <- col[1:maxGenes]
  return(names(col))
})

hrtGenes <- ezRead.table("/srv/GT/databases/HRT/Human_Mouse_Common.csv",
                         sep = ";", row.names = NULL)
hrtGenes <- c(hrtGenes$Human, hrtGenes$Mouse)
topMarkers <- list()
for (j in 1:ncol(topMarkersPerSample)) {
  topMarkers[[colnames(topMarkersPerSample)[j]]] <-
    topMarkersPerSample[!topMarkersPerSample[, j] %in% hrtGenes, j]
}
topMarkers[["AllSamples"]] <- unique(unlist(lapply(topMarkers, head, n = maxGenes / 2),
                                            use.names = FALSE))
topMarkers <- topMarkers[c("AllSamples", colnames(topMarkersPerSample))]

jsCall <- paste0('enrich({list: "', sapply(topMarkers, paste, collapse = "\\n"),
                 '", popup: true});')
enrichrCalls <- paste0("<a href='javascript:void(0)' onClick='", jsCall,
                       "'>Analyse at Enrichr website</a>")
enrichrTable <- tibble(Sample              = names(topMarkers),
                       "# of posMarkers"   = lengths(topMarkers),
                       "Enrichr link"      = enrichrCalls)
enrichrKable <- kable(enrichrTable, format = "html", escape = FALSE,
      caption = paste0("GeneSet enrichment")) |>
  kable_styling("striped", full_width = F, position = "left")
# Blank-line-delimited raw HTML block (see GO table above): auto-printing a
# kableExtra table as the tail of a results:asis chunk leaves Pandoc verbatim
# and swallows everything after into an unclosed <pre><code>.
cat("\n\n", as.character(enrichrKable), "\n\n", sep = "")
```

# Input Dataset

```{r input-dataset}
#| echo: false
ezInteractiveTableRmd(values = dataset, digits = 4)
```

```{r save-qs2}
#| include: false
qs2::qs_save(rawData, "rawData_final.qs2", nthreads=param$cores)
```

# Session Info

::: {.panel-tabset}

## Report provenance

```{r report-metadata-marker}
#| echo: false
#| results: asis
## Hidden marker the Download panel reads to name its ZIP (order_/workunit_).
## Emitted only for ids that actually resolved, so the panel falls back to the
## report title rather than embedding an empty attribute.
.attrs <- c(
  if (nzchar(.fgcz_project)) sprintf(' data-project-id="%s"', .fgcz_project),
  if (nzchar(.fgcz_order))   sprintf(' data-order-id="%s"',   .fgcz_order),
  sprintf(' data-generated-by="%s"', "ezRun CountQC"),
  sprintf(' data-generated-at="%s"', .fgcz_rendered)
)
cat('<div id="fgcz-report-metadata" hidden', paste(.attrs, collapse = ""), '></div>')
```

```{r report-provenance}
#| echo: false
## Who/when/what this report is, above the R session dump. Rows with no value
## (e.g. no order number could be resolved) are dropped rather than shown blank.
.prov <- c(
  "Project"    = .fgcz_project,
  "Order"      = .fgcz_order,
  "Rendered by" = .fgcz_user,
  "Rendered at" = .fgcz_rendered,
  "Host"       = unname(Sys.info()[["nodename"]]),
  "R version"  = paste(R.version$major, R.version$minor, sep = ".")
)
.prov <- .prov[nzchar(.prov)]
knitr::kable(
  data.frame(Field = names(.prov), Value = unname(.prov), check.names = FALSE),
  caption = "Provenance for this rendered report."
)
```

## R session info

```{r session-info}
sessionInfo()
```

:::

:::

```{r agent-metadata}
#| include: false
.write_metadata(
  `@context`  = "https://fgcz.ch/report-schema/v1",
  reportType  = "count-qc",
  omicsType   = "rnaseq",
  organism    = if (!is.null(param$refBuild)) param$refBuild else NA,
  nSamples    = ncol(rawData),
  nFeatures   = nrow(rawData),
  normMethod  = param$normMethod,
  files       = list(
    countFile  = countFile,
    signalFile = signalFile,
    plots      = "plots/"
  )
)
```
