---
title: "`r if (exists('reportTitle')) reportTitle else 'SUSHI Report'`"
output: 
  html_document:
    mathjax: https://fgcz-gstore.uzh.ch/reference/mathjax.js
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package="ezRun")
    css: !expr system.file("templates/fgcz.css", package="ezRun")
editor_options: 
  chunk_output_type: inline
---

# {.tabset}


```{r setup, include=FALSE}
library(ezRun)
library(DT)
library(htmltools)
library(pheatmap)
library(ggplot2)
library(DescTools)
debug <- FALSE
#dict <- ezRead.table('/srv/gstore/projects/p33200/CountSpacer_Baggen_2021_2024-04-07--14-01-14/QC_Screen/QC_Screen-sgRNA_counts.txt', row.names = #NULL)
dict <- readRDS('dict.rds')
res <- dict[!dict$isControl,]
param <- readRDS('param.rds')
output <- readRDS("output.rds")
sortedCounts = log2(1+sort(res$Count))
meanCounts <- mean(sortedCounts)
upperCutOff = meanCounts + param$diffToLogMeanThreshold
lowerCutOff = meanCounts - param$diffToLogMeanThreshold
up_sgRNAs = sum(sortedCounts>upperCutOff)
relUp_sgRNA = round(100 * (up_sgRNAs/length(sortedCounts)), digits = 2)
down_sgRNAs = sum(sortedCounts<lowerCutOff)
relDown_sgRNA = round(100 * (down_sgRNAs/length(sortedCounts)), digits = 2)


stats <- readRDS('stats.rds')
data = data.frame(group = rep('Count', nrow(dict)), counts = c(dict$Count))
sampleName <- output$getNames()
```

## Overview

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

* Total Number of sgRNAs: `r nrow(res)`
* Median Reads per sgRNA = `r median(res$Count)`
* Mean Reads per sgRNA = `r mean(res$Count)`

## Diagnostic plots

```{r, fig.width=7, fig.height=6, echo=FALSE, message=FALSE, warning=FALSE, eval=!debug}
readFilteringStats = unlist(stats)[1:3]/10^6
barplot(readFilteringStats, main = 'Read Filtering Stats', ylab = 'Reads in [M]', col = c('darkblue', 'royalblue', 'lightblue'))
mappingStats = unlist(stats)[4:length(stats)]
par(mar=c(5.1,6.1,4.1,2.1))
barplot(mappingStats/10^6, las  = 1, main = 'Read Mapping Stats', ylab = 'Reads in [M]', col = 'royalblue')
par(mar=c(5.1,4.1,4.1,2.1))
```

```{r, fig.width=6, fig.height=5, echo=FALSE, message=FALSE, warning=FALSE, eval=!debug}
p <- ggplot(data, aes(x=group, y=counts))
  p <- p + geom_violin(fill="royalblue", alpha= 0.5, trim = FALSE, adjust = 0.5) 
  p <- p + geom_boxplot(width = 0.1)
  p <- p +  ggtitle(paste0(sampleName, '-ReadCount Distribution')) + ylab('ReadCount per sgRNA')
#  p <- p + theme(plot.title = element_text(size=12, face="bold"), axis.title.x =element_blank(), axis.text.x = element_text(angle=45,hjust=1))
  print(p)
```

```{r, fig.width=8, fig.height=5, echo=FALSE, message=FALSE, warning=FALSE, eval=!debug}
  h <- ggplot(dict, aes(x=log2(1+Count))) + geom_histogram(binwidth=0.1)
  h <- h +  ggtitle(paste0(sampleName, '-Histogram')) + ylab('Number of sgRNAs') + xlab('Log2 count per sgRNA')
  print(h)
```
  
```{r, fig.width=8, fig.height=5, echo=FALSE, message=FALSE, warning=FALSE, eval=!debug}
  mu = round(meanCounts, 2)
  plot(sortedCounts, pch = c(15), cex = 0.7, main = paste(sampleName, '- sgCount Overview'), ylab = 'log2(sgRNA Count)', ylim = c(0, max(max(sortedCounts), 1.05*upperCutOff)))
  abline(h = mean(sortedCounts))
  abline(h = upperCutOff, lty = 2)
  abline(h = lowerCutOff, lty = 2)
  text(length(sortedCounts)*0.05, 1.05*mean(sortedCounts), bquote(mu==.(mu)), cex = 0.8)
  text(length(sortedCounts)*0.8, 1.02*upperCutOff, paste0('#',up_sgRNAs, ' (',relUp_sgRNA,'%)' ), cex = 0.8)
  text(length(sortedCounts)*0.15, 0.96*lowerCutOff, paste0('#',down_sgRNAs, ' (',relDown_sgRNA,'%)' ), cex = 0.8)
  text(length(sortedCounts)*0.1, max(sortedCounts)*0.95, paste0('Gini-index=', round(Gini(res$Count),digits=3)), cex = 0.8)
```

## Data availability

**Expression matrix**

* raw counts allowing 0 and 1 mismatches: [`r paste0(sampleName,'-result.xlsx')`](`r paste0(sampleName,'-result.xlsx')`)

**Target based matrix**

[Underrepresented Targets - htmlFile](underrepresentedTargets.html)

[`r paste0(sampleName,'-targetBasedResult.xlsx')`](`r paste0(sampleName,'-targetBasedResult.xlsx')`)


## Methods 

* LibraryPath: `r param[['dictPath']]`
* LeftPattern: `r param[['leftPattern']]`
* RightPattern: `r param[['rightPattern']]`
* Alignment method: bowtie

## SessionInfo
```{r, echo=FALSE}
ezSessionInfo()
```
