---
title:  "CountSpacer - `r sampleName`"
author: "Functional Genomics Center Zurich"
output: 
  html_document:
    self_contained: false
    includes:
      in_header: fgcz_header.html
    css: fgcz.css
editor_options: 
  chunk_output_type: console
---

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

```{r setup, include=FALSE}
library(ezRun)
library(DT)
library(htmltools)
library(pheatmap)
library(ggplot2)
## ----------------------------------------------
## debug
# title:  "`r metadata(sce)$param$name`"
# sce <- readRDS("/srv/gstore/projects/p3271/SCCountQC_41088_2019-11-12--14-43-47/1834_SCCountQC/sce.rds")

debug <- FALSE

#param <- metadata(sce)$param
```

## CountSpacer Result {.tabset}

### Overview

* Total Number of sgRNAs: `r nrow(dict)`
* Median Reads per sgRNA = `r median(dict$Count)`
* Mean Reads per sgRNA = `r mean(dict$Count)`

### Diagnostic plots

```{r plots, fig.width=10, fig.height=10, echo=FALSE, message=FALSE, warning=FALSE, eval=!debug}
readFilteringStats = unlist(stats)[1:3]/10^6
barplot(readFilteringStats, main = 'Read Filtering Stats', ylab = 'Reads in [M]', col = c('darkblue', 'royalblue', 'lightblue'))

mappingStats = unlist(stats)[4:length(stats)]
barplot(mappingStats, las  = 1, main = 'Read Mapping Stats', ylab = 'Reads', col = 'royalblue')


p <- ggplot(data, aes(x=group, y=counts))
  p <- p + geom_violin(fill="royalblue", alpha= 0.5, trim = FALSE, adjust = 0.5) 
  p <- p + geom_boxplot(width = 0.1)
  p <- p +  ggtitle(paste0(sampleName, '-ReadCount Distribution')) + ylab('ReadCount per sgRNA')
  p <- p + theme(plot.title = element_text(size=15, face="bold"), axis.title.x =element_blank(), axis.text.x = element_text(angle=45,hjust=1))
  print(p)

  h <- ggplot(dict, aes(x=log2(1+Count))) + geom_histogram(binwidth=0.1)
  h <- h +  ggtitle(paste0(sampleName, '-Histogram')) + ylab('Number of sgRNAs') + xlab('Log2 count per sgRNA')
  print(h)
  
  plot(sortedCounts, pch = c(15), cex = 0.7, main = paste(sampleName, '- sgCount Overview'), ylab = 'log2(sgRNA Count)', ylim = c(0, max(max(sortedCounts), 1.05*upperCutOff)))
  abline(h = mean(sortedCounts))
  abline(h = upperCutOff, lty = 2)
  abline(h = lowerCutOff, lty = 2)
  text(length(sortedCounts)*0.05, 1.05*mean(sortedCounts), bquote(mu==.(mu)), cex = 0.8)
  text(length(sortedCounts)*0.8, 1.02*upperCutOff, paste0('#',up_sgRNAs, ' (',relUp_sgRNA,'%)' ), cex = 0.8)
  text(length(sortedCounts)*0.15, 0.96*lowerCutOff, paste0('#',down_sgRNAs, ' (',relDown_sgRNA,'%)' ), cex = 0.8)
```

### Data availability

**Expression matrix**

The raw counts allowing 0 and 1 mismatches are available here:

```{r raw count matrix, echo=FALSE, results='asis', message=FALSE, eval=!debug}
# Raw counts
zip("rawCount.zip", files=paste0(sampleName,'-result.txt'))
cat("\n")
cat(paste0("[", "rawCount.zip", "](", "rawCount.zip", ")"))
cat("\n")
```

**Target based matrix**

[Underrepresented Targets - htmlFile](underrepresentedTargets.html)

```{r full result, echo=FALSE, results='asis', message=FALSE, eval=!debug}
# Full result
zip("fullResult_targetView.zip", files=paste0(sampleName,'-targetBasedResult.txt'))
cat("\n")
cat(paste0("[", "fullResult_targetView.zip", "](", "fullResult_targetView.zip", ")"))
cat("\n")
```

### Methods 

LibraryPath: `r param[['dictPath']]`

LeftPattern: `r sub('\\.', '', param[['leftPattern']])`

RightPattern: `r param[['rightPattern']]`

Alignment method: bowtie

### SessionInfo
```{r, echo=FALSE}
sessionInfo()
```
