---
title: "`r if (exists('reportTitle')) reportTitle else 'SUSHI Report'`"
output: 
  html_document:
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package="ezRun")
    css: !expr system.file("templates/fgcz.css", package="ezRun")
editor_options: 
  chunk_output_type: inline
---

```{r setup countQC, include=FALSE}
library(knitr)
library(kableExtra)
library(SummarizedExperiment)
library(WGCNA)
library(plotly)
library(matrixStats)
library(reshape2)
library(tidyverse)
library(ezRun)
library(writexl)
library(gridExtra)
library(cowplot)
library(ggpubr)
#source("../../R/plots-reports.R")
```

```{r prepare data countQC, include=FALSE}
debug <- FALSE
if(!exists("rawData")){
  rawData <- readRDS("rawData.rds")
  # rawData <- readRDS("/srv/gstore/projects/p3682/CountQC_57360_2021-07-18--01-48-29/Count_QC/rawData.rds")
  # ctrl+enter in RStudio set cwd in the Rmd folder.
}
output <- metadata(rawData)$output
param <- metadata(rawData)$param
seqAnno <- data.frame(rowData(rawData), row.names=rownames(rawData),
                      check.names = FALSE)
dataset <- data.frame(colData(rawData), row.names=colnames(rawData),
                      check.names = FALSE)

types <- data.frame(row.names=rownames(seqAnno))
for (nm in setdiff(na.omit(unique(seqAnno$type)), "")){
  types[[nm]] = seqAnno$type == nm
}

metadata(rawData)$design <- ezDesignFromDataset(dataset, param)
colData(rawData)$conds <- ezConditionsFromDesign(metadata(rawData)$design,
                                                 maxFactors = 2)
metadata(rawData)$condColors <- lapply(metadata(rawData)$design, getCondsColors)
metadata(rawData)$sampleColors <- list()
for(cond in colnames(metadata(rawData)$design)){
  metadata(rawData)$sampleColors[[cond]] <- set_names(metadata(rawData)$condColors[[cond]][metadata(rawData)$design[[cond]]],
                                                     rownames(metadata(rawData)$design))
}
```

```{r check enough samples, echo=FALSE, results='asis'}
if (ncol(rawData) < 2){
  cat("Note: Statistics and Plots are not available for single sample experiments.", "\n")
  cat("Run the report again with multiple samples selected.", "\n")
  knit_exit()
}
```


```{r prepare signal, include=FALSE}
assays(rawData)$signal <- shiftZeros(ezNorm(assays(rawData)$counts,
                                            presentFlag=assays(rawData)$presentFlag,
                                            method=param$normMethod),
                                     param$minSignal)
signalRange <- range(assays(rawData)$signal, na.rm=TRUE)

signalCond <- rowsum(t(log2(assays(rawData)$signal)), group=colData(rawData)$conds)
signalCond <- 2^t(sweep(signalCond, 1,
                        table(colData(rawData)$conds)[rownames(signalCond)], FUN="/"))

isPresentCond <- rowsum(t(assays(rawData)$presentFlag * 1), group=colData(rawData)$conds)
isPresentCond <- t(sweep(isPresentCond, 1,
                         table(colData(rawData)$conds)[rownames(isPresentCond)], FUN="/")) >= 0.5
rowData(rawData)$isValid <- rowMeans(isPresentCond) >= 0.5
```

# {.tabset}

## Settings
```{r setting, echo=FALSE}
settings = character()
settings["Reference:"] = param$refBuild
settings["Normalization method:"] = param$normMethod
if (param$useSigThresh){
  settings["Log2 signal threshold:"] = signif(log2(param$sigThresh), digits=4)
  settings["Linear signal threshold:"] = signif(param$sigThresh, digits=4)
}
settings["Feature level:"] = metadata(rawData)$featureLevel
settings["Number of features:"] = nrow(rawData)
settings["Data Column Used:"] = metadata(rawData)$countName
  
kable(settings, row.names=TRUE, 
      col.names="Setting", format="html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F,
                position = "left")
```

```{r output live report, echo=FALSE, results='asis'}
if (exists("output") && !is.null(output)){
  liveReportLink = output$getColumn("Live Report")
  result = EzResult(se=rawData)
  result$saveToFile(basename(output$getColumn("Live Report")))
  cat(paste0("[Live Report and Visualizations](", liveReportLink, ")", "{target=\"_blank\"}"), "\n")
}
```

`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`


## Data Files

The data files are in tabular text format and can also be opened with a spreadsheet program (e.g. Excel).

When opening with Excel, do make sure that the Gene symbols are loaded into a column formatted as 'text' that prevents
conversion of the symbols to dates). See

(https://www.genenames.org/help/importing-gene-symbol-data-into-excel-correctly)


```{r write data files, include=FALSE}
if(!is.null(assays(rawData)$presentFlag)){
  combined = interleaveMatricesByColumn(assays(rawData)$signal, 
                                        assays(rawData)$presentFlag)
}else{
  combined = assays(rawData)$signal
}
if(!is.null(seqAnno)){
  combined = cbind(seqAnno[rownames(combined), , drop=FALSE], combined)
}
if(!is.null(combined$featWidth)){
  combined$featWidth = as.integer(combined$featWidth)
}
# Raw counts
countFile <- paste0(ezValidFilename(param$name), "-raw-count.xlsx")
counts <- as_tibble(assays(rawData)$counts, rownames="Feature ID")
if(!is.null(rowData(rawData)$gene_name)){
  counts <- left_join(counts,
                      as_tibble(rowData(rawData)[ ,"gene_name", drop=FALSE],
                                rownames="Feature ID"))
  counts <- dplyr::select(counts, "Feature ID", gene_name, everything())
}
write_xlsx(counts, path=countFile)
# normalized signal
signalFile <- paste0(ezValidFilename(param$name), "-normalized-signal.xlsx")
write_xlsx(combined, path=signalFile)

selectSignals = grepl("Signal", colnames(combined))
combined$"Mean signal" = rowMeans(combined[, selectSignals], na.rm=TRUE)
combined$"Maximum signal" = rowMaxs(as.matrix(combined[, selectSignals]),
                                    na.rm = TRUE)
topGenesPerSample = apply(combined[, selectSignals], 2, function(col){
  col = sort(col, decreasing = TRUE)
  if (length(col) > 100) col = col[1:100]
    return(names(col))
  })
    
topGenes = unique(as.character(topGenesPerSample))
    
combined = combined[order(combined$"Maximum signal", decreasing = TRUE), ,
                    drop=FALSE]
useInInteractiveTable = c("seqid", "gene_name", "Maximum signal", 
                          "Mean signal", "description", "featWidth", "gc")
useInInteractiveTable = intersect(useInInteractiveTable, colnames(combined))
tableLink = sub(".xlsx", "-viewHighExpressionGenes.html", signalFile)
## select top genes
combinedTopGenes = combined[which(rownames(combined) %in% topGenes),]
## restrict number of table rows if necessary
interactiveTable = head(combinedTopGenes[, useInInteractiveTable, drop=FALSE], 
                        param$maxTableRows)
nRows = ifelse(length(topGenes)>=param$maxTableRows, param$maxTableRows,
               length(topGenes))
ezInteractiveTable(interactiveTable, tableLink=tableLink, digits=3, 
                   colNames=c("ID", colnames(interactiveTable)),
                   title=paste("Showing the top", nRows, 
                               "genes with the highest expression"))

rpkmFile <- paste0(ezValidFilename(param$name), "-rpkm.xlsx")
rpkm <- as_tibble(getRpkm(rawData), rownames="Feature ID")
if(!is.null(rowData(rawData)$gene_name)){
  rpkm <- left_join(rpkm,
                    as_tibble(rowData(rawData)[ ,"gene_name", drop=FALSE],
                              rownames="Feature ID"))
  rpkm <- dplyr::select(rpkm, "Feature ID", gene_name, everything())
}
write_xlsx(rpkm, path=rpkmFile)

tpmFile <- paste0(ezValidFilename(param$name), "-tpm.xlsx")
tpm <- as_tibble(getTpm(rawData), rownames="Feature ID")
if(!is.null(rowData(rawData)$gene_name)){
  tpm <- left_join(tpm,
                   as_tibble(rowData(rawData)[ ,"gene_name", drop=FALSE],
                             rownames="Feature ID"))
  tpm <- dplyr::select(tpm, "Feature ID", gene_name, everything())
}
write_xlsx(tpm, path=tpmFile)

```

```{r add data files link, echo=FALSE, results='asis', message=FALSE}
for(each in c(countFile, signalFile, rpkmFile, tpmFile, tableLink)){
  cat("\n")
  cat(paste0("[", each, "](", each, ")"))
  cat("\n")
}
```

## Count Statistics
```{r count statistics, echo=FALSE, fig.width=min(7+(ncol(rawData)-10)*0.3, 30), message=FALSE, warning=FALSE}
toPlot <- tibble(samples=colnames(rawData),
                 totalCounts=signif(colSums(assays(rawData)$counts) / 1e6,
                                    digits=3),
                 presentCounts=colSums(assays(rawData)$presentFlag),
                 percentages=paste(signif(100*presentCounts/nrow(rawData),
                                          digits=2), "%"))
m <- list(
  l = 80,
  r = 80,
  b = 200,
  t = 100,
  pad = 0
)
## Total reads
plot_ly(toPlot, x=~samples, 
        y=~totalCounts, type="bar") %>%
  layout(title="Total reads",
         yaxis = list(title = "Counts [Mio]"),
         margin = m
  )
plot_ly(toPlot, x=~samples, y=~presentCounts, type="bar",
        text=~percentages, textposition = 'auto') %>%
  layout(title="Genomic features with reads above threshold",
         yaxis = list(title = "Counts"),
         margin = m
  )
```

### Gini Index per sample 
```{r gini statistics, echo=FALSE, fig.width=min(7+(ncol(rawData)-10)*0.3, 30), message=FALSE, warning=FALSE}
require(DescTools)
giniPerSample <- apply(assays(rawData)$counts,2, Gini)

par(mar=c(15.1,4.1,4.1,2.1))
barplot(giniPerSample, col='royalblue', las = 2, ylab = 'Gini-index for raw counts')
par(mar=c(5.1,4.1,4.1,2.1))
```

## Correlation/Clustering Plot

```{r prepare data correlation, include=FALSE}
if (!is.null(seqAnno$IsControl)){
  rowData(rawData)$isValid <- rowData(rawData)$isValid & !seqAnno$IsControl
}
  
if(sum(rowData(rawData)$isValid) < 10){
  cat("Not enough valid features for further plots", "\n")
  knit_exit()
}

rawDataAll <- rawData
rawData <- rawData[rowData(rawData)$isValid, ]
assays(rawData)$log2signal <- log2(assays(rawData)$signal +
                                     param$backgroundExpression)
assays(rawData)$log2signalCentered <- sweep(assays(rawData)$log2signal, 1 ,
                                            rowMeans(assays(rawData)$log2signal))
metadata(rawData)$topGenes <- rownames(rawData)[head(order(rowSds(assays(rawData)$log2signal, na.rm=TRUE),
                                         decreasing = TRUE), param$topGeneSize)]
```

### Sample correlation
```{r sample correlation data, include=FALSE}
figure.height <- min(max(7, ncol(rawData) * 0.3), 30)
```

```{r plot sample correlation, echo=FALSE, fig.height=figure.height, fig.width=figure.height*1.25}
## All genes
ezCorrelationPlot(cor(assays(rawData)$log2signal, use="complete.obs"),
                  cond=colData(rawData)$conds, condLabels=colData(rawData)$conds,
                  main=paste0("all present genes (",
                              nrow(rawData), ")"), 
                  colors=metadata(rawData)$sampleColors[[1]])

## Top genes
ezCorrelationPlot(cor(assays(rawData)$log2signal[metadata(rawData)$topGenes, ], use="complete.obs"),
                  cond=colData(rawData)$conds, condLabels=colData(rawData)$conds,
                  main=paste0("top genes (", length(metadata(rawData)$topGenes), ")"),
                  colors=metadata(rawData)$sampleColors[[1]])
```

```{r sample clustering, echo=FALSE, fig.height=7, fig.width=min(7 + (ncol(rawData)-10)*0.3, 30), results='asis'}
if (ncol(rawData) > 3){
  ## All genes
  d <- as.dist(1-cor(assays(rawData)$log2signal, use="complete.obs"))
  hc <- hclust(d, method="ward.D2")
  plotDendroAndColors(hc, 
                      colors=as.data.frame(bind_cols(metadata(rawData)$sampleColors)),
                      autoColorHeight=TRUE, hang = -0.1,
                      main="all present genes")

  ## Top genes
  d <- as.dist(1-cor(assays(rawData)$log2signal[metadata(rawData)$topGenes, ], use="complete.obs"))
  hc <- hclust(d, method="ward.D2")
  plotDendroAndColors(hc, 
                      colors=as.data.frame(bind_cols(metadata(rawData)$sampleColors)),
                      autoColorHeight=TRUE, hang = -0.1,
                      main=paste("top", length(metadata(rawData)$topGenes), " genes"))
}
```

## Clustering of High Variance Features

```{r Clustering of High Variance Features, echo=FALSE, message=FALSE, fig.width=max(8,4+0.15*ncol(rawData)), fig.height=10, results='asis', eval=!debug}
if(ncol(rawData) > 3){
  varFeatures <- rownames(rawData)[head(order(rowSds(assays(rawData)$log2signal),
                                              decreasing=TRUE),
                                        param$maxGenesForClustering)]
  notVarFeatures <- rownames(rawData)[which(rowSds(assays(rawData)$log2signal) <=
                                              param$highVarThreshold)]
  varFeatures <- setdiff(varFeatures, notVarFeatures)
  param$highVarThreshold <- signif(min(rowSds(assays(rawData)$log2signal[varFeatures, ])),
                                   digits=3)
    
  if(length(varFeatures) > param$minGenesForClustering){
    cat(paste("Threshold for std. dev. of log2 signal across samples:",
              param$highVarThreshold), "\n")
    cat("\n")
    cat(paste("Number of features with high std. dev.:", length(varFeatures)), "\n")
    cat("\n")
    clusterColors <- hcl.colors(param$nSampleClusters, palette = "Spectral")
    clusterResult <-
      clusterPheatmap(x=assays(rawData)$log2signalCentered[varFeatures, ], 
                      design = metadata(rawData)$design,
                      param = param, clusterColors = clusterColors,
                      lim = c(-param$logColorRange, param$logColorRange),
                      doClusterColumns = TRUE,
                      condColors = metadata(rawData)$condColors)

    xTmp <- enframe(clusterResult$clusterNumbers, "feature_id", value = "Cluster")
    xTmp <- left_join(xTmp, as_tibble(rowData(rawData)[, "gene_name", drop=FALSE],
                                      rownames="feature_id"))
    xTmp <- arrange(xTmp, Cluster)
    write_tsv(xTmp, file="cluster-heatmap-clusterMembers.txt")

    clusterResult$pheatmap$gtable
  
    if (doGo(param, seqAnno)){
      clusterResult = goClusterResults(assays(rawData)$log2signalCentered[varFeatures, ],
                                       param, clusterResult, 
                                       seqAnno=seqAnno,
                                       universeProbeIds=rownames(seqAnno))
    }
    
    if (!is.null(clusterResult$GO)){
      goTables = goClusterTableRmd(param, clusterResult, seqAnno)
      if (doEnrichr(param)){
        goAndEnrichr = cbind(goTables$linkTable, goTables$enrichrTable)
      } else {
        goAndEnrichr = goTables$linkTable
      }
      bgColors = rep(gsub("FF$", "", clusterResult$clusterColors))
      rownames(goAndEnrichr) <- paste0('<font color="', bgColors,
                                     '">Cluster ', rownames(goAndEnrichr),
                                     '</font>')
    kable(goAndEnrichr, escape=FALSE, row.names=TRUE, format = "html",
          caption="GO categories of feature clusters") %>%
      kable_styling(bootstrap_options = "striped",
                    full_width = F, position = "float_right") %>%
      footnote(general="Cluster font color corresponds to the row colors in the heatmap plot.")
    } else {
      cat("No information available", "\n")
    }
  }
}
```

```{r render go cluster, echo=FALSE, results='hide', message=FALSE, warning=FALSE, eval=!debug}
if (ncol(rawData) > 3){
  if (length(varFeatures) > param$minGenesForClustering){
    if (!is.null(clusterResult$GO)){
    ## GO Cluster tables
    file <- file.path(system.file("templates", package="ezRun"), 
                      "CountQC_goClusterTable.Rmd")
    file.copy(from=file, to=basename(file), overwrite=TRUE)
    rmarkdown::render(input=basename(file), envir = new.env(),
                      output_dir=".", output_file="goClusterTable.html",
                      quiet=TRUE)
    }
  }
}
```

```{r go cluster table link, echo=FALSE, results='asis', eval=!debug}
if (file.exists("goClusterTable.html")){
  cat(paste0("[GO cluster tables](", "goClusterTable.html", ")"), "\n")
  cat("\n")
}
```

```{r cluster heatmap members, echo=FALSE, results="asis", eval=!debug}
if(file.exists("cluster-heatmap-clusterMembers.txt")){
  cat(paste0("[Heatmap cluster members](", "cluster-heatmap-clusterMembers.txt", ")"), 
      "\n")
  cat("\n")
}
```

## MDS Plot
```{r MDS section, echo=FALSE, results='asis', message=FALSE, eval=TRUE}
if(ncol(rawData) <= 3){
  cat("MDS plot is not possible with fewer than 4 samples. \n")
}
```

```{r MDS plot 3D present, echo=FALSE, message=FALSE, warning=FALSE, fig.height=7, fig.width=9, eval=TRUE}
if(ncol(rawData) > 3){
  ## 3D scatter plot is strange. The plots are messy when use htmltools::tagList()
  ## subplot doesn't work either.
  ezMdsPlotly(assays(rawData)$log2signal,
              design=metadata(rawData)$design,
              ndim=3, main="mdsPlot_PresentGenes 3D",
              condColors=metadata(rawData)$condColors[[1]])
}
if(ncol(rawData) > 3){
  ezMdsPlotly(assays(rawData)$log2signal[metadata(rawData)$topGenes, ],
              design=metadata(rawData)$design, ndim=3,
              main="mdsPlot_TopGenes 3D",
              condColors=metadata(rawData)$condColors[[1]])
}
if(ncol(rawData) > 3){
  ezMdsPlotly(assays(rawData)$log2signal,
              design=metadata(rawData)$design, ndim=2,
              main="mdsPlot_PresentGenes 2D",
              condColors=metadata(rawData)$condColors[[1]])
}
if(ncol(rawData) > 3){
  ezMdsPlotly(assays(rawData)$log2signal[metadata(rawData)$topGenes, ],
              design=metadata(rawData)$design, ndim=2,
              main="mdsPlot_TopGenes 2D",
              condColors=metadata(rawData)$condColors[[1]])
}
```

## Scatter Plots by Conditions

```{r scatter plot setup, eval=TRUE, echo=FALSE, message=FALSE, results='asis'}
scatterPlotData <- countQcScatterPlots(param, metadata(rawData)$design,
                                        colData(rawData)$conds, rawDataAll,
                                        signalCond, isPresentCond, types=types)
nPlots <- max(scatterPlotData$nPlots, 1)
```

```{r scatter plot all pairs, echo=FALSE, message=FALSE, results='asis', fig.width=min(max(ncol(signalCond) * 2, 4), 20), fig.height=min(max(ncol(signalCond) * 2, 4), 20) * (16 / 15), eval=ncol(rawData) > 3 & param$writeScatterPlots}

qcScatterFiles <- scatterPlotData$allPairs

cat("\n")
cat("#### allPairs \n")
for(gridOfPlots in qcScatterFiles){
  # Get the relevant objects
  scatterPlots <- gridOfPlots$scatter
  axisLabels <- gridOfPlots$axisLabels
  nItems <- gridOfPlots$nItems
  
  # Prepare variables for adding axis labels and plots to the grob list
  grobList <- list()
  idxLabel <- 1
  idxScatterPlot <- 1
  fontSize <- max(4, 10 - nItems)

  # Add first col of labels
  for (idxRow in 1:(nItems + 1)) {
    if (idxRow == nItems + 1) {
      axisLabel <- ""  # Empty square at bottom-left
    } else {
      axisLabel <- axisLabels[idxLabel]
      idxLabel <- idxLabel + 1
    }
    colLabel <- text_grob(axisLabel, rot=90, size=fontSize)
    grobList <- c(grobList, list(colLabel))
  }
  
  # Add the last row of labels
  for (rowIndex in 1:nItems) {
    for (colIndex in 1:nItems) {
      if (colIndex == nItems) {  # We are in the last row, so add label after
        colLabel <- text_grob(axisLabels[idxLabel], size=fontSize)
        grobList <- c(grobList, list(scatterPlots[[idxScatterPlot]]), list(colLabel))
        idxLabel <- idxLabel + 1
      } else {
        grobList <- c(grobList, list(scatterPlots[[idxScatterPlot]]))
      }
      idxScatterPlot <- idxScatterPlot + 1
    }
  }
  
  # Create the grob associated with the plots
  plotGrob <- plot_grid(
    plotlist=grobList,
    nrow=nItems + 1,
    ncol=nItems + 1,
    rel_heights=c(rep(1, nItems), 0.1),
    rel_widths=c(0.1, rep(1, nItems)),
    byrow=FALSE
  )
  
  # Create the title grob
  title <- ggdraw() +
  draw_label(
    gridOfPlots$main,
    x = 0,
    hjust = 0
  ) +
  theme(
    # add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7)
  )
  
  # Combine title and plot grobs and plot
  print(plot_grid(
    title,
    plotGrob,
    nrow=2,
    ncol=1,
    rel_heights=c(1, 15)
  ))
  cat("\n")
}
cat("\n")
```

```{r scatter plot narrow, echo=FALSE, message=FALSE, results='asis', eval=ncol(rawData) > 3 & param$writeScatterPlots, fig.width=min(nPlots, 6) * 5, fig.height=max(ceiling(nPlots/6), 1) * 5, out.width=paste0(min(nPlots + 1, 6) * 167, "px")}


qcScatterFiles <- scatterPlotData$narrowPlots

if(length(qcScatterFiles) > 0){
  for(i in 1:length(qcScatterFiles)){ #
    cat("\n")
    cat("####", names(qcScatterFiles)[i], "\n")
    rowsOfPlots <- qcScatterFiles[[i]]
    for(rowOfPlots in rowsOfPlots){
      
      # Get scatters
      scatterPlots <- rowOfPlots$scatters
      
      # Extract legend
      legend_temp <- get_legend(
        scatterPlots[[1]] + 
          theme(legend.text = element_text(size = 10)) +
          guides(color = guide_legend(override.aes = list(size = 5)))
      )
      
      # Set plot themes
      scatterPlots <- lapply(scatterPlots, function(x){
        x  + 
          coord_fixed(xlim = c(1e1, NA), ylim = c(1e1, NA), expand = TRUE) +
          theme(legend.position="none", 
                text = element_text(size=10), 
                aspect.ratio = 1)
      })
      
      # Set legend at end
      scatterPlots[["legend"]] <- legend_temp
      
      # Get number of final columns and rows
      numRowsFinal <- rowOfPlots$nrow
      if (rowOfPlots$ncol * rowOfPlots$nrow > length(rowOfPlots$scatters)) {
        numColsFinal <- rowOfPlots$ncol
      } else {
        numColsFinal <- rowOfPlots$ncol + 1
        if (length(rowOfPlots$scatters) < numColsFinal * (numRowsFinal - 1)) {
          numRowsFinal <- numRowsFinal - 1
        }
      }
      
      grid.arrange(grobs=scatterPlots, 
                   nrow=numRowsFinal,
                   ncol=numColsFinal,
                   widths=rep(6, numColsFinal),
                   heights=rep(6, numRowsFinal))
      cat("\n")
    }
    cat("\n")
  }
}

```

## Input Dataset
```{r input, echo=FALSE}
ezInteractiveTableRmd(values=dataset, digits=4)
```

```{r saveRDS, include=FALSE}
saveRDS(rawData, "rawData_final.rds")
```

## SessionInfo
```{r, echo=FALSE}
format(Sys.time(), '%Y-%m-%d %H:%M:%S')
ezSessionInfo()
```
