---
params:
  reportTitle: "CRISPR Screen QC"
  # B-Fabric identifiers, shown in Session Info > Report provenance and used to
  # name the ZIP the Download panel builds. Leave empty to auto-detect from the
  # render path / SUSHI job description.
  project: ""
  order: ""
title: "`r params$reportTitle`"
---

<!-- Styling comes from the fgczQuartoTemplate assets staged next to this file by
     ezRun::makeQuartoReport() -> fgczQuartoTemplate::fgcz_render(); there is no
     format: block to declare here. Tab colour/numbering and the Find/Download
     toolbar are render-time arguments of makeQuartoReport(), not params. -->

```{r setup}
#| include: false
library(ezRun)
library(seqLogo)
library(ggplot2)
library(DT)
library(knitr)
library(scales)

## ── Data loading (qs2 with rds fallback, checks data/ first) ──
.fgcz_load <- function(name, nthreads = 4L) {
  paths <- c(
    file.path("data", paste0(name, ".qs2")),
    paste0(name, ".qs2"),
    file.path("data", paste0(name, ".rds")),
    paste0(name, ".rds")
  )
  for (p in paths) {
    if (file.exists(p)) {
      if (grepl("\\.qs2$", p)) return(qs2::qs_read(p, nthreads = nthreads))
      return(readRDS(p))
    }
  }
  stop("No .qs2 or .rds file found for: ", name)
}

## ── Compact DT helper (search box + horizontal scroll, no pagination noise) ──
.dt <- function(df, caption = NULL, pageLength = 10L,
                filter = "none", rownames = FALSE, escape = TRUE, ...) {
  DT::datatable(
    data = df, caption = caption, filter = filter,
    rownames = rownames, escape = escape,
    class = "compact stripe",
    options = list(
      pageLength = pageLength, scrollX = TRUE, dom = "tip",
      columnDefs = list(list(targets = "_all", className = "dt-left"))
    ),
    ...
  )
}

## ── Colourblind-safe palette (Okabe-Ito; report author has red-green CVD) ──
.cvd <- c("#0072B2", "#E69F00", "#009E73", "#CC79A7",
          "#56B4E9", "#D55E00", "#F0E442", "#000000")

## ── Gini index of a count vector (0 = perfectly even, 1 = maximally skewed) ──
## x is coerced to double: a library has tens of thousands of sgRNAs, so the
## integer products in the formula overflow 32-bit and silently return NA.
.gini <- function(x) {
  x <- as.numeric(sort(x[!is.na(x)]))
  n <- length(x)
  if (n == 0L || sum(x) == 0) return(NA_real_)
  sum((2 * seq_len(n) - n - 1) * x) / (n * sum(x))
}

## ── Lorenz curve points for a count vector ──
.lorenz <- function(x) {
  x <- as.numeric(sort(x[!is.na(x)]))
  if (length(x) == 0L || sum(x) == 0) {
    return(data.frame(p = c(0, 1), L = c(0, 1)))
  }
  data.frame(p = c(0, seq_along(x) / length(x)),
             L = c(0, cumsum(x) / sum(x)))
}

output <- .fgcz_load("output")
param  <- .fgcz_load("param")
input  <- .fgcz_load("input")
data   <- .fgcz_load("data")

dataset <- input$meta
samples <- names(data$matchedLibCounts)
if (is.null(samples)) samples <- names(data$countsPerLib)
nReads  <- suppressWarnings(as.numeric(param$nReads))

## Sequence-logo width scales with read length (one PWM column per base): short
## reads (~36 bp) stay compact, long reads (150 bp) get a wide enough figure that
## individual positions remain legible. Rendered full width; the template's
## lightbox lets the reader click to zoom to base resolution.
.readLen <- tryCatch(
  max(vapply(data$PWMs, function(p) length(ic(p)), integer(1))),
  error = function(e) 50L)
.seqlogoWidth <- max(9, min(32, round(0.16 * .readLen)))

theme_set(theme_bw(base_size = 12))
```

```{r provenance}
#| include: false
## Minimal B-Fabric provenance (project/order/user/time), resolved once for the
## Session Info tab and the Download-panel ZIP name. Mirrors CountQC.qmd.
`%||%` <- function(a, b) if (is.null(a) || !length(a) || !nzchar(a[1])) b else a
.fgcz_meta <- function(col) {
  v <- tryCatch(unique(output$meta[[col]]), error = function(e) NULL)
  v <- as.character(v[!is.na(v)]); v <- v[nzchar(v)]
  if (length(v)) v[1] else ""
}
.fgcz_id <- function(explicit, prefix) {
  x <- as.character(explicit %||% "")
  if (!nzchar(x)) {
    sources <- c(as.character(param$resultDir %||% ""),
                 tryCatch(as.character(output$getColumn("Report")),
                          error = function(e) ""),
                 getwd())
    pat <- sprintf("(?<![A-Za-z0-9])%s[0-9]{3,6}(?![0-9])", prefix)
    for (s in sources) {
      if (!nzchar(s)) next
      hits <- regmatches(s, gregexpr(pat, s, perl = TRUE))[[1]]
      if (length(hits)) { x <- hits[length(hits)]; break }
    }
  }
  if (!nzchar(x)) return("")
  if (!grepl(sprintf("^%s", prefix), x)) x <- paste0(prefix, x)
  x
}
.fgcz_order_explicit <- as.character(params$order %||% "")
if (!nzchar(.fgcz_order_explicit)) {
  .fgcz_order_explicit <- .fgcz_meta("Order Id [B-Fabric]")
}
.fgcz_project <- .fgcz_id(params$project, "p")
.fgcz_order   <- .fgcz_id(.fgcz_order_explicit, "o")
.fgcz_user <- unname(Sys.info()[["user"]])
if (!nzchar(.fgcz_user %||% "")) .fgcz_user <- Sys.getenv("USER", "unknown")
.fgcz_rendered <- format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")
```

```{r per-sample-metrics}
#| include: false
## Per-sample QC metrics, computed on the MATCHED library (the reference the
## reads actually map to) so that representation numbers are not diluted by the
## dozens of unrelated libraries in the merged MAGeCK reference.
summaryDf <- NULL
for (nm in samples) {
  mlc <- data$matchedLibCounts[[nm]]
  if (is.null(mlc)) next
  matchedLib <- data$matchedLibs[[nm]]
  nSgRNA     <- length(mlc)
  mappedMatched <- sum(mlc)
  totalReads <- if (!is.null(data$countSummary) &&
                    nm %in% data$countSummary$Sample) {
    as.numeric(data$countSummary$Reads[data$countSummary$Sample == nm][1])
  } else nReads
  summaryDf <- rbind(summaryDf, data.frame(
    Sample          = nm,
    `Matched library` = matchedLib,
    `Reads`         = totalReads,
    `Mapped to library` = mappedMatched,
    `Mapping rate (%)`  = 100 * mappedMatched / max(totalReads, 1),
    `sgRNAs in library` = nSgRNA,
    `Zero-count sgRNAs (%)` = 100 * mean(mlc == 0),
    `Mean reads / sgRNA` = mappedMatched / max(nSgRNA, 1),
    ## Zeros expected from pure Poisson sampling at this depth: exp(-mean
    ## reads/sgRNA). The sampling "floor" a perfect library cannot beat, so
    ## observed zero-count above this is real dropout, not just shallow reads.
    `Expected zero (%)` = 100 * exp(-mappedMatched / max(nSgRNA, 1)),
    `Gini index`    = .gini(mlc),
    check.names = FALSE, stringsAsFactors = FALSE
  ))
}
```

::: {.panel-tabset}

# Overview

::: {.panel-tabset}

## At a glance

::: {.callout-note title="What this report is"}
**CRISPR Screen QC** checks whether a pooled CRISPR-screen sequencing sample is
sound *before* any screen analysis (MAGeCK test / MLE) begins. Reads are
subsampled, adapter/length-trimmed with **fastp**, and counted against every
installed sgRNA library with **MAGeCK count**. Three questions matter:

1. **Identity & mapping** — do the reads map, and to the *intended* library?
   (Overview + Library representation → *Mapping per library*.)
2. **Representation** — is the library evenly covered, or have many sgRNAs
   dropped to zero? A skewed library biases every downstream hit call.
   (Library representation → *sgRNA representation*.)
3. **Read structure** — do the reads show the expected constant scaffold around
   a variable spacer? (Read patterns.)

*Reading the numbers — mind the subsampling:* this report is normally generated
from **subsampled** reads (see *Parameters & inputs*), typically only ~1 read per
sgRNA. **Mapping rate** is a proportion and so is depth-robust — it should be high
(screen libraries usually map well above 60 %, the dashed guide). **Zero-count
fraction** and **Gini index**, in contrast, are inflated by shallow depth alone:
at ~1 read/sgRNA even a flawless library shows ~30 % zeros and a high Gini.
So there is *no* fixed pass/fail cut-off for these two here — read them as
**relative** comparisons between samples sequenced at the same depth. On the
zero-count panel the diamonds mark the zeros expected from pure Poisson sampling
at each sample's depth; a bar sitting well above its diamond is real dropout
beyond sampling, and a sample that is an outlier from the rest is the flag to act
on.
:::

The figure below summarises the key numbers per sample, each computed on the
library the sample actually maps to.

```{r glance-plot}
#| echo: false
#| fig-width: 8
#| fig-height: 4.5
#| out-width: "72%"
#| fig-cap: "Per-sample QC on the matched library. Dashed line: mapping-rate guide (depth-robust). Diamonds on the zero-count panel: zeros expected from Poisson sampling at each sample's depth — no fixed guide is drawn for zero-count or Gini because both depend on sequencing depth (see note above)."
if (!is.null(summaryDf)) {
  metricLevels <- c("Mapping rate (%)", "Zero-count sgRNAs (%)", "Gini index")
  long <- rbind(
    data.frame(Sample = summaryDf$Sample, metric = "Mapping rate (%)",
               value = summaryDf$`Mapping rate (%)`),
    data.frame(Sample = summaryDf$Sample, metric = "Zero-count sgRNAs (%)",
               value = summaryDf$`Zero-count sgRNAs (%)`),
    data.frame(Sample = summaryDf$Sample, metric = "Gini index",
               value = summaryDf$`Gini index`)
  )
  long$metric <- factor(long$metric, levels = metricLevels)
  long$Sample <- factor(long$Sample, levels = rev(summaryDf$Sample))
  ## Only mapping rate gets a fixed guide; zero-count and Gini are depth-dependent
  ## so no absolute cut-off is drawn (see the note above the figure).
  guides <- data.frame(
    metric = factor("Mapping rate (%)", levels = metricLevels),
    xint = 60)
  ## Depth-aware reference for the zero-count panel: Poisson-expected zeros.
  expZero <- data.frame(
    metric = factor("Zero-count sgRNAs (%)", levels = metricLevels),
    Sample = factor(summaryDf$Sample, levels = rev(summaryDf$Sample)),
    value = summaryDf$`Expected zero (%)`)
  ## Per-facet axis domains: percentages on [0,100], Gini on [0,1]. facet
  ## free_x auto-ranges from 0; invisible points at the upper bound fix each
  ## panel's maximum (ggh4x is not available here).
  lims <- data.frame(
    metric = factor(metricLevels, levels = metricLevels),
    Sample = long$Sample[1],
    value = c(100, 100, 1))
  ggplot(long, aes(x = value, y = Sample)) +
    geom_col(width = 0.7, fill = .cvd[1], orientation = "y") +
    geom_vline(data = guides, aes(xintercept = xint),
               linetype = "dashed", colour = "grey40") +
    geom_point(data = expZero, aes(x = value, y = Sample),
               shape = 18, size = 2, colour = "grey20") +
    geom_blank(data = lims, aes(x = value, y = Sample)) +
    facet_wrap(~metric, scales = "free_x", nrow = 1) +
    labs(x = NULL, y = NULL)
} else {
  cat("No per-sample metrics available.")
}
```

::: {.callout-note collapse="true" title="Exact per-sample QC numbers"}
```{r glance-table}
#| echo: false
if (!is.null(summaryDf)) {
  showDf <- summaryDf
  showDf$`Mapping rate (%)`       <- round(showDf$`Mapping rate (%)`, 1)
  showDf$`Zero-count sgRNAs (%)`  <- round(showDf$`Zero-count sgRNAs (%)`, 1)
  showDf$`Expected zero (%)`      <- round(showDf$`Expected zero (%)`, 1)
  showDf$`Mean reads / sgRNA`     <- round(showDf$`Mean reads / sgRNA`, 2)
  showDf$`Gini index`             <- round(showDf$`Gini index`, 3)
  .dt(showDf, caption = "Per-sample QC on the matched library.")
} else {
  cat("Not available.")
}
```
:::

::: {.callout-note collapse="true" title="MAGeCK count summary (all merged libraries)"}
These are MAGeCK's own numbers, computed across the *whole* merged reference
(every installed library at once). Because most libraries are irrelevant to any
given sample, the overall Gini and zero-count values here are expected to be
high and are **not** a measure of the real library's quality — use the
matched-library numbers above for that.

```{r mageck-summary-table}
#| echo: false
if (!is.null(data$countSummary)) {
  .dt(data$countSummary, caption = "MAGeCK count summary (merged reference).")
} else {
  cat("MAGeCK count summary not available for this run.")
}
```
:::

## Parameters & inputs

```{r params-table}
#| echo: false
.pv <- c(
  "Subsampled reads"        = as.character(param$nReads),
  "Minimum read length"     = as.character(param$length_required),
  "Adapter trimming (fastp)" = "on",
  "5' trim (front1)"        = as.character(param$trim_front1),
  "3' trim (tail1)"         = as.character(param$trim_tail1),
  "Top sgRNAs listed"       = as.character(param$topFeatures),
  "sgRNA library path"      = as.character(param$libPath)
)
.pv <- .pv[!is.na(.pv) & nzchar(.pv)]
knitr::kable(
  data.frame(Parameter = names(.pv), Value = unname(.pv), check.names = FALSE),
  caption = "Processing parameters for this run.")
```

```{r input-summary}
#| echo: false
knitr::kable(
  data.frame(
    Field = c("Samples", "Libraries screened against",
              "sgRNAs in reference (merged)"),
    Value = c(length(samples),
              length(data$sgRNAPerLib),
              format(sum(data$sgRNAPerLib), big.mark = ",")),
    check.names = FALSE),
  caption = "Inputs to this run.")
```

::: {.callout-note collapse="true" title="Input dataset table"}
```{r input-dataset}
#| echo: false
.dt(dataset, caption = "Input dataset for this run.", pageLength = 20L)
```
:::

:::

# Library representation

::: {.panel-tabset}

## Mapping per library

Each sample is counted against **all** installed libraries at once. The reads of
a sound sample should pile up on a single, intended library; substantial signal
spread across several libraries points to contamination or a mislabelled sample.

```{r mapping-per-lib}
#| echo: false
#| fig-width: 8
#| fig-height: 4
#| out-width: "80%"
#| fig-cap: "Percentage of subsampled reads mapping to each sgRNA library (heatmap; only libraries reaching >=0.05% in at least one sample are shown). A sound sample is bright on a single row."
mapLong <- NULL
for (nm in samples) {
  cpl <- data$countsPerLib[[nm]]
  if (is.null(cpl)) next
  mapLong <- rbind(mapLong, data.frame(
    Sample = nm, Library = names(cpl),
    Pct = 100 * as.numeric(cpl) / max(nReads, 1),
    stringsAsFactors = FALSE))
}
keep <- with(mapLong, tapply(Pct, Library, max))
keep <- names(keep)[keep >= 0.05]
mapPlot <- mapLong[mapLong$Library %in% keep, ]
mapPlot$Library <- factor(mapPlot$Library,
  levels = names(sort(tapply(mapPlot$Pct, mapPlot$Library, max))))
mapPlot$Sample <- factor(mapPlot$Sample, levels = samples)
ggplot(mapPlot, aes(x = Sample, y = Library, fill = Pct)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  ## Label only non-trace cells, and flip text colour so it always contrasts
  ## with the viridis fill (dark text on the bright high-value cells, white on
  ## the dark low-value ones).
  geom_text(aes(label = ifelse(Pct >= 1, round(Pct), ""),
                colour = Pct > 55), size = 2.6, show.legend = FALSE) +
  scale_colour_manual(values = c(`TRUE` = "grey10", `FALSE` = "white"),
                      guide = "none") +
  scale_fill_viridis_c(name = "Reads (%)", limits = c(0, 100)) +
  labs(x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## sgRNA representation

For the matched library of each sample, how evenly are reads spread across its
sgRNAs? The **distribution** should be a single, reasonably tight peak; a long
left tail toward zero means many guides are poorly represented. The **Lorenz
curve** restates the Gini index visually — the further it bows below the
diagonal, the more skewed the library. Note that at the shallow **subsampled**
depth used here even a perfect library looks skewed, so compare curves *between
samples* (same depth) rather than against an absolute standard.

```{r count-distribution}
#| echo: false
#| fig-width: 7.5
#| fig-height: 4.5
#| out-width: "72%"
#| fig-cap: "Per-sgRNA read-count distribution (log10, +1 pseudocount) within each sample's matched library, as a violin with an inset boxplot. A tight box high above zero is good; a long tail to 1 means many under-represented guides."
distLong <- NULL
for (nm in samples) {
  mlc <- data$matchedLibCounts[[nm]]
  if (is.null(mlc)) next
  distLong <- rbind(distLong, data.frame(
    Sample = nm, count = as.numeric(mlc), stringsAsFactors = FALSE))
}
distLong$Sample <- factor(distLong$Sample, levels = rev(samples))
ggplot(distLong, aes(x = Sample, y = count + 1)) +
  geom_violin(fill = .cvd[1], colour = NA, alpha = 0.55, scale = "width") +
  geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.2,
               outlier.colour = "grey60") +
  scale_y_log10(labels = scales::comma) +
  coord_flip() +
  labs(x = NULL, y = "Read count per sgRNA (+1, log10)")
```

```{r lorenz-curve}
#| echo: false
#| fig-width: 5.5
#| fig-height: 5
#| out-width: "56%"
#| fig-cap: "Lorenz curve of sgRNA counts within each matched library, coloured by Gini index. The dashed diagonal is perfectly even representation; the further a curve bows below it, the more skewed the library."
lz <- NULL
for (nm in samples) {
  mlc <- data$matchedLibCounts[[nm]]
  if (is.null(mlc)) next
  d <- .lorenz(mlc)
  d$Sample <- nm
  d$Gini <- .gini(mlc)
  lz <- rbind(lz, d)
}
ggplot(lz, aes(x = p, y = L, group = Sample, colour = Gini)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.7) +
  scale_colour_viridis_c(name = "Gini") +
  coord_equal() +
  labs(x = "Cumulative fraction of sgRNAs",
       y = "Cumulative fraction of reads")
```

::: {.callout-note collapse="true" title="Screened libraries (sgRNA counts per reference)"}
```{r screened-libs}
#| echo: false
sgTab <- data.frame(Library = names(data$sgRNAPerLib),
                    `#sgRNAs` = as.integer(data$sgRNAPerLib),
                    check.names = FALSE)
sgTab <- sgTab[order(-sgTab$`#sgRNAs`), ]
knitr::kable(sgTab, row.names = FALSE,
             caption = "Number of sgRNAs per installed reference library.")
```
:::

:::

# Top sgRNAs

How concentrated are the reads? The bar below is the share of each matched
library's mapped reads captured by its most abundant sgRNAs. A high share means
a few guides dominate the library — a sign of skew or a jackpot artefact. The
individual top sgRNAs (with gene targets) are in the collapsible table.

```{r top-sgrna-plot}
#| echo: false
#| fig-width: 7.5
#| fig-height: 4
#| out-width: "70%"
#| fig-cap: "Share of each matched library's mapped reads captured by its top sgRNAs."
topN <- suppressWarnings(as.integer(param$topFeatures))
concDf <- NULL
for (nm in names(data$topFeatureResults)) {
  tf  <- data$topFeatureResults[[nm]]
  mlc <- data$matchedLibCounts[[nm]]
  if (is.null(tf) || nrow(tf) == 0 || is.null(mlc) || sum(mlc) == 0) next
  concDf <- rbind(concDf, data.frame(
    Sample = nm,
    Share = 100 * sum(tf$sample1) / sum(as.numeric(mlc)),
    stringsAsFactors = FALSE))
}
concDf$Sample <- factor(concDf$Sample, levels = concDf$Sample[order(concDf$Share)])
ggplot(concDf, aes(x = Sample, y = Share)) +
  geom_col(width = 0.7, fill = .cvd[2]) +
  coord_flip() +
  labs(x = NULL,
       y = sprintf("Reads in top %s sgRNAs (%% of matched library)", topN))
```

::: {.callout-note collapse="true" title="Top sgRNAs table"}
```{r top-sgrna-table}
#| echo: false
myTable <- NULL
for (nm in names(data$topFeatureResults)) {
  tf <- data.frame(data$topFeatureResults[[nm]])
  if (nrow(tf) > 0) {
    tf[["Sample"]] <- nm
    colnames(tf) <- c("sgRNA", "Gene", "Count", "Library", "Sample")
    myTable <- rbind(myTable, tf)
  }
}
.dt(myTable, caption = "Top sgRNAs per sample.", pageLength = 20L)
```
:::

# Read patterns

Base composition of the reads as a sequence logo. A well-behaved screen shows a
constant scaffold (tall, single-letter columns) flanking the variable spacer
region (mixed columns). Loss of the expected constant bases indicates a trimming
or library-prep problem. The figure width scales with read length; for long
reads click a logo to zoom to base resolution.

```{r seqlogo}
#| echo: false
#| results: asis
#| fig-width: !expr .seqlogoWidth
#| fig-height: 3.6
#| fig-dpi: 100
#| out-width: "100%"
consensusSeqs <- character()
for (nm in names(data$PWMs)) {
  cat("\n\n####", nm, "\n\n")
  seqLogo(data$PWMs[[nm]])
  cons <- limma::strsplit2(consensus(data$PWMs[[nm]]), split = "")
  cons[ic(data$PWMs[[nm]]) < 1] <- "N"
  consensusSeqs[nm] <- paste(cons, collapse = "")
  cat("\n\n")
}
```

::: {.callout-note collapse="true" title="Consensus sequences"}
```{r consensus-table}
#| echo: false
knitr::kable(
  data.frame(Sample = names(consensusSeqs), Consensus = consensusSeqs,
             row.names = NULL, check.names = FALSE),
  caption = "Per-sample consensus (low-information positions shown as N).")
```
:::

# Session Info

::: {.panel-tabset}

## Report provenance

```{r report-metadata-marker}
#| echo: false
#| results: asis
.attrs <- c(
  if (nzchar(.fgcz_project)) sprintf(' data-project-id="%s"', .fgcz_project),
  if (nzchar(.fgcz_order))   sprintf(' data-order-id="%s"',   .fgcz_order),
  sprintf(' data-generated-by="%s"', "ezRun CrisprScreenQC"),
  sprintf(' data-generated-at="%s"', .fgcz_rendered)
)
cat('<div id="fgcz-report-metadata" hidden', paste(.attrs, collapse = ""), '></div>')
```

```{r report-provenance}
#| echo: false
.prov <- c(
  "Project"     = .fgcz_project,
  "Order"       = .fgcz_order,
  "Rendered by" = .fgcz_user,
  "Rendered at" = .fgcz_rendered,
  "Host"        = unname(Sys.info()[["nodename"]]),
  "R version"   = paste(R.version$major, R.version$minor, sep = ".")
)
.prov <- .prov[nzchar(.prov)]
knitr::kable(
  data.frame(Field = names(.prov), Value = unname(.prov), check.names = FALSE),
  caption = "Provenance for this rendered report.")
```

## R session info

```{r session-info}
#| echo: false
ezSessionInfo()
```

:::

:::
