---
params:
  reportTitle: "Differential Peak Analysis"
title: "`r params$reportTitle`"
---

<!-- Styling comes from the fgczQuartoTemplate assets staged next to this file
     (_metadata.yml is applied by Quarto automatically). ezRun stages them via
     makeQuartoReport(), which calls fgczQuartoTemplate::fgcz_render(). -->

```{r setup}
#| include: false
library(ezRun)
library(DESeq2)
library(SummarizedExperiment)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(ggplot2)
library(ggrepel)
library(ComplexHeatmap)
library(RColorBrewer)
library(DT)
library(htmltools)

## ── Data loading (qs2 with rds fallback, checks data/ first) ──────────────
.fgcz_load <- function(name, nthreads = 4L) {
  paths <- c(
    file.path("data", paste0(name, ".qs2")),
    paste0(name, ".qs2"),
    file.path("data", paste0(name, ".rds")),
    paste0(name, ".rds")
  )
  for (p in paths) {
    if (file.exists(p)) {
      if (grepl("\\.qs2$", p)) return(qs2::qs_read(p, nthreads = nthreads))
      return(readRDS(p))
    }
  }
  stop("No .qs2 or .rds file found for: ", name)
}

dds      <- .fgcz_load("dds")
peakAnno <- .fgcz_load("peakAnno")
param    <- .fgcz_load("param")
output   <- .fgcz_load("output")

sampleGroup  <- param$sampleGroup
refGroup     <- param$refGroup
lfcThreshold <- 1
fdrThreshold <- 0.05

## Robust fit (handles comparisons without replicates); see ezRun app code.
deseqOut     <- runDiffPeakDESeq(dds, sampleGroup, refGroup, peakAnno)
mlDds        <- deseqOut$dds
res          <- deseqOut$res
noReplicates <- deseqOut$noReplicates

peakTable <- makeDiffPeakTable(res, noReplicates, lfcThreshold, fdrThreshold)
hasAnno   <- "geneName" %in% colnames(peakTable)

## Variance-stabilized signal, computed once and reused by PCA and the heatmap.
vsd <- diffPeakVST(mlDds, noReplicates)

groupColors <- c("#009E73", "#CC79A7")
names(groupColors) <- c(refGroup, sampleGroup)

## ── Summary counts ────────────────────────────────────────────────────────
nTested   <- sum(!is.na(peakTable$log2FoldChange))
nCand     <- sum(peakTable$candidate, na.rm = TRUE)
nUp       <- sum(peakTable$candidate & peakTable$direction == "up", na.rm = TRUE)
nDown     <- sum(peakTable$candidate & peakTable$direction == "down", na.rm = TRUE)

## ── Full result table for download ────────────────────────────────────────
resultsFile <- "differential_peaks.txt"
utils::write.table(
  peakTable |> dplyr::arrange(padj),
  file = resultsFile, sep = "\t", quote = FALSE, row.names = FALSE
)

## ── Candidate gene lists for Enrichr ──────────────────────────────────────
candGenes <- function(dir = NULL) {
  if (!hasAnno) return(character(0))
  tbl <- peakTable |> dplyr::filter(candidate)
  if (!is.null(dir)) tbl <- tbl |> dplyr::filter(direction == dir)
  g <- unique(tbl$geneName)
  g[!is.na(g) & g != ""]
}

## An HTML form that submits a gene list to Enrichr in a new tab. This is the
## documented Enrichr "add gene list from your website" pattern; it needs no
## external JS library, so it stays valid in self-contained HTML.
enrichrForm <- function(genes, description, label, maxGenes = 500L) {
  if (length(genes) == 0) {
    return(tags$p(tags$em(paste0("No candidate genes for ", label, "."))))
  }
  genes <- head(genes, maxGenes)
  tags$form(
    action = "https://maayanlab.cloud/Enrichr/enrich",
    method = "post", target = "_blank", enctype = "multipart/form-data",
    style = "display:inline-block; margin:0 10px 10px 0;",
    tags$input(type = "hidden", name = "list",
               value = paste(genes, collapse = "\n")),
    tags$input(type = "hidden", name = "description", value = description),
    tags$button(
      type = "submit",
      style = "padding:6px 12px; cursor:pointer; border-radius:6px; border:1px solid #888;",
      paste0(label, " (", length(genes), " genes) → Enrichr")
    )
  )
}

## ── Plot helpers (right-sized, single comparison) ─────────────────────────
compLabel <- paste(sampleGroup, "vs", refGroup)

plotVolcano <- function() {
  d <- peakTable |> dplyr::filter(!is.na(padj))
  d$class <- dplyr::case_when(
    d$candidate & d$direction == "up"   ~ "Up",
    d$candidate & d$direction == "down" ~ "Down",
    TRUE ~ "Not sig"
  )
  top <- d |> dplyr::filter(candidate)
  if (hasAnno) top <- top |> dplyr::filter(!is.na(geneName) & geneName != "")
  top <- top |> dplyr::group_by(direction) |>
    dplyr::slice_min(padj, n = 10, with_ties = FALSE) |> dplyr::ungroup()

  p <- ggplot(d, aes(log2FoldChange, -log10(padj), color = class)) +
    geom_point(size = 0.9, alpha = 0.5) +
    scale_color_manual(name = "",
      values = c("Up" = "#D55E00", "Down" = "#0072B2", "Not sig" = "grey75"),
      breaks = c("Up", "Not sig", "Down")) +
    geom_vline(xintercept = c(-lfcThreshold, lfcThreshold),
               linetype = "dashed", color = "grey40") +
    geom_hline(yintercept = -log10(fdrThreshold),
               linetype = "dashed", color = "grey40") +
    labs(x = expression(Log[2]~fold~change), y = expression(-Log[10]~adjusted~p),
         title = paste("Volcano:", compLabel)) +
    theme_bw(base_size = 11) + theme(plot.title = element_text(hjust = 0.5))
  if (hasAnno && nrow(top) > 0) {
    p <- p + geom_label_repel(data = top, aes(label = geneName),
      color = "black", fill = "white", size = 3, min.segment.length = 0,
      max.overlaps = Inf, show.legend = FALSE)
  }
  p
}

plotMA <- function() {
  d <- peakTable |> dplyr::filter(!is.na(padj)) |>
    dplyr::mutate(significant = dplyr::if_else(candidate, "Candidate", "Not sig"),
                  baseMeanLog = log10(baseMean + 1))
  ggplot(d, aes(baseMeanLog, log2FoldChange, color = significant)) +
    geom_point(size = 0.9, alpha = 0.5) +
    scale_color_manual(name = "",
      values = c("Candidate" = "#0072B2", "Not sig" = "grey65"),
      breaks = c("Candidate", "Not sig")) +
    geom_hline(yintercept = c(-lfcThreshold, lfcThreshold),
               linetype = "dashed", color = "grey40") +
    geom_hline(yintercept = 0, color = "grey30") +
    labs(x = expression(Log[10]~mean~normalized~count),
         y = expression(Log[2]~fold~change), title = paste("MA:", compLabel)) +
    theme_bw(base_size = 11) + theme(plot.title = element_text(hjust = 0.5))
}

plotPca <- function() {
  pcaData <- DESeq2::plotPCA(vsd, intgroup = "Condition", returnData = TRUE)
  pv <- round(100 * attr(pcaData, "percentVar"))
  ggplot(pcaData, aes(PC1, PC2, color = group, label = name)) +
    geom_point(size = 3) +
    ggrepel::geom_text_repel(size = 3, show.legend = FALSE) +
    scale_color_manual(values = groupColors, name = "Group") +
    labs(title = paste("PCA:", compLabel),
         x = paste0("PC1 (", pv[1], "%)"), y = paste0("PC2 (", pv[2], "%)")) +
    theme_bw(base_size = 11) + theme(plot.title = element_text(hjust = 0.5))
}

plotHeat <- function(nTop = 1000) {
  ## Rank by the same criterion used to call candidates: fold change when there
  ## are no replicates (p-values unreliable), adjusted p-value otherwise.
  ord <- if (noReplicates) {
    order(-abs(peakTable$log2FoldChange), na.last = NA)
  } else {
    order(peakTable$padj, na.last = NA)
  }
  ord <- head(ord, min(nTop, length(ord)))
  if (length(ord) < 2) return(NULL)
  ids <- peakTable$peakId[ord]
  mat <- assay(vsd)[ids, , drop = FALSE]
  ## Row z-scaling and row clustering are undefined for rows with zero or
  ## non-finite variance (common with few samples): they become NaN and make
  ## hclust() fail. Drop them so the heatmap is always well defined.
  rowSd <- apply(mat, 1, stats::sd)
  mat <- mat[is.finite(rowSd) & rowSd > 0, , drop = FALSE]
  if (nrow(mat) < 2) return(NULL)
  annotation_col <- data.frame(Condition = mlDds$Condition,
                               row.names = colnames(mat))
  ComplexHeatmap::pheatmap(mat, scale = "row", show_rownames = FALSE,
    annotation_col = annotation_col,
    annotation_colors = list(Condition = groupColors),
    color = colorRampPalette(rev(brewer.pal(7, "RdYlBu")))(100),
    clustering_method = "complete", angle_col = "45",
    name = "z-score",
    column_title = paste0("Top ", nrow(mat), " differential peaks"))
}

## Genomic feature and TSS-distance columns depend on the annotation tool.
featureCol <- intersect(c("Annotation", "annotation", "insideFeature"),
                        colnames(peakTable))
featureCol <- if (length(featureCol)) featureCol[1] else NA_character_
distCol    <- intersect(c("Distance to TSS", "distanceToTSS",
                          "shortestDistance", "distancetoFeature"),
                        colnames(peakTable))
distCol    <- if (length(distCol)) distCol[1] else NA_character_

plotFeatureDist <- function() {
  if (is.na(featureCol)) return(NULL)
  d <- peakTable |>
    dplyr::mutate(feature = str_replace(as.character(.data[[featureCol]]),
                                        " \\(.*$", ""),
                  set = dplyr::if_else(candidate, "Candidates", "All peaks")) |>
    dplyr::filter(!is.na(feature) & feature != "")
  if (nrow(d) == 0) return(NULL)
  d <- d |> dplyr::count(set, feature) |>
    dplyr::group_by(set) |> dplyr::mutate(frac = n / sum(n)) |> dplyr::ungroup()
  ggplot(d, aes(reorder(feature, frac), frac, fill = set)) +
    geom_col(position = "dodge") + coord_flip() +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_manual(values = c("All peaks" = "grey60",
                                 "Candidates" = "#D55E00"), name = "") +
    labs(x = "Genomic feature", y = "Fraction of peaks",
         title = "Peak genomic-feature distribution") +
    theme_bw(base_size = 11) + theme(plot.title = element_text(hjust = 0.5))
}

plotTssDist <- function() {
  if (is.na(distCol)) return(NULL)
  d <- peakTable |>
    dplyr::mutate(dist = suppressWarnings(as.numeric(.data[[distCol]])),
                  set = dplyr::if_else(candidate, "Candidates", "All peaks")) |>
    dplyr::filter(!is.na(dist) & abs(dist) <= 1e5)
  if (nrow(d) == 0) return(NULL)
  ggplot(d, aes(dist / 1e3, fill = set)) +
    geom_histogram(bins = 60, position = "identity", alpha = 0.5) +
    scale_fill_manual(values = c("All peaks" = "grey60",
                                 "Candidates" = "#D55E00"), name = "") +
    labs(x = "Distance to TSS (kb)", y = "Number of peaks",
         title = "Distance of peaks to nearest TSS") +
    theme_bw(base_size = 11) + theme(plot.title = element_text(hjust = 0.5))
}
```

:::::: {.panel-tabset}

# Overview

::: {.callout-note}
## Analysis summary

Differential accessibility/binding analysis of consensus peaks with **DESeq2**,
comparing **`r sampleGroup`** over **`r refGroup`**
(grouping column *`r param$grouping`*, annotation *`r param$annotationMethod`*).
Peak counts were variance-stabilized (VST). A peak is called a **candidate** at
|log2 fold change| ≥ `r lfcThreshold`
`r if (noReplicates) "" else paste0("and adjusted p-value < ", fdrThreshold)`.
Positive log2 fold change means higher signal in **`r sampleGroup`**.
:::

```{r no-rep-warning}
#| output: asis
if (noReplicates) {
  cat(
    ":::: {.callout-warning}\n",
    "## No biological replicates in this comparison\n\n",
    "At least one group has a single sample, so DESeq2 cannot estimate ",
    "within-group dispersion in the usual way. Dispersion was estimated with a ",
    "blind (`~1`) design that treats the two conditions as one group, which ",
    "makes the Wald test **conservative**: adjusted p-values are typically not ",
    "significant and should not be trusted. Candidates below are therefore ",
    "ranked by **fold change only** (|log2FC| ≥ ", lfcThreshold, "). ",
    "Add replicates for a statistically robust analysis.\n",
    "::::\n", sep = ""
  )
}
```

```{r summary-table}
summaryTab <- data.frame(
  Metric = c("Peaks tested", "Candidate peaks",
             paste0("Up in ", sampleGroup), paste0("Down in ", sampleGroup)),
  Count  = c(nTested, nCand, nUp, nDown)
)
DT::datatable(summaryTab, rownames = FALSE, options = list(dom = "t"),
              caption = "Peak-level summary")
```

# Pathway enrichment

Submit the candidate genes associated with differential peaks to
[Enrichr](https://maayanlab.cloud/Enrichr/) (opens in a new tab). Up to 500
genes per list are sent.

```{r enrichr-buttons}
if (!hasAnno) {
  htmltools::tags$p(htmltools::tags$em(
    "Peak annotation is unavailable, so gene-based enrichment links cannot be built."))
} else {
  htmltools::tagList(
    enrichrForm(candGenes("up"),  paste0(compLabel, " - up candidates"),   "Up candidates"),
    enrichrForm(candGenes("down"),paste0(compLabel, " - down candidates"), "Down candidates"),
    enrichrForm(candGenes(),      paste0(compLabel, " - all candidates"),  "All candidates")
  )
}
```

# Differential peaks

::: {.panel-tabset}

## Volcano

```{r volcano}
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Volcano plot. Dashed lines mark the fold-change and significance thresholds; the top 10 up- and down-regulated annotated peaks are labelled."
plotVolcano()
```

## MA

```{r ma}
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "MA plot of log2 fold change against mean normalized count."
plotMA()
```

## PCA

```{r pca}
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Principal component analysis of variance-stabilized peak counts."
res_pca <- tryCatch(plotPca(), error = function(e) NULL)
if (is.null(res_pca)) cat("PCA could not be computed for this comparison.") else res_pca
```

## Heatmap

```{r heatmap}
#| fig-width: 7
#| fig-height: 7
#| fig-cap: "Row-scaled variance-stabilized signal of the top differential peaks."
h <- tryCatch(plotHeat(), error = function(e) NULL)
if (is.null(h)) cat("Not enough peaks to draw a heatmap.") else h
```

:::

# Peak annotation

::: {.panel-tabset}

## Feature distribution

```{r feature-dist}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Distribution of peaks across genomic features, comparing all peaks with the differential candidates."
fd <- tryCatch(plotFeatureDist(), error = function(e) NULL)
if (is.null(fd)) cat("Genomic-feature annotation is not available for this run.") else fd
```

## Distance to TSS

```{r tss-dist}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Distribution of peak distance to the nearest transcription start site (within 100 kb)."
td <- tryCatch(plotTssDist(), error = function(e) NULL)
if (is.null(td)) cat("Distance-to-TSS annotation is not available for this run.") else td
```

:::

# Results table

The interactive table lists the top 1000 candidate peaks. The complete table
for all tested peaks is written to
[`differential_peaks.txt`](differential_peaks.txt).

```{r results-table}
displayCols <- intersect(
  c("peakId", "geneName", featureCol, distCol, "baseMean",
    "log2FoldChange", "lfcSE", "stat", "pvalue", "padj", "direction"),
  colnames(peakTable))
tableTib <- peakTable |>
  dplyr::filter(candidate) |>
  ## rank by fold change without replicates (p-values unreliable), else by FDR
  dplyr::arrange(if (noReplicates) dplyr::desc(abs(log2FoldChange)) else padj) |>
  dplyr::select(dplyr::all_of(displayCols))
if (hasAnno) {
  tableTib <- tableTib |> dplyr::distinct(geneName, .keep_all = TRUE)
}
tableTib <- head(tableTib, 1000)
ezInteractiveTableRmd(tableTib, digits = 3)
```

# Session Info

```{r session-info}
ezSessionInfo()
```

:::::
