## ALDEx2 {.tabset}

CLR-based compositional test. `r if (length(da_covariates)) "Covariate-adjusted GLM path (aldex.glm)." else "Classic t-test path (aldex)."`

### Species {.tabset}
```{r aldex_species, eval=has_species}
aldex_sp <- run_aldex2(physeq_Species, "Species")
```
```{r aldex_species_vdat, include=FALSE, eval=has_species}
aldex_sp_vd_adj <- mk_volcano_df(aldex_sp, "effect", "wi.eBH", "Taxonomy")
aldex_sp_vd_raw <- mk_volcano_df(aldex_sp, "effect", "wi.ep",  "Taxonomy")
```

#### Volcano — Adjusted p (Static)
```{r aldex_species_vstat_adj, eval=has_species, fig.width=11, fig.height=8}
show_volcano_static(aldex_sp_vd_adj, "ALDEx2 Species (effect vs BH-adj p)",
                    "ALDEx2 effect / coef",
                    ylab = "-log10(Wilcoxon BH-adjusted p)")
```

#### Volcano — Adjusted p (Interactive)
```{r aldex_species_vint_adj, eval=has_species}
show_volcano_interactive(aldex_sp_vd_adj, "ALDEx2 Species (effect vs BH-adj p)",
                         "ALDEx2 effect / coef",
                         ylab = "-log10(Wilcoxon BH-adjusted p)")
```

#### Volcano — Raw p (Static)
```{r aldex_species_vstat_raw, eval=has_species, fig.width=11, fig.height=8}
show_volcano_static(aldex_sp_vd_raw, "ALDEx2 Species (effect vs raw p)",
                    "ALDEx2 effect / coef",
                    ylab = "-log10(Wilcoxon raw p)")
```

#### Volcano — Raw p (Interactive)
```{r aldex_species_vint_raw, eval=has_species}
show_volcano_interactive(aldex_sp_vd_raw, "ALDEx2 Species (effect vs raw p)",
                         "ALDEx2 effect / coef",
                         ylab = "-log10(Wilcoxon raw p)")
```

#### Table
```{r aldex_species_tab, eval=has_species}
dt(aldex_sp, "ALDEx2 — Species")
```

### Genus {.tabset}
```{r aldex_genus}
aldex_ge <- run_aldex2(physeq_Genus, "Genus")
```
```{r aldex_genus_vdat, include=FALSE}
aldex_ge_vd_adj <- mk_volcano_df(aldex_ge, "effect", "wi.eBH", "Taxonomy")
aldex_ge_vd_raw <- mk_volcano_df(aldex_ge, "effect", "wi.ep",  "Taxonomy")
```

#### Volcano — Adjusted p (Static)
```{r aldex_genus_vstat_adj, fig.width=11, fig.height=8}
show_volcano_static(aldex_ge_vd_adj, "ALDEx2 Genus (effect vs BH-adj p)",
                    "ALDEx2 effect / coef",
                    ylab = "-log10(Wilcoxon BH-adjusted p)")
```

#### Volcano — Adjusted p (Interactive)
```{r aldex_genus_vint_adj}
show_volcano_interactive(aldex_ge_vd_adj, "ALDEx2 Genus (effect vs BH-adj p)",
                         "ALDEx2 effect / coef",
                         ylab = "-log10(Wilcoxon BH-adjusted p)")
```

#### Volcano — Raw p (Static)
```{r aldex_genus_vstat_raw, fig.width=11, fig.height=8}
show_volcano_static(aldex_ge_vd_raw, "ALDEx2 Genus (effect vs raw p)",
                    "ALDEx2 effect / coef",
                    ylab = "-log10(Wilcoxon raw p)")
```

#### Volcano — Raw p (Interactive)
```{r aldex_genus_vint_raw}
show_volcano_interactive(aldex_ge_vd_raw, "ALDEx2 Genus (effect vs raw p)",
                         "ALDEx2 effect / coef",
                         ylab = "-log10(Wilcoxon raw p)")
```

#### Table
```{r aldex_genus_tab}
dt(aldex_ge, "ALDEx2 — Genus")
```
