## MaAsLin 3 {.tabset}

GLMs testing **abundance** and **prevalence** associations. Formula:
`~ Group [+ covariates] + ReadDepth`, TSS / LOG transform, max_significance =
`r maaslin_max_sig`. MaAsLin3 writes its own `summary_plot.pdf` and per-
association PNGs into `r results_dir`/maaslin3_*/.

### Species {.tabset}
```{r maaslin_species, eval=has_species, results='hide'}
ml_sp <- run_maaslin3(physeq_Species, "Species")
```

#### Significant associations
```{r maaslin_species_tab, eval=has_species}
if (!is.null(ml_sp) && !is.null(ml_sp$res)) {
  sig <- ml_sp$res %>% filter(.data$qval_individual <= maaslin_max_sig)
  dt(if (nrow(sig)) sig else ml_sp$res,
     paste0("MaAsLin3 Species — ",
            if (nrow(sig)) paste0("significant (q<=", maaslin_max_sig, ")")
            else "all (none significant)"))
} else cat("MaAsLin3 Species produced no results.\n")
```

#### Coefficient plot
```{r maaslin_species_plot, eval=has_species, fig.width=12, fig.height=8}
if (!is.null(ml_sp) && !is.null(ml_sp$res)) {
  d <- ml_sp$res %>%
       filter(.data$qval_individual <= maaslin_max_sig,
              .data$name != "ReadDepth") %>%
       arrange(coef)
  if (nrow(d) > 0) {
    d$lab <- factor(paste(d$Taxonomy, d$model, sep = " | "),
                    levels = paste(d$Taxonomy, d$model, sep = " | "))
    print(ggplot(d, aes(coef, lab, color = model)) +
            geom_point(size = 3) + geom_vline(xintercept = 0, linetype = 2) +
            geom_errorbarh(aes(xmin = coef - stderr, xmax = coef + stderr), height = .2) +
            labs(x = "Coefficient", y = NULL,
                 title = "MaAsLin3 Species — significant") + theme_bw())
  } else cat("No significant MaAsLin3 Species associations.\n")
}
```
```{r maaslin_species_vdat, include=FALSE, eval=has_species}
ml_sp_vd_adj <- NULL; ml_sp_vd_raw <- NULL
if (!is.null(ml_sp) && !is.null(ml_sp$res)) {
  .r <- ml_sp$res[!is.na(ml_sp$res$metadata) & ml_sp$res$metadata == da_group, , drop = FALSE]
  .r <- .r[order(.r$feature, .r$model != "abundance"), , drop = FALSE]
  .r <- .r[!duplicated(.r$feature), , drop = FALSE]
  .pcol <- if ("pval_joint" %in% colnames(.r)) "pval_joint" else "pval_individual"
  .qcol <- if ("qval_joint" %in% colnames(.r)) "qval_joint" else "qval_individual"
  ml_sp_vd_adj <- mk_volcano_df(.r, "coef", .qcol, "Taxonomy")
  ml_sp_vd_raw <- mk_volcano_df(.r, "coef", .pcol, "Taxonomy")
}
```

#### Volcano — Adjusted p (Static)
```{r maaslin_species_vstat_adj, eval=has_species, fig.width=11, fig.height=8}
show_volcano_static(ml_sp_vd_adj, "MaAsLin3 Species (coef vs joint q)",
                    "MaAsLin3 coefficient",
                    ylab = "-log10(joint BH-adjusted q)", fc = 0)
```

#### Volcano — Adjusted p (Interactive)
```{r maaslin_species_vint_adj, eval=has_species}
show_volcano_interactive(ml_sp_vd_adj, "MaAsLin3 Species (coef vs joint q)",
                         "MaAsLin3 coefficient",
                         ylab = "-log10(joint BH-adjusted q)", fc = 0)
```

#### Volcano — Raw p (Static)
```{r maaslin_species_vstat_raw, eval=has_species, fig.width=11, fig.height=8}
show_volcano_static(ml_sp_vd_raw, "MaAsLin3 Species (coef vs joint raw p)",
                    "MaAsLin3 coefficient",
                    ylab = "-log10(joint raw p)", fc = 0)
```

#### Volcano — Raw p (Interactive)
```{r maaslin_species_vint_raw, eval=has_species}
show_volcano_interactive(ml_sp_vd_raw, "MaAsLin3 Species (coef vs joint raw p)",
                         "MaAsLin3 coefficient",
                         ylab = "-log10(joint raw p)", fc = 0)
```

### Genus {.tabset}
```{r maaslin_genus, results='hide'}
ml_ge <- run_maaslin3(physeq_Genus, "Genus")
```

#### Significant associations
```{r maaslin_genus_tab}
if (!is.null(ml_ge) && !is.null(ml_ge$res)) {
  sig <- ml_ge$res %>% filter(.data$qval_individual <= maaslin_max_sig)
  dt(if (nrow(sig)) sig else ml_ge$res,
     paste0("MaAsLin3 Genus — ",
            if (nrow(sig)) paste0("significant (q<=", maaslin_max_sig, ")")
            else "all (none significant)"))
} else cat("MaAsLin3 Genus produced no results.\n")
```

#### Coefficient plot
```{r maaslin_genus_plot, fig.width=12, fig.height=8}
if (!is.null(ml_ge) && !is.null(ml_ge$res)) {
  d <- ml_ge$res %>%
       filter(.data$qval_individual <= maaslin_max_sig,
              .data$name != "ReadDepth") %>%
       arrange(coef)
  if (nrow(d) > 0) {
    d$lab <- factor(paste(d$Taxonomy, d$model, sep = " | "),
                    levels = paste(d$Taxonomy, d$model, sep = " | "))
    print(ggplot(d, aes(coef, lab, color = model)) +
            geom_point(size = 3) + geom_vline(xintercept = 0, linetype = 2) +
            geom_errorbarh(aes(xmin = coef - stderr, xmax = coef + stderr), height = .2) +
            labs(x = "Coefficient", y = NULL,
                 title = "MaAsLin3 Genus — significant") + theme_bw())
  } else cat("No significant MaAsLin3 Genus associations.\n")
}
```
```{r maaslin_genus_vdat, include=FALSE}
ml_ge_vd_adj <- NULL; ml_ge_vd_raw <- NULL
if (!is.null(ml_ge) && !is.null(ml_ge$res)) {
  .r <- ml_ge$res[!is.na(ml_ge$res$metadata) & ml_ge$res$metadata == da_group, , drop = FALSE]
  .r <- .r[order(.r$feature, .r$model != "abundance"), , drop = FALSE]
  .r <- .r[!duplicated(.r$feature), , drop = FALSE]
  .pcol <- if ("pval_joint" %in% colnames(.r)) "pval_joint" else "pval_individual"
  .qcol <- if ("qval_joint" %in% colnames(.r)) "qval_joint" else "qval_individual"
  ml_ge_vd_adj <- mk_volcano_df(.r, "coef", .qcol, "Taxonomy")
  ml_ge_vd_raw <- mk_volcano_df(.r, "coef", .pcol, "Taxonomy")
}
```

#### Volcano — Adjusted p (Static)
```{r maaslin_genus_vstat_adj, fig.width=11, fig.height=8}
show_volcano_static(ml_ge_vd_adj, "MaAsLin3 Genus (coef vs joint q)",
                    "MaAsLin3 coefficient",
                    ylab = "-log10(joint BH-adjusted q)", fc = 0)
```

#### Volcano — Adjusted p (Interactive)
```{r maaslin_genus_vint_adj}
show_volcano_interactive(ml_ge_vd_adj, "MaAsLin3 Genus (coef vs joint q)",
                         "MaAsLin3 coefficient",
                         ylab = "-log10(joint BH-adjusted q)", fc = 0)
```

#### Volcano — Raw p (Static)
```{r maaslin_genus_vstat_raw, fig.width=11, fig.height=8}
show_volcano_static(ml_ge_vd_raw, "MaAsLin3 Genus (coef vs joint raw p)",
                    "MaAsLin3 coefficient",
                    ylab = "-log10(joint raw p)", fc = 0)
```

#### Volcano — Raw p (Interactive)
```{r maaslin_genus_vint_raw}
show_volcano_interactive(ml_ge_vd_raw, "MaAsLin3 Genus (coef vs joint raw p)",
                         "MaAsLin3 coefficient",
                         ylab = "-log10(joint raw p)", fc = 0)
```
