---
title: "`r if (exists('reportTitle')) reportTitle else 'SUSHI Report'`"
output: 
  html_document:
    mathjax: https://fgcz-gstore.uzh.ch/reference/mathjax.js
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package="ezRun")
    css: !expr system.file("templates/fgcz.css", package="ezRun")
editor_options: 
  chunk_output_type: inline
---


# {.tabset}
`r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

## Differential Peak Analysis {.tabset}

DESeq2 differential peak analysis performed across all possible pairwise comparisons. Peak counts were normalized using a variance-stabilizing transformation (VST) to mitigate the dependency of the variance on the mean expression level.

For each two-group comparison we show:

* a principal component analysis (PCA) to asses whether samples segregate according to biological conditions.
* a volcano plot with the top 10 up- and down-regulated genes associated with peaks. Differentially expressed genes (DEGs) are called at an FDR (false discovery rate) of 5% (Benjamini-Hochberg adjusted p-value < 0.05) and effect size (absolute log2 fold change cutoff) of 1.
* a MA plot that highlights statistically significant peak, called at an FDR of 5% (Benjamini-Hochberg adjusted p-value < 0.05).
* a heatmap of the top 1000 differential peaks.
* an interactive table listing the top 1000 differential peaks.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, knitr.table.format = "html")
## generate all graphs as PDF
## knitr::opts_chunk$set(dev = "cairo_pdf")
```

```{r packages, message=FALSE}
library(ezRun)
library(ComplexHeatmap)
library(DESeq2)
library(RColorBrewer)
library(ggrepel)
library(ggplot2)
library(dplyr)
library(patchwork)
library(kableExtra)
```

```{r load_data}
dds <- qs2::qs_read('dds.qs2')
peakAnno <- qs2::qs_read('peakAnno.qs2')
```

```{r get_results}
getComparisonMatrix <- function(dds){
  conditions <- dds$Condition |> levels()

  compMat <- combn(conditions, 2)
  colnames(compMat) <- paste0("pair", 1:ncol(compMat))
  rownames(compMat) <- c("group1", "group2")

  return(compMat)
}

getResults <- function(dds, peakAnno, groupPair){
  ml_dds <- DESeq(dds)
  
  ## alph. group: colData(ml_subset)$Condition |> levels()
  contrast <- c('group', unname(groupPair))
  ml_raw <- results(ml_dds, contrast = contrast)
  
  ml_res <-ml_raw |>
    as.data.frame() |>
    rownames_to_column("peakId") |>
    as_tibble() |>
    left_join(peakAnno, by = "peakId")
  
  compname <- paste0(contrast[2], "--vs--" ,contrast[3])
  
  res <- list()
  res[[compname]][['dds']] <- ml_dds
  res[[compname]][['res']] <- ml_res
  return(res)
}
```

```{r utils}
groupFormatter <- function(groupName){
  str_replace(groupName, '--vs--', ' vs ')
}
```

```{r peak_table}
make_peak_table <- function(res, groupName){
  res[[groupName]]$res |>
    mutate(
      # pvalue = pmax(padj, 1e-50),
      # neg_log10_padj = -log10(padj),
      direction = if_else(log2FoldChange > 0, "up", "down", NA_character_),
      candidate = abs(log2FoldChange) >= 1 & padj < 0.05) |> 
    filter(!is.na(log2FoldChange) & !is.na(padj))
}
```

```{r plot_functions}
plot_volcano <- function(res, groupName){
  ml_res <- make_peak_table(res, groupName)
  
  top_genes <- ml_res |> 
    filter(candidate) |> 
    group_by(direction) |> 
    slice_min(padj, n = 10) |> 
    ungroup() |>
    filter(!is.na(`geneName`))
  
  ggplot(ml_res, aes(x = log2FoldChange, y = -log10(padj))) +
    ## no legend ...
    #geom_point(data = filter(ml_res, !candidate), color = "lightgrey", size = 0.8) +
    #geom_point(data = filter(ml_res, candidate & direction == "up"), color = "#FF7777", alpha = 0.5, size = 1) +
    #geom_point(data = filter(ml_res, candidate & direction == "down"), color = "#7DA8E6", alpha = 0.5, size = 1) +
    geom_point(aes(color = case_when(
      candidate & direction == "up" ~ "Up",
      candidate & direction == "down" ~ "Down",
      !candidate ~ "Not Sig",
      TRUE ~ "Not significant")), 
     size = 1, #ifelse(ml_res$candidate, 1, 0.1),
     alpha = 0.5) + #ifelse(ml_res$candidate, 1, 0)) +
    scale_color_manual(
      name = "",
      values = c("Not Sig" = "lightgrey", "Up" = "#FF7777", "Down" = "#7DA8E6"),
      breaks = c("Up", "Not Sig", "Down") ## controls order legend items
    ) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
    geom_label_repel(data = top_genes, aes(label = `geneName`), color = "black", fill = "white", min.segment.length = 0, max.overlaps = Inf) +
                     # label.padding = unit(0.2, "lines"), label.r = unit(0.1, "lines"), box.padding = 0.3, point.padding = 0.3 
    labs(x = expression(Log[2]~fold~change), y = expression(-Log[10]~Pvalue), title = paste('DEG -', groupFormatter(groupName))) +
    theme_bw() + theme(plot.title = element_text(hjust = 0.5))
}

plot_ma <- function(res, groupName){
  ## deseq way -- no ggplot+patchwork compatible
  # groupPair <- groupName |> str_split('_vs_') |> pluck(1)
  # ma_raw <- results(res[[groupName]]$dds, contrast=c("Condition", groupPair))
  # DESeq2::plotMA(ma_raw)
  
  ml_res <- res[[groupName]]$res |>
    as_tibble() |>
    mutate(
      significant = ifelse(!is.na(padj) & padj < 0.05, "Significant", "Not Significant"),
      baseMeanLog = log10(baseMean+1) ## if_else(baseMean == 0, log10(1), log10(baseMean)))
    ) |> 
    filter(!is.na(log2FoldChange) & !is.na(padj))
  
  ggplot(ml_res, aes(x = baseMeanLog, y = log2FoldChange, color = significant)) +
    geom_point(alpha = 0.5) +
    scale_color_manual(name = "", 
      values = c("Significant" = "#0072B2", "Not Significant" = "gray55"),
      breaks = c("Significant", "Not Significant")) +
    geom_smooth(inherit.aes = FALSE, aes(x = baseMeanLog, y = log2FoldChange), method = "loess", se = FALSE, colour = "purple") +
    geom_hline(yintercept = 0, colour = 'gray30') +
    labs(x = expression(Log[10]~Average~Normalized~Count), y = expression(Log[2]~Fold~Change), 
      color = "Significant", title = paste('MA Plot:', groupFormatter(groupName))) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5))
}

plot_pca <- function(res, groupName){
  ml_dds <- res[[groupName]]$dds
  vsd <- vst(ml_dds, blind=FALSE)
  pca_data <- DESeq2::plotPCA(vsd, intgroup="Condition", returnData = TRUE)
  pc_val <- round(100 * attr(pca_data, "percentVar"))
  
  ggplot(pca_data, aes(x = PC1, y = PC2, color = group)) +
    geom_point(size = 3) + 
    labs(title = paste("PCA plot:", groupFormatter(groupName)),
      x = paste0("PC1 (", pc_val[1], "% variance)"),
      y = paste0("PC2 (", pc_val[2], "% variance)"),
      color = "Group") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5))
}

plot_heatmap <- function(res, groupName){
  ml_res <- res[[groupName]]$res
  ml_dds <- res[[groupName]]$dds
  
  vsd <- vst(ml_dds, blind=FALSE)
  top_peaks <- order(ml_res$padj, na.last=TRUE)[1:1000]
  mat <-assay(vsd)[ml_res[top_peaks,][['peakId']],]
  rownames(mat) <- rownames(ml_res)[top_peaks]
  
  # Create annotation data frame
  annotation_col <- data.frame(
    Condition = ml_dds$Condition,
    row.names = colnames(mat)
  )
  
  condition_colors <- c("#009E73", "#CC79A7")
  names(condition_colors) <- levels(annotation_col$Condition)
  annotation_colors <- list(Condition = condition_colors)
  
  # Create color palette
  heat_colors <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(100)
  # heat_colors <- colorRampPalette(rev(brewer.pal(n = 9, name = "Purples")))(255)
  
  # Plot heatmap
  gg <- ComplexHeatmap::pheatmap(mat,
    scale = "row",
    show_rownames = FALSE,
    annotation_col = annotation_col,
    annotation_colors = annotation_colors,
    color = heat_colors,
    clustering_distance_rows = "euclidean",
    clustering_distance_cols = "euclidean",
    clustering_method = "complete",
    angle_col = "45"
  ) %>% ggplotify::as.ggplot()
  
  gg + 
    labs(title = paste("Top 1000 Differential Peaks -", groupFormatter(groupName))) +
    theme(plot.title = element_text(hjust = 0.5))
  
  ## complexheatmap way -- no compatiple with patchwork
  # draw(gg,
  #   column_title = paste("Top 1000 Differential Peaks -", groupFormatter(groupName)),
  #   column_title_gp = gpar(fontsize = 16))
}
```

```{r prepare_results, include="false"}
compMat <- getComparisonMatrix(dds)
curres <- getResults(dds=dds, peakAnno, groupPair=compMat[,1])
groupName <- names(curres)
pander::pandoc.p('')
pander::pandoc.header(groupFormatter(groupName), level = 3)
pander::pandoc.p('')
```

### PCA plot

```{r display_pca, results='asis', fig.width=15, fig.height=10}
p1 <- plot_pca(curres, groupName)
print(p1)
```

### DEG

```{r display_deg, results='asis', fig.width=15, fig.height=10}
p2 <- plot_volcano(curres, groupName)
print(p2)
```

### MA Plot

```{r display_ma, results='asis', fig.width=15, fig.height=10}
p3 <- plot_ma(curres, groupName)
print(p3)
```

### Heatmap

```{r display_heatmap, results='asis', fig.width=15, fig.height=10}
p4 <- plot_heatmap(curres, groupName)
print(p4)
```

### Result table

```{r display_table, results='asis', fig.width=15, fig.height=10}
peak_tib <- make_peak_table(curres, groupName) |>
  filter(candidate) |>
  arrange(padj) |>
  distinct(`geneName`, .keep_all = TRUE) |>
  head(1000)
print(htmltools::tagList(ezInteractiveTableRmd(peak_tib, digits = 3)))
# |> DT::formatSignif(c("baseMean", "log2FoldChange", "lfcSE", "stat", "pvalue", "padj"), digits=3)
## NBB: this dummy table triggers/forces rendering, otherwise no table will be displayed ¯\_(ツ)_/¯
ezInteractiveTableRmd(data.frame())
```
