---
title: "`r if (exists('reportTitle')) reportTitle else 'Shotgun Functional DA (MaAsLin3)'`"
output:
  html_document:
    highlight: pygments
    theme: sand
    code_folding: hide
    toc: yes
    toc_depth: 4
    df_print: paged
    self_contained: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE,
                      fig.width = 12, fig.height = 8, cache = FALSE)
knitr::opts_knit$set(root.dir = ".")

library(ggplot2)
library(dplyr)
library(tibble)
library(stringr)
library(DT)
library(kableExtra)
library(maaslin3)
```

```{r register_widget_deps, include=FALSE}
## Force-register DT JS dependencies at top level. DT tables produced deep
## inside a results='asis' loop don't propagate their deps back through
## knit_meta, leaving the widgets as blank containers.
local({
  for (d in htmltools::findDependencies(DT::datatable(data.frame(x = 1),
                                                      extensions = "Buttons")))
    knitr::knit_meta_add(list(d))
})
```

```{r load_params, include=FALSE}
library(ezRun)
param  <- ezLoadRobj("param")
output <- ezLoadRobj("output")

`%||%` <- function(a, b) {
  if (is.null(a)) return(b)
  if (is.character(a) && length(a) == 1 && nchar(a) == 0) return(b)
  a
}
.split_csv <- function(s) {
  if (is.null(s)) return(character(0))
  if (length(s) > 1) {
    out <- trimws(as.character(s)); return(out[nzchar(out)])
  }
  if (is.character(s) && nchar(s) == 0) return(character(0))
  out <- trimws(unlist(strsplit(as.character(s), "[,;]")))
  out[nzchar(out)]
}
.as_num <- function(x, default) {
  v <- suppressWarnings(as.numeric(x))
  if (length(v) == 0 || is.na(v)) default else v
}

## --- params --------------------------------------------------------
funcTables   <- param$functionalTables       # named char: tag -> joined TSV
grouping     <- as.character(param$grouping %||% "")
sampleGroup  <- as.character(param$sampleGroup %||% "")
refGroup     <- as.character(param$refGroup %||% "")
covariates   <- .split_csv(param$daCovariates %||% "")
samplesDrop  <- .split_csv(param$samplesToDrop %||% "")
prevMin      <- .as_num(param$prevalenceMin   %||% 0.10, 0.10)
relabMin     <- .as_num(param$relabundMin     %||% 0.10, 0.10)
maaslinMaxSig <- .as_num(param$maaslinMaxSig  %||% 0.10, 0.10)
pValueThresh  <- .as_num(param$pValueThresh   %||% 0.05, 0.05)

if (!nzchar(grouping) || !nzchar(sampleGroup) || !nzchar(refGroup))
  stop("grouping, sampleGroup and refGroup are required.")
if (sampleGroup == refGroup)
  stop("sampleGroup must differ from refGroup.")
```

```{r load_metadata, include=FALSE}
metadataFile <- "sample-metadata.tsv"
if (!file.exists(metadataFile))
  stop("sample-metadata.tsv missing - ezMethodDiffShot should have produced it.")
md <- read.table(metadataFile, header = TRUE, sep = "\t",
                 stringsAsFactors = FALSE, check.names = FALSE,
                 quote = "", na.strings = c("", "NA"))
rownames(md) <- md$SampleID
md$SampleID  <- NULL
if (length(samplesDrop) > 0) md <- md[!rownames(md) %in% samplesDrop, , drop = FALSE]
if (!(grouping %in% colnames(md)))
  stop(sprintf("grouping column '%s' not found in sample metadata. ",
               "Available [Factor] columns: %s",
               grouping, paste(colnames(md), collapse = ", ")))
keepGroup <- md[[grouping]] %in% c(sampleGroup, refGroup)
md <- md[keepGroup, , drop = FALSE]
md[[grouping]] <- factor(md[[grouping]], levels = c(refGroup, sampleGroup))

## Rendered as a markdown snippet by the Input section below. Kept out of
## the inline `r ...` expression on purpose: backticks inside inline R
## terminate the expression early and break knitr's parser.
funcTablesLabel <- paste(sprintf("`%s`", names(funcTables)), collapse = ", ")
covariatesLabel <- if (length(covariates)) paste(covariates, collapse = ", ") else "(none beyond ReadDepth)"
```

# Input

- Source type: **HUMAnN functional tables**
- Grouping: **`r grouping`** (contrast: `r sampleGroup` vs `r refGroup`)
- Covariates: `r covariatesLabel`
- Samples after filters: `r nrow(md)`
- Feature tables available: `r funcTablesLabel`

# Per-table MaAsLin3 {.tabset}

```{r helpers, include=FALSE}
## Persisted CSV outputs live here so users can download them
## alongside the report (mirrors DiffShot.Rmd/results_out).
results_dir <- "results_out"
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)

## Standard interactive table widget with copy/CSV/Excel download
## buttons, column filters, and significant-digit formatting on
## non-integer numeric columns. Same shape as DiffShot.Rmd::dt().
dt <- function(df, caption, sig = 3, length_menu = TRUE) {
  dom <- if (length_menu) "lftBip" else "ftBip"
  w <- DT::datatable(df, caption = caption, extensions = "Buttons",
                     rownames = FALSE, filter = "top",
                     options = list(scrollX = TRUE, dom = dom,
                                    buttons = c("copy", "csv", "excel"),
                                    pageLength = 10,
                                    lengthMenu = c(10, 20, 100, 200)))
  num_cols <- names(df)[vapply(df, function(x) {
    if (!is.numeric(x)) return(FALSE)
    xf <- x[is.finite(x)]
    length(xf) > 0 && !isTRUE(all(xf == round(xf)))
  }, logical(1))]
  if (length(num_cols)) w <- DT::formatSignif(w, columns = num_cols, digits = sig)
  w
}

save_csv <- function(df, fname) {
  fp <- file.path(results_dir, fname)
  write.csv(df, fp, row.names = FALSE)
  fp
}

## Read a humann_join_tables output, recover bare sample IDs.
load_joined <- function(path) {
  m <- read.table(path, header = TRUE, sep = "\t", check.names = FALSE,
                  comment.char = "", quote = "", row.names = 1)
  ## humann_join_tables column names look like "<sample>_concat_<N>_<tag>"
  ## or "<sample>_concat_Abundance" - strip from "_concat" onward.
  colnames(m) <- sub("_concat.*$", "", colnames(m))
  ## Belt-and-suspenders: drop common HUMAnN suffixes too
  colnames(m) <- sub("_(Abundance|Abundance-CPM|Coverage|Abundance-RPKs).*$",
                    "", colnames(m))
  ## Drop HUMAnN special rows and SGB-stratified rows.
  drop <- c("UNMAPPED", "UNINTEGRATED", "UNGROUPED", "READS_UNMAPPED")
  m <- m[!rownames(m) %in% drop, , drop = FALSE]
  m <- m[!grepl("\\|", rownames(m)), , drop = FALSE]
  as.matrix(m)
}

## Prevalence + pooled-abundance filter.
filter_features <- function(mat, prevMin, relabMin) {
  if (nrow(mat) == 0 || ncol(mat) == 0) return(mat)
  prev  <- rowSums(mat > 0) / ncol(mat)
  total <- sum(mat)
  if (total == 0) return(mat[FALSE, , drop = FALSE])
  ra    <- rowSums(mat) / total * 100
  keep  <- (prev >= prevMin) & (ra >= relabMin)
  mat[keep, , drop = FALSE]
}

## One-shot MaAsLin3 + return tidy result.
run_one <- function(mat, md, tableTag) {
  shared <- intersect(colnames(mat), rownames(md))
  if (length(shared) < 3) return(list(err = sprintf(
    "Only %d samples shared between feature table and metadata; skip.",
    length(shared))))
  mat <- mat[, shared, drop = FALSE]
  md_sub <- md[shared, , drop = FALSE]
  md_sub$ReadDepth <- as.numeric(colSums(mat))

  fixedEffects <- c(grouping, intersect(covariates, colnames(md_sub)), "ReadDepth")
  reference    <- sprintf("%s,%s", grouping, refGroup)

  outDir <- sprintf("maaslin3_%s", tableTag)
  unlink(outDir, recursive = TRUE, force = TRUE)
  dir.create(outDir, showWarnings = FALSE)

  ## MaAsLin3 emits a chatty INFO::/WARN:: log stream via cat() (one
  ## line per feature fitted). In a results='asis' chunk those lines
  ## leak into the report body and pandoc turns "INFO::Fitting" into a
  ## bogus <a href="INFO::..."> link. Sink stdout for the duration of
  ## the call to keep the report layout clean.
  sink(tempfile())
  res <- tryCatch(
    suppressMessages(suppressWarnings(
      maaslin3::maaslin3(
        input_data        = t(mat),
        input_metadata    = md_sub,
        output            = outDir,
        fixed_effects     = fixedEffects,
        reference         = reference,
        normalization     = "NONE",      # HUMAnN already CPM
        transform         = "LOG",
        standardize       = TRUE,
        max_significance  = maaslinMaxSig,
        plot_summary_plot = TRUE,
        plot_associations = TRUE
      )
    )),
    error   = function(e) list(error = conditionMessage(e)),
    finally = sink()
  )
  if (!is.null(res$error)) return(list(err = res$error))

  ## MaAsLin3 writes all_results.tsv into outDir.
  resPath <- file.path(outDir, "all_results.tsv")
  if (!file.exists(resPath)) return(list(err = "all_results.tsv missing"))
  resDf <- read.table(resPath, header = TRUE, sep = "\t", quote = "\"",
                      comment.char = "", check.names = FALSE,
                      stringsAsFactors = FALSE)
  list(res = resDf, outDir = outDir, nFeatures = nrow(mat), nSamples = ncol(mat))
}

volcano_plot <- function(resDf, title) {
  if (is.null(resDf) || nrow(resDf) == 0) return(NULL)
  sub <- resDf[resDf$metadata == grouping &
               is.finite(resDf$coef) &
               !is.na(resDf$qval_individual), , drop = FALSE]
  if (nrow(sub) == 0) return(NULL)
  sub$nlog10q <- -log10(pmax(sub$qval_individual, 1e-300))
  sub$signif  <- sub$qval_individual < maaslinMaxSig
  top <- head(sub[order(sub$qval_individual), ], 10)
  ggplot(sub, aes(x = coef, y = nlog10q, color = signif)) +
    geom_point(alpha = 0.6, size = 1.8) +
    geom_text(data = top, aes(label = feature),
              size = 2.6, vjust = -0.6, check_overlap = TRUE) +
    geom_hline(yintercept = -log10(maaslinMaxSig),
               linetype = "dashed", color = "grey50") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey80") +
    scale_color_manual(values = c(`FALSE` = "grey60", `TRUE` = "firebrick"),
                       guide = "none") +
    labs(title = title, x = "MaAsLin3 coefficient",
         y = expression(-log[10](q[ind]))) +
    theme_minimal(base_size = 12)
}

coef_plot <- function(resDf, title) {
  if (is.null(resDf) || nrow(resDf) == 0) return(NULL)
  d <- resDf[resDf$metadata == grouping &
             !is.na(resDf$qval_individual) &
             resDf$qval_individual <= maaslinMaxSig, , drop = FALSE]
  if (nrow(d) == 0) return(NULL)
  d <- d[order(d$coef), , drop = FALSE]
  d$feature <- factor(d$feature, levels = unique(d$feature))
  modelCol  <- if ("model" %in% colnames(d)) d$model else "fit"
  ggplot(d, aes(x = coef, y = feature, color = modelCol)) +
    geom_point(size = 3) +
    geom_vline(xintercept = 0, linetype = 2) +
    geom_errorbarh(aes(xmin = coef - stderr, xmax = coef + stderr), height = 0.2) +
    labs(x = "MaAsLin3 coefficient", y = NULL,
         title = title, color = "model") +
    theme_bw(base_size = 11)
}

label_for <- function(tag) {
  switch(tag,
         pathabundance     = "MetaCyc pathways",
         reactions         = "MetaCyc reactions",
         genefamilies_ko   = "KEGG KO",
         genefamilies      = "UniRef90-EC gene families",
         tag)
}
```

```{r per_table_loop, results='asis'}
for (tag in names(funcTables)) {
  cat(sprintf("\n## %s\n\n", label_for(tag)))
  mat_raw <- load_joined(funcTables[[tag]])
  mat <- filter_features(mat_raw, prevMin, relabMin)
  cat(sprintf("- Features (raw / after filter): **%d / %d**\n",
              nrow(mat_raw), nrow(mat)))
  cat(sprintf("- Samples (in joined table): %d\n\n", ncol(mat_raw)))
  if (nrow(mat) == 0 || ncol(mat) < 3) {
    cat("Too few features/samples after filtering — skipping MaAsLin3.\n\n")
    next
  }
  r <- run_one(mat, md, tag)
  if (!is.null(r$err)) {
    cat(sprintf("**MaAsLin3 error:** %s\n\n", r$err)); next
  }

  cat(sprintf("- MaAsLin3 input: %d features × %d samples\n", r$nFeatures, r$nSamples))
  cat(sprintf("- Native output dir: `%s/`\n\n", r$outDir))

  resDf <- r$res
  sig <- resDf[!is.na(resDf$qval_individual) &
               resDf$qval_individual <= maaslinMaxSig &
               resDf$metadata == grouping, , drop = FALSE]
  cat(sprintf("- Significant associations (q ≤ %g): **%d**\n\n",
              maaslinMaxSig, nrow(sig)))

  ## Persist per-tag CSVs (all results + significant subset) so the
  ## tables are downloadable from the report folder without opening the
  ## report first.
  allCsv <- save_csv(resDf, sprintf("maaslin3_%s_all_results.csv", tag))
  sigCsv <- save_csv(sig,   sprintf("maaslin3_%s_significant.csv", tag))
  cat(sprintf("- Downloads: [all results](%s), [significant only](%s)\n\n",
              allCsv, sigCsv))

  ## Interactive table with copy/CSV/Excel buttons and column filters.
  showTbl <- if (nrow(sig) > 0) sig else
             head(resDf[order(resDf$qval_individual), ], 30)
  showCaption <- if (nrow(sig) > 0)
    sprintf("MaAsLin3 significant associations - %s (q <= %g)",
            label_for(tag), maaslinMaxSig)
  else
    sprintf("MaAsLin3 top 30 by q-value - %s (no association passed q <= %g)",
            label_for(tag), maaslinMaxSig)
  print(htmltools::tagList(dt(showTbl, caption = showCaption)))
  cat("\n\n")

  ## Coefficient plot
  cp <- coef_plot(resDf, sprintf("%s — significant associations", label_for(tag)))
  if (!is.null(cp)) print(cp)
  cat("\n\n")

  ## Volcano plot
  vp <- volcano_plot(resDf, sprintf("%s — volcano (coef vs q)", label_for(tag)))
  if (!is.null(vp)) print(vp)
  cat("\n\n")
}
```

# Method notes

- Input: HUMAnN 4 alpha per-sample outputs joined via `humann_join_tables` (substring match on `genefamilies` / `reactions` / `pathabundance`). Stratified rows (`|s__`, `|t__SGB`) and HUMAnN special rows (`UNMAPPED` / `UNINTEGRATED` / `UNGROUPED` / `READS_UNMAPPED`) are dropped.
- Filter: prevalence ≥ `r prevMin` AND pooled relative abundance ≥ `r relabMin`%.
- Model: `~ grouping + covariates + ReadDepth` (ReadDepth = column sum, always added).
- MaAsLin3 options: `normalization = "NONE"` (HUMAnN tables are already CPM-normalised), `transform = "LOG"`, `standardize = TRUE`, `max_significance = `r maaslinMaxSig``.
- One MaAsLin3 run per available functional table; results dirs sit at `maaslin3_<tag>/`.
