## LEfSe {.tabset}

> :information_source: **LEfSe runs unadjusted by design** — non-parametric
> Kruskal–Wallis + LDA on relative abundance with no covariate term. This SUSHI
> app therefore does not expose a covariate selector. If you need adjustment
> for batch / age / sex, run DiffShotALDEx, DiffShotANCOMBC or DiffShotMaAsLin3
> on the same dataset instead.

Significant = Kruskal–Wallis p < `r p_thresh` AND |LDA| ≥ 2.0.

### Species {.tabset}
```{r lefse_species, eval=has_species}
lefse_sp <- run_lefse_tab(physeq_Species, "Species")
```

#### Barplot
```{r lefse_species_bar, eval=has_species, fig.width=14, fig.height=8}
if (!is.null(lefse_sp)) {
  d <- lefse_sp %>% filter(Significant) %>% arrange(ef_lda)
  if (nrow(d) > 0) {
    d$feature <- factor(d$feature, levels = d$feature)
    print(ggplot(d, aes(feature, ef_lda, fill = enrich_group)) + geom_col() +
            coord_flip() + labs(x = NULL, y = "LDA score (log10)",
            title = "LEfSe Species — significant markers",
            fill = "Enriched in") + theme_bw())
  } else cat("No significant LEfSe Species markers.\n")
}
```
```{r lefse_species_vdat, include=FALSE, eval=has_species}
lefse_sp_vd_adj <- NULL; lefse_sp_vd_raw <- NULL
if (!is.null(lefse_sp)) {
  .lv <- sort(unique(as.character(lefse_sp$enrich_group)))
  .d  <- lefse_sp
  .d$eff_signed <- ifelse(.d$enrich_group == .lv[length(.lv)], .d$ef_lda, -.d$ef_lda)
  .d$padj <- p.adjust(.d$pvalue, method = "BH")
  lefse_sp_vd_adj <- mk_volcano_df(.d, "eff_signed", "padj",   "feature")
  lefse_sp_vd_raw <- mk_volcano_df(.d, "eff_signed", "pvalue", "feature")
}
```

#### Volcano — Adjusted p (Static)
```{r lefse_species_vstat_adj, eval=has_species, fig.width=11, fig.height=8}
show_volcano_static(lefse_sp_vd_adj, "LEfSe Species (signed LDA vs BH-adj p)",
                    "signed LDA score", ylab = "-log10(BH-adjusted p)", fc = 2)
```

#### Volcano — Adjusted p (Interactive)
```{r lefse_species_vint_adj, eval=has_species}
show_volcano_interactive(lefse_sp_vd_adj, "LEfSe Species (signed LDA vs BH-adj p)",
                         "signed LDA score", ylab = "-log10(BH-adjusted p)", fc = 2)
```

#### Volcano — Raw p (Static)
```{r lefse_species_vstat_raw, eval=has_species, fig.width=11, fig.height=8}
show_volcano_static(lefse_sp_vd_raw, "LEfSe Species (signed LDA vs raw p)",
                    "signed LDA score", ylab = "-log10(raw KW/Wilcoxon p)", fc = 2)
```

#### Volcano — Raw p (Interactive)
```{r lefse_species_vint_raw, eval=has_species}
show_volcano_interactive(lefse_sp_vd_raw, "LEfSe Species (signed LDA vs raw p)",
                         "signed LDA score", ylab = "-log10(raw KW/Wilcoxon p)", fc = 2)
```

#### Table
```{r lefse_species_tab, eval=has_species}
if (!is.null(lefse_sp)) dt(lefse_sp, "LEfSe — Species (all tested taxa)")
```

### Genus {.tabset}
```{r lefse_genus}
lefse_ge <- run_lefse_tab(physeq_Genus, "Genus")
```

#### Barplot
```{r lefse_genus_bar, fig.width=14, fig.height=8}
if (!is.null(lefse_ge)) {
  d <- lefse_ge %>% filter(Significant) %>% arrange(ef_lda)
  if (nrow(d) > 0) {
    d$feature <- factor(d$feature, levels = d$feature)
    print(ggplot(d, aes(feature, ef_lda, fill = enrich_group)) + geom_col() +
            coord_flip() + labs(x = NULL, y = "LDA score (log10)",
            title = "LEfSe Genus — significant markers",
            fill = "Enriched in") + theme_bw())
  } else cat("No significant LEfSe Genus markers.\n")
}
```
```{r lefse_genus_vdat, include=FALSE}
lefse_ge_vd_adj <- NULL; lefse_ge_vd_raw <- NULL
if (!is.null(lefse_ge)) {
  .lv <- sort(unique(as.character(lefse_ge$enrich_group)))
  .d  <- lefse_ge
  .d$eff_signed <- ifelse(.d$enrich_group == .lv[length(.lv)], .d$ef_lda, -.d$ef_lda)
  .d$padj <- p.adjust(.d$pvalue, method = "BH")
  lefse_ge_vd_adj <- mk_volcano_df(.d, "eff_signed", "padj",   "feature")
  lefse_ge_vd_raw <- mk_volcano_df(.d, "eff_signed", "pvalue", "feature")
}
```

#### Volcano — Adjusted p (Static)
```{r lefse_genus_vstat_adj, fig.width=11, fig.height=8}
show_volcano_static(lefse_ge_vd_adj, "LEfSe Genus (signed LDA vs BH-adj p)",
                    "signed LDA score", ylab = "-log10(BH-adjusted p)", fc = 2)
```

#### Volcano — Adjusted p (Interactive)
```{r lefse_genus_vint_adj}
show_volcano_interactive(lefse_ge_vd_adj, "LEfSe Genus (signed LDA vs BH-adj p)",
                         "signed LDA score", ylab = "-log10(BH-adjusted p)", fc = 2)
```

#### Volcano — Raw p (Static)
```{r lefse_genus_vstat_raw, fig.width=11, fig.height=8}
show_volcano_static(lefse_ge_vd_raw, "LEfSe Genus (signed LDA vs raw p)",
                    "signed LDA score", ylab = "-log10(raw KW/Wilcoxon p)", fc = 2)
```

#### Volcano — Raw p (Interactive)
```{r lefse_genus_vint_raw}
show_volcano_interactive(lefse_ge_vd_raw, "LEfSe Genus (signed LDA vs raw p)",
                         "signed LDA score", ylab = "-log10(raw KW/Wilcoxon p)", fc = 2)
```

#### Table
```{r lefse_genus_tab}
if (!is.null(lefse_ge)) dt(lefse_ge, "LEfSe — Genus (all tested taxa)")
```
