---
params:
  reportTitle: "Differential Peak Analysis"
title: "`r params$reportTitle`"
---

<!-- Styling comes from the fgczQuartoTemplate assets staged next to this file
     (_metadata.yml is applied by Quarto automatically). ezRun stages them via
     makeQuartoReport(), which calls fgczQuartoTemplate::fgcz_render().
     All statistics are computed in ezMethodDiffPeakAnalysis; this report only
     reads the saved objects and draws them. -->

```{r setup}
#| include: false
library(ezRun)
library(DESeq2)
library(SummarizedExperiment)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(ggplot2)
library(ggrepel)
library(ComplexHeatmap)
library(RColorBrewer)
library(DT)
library(htmltools)

## -- Data loading (qs2 with rds fallback, checks data/ first) ---------------
.fgcz_load <- function(name, nthreads = 4L, optional = FALSE) {
  paths <- c(
    file.path("data", paste0(name, ".qs2")),
    paste0(name, ".qs2"),
    file.path("data", paste0(name, ".rds")),
    paste0(name, ".rds")
  )
  for (p in paths) {
    if (file.exists(p)) {
      if (grepl("\\.qs2$", p)) return(qs2::qs_read(p, nthreads = nthreads))
      return(readRDS(p))
    }
  }
  if (optional) return(NULL)
  stop("No .qs2 or .rds file found for: ", name)
}

mlDds     <- .fgcz_load("dds")
vsd       <- .fgcz_load("vsd")
peakTable <- .fgcz_load("peakTable")
deseqInfo <- .fgcz_load("deseqInfo")
param     <- .fgcz_load("param")
output    <- .fgcz_load("output")
normDiag  <- .fgcz_load("normDiag", optional = TRUE)
peakGc    <- .fgcz_load("peakGc", optional = TRUE)
motifRes  <- .fgcz_load("motifRes", optional = TRUE)
goRes     <- .fgcz_load("goRes", optional = TRUE)
profiles  <- .fgcz_load("profiles", optional = TRUE)

sampleGroup   <- param$sampleGroup
refGroup      <- param$refGroup
lfcThreshold  <- param$lfcThreshold
fdrThreshold  <- param$fdrThreshold
noReplicates  <- isTRUE(deseqInfo$noReplicates)
blockingUsed  <- isTRUE(deseqInfo$blockingUsed)
grouping2Name <- deseqInfo$grouping2Name
hasAnno       <- "geneName" %in% colnames(peakTable)
hasShrunk     <- "log2FoldChange_shrunk" %in% colnames(peakTable) &&
  any(!is.na(peakTable$log2FoldChange_shrunk))

groupColors <- c("#009E73", "#CC79A7")
names(groupColors) <- c(refGroup, sampleGroup)
dirColors <- c("up" = "#D55E00", "down" = "#0072B2")

## -- Summary counts ---------------------------------------------------------
nTested   <- sum(!is.na(peakTable$log2FoldChange))
nCand     <- sum(peakTable$candidate, na.rm = TRUE)
nUp       <- sum(peakTable$candidate & peakTable$direction == "up", na.rm = TRUE)
nDown     <- sum(peakTable$candidate & peakTable$direction == "down", na.rm = TRUE)

## Full result table for download, written by the app
resultsFile <- "differential_peaks.txt"

## -- Candidate gene lists for Enrichr ---------------------------------------
candGenes <- function(dir = NULL) {
  if (!hasAnno) return(character(0))
  tbl <- peakTable |> dplyr::filter(candidate)
  if (!is.null(dir)) tbl <- tbl |> dplyr::filter(direction == dir)
  g <- unique(tbl$geneName)
  g[!is.na(g) & g != ""]
}

## An HTML form that submits a gene list to Enrichr in a new tab. This is the
## documented Enrichr "add gene list from your website" pattern; it needs no
## external JS library, so it stays valid in self-contained HTML.
enrichrForm <- function(genes, description, label, maxGenes = 500L) {
  if (length(genes) == 0) {
    return(tags$p(tags$em(paste0("No candidate genes for ", label, "."))))
  }
  genes <- head(genes, maxGenes)
  tags$form(
    action = "https://maayanlab.cloud/Enrichr/enrich",
    method = "post", target = "_blank", enctype = "multipart/form-data",
    style = "display:inline-block; margin:0 10px 10px 0;",
    tags$input(type = "hidden", name = "list",
               value = paste(genes, collapse = "\n")),
    tags$input(type = "hidden", name = "description", value = description),
    tags$button(
      type = "submit",
      style = "padding:6px 12px; cursor:pointer; border-radius:6px; border:1px solid #888;",
      paste0(label, " (", length(genes), " genes) &rarr; Enrichr") |> HTML()
    )
  )
}

## -- Plot helpers (right-sized, single comparison) --------------------------
compLabel <- paste(sampleGroup, "vs", refGroup)

plotVolcano <- function() {
  d <- peakTable |> dplyr::filter(!is.na(padj))
  d$class <- dplyr::case_when(
    d$candidate & d$direction == "up"   ~ "Up",
    d$candidate & d$direction == "down" ~ "Down",
    TRUE ~ "Not sig"
  )
  top <- d |> dplyr::filter(candidate)
  if (hasAnno) top <- top |> dplyr::filter(!is.na(geneName) & geneName != "")
  top <- top |> dplyr::group_by(direction) |>
    dplyr::slice_min(padj, n = 10, with_ties = FALSE) |> dplyr::ungroup()

  p <- ggplot(d, aes(log2FoldChange, -log10(padj), color = class)) +
    geom_point(size = 0.9, alpha = 0.5) +
    scale_color_manual(name = "",
      values = c("Up" = "#D55E00", "Down" = "#0072B2", "Not sig" = "grey75"),
      breaks = c("Up", "Not sig", "Down")) +
    geom_vline(xintercept = c(-lfcThreshold, lfcThreshold),
               linetype = "dashed", color = "grey40") +
    geom_hline(yintercept = -log10(fdrThreshold),
               linetype = "dashed", color = "grey40") +
    labs(x = expression(Log[2]~fold~change), y = expression(-Log[10]~adjusted~p),
         title = paste("Volcano:", compLabel)) +
    theme_bw(base_size = 11) + theme(plot.title = element_text(hjust = 0.5))
  if (hasAnno && nrow(top) > 0) {
    p <- p + geom_label_repel(data = top, aes(label = geneName),
      color = "black", fill = "white", size = 3, min.segment.length = 0,
      max.overlaps = Inf, show.legend = FALSE)
  }
  p
}

plotMA <- function() {
  d <- peakTable |> dplyr::filter(!is.na(padj)) |>
    dplyr::mutate(significant = dplyr::if_else(candidate, "Candidate", "Not sig"),
                  baseMeanLog = log10(baseMean + 1))
  ggplot(d, aes(baseMeanLog, log2FoldChange, color = significant)) +
    geom_point(size = 0.9, alpha = 0.5) +
    scale_color_manual(name = "",
      values = c("Candidate" = "#0072B2", "Not sig" = "grey65"),
      breaks = c("Candidate", "Not sig")) +
    geom_hline(yintercept = c(-lfcThreshold, lfcThreshold),
               linetype = "dashed", color = "grey40") +
    geom_hline(yintercept = 0, color = "grey30") +
    geom_smooth(aes(group = 1), method = "loess", se = FALSE, color = "black",
                linewidth = 0.8,
                data = d[sample(nrow(d), min(nrow(d), 20000)), ]) +
    labs(x = expression(Log[10]~mean~normalized~count),
         y = expression(Log[2]~fold~change), title = paste("MA:", compLabel)) +
    theme_bw(base_size = 11) + theme(plot.title = element_text(hjust = 0.5))
}

plotPca <- function() {
  pcaData <- DESeq2::plotPCA(vsd, intgroup = "Condition", returnData = TRUE)
  pv <- round(100 * attr(pcaData, "percentVar"))
  ## when a blocking factor was used, encode it as point shape
  if (blockingUsed && "grouping2" %in% colnames(SummarizedExperiment::colData(vsd))) {
    pcaData$block <- as.character(
      SummarizedExperiment::colData(vsd)[pcaData$name, "grouping2"])
    p <- ggplot(pcaData, aes(PC1, PC2, color = group, shape = block, label = name)) +
      geom_point(size = 3) +
      scale_shape_discrete(name = grouping2Name)
  } else {
    p <- ggplot(pcaData, aes(PC1, PC2, color = group, label = name)) +
      geom_point(size = 3)
  }
  p +
    ggrepel::geom_text_repel(size = 3, show.legend = FALSE) +
    scale_color_manual(values = groupColors, name = "Group") +
    labs(title = paste("PCA (all peaks):", compLabel),
         x = paste0("PC1 (", pv[1], "%)"), y = paste0("PC2 (", pv[2], "%)")) +
    theme_bw(base_size = 11) + theme(plot.title = element_text(hjust = 0.5))
}

## Candidate-based heatmap. Returns a list(plot, caption) or NULL when there are
## no candidate peaks (the chunk then prints a 'no candidates' message).
heatMaxRows <- 2000
plotHeat <- function() {
  cand <- peakTable |> dplyr::filter(candidate)
  nCandTotal <- nrow(cand)
  if (nCandTotal == 0) return(NULL)
  ## rank candidates: fold change without replicates, adjusted p-value otherwise
  ord <- if (noReplicates) {
    order(-abs(cand$log2FoldChange))
  } else {
    order(cand$padj, na.last = TRUE)
  }
  capped <- nCandTotal > heatMaxRows
  cand <- cand[head(ord, min(heatMaxRows, nCandTotal)), , drop = FALSE]
  mat <- assay(vsd)[cand$peakId, , drop = FALSE]
  ## Row z-scaling / clustering are undefined for zero- or non-finite-variance
  ## rows (they become NaN and break hclust); drop them.
  rowSd <- apply(mat, 1, stats::sd)
  mat <- mat[is.finite(rowSd) & rowSd > 0, , drop = FALSE]
  if (nrow(mat) < 2) return(NULL)
  annotation_col <- data.frame(Condition = vsd$Condition,
                               row.names = colnames(mat))
  caption <- if (capped) {
    paste0("Row z-scored VST of the top ", nrow(mat), " of ", nCandTotal,
           " candidate peaks.")
  } else {
    paste0("Row z-scored VST of ", nrow(mat), " candidate peaks.")
  }
  gg <- ComplexHeatmap::pheatmap(mat, scale = "row", show_rownames = FALSE,
    annotation_col = annotation_col,
    annotation_colors = list(Condition = groupColors),
    color = colorRampPalette(rev(brewer.pal(7, "RdYlBu")))(100),
    clustering_method = "complete", angle_col = "45",
    name = "z-score",
    column_title = paste0(nrow(mat), " candidate peaks"))
  list(plot = gg, caption = caption)
}

## Genomic feature and TSS-distance columns depend on the annotation tool.
featureCol <- intersect(c("Annotation", "annotation", "insideFeature"),
                        colnames(peakTable))
featureCol <- if (length(featureCol)) featureCol[1] else NA_character_
distCol    <- intersect(c("Distance to TSS", "distanceToTSS",
                          "shortestDistance", "distancetoFeature"),
                        colnames(peakTable))
distCol    <- if (length(distCol)) distCol[1] else NA_character_

plotFeatureDist <- function() {
  if (is.na(featureCol)) return(NULL)
  d <- peakTable |>
    dplyr::mutate(feature = str_replace(as.character(.data[[featureCol]]),
                                        " \\(.*$", ""),
                  set = dplyr::if_else(candidate, "Candidates", "All peaks")) |>
    dplyr::filter(!is.na(feature) & feature != "")
  if (nrow(d) == 0) return(NULL)
  d <- d |> dplyr::count(set, feature) |>
    dplyr::group_by(set) |> dplyr::mutate(frac = n / sum(n)) |> dplyr::ungroup()
  ggplot(d, aes(reorder(feature, frac), frac, fill = set)) +
    geom_col(position = "dodge") + coord_flip() +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_manual(values = c("All peaks" = "grey60",
                                 "Candidates" = "#D55E00"), name = "") +
    labs(x = "Genomic feature", y = "Fraction of peaks",
         title = "Peak genomic-feature distribution") +
    theme_bw(base_size = 11) + theme(plot.title = element_text(hjust = 0.5))
}

plotTssDist <- function() {
  if (is.na(distCol)) return(NULL)
  d <- peakTable |>
    dplyr::mutate(dist = suppressWarnings(as.numeric(.data[[distCol]])),
                  set = dplyr::if_else(candidate, "Candidates", "All peaks")) |>
    dplyr::filter(!is.na(dist) & abs(dist) <= 1e5)
  if (nrow(d) == 0) return(NULL)
  ggplot(d, aes(dist / 1e3, fill = set)) +
    geom_histogram(bins = 60, position = "identity", alpha = 0.5) +
    scale_fill_manual(values = c("All peaks" = "grey60",
                                 "Candidates" = "#D55E00"), name = "") +
    labs(x = "Distance to TSS (kb)", y = "Number of peaks",
         title = "Distance of peaks to nearest TSS") +
    theme_bw(base_size = 11) + theme(plot.title = element_text(hjust = 0.5))
}

## -- Motifs -----------------------------------------------------------------
plotMotifDots <- function(tab, direction, nTop = 20) {
  if (is.null(tab) || nrow(tab) == 0) return(NULL)
  d <- head(tab[order(-tab$log10P), ], nTop)
  d$label <- factor(d$motif, levels = rev(unique(d$motif)))
  ggplot(d, aes(log2(enrichment), label)) +
    geom_segment(aes(x = 0, xend = log2(enrichment), yend = label),
                 color = "grey80") +
    geom_point(aes(size = pctTarget, fill = pmin(log10P, 300)), shape = 21,
               color = "white") +
    scale_fill_gradient(low = "#C6DBEF", high = "#08306B",
                        name = expression(-log[10]~p)) +
    scale_size_area(max_size = 7, name = "% targets") +
    guides(size = guide_legend(override.aes = list(fill = "grey50"))) +
    geom_vline(xintercept = 0, color = "grey50") +
    labs(x = expression(log[2]~("% targets" / "% background")), y = NULL,
         title = paste0("Top known motifs, ", direction, " candidates")) +
    theme_bw(base_size = 11) + theme(plot.title = element_text(hjust = 0.5))
}

motifTable <- function(tab, nTop = 100) {
  if (is.null(tab) || nrow(tab) == 0) return(NULL)
  d <- head(tab[order(-tab$log10P), ], nTop)
  d <- d[, c("rank", "motif", "family", "consensus", "log10P", "qValue",
             "pctTarget", "pctBackground", "enrichment", "source")]
  colnames(d) <- c("Rank", "Motif", "Family", "Consensus", "-log10 p", "q-value",
                   "% targets", "% background", "Enrichment", "Source")
  ezInteractiveTableRmd(d, digits = 3)
}

motifLink <- function(direction) {
  dirName <- motifRes$dirs[[direction]]
  links <- list()
  if (file.exists(file.path(dirName, "knownResults.html"))) {
    links <- c(links, list(tags$a(href = file.path(dirName, "knownResults.html"),
      target = "_blank", "HOMER known-motif report (with logos)")))
  }
  if (file.exists(file.path(dirName, "homerResults.html"))) {
    links <- c(links, list(" | ", tags$a(href = file.path(dirName, "homerResults.html"),
      target = "_blank", "HOMER de novo motifs")))
  }
  if (length(links) == 0) return(NULL)
  tags$p(links)
}

## -- Signal profiles --------------------------------------------------------
plotMetaProfile <- function() {
  if (is.null(profiles)) return(NULL)
  d <- dplyr::bind_rows(lapply(c("up", "down"), function(dir) {
    p <- profiles[[dir]]
    if (is.null(p)) return(NULL)
    dplyr::mutate(p$meta, direction = paste0(dir, " candidates (n = ",
                                            length(p$peakIds), ")"))
  }))
  if (nrow(d) == 0) return(NULL)
  ggplot(d, aes(position / 1e3, signal, color = group, group = sample)) +
    geom_line(linewidth = 0.6, alpha = 0.8) +
    scale_color_manual(values = groupColors, name = "Group") +
    facet_wrap(~direction, scales = "free_y") +
    labs(x = "Distance to peak centre (kb)", y = "Mean BigWig signal",
         title = "Signal around differential peaks (one line per sample)") +
    theme_bw(base_size = 11) + theme(plot.title = element_text(hjust = 0.5))
}

drawProfileHeatmaps <- function(dir) {
  p <- profiles[[dir]]
  if (is.null(p)) return(NULL)
  mats <- p$groupMats[intersect(c(refGroup, sampleGroup), names(p$groupMats))]
  upper <- stats::quantile(unlist(lapply(mats, unclass)), 0.99, na.rm = TRUE)
  colFun <- circlize::colorRamp2(c(0, upper), c("white", "#08306B"))
  hl <- NULL
  for (g in names(mats)) {
    h <- EnrichedHeatmap::EnrichedHeatmap(
      mats[[g]], col = colFun, name = g, column_title = g,
      show_heatmap_legend = identical(g, names(mats)[1]),
      heatmap_legend_param = list(title = "signal"),
      top_annotation = ComplexHeatmap::HeatmapAnnotation(
        lines = EnrichedHeatmap::anno_enriched(gp = grid::gpar(col = groupColors[[g]]))),
      axis_name = c(paste0("-", profiles$extend / 1e3, "kb"), "centre",
                    paste0(profiles$extend / 1e3, "kb")),
      use_raster = TRUE)
    hl <- if (is.null(hl)) h else hl + h
  }
  ComplexHeatmap::draw(hl, column_title = paste0(dir, " candidates, ranked by significance"))
}

## -- Diagnostics ------------------------------------------------------------
binnedLfc <- function(x, lfc, nBins = 20) {
  ok <- is.finite(x) & is.finite(lfc)
  br <- unique(stats::quantile(x[ok], probs = seq(0, 1, length.out = nBins + 1)))
  bin <- cut(x[ok], br, include.lowest = TRUE)
  d <- data.frame(x = x[ok], lfc = lfc[ok], bin = bin)
  d |> dplyr::group_by(bin) |>
    dplyr::summarise(x = stats::median(x), median = stats::median(lfc),
                     q25 = stats::quantile(lfc, 0.25), q75 = stats::quantile(lfc, 0.75),
                     n = dplyr::n(), .groups = "drop")
}

plotBias <- function(var, label) {
  if (is.null(peakGc)) return(NULL)
  d <- dplyr::inner_join(peakTable[, c("peakId", "log2FoldChange")], peakGc, by = "peakId")
  b <- binnedLfc(d[[var]], d$log2FoldChange)
  p <- ggplot(b, aes(x, median)) +
    geom_hline(yintercept = 0, color = "grey50") +
    geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#C6DBEF") +
    geom_line(color = "#08519C", linewidth = 0.8) + geom_point(color = "#08519C", size = 1.5) +
    labs(x = label, y = "log2 fold change (median, IQR)",
         title = paste("Fold change vs", label)) +
    theme_bw(base_size = 11) + theme(plot.title = element_text(hjust = 0.5))
  if (var == "width") p <- p + scale_x_log10()
  p
}
```

:::::: {.panel-tabset}

# Overview

::: {.callout-note}
## Analysis summary

Differential accessibility/binding analysis of consensus peaks with **DESeq2**,
comparing **`r sampleGroup`** over **`r refGroup`**
(grouping column *`r param$grouping`*, annotation *`r param$annotationMethod`*).
Peak counts were variance-stabilized (VST). A peak is called a **candidate** at
|log2 fold change| &ge; `r lfcThreshold`
`r if (noReplicates) "" else paste0("and adjusted p-value < ", fdrThreshold)`.
Positive log2 fold change means higher signal in **`r sampleGroup`**.
:::

```{r design-note}
#| output: asis
if (blockingUsed) {
  cat("**Design:** paired/blocked - `~ group + ", grouping2Name, "`.\n", sep = "")
} else if (ezIsSpecified(param$grouping2)) {
  cat("**Design:** single-factor `~ group`. The requested second factor **",
      param$grouping2, "** was dropped (single level, confounded with the ",
      "groups, or too few residual degrees of freedom).\n", sep = "")
} else {
  cat("**Design:** single-factor `~ group`.\n", sep = "")
}
```

```{r no-rep-warning}
#| output: asis
if (noReplicates) {
  cat(
    ":::: {.callout-warning}\n",
    "## No biological replicates in this comparison\n\n",
    "At least one group has a single sample, so DESeq2 cannot estimate ",
    "within-group dispersion in the usual way. Dispersion was estimated with a ",
    "blind (`~1`) design that treats the two conditions as one group, which ",
    "makes the Wald test **conservative**: adjusted p-values are typically not ",
    "significant and should not be trusted. Candidates below are therefore ",
    "ranked by **fold change only** (|log2FC| &ge; ", lfcThreshold, "). ",
    "Add replicates for a statistically robust analysis.\n",
    "::::\n", sep = ""
  )
}
```

```{r summary-table}
summaryTab <- data.frame(
  Metric = c("Peaks tested", "Candidate peaks",
             paste0("Up in ", sampleGroup), paste0("Down in ", sampleGroup)),
  Count  = c(nTested, nCand, nUp, nDown)
)
DT::datatable(summaryTab, rownames = FALSE, options = list(dom = "t"),
              caption = "Peak-level summary")
```

```{r settings-table}
settingsTab <- data.frame(
  Setting = c("Size factors (normMethod)", "Samples in the fit",
              "Test", "Fold-change shrinkage (extra column)",
              "|log2FC| threshold", "FDR threshold",
              "Motif / GO input (top candidates per direction)"),
  Value = c(
    deseqInfo$normMethod,
    paste0(deseqInfo$nFitSamples,
           if (isTRUE(deseqInfo$fitAllSamples)) " (all samples of the dataset)" else " (the two groups)"),
    if (isTRUE(deseqInfo$lfcTest)) paste0("Wald, H0: |log2FC| <= ", lfcThreshold) else "Wald, H0: log2FC = 0",
    deseqInfo$lfcShrinkType,
    lfcThreshold, fdrThreshold,
    param$enrichMaxPeaks
  )
)
DT::datatable(settingsTab, rownames = FALSE, options = list(dom = "t"),
              caption = "Analysis settings")
```

# Explorative analysis

Unsupervised views based on **all** consensus peaks, independent of the
differential test.

::: {.panel-tabset}

## PCA

```{r pca}
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Principal component analysis of variance-stabilized counts over all peaks."
res_pca <- tryCatch(plotPca(), error = function(e) NULL)
if (is.null(res_pca)) cat("PCA could not be computed for this comparison.") else res_pca
```

## Feature distribution

```{r feature-dist}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Distribution of peaks across genomic features, comparing all peaks with the differential candidates."
fd <- tryCatch(plotFeatureDist(), error = function(e) NULL)
if (is.null(fd)) cat("Genomic-feature annotation is not available for this run.") else fd
```

## Distance to TSS

```{r tss-dist}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Distribution of peak distance to the nearest transcription start site (within 100 kb)."
td <- tryCatch(plotTssDist(), error = function(e) NULL)
if (is.null(td)) cat("Distance-to-TSS annotation is not available for this run.") else td
```

:::

# Differential peaks

::: {.panel-tabset}

## Volcano

```{r volcano}
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Volcano plot. Dashed lines mark the fold-change and significance thresholds; the top 10 up- and down-regulated annotated peaks are labelled."
plotVolcano()
```

## MA

```{r ma}
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "MA plot of log2 fold change against mean normalized count. The black line is a loess fit; it should stay close to 0 unless the size factors do not match a global shift (see Diagnostics)."
plotMA()
```

## Heatmap

```{r heatmap}
#| fig-width: 7
#| fig-height: 8
h <- tryCatch(plotHeat(), error = function(e) NULL)
if (is.null(h)) {
  cat("**No candidate peaks - the heatmap is only shown when differential candidates exist.**")
} else {
  ComplexHeatmap::draw(h$plot)
}
```

```{r heatmap-caption}
#| output: asis
if (!is.null(h)) cat("*", h$caption, "*\n", sep = "")
```

## Signal profiles

```{r meta-profile}
#| fig-width: 10
#| fig-height: 4
#| out-width: "80%"
#| fig-cap: "Mean BigWig signal +-2 kb around the centre of the top up and down candidates. The groups should separate in the expected direction."
mp <- tryCatch(plotMetaProfile(), error = function(e) NULL)
if (is.null(mp)) cat("BigWig tracks are not available for this dataset.") else mp
```

```{r profile-heat-up}
#| fig-width: 6
#| fig-height: 7
#| out-width: "45%"
if (!is.null(profiles) && !is.null(profiles$up)) {
  tryCatch(drawProfileHeatmaps("up"), error = function(e) cat("Heatmap failed:", conditionMessage(e)))
}
```

```{r profile-heat-down}
#| fig-width: 6
#| fig-height: 7
#| out-width: "45%"
if (!is.null(profiles) && !is.null(profiles$down)) {
  tryCatch(drawProfileHeatmaps("down"), error = function(e) cat("Heatmap failed:", conditionMessage(e)))
}
```

:::

# Motifs

```{r motif-intro}
#| output: asis
if (is.null(motifRes)) {
  cat("Motif enrichment was not run (or failed) for this comparison.\n")
} else {
  cat("Known-motif enrichment with **HOMER** (`findMotifsGenome.pl`, motif set *",
      motifRes$motifSet, "*) in 200 bp windows centred on the top ",
      param$enrichMaxPeaks, " candidates per direction. The background is ",
      format(motifRes$nBackground, big.mark = ","), " tested, non-candidate ",
      "consensus peaks, so enrichment is relative to other accessible regions ",
      "rather than the whole genome. Motifs of one family (e.g. HIF-1a/HIF2a, ",
      "AP-1 members) share a consensus and cannot be told apart.\n", sep = "")
}
```

::: {.panel-tabset}

## Up in `r sampleGroup`

```{r motif-up-plot}
#| fig-width: 8
#| fig-height: 6
#| out-width: "60%"
mu <- if (!is.null(motifRes)) plotMotifDots(motifRes$up, "up") else NULL
if (is.null(mu)) cat("No motif results for up candidates (too few candidates or not run).") else mu
```

```{r motif-up-link}
if (!is.null(motifRes)) motifLink("up")
```

```{r motif-up-table}
if (!is.null(motifRes)) motifTable(motifRes$up)
```

## Down in `r sampleGroup`

```{r motif-down-plot}
#| fig-width: 8
#| fig-height: 6
#| out-width: "60%"
md <- if (!is.null(motifRes)) plotMotifDots(motifRes$down, "down") else NULL
if (is.null(md)) cat("No motif results for down candidates (too few candidates or not run).") else md
```

```{r motif-down-link}
if (!is.null(motifRes)) motifLink("down")
```

```{r motif-down-table}
if (!is.null(motifRes)) motifTable(motifRes$down)
```

:::

# Pathway enrichment

```{r go-intro}
#| output: asis
if (is.null(goRes)) {
  cat("GO over-representation was not run (only mouse and human are supported, ",
      "or no gene annotation was available).\n\n", sep = "")
} else {
  cat("GO ", goRes$ont, " over-representation (clusterProfiler, ", goRes$orgDb,
      ") of the nearest genes of the top ", param$enrichMaxPeaks,
      " candidates per direction. The universe is the ",
      format(goRes$nUniverse, big.mark = ","), " nearest genes of all tested ",
      "peaks, which corrects for genes that simply carry many peaks.\n\n", sep = "")
}
```

::: {.panel-tabset}

## GO: up in `r sampleGroup`

```{r go-up}
#| fig-width: 9
#| fig-height: 8
#| out-width: "60%"
goPlot <- function(ego, dir) {
  if (is.null(ego) || nrow(as.data.frame(ego)) == 0) {
    cat("No significant GO terms for", dir, "candidates.")
    return(invisible(NULL))
  }
  print(enrichplot::dotplot(ego, showCategory = 20, label_format = 50) +
          ggtitle(paste("GO", goRes$ont, "-", dir, "candidates")) +
          theme(axis.text.y = element_text(size = 9)))
}
if (!is.null(goRes)) goPlot(goRes$up, "up")
```

```{r go-up-table}
goTable <- function(ego) {
  if (is.null(ego) || nrow(as.data.frame(ego)) == 0) return(NULL)
  d <- as.data.frame(ego)[, c("ID", "Description", "GeneRatio", "BgRatio",
                              "pvalue", "p.adjust", "geneID")]
  ezInteractiveTableRmd(head(d, 200), digits = 3)
}
if (!is.null(goRes)) goTable(goRes$up)
```

## GO: down in `r sampleGroup`

```{r go-down}
#| fig-width: 9
#| fig-height: 8
#| out-width: "60%"
if (!is.null(goRes)) goPlot(goRes$down, "down")
```

```{r go-down-table}
if (!is.null(goRes)) goTable(goRes$down)
```

## Enrichr

Submit the candidate genes associated with differential peaks to
[Enrichr](https://maayanlab.cloud/Enrichr/) (opens in a new tab). Up to 500
genes per list are sent. Enrichr uses its own background, not the peak
universe used above.

```{r enrichr-buttons}
if (!hasAnno) {
  htmltools::tags$p(htmltools::tags$em(
    "Peak annotation is unavailable, so gene-based enrichment links cannot be built."))
} else {
  htmltools::tagList(
    enrichrForm(candGenes("up"),  paste0(compLabel, " - up candidates"),   "Up candidates"),
    enrichrForm(candGenes("down"),paste0(compLabel, " - down candidates"), "Down candidates"),
    enrichrForm(candGenes(),      paste0(compLabel, " - all candidates"),  "All candidates")
  )
}
```

:::

# Diagnostics

::: {.panel-tabset}

## Normalisation

```{r norm-intro}
#| output: asis
cat("Size factors decide what \"no change\" means. DESeq2 median-of-ratios ",
    "assumes most peaks do not change; reads-in-peaks assumes the total signal ",
    "in peaks is comparable; TMM trims extreme peaks. When the methods disagree, ",
    "a global shift of accessibility is likely and the number and direction of ",
    "candidates depend on the choice. The table below approximates the ",
    "candidate counts under each method **without refitting** (all fold changes ",
    "shift by the same offset; Wald statistics are recomputed with the fitted ",
    "standard errors). Used here: **", deseqInfo$normMethod, "**.\n", sep = "")
```

```{r norm-balance}
if (!is.null(normDiag)) {
  b <- normDiag$balance
  b$lfcOffset <- round(b$lfcOffset, 3)
  b$medianLfc <- round(b$medianLfc, 3)
  colnames(b) <- c("Size factors", "log2FC offset vs used", "Median log2FC",
                   "Up candidates (approx.)", "Down candidates (approx.)")
  DT::datatable(b, rownames = FALSE, options = list(dom = "t"),
                caption = "Candidate balance under alternative size factors")
}
```

```{r norm-sf}
#| fig-width: 8
#| fig-height: 4
#| out-width: "60%"
#| fig-cap: "Size factors per sample under the three methods (geometric mean 1 for reads-in-peaks and TMM)."
if (!is.null(normDiag)) {
  sf <- normDiag$sizeFactors
  sfLong <- tidyr::pivot_longer(sf, cols = c("DESeq2", "readsInPeaks", "TMM"),
                                names_to = "method", values_to = "sizeFactor")
  ggplot(sfLong, aes(sample, sizeFactor, color = method, group = method)) +
    geom_point(size = 2) + geom_line(alpha = 0.5) +
    scale_color_manual(values = c("DESeq2" = "#0072B2", "readsInPeaks" = "#E69F00",
                                  "TMM" = "#009E73"), name = "Method") +
    facet_grid(~group, scales = "free_x", space = "free_x") +
    labs(x = NULL, y = "Size factor") +
    theme_bw(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## GC and width

```{r gc-bias}
#| fig-width: 7
#| fig-height: 4
#| out-width: "50%"
#| fig-cap: "Median log2 fold change per GC-content bin (ribbon: interquartile range). A monotone trend indicates a GC-dependent technical difference between the groups (e.g. PCR or Tn5 batch)."
g1 <- tryCatch(plotBias("gc", "peak GC content"), error = function(e) NULL)
if (is.null(g1)) cat("GC content not available.") else g1
```

```{r width-bias}
#| fig-width: 7
#| fig-height: 4
#| out-width: "50%"
#| fig-cap: "Median log2 fold change per peak-width bin. Narrow peaks are noisier; a strong trend points to differences in fragment-size distribution between the groups."
g2 <- tryCatch(plotBias("width", "peak width (bp)"), error = function(e) NULL)
if (is.null(g2)) cat("Peak width not available.") else g2
```

:::

# Results table

The interactive table lists the top 1000 candidate peaks. The complete table
for all tested peaks is written to
[`differential_peaks.txt`](differential_peaks.txt).
`r if (hasShrunk) paste0("The column *log2FoldChange_shrunk* holds the ", deseqInfo$lfcShrinkType, "-shrunken estimate; candidates use the unshrunken *log2FoldChange*.") else ""`

```{r results-table}
displayCols <- intersect(
  c("peakId", "geneName", featureCol, distCol, "baseMean",
    "log2FoldChange", "log2FoldChange_shrunk", "lfcSE", "stat", "pvalue",
    "padj", "direction"),
  colnames(peakTable))
if (!hasShrunk) displayCols <- setdiff(displayCols, "log2FoldChange_shrunk")
tableTib <- peakTable |>
  dplyr::filter(candidate) |>
  ## rank by fold change without replicates (p-values unreliable), else by FDR
  dplyr::arrange(if (noReplicates) dplyr::desc(abs(log2FoldChange)) else padj) |>
  dplyr::select(dplyr::all_of(displayCols))
if (hasAnno) {
  tableTib <- tableTib |> dplyr::distinct(geneName, .keep_all = TRUE)
}
tableTib <- head(tableTib, 1000)
ezInteractiveTableRmd(tableTib, digits = 3)
```

# Session Info

```{r session-info}
ezSessionInfo()
```

::::::
