---
title: "`r if (exists('reportTitle')) reportTitle else 'SUSHI Report'`"
output:
  html_document:
    mathjax: https://fgcz-gstore.uzh.ch/reference/mathjax.js
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package="ezRun")
    css: !expr system.file("templates/fgcz.css", package="ezRun")
editor_options:
  chunk_output_type: inline
---

# {.tabset}

```{r setup DnaBamStats, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
library(DT)
library(htmltools)
library(knitr)
library(plotly)
library(tidyverse)
library(ezRun)
```

```{r style DnaBamStats, include=FALSE, results='asis'}
cat(
  "<style>",
  ".dna-bamstats-table { display: inline-block; width: auto; max-width: 100%; }",
  "</style>",
  sep = ""
)
```

```{r prepare data DnaBamStats, include=FALSE}
if (!exists("dataset")) {
  dataset <- ezLoadRobj("dataset")
}
if (!exists("param")) {
  param <- ezLoadRobj("param")
}
if (!exists("resultList")) {
  resultList <- ezLoadRobj("resultList")
}
if (!exists("reportData") && file.exists("reportData.qs2")) {
  reportData <- ezLoadRobj("reportData")
}
if (!exists("reportData")) {
  reportData <- list()
}

samples <- rownames(dataset)
if (is.null(reportData$reviewOrder)) {
  reportData$reviewOrder <- samples
}
if (is.null(reportData$fullSummaryTable)) {
  reportData$fullSummaryTable <- reportData$overviewTable
}
if (is.null(reportData$quickViewTable)) {
  reportData$quickViewTable <- reportData$fullSummaryTable
}
if (is.null(reportData$resourceTable)) {
  reportData$resourceTable <- data.frame(Sample = samples, stringsAsFactors = FALSE)
}
if (is.null(reportData$inputDatasetTable)) {
  reportData$inputDatasetTable <- data.frame(Sample = samples, stringsAsFactors = FALSE)
}
if (is.null(reportData$flagSummaryTable)) {
  reportData$flagSummaryTable <- data.frame(
    Flag = character(),
    SamplesAffected = numeric()
  )
}
if (is.null(reportData$qcThresholds)) {
  reportData$qcThresholds <- data.frame(
    Flag = character(),
    Metric = character(),
    Direction = character(),
    Threshold = numeric(),
    Unit = character()
  )
}
if (is.null(reportData$legacyAvailabilityNotes)) {
  reportData$legacyAvailabilityNotes <- list(
    alignment = character(),
    duplicates = character(),
    resources = character()
  )
}

fullSummaryTable <- reportData$fullSummaryTable
quickViewTable <- reportData$quickViewTable
resourceTable <- reportData$resourceTable
inputDatasetTable <- reportData$inputDatasetTable
flagSummaryTable <- reportData$flagSummaryTable
qcThresholds <- reportData$qcThresholds
reviewOrder <- reportData$reviewOrder
legacyAvailabilityNotes <- reportData$legacyAvailabilityNotes

metricPlotHeight <- min(max(420, 80 + 20 * length(reviewOrder)), 1200)
hasMmCounts <- FALSE
mmPlotWidth <- 8
```

```{r helper functions DnaBamStats, include=FALSE}
make_resource_link <- function(path) {
  if (is.null(path) || is.na(path) || path == "" || !file.exists(path)) {
    return("")
  }
  paste0('<a href="', path, '">', basename(path), "</a>")
}

make_qualimap_link <- function(path) {
  if (is.null(path) || is.na(path) || path == "" || !file.exists(path)) {
    return("")
  }
  paste0('<a href="', path, '">', basename(dirname(path)), "/qualimapReport.html</a>")
}

get_threshold_row <- function(metricName) {
  idx <- which(qcThresholds$Metric == metricName)
  if (length(idx) == 0) {
    return(NULL)
  }
  qcThresholds[idx[1], , drop = FALSE]
}

alignmentCountBarPlotDna <- function(mmCounts, relative = FALSE, file = NULL) {
  multiCountColors <- c(
    "0 hit(s)" = "#9e9e9e",
    "1 hit(s)" = "#1f77b4",
    "2 hit(s)" = "#ff7f0e",
    "3 hit(s)" = "#2ca02c",
    ">3 hit(s)" = "#d62728"
  )
  alignmentCountBarPlot(
    mmCounts = mmCounts,
    relative = relative,
    file = file,
    multiCountColors = multiCountColors
  )
}

plotDataBase <- fullSummaryTable[
  match(reviewOrder, fullSummaryTable$Sample),
  ,
  drop = FALSE
]
plotDataBase$SampleFactor <- factor(
  plotDataBase$Sample,
  levels = rev(reviewOrder)
)

make_metric_plot <- function(
  metricName,
  title,
  unit = "",
  showLabels = FALSE,
  thresholded = TRUE
) {
  plotData <- plotDataBase

  thresholdRow <- if (thresholded) {
    get_threshold_row(metricName)
  } else {
    NULL
  }
  if (!is.null(thresholdRow)) {
    flagName <- thresholdRow$Flag[1]
    plotData$PointColor <- ifelse(plotData[[flagName]], "#b22222", "#7f7f7f")
    thresholdText <- paste0(
      ifelse(thresholdRow$Direction[1] == "below", "< ", "> "),
      sprintf("%.1f", thresholdRow$Threshold[1]),
      thresholdRow$Unit[1]
    )
  } else {
    plotData$PointColor <- "#7f7f7f"
    thresholdText <- NULL
  }

  metricValues <- plotData[[metricName]]
  metricLabels <- format_dna_bamstats_metric(metricValues, unit = unit)
  hoverText <- if (!is.null(thresholdText)) {
    paste(
      "Sample:", plotData$Sample,
      "<br>Value:", metricLabels,
      "<br>Threshold:", thresholdText
    )
  } else {
    paste(
      "Sample:", plotData$Sample,
      "<br>Value:", metricLabels
    )
  }

  plotObj <- plot_ly(
    data = plotData,
    x = metricValues,
    y = ~SampleFactor,
    type = "scatter",
    mode = "markers",
    text = hoverText,
    hovertemplate = "%{text}<extra></extra>",
    marker = list(
      size = 10,
      color = plotData$PointColor,
      line = list(color = "#333333", width = 0.7)
    ),
    hoverlabel = list(
      bgcolor = "rgba(255,255,255,0.78)",
      bordercolor = "#666666",
      font = list(color = "#222222")
    ),
    showlegend = FALSE
  ) %>%
    layout(
      height = metricPlotHeight,
      xaxis = list(
        title = if (nzchar(unit)) paste0(title, " (", trimws(unit), ")") else title,
        rangemode = "tozero"
      ),
      yaxis = list(
        title = "",
        showticklabels = showLabels
      ),
      margin = list(l = if (showLabels) 130 else 40, r = 20, t = 30, b = 60)
    )

  if (!is.null(thresholdRow)) {
    plotObj <- plotObj %>%
      layout(
        shapes = list(
          list(
            type = "line",
            x0 = thresholdRow$Threshold[1],
            x1 = thresholdRow$Threshold[1],
            y0 = 0,
            y1 = 1,
            yref = "paper",
            line = list(color = "#333333", dash = "dash", width = 1.2)
          )
        )
      )
  }

  plotObj
}

resourceTableDisplay <- resourceTable
if (nrow(resourceTableDisplay) > 0) {
  resourceTableDisplay$QualimapReport <- vapply(
    resourceTable$QualimapReport,
    make_qualimap_link,
    character(1)
  )
  resourceTableDisplay$PicardMetrics <- vapply(
    resourceTable$PicardMetrics,
    make_resource_link,
    character(1)
  )
  resourceTableDisplay$FragmentSizePlot <- vapply(
    resourceTable$FragmentSizePlot,
    make_resource_link,
    character(1)
  )
  resourceTableDisplay$LibraryComplexityPlot <- vapply(
    resourceTable$LibraryComplexityPlot,
    make_resource_link,
    character(1)
  )
}

fullSummaryTableDisplay <- fullSummaryTable
if ("TriggeredFlags" %in% colnames(fullSummaryTableDisplay)) {
  fullSummaryTableDisplay$TriggeredFlags <- vapply(
    seq_len(nrow(fullSummaryTableDisplay)),
    function(i) {
      flagText <- get_dna_bamstats_triggered_flags_text(
        sampleRow = fullSummaryTableDisplay[i, , drop = FALSE],
        qcThresholds = qcThresholds,
        formatted = TRUE
      )
      if (is.na(flagText) || flagText == "") {
        return("NoFlags")
      }
      gsub("; ", "<br>", flagText, fixed = TRUE)
    },
    character(1)
  )
}
```

## Quick View
`r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

```{r quick view table, echo=FALSE}
htmltools::div(
  class = "dna-bamstats-table",
  DT::datatable(
    quickViewTable,
    rownames = FALSE,
    options = list(
      dom = "ftip",
      pageLength = min(max(nrow(quickViewTable), 5), 25),
      order = list(list(1, "desc"), list(0, "asc")),
      autoWidth = TRUE,
      columnDefs = list(
        list(
          targets = 2,
          render = JS(
            "function(data, type, row, meta) {",
            "  if (type !== 'display') { return data; }",
            "  if (data === null || data === '' || isNaN(data)) { return 'NA'; }",
            "  return Number(data).toFixed(1) + '%';",
            "}"
          )
        ),
        list(
          targets = 3,
          render = JS(
            "function(data, type, row, meta) {",
            "  if (type !== 'display') { return data; }",
            "  if (data === null || data === '' || isNaN(data)) { return 'NA'; }",
            "  return Number(data).toFixed(1) + '%';",
            "}"
          )
        ),
        list(
          targets = 4,
          render = JS(
            "function(data, type, row, meta) {",
            "  if (type !== 'display') { return data; }",
            "  if (data === null || data === '' || isNaN(data)) { return 'NA'; }",
            "  return Number(data).toFixed(1) + '%';",
            "}"
          )
        ),
        list(
          targets = 5,
          render = JS(
            "function(data, type, row, meta) {",
            "  if (type !== 'display') { return data; }",
            "  if (data === null || data === '' || isNaN(data)) { return 'NA'; }",
            "  return Number(data).toFixed(1) + 'x';",
            "}"
          )
        )
      )
    ),
    callback = JS(
      "var container = $(table.table().container());",
      "container.css({'width':'auto','display':'inline-block'});",
      "table.rows().every(function(){",
      "  var row = this.node();",
      "  var data = this.data();",
      "  if (parseFloat(data[1]) > 0) {",
      "    $('td', row).css('background-color', 'rgba(178, 34, 34, 0.08)');",
      "  }",
      "});"
    )
  )
)
```

### Flag thresholds

```{r quick view threshold bullets, echo=FALSE, results='asis'}
for (i in seq_len(nrow(qcThresholds))) {
  cat(
    "- `",
    qcThresholds$Flag[i],
    "`: ",
    ifelse(qcThresholds$Direction[i] == "below", "< ", "> "),
    sprintf("%.1f", qcThresholds$Threshold[i]),
    qcThresholds$Unit[i],
    "\n",
    sep = ""
  )
}
```

```{r quick view flag summary, echo=FALSE, eval=nrow(flagSummaryTable) > 0}
maxAffected <- max(flagSummaryTable$SamplesAffected, 0, na.rm = TRUE)
yDtick <- max(1, ceiling(maxAffected / 8))
plot_ly(
  data = flagSummaryTable,
  x = ~Flag,
  y = ~SamplesAffected,
  type = "bar",
  marker = list(color = "#4c78a8"),
  hovertemplate = paste(
    "Flag: %{x}",
    "<br>Samples affected: %{y}",
    "<extra></extra>"
  )
) %>%
  layout(
    title = "Samples affected per QC flag",
    xaxis = list(title = "", tickangle = 45),
    yaxis = list(
      title = "Samples affected",
      tickmode = "linear",
      tick0 = 0,
      dtick = yDtick,
      tickformat = "d",
      range = c(0, maxAffected + 2)
    )
  )
```

## Alignment & Coverage

Review-priority ordering is shared across plots in this tab. Dashed line marks the review threshold.

```{r alignment legacy notes, echo=FALSE, results='asis', eval=length(legacyAvailabilityNotes$alignment) > 0}
for (noteText in legacyAvailabilityNotes$alignment) {
  cat("Note:", noteText, "\n\n")
}
```

```{r alignment and coverage plots, echo=FALSE}
subplot(
  make_metric_plot(
    metricName = "MappingRate",
    title = "Mapping rate",
    unit = "%",
    showLabels = TRUE,
    thresholded = TRUE
  ),
  make_metric_plot(
    metricName = "AverageCoverage",
    title = "Average coverage",
    unit = "x",
    showLabels = FALSE,
    thresholded = TRUE
  ),
  make_metric_plot(
    metricName = "PctCov10x",
    title = "Reference covered ≥ 10x",
    unit = "%",
    showLabels = FALSE,
    thresholded = FALSE
  ),
  nrows = 1,
  shareY = TRUE,
  margin = 0.04,
  widths = c(0.34, 0.33, 0.33),
  titleX = TRUE,
  titleY = TRUE
)
```

### Library metrics

Descriptive in v1 (no QC flags). Insert-size metrics are shown for paired-end data only.

```{r library metric plots, echo=FALSE}
subplot(
  make_metric_plot(
    metricName = "MeanInsertSize",
    title = "Mean insert size",
    unit = " bp",
    showLabels = TRUE,
    thresholded = FALSE
  ),
  make_metric_plot(
    metricName = "GcContent",
    title = "GC content",
    unit = "%",
    showLabels = FALSE,
    thresholded = FALSE
  ),
  make_metric_plot(
    metricName = "MeanMapQuality",
    title = "Mean mapping quality",
    unit = "",
    showLabels = FALSE,
    thresholded = FALSE
  ),
  nrows = 1,
  shareY = TRUE,
  margin = 0.04,
  widths = c(0.34, 0.33, 0.33),
  titleX = TRUE,
  titleY = TRUE
)
```

### Multi-Matching Reported In BAM File

```{r multi-matching data, echo=FALSE}
multiMatchList <- lapply(resultList, function(x) x$multiMatchInFileTable)
hasMultiMatch <- any(!vapply(multiMatchList, is.null, logical(1)))

if (hasMultiMatch) {
  mmValues <- integer()
  for (sm in samples) {
    mm <- multiMatchList[[sm]]
    if (!is.null(mm)) {
      mmValues <- union(mmValues, as.integer(names(mm)))
    }
  }

  mmCounts <- ezMatrix(0, rows = reviewOrder, cols = sort(mmValues))
  for (sm in reviewOrder) {
    mm <- multiMatchList[[sm]]
    if (!is.null(mm)) {
      mmCounts[sm, names(mm)] <- mm
    }
  }
  hasMmCounts <- TRUE
  mmPlotWidth <- min(max(8, 8 + (nrow(mmCounts) - 20) * 0.3), 30)
}
```

```{r multi-matching plot, echo=FALSE, eval=hasMmCounts, fig.height=8, fig.width=mmPlotWidth}
alignmentCountBarPlotDna(mmCounts, relative = FALSE)
alignmentCountBarPlotDna(mmCounts, relative = TRUE)
```

```{r multi-matching fallback, echo=FALSE, results='asis', eval=!hasMmCounts}
cat("Multi-matching counts are not yet available for this dataset.")
```

## Duplicates

Review-priority ordering is shared across plots in this tab. Dashed line marks the review threshold.

```{r duplicate legacy notes, echo=FALSE, results='asis', eval=length(legacyAvailabilityNotes$duplicates) > 0}
for (noteText in legacyAvailabilityNotes$duplicates) {
  cat("Note:", noteText, "\n\n")
}
```

```{r duplicates plots, echo=FALSE}
subplot(
  make_metric_plot(
    metricName = "DuplicationRate",
    title = "Duplication rate",
    unit = "%",
    showLabels = TRUE,
    thresholded = TRUE
  ),
  make_metric_plot(
    metricName = "OpticalDupRate",
    title = "Optical duplication rate",
    unit = "%",
    showLabels = FALSE,
    thresholded = TRUE
  ),
  nrows = 1,
  shareY = TRUE,
  margin = 0.04,
  widths = c(0.5, 0.5),
  titleX = TRUE,
  titleY = TRUE
)
```

Library complexity drilldowns are available in the `Resources` tab.

## Error Profile

These metrics are descriptive in v1 and do not currently trigger QC review flags. Review-priority ordering is shared across plots in this tab.

```{r error profile row, echo=FALSE}
subplot(
  make_metric_plot(
    metricName = "ErrorRate",
    title = "Error rate",
    unit = "%",
    showLabels = TRUE,
    thresholded = FALSE
  ),
  make_metric_plot(
    metricName = "InsertionRate",
    title = "Insertion rate",
    unit = "%",
    showLabels = FALSE,
    thresholded = FALSE
  ),
  make_metric_plot(
    metricName = "DeletionRate",
    title = "Deletion rate",
    unit = "%",
    showLabels = FALSE,
    thresholded = FALSE
  ),
  nrows = 1,
  shareY = TRUE,
  margin = 0.04,
  widths = c(0.34, 0.33, 0.33),
  titleX = TRUE,
  titleY = TRUE
)
```

## Tables

### Input Dataset

```{r input dataset table, echo=FALSE}
htmltools::div(
  class = "dna-bamstats-table",
  DT::datatable(
    inputDatasetTable,
    rownames = FALSE,
    options = list(
      dom = "ftip",
      pageLength = min(max(nrow(inputDatasetTable), 5), 25),
      autoWidth = TRUE
    ),
    callback = JS(
      "var container = $(table.table().container());",
      "container.css({'width':'auto','display':'inline-block'});"
    )
  )
)
```

### Full Sample Summary

```{r full summary table, echo=FALSE}
htmltools::div(
  class = "dna-bamstats-table",
  DT::datatable(
    fullSummaryTableDisplay,
    rownames = FALSE,
    extensions = "Buttons",
    escape = FALSE,
    options = list(
      dom = "Bfrtip",
      buttons = c("excel"),
      pageLength = min(max(nrow(fullSummaryTable), 5), 25),
      order = list(list(1, "desc"), list(0, "asc")),
      autoWidth = TRUE
    ),
    callback = JS(
      "var container = $(table.table().container());",
      "container.css({'width':'auto','display':'inline-block'});"
    )
  ) %>%
    DT::formatRound(
      columns = c(
        "MappingRate",
        "DuplicationRate",
        "OpticalDupRate",
        "AverageCoverage"
      ),
      digits = 1
    )
)
```

Canonical machine-readable summary: `metrics_summary.tsv`

## Resources

Qualimap report: per-sample alignment and coverage drilldown.

Picard metrics: duplicate metrics used in the summary tables.

Fragment size plot: per-sample insert-size distribution.

Library complexity plot: per-sample duplication complexity diagnostic.

```{r resources legacy notes, echo=FALSE, results='asis', eval=length(legacyAvailabilityNotes$resources) > 0}
for (noteText in legacyAvailabilityNotes$resources) {
  cat("Note:", noteText, "\n\n")
}
```

```{r multi-sample qualimap link, echo=FALSE, results='asis'}
if (!is.null(reportData$qualimapMulti$reportFile)) {
  cat(
    '[Multi-sample Qualimap report](',
    reportData$qualimapMulti$reportFile,
    ')',
    sep = ""
  )
}
```

```{r resource table, echo=FALSE}
DT::datatable(
  resourceTableDisplay,
  escape = FALSE,
  rownames = FALSE,
  options = list(
    dom = "ftip",
    pageLength = min(max(nrow(resourceTableDisplay), 5), 25),
    autoWidth = TRUE
  )
)
```

Summary artifact: `metrics_summary.tsv`. Drilldown artifacts: Qualimap, Picard, fragment size, and library complexity outputs.

## SessionInfo

```{r session info, echo=FALSE}
ezSessionInfo()
```
