---
params:
  reportTitle: "ExploreMageckCounts"
  project: ""
  order: ""
title: "`r params$reportTitle`"
---

<!-- Styling comes from the fgczQuartoTemplate assets staged next to this file by
     ezRun::makeQuartoReport() -> fgczQuartoTemplate::fgcz_render(); no format:
     block here. Tab colour/numbering and the toolbar are makeQuartoReport args. -->

```{r setup}
#| include: false
library(ezRun)
library(SummarizedExperiment)
library(ggplot2)
library(knitr)
library(DT)
library(scales)
library(pheatmap)

## ── Data loading (qs2 with rds fallback, checks data/ first) ──
.fgcz_load <- function(name, nthreads = 4L) {
  paths <- c(file.path("data", paste0(name, ".qs2")), paste0(name, ".qs2"),
             file.path("data", paste0(name, ".rds")), paste0(name, ".rds"))
  for (p in paths) {
    if (file.exists(p)) {
      if (grepl("\\.qs2$", p)) return(qs2::qs_read(p, nthreads = nthreads))
      return(readRDS(p))
    }
  }
  stop("No .qs2 or .rds file found for: ", name)
}

## ── Compact DT helper ─────────────────────────────────────────
.dt <- function(df, caption = NULL, pageLength = 15L, filter = "none",
                rownames = FALSE, escape = TRUE, ...) {
  DT::datatable(df, caption = caption, filter = filter, rownames = rownames,
    escape = escape, class = "compact stripe",
    options = list(pageLength = pageLength, scrollX = TRUE, dom = "tip",
      columnDefs = list(list(targets = "_all", className = "dt-left"))), ...)
}

## ── Colourblind-safe palette (Okabe-Ito; report author has red-green CVD) ──
.cvd <- c("#0072B2", "#E69F00", "#009E73", "#CC79A7",
          "#56B4E9", "#D55E00", "#F0E442", "#000000")

## ── Gini (0 even, 1 skewed); x coerced to double to avoid 32-bit overflow ──
.gini <- function(x) {
  x <- as.numeric(sort(x[!is.na(x)])); n <- length(x)
  if (n == 0L || sum(x) == 0) return(NA_real_)
  sum((2 * seq_len(n) - n - 1) * x) / (n * sum(x))
}
.lorenz <- function(x) {
  x <- as.numeric(sort(x[!is.na(x)]))
  if (length(x) == 0L || sum(x) == 0) return(data.frame(p = c(0, 1), L = c(0, 1)))
  data.frame(p = c(0, seq_along(x) / length(x)), L = c(0, cumsum(x) / sum(x)))
}
## Mann-Whitney AUC: P(score of a positive > score of a negative).
.auc <- function(score, isPos) {
  isPos <- isPos[!is.na(score)]; score <- score[!is.na(score)]
  np <- sum(isPos); nn <- sum(!isPos)
  if (np == 0 || nn == 0) return(NA_real_)
  r <- rank(score)
  (sum(r[isPos]) - np * (np + 1) / 2) / (np * nn)
}

theme_set(theme_bw(base_size = 12))
```

```{r load}
#| include: false
rawData <- .fgcz_load("rawData")
param   <- metadata(rawData)$param
counts  <- assays(rawData)$counts
gene    <- rowData(rawData)$Gene
isControl <- rowData(rawData)$isControl
samples <- colnames(counts)
nSamp   <- length(samples)
conds   <- as.character(colData(rawData)$conds)
if (is.null(conds) || all(is.na(conds))) conds <- rep("all", nSamp)
bg      <- suppressWarnings(as.numeric(param$backgroundExpression)); if (is.na(bg)) bg <- 5
normMethod <- if (!is.null(param$normMethod)) tolower(param$normMethod) else "deseq2"
topN    <- suppressWarnings(as.integer(param$topGeneSize)); if (is.na(topN)) topN <- 100L

## ── Normalisation (size factors = the median-of-ratios scheme MAGeCK uses) ──
normFactors <- tryCatch({
  if (normMethod %in% c("tmm", "cpm")) {
    dge <- edgeR::DGEList(counts = counts)
    dge <- edgeR::calcNormFactors(dge, method = if (normMethod == "tmm") "TMM" else "none")
    (dge$samples$lib.size * dge$samples$norm.factors) / 1e6   # per-million scaling
  } else if (normMethod == "logmean") {
    colSums(counts) / mean(colSums(counts))
  } else {
    DESeq2::estimateSizeFactorsForMatrix(counts)              # deseq2 (default)
  }
}, error = function(e) colSums(counts) / mean(colSums(counts)))
normFactors[!is.finite(normFactors) | normFactors <= 0] <- 1
normCounts <- sweep(counts, 2, normFactors, "/")
logNorm    <- log2(normCounts + bg)

## Variance-stabilised matrix for ordination; fall back to logNorm if vst fails.
mat <- tryCatch({
  if (normMethod == "deseq2" && nSamp >= 2) {
    dds <- DESeq2::DESeqDataSetFromMatrix(round(counts), DataFrame(row.names = samples), ~1)
    SummarizedExperiment::assay(DESeq2::vst(dds, blind = TRUE))
  } else logNorm
}, error = function(e) logNorm)

## ── Essential / non-essential reference gene sets (Hart CEGv2 / NEGv1) ──
.readSet <- function(f) {
  p <- system.file(file.path("extdata/crispr", f), package = "ezRun")
  if (!nzchar(p) || !file.exists(p)) return(character(0))
  d <- try(read.delim(p, stringsAsFactors = FALSE), silent = TRUE)
  if (inherits(d, "try-error") || !nrow(d)) return(character(0))
  toupper(trimws(d[[1]]))
}
essSet <- .readSet("CEGv2.txt")
negSet <- .readSet("NEGv1.txt")
geneU  <- toupper(trimws(gene))
geneClass <- ifelse(isControl, "control",
              ifelse(geneU %in% essSet, "essential",
              ifelse(geneU %in% negSet, "non-essential", "other")))

## ── Reference condition (explicit param$referenceCondition) for depletion ──
## Matched against the condition labels first, then any sample-annotation column
## value, so the user can name the reference as it appears in their design.
refGroup <- if (!is.null(param$referenceCondition)) trimws(as.character(param$referenceCondition)) else ""
refGroupSpecified <- nzchar(refGroup)
refIdx <- integer(0)
if (refGroupSpecified) {
  refIdx <- which(conds == refGroup)
  if (length(refIdx) == 0L) {
    cd <- as.data.frame(colData(rawData))
    hit <- apply(cd, 1, function(r) any(as.character(r) == refGroup))
    refIdx <- which(hit)
  }
}
hasRef <- length(refIdx) >= 1 && length(refIdx) < nSamp

## Per-sample library-representation table (targeting sgRNAs only).
repTab <- do.call(rbind, lapply(seq_len(nSamp), function(j) {
  cj  <- counts[!isControl, j]
  tot <- sum(counts[, j])
  data.frame(
    Sample = samples[j],
    Condition = conds[j],
    `Total reads` = tot,
    `Targeting sgRNAs` = length(cj),
    `Detected (%)` = 100 * mean(cj > 0),
    `Zero-count (%)` = 100 * mean(cj == 0),
    `Median reads/sgRNA` = stats::median(cj),
    `Gini` = .gini(cj),
    check.names = FALSE)
}))
```

```{r provenance}
#| include: false
`%||%` <- function(a, b) if (is.null(a) || !length(a) || !nzchar(as.character(a)[1])) b else a
.fgcz_rendered <- format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")
.fgcz_user <- unname(Sys.info()[["user"]]); if (!nzchar(.fgcz_user %||% "")) .fgcz_user <- Sys.getenv("USER", "unknown")
```

::: {.panel-tabset}

# Overview

::: {.callout-note title="What this report is"}
An exploratory QC overview of a pooled **CRISPR screen** after sgRNA counting
(**MAGeCK count**), *before* differential hit-calling. It answers three questions:

1. **Is the library well represented?** — evenness of guide counts (Gini, Lorenz),
   zero-count guides, reads per guide.
2. **Do the samples group as expected?** — sample correlation, clustering and PCA
   on normalised counts (needs ≥ 3 samples for clustering/PCA).
3. **Does the screen behave biologically?** — where known **core-essential**
   (Hart CEGv2) and **non-essential** (NEGv1) genes and the library's control
   guides fall, and (with a reference/T0 sample) whether essential genes deplete.
:::

```{r overview-summary}
#| echo: false
info <- data.frame(
  Item = c("Samples", "sgRNAs (guides)", "Genes", "Control guides",
           "Essential genes (CEGv2) present", "Non-essential (NEGv1) present",
           "Normalisation", "Reference condition"),
  Value = c(
    nSamp, nrow(counts), length(unique(gene[!isControl])), sum(isControl),
    length(intersect(geneU[!isControl], essSet)),
    length(intersect(geneU[!isControl], negSet)),
    normMethod,
    if (hasRef) sprintf("%s (%s)", refGroup, paste(samples[refIdx], collapse = ", "))
    else if (refGroupSpecified) sprintf("'%s' (not matched)", refGroup)
    else "not set"),
  check.names = FALSE)
knitr::kable(info)
```

```{r few-samples-callout}
#| echo: false
#| results: asis
if (nSamp < 3) {
  cat("::: {.callout-warning}\n",
      sprintf("Only **%d** sample(s). Sample clustering and PCA need **≥ 3** samples; ", nSamp),
      "those panels are skipped and only library representation",
      if (nSamp == 2) " and pairwise correlation" else "",
      " are shown.\n:::\n", sep = "")
}
```

# Library representation

::: {.callout-note}
Even guide representation is a core quality signal: a low **Gini** (≈ 0 even, 1 skewed),
few zero-count guides and adequate reads/guide indicate the library was not bottlenecked.
Metrics below are for **targeting** guides (controls excluded).
:::

```{r rep-table}
#| echo: false
disp <- repTab
disp$`Total reads` <- format(round(disp$`Total reads`), big.mark = ",")
disp$`Detected (%)` <- sprintf("%.1f", disp$`Detected (%)`)
disp$`Zero-count (%)` <- sprintf("%.1f", disp$`Zero-count (%)`)
disp$Gini <- sprintf("%.3f", disp$Gini)
.dt(disp, caption = "Per-sample library representation (targeting sgRNAs).")
```

```{r lorenz}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
#| out-width: "70%"
#| fig-cap: "Lorenz curves of guide counts per sample. The diagonal is perfectly even representation; more sag = more skew (higher Gini)."
lor <- do.call(rbind, lapply(seq_len(nSamp), function(j) {
  d <- .lorenz(counts[!isControl, j]); d$Sample <- samples[j]; d
}))
ggplot(lor, aes(p, L, colour = Sample)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey60") +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = rep(.cvd, length.out = nSamp)) +
  labs(x = "Cumulative fraction of sgRNAs (low → high count)",
       y = "Cumulative fraction of reads")
```

# Controls & essential genes

::: {.callout-note}
Biological sanity checks. **Non-targeting/control** guides should behave neutrally.
Known **core-essential** genes (positive controls for a viability screen) should be
depleted relative to **non-essential** genes when a reference/T0 sample is present.
:::

```{r class-dist}
#| echo: false
#| fig-width: 8
#| fig-height: 4.5
#| out-width: "85%"
#| fig-cap: "Distribution of log2 normalised counts by gene class, per sample."
long <- do.call(rbind, lapply(seq_len(nSamp), function(j)
  data.frame(Sample = samples[j], class = geneClass, value = logNorm[, j])))
long <- long[long$class != "other", ]
long$class <- factor(long$class, levels = c("control", "non-essential", "essential"))
ggplot(long, aes(class, value, fill = class)) +
  geom_violin(scale = "width", alpha = 0.85) +
  geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white") +
  facet_wrap(~ Sample) +
  scale_fill_manual(values = c(control = .cvd[8], `non-essential` = .cvd[1], essential = .cvd[6]),
                    guide = "none") +
  labs(x = NULL, y = "log2 normalised count") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r essential-depletion}
#| echo: false
#| results: asis
if (!hasRef) {
  msg <- if (!refGroupSpecified) {
    "No **reference condition** was set, so essential-gene **depletion** and the ROC are not computed. Set the `referenceCondition` parameter to the reference/plasmid/T0 condition to enable them."
  } else if (length(refIdx) == 0L) {
    sprintf("The specified **reference condition** '%s' matched no sample, so depletion/ROC are skipped. Available conditions: %s.",
            refGroup, paste(unique(conds), collapse = ", "))
  } else {
    sprintf("The specified **reference condition** '%s' matched *all* samples; non-reference samples are needed too, so depletion/ROC are skipped.", refGroup)
  }
  cat(sprintf("::: {.callout-note}\n%s\n:::\n", msg))
}
```

```{r essential-roc}
#| echo: false
#| eval: !expr "hasRef"
#| fig-width: 5.5
#| fig-height: 5
#| out-width: "55%"
#| fig-cap: "Separation of core-essential from non-essential genes by depletion vs the reference sample. A screen that worked pushes essentials to negative LFC (AUC → 1)."
refMean <- rowMeans(logNorm[, refIdx, drop = FALSE])
smpMean <- rowMeans(logNorm[, setdiff(seq_len(nSamp), refIdx), drop = FALSE])
lfc <- smpMean - refMean                      # depletion is negative
geneLFC <- tapply(lfc[!isControl], gene[!isControl], mean)
gU <- toupper(trimws(names(geneLFC)))
sel <- gU %in% essSet | gU %in% negSet
score <- -geneLFC[sel]                          # higher score = more depleted
isEss <- gU[sel] %in% essSet
auc <- .auc(score, isEss)
roc <- data.frame(lfc = geneLFC[sel], ess = ifelse(isEss, "essential", "non-essential"))
ggplot(roc, aes(lfc, fill = ess)) +
  geom_density(alpha = 0.6) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  scale_fill_manual(values = c(essential = .cvd[6], `non-essential` = .cvd[1]), name = NULL) +
  labs(x = "gene mean log2 fold change vs reference",
       title = sprintf("Essential-gene depletion (AUC = %.3f)", auc))
```

# Sample correlation

```{r correlation}
#| echo: false
#| eval: !expr "nSamp >= 2"
#| fig-width: !expr "max(6, min(16, 3.5 + 0.55 * nSamp))"
#| fig-height: !expr "max(5, min(15, 3 + 0.55 * nSamp))"
#| out-width: !expr "if (nSamp <= 8) '60%' else '90%'"
#| fig-cap: "Pearson correlation of samples on normalised counts (all targeting sgRNAs)."
cm <- cor(mat[!isControl, ], use = "complete.obs")
ann <- data.frame(Condition = conds, row.names = samples)
## Fixed cell size (points) keeps the grid readable when there are few samples,
## so labels/title no longer dominate a tiny plotting area.
cellSz <- if (nSamp <= 20) 32 else NA
pheatmap(cm, annotation_col = ann,
         display_numbers = nSamp <= 15, number_format = "%.2f",
         number_color = "grey15", fontsize = 9, fontsize_number = 8,
         cellwidth = cellSz, cellheight = cellSz, border_color = "white",
         color = colorRampPalette(c("#f7fbff", .cvd[1]))(50),
         main = "Sample correlation (Pearson)")
```

```{r correlation-skip}
#| echo: false
#| results: asis
if (nSamp < 2) cat("Correlation needs at least 2 samples.\n")
```

# Clustering & PCA

```{r clustering}
#| echo: false
#| eval: !expr "nSamp >= 3"
#| fig-width: !expr "max(6, min(14, 4 + 0.4 * nSamp))"
#| fig-height: 5
#| out-width: "80%"
#| fig-cap: "Hierarchical clustering of samples (Ward D2, 1 − Pearson on normalised counts)."
d  <- as.dist(1 - cor(mat[!isControl, ], use = "complete.obs"))
hc <- hclust(d, method = "ward.D2")
op <- par(mar = c(7, 4, 3, 1)); plot(hc, main = "Sample clustering", xlab = "", sub = ""); par(op)
```

```{r pca}
#| echo: false
#| eval: !expr "nSamp >= 3"
#| fig-width: 6.5
#| fig-height: 5
#| out-width: "65%"
#| fig-cap: "PCA on the most variable sgRNAs (variance-stabilised counts)."
v <- matrixStats::rowVars(mat[!isControl, ])
sel <- head(order(v, decreasing = TRUE), min(2000L, sum(!isControl)))
pc <- prcomp(t(mat[!isControl, ][sel, ]), scale. = FALSE)
ve <- 100 * pc$sdev^2 / sum(pc$sdev^2)
pcd <- data.frame(PC1 = pc$x[, 1], PC2 = pc$x[, 2], Sample = samples, Condition = conds)
ggplot(pcd, aes(PC1, PC2, colour = Condition, label = Sample)) +
  geom_point(size = 3) + ggrepel::geom_text_repel(size = 3, show.legend = FALSE) +
  scale_colour_manual(values = rep(.cvd, length.out = length(unique(conds)))) +
  labs(x = sprintf("PC1 (%.1f%%)", ve[1]), y = sprintf("PC2 (%.1f%%)", ve[2]))
```

```{r highvar-heatmap}
#| echo: false
#| eval: !expr "nSamp >= 3"
#| fig-width: !expr "max(6, 3 + 0.4 * nSamp)"
#| fig-height: 8
#| out-width: "70%"
#| fig-cap: "Top variable sgRNAs (row-centred normalised counts). Samples as columns."
v <- matrixStats::rowVars(logNorm[!isControl, ])
sel <- head(order(v, decreasing = TRUE), min(topN, sum(!isControl)))
hm <- logNorm[!isControl, ][sel, ]
hm <- hm - rowMeans(hm)
nClust <- suppressWarnings(as.integer(param$nSampleClusters)); if (is.na(nClust)) nClust <- 6L
pheatmap(hm, annotation_col = data.frame(Condition = conds, row.names = samples),
         show_rownames = FALSE, scale = "none",
         color = colorRampPalette(c(.cvd[1], "white", .cvd[6]))(50),
         main = sprintf("Top %d variable sgRNAs", nrow(hm)), silent = FALSE)
```

```{r clustering-skip}
#| echo: false
#| results: asis
if (nSamp < 3) cat("Clustering and PCA need at least 3 samples — skipped.\n")
```

# Gene level

::: {.callout-note}
Guide counts aggregated to genes (sum of targeting guides per gene). Controls are excluded.
:::

```{r gene-level}
#| echo: false
keep <- !isControl
geneCounts <- rowsum(counts[keep, , drop = FALSE], group = gene[keep])
geneNorm   <- sweep(geneCounts, 2, normFactors, "/")
geneLog    <- log2(geneNorm + bg)
gRep <- data.frame(
  Sample = samples, Condition = conds,
  `Genes` = nrow(geneCounts),
  `Detected (%)` = sprintf("%.1f", 100 * colMeans(geneCounts > 0)),
  `Gini` = sprintf("%.3f", apply(geneCounts, 2, .gini)),
  check.names = FALSE)
.dt(gRep, caption = "Per-sample gene-level representation.")
```

```{r gene-lorenz}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
#| out-width: "70%"
#| fig-cap: "Gene-level Lorenz curves (summed guide counts per gene)."
glor <- do.call(rbind, lapply(seq_len(nSamp), function(j) {
  d <- .lorenz(geneCounts[, j]); d$Sample <- samples[j]; d
}))
ggplot(glor, aes(p, L, colour = Sample)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey60") +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = rep(.cvd, length.out = nSamp)) +
  labs(x = "Cumulative fraction of genes", y = "Cumulative fraction of reads")
```

```{r gene-clustering}
#| echo: false
#| eval: !expr "nSamp >= 3"
#| fig-width: !expr "max(6, min(14, 4 + 0.4 * nSamp))"
#| fig-height: 5
#| out-width: "80%"
#| fig-cap: "Gene-level sample clustering (Ward D2, 1 − Pearson)."
d  <- as.dist(1 - cor(geneLog, use = "complete.obs"))
hc <- hclust(d, method = "ward.D2")
op <- par(mar = c(7, 4, 3, 1)); plot(hc, main = "Gene-level sample clustering", xlab = "", sub = ""); par(op)
```

# Data & Session

```{r write-data}
#| include: false
outDir <- "."
sgFile <- "sgRNA_counts.xlsx"; geFile <- "gene_counts.xlsx"
sgOut <- data.frame(sgRNA = rownames(counts), Gene = gene, isControl = isControl,
                    counts, check.names = FALSE)
geOut <- data.frame(Gene = rownames(geneCounts), geneCounts, check.names = FALSE)
writexl::write_xlsx(sgOut, file.path(outDir, sgFile))
writexl::write_xlsx(geOut, file.path(outDir, geFile))
```

Raw count tables: [`r sgFile`](`r sgFile`){target="_blank"} (sgRNA level) ·
[`r geFile`](`r geFile`){target="_blank"} (gene level).

**Rendered:** `r .fgcz_rendered` by `r .fgcz_user`.
`r if (nzchar(params$project)) paste0("Project: ", params$project, ".")` `r if (nzchar(params$order)) paste0("Order: ", params$order, ".")`

```{r session}
#| echo: false
ezSessionInfo()
```

:::
