---
title: "`r paste('FASTQC:', param$name)`"
author: "Functional Genomics Center Zurich"
output: 
  html_document:
    self_contained: false
    includes:
      in_header: fgcz_header.html
    css: fgcz.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

## FastQC_Result {.tabset}

### Read Counts
The read counts in each sample.

```{r readCounts, echo=FALSE}
readCount <- ans4Report[["Read Counts"]]
par(mar=c(10.1, 4.1, 4.1, 2.1))
bp = barplot(readCount, ylab="Counts [Mio]", main="total reads", 
             names.arg = rep('',length(readCount)))
text(x = bp, y = par("usr")[3]-0.2, srt = 45, adj = 1, 
     labels = names(readCount), xpd = TRUE)
```

### Fastqc Quality Measures
The summary of [FastQC](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/) reports.

```{r quality, echo=FALSE, results='asis'}
tbl <- ans4Report[["Fastqc quality measures"]]
knitr::kable(tbl, col.names=colnames(tbl), row.names=TRUE)
```

### Per Base Read Quality
The black and white heatmaps show the average reads quality  
over all samples per base.
The color scale corresponds to the percentage of reads at each reads quality.

```{r avgQualColors, fig.width=16, fig.height=2, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
qualMatrixList <- ans4Report[["Per Base Read Quality"]]
plotList <- plotQualityMatrixAsHeatmapGG2(qualMatrixList, 
                                          isR2=grepl("_R2", names(qualMatrixList)))
isPaired <- any(grepl("_R2", names(qualMatrixList)))
# To plot in the order of R1, R2 in row
library(grid)
library(gridExtra)
## This is quite stupid.
if(isPaired){
  grid.arrange(plotList[["R1"]][["Avg Qual Colors"]],
               plotList[["R2"]][["Avg Qual Colors"]], ncol = 2)
}else{
  grid.arrange(plotList[["R1"]][["Avg Qual Colors"]], ncol = 2)
}
```

```{r average, fig.width=16, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
if(isPaired){
  grid.arrange(plotList[["R1"]][["Average"]],
               plotList[["R2"]][["Average"]], ncol = 2)
}else{
  grid.arrange(plotList[["R1"]][["Average"]], ncol = 2)
}
```

The blue and read heatmaps show the difference between each sample's reads quality
and the average reads quality per base.
The red represents a higher percentage of reads over the average, 
while the blue represents a lower percentage of reads over the average.

```{r diffQualColors, fig.width=16, fig.height=2, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
if(isPaired){
  grid.arrange(plotList[["R1"]][["Diff Qual Colors"]],
               plotList[["R2"]][["Diff Qual Colors"]], ncol = 2)
}else{
  grid.arrange(plotList[["R1"]][["Diff Qual Colors"]], ncol = 2)
}
```

```{r sampleDiffQualColors, fig.width=16, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
sampleIndices <- which(!(names(plotList[["R1"]]) %in%
                       c("Avg Qual Colors", "Average", "Diff Qual Colors")))
for(sampleIndex in sampleIndices){
  if(isPaired){
    grid.arrange(plotList[["R1"]][[sampleIndex]],
                 plotList[["R2"]][[sampleIndex]], ncol = 2)
  }else{
    grid.arrange(plotList[["R1"]][[sampleIndex]], ncol = 2)
  }
}
```

### Correlation
Correlation between Library concentration measurements and ReadCounts.
It's only shown when "LibConc_qPCR" or "LibConc_100_800bp" columns exit in `dataset`.

```{r correlation, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
dataset <- ans4Report[["dataset"]]
if(nrow(dataset) > 1){
  plotReadCountToLibConcRmd(dataset, 
                            colname='LibConc_qPCR [Characteristic]')
  plotReadCountToLibConcRmd(dataset, 
                            colname='LibConc_100_800bp [Characteristic]')
}
```

### Input Dataset
```{r, echo=FALSE}
ezInteractiveTableRmd(values=ans4Report[["dataset"]])
```

### SessionInfo
```{r, echo=FALSE}
sessionInfo()
```

