---
title: "FastQC AI Interpretation"
output:
  html_document:
    mathjax: https://fgcz-gstore.uzh.ch/reference/mathjax.js
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package="ezRun")
    css: !expr system.file("templates/fgcz.css", package="ezRun")
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

rds_candidates <- c(
  "ai_interpretation.rds",
  file.path("..", "multi_FastQC", "ai_interpretation.rds")
)
rds_path <- rds_candidates[file.exists(rds_candidates)][1]
stopifnot(!is.na(rds_path) && file.exists(rds_path))
ai <- readRDS(rds_path)

redact_endpoint <- function(url) sub("^(https?://[^/]+).*", "\\1", as.character(url))

style_directives <- function(txt) {
  if (is.null(txt) || !nzchar(txt)) return("")
  for (sev in c("red", "orange", "yellow", "green")) {
    pat_span   <- sprintf(":span\\[([^]]*)\\]\\{\\.text-%s\\}", sev)
    pat_sample <- sprintf(":sample\\[([^]]*)\\]\\{\\.text-%s\\}", sev)
    txt <- gsub(pat_span,   sprintf("<span class=\"ai-severity-%s\">\\1</span>", sev), txt, perl = TRUE)
    txt <- gsub(pat_sample, sprintf("<span class=\"ai-sample ai-severity-%s\">\\1</span>", sev), txt, perl = TRUE)
  }
  txt <- gsub(":span\\[([^]]*)\\]\\{[^}]*\\}",   "\\1", txt, perl = TRUE)
  txt <- gsub(":sample\\[([^]]*)\\]\\{[^}]*\\}", "<span class=\"ai-sample\">\\1</span>", txt, perl = TRUE)
  txt
}

prettify_section <- function(name) {
  name <- sub("^fastqc[-_]?", "", name)
  name <- sub("[-_]plot$|[-_]table$|[-_]heatmap$", "", name)
  gsub("[_-]", " ", name)
}
```

<style>
.ai-banner { padding: 10px 14px; margin: 10px 0 20px 0; background: #eef4fb; border-left: 4px solid #4a90d9; font-family: monospace; font-size: 0.95em; }
.ai-placeholder { padding: 12px 16px; margin: 10px 0; background: #fafafa; border-left: 4px solid #ccc; color: #555; font-style: italic; }
.ai-section { padding: 8px 12px; margin: 12px 0; border-left: 3px solid #4a90d9; }
.ai-section h4 { margin: 0 0 6px 0; }
.ai-section .ai-dur { color: #888; font-size: 0.85em; font-weight: normal; }
.ai-severity-red    { color: #c0392b; font-weight: 600; }
.ai-severity-orange { color: #d35400; font-weight: 600; }
.ai-severity-yellow { color: #b7950b; font-weight: 600; }
.ai-severity-green  { color: #1e8449; font-weight: 600; }
.ai-sample { font-weight: 700; font-style: italic; }
pre.ai-raw { white-space: pre-wrap; word-break: break-word; max-height: 70vh; overflow-y: auto; }
</style>

```{r banner, results='asis'}
lines <- c(
  sprintf("MultiQC (incl. its inline global AI call): <b>%.1f s</b>", ai$mqc_dur_s)
)
if (isTRUE(ai$flags$bake_sections)) {
  lines <- c(lines, sprintf("Per-section AI calls (N=%d): <b>%.1f s</b> total",
                            length(ai$section_summaries), ai$sec_total_dur_s))
}
if (isTRUE(ai$flags$extra_call)) {
  lines <- c(lines, sprintf("FGCZ Follow-up AI call: <b>%.1f s</b>", ai$extra_dur_s))
}
lines <- c(lines, sprintf("Endpoint: <code>%s</code>, model: <code>%s</code>",
                          redact_endpoint(ai$endpoint), htmltools::htmlEscape(ai$model)))
cat('<div class="ai-banner">', paste(lines, collapse = "<br/>"), '</div>', sep = "")
```

# {.tabset}

## MultiQC Global Summary

```{r mqc-global, results='asis'}
have_global <- nzchar(ai$mqc_global_summary_html) || nzchar(ai$mqc_detailed_html)
if (have_global) {
  cat(ai$mqc_global_summary_html, "\n", ai$mqc_detailed_html, sep = "")
} else {
  cat('<div class="ai-placeholder">MultiQC did not populate the global summary divs in <code>multiqc_report.html</code>. Open that report directly to see whether the AI summary is present.</div>')
}
```

## Per-Section Summaries

```{r per-section, results='asis'}
if (isTRUE(ai$flags$bake_sections) && length(ai$section_summaries) > 0) {
  for (sec_name in names(ai$section_summaries)) {
    s <- ai$section_summaries[[sec_name]]
    cat(sprintf('<div class="ai-section"><h4>%s <span class="ai-dur">(%.1f s)</span></h4>',
                htmltools::htmlEscape(prettify_section(sec_name)), s$dur_s))
    cat(style_directives(s$response_text))
    cat('</div>')
  }
} else {
  cat('<div class="ai-placeholder">Per-section summaries were not generated. Tick the <b>Bake per-section AI summaries</b> checkbox in the SUSHI app to enable this tab. Alternatively, open <code>multiqc_report.html</code> and click each "Summarize plot" / "Summarize table" button in the MultiQC toolbox.</div>')
}
```

## FGCZ Follow-up

```{r follow-up, results='asis'}
if (isTRUE(ai$flags$extra_call) && !is.null(ai$extra_response_text)) {
  cat('<details><summary>User\'s extra prompt</summary><pre class="ai-raw">',
      htmltools::htmlEscape(ai$extra_prompt), '</pre></details>', sep = "")
  cat(style_directives(ai$extra_response_text))
} else {
  cat('<div class="ai-placeholder">No FGCZ follow-up call was made. Tick <b>Extra AI call</b> and fill <b>Extra AI prompt</b> in the SUSHI app to enable this tab.</div>')
}
```

## Raw Prompt

```{r raw-prompt, results='asis'}
if (nzchar(ai$prompt)) {
  cat('<pre class="ai-raw">', htmltools::htmlEscape(ai$prompt), '</pre>', sep = "")
} else {
  cat('<div class="ai-placeholder"><code>multiqc_data/llms-full.txt</code> was not found.</div>')
}
```

## Raw Follow-up Response

```{r raw-followup, results='asis'}
if (isTRUE(ai$flags$extra_call) && !is.null(ai$extra_raw_body)) {
  cat('<pre class="ai-raw">', htmltools::htmlEscape(ai$extra_raw_body), '</pre>', sep = "")
} else {
  cat('<div class="ai-placeholder">No follow-up response to display. Enable <b>Extra AI call</b> in the SUSHI app to populate this tab.</div>')
}
```
