---
title: "HUMAnN 4 Functional Profiling Report"
params:
  sample: "Sample"
  mode: "translated"
  profileSource: "MetaPhlAnProfile"
  detectedDb: ""
  forced: FALSE
  humannLog: ""
  gfCpmFile: ""
  rxCpmFile: ""
  paFile: ""
  koFile: NA
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 6, dev = "png")
library(ggplot2)
```

## Sample: `r params$sample`

```{r mode_box, results = "asis"}
mode_label <- if (identical(params$mode, "full")) {
  "**Full HUMAnN** — tier-1 (ChocoPhlAn pangenomes, SGB-stratified) + tier-2 (UniRef90-EC DIAMOND)."
} else {
  "**Translated-only HUMAnN** — tier-2 only (UniRef90-EC DIAMOND), community-level output."
}
cat("- Mode: ", mode_label, "\n", sep = "")
cat("- Upstream profile source: `", params$profileSource, "`\n", sep = "")
cat("- Detected DB tag: `", params$detectedDb, "`\n", sep = "")
if (isTRUE(params$forced)) {
  cat("- **Note:** `forceTranslatedOnly` parameter was TRUE — any vOct22 input was downgraded.\n")
}
```

## Run summary

```{r log_summary, results = "asis"}
log_path <- params$humannLog
if (nzchar(log_path) && file.exists(log_path)) {
  ln <- readLines(log_path, warn = FALSE)
  pick <- function(pat) {
    m <- grep(pat, ln, value = TRUE)
    if (length(m) == 0) return("(not found in log)")
    sub("^[^:]*:\\s*", "", tail(m, 1))
  }
  rows <- list(
    c("Total reads (input)",                          pick("Total reads")),
    c("Reads aligned (nucleotide tier-1, full only)", pick("Unaligned reads after nucleotide alignment")),
    c("Reads aligned (after translated tier-2)",      pick("Unaligned reads after translated alignment")),
    c("Total gene families",                          pick("^Total gene families"))
  )
  tbl <- do.call(rbind, lapply(rows, function(x) data.frame(Metric = x[1], Value = x[2])))
  knitr::kable(tbl, row.names = FALSE)
} else {
  cat("HUMAnN log not found at `", log_path, "`\n", sep = "")
}
```

## Top features

```{r helpers, include = FALSE}
read_humann <- function(path) {
  if (is.na(path) || !nzchar(path) || !file.exists(path)) return(NULL)
  m <- read.table(path, header = TRUE, sep = "\t", check.names = FALSE,
                  comment.char = "", quote = "", row.names = 1)
  colnames(m) <- sub("_(Abundance|Abundance-CPM|Coverage|Abundance-RPKs).*$", "",
                    colnames(m))
  m <- m[!rownames(m) %in% c("UNMAPPED", "UNINTEGRATED", "UNGROUPED",
                             "READS_UNMAPPED"), , drop = FALSE]
  m <- m[!grepl("\\|", rownames(m)), , drop = FALSE]   # community-level only
  m
}

top_n_table <- function(m, n = 20) {
  if (is.null(m) || nrow(m) == 0) return(NULL)
  v <- m[, 1]
  ord <- order(v, decreasing = TRUE)
  head(data.frame(Feature = rownames(m)[ord], Value = v[ord],
                  stringsAsFactors = FALSE), n)
}

bar_top <- function(m, title, n = 20) {
  d <- top_n_table(m, n = n)
  if (is.null(d)) return(NULL)
  d$Feature <- factor(d$Feature, levels = rev(d$Feature))
  ggplot(d, aes(x = Feature, y = Value)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    labs(title = title, x = NULL, y = "CPM") +
    theme_minimal(base_size = 11) +
    theme(axis.text.y = element_text(size = 8))
}

pa <- read_humann(params$paFile)
rx <- read_humann(params$rxCpmFile)
ko <- read_humann(params$koFile)
gf <- read_humann(params$gfCpmFile)
```

### MetaCyc pathways

`r if (is.null(pa)) "(no pathway abundance table)" else sprintf("%d community-level pathways detected.", nrow(pa))`

```{r pathway_plot, fig.height = 6}
if (!is.null(pa)) print(bar_top(pa, "Top 20 MetaCyc pathways"))
```

```{r pathway_table}
if (!is.null(pa)) knitr::kable(top_n_table(pa, 20), row.names = FALSE,
                                col.names = c("Pathway", "Abundance"))
```

### MetaCyc reactions

`r if (is.null(rx)) "(no reactions table)" else sprintf("%d community-level reactions detected.", nrow(rx))`

```{r reactions_plot, fig.height = 6}
if (!is.null(rx)) print(bar_top(rx, "Top 20 MetaCyc reactions"))
```

### KEGG KO

`r if (is.null(ko)) "(KEGG KO regroup not run or failed; HUMAnN 4 alpha utility_mapping covers EC-annotated KOs only.)" else sprintf("%d community-level KOs detected (EC-mapped subset only).", nrow(ko))`

```{r ko_plot, fig.height = 6}
if (!is.null(ko)) print(bar_top(ko, "Top 20 KEGG KOs"))
```

### Gene families (community level)

`r if (is.null(gf)) "(no gene families table)" else sprintf("%d community-level UniRef90-EC + UniClust90 gene families detected.", nrow(gf))`

```{r gf_plot, fig.height = 6}
if (!is.null(gf)) print(bar_top(gf, "Top 20 gene families"))
```

## Output files

```{r files_table}
fs <- c(params$gfCpmFile, params$rxCpmFile, params$paFile, params$koFile)
fs <- fs[!is.na(fs) & nzchar(fs)]
fs <- fs[file.exists(fs)]
if (length(fs) > 0) {
  info <- file.info(fs)
  knitr::kable(data.frame(File = basename(fs),
                          Size_KB = round(info$size / 1024, 1)),
               row.names = FALSE)
}
```

## Method notes

HUMAnN **v4.0.0.alpha.2** with these caveats:

- **DB compatibility.** ChocoPhlAn pangenomes are pinned to `mpa_vOct22_CHOCOPhlAnSGB_202403`. MetaPhlAn profiles from newer DBs (e.g. `mpa_vJan25_CHOCOPhlAnSGB_202503`) can only drive translated-only mode — SGB IDs don't align across vintages.
- **Profile is always supplied externally.** MetaPhlAn ≥ 4.2 has an API change HUMAnN 4 alpha cannot drive itself; this app passes `--taxonomic-profile` so the MetaPhlAn binary version is irrelevant.
- **EC-filtered UniRef DB only.** Gene-family quantitation is restricted to enzymes. MetaCyc pathways work as designed; KEGG KO regroup covers the EC-mapped subset only (no transporters / regulators / structural).
- **`READS_UNMAPPED` strip.** HUMAnN 4 renamed `UNMAPPED` → `READS_UNMAPPED` in gene families; `humann_renorm_table --special n` doesn't know the new name. We strip it before CPM-renorm so it doesn't dominate the denominator.

See [biobakery forum #8523](https://forum.biobakery.org/t/metaphlan-4-humann-4-compatibility/8523) for compatibility background.
