---
title: "HUMAnN 4 Functional Profiling Report"
params:
  sample: "Sample"
  mode: "translated"
  profileSource: "MetaPhlAnProfile"
  detectedDb: ""
  forced: FALSE
  humannLog: ""
  gfCpmFile: ""
  rxCpmFile: ""
  paFile: ""
  koFile: NA
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 6, dev = "png")
library(ggplot2)
```

## Sample: `r params$sample`

```{r mode_box, results = "asis"}
mode_label <- if (identical(params$mode, "full")) {
  "**Full HUMAnN** — tier-1 (ChocoPhlAn pangenomes, SGB-stratified) + tier-2 (UniRef90-EC DIAMOND)."
} else {
  "**Translated-only HUMAnN** — tier-2 only (UniRef90-EC DIAMOND), community-level output."
}
cat("- Mode: ", mode_label, "\n", sep = "")
cat("- Upstream profile source: `", params$profileSource, "`\n", sep = "")
cat("- Detected DB tag: `", params$detectedDb, "`\n", sep = "")
if (isTRUE(params$forced)) {
  cat("- **Note:** `forceTranslatedOnly` parameter was TRUE — any vOct22 input was downgraded.\n")
}
```

## Run summary

```{r log_summary, results = "asis"}
log_path <- params$humannLog
if (nzchar(log_path) && file.exists(log_path)) {
  ln <- readLines(log_path, warn = FALSE)
  pick <- function(pat) {
    m <- grep(pat, ln, value = TRUE)
    if (length(m) == 0) return("(not found in log)")
    sub("^[^:]*:\\s*", "", tail(m, 1))
  }
  rows <- list(
    c("Total reads (input)",                          pick("Total reads")),
    c("Reads aligned (nucleotide tier-1, full only)", pick("Unaligned reads after nucleotide alignment")),
    c("Reads aligned (after translated tier-2)",      pick("Unaligned reads after translated alignment")),
    c("Total gene families",                          pick("^Total gene families"))
  )
  tbl <- do.call(rbind, lapply(rows, function(x) data.frame(Metric = x[1], Value = x[2])))
  knitr::kable(tbl, row.names = FALSE)
} else {
  cat("HUMAnN log not found at `", log_path, "`\n", sep = "")
}
```

## Top features

```{r helpers, include = FALSE}
read_humann <- function(path) {
  if (is.na(path) || !nzchar(path) || !file.exists(path)) return(NULL)
  m <- read.table(path, header = TRUE, sep = "\t", check.names = FALSE,
                  comment.char = "", quote = "", row.names = 1)
  colnames(m) <- sub("_(Abundance|Abundance-CPM|Coverage|Abundance-RPKs).*$", "",
                    colnames(m))
  m <- m[!rownames(m) %in% c("UNMAPPED", "UNINTEGRATED", "UNGROUPED",
                             "READS_UNMAPPED"), , drop = FALSE]
  m <- m[!grepl("\\|", rownames(m)), , drop = FALSE]   # community-level only
  m
}

top_n_table <- function(m, n = 20) {
  if (is.null(m) || nrow(m) == 0) return(NULL)
  v <- m[, 1]
  ord <- order(v, decreasing = TRUE)
  head(data.frame(Feature = rownames(m)[ord], Value = v[ord],
                  stringsAsFactors = FALSE), n)
}

bar_top <- function(m, title, n = 20) {
  d <- top_n_table(m, n = n)
  if (is.null(d)) return(NULL)
  d$Feature <- factor(d$Feature, levels = rev(d$Feature))
  ggplot(d, aes(x = Feature, y = Value)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    labs(title = title, x = NULL, y = "CPM") +
    theme_minimal(base_size = 11) +
    theme(axis.text.y = element_text(size = 8))
}

pa <- read_humann(params$paFile)
rx <- read_humann(params$rxCpmFile)
ko <- read_humann(params$koFile)
gf <- read_humann(params$gfCpmFile)
```

### MetaCyc pathways

`r if (is.null(pa)) "(no pathway abundance table)" else sprintf("%d community-level pathways detected.", nrow(pa))`

```{r pathway_plot, fig.height = 6}
if (!is.null(pa)) print(bar_top(pa, "Top 20 MetaCyc pathways"))
```

```{r pathway_table}
if (!is.null(pa)) knitr::kable(top_n_table(pa, 20), row.names = FALSE,
                                col.names = c("Pathway", "Abundance"))
```

### MetaCyc reactions

`r if (is.null(rx)) "(no reactions table)" else sprintf("%d community-level reactions detected.", nrow(rx))`

```{r reactions_plot, fig.height = 6}
if (!is.null(rx)) print(bar_top(rx, "Top 20 MetaCyc reactions"))
```

### KEGG KO

`r if (is.null(ko)) "(KEGG KO regroup not run or failed; HUMAnN 4 alpha utility_mapping covers EC-annotated KOs only.)" else sprintf("%d community-level KOs detected (EC-mapped subset only).", nrow(ko))`

```{r ko_plot, fig.height = 6}
if (!is.null(ko)) print(bar_top(ko, "Top 20 KEGG KOs"))
```

### Gene families (community level)

`r if (is.null(gf)) "(no gene families table)" else sprintf("%d community-level UniRef90-EC + UniClust90 gene families detected.", nrow(gf))`

```{r gf_plot, fig.height = 6}
if (!is.null(gf)) print(bar_top(gf, "Top 20 gene families"))
```

## Stratification per organism

```{r strat_helpers, include = FALSE}
# Read a HUMAnN CPM TSV keeping stratified rows (feature|species).
# Community-level rows and specials are dropped so the section shows only
# per-organism contributions.
read_humann_strat <- function(path) {
  if (is.na(path) || !nzchar(path) || !file.exists(path)) return(NULL)
  m <- read.table(path, header = TRUE, sep = "\t", check.names = FALSE,
                  comment.char = "", quote = "", row.names = 1)
  colnames(m) <- sub("_(Abundance|Abundance-CPM|Coverage|Abundance-RPKs).*$",
                     "", colnames(m))
  strat <- m[grepl("\\|", rownames(m)), , drop = FALSE]
  if (nrow(strat) == 0) return(NULL)
  parts <- strsplit(rownames(strat), "|", fixed = TRUE)
  base    <- vapply(parts, `[`, character(1), 1)
  species <- vapply(parts, function(x)
                     if (length(x) >= 2) x[[2]] else NA_character_,
                     character(1))
  data.frame(BaseFeature = base,
             Species     = species,
             SpeciesLabel = gsub("_", " ", sub("^[a-z]__", "", species, perl = TRUE)),
             Value       = as.numeric(strat[, 1]),
             stringsAsFactors = FALSE)
}

top_organisms <- function(df, n = 20) {
  if (is.null(df) || nrow(df) == 0) return(NULL)
  agg <- aggregate(Value ~ SpeciesLabel, data = df, FUN = sum)
  agg <- agg[order(agg$Value, decreasing = TRUE), , drop = FALSE]
  head(agg, n)
}

per_feature_top_org <- function(df, top_features = 5, top_org_per = 5) {
  if (is.null(df) || nrow(df) == 0) return(NULL)
  # Top features by summed stratified CPM
  feat_totals <- aggregate(Value ~ BaseFeature, data = df, FUN = sum)
  feat_totals <- feat_totals[order(feat_totals$Value, decreasing = TRUE), ]
  top_feats <- head(feat_totals$BaseFeature, top_features)
  sub <- df[df$BaseFeature %in% top_feats, , drop = FALSE]
  # Keep only top organisms per feature
  do.call(rbind, lapply(top_feats, function(f) {
    r <- sub[sub$BaseFeature == f, ]
    r <- r[order(r$Value, decreasing = TRUE), ]
    head(r, top_org_per)
  }))
}

pa_strat <- read_humann_strat(params$paFile)
rx_strat <- read_humann_strat(params$rxCpmFile)
gf_strat <- read_humann_strat(params$gfCpmFile)
ko_strat <- read_humann_strat(params$koFile)

any_strat <- any(!vapply(list(pa_strat, rx_strat, gf_strat, ko_strat),
                         is.null, logical(1)))
```

`r if (!any_strat) sprintf("No stratified rows detected. Full HUMAnN mode (tier-1 + tier-2, vOct22 upstream) is required to see per-organism contributions; this sample ran in **%s** mode.", params$mode)`

```{r strat_summary, results = "asis", eval = any_strat}
tbls <- list(Pathways = pa_strat, Reactions = rx_strat,
             `Gene families` = gf_strat, `KEGG KOs` = ko_strat)
tbls <- Filter(Negate(is.null), tbls)
summary_df <- do.call(rbind, lapply(names(tbls), function(nm) {
  d <- tbls[[nm]]
  data.frame(Table = nm,
             `Stratified rows` = nrow(d),
             `Unique organisms` = length(unique(d$SpeciesLabel)),
             check.names = FALSE)
}))
cat("**Overall:**\n\n")
print(knitr::kable(summary_df, row.names = FALSE))
```

### Top organisms overall (pathways)

```{r strat_bar_pa, fig.height = 6, eval = !is.null(pa_strat)}
top_pa <- top_organisms(pa_strat, n = 20)
if (!is.null(top_pa) && nrow(top_pa) > 0) {
  top_pa$SpeciesLabel <- factor(top_pa$SpeciesLabel, levels = rev(top_pa$SpeciesLabel))
  ggplot(top_pa, aes(x = SpeciesLabel, y = Value)) +
    geom_col(fill = "#527085") +
    coord_flip() +
    labs(title = "Top 20 organisms (summed CPM across all pathways)",
         x = NULL, y = "Summed CPM") +
    theme_minimal(base_size = 11) +
    theme(axis.text.y = element_text(size = 8, face = "italic"))
}
```

`r if (is.null(pa_strat)) "(no stratified pathway data)"`

### Top organism contributions to top pathways

```{r strat_per_feat_pa, eval = !is.null(pa_strat)}
pf <- per_feature_top_org(pa_strat, top_features = 5, top_org_per = 5)
if (!is.null(pf) && nrow(pf) > 0) {
  disp <- data.frame(Pathway = pf$BaseFeature,
                     Organism = pf$SpeciesLabel,
                     `CPM` = round(pf$Value, 2),
                     check.names = FALSE)
  knitr::kable(disp, row.names = FALSE)
}
```

### Top organism contributions to top reactions

```{r strat_per_feat_rx, eval = !is.null(rx_strat)}
pf <- per_feature_top_org(rx_strat, top_features = 5, top_org_per = 5)
if (!is.null(pf) && nrow(pf) > 0) {
  disp <- data.frame(Reaction = pf$BaseFeature,
                     Organism = pf$SpeciesLabel,
                     `CPM` = round(pf$Value, 2),
                     check.names = FALSE)
  knitr::kable(disp, row.names = FALSE)
}
```

### Top organisms overall (reactions)

```{r strat_bar_rx, fig.height = 6, eval = !is.null(rx_strat)}
top_rx <- top_organisms(rx_strat, n = 20)
if (!is.null(top_rx) && nrow(top_rx) > 0) {
  top_rx$SpeciesLabel <- factor(top_rx$SpeciesLabel, levels = rev(top_rx$SpeciesLabel))
  ggplot(top_rx, aes(x = SpeciesLabel, y = Value)) +
    geom_col(fill = "#8FA9B8") +
    coord_flip() +
    labs(title = "Top 20 organisms (summed CPM across all reactions)",
         x = NULL, y = "Summed CPM") +
    theme_minimal(base_size = 11) +
    theme(axis.text.y = element_text(size = 8, face = "italic"))
}
```

`r if (is.null(rx_strat)) "(no stratified reactions data)"`

### Top organisms overall (gene families)

```{r strat_bar_gf, fig.height = 6, eval = !is.null(gf_strat)}
top_gf <- top_organisms(gf_strat, n = 20)
if (!is.null(top_gf) && nrow(top_gf) > 0) {
  top_gf$SpeciesLabel <- factor(top_gf$SpeciesLabel, levels = rev(top_gf$SpeciesLabel))
  ggplot(top_gf, aes(x = SpeciesLabel, y = Value)) +
    geom_col(fill = "#3A5060") +
    coord_flip() +
    labs(title = "Top 20 organisms (summed CPM across all gene families)",
         x = NULL, y = "Summed CPM") +
    theme_minimal(base_size = 11) +
    theme(axis.text.y = element_text(size = 8, face = "italic"))
}
```

`r if (is.null(gf_strat)) "(no stratified gene families data)"`

### Top organism contributions to top gene families

```{r strat_per_feat_gf, eval = !is.null(gf_strat)}
pf <- per_feature_top_org(gf_strat, top_features = 5, top_org_per = 5)
if (!is.null(pf) && nrow(pf) > 0) {
  disp <- data.frame(`Gene family` = pf$BaseFeature,
                     Organism = pf$SpeciesLabel,
                     `CPM` = round(pf$Value, 2),
                     check.names = FALSE)
  knitr::kable(disp, row.names = FALSE)
}
```

### Top organisms overall (KEGG KOs)

```{r strat_bar_ko, fig.height = 6, eval = !is.null(ko_strat)}
top_ko <- top_organisms(ko_strat, n = 20)
if (!is.null(top_ko) && nrow(top_ko) > 0) {
  top_ko$SpeciesLabel <- factor(top_ko$SpeciesLabel, levels = rev(top_ko$SpeciesLabel))
  ggplot(top_ko, aes(x = SpeciesLabel, y = Value)) +
    geom_col(fill = "#6E8D9E") +
    coord_flip() +
    labs(title = "Top 20 organisms (summed CPM across all KEGG KOs)",
         x = NULL, y = "Summed CPM") +
    theme_minimal(base_size = 11) +
    theme(axis.text.y = element_text(size = 8, face = "italic"))
}
```

`r if (is.null(ko_strat)) "(no stratified KEGG KO data — KO regroup may not have been run, or the run was translated-only)"`

### Top organism contributions to top KEGG KOs

```{r strat_per_feat_ko, eval = !is.null(ko_strat)}
pf <- per_feature_top_org(ko_strat, top_features = 5, top_org_per = 5)
if (!is.null(pf) && nrow(pf) > 0) {
  disp <- data.frame(`KEGG KO` = pf$BaseFeature,
                     Organism = pf$SpeciesLabel,
                     `CPM` = round(pf$Value, 2),
                     check.names = FALSE)
  knitr::kable(disp, row.names = FALSE)
}
```

## Output files

```{r files_table}
fs <- c(params$gfCpmFile, params$rxCpmFile, params$paFile, params$koFile)
fs <- fs[!is.na(fs) & nzchar(fs)]
fs <- fs[file.exists(fs)]
if (length(fs) > 0) {
  info <- file.info(fs)
  knitr::kable(data.frame(File = basename(fs),
                          Size_KB = round(info$size / 1024, 1)),
               row.names = FALSE)
}
```

## Method notes

HUMAnN **v4.0.0.alpha.2** with these caveats:

- **DB compatibility.** ChocoPhlAn pangenomes are pinned to `mpa_vOct22_CHOCOPhlAnSGB_202403`. MetaPhlAn profiles from newer DBs (e.g. `mpa_vJan25_CHOCOPhlAnSGB_202503`) can only drive translated-only mode — SGB IDs don't align across vintages.
- **Profile is always supplied externally.** MetaPhlAn ≥ 4.2 has an API change HUMAnN 4 alpha cannot drive itself; this app passes `--taxonomic-profile` so the MetaPhlAn binary version is irrelevant.
- **EC-filtered UniRef DB only.** Gene-family quantitation is restricted to enzymes. MetaCyc pathways work as designed; KEGG KO regroup covers the EC-mapped subset only (no transporters / regulators / structural).
- **`READS_UNMAPPED` strip.** HUMAnN 4 renamed `UNMAPPED` → `READS_UNMAPPED` in gene families; `humann_renorm_table --special n` doesn't know the new name. We strip it before CPM-renorm so it doesn't dominate the denominator.

See [biobakery forum #8523](https://forum.biobakery.org/t/metaphlan-4-humann-4-compatibility/8523) for compatibility background.
