---
params:
  reportTitle: "MAGeCK Count QC"
  project: ""
  order: ""
title: "`r params$reportTitle`"
---

<!-- Styling comes from the fgczQuartoTemplate assets staged next to this file by
     ezRun::makeQuartoReport() -> fgczQuartoTemplate::fgcz_render(); no format:
     block here. Tab colour/numbering and the toolbar are makeQuartoReport args. -->

```{r setup}
#| include: false
library(ezRun)
library(ggplot2)
library(knitr)
library(scales)

## ── Data loading (qs2 with rds fallback, checks data/ first) ──
.fgcz_load <- function(name, nthreads = 4L) {
  paths <- c(file.path("data", paste0(name, ".qs2")), paste0(name, ".qs2"),
             file.path("data", paste0(name, ".rds")), paste0(name, ".rds"))
  for (p in paths) {
    if (file.exists(p)) {
      if (grepl("\\.qs2$", p)) return(qs2::qs_read(p, nthreads = nthreads))
      return(readRDS(p))
    }
  }
  stop("No .qs2 or .rds file found for: ", name)
}

## ── Colourblind-safe palette (Okabe-Ito; report author has red-green CVD) ──
.cvd <- c("#0072B2", "#E69F00", "#009E73", "#CC79A7",
          "#56B4E9", "#D55E00", "#F0E442", "#000000")

theme_set(theme_bw(base_size = 12))
```

```{r load}
#| include: false
param <- .fgcz_load("param")
sampleName <- param$sampleName
summaryFile <- paste0(sampleName, ".countsummary.txt")
countSummary <- ezRead.table(summaryFile, row.names = NULL)

## Tolerant column access across mageck versions.
pick <- function(df, candidates) {
  hit <- intersect(candidates, colnames(df))
  if (length(hit) >= 1L) df[[hit[1]]][1] else NA
}
reads  <- suppressWarnings(as.numeric(pick(countSummary, "Reads")))
mapped <- suppressWarnings(as.numeric(pick(countSummary, "Mapped")))
pct    <- suppressWarnings(as.numeric(pick(countSummary, "Percentage")))
totalG <- suppressWarnings(as.numeric(pick(countSummary, c("TotalsgRNAs", "TotalsgRNA"))))
zeros  <- suppressWarnings(as.numeric(pick(countSummary, c("Zerocounts", "ZeroCounts"))))
gini   <- suppressWarnings(as.numeric(pick(countSummary, "GiniIndex")))

## mageck reports Percentage as a fraction (0-1); normalise to a rate in [0,1].
mapFrac <- if (!is.na(pct)) ifelse(pct > 1, pct / 100, pct) else
  if (!is.na(mapped) && !is.na(reads) && reads > 0) mapped / reads else NA
zeroFrac <- if (!is.na(zeros) && !is.na(totalG) && totalG > 0) zeros / totalG else NA
```

::: {.panel-tabset}

# Overview

::: {.callout-note title="What this report is"}
**MAGeCK count** maps the raw reads of a single sample against the sgRNA library
and reports the per-sample metrics below. The three that decide whether a sample
is usable in a pooled CRISPR screen:

1. **Mapping rate** — the fraction of reads assigned to a library sgRNA. A low
   rate usually means the sgRNA length / 5' constant region / strand is off, not
   a bad library; tune it via the `cmdOptions` parameter (e.g. `--sgrna-len`).
2. **Zero-count sgRNAs** — guides never observed. A large fraction signals
   dropout or under-sequencing.
3. **Gini index** — evenness of guide representation (0 = perfectly even,
   1 = fully skewed). High values are expected under strong selection but worth
   checking for reference samples.
:::

```{r glance-callout}
#| echo: false
#| results: asis
ct  <- if (!is.na(mapFrac) && mapFrac < 0.65) "warning" else "note"
msg <- if (!is.na(mapFrac) && mapFrac < 0.65)
  sprintf("⚠ Only **%.1f%%** of reads mapped — likely an sgRNA-length / 5' constant-region / strand problem; see the `cmdOptions` parameter.", mapFrac * 100)
else if (!is.na(mapFrac))
  sprintf("Mapping rate: **%.1f%%** of reads mapped to the library.", mapFrac * 100)
else "Mapping rate could not be determined from the countsummary file."
cat(sprintf("::: {.callout-%s}\n%s\n:::\n", ct, msg))
```

```{r glance-metrics}
#| echo: false
gm <- data.frame(
  Metric = c("Total reads", "Mapped reads", "Mapping rate",
             "Total sgRNAs", "Zero-count sgRNAs", "Gini index"),
  Value = c(
    format(reads, big.mark = ",", scientific = FALSE),
    format(mapped, big.mark = ",", scientific = FALSE),
    if (is.na(mapFrac)) "NA" else sprintf("%.1f %%", mapFrac * 100),
    format(totalG, big.mark = ",", scientific = FALSE),
    if (is.na(zeros)) "NA" else sprintf("%s (%.1f %%)",
        format(zeros, big.mark = ",", scientific = FALSE),
        ifelse(is.na(zeroFrac), NA, zeroFrac * 100)),
    if (is.na(gini)) "NA" else sprintf("%.3f", gini)),
  check.names = FALSE)
knitr::kable(gm, caption = sprintf("Per-sample count QC — %s", sampleName))
```

```{r mapping-bar}
#| echo: false
#| fig-width: 7
#| fig-height: 2.2
#| out-width: "70%"
#| fig-cap: "Reads assigned to the library vs. unmapped."
if (!is.na(reads) && !is.na(mapped)) {
  bar <- data.frame(
    stage = factor(c("Mapped", "Unmapped"), levels = c("Unmapped", "Mapped")),
    reads = c(mapped, max(reads - mapped, 0)))
  bar$lab <- sprintf("%s  (%.1f%%)", format(round(bar$reads), big.mark = ","),
                     100 * bar$reads / max(reads, 1))
  ggplot(bar, aes(x = reads, y = stage, fill = stage)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = lab), hjust = -0.03, size = 3.2) +
    scale_fill_manual(values = c(Mapped = .cvd[3], Unmapped = .cvd[6]), guide = "none") +
    scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.3))) +
    labs(x = "Reads", y = NULL)
}
```

The raw mageck output: [`r summaryFile`](`r summaryFile`){target="_blank"}.

# Parameters

```{r params-table}
#| echo: false
keep <- intersect(c("libName", "cmdOptions", "sampleName", "name"), names(param))
knitr::kable(data.frame(Parameter = keep,
                        Value = vapply(param[keep], function(x) paste(as.character(x), collapse = ", "), ""),
                        row.names = NULL, check.names = FALSE))
```

# Session info

```{r session}
#| echo: false
ezSessionInfo()
```

:::
