---
title: "`r if (exists('reportTitle')) reportTitle else 'SUSHI Report'`"
output:
  html_document:
    mathjax: https://fgcz-gstore.uzh.ch/reference/mathjax.js
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package="ezRun")
    css: !expr system.file("templates/fgcz.css", package="ezRun")
editor_options:
  chunk_output_type: inline
---

# {.tabset}

`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`

```{r setup, include=FALSE}
library(kableExtra)
library(ezRun)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, knitr.table.format = "html")
```

```{r load}
param <- ezLoadRobj("param")
sampleName <- param$sampleName
summaryFile <- paste0(sampleName, ".countsummary.txt")
countSummary <- ezRead.table(summaryFile, row.names = NULL)
```

## Count QC

`mageck count` maps the raw reads of each sample against the sgRNA library and
reports the per-sample quality metrics below. The most important indicators of a
healthy CRISPR screen are a high **mapping rate**, a low fraction of
**zero-count sgRNAs**, and a low **Gini index** (even representation of guides).

```{r qc-table, results='asis'}
## Present the metrics that matter, tolerant to column-name variation across
## mageck versions.
pick <- function(df, candidates) {
  hit <- intersect(candidates, colnames(df))
  if (length(hit) >= 1L) df[[hit[1]]] else NA
}

reads     <- pick(countSummary, c("Reads"))
mapped    <- pick(countSummary, c("Mapped"))
pct       <- pick(countSummary, c("Percentage"))
totalG    <- pick(countSummary, c("TotalsgRNAs", "TotalsgRNA"))
zeros     <- pick(countSummary, c("Zerocounts", "ZeroCounts"))
gini      <- pick(countSummary, c("GiniIndex"))

pctNum <- suppressWarnings(as.numeric(pct))
giniNum <- suppressWarnings(as.numeric(gini))

qc <- data.frame(
  Metric = c("Total reads", "Mapped reads", "Mapping rate",
             "Total sgRNAs", "Zero-count sgRNAs", "Gini index"),
  Value = c(
    format(reads, big.mark = ","),
    format(mapped, big.mark = ","),
    ifelse(is.na(pctNum), as.character(pct),
           paste0(signif(ifelse(pctNum <= 1, pctNum * 100, pctNum), 4), " %")),
    format(totalG, big.mark = ","),
    format(zeros, big.mark = ","),
    ifelse(is.na(giniNum), as.character(gini), signif(giniNum, 4))
  ),
  stringsAsFactors = FALSE
)

kable(qc, escape = FALSE, row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r flags, results='asis'}
notes <- c()
mapFrac <- if (!is.na(pctNum)) ifelse(pctNum > 1, pctNum / 100, pctNum) else NA
if (!is.na(mapFrac) && mapFrac < 0.65) {
  notes <- c(notes, sprintf("- **Low mapping rate (%.1f%%).** Check the sgRNA length / 5' constant region / strand (see the `cmdOptions` parameter, e.g. `--sgrna-len`).", mapFrac * 100))
}
if (!is.na(giniNum) && giniNum > 0.35) {
  notes <- c(notes, sprintf("- **High Gini index (%.3f).** Uneven guide representation; expected for late timepoints/strong selection but worth checking for reference samples.", giniNum))
}
if (length(notes) > 0) {
  cat("### Flags\n\n")
  cat(paste(notes, collapse = "\n"), "\n")
}
```

#### Raw countsummary

[countsummary file](`r summaryFile`){target="_blank"}

## Parameters
```{r report parameters, echo=FALSE}
param[intersect(c("libName", "cmdOptions", "sampleName"), names(param))]
```

## SessionInfo
```{r, echo=FALSE}
ezSessionInfo()
```
