---
title: "`r paste('Mothur Data Clean')`"
author: "Functional Genomics Center Zurich"
output: 
  html_document:
    self_contained: false
    includes:
      in_header: fgcz_header.html
    css: fgcz.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, fig.width=18, fig.height=12)
## This report requires summary files form mothur

require(knitr)
require(kableExtra)
require(SummarizedExperiment)
require(webshot)
require(htmlwidgets)
debug <- FALSE
setEnvironments("phantomjs")
```

```{r prepare data, include=FALSE}
### create plots: 1. abundance
mergedAbPlotIndPhylum <- plot_bar(physeqMerged, "Abundance", "Phylum") + geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="stack")
mergedAbPlotGroupGenus <- plot_bar(physeqMerged, "Group", "Abundance", "Genus") + geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="stack")


### create plots: 2. ordination
mergedOrd <- ordinate(physeqMerged, "NMDS", "bray")
plotOrdTaxaMerged = plot_ordination(physeqMerged, mergedOrd, type="taxa", color="Phylum") + facet_wrap(~Phylum, 3) + ggtitle("Taxa ordination, all samples")
plotOrdSamplesMerged = plot_ordination(physeqMerged, mergedOrd, type="samples", color="Group", shape = "Technology") 
plotOrdSamplesMerged = plotOrdSamplesMerged + geom_polygon(aes(fill=Group)) + geom_point(size=3) + ggtitle("Sample ordination, all samples")

illOrd <- ordinate(physeqIllNoTree, "NMDS", "bray")
plotOrdTaxaIll = plot_ordination(physeqIllNoTree, illOrd, type="taxa", color="Phylum") + facet_wrap(~Phylum, 3) + ggtitle("Taxa ordination, only Illumina samples")
plotOrdSamplesIll = plot_ordination(physeqIllNoTree, illOrd, type="samples", color="Group") 
plotOrdSamplesIll = plotOrdSamplesIll + geom_polygon(aes(fill=Group)) + geom_point(size=5) + ggtitle("Sample ordination, only Illumina samples")

### create plots: 3. richness 
plotRichMerged <- plot_richness(physeqMerged, x="Technology", color = "Group", measures=c("Chao1", "Shannon"))

### create plots: distribution deepest taxa
plotDeepDistPB  <- phyloSeqDivPlotAndPercUnclassified(basename(input$getFullPaths("Taxonomy_pacbio")),"Genus")
plotDeepDistIll <- phyloSeqDivPlotAndPercUnclassified(basename(input$getFullPaths("Taxonomy_Illumina")),"Genus")

### create plots: 4. tree
plotTreeMerged <- plot_tree(physeqMergedPruned , ladderize="left", color="Group", shape="Technology", label.tips="Genus")
plotTreeRadialMerged <- plot_tree(physeqMergedPruned,  color="Group", shape="Technology", ladderize="left") + coord_polar(theta="y")

### create plots:5. heatmap
plotHeatmapMerged <- plot_heatmap(physeqMergedPruned, taxa.label="Genus")

### 6: compare groups only Illumina for now
deseqResults <- phyloSeqToDeseq2_tableAndPlots(physeqIllNoTree)
tableToPlot <- deseqResults$table
tableToPlot <- tableToPlot[order(tableToPlot$padj),][1:20,]

```

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

## Data pre-processing {.tabset}

### Abundance barplots
#### Community composition by sample: phyla
```{r abundPBSamplePhylum, echo=FALSE}
plot(mergedAbPlotIndPhylum)
```

#### Community composition by group: genera
```{r abundPBGroupGenus, echo=FALSE}
plot(mergedAbPlotGroupGenus)
```

### Ordination plots
#### Taxa level
```{r ordPBtaxa, echo=FALSE}
plot(plotOrdTaxaMerged)
plot(plotOrdTaxaIll)
```

#### Sample level
```{r ordPB, echo=FALSE}
plot(plotOrdSamplesMerged)
plot(plotOrdSamplesIll)

```

### Richness plots
#### Community alpha-diversity
```{r richPB, echo=FALSE}
plot(plotRichMerged)
```

### Distribution of the deepest taxa
#### PacBio
```{r distPB, echo=FALSE}
plot(plotDeepDistPB)
```

#### Illumina
```{r distIll, echo=FALSE}
plot(plotDeepDistIll)
```

### Tree plots
#### Classical tree
```{r treePB, echo=FALSE}
plot(plotTreeMerged)
```

#### Radial tree
```{r radialTreePB, echo=FALSE}
plot(plotTreeRadialMerged)
```

### Heatmaps
#### Abundances
```{r heatPB, echo=FALSE}
plot(plotHeatmapMerged)
```

### Two-goups comparison (Illumina real data)
#### Volcano plot
```{r volcano, echo=FALSE}
plot(deseqResults$vPlot)
```

#### Log-fold changes 
```{r logfold, echo=FALSE}
plot(deseqResults$logPlot)
```

#### Summary table
```{r table, echo=FALSE}
kable(tableToPlot, row.names=FALSE, 
      col.names=colnames(tableToPlot), format="html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F,
                position = "left")
```

### SessionInfo
```{r, echo=FALSE}
sessionInfo()
```
