---
title: "`r if (exists('reportTitle')) reportTitle else 'QIIME2 SUSHI Report'`"
output:
  html_document:
    mathjax: https://fgcz-gstore.uzh.ch/reference/mathjax.js
    highlight: pygments
    theme: sand
    code_folding: hide
    toc: yes
    toc_depth: 4
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package="ezRun")
    css: !expr system.file("templates/fgcz.css", package="ezRun")
editor_options:
  chunk_output_type: inline
---

<style>
/* fgcz.css sets h1 to 8px (used across FGCZ reports where sections start at
   h2). This report's sections are h1, so restore readable header sizes
   without touching the shared CSS. */
h1 { font-size: 1.9em; font-weight: 700; margin-top: 1.2em; }
h2 { font-size: 1.5em; font-weight: 700; margin-top: 1.0em; }
h3 { font-size: 1.2em; font-weight: 700; }
h4 { font-size: 1.05em; font-weight: 700; }
details > summary { cursor: pointer; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment    = '',
  fig.width  = 12,
  fig.height = 8,
  warning    = FALSE,
  error      = FALSE,
  message    = FALSE,
  cache      = TRUE,
  echo       = FALSE
)
```

```{r load_pkgs, include=FALSE}
## Mandatory packages -- the SUSHI app already enforces these in its renv/module
suppressMessages(suppressWarnings({
  library(phyloseq)
  library(qiime2R)
  library(microbiome)
  library(vegan)
  library(pairwiseAdonis)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(tibble)
  library(reshape2)
  library(ggplot2)
  library(ggpubr)
  library(ggrepel)
  library(patchwork)
  library(cowplot)
  library(scales)
  library(RColorBrewer)
  library(rstatix)
  library(kableExtra)
  library(DT)
  library(plotly)
  library(htmlwidgets)
  library(htmltools)
  library(ggvenn)
  library(ComplexHeatmap)
  library(DESeq2)
  library(edgeR)
}))

## Optional packages -- gracefully degrade when not available
.have_ancombc        <- requireNamespace("ANCOMBC",        quietly = TRUE)
.have_aldex2         <- requireNamespace("ALDEx2",         quietly = TRUE)
.have_ggpicrust2     <- requireNamespace("ggpicrust2",     quietly = TRUE)
.have_microbiomeMrk  <- requireNamespace("microbiomeMarker", quietly = TRUE)
.have_dada2          <- requireNamespace("dada2",          quietly = TRUE)
.have_shortread      <- requireNamespace("ShortRead",      quietly = TRUE)
.have_biostrings     <- requireNamespace("Biostrings",     quietly = TRUE)
.have_biomformat     <- requireNamespace("biomformat",     quietly = TRUE)
.have_phangorn       <- requireNamespace("phangorn",       quietly = TRUE)
.have_picante        <- requireNamespace("picante",        quietly = TRUE)
.have_gunifrac       <- requireNamespace("GUniFrac",       quietly = TRUE)
.have_enhancedvolc   <- requireNamespace("EnhancedVolcano",quietly = TRUE)
.have_scatter3d      <- requireNamespace("scatterplot3d",  quietly = TRUE)
.have_microViz       <- requireNamespace("microViz",       quietly = TRUE)
.have_ggtree         <- requireNamespace("ggtree",         quietly = TRUE)
.have_ape            <- requireNamespace("ape",            quietly = TRUE)
.have_metagenomeSeq  <- requireNamespace("metagenomeSeq",  quietly = TRUE)
.have_car            <- requireNamespace("car",            quietly = TRUE)

if (.have_biostrings) suppressMessages(suppressWarnings(library(Biostrings)))
if (.have_shortread) suppressMessages(suppressWarnings(library(ShortRead)))
if (.have_dada2)     suppressMessages(suppressWarnings(library(dada2)))
if (.have_biomformat) suppressMessages(suppressWarnings(library(biomformat)))
if (.have_phangorn)  suppressMessages(suppressWarnings(library(phangorn)))
if (.have_picante)   suppressMessages(suppressWarnings(library(picante)))
if (.have_gunifrac)  suppressMessages(suppressWarnings(library(GUniFrac)))
if (.have_ape)       suppressMessages(suppressWarnings(library(ape)))
if (.have_ggtree)    suppressMessages(suppressWarnings(library(ggtree)))
if (.have_microViz)  suppressMessages(suppressWarnings(library(microViz)))

## shiny::validate masking guard from the project memory
## (jsonlite gets loaded transitively and masks validate)
.qiime2_validate <- function(...) if (requireNamespace("shiny", quietly = TRUE)) shiny::validate(...) else invisible(NULL)
```

```{r load_passed_objects, include=FALSE}
## makeRmdReport() in ezRun (util.R) serialises every named argument it
## receives to <name>.qs2 in the working directory and then renders the Rmd
## in a fresh env. The Rmd must read those objects back in -- nothing is
## auto-bound. Pull in every .qs2 sitting next to the report.
suppressMessages(suppressWarnings(library(qs2)))
.passed_files <- list.files(".", pattern = "\\.qs2$", full.names = FALSE)
for (.f in .passed_files) {
  .nm <- sub("\\.qs2$", "", .f)
  tryCatch(
    assign(.nm, qs_read(.f), envir = knitr::knit_global()),
    error = function(e) message("Failed to qs_read ", .f, ": ", conditionMessage(e))
  )
}
suppressWarnings(rm(list = intersect(c(".passed_files", ".f", ".nm"), ls())))
```

# Configuration

```{r config}
## Pull configuration from the SUSHI `param` object passed by makeRmdReport.
## All knobs are optional: sensible defaults kick in when missing.
if (!exists("param") || is.null(param)) param <- list()

.cfg <- list(
  grouping        = if (!is.null(param$grouping) && nzchar(param$grouping)) param$grouping else "Group",
  group_cols      = if (!is.null(param$grouping) && nzchar(param$grouping)) strsplit(param$grouping, ",")[[1]] else character(0),
  samples_to_prune= if (!is.null(param$samples_to_prune)) param$samples_to_prune else character(0),
  prevalence_frac = if (!is.null(param$prevalence_frac))  param$prevalence_frac  else 0.10,
  rel_abund_pct   = if (!is.null(param$rel_abund_pct))    param$rel_abund_pct    else 0.10,
  forward_primer  = if (!is.null(param$forward_primer))   param$forward_primer   else "CCTACGGGNGGCWGCAG",
  reverse_primer  = if (!is.null(param$reverse_primer))   param$reverse_primer   else "GACTACHVGGGTATCTAATCC",
  run_picrust2    = isTRUE(param$run_picrust2),
  run_fastp       = isTRUE(param$run_fastp)
)

## Resolve output directory. The R backend calls setwdNew(basename(ResultDir))
## and renders the Rmd from inside that working directory, so getwd() is the
## canonical place to drop CSV exports. Do NOT use output$getColumn("ResultDir")
## here -- that returns the gstore destination path that doesn't exist yet at
## render time (SUSHI copies results there only after the job finishes).
.report_dir <- getwd()

## Build a final `grouping` scalar (single column carrying the Group factor inside physeq).
grouping <- "Group"

## Verbose echo for traceability
cat("Configuration:\n")
cat("  Grouping column   :", .cfg$grouping, "\n")
cat("  Group sub-columns :", paste(.cfg$group_cols, collapse = ", "), "\n")
cat("  Prevalence filter :", .cfg$prevalence_frac * 100, "% of samples\n")
cat("  Abundance filter  :", .cfg$rel_abund_pct, "% of total reads\n")
cat("  Forward primer    :", .cfg$forward_primer, "\n")
cat("  Reverse primer    :", .cfg$reverse_primer, "\n")
cat("  PICRUSt2 enabled  :", .cfg$run_picrust2, "\n")
cat("  fastp enabled     :", .cfg$run_fastp, "\n")
```

```{r helpers}
## ---------------------------------------------------------------------------
## Helper functions used across the report (defined once, reused everywhere).
## ---------------------------------------------------------------------------

## Strip square brackets, fix common SILVA quirks (UCG-, A2, group, uncultured).
clean_taxonomy <- function(ps) {
  tax <- as.data.frame(tax_table(ps), stringsAsFactors = FALSE)
  if ("Genus" %in% colnames(tax)) {
    tax$Genus <- gsub("\\[|\\]", "", tax$Genus)
    tax$Genus[tax$Genus == "uncultured uncultured"] <- "unknown uncultured"
    idx_grp <- grepl("group", tax$Genus)
    tax$Genus[idx_grp] <- str_remove(tax$Genus[idx_grp], "_group")
    tax$Genus[tax$Genus == "UCG-010"] <- "Oscillospirales UCG-010"
    if ("Family" %in% colnames(tax)) {
      idx <- grepl("^UCG", tax$Genus) & !grepl("^UCG", tax$Family) & !is.na(tax$Family)
      tax$Genus[idx] <- paste0(tax$Family[idx], "_", tax$Genus[idx])
    }
    tax$Genus[tax$Genus == "A2"] <- "Lachnospiraceae_A2"
  }
  if ("Family" %in% colnames(tax)) tax$Family <- gsub("\\[|\\]", "", tax$Family)
  tax_table(ps) <- tax_table(as.matrix(tax))
  ps
}

## NULL if no taxa survive after pruning zero-sum taxa.
safe_prune_nonzero <- function(physeq) {
  if (is.null(physeq)) return(NULL)
  keep <- taxa_sums(physeq) > 0
  if (sum(keep) == 0) return(NULL)
  prune_taxa(keep, physeq)
}

## Standard DT widget matching the rest of the FGCZ reports.
fgcz_dt <- function(x, caption = NULL, escape = TRUE, page_length = 10) {
  datatable(x,
            caption    = caption,
            escape     = escape,
            extensions = "Buttons",
            rownames   = FALSE,
            options    = list(scrollX    = TRUE,
                              dom        = "lftBip",
                              buttons    = c("copy", "csv", "excel"),
                              pageLength = page_length,
                              lengthMenu = c(10, 20, 100, 200)))
}

## Save a ggplot/plotly widget as a self-contained HTML alongside the report.
save_interactive_widget <- function(widget, fname) {
  fpath <- file.path(.report_dir, fname)
  tryCatch({
    htmlwidgets::saveWidget(widget, fpath, selfcontained = TRUE)
    fname
  }, error = function(e) { message("saveWidget failed: ", e$message); "" })
}

## Aggregate to a rank and return a long df with per-sample rel. abundance.
taxa_aggregate_rank <- function(ps, rank) {
  df <- psmelt(ps)
  df <- df[!is.na(df[[rank]]), , drop = FALSE]
  df %>%
    dplyr::group_by(Sample, Group, !!rlang::sym(rank)) %>%
    dplyr::summarise(Abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
    dplyr::group_by(Sample) %>%
    dplyr::mutate(SampleTotal  = sum(Abundance, na.rm = TRUE),
                  RelAbundance = ifelse(SampleTotal > 0, Abundance / SampleTotal, 0)) %>%
    dplyr::select(-SampleTotal) %>%
    dplyr::ungroup()
}

plot_taxa_bar_all <- function(ps, rank, ylab) {
  agg <- taxa_aggregate_rank(ps, rank)
  if (nrow(agg) == 0) return(ggplot() + theme_void() + ggtitle(paste("No", rank, "data")))
  cols <- colorRampPalette(brewer.pal(12, "Set3"))(length(unique(agg[[rank]])))
  ggplot(agg, aes(x = Sample, y = Abundance, fill = .data[[rank]])) +
    facet_grid(~Group, scales = "free_x") +
    geom_bar(stat = "identity", position = "fill") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom") +
    labs(y = ylab) +
    scale_fill_manual(values = cols)
}

plot_taxa_bar_topN <- function(ps, rank, N, ylab) {
  agg <- taxa_aggregate_rank(ps, rank)
  if (nrow(agg) == 0) return(ggplot() + theme_void() + ggtitle(paste("No", rank, "data")))
  ranking <- agg %>%
    dplyr::group_by(.data[[rank]]) %>%
    dplyr::summarise(MeanRel = mean(RelAbundance, na.rm = TRUE), .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(MeanRel))
  top <- as.character(ranking[[rank]][seq_len(min(N, nrow(ranking)))])
  agg$TaxonDisp <- ifelse(as.character(agg[[rank]]) %in% top,
                          as.character(agg[[rank]]), "Other")
  agg <- agg %>%
    dplyr::group_by(Sample, Group, TaxonDisp) %>%
    dplyr::summarise(Abundance = sum(Abundance, na.rm = TRUE), .groups = "drop")
  lvls <- c(top, "Other")
  lvls <- lvls[lvls %in% unique(agg$TaxonDisp)]
  agg$TaxonDisp <- factor(agg$TaxonDisp, levels = lvls)
  cols <- colorRampPalette(brewer.pal(12, "Set3"))(length(lvls))
  if ("Other" %in% lvls) cols[length(lvls)] <- "grey70"
  ggplot(agg, aes(x = Sample, y = Abundance, fill = TaxonDisp)) +
    facet_grid(~Group, scales = "free_x") +
    geom_bar(stat = "identity", position = "fill") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom") +
    labs(y = ylab, fill = NULL) +
    scale_fill_manual(values = cols)
}

make_taxa_wide_table <- function(ps, rank) {
  agg <- taxa_aggregate_rank(ps, rank) %>%
    dplyr::mutate(RelAbundancePct = round(RelAbundance * 100, 3)) %>%
    dplyr::select(Sample, !!rlang::sym(rank), RelAbundancePct)
  wide <- tidyr::pivot_wider(agg, names_from = Sample,
                             values_from = RelAbundancePct, values_fill = 0)
  sample_cols <- setdiff(colnames(wide), rank)
  mat <- as.matrix(wide[, sample_cols, drop = FALSE]); mode(mat) <- "numeric"
  wide$`Mean%`       <- round(rowMeans(mat, na.rm = TRUE), 3)
  wide$`Prevalence%` <- round(rowSums(mat > 0, na.rm = TRUE) / length(sample_cols) * 100, 1)
  wide <- wide[, c(rank, "Mean%", "Prevalence%", sample_cols), drop = FALSE]
  wide[order(-wide$`Mean%`), , drop = FALSE]
}

## Per-sample relative abundance transform that doesn't divide by zero.
to_rel <- function(ps) {
  transform_sample_counts(ps, function(x) {
    s <- sum(x, na.rm = TRUE)
    if (s == 0) return(x)
    x / s * 100
  })
}

plot_level_heatmap <- function(ps, level = "Genus", title = "") {
  df <- psmelt(to_rel(ps))
  df <- df[!is.na(df[[level]]), , drop = FALSE]
  if (nrow(df) == 0) return(ggplot() + theme_void() + ggtitle(paste("No", level, "data")))
  df <- df %>%
    dplyr::group_by(Sample, Group, .data[[level]]) %>%
    dplyr::summarise(Abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
    dplyr::mutate(NormAbundance = log10(Abundance + 0.01))
  ggplot(df, aes(x = Sample, y = .data[[level]], fill = NormAbundance)) +
    geom_tile() +
    facet_grid(~Group, scales = "free_x") +
    scale_fill_viridis_c(name = "Log10 Abundance") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
    labs(y = level, x = "Sample ID", title = title)
}

plot_topN_heatmap <- function(ps, N = 30, title = "") {
  ps_rel <- to_rel(ps)
  taxa_means <- taxa_sums(ps_rel) / nsamples(ps_rel)
  N <- min(N, length(taxa_means))
  topN <- names(sort(taxa_means, decreasing = TRUE)[seq_len(N)])
  df <- psmelt(ps_rel) %>%
    dplyr::mutate(
      CleanGenus  = dplyr::case_when(
        !is.na(Genus)  & Genus  != "" ~ as.character(Genus),
        !is.na(Family) & Family != "" ~ paste0("f__", Family),
        TRUE ~ as.character(OTU)
      ),
      UniqueTaxa  = paste0(CleanGenus, "_", substr(OTU, 1, 6)),
      DisplayTaxa = dplyr::if_else(OTU %in% topN, UniqueTaxa, "Remainder")
    ) %>%
    dplyr::group_by(Sample, Group, DisplayTaxa) %>%
    dplyr::summarise(Abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
    dplyr::mutate(NormAbundance = log10(Abundance + 0.01))
  ggplot(df, aes(x = Sample, y = DisplayTaxa, fill = NormAbundance)) +
    geom_tile() +
    facet_grid(~Group, scales = "free_x") +
    scale_fill_viridis_c(name = "Log10 Abundance") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(y = paste("Top", N, "Taxa"), x = "Sample ID", title = title)
}

## CLR transform helper: works column-wise (per sample) on a feature×sample matrix.
clr_transform_matrix <- function(mat) {
  mat <- as.matrix(mat) + 0.5  # pseudocount
  apply(mat, 2, function(col) log(col) - mean(log(col)))
}

## Geometric mean for DESeq2 size factor estimation on sparse OTU tables.
gm_mean <- function(x, na.rm = TRUE) exp(sum(log(x[x > 0]), na.rm = na.rm) / length(x))

## Resolve an ANCOM-BC taxon row id to a readable taxonomy.
resolve_taxon_da <- function(ids, tax_tbl) {
  tax_df <- as.data.frame(tax_tbl)
  sapply(ids, function(id) {
    if (id %in% rownames(tax_df)) {
      tx <- tax_df[id, ]
      if (!is.null(tx$Species) && !is.na(tx$Species) && tx$Species != "")
        return(paste0(tx$Genus, " ", tx$Species))
      if (!is.null(tx$Genus)   && !is.na(tx$Genus)   && tx$Genus != "")   return(tx$Genus)
      if (!is.null(tx$Family)  && !is.na(tx$Family)  && tx$Family != "")  return(paste0("f__", tx$Family))
      if (!is.null(tx$Phylum)  && !is.na(tx$Phylum)  && tx$Phylum != "")  return(paste0("p__", tx$Phylum))
    }
    id
  })
}
```

# Data Loading

```{r load_data}
## physeq is supplied by the R backend via makeRmdReport.
## Apply prevalence + relative abundance filters and glom up the standard ranks.

if (!exists("physeq")) stop("`physeq` was not passed to the Rmd by makeRmdReport.")

## Inject combined Group factor if grouping columns are supplied.
if (length(.cfg$group_cols) > 0) {
  sd <- as(sample_data(physeq), "data.frame")
  present <- intersect(.cfg$group_cols, colnames(sd))
  if (length(present) > 0) {
    sd$Group <- do.call(paste, c(lapply(present, function(col) sd[[col]]), sep = "-"))
    sample_data(physeq) <- sample_data(sd)
  } else if (!"Group" %in% colnames(sd)) {
    ## Fall back: pretend everything is one group so the rest of the report still renders.
    sd$Group <- "all"
    sample_data(physeq) <- sample_data(sd)
  }
}

physeq <- clean_taxonomy(physeq)

## Filter 1: prevalence — keep ASVs present (count > 1) in at least N% of samples.
physeq_filter <- genefilter_sample(physeq,
                                   filterfun_sample(function(x) x > 1),
                                   A = .cfg$prevalence_frac * nsamples(physeq))
physeq_prev <- prune_taxa(physeq_filter, physeq)

## Filter 2: relative abundance — drop ASVs below .cfg$rel_abund_pct of total reads.
total_reads <- sum(sample_sums(physeq_prev))
if (total_reads > 0) {
  asv_rel_abund <- taxa_sums(physeq_prev) / total_reads * 100
  physeq_filtered <- prune_taxa(asv_rel_abund >= .cfg$rel_abund_pct, physeq_prev)
} else {
  physeq_filtered <- physeq_prev
}

## Optional manual sample drop list.
if (length(.cfg$samples_to_prune) > 0) {
  drop <- intersect(sample_names(physeq_filtered), .cfg$samples_to_prune)
  if (length(drop) > 0) {
    physeq_filtered <- prune_samples(!sample_names(physeq_filtered) %in% drop, physeq_filtered)
  }
}

## Relative abundance transform of the post-filter object.
physeq_transformed <- to_rel(physeq_filtered)

## Glom to standard ranks (Species optional).
has_species <- "Species" %in% colnames(tax_table(physeq_filtered)) &&
  any(!is.na(tax_table(physeq_filtered)[, "Species"]) &
        tax_table(physeq_filtered)[, "Species"] != "")
physeq_Species <- if (has_species) tax_glom(physeq_filtered, "Species", NArm = TRUE) else NULL
physeq_Genus   <- tax_glom(physeq_filtered, "Genus",   NArm = TRUE)
physeq_Family  <- tax_glom(physeq_filtered, "Family",  NArm = TRUE)
physeq_Phylum  <- tax_glom(physeq_filtered, "Phylum",  NArm = TRUE)

## Build a metadata data.frame to use across the report.
if (!exists("metadata") || is.null(metadata)) {
  metadata <- data.frame(sample_data(physeq_filtered))
  metadata$SampleID <- rownames(metadata)
} else {
  metadata <- as.data.frame(metadata)
  ## Find the sample-id column under any of the common aliases (QIIME2 uses
  ## "sample-id"; R's read.table with default check.names mangles that to
  ## "sample.id"; qiime2R/phyloseq pipelines sometimes pass "SampleID" or "#SampleID").
  if (!"SampleID" %in% colnames(metadata)) {
    .id_aliases <- c("sample-id", "sample.id", "sampleid", "Sample.ID",
                     "Sample-ID", "#SampleID", "sample_id")
    .id_match <- intersect(.id_aliases, colnames(metadata))
    if (length(.id_match) > 0) {
      metadata$SampleID <- as.character(metadata[[.id_match[1]]])
    } else if (!is.null(rownames(metadata)) &&
               !all(rownames(metadata) == as.character(seq_len(nrow(metadata))))) {
      ## Only use rownames if they are NOT the default 1..N integer index.
      metadata$SampleID <- rownames(metadata)
    } else {
      ## Last resort: assume the first column is the sample id.
      metadata$SampleID <- as.character(metadata[[1]])
    }
  }
  if (!"Group" %in% colnames(metadata)) {
    metadata$Group <- as.character(sample_data(physeq_filtered)$Group[match(metadata$SampleID, sample_names(physeq_filtered))])
  }
}

## Levels of the grouping factor — drives all per-subgroup tabsets.
group_levels <- levels(factor(sample_data(physeq_filtered)$Group))
```

# Filtering Summary

```{r filtering_summary}
filt_tbl <- data.frame(
  Stage  = c("Raw input",
             paste0("After ", .cfg$prevalence_frac * 100, "% prevalence filter"),
             paste0("After ", .cfg$rel_abund_pct, "% abundance filter"),
             "Species-level glom",
             "Genus-level glom",
             "Family-level glom",
             "Phylum-level glom"),
  ASVs   = c(ntaxa(physeq), ntaxa(physeq_prev), ntaxa(physeq_filtered),
             if (!is.null(physeq_Species)) ntaxa(physeq_Species) else NA_integer_,
             ntaxa(physeq_Genus), ntaxa(physeq_Family), ntaxa(physeq_Phylum)),
  Samples = c(nsamples(physeq), nsamples(physeq_prev), nsamples(physeq_filtered),
              if (!is.null(physeq_Species)) nsamples(physeq_Species) else NA_integer_,
              nsamples(physeq_Genus), nsamples(physeq_Family), nsamples(physeq_Phylum))
)
kable(filt_tbl, caption = "Filtering pipeline summary") %>% kable_styling()

if (!has_species) {
  cat("\nNote: no Species-level annotations survive filtering -- all Species sub-tabs are skipped.\n")
}
```

# QC {.tabset}

## Library sizes (pre-filter) {.tabset}

### Plot

```{r lib_sizes_pre, fig.width=12, fig.height=8}
## Prefer the pre-computed `library_sizes_pre` object from the R backend
## (single source of truth) and fall back to local recomputation if missing.
.asvs_detected_pre <- colSums(as(otu_table(physeq), "matrix") > 0)[sample_names(physeq)]
if (exists("library_sizes_pre") && !is.null(library_sizes_pre)) {
  .lp <- if (is.data.frame(library_sizes_pre)) {
    if ("SampleID" %in% colnames(library_sizes_pre)) {
      stats::setNames(library_sizes_pre[[setdiff(colnames(library_sizes_pre), "SampleID")[1]]],
                      as.character(library_sizes_pre$SampleID))
    } else stats::setNames(library_sizes_pre[[1]], rownames(library_sizes_pre))
  } else library_sizes_pre
  samp_ids_pre <- intersect(names(.lp), sample_names(physeq))
  if (length(samp_ids_pre) == 0) samp_ids_pre <- sample_names(physeq)
  lib_size_pre_df <- data.frame(
    Sample = samp_ids_pre,
    Reads  = as.numeric(.lp[samp_ids_pre]),
    ASVs_detected = as.integer(.asvs_detected_pre[samp_ids_pre])
  )
} else {
  lib_size_pre_df <- data.frame(
    Sample = sample_names(physeq),
    Reads  = sample_sums(physeq),
    ASVs_detected = .asvs_detected_pre
  )
}
lib_size_pre_df$Sample <- factor(lib_size_pre_df$Sample,
                                 levels = lib_size_pre_df$Sample[order(lib_size_pre_df$Reads)])
p_lib_pre <- ggplot(lib_size_pre_df,
                    aes(x = Reads, y = Sample,
                        text = paste("Sample:", Sample, "<br>Reads:", scales::comma(Reads)))) +
  geom_col(fill = "#00A6A6") +
  geom_text(aes(label = scales::comma(Reads)), hjust = -0.1, size = 3) +
  geom_vline(xintercept = 10000, color = "red", linetype = "dashed") +
  labs(x = "Read Count", y = "", title = "Library Sizes (before filtering)") +
  theme_bw() +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15)))
ggplotly(p_lib_pre, tooltip = "text")
```

### Table

```{r lib_sizes_pre_table}
lib_pre_out <- data.frame(
  SampleID = as.character(lib_size_pre_df$Sample),
  Reads    = lib_size_pre_df$Reads,
  ASVs_detected = lib_size_pre_df$ASVs_detected
)
write.csv(lib_pre_out, file.path(.report_dir, "library_sizes_pre.csv"), row.names = FALSE)
fgcz_dt(lib_pre_out, "Library sizes (pre-filtering)")
```

## Library sizes (post-filter) {.tabset}

### Plot

```{r lib_sizes_post, fig.width=12, fig.height=8}
## Prefer the pre-computed `library_sizes_post` object from the R backend
## (single source of truth) and fall back to local recomputation if missing.
.asvs_detected_post <- colSums(as(otu_table(physeq_filtered), "matrix") > 0)[sample_names(physeq_filtered)]
if (exists("library_sizes_post") && !is.null(library_sizes_post)) {
  .lpo <- if (is.data.frame(library_sizes_post)) {
    if ("SampleID" %in% colnames(library_sizes_post)) {
      stats::setNames(library_sizes_post[[setdiff(colnames(library_sizes_post), "SampleID")[1]]],
                      as.character(library_sizes_post$SampleID))
    } else stats::setNames(library_sizes_post[[1]], rownames(library_sizes_post))
  } else library_sizes_post
  samp_ids_post <- intersect(names(.lpo), sample_names(physeq_filtered))
  if (length(samp_ids_post) == 0) samp_ids_post <- sample_names(physeq_filtered)
  lib_size_df <- data.frame(
    Sample = samp_ids_post,
    Reads  = as.numeric(.lpo[samp_ids_post]),
    ASVs_detected = as.integer(.asvs_detected_post[samp_ids_post])
  )
} else {
  lib_size_df <- data.frame(
    Sample = sample_names(physeq_filtered),
    Reads  = sample_sums(physeq_filtered),
    ASVs_detected = .asvs_detected_post
  )
}
lib_size_df$Sample <- factor(lib_size_df$Sample,
                             levels = lib_size_df$Sample[order(lib_size_df$Reads)])
p_lib <- ggplot(lib_size_df,
                aes(x = Reads, y = Sample,
                    text = paste("Sample:", Sample, "<br>Reads:", scales::comma(Reads)))) +
  geom_col(fill = "#00A6A6") +
  geom_text(aes(label = scales::comma(Reads)), hjust = -0.1, size = 3) +
  geom_vline(xintercept = 10000, color = "red", linetype = "dashed") +
  labs(x = "Read Count", y = "", title = "Library Sizes (after filtering)") +
  theme_bw() +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15)))
ggplotly(p_lib, tooltip = "text")
```

### Table

```{r lib_sizes_post_table}
lib_post_out <- data.frame(
  SampleID = as.character(lib_size_df$Sample),
  Reads    = lib_size_df$Reads,
  ASVs_detected = lib_size_df$ASVs_detected
)
write.csv(lib_post_out, file.path(.report_dir, "library_sizes_post.csv"), row.names = FALSE)
fgcz_dt(lib_post_out, "Library sizes (after filtering)")
```

## Read quality profiles

### Forward reads

```{r qc_forward, fig.width=9, fig.height=4.5}
.fnFs <- character(0)
if (exists("raw_reads_path") && length(raw_reads_path) > 0 && nzchar(raw_reads_path[1]) && dir.exists(raw_reads_path[1])) {
  .fnFs <- sort(list.files(raw_reads_path[1],
                           pattern = "_R1\\.fastq\\.gz$|_R1_001\\.fastq\\.gz$",
                           full.names = TRUE))
}
if (length(.fnFs) > 0 && .have_dada2) {
  qc_F <- dada2::plotQualityProfile(.fnFs, aggregate = TRUE) +
    ggtitle("Forward Reads -- Aggregated Quality Profile")
  print(qc_F)
} else {
  cat("Forward-read FASTQs not available -- skipping quality profile.\n")
}
```

### Reverse reads

```{r qc_reverse, fig.width=9, fig.height=4.5}
.fnRs <- character(0)
if (exists("raw_reads_path") && length(raw_reads_path) > 0 && nzchar(raw_reads_path[1]) && dir.exists(raw_reads_path[1])) {
  .fnRs <- sort(list.files(raw_reads_path[1],
                           pattern = "_R2\\.fastq\\.gz$|_R2_001\\.fastq\\.gz$",
                           full.names = TRUE))
}
if (length(.fnRs) > 0 && .have_dada2) {
  qc_R <- dada2::plotQualityProfile(.fnRs, aggregate = TRUE) +
    ggtitle("Reverse Reads -- Aggregated Quality Profile")
  print(qc_R)
} else {
  cat("Reverse-read FASTQs not available -- skipping quality profile.\n")
}
```

## Primer presence check

```{r primer_check}
.primer_check_df <- NULL
if (exists("primer_check") && !is.null(primer_check) && is.data.frame(primer_check)) {
  .primer_check_df <- primer_check
} else if (exists("dada2_rep_seqs_fasta") &&
           length(dada2_rep_seqs_fasta) > 0 && nzchar(dada2_rep_seqs_fasta[1]) &&
           file.exists(dada2_rep_seqs_fasta[1]) && .have_shortread) {
  seqs <- ShortRead::sread(ShortRead::readFasta(dada2_rep_seqs_fasta[1]))
  n_seqs <- length(seqs)
  fwd_hits <- Biostrings::vcountPattern(.cfg$forward_primer, seqs, fixed = FALSE)
  rev_hits <- Biostrings::vcountPattern(.cfg$reverse_primer, seqs, fixed = FALSE)
  rev_comp_r <- as.character(Biostrings::reverseComplement(Biostrings::DNAString(.cfg$reverse_primer)))
  rev_comp_hits <- Biostrings::vcountPattern(rev_comp_r, seqs, fixed = FALSE)
  .primer_check_df <- data.frame(
    Primer        = c("Forward (5')", "Reverse (3')", "Reverse complement"),
    Sequence      = c(.cfg$forward_primer, .cfg$reverse_primer, rev_comp_r),
    Seqs_with_hit = c(sum(fwd_hits > 0), sum(rev_hits > 0), sum(rev_comp_hits > 0)),
    Percent       = c(round(sum(fwd_hits > 0) / n_seqs * 100, 2),
                      round(sum(rev_hits > 0) / n_seqs * 100, 2),
                      round(sum(rev_comp_hits > 0) / n_seqs * 100, 2))
  )
}

if (!is.null(.primer_check_df)) {
  kable(.primer_check_df, caption = "Primer presence in representative sequences") %>% kable_styling()
  if (any(.primer_check_df$Percent > 5, na.rm = TRUE)) {
    cat("\nWARNING: primers detected in >5% of sequences -- consider revising trim parameters.\n")
  } else if (all(.primer_check_df$Percent < 1, na.rm = TRUE)) {
    cat("\nPrimers successfully removed (<1% residual).\n")
  }
} else {
  cat("No representative-sequence FASTA available -- skipping primer check.\n")
}
```

## DADA2 denoise stats

```{r dada_denoise_table}
.denoise_df <- NULL
if (exists("denoise_stats_raw") && !is.null(denoise_stats_raw)) {
  .denoise_df <- as.data.frame(denoise_stats_raw)
} else if (exists("dada2") && !is.null(dada2)) {
  .denoise_df <- tryCatch({
    if (is.list(dada2) && !is.null(dada2$data)) as.data.frame(dada2$data) else as.data.frame(dada2)
  }, error = function(e) NULL)
}
if (!is.null(.denoise_df)) {
  if (!"SampleID" %in% colnames(.denoise_df)) {
    .denoise_df <- tibble::rownames_to_column(.denoise_df, "SampleID")
  }
  write.csv(.denoise_df, file.path(.report_dir, "denoise_stats.csv"), row.names = FALSE)
  fgcz_dt(.denoise_df, "DADA2 denoising stats per sample")
} else {
  cat("DADA2 denoise stats not available.\n")
}
```

```{r dada_denoise_plot, fig.width=12, fig.height=6}
if (!is.null(.denoise_df) && "non-chimeric" %in% colnames(.denoise_df) ||
    !is.null(.denoise_df) && "non.chimeric" %in% colnames(.denoise_df) ||
    !is.null(.denoise_df) && "non_chimeric" %in% colnames(.denoise_df)) {
  ncol_name <- intersect(c("non-chimeric", "non.chimeric", "non_chimeric"), colnames(.denoise_df))[1]
  d2 <- .denoise_df
  d2$.nonchim <- suppressWarnings(as.numeric(as.character(d2[[ncol_name]])))
  d2 <- d2[!is.na(d2$.nonchim), , drop = FALSE]
  if (nrow(d2) > 0) {
    p_reads <- ggplot(d2, aes(x = SampleID, y = .nonchim, fill = SampleID)) +
      geom_col() +
      theme_bw() +
      geom_text(aes(label = scales::comma(.nonchim)), vjust = -0.3, size = 3) +
      labs(title = "Reads retained after DADA2", x = NULL, y = "Non-chimeric reads") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
    print(p_reads)
  }
}
```

## fastp QC

```{r fastp_qc, results='asis'}
## Only rendered when the R backend wired in `fastp_reports`.
## Accepts either a character vector / list of paths or a data.frame with a
## path-like column. Tries to summarise per-sample reads-before/after from any
## fastp JSON it can find; otherwise falls back to a link-only table.
.fastp_ok <- exists("fastp_reports") && !is.null(fastp_reports) && length(fastp_reports) > 0
if (.fastp_ok) {
  .fastp_paths <- if (is.data.frame(fastp_reports)) {
    path_col <- intersect(c("path", "Path", "report", "Report", "html", "HTML",
                            "json", "JSON", "file", "File"),
                          colnames(fastp_reports))
    if (length(path_col) > 0) as.character(fastp_reports[[path_col[1]]])
    else as.character(fastp_reports[[ncol(fastp_reports)]])
  } else {
    as.character(unlist(fastp_reports))
  }
  .fastp_paths <- .fastp_paths[nzchar(.fastp_paths)]

  ## Split HTML vs JSON reports.
  .fastp_html <- .fastp_paths[grepl("\\.html?$", .fastp_paths, ignore.case = TRUE)]
  .fastp_json <- .fastp_paths[grepl("\\.json$",  .fastp_paths, ignore.case = TRUE)]

  ## Best-effort sample naming -- strip the common fastp suffix patterns.
  derive_sample <- function(p) {
    bn <- basename(p)
    bn <- sub("\\.fastp\\.(html|json)$", "", bn, ignore.case = TRUE)
    bn <- sub("_fastp\\.(html|json)$",  "", bn, ignore.case = TRUE)
    bn <- sub("\\.(html|json)$",         "", bn, ignore.case = TRUE)
    bn
  }

  ## Try to summarise reads before/after from JSON, defensively.
  .fastp_summary <- NULL
  if (length(.fastp_json) > 0 && requireNamespace("jsonlite", quietly = TRUE)) {
    rows <- lapply(.fastp_json, function(jp) {
      js <- tryCatch(jsonlite::fromJSON(jp, simplifyVector = TRUE),
                     error = function(e) NULL)
      if (is.null(js)) return(NULL)
      rb <- tryCatch(js$summary$before_filtering$total_reads, error = function(e) NA)
      ra <- tryCatch(js$summary$after_filtering$total_reads,  error = function(e) NA)
      if (is.null(rb)) rb <- NA
      if (is.null(ra)) ra <- NA
      data.frame(Sample      = derive_sample(jp),
                 Reads_before = suppressWarnings(as.numeric(rb)),
                 Reads_after  = suppressWarnings(as.numeric(ra)),
                 stringsAsFactors = FALSE)
    })
    rows <- rows[!sapply(rows, is.null)]
    if (length(rows) > 0) {
      .fastp_summary <- do.call(rbind, rows)
      .fastp_summary$Pct_kept <- ifelse(
        is.na(.fastp_summary$Reads_before) | .fastp_summary$Reads_before == 0, NA,
        round(.fastp_summary$Reads_after / .fastp_summary$Reads_before * 100, 2)
      )
    }
  }

  if (!is.null(.fastp_summary) && nrow(.fastp_summary) > 0) {
    cat("Per-sample reads before and after fastp filtering:\n\n")
    write.csv(.fastp_summary,
              file.path(.report_dir, "fastp_summary.csv"), row.names = FALSE)
    print(fgcz_dt(.fastp_summary, "fastp summary (per sample)"))
  } else {
    cat("Per-sample fastp JSON summaries unavailable (JSON shape unfamiliar or jsonlite missing). ",
        "Showing report links only.\n\n", sep = "")
  }

  ## HTML report links table.
  if (length(.fastp_html) > 0) {
    .fastp_links_df <- data.frame(
      Sample = derive_sample(.fastp_html),
      Report = sprintf('<a href="%s" target="_blank" rel="noopener">view</a>',
                       .fastp_html),
      Path   = .fastp_html,
      stringsAsFactors = FALSE
    )
    print(DT::datatable(
      .fastp_links_df,
      caption = "fastp HTML reports",
      escape  = FALSE,
      rownames = FALSE,
      extensions = "Buttons",
      options = list(scrollX = TRUE, dom = "lftBip",
                     buttons = c("copy", "csv", "excel"),
                     pageLength = 10, lengthMenu = c(10, 20, 100, 200))
    ))
  } else if (length(.fastp_json) > 0) {
    ## No HTML present -- still expose the JSON paths for traceability.
    print(fgcz_dt(data.frame(Sample = derive_sample(.fastp_json),
                             JSON   = .fastp_json,
                             stringsAsFactors = FALSE),
                  "fastp JSON reports"))
  }
} else {
  cat("fastp not run (run_fastp = false) or no reports available -- skipping fastp QC.\n")
}
```

# ASV table with BLAST links

```{r asv_table_blast}
## Fast-path: if the R backend already passed a fully-built `asv_taxonomy_df`
## (with ASV_ID + per-sample counts + BLAST link), use it directly.
.use_passed_asv_df <- exists("asv_taxonomy_df") && !is.null(asv_taxonomy_df) &&
  is.data.frame(asv_taxonomy_df) &&
  any(c("ASV_ID", "Feature.ID", "Feature ID", "FeatureID") %in% colnames(asv_taxonomy_df))

## physeq_all is needed for downstream chunks too -- compute once here.
physeq_all <- if (exists("physeq_unfiltered") && !is.null(physeq_unfiltered)) {
  clean_taxonomy(physeq_unfiltered)
} else if (exists("table_unfiltered") && !is.null(table_unfiltered)) {
  clean_taxonomy(table_unfiltered)
} else {
  physeq
}

if (.use_passed_asv_df &&
    "ASV_ID" %in% colnames(asv_taxonomy_df) &&
    "BLAST"  %in% colnames(asv_taxonomy_df) &&
    any(sample_names(physeq) %in% colnames(asv_taxonomy_df))) {
  ## Backend provided a complete, render-ready table. Just emit it.
  asv_tax_df <- as.data.frame(asv_taxonomy_df, stringsAsFactors = FALSE)
  write.csv(asv_tax_df[, setdiff(colnames(asv_tax_df), "BLAST"), drop = FALSE],
            file.path(.report_dir, "asv_ids_taxonomy.csv"), row.names = FALSE)
  print(DT::datatable(asv_tax_df,
    caption = paste0("All ASVs (", nrow(asv_tax_df), " total)."),
    escape  = FALSE,
    extensions = "Buttons", rownames = FALSE,
    options = list(scrollX = TRUE, dom = "lftBip",
                   buttons = c("copy", "csv", "excel"),
                   pageLength = 10, lengthMenu = c(10, 20, 100, 200))))
} else {

asv_tax_df <- data.frame(
  ASV_ID = taxa_names(physeq_all),
  as.data.frame(tax_table(physeq_all)),
  stringsAsFactors = FALSE
)

## Merge in any extra columns from the backend-provided asv_taxonomy_df
## (e.g. the QIIME2 raw Taxon string / Confidence) so users can still see
## the source taxonomy alongside the cleaned phyloseq tax_table.
if (.use_passed_asv_df) {
  extra <- as.data.frame(asv_taxonomy_df, stringsAsFactors = FALSE)
  id_col <- intersect(c("ASV_ID", "Feature.ID", "Feature ID", "FeatureID"),
                      colnames(extra))[1]
  if (!is.null(id_col) && id_col != "ASV_ID") {
    extra$ASV_ID <- as.character(extra[[id_col]])
  }
  keep_cols <- setdiff(colnames(extra),
                       c(colnames(asv_tax_df), id_col))
  if (length(keep_cols) > 0) {
    extra <- extra[, c("ASV_ID", keep_cols), drop = FALSE]
    asv_tax_df <- dplyr::left_join(asv_tax_df, extra, by = "ASV_ID")
  }
}

otu_pre_mat <- as(otu_table(physeq_all), "matrix")
if (!taxa_are_rows(physeq_all)) otu_pre_mat <- t(otu_pre_mat)
otu_pre_mat <- otu_pre_mat[asv_tax_df$ASV_ID, , drop = FALSE]
sample_cols <- colnames(otu_pre_mat)
asv_tax_df$Total_reads <- as.integer(rowSums(otu_pre_mat))
asv_tax_df$N_samples   <- as.integer(rowSums(otu_pre_mat > 0))
asv_tax_df$In_filtered <- asv_tax_df$ASV_ID %in% taxa_names(physeq_filtered)

## Attach representative sequences + BLAST link if available.
.rep_seqs <- NULL
if (exists("rep_seqs") && !is.null(rep_seqs)) {
  .rep_seqs <- if (inherits(rep_seqs, "DNAStringSet")) rep_seqs
               else if (is.list(rep_seqs) && !is.null(rep_seqs$data)) rep_seqs$data
               else NULL
} else if (exists("dada2_rep_seqs_fasta") &&
           length(dada2_rep_seqs_fasta) > 0 && file.exists(dada2_rep_seqs_fasta[1]) &&
           .have_biostrings) {
  .rep_seqs <- tryCatch(Biostrings::readDNAStringSet(dada2_rep_seqs_fasta[1]),
                        error = function(e) NULL)
}
if (!is.null(.rep_seqs) && length(.rep_seqs) > 0) {
  seq_lookup <- setNames(as.character(.rep_seqs), names(.rep_seqs))
  asv_tax_df$Sequence <- seq_lookup[asv_tax_df$ASV_ID]
  asv_tax_df$Length   <- nchar(asv_tax_df$Sequence)
  asv_tax_df$BLAST    <- ifelse(
    is.na(asv_tax_df$Sequence), "",
    sprintf('<a href="https://blast.ncbi.nlm.nih.gov/Blast.cgi?PROGRAM=blastn&PAGE_TYPE=BlastSearch&DATABASE=nt&QUERY=%s" target="_blank" rel="noopener">BLAST</a>',
            asv_tax_df$Sequence)
  )
}

## Per-sample read counts at the end so BLAST stays visible.
asv_tax_df[, sample_cols] <- as.data.frame(otu_pre_mat, stringsAsFactors = FALSE)

## CSV without the HTML BLAST column.
write.csv(asv_tax_df[, setdiff(colnames(asv_tax_df), "BLAST")],
          file.path(.report_dir, "asv_ids_taxonomy.csv"), row.names = FALSE)

print(datatable(asv_tax_df,
  caption = paste0("All ASVs (", nrow(asv_tax_df),
                   " total). In_filtered=TRUE marks the ",
                   sum(asv_tax_df$In_filtered),
                   " ASVs that survived the prevalence + abundance filter."),
  escape = FALSE,
  extensions = "Buttons", rownames = FALSE,
  options = list(scrollX = TRUE, dom = "lftBip",
                 buttons = c("copy", "csv", "excel"),
                 pageLength = 10, lengthMenu = c(10, 20, 100, 200))))
}
```

# Taxa ASVs abundance per sample {.tabset}

## Pre-filtering

```{r prevalence_pre, fig.width=15, fig.height=7}
otu_full <- as(otu_table(physeq_all), "matrix")
if (!taxa_are_rows(physeq_all)) otu_full <- t(otu_full)
asv_per_sample <- colSums(otu_full)
asv_df <- data.frame(
  SampleID      = names(asv_per_sample),
  Total_reads   = asv_per_sample,
  ASVs_detected = colSums(otu_full > 0)
)
p_pre <- ggplot(asv_df, aes(x = SampleID, y = Total_reads, fill = SampleID)) +
  geom_col() +
  theme_bw() +
  geom_text(aes(label = scales::comma(Total_reads)), vjust = -0.3, size = 3) +
  labs(title = "Total ASV abundance per sample (pre-filter)",
       x = NULL, y = "Total ASV abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
print(p_pre)
write.csv(asv_df, file.path(.report_dir, "asv_per_sample_pre.csv"), row.names = FALSE)
fgcz_dt(asv_df, "Per-sample read totals and ASV richness (pre-filter)")
```

## After filtering

```{r prevalence_post, fig.width=15, fig.height=7}
otu_filt <- as(otu_table(physeq_filtered), "matrix")
if (!taxa_are_rows(physeq_filtered)) otu_filt <- t(otu_filt)
asv_df_filt <- data.frame(
  SampleID      = colnames(otu_filt),
  Total_reads   = colSums(otu_filt),
  ASVs_detected = colSums(otu_filt > 0)
)
p_post <- ggplot(asv_df_filt, aes(x = SampleID, y = Total_reads, fill = SampleID)) +
  geom_col() +
  theme_bw() +
  geom_text(aes(label = scales::comma(Total_reads)), vjust = -0.3, size = 3) +
  labs(title = "Total ASV abundance per sample (after filtering)",
       x = NULL, y = "Total ASV abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
print(p_post)
write.csv(asv_df_filt, file.path(.report_dir, "asv_per_sample_post.csv"), row.names = FALSE)
fgcz_dt(asv_df_filt, "Per-sample read totals and ASV richness (after filter)")
```

# Taxa prevalence {.tabset}

Taxa prevalence is how widespread a taxon is across samples (independent of abundance).
The All samples tabset shows the global view; per-group sub-tabsets are generated
automatically for every level of the grouping factor.

```{r prevalence_helpers}
remove_na_rank <- function(ps, rank) {
  tax <- as.data.frame(tax_table(ps))
  keep <- !is.na(tax[[rank]])
  if (sum(keep) == 0) return(NULL)
  prune_taxa(rownames(tax)[keep], ps)
}

safe_prev_plot <- function(ps, rank, title) {
  if (is.null(ps)) { cat(title, "-- no data\n"); return(invisible(NULL)) }
  tryCatch({
    if (rank %in% c("Genus", "Species")) {
      ps_clean <- remove_na_rank(ps, rank)
      if (is.null(ps_clean)) { cat(title, "-- no data after NA removal\n"); return(invisible(NULL)) }
      p <- plot_taxa_prevalence(ps_clean, rank) + ggtitle(title) + theme(legend.position = "none")
    } else {
      p <- plot_taxa_prevalence(ps, rank) + ggtitle(title) + theme(legend.position = "none")
    }
    print(p)
  }, error = function(e) cat(title, "skipped: ", e$message, "\n"))
}
```

## Pre-filtering (all samples) {.tabset}

### Genus
```{r prev_pre_genus_all, fig.width=15, fig.height=7}
safe_prev_plot(physeq, "Genus", "Genus -- All samples (pre)")
```

### Family
```{r prev_pre_family_all, fig.width=15, fig.height=7}
safe_prev_plot(physeq, "Family", "Family -- All samples (pre)")
```

### Phylum
```{r prev_pre_phylum_all, fig.width=15, fig.height=7}
safe_prev_plot(physeq, "Phylum", "Phylum -- All samples (pre)")
```

### Species
```{r prev_pre_species_all, fig.width=15, fig.height=7, eval=has_species}
safe_prev_plot(physeq, "Species", "Species -- All samples (pre)")
```

## After filtering (all samples) {.tabset}

### Genus
```{r prev_post_genus_all, fig.width=15, fig.height=7}
safe_prev_plot(physeq_filtered, "Genus", "Genus -- All samples (post)")
```

### Family
```{r prev_post_family_all, fig.width=15, fig.height=7}
safe_prev_plot(physeq_filtered, "Family", "Family -- All samples (post)")
```

### Phylum
```{r prev_post_phylum_all, fig.width=15, fig.height=7}
safe_prev_plot(physeq_filtered, "Phylum", "Phylum -- All samples (post)")
```

### Species
```{r prev_post_species_all, fig.width=15, fig.height=7, eval=has_species}
safe_prev_plot(physeq_filtered, "Species", "Species -- All samples (post)")
```

## Per-group tabsets (post-filter) {.tabset}

```{r prev_per_group_dynamic, results='asis', fig.width=15, fig.height=7}
for (grp in group_levels) {
  ## Use prune_samples (no NSE) instead of subset_samples to avoid the
  ## loop-variable scope issue: phyloseq::subset_samples evaluates the
  ## predicate inside sample_data first, and knitr's chunk env hides the
  ## loop var on the parent-frame fallback.
  .keep <- as.character(sample_data(physeq_filtered)$Group) == grp
  ps_grp <- safe_prune_nonzero(prune_samples(.keep, physeq_filtered))
  if (is.null(ps_grp)) {
    cat(sprintf("### %s\n\n_No data in this group._\n\n", grp))
    next
  }
  cat(sprintf("### %s {.tabset}\n\n", grp))
  for (rank in c("Genus", "Family", "Phylum", "Species")) {
    if (rank == "Species" && !has_species) next
    cat(sprintf("#### %s\n\n", rank))
    cat("```{r prev_grp_", make.names(grp), "_", rank, ", fig.width=15, fig.height=7}\n", sep = "")
    cat(sprintf("safe_prev_plot(safe_prune_nonzero(subset_samples(physeq_filtered, Group == \"%s\")), \"%s\", \"%s -- %s\")\n",
                grp, rank, rank, grp))
    cat("```\n\n")
  }
}
```

# Heatmaps {.tabset}

Log10-transformed relative abundance heatmaps. Each value is `log10(rel% + 0.01)`; absent taxa render at `-2`.

## Genus level (pre-filter)
```{r heatmap_genus_pre, fig.width=20, fig.height=12}
plot_level_heatmap(physeq, level = "Genus", title = "Genus -- Pre-filter")
```

## Genus level (post-filter)
```{r heatmap_genus_post, fig.width=20, fig.height=12}
plot_level_heatmap(physeq_filtered, level = "Genus", title = "Genus -- Post-filter")
```

## Family level (post-filter)
```{r heatmap_family_post, fig.width=20, fig.height=12}
plot_level_heatmap(physeq_filtered, level = "Family", title = "Family -- Post-filter")
```

## Phylum level (post-filter)
```{r heatmap_phylum_post, fig.width=20, fig.height=12}
plot_level_heatmap(physeq_filtered, level = "Phylum", title = "Phylum -- Post-filter")
```

## Species level (post-filter)
```{r heatmap_species_post, fig.width=20, fig.height=12, eval=has_species}
plot_level_heatmap(physeq_filtered, level = "Species", title = "Species -- Post-filter")
```

## Top 30 ASVs (post-filter)
```{r heatmap_top30_post, fig.width=20, fig.height=14}
plot_topN_heatmap(physeq_filtered, N = 30, title = "Top 30 ASVs -- Post-filter")
```

## Interactive heatmaps (HTML files)

```{r heatmap_interactive, results='asis'}
items <- list(
  list("Genus -- Pre",  function() plot_level_heatmap(physeq, "Genus", "Genus -- Pre"),  "interactive_heatmap_genus_pre.html",  TRUE),
  list("Genus -- Post", function() plot_level_heatmap(physeq_filtered, "Genus", "Genus -- Post"), "interactive_heatmap_genus_post.html", TRUE),
  list("Family -- Post",function() plot_level_heatmap(physeq_filtered, "Family", "Family -- Post"), "interactive_heatmap_family_post.html", TRUE),
  list("Phylum -- Post",function() plot_level_heatmap(physeq_filtered, "Phylum", "Phylum -- Post"), "interactive_heatmap_phylum_post.html", TRUE),
  list("Top 30 -- Post",function() plot_topN_heatmap(physeq_filtered, 30, "Top 30 ASVs -- Post"), "interactive_heatmap_top30_post.html", TRUE),
  list("Species -- Post",function() plot_level_heatmap(physeq_filtered, "Species", "Species -- Post"), "interactive_heatmap_species_post.html", has_species)
)
for (it in items) {
  if (!it[[4]]) next
  tryCatch({
    widget <- ggplotly(it[[2]]())
    fn <- save_interactive_widget(widget, it[[3]])
    if (nzchar(fn)) cat(sprintf("- [%s](%s)\n", it[[1]], fn))
  }, error = function(e) cat(sprintf("- %s skipped: %s\n", it[[1]], e$message)))
}
```

# Composition barplots {.tabset}

Relative abundance per sample, faceted by Group. The All tab shows every taxon at the rank;
Top 20 / Top 50 lump the rest into "Other".

```{r barplots_setup}
.do_bar_rank <- function(rank) {
  cat(sprintf("\n### %s {.tabset}\n\n", rank))
  cat("#### All\n\n")
  print(plot_taxa_bar_all(physeq, rank, paste("Relative Abundance", rank, "(pre)")))
  print(plot_taxa_bar_all(physeq_filtered, rank, paste("Relative Abundance", rank, "(post)")))
  cat("\n\n#### Top 20\n\n")
  print(plot_taxa_bar_topN(physeq, rank, 20, paste("Top 20", rank, "(pre)")))
  print(plot_taxa_bar_topN(physeq_filtered, rank, 20, paste("Top 20", rank, "(post)")))
  cat("\n\n#### Top 50\n\n")
  print(plot_taxa_bar_topN(physeq, rank, 50, paste("Top 50", rank, "(pre)")))
  print(plot_taxa_bar_topN(physeq_filtered, rank, 50, paste("Top 50", rank, "(post)")))
}
```

## Genus
```{r barplot_genus, fig.width=20, fig.height=12}
print(plot_taxa_bar_all(physeq_filtered, "Genus", "Relative Abundance Genus (post)"))
print(plot_taxa_bar_topN(physeq_filtered, "Genus", 20, "Top 20 Genus (post)"))
print(plot_taxa_bar_topN(physeq_filtered, "Genus", 50, "Top 50 Genus (post)"))
genus_tbl_post <- make_taxa_wide_table(physeq_filtered, "Genus")
write.csv(genus_tbl_post, file.path(.report_dir, "taxa_abundance_genus_post.csv"), row.names = FALSE)
fgcz_dt(genus_tbl_post, "Genus -- Relative abundance (%) per sample, post-filter")
```

## Family
```{r barplot_family, fig.width=20, fig.height=12}
print(plot_taxa_bar_all(physeq_filtered, "Family", "Relative Abundance Family (post)"))
print(plot_taxa_bar_topN(physeq_filtered, "Family", 20, "Top 20 Family (post)"))
print(plot_taxa_bar_topN(physeq_filtered, "Family", 50, "Top 50 Family (post)"))
family_tbl_post <- make_taxa_wide_table(physeq_filtered, "Family")
write.csv(family_tbl_post, file.path(.report_dir, "taxa_abundance_family_post.csv"), row.names = FALSE)
fgcz_dt(family_tbl_post, "Family -- Relative abundance (%) per sample, post-filter")
```

## Phylum
```{r barplot_phylum, fig.width=20, fig.height=12}
print(plot_taxa_bar_all(physeq_filtered, "Phylum", "Relative Abundance Phylum (post)"))
print(plot_taxa_bar_topN(physeq_filtered, "Phylum", 20, "Top 20 Phylum (post)"))
phylum_tbl_post <- make_taxa_wide_table(physeq_filtered, "Phylum")
write.csv(phylum_tbl_post, file.path(.report_dir, "taxa_abundance_phylum_post.csv"), row.names = FALSE)
fgcz_dt(phylum_tbl_post, "Phylum -- Relative abundance (%) per sample, post-filter")
```

## Species
```{r barplot_species, fig.width=20, fig.height=12, eval=has_species}
print(plot_taxa_bar_all(physeq_filtered, "Species", "Relative Abundance Species (post)"))
print(plot_taxa_bar_topN(physeq_filtered, "Species", 20, "Top 20 Species (post)"))
print(plot_taxa_bar_topN(physeq_filtered, "Species", 50, "Top 50 Species (post)"))
species_tbl_post <- make_taxa_wide_table(physeq_filtered, "Species")
write.csv(species_tbl_post, file.path(.report_dir, "taxa_abundance_species_post.csv"), row.names = FALSE)
fgcz_dt(species_tbl_post, "Species -- Relative abundance (%) per sample, post-filter")
```

## Interactive (HTML files)

```{r barplot_interactive, results='asis'}
ranks_iv <- c("Phylum", "Family", "Genus")
if (has_species) ranks_iv <- c("Species", ranks_iv)
for (rank in ranks_iv) {
  for (state in c("pre", "post")) {
    ps_use <- if (state == "pre") physeq else physeq_filtered
    tryCatch({
      p <- plot_taxa_bar_all(ps_use, rank, paste("Relative Abundance", rank, "(", state, ")"))
      widget <- ggplotly(p)
      fn <- save_interactive_widget(widget,
                                    paste0("interactive_barplot_", tolower(rank), "_", state, ".html"))
      if (nzchar(fn)) cat(sprintf("- [%s -- %s-filter](%s)\n", rank, state, fn))
    }, error = function(e) cat(sprintf("- %s %s skipped: %s\n", rank, state, e$message)))
  }
}
```

# Phylogenetic tree

```{r phylo_tree, fig.width=15, fig.height=10}
.tree_obj <- NULL
if (exists("tree") && !is.null(tree)) {
  .tree_obj <- tree
} else if (!is.null(phy_tree(physeq_filtered, errorIfNULL = FALSE))) {
  .tree_obj <- phy_tree(physeq_filtered)
}

if (!is.null(.tree_obj)) {
  if (.have_ggtree) {
    tryCatch({
      melt_simple <- psmelt(physeq_filtered) %>%
        dplyr::filter(Abundance < 50) %>%
        dplyr::select(OTU, val = Abundance)
      p <- ggtree::ggtree(physeq_filtered, layout = "fan", open.angle = 10) +
        ggtree::geom_tippoint(mapping = aes(color = Genus), size = 1.5, show.legend = TRUE)
      print(p + ggtitle("Phylogenetic tree (ggtree)"))
    }, error = function(e) {
      cat("ggtree plot failed: ", e$message, "\n")
      if (.have_ape) ape::plot.phylo(.tree_obj, type = "fan", show.tip.label = FALSE,
                                     main = "Phylogenetic tree (ape fallback)")
    })
  } else if (.have_ape) {
    ape::plot.phylo(.tree_obj, type = "fan", show.tip.label = FALSE,
                    main = "Phylogenetic tree")
  } else {
    cat("Neither ggtree nor ape available -- skipping phylogenetic tree.\n")
  }
} else {
  cat("No tree object available -- skipping phylogenetic tree.\n")
}
```

# Venn diagram across grouping levels

```{r venn, fig.width=20, fig.height=9}
otu_pre  <- as(otu_table(physeq), "matrix")
if (!taxa_are_rows(physeq)) otu_pre <- t(otu_pre)
otu_post <- as(otu_table(physeq_filtered), "matrix")
if (!taxa_are_rows(physeq_filtered)) otu_post <- t(otu_post)

groups_post <- as.character(sample_data(physeq_filtered)$Group)
names(groups_post) <- sample_names(physeq_filtered)

make_group_sets <- function(otu_mat, samp_groups) {
  setNames(lapply(unique(samp_groups), function(g) {
    samp_in <- names(samp_groups)[samp_groups == g]
    samp_in <- intersect(samp_in, colnames(otu_mat))
    if (length(samp_in) == 0) return(character(0))
    asvs <- rowSums(otu_mat[, samp_in, drop = FALSE] >= 1) > 0
    rownames(otu_mat)[asvs]
  }), unique(samp_groups))
}

sets_post <- make_group_sets(otu_post, groups_post)

if (length(sets_post) >= 2 && length(sets_post) <= 5) {
  print(ggvenn(sets_post, set_name_size = 4, text_size = 4) +
          ggtitle(paste("Shared ASVs across", length(sets_post), "groups")))
} else if (length(sets_post) > 5) {
  m <- make_comb_mat(rapply(sets_post, na.omit, how = "replace"), mode = "distinct")
  print(UpSet(m,
              comb_order = order(comb_size(m), decreasing = TRUE),
              top_annotation = upset_top_annotation(m, add_numbers = TRUE),
              right_annotation = upset_right_annotation(m, add_numbers = TRUE)))
}

## Membership table
membership <- as.data.frame(
  do.call(cbind, lapply(sets_post, function(asvs) as.integer(rownames(otu_post) %in% asvs))),
  stringsAsFactors = FALSE
)
membership$ASV_ID    <- rownames(otu_post)
membership$N_groups  <- rowSums(membership[, names(sets_post), drop = FALSE])
write.csv(membership, file.path(.report_dir, "venn_membership_post.csv"), row.names = FALSE)
fgcz_dt(head(membership[order(-membership$N_groups), ], 200),
        "Top 200 ASVs by group-membership count (post-filter)")
```

# Alpha diversity {.tabset}

Shannon -- richness + evenness combined. Observed -- raw ASV count. Faith's PD -- phylogenetic
breadth (summed branch lengths). Rarefaction equalises sequencing depth across samples.

## Rarefaction curve

```{r rarecurve, fig.width=6, fig.height=6}
otu_tab <- t(as(otu_table(physeq_filtered), "matrix"))
if (taxa_are_rows(physeq_filtered)) otu_tab <- as(otu_table(physeq_filtered), "matrix") %>% t()
tryCatch(rarecurve(otu_tab, step = 50, cex = 0.5),
         error = function(e) cat("rarecurve failed: ", e$message, "\n"))
```

## All measures (rarefied)

```{r alpha_div_rarefied, fig.width=14, fig.height=10}
set.seed(123)
ps_clean <- prune_samples(sample_sums(physeq_filtered) > 0, physeq_filtered)
ps_clean <- prune_taxa(taxa_sums(ps_clean) > 0, ps_clean)
min_lib  <- min(sample_sums(ps_clean))

ps_rarefied <- tryCatch(rarefy_even_depth(ps_clean, sample.size = min_lib, rngseed = 123,
                                          replace = FALSE, verbose = FALSE),
                        error = function(e) NULL)

richness_rar <- NULL
if (!is.null(ps_rarefied)) {
  richness_rar <- estimate_richness(ps_rarefied,
                                    measures = c("Observed", "Chao1", "Shannon", "Simpson")) %>%
    tibble::rownames_to_column("SampleID")
  richness_rar$SampleID <- str_remove(richness_rar$SampleID, "^X")

  ## Prefer QIIME2-supplied evenness vector (Pielou's) from `evenness$data`
  ## over re-computing with microbiome::evenness.
  pielou_rar <- NULL
  if (exists("evenness") && !is.null(evenness)) {
    .ev_df <- if (is.list(evenness) && !is.null(evenness$data)) evenness$data else evenness
    .ev_df <- as.data.frame(.ev_df, stringsAsFactors = FALSE)
    if (!"SampleID" %in% colnames(.ev_df)) {
      .ev_df <- tibble::rownames_to_column(.ev_df, "SampleID")
    }
    ## QIIME2 column is usually `pielou_evenness`; rename to `pielou` to match
    ## downstream alpha_metrics intersect().
    ev_val_col <- setdiff(colnames(.ev_df), "SampleID")[1]
    if (!is.null(ev_val_col) && ev_val_col != "pielou") {
      names(.ev_df)[names(.ev_df) == ev_val_col] <- "pielou"
    }
    pielou_rar <- .ev_df[, c("SampleID", "pielou"), drop = FALSE]
  } else {
    pielou_rar <- microbiome::evenness(ps_rarefied, index = "pielou") %>%
      tibble::rownames_to_column("SampleID")
  }

  ## Prefer QIIME2-supplied observed-features vector from `richness$data` over
  ## the locally-estimated Observed column (overwrite when available).
  if (exists("richness") && !is.null(richness)) {
    .rich_df <- if (is.list(richness) && !is.null(richness$data)) richness$data else richness
    .rich_df <- as.data.frame(.rich_df, stringsAsFactors = FALSE)
    if (!"SampleID" %in% colnames(.rich_df)) {
      .rich_df <- tibble::rownames_to_column(.rich_df, "SampleID")
    }
    rich_val_col <- setdiff(colnames(.rich_df), "SampleID")[1]
    if (!is.null(rich_val_col)) {
      names(.rich_df)[names(.rich_df) == rich_val_col] <- "Observed.qiime2"
      richness_rar <- dplyr::left_join(richness_rar, .rich_df[, c("SampleID", "Observed.qiime2")],
                                       by = "SampleID")
      ## Replace Observed with QIIME2's value where available.
      ok <- !is.na(richness_rar$Observed.qiime2)
      richness_rar$Observed[ok] <- richness_rar$Observed.qiime2[ok]
      richness_rar$Observed.qiime2 <- NULL
    }
  }

  meta_rar   <- as(sample_data(ps_rarefied), "data.frame") %>%
    tibble::rownames_to_column("SampleID")
  richness_rar <- richness_rar %>%
    dplyr::left_join(pielou_rar, by = "SampleID") %>%
    dplyr::left_join(meta_rar,   by = "SampleID")
  richness_rar$Shannon.Effective <- round(exp(richness_rar$Shannon), 2)

  ## Prefer QIIME2-supplied Faith's PD vector over the picante recomputation.
  if (exists("faith_pd") && !is.null(faith_pd)) {
    .fp_df <- if (is.list(faith_pd) && !is.null(faith_pd$data)) faith_pd$data else faith_pd
    .fp_df <- as.data.frame(.fp_df, stringsAsFactors = FALSE)
    if (!"SampleID" %in% colnames(.fp_df)) {
      .fp_df <- tibble::rownames_to_column(.fp_df, "SampleID")
    }
    fp_val_col <- setdiff(colnames(.fp_df), "SampleID")[1]
    if (!is.null(fp_val_col) && fp_val_col != "Faiths.PD") {
      names(.fp_df)[names(.fp_df) == fp_val_col] <- "Faiths.PD"
    }
    richness_rar <- dplyr::left_join(richness_rar,
                                     .fp_df[, c("SampleID", "Faiths.PD")],
                                     by = "SampleID")
  } else if (.have_phangorn && !is.null(phy_tree(ps_rarefied, errorIfNULL = FALSE)) && .have_picante) {
    my_tree <- phy_tree(ps_rarefied)
    if (!ape::is.rooted(my_tree)) my_tree <- phangorn::midpoint(my_tree)
    faith_rar <- tryCatch(picante::pd(t(as.data.frame(otu_table(ps_rarefied))),
                                      my_tree, include.root = TRUE),
                          error = function(e) NULL)
    if (!is.null(faith_rar)) {
      faith_rar$SampleID <- rownames(faith_rar)
      richness_rar <- richness_rar %>%
        dplyr::left_join(faith_rar[, c("SampleID", "PD")], by = "SampleID") %>%
        dplyr::rename(Faiths.PD = PD)
    }
  }

  lev <- levels(as.factor(richness_rar$Group))
  L.pairs <- if (length(lev) >= 2) combn(seq_along(lev), 2, simplify = FALSE, FUN = function(i) lev[i]) else list()
  alpha_metrics <- intersect(c("Observed", "Chao1", "Shannon", "Simpson", "pielou", "Shannon.Effective", "Faiths.PD"),
                             colnames(richness_rar))
  alpha_plots <- lapply(alpha_metrics, function(metric) {
    p <- ggviolin(richness_rar, x = "Group", y = metric, fill = "Group",
                  add = "boxplot", add.params = list(fill = "white", outlier.shape = NA)) +
      theme_bw() + ggtitle(paste(metric, "-- Rarefied")) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    if (length(L.pairs) > 0) {
      p <- p + stat_compare_means(comparisons = L.pairs, method = "wilcox.test", label = "p.format")
    }
    p
  })
  if (length(alpha_plots) > 0) print(wrap_plots(alpha_plots, ncol = 3))
} else {
  cat("Rarefaction failed -- skipping rarefied alpha diversity.\n")
}
```

## Shannon Kruskal-Wallis

```{r kw_test}
.use_alpha <- NULL
if (!is.null(richness_rar) && "Shannon" %in% colnames(richness_rar)) {
  .use_alpha <- richness_rar
} else if (exists("shannon") && !is.null(shannon)) {
  s <- if (is.list(shannon) && !is.null(shannon$data)) shannon$data else as.data.frame(shannon)
  .use_alpha <- s %>% tibble::rownames_to_column("SampleID")
  if (!"Shannon" %in% colnames(.use_alpha) && "shannon_entropy" %in% colnames(.use_alpha)) {
    .use_alpha$Shannon <- .use_alpha$shannon_entropy
  }
  .use_alpha <- dplyr::left_join(.use_alpha, metadata, by = "SampleID")
}

if (!is.null(.use_alpha) && "Shannon" %in% colnames(.use_alpha) && "Group" %in% colnames(.use_alpha)) {
  if (length(unique(.use_alpha$Group)) >= 2) {
    kw <- kruskal.test(Shannon ~ as.factor(Group), data = .use_alpha)
    cat("Global Kruskal-Wallis chi-squared = ", round(kw$statistic, 3), "\n")
    cat("Global Kruskal-Wallis p-value     = ", signif(kw$p.value, 4), "\n\n")

    pw <- tryCatch(pairwise_wilcox_test(data = .use_alpha,
                                        formula = Shannon ~ Group,
                                        p.adjust.method = "BH"),
                   error = function(e) NULL)
    if (!is.null(pw)) {
      write.csv(pw, file.path(.report_dir, "alpha_diversity_pairwise_wilcoxon.csv"), row.names = FALSE)
      fgcz_dt(pw[, intersect(c("group1", "group2", "n1", "n2", "statistic", "p", "p.adj", "p.adj.signif"),
                              colnames(pw))],
              "Pairwise Wilcoxon (Shannon)")
    }
  } else {
    cat("Only one group level present -- skipping KW test.\n")
  }
} else {
  cat("Shannon values not available -- skipping KW test.\n")
}
```

## Faith's PD

```{r faith_pd_tab, fig.width=8, fig.height=6}
## Render a Faith's PD violin/boxplot by Group with Kruskal-Wallis, using the
## QIIME2-supplied `faith_pd` vector. Guarded so the section disappears when
## the input is missing (e.g. older runs without faith_pd_vector.qza).
if (exists("faith_pd") && !is.null(faith_pd)) {
  .fp_df_tab <- if (is.list(faith_pd) && !is.null(faith_pd$data)) faith_pd$data else faith_pd
  .fp_df_tab <- as.data.frame(.fp_df_tab, stringsAsFactors = FALSE)
  if (!"SampleID" %in% colnames(.fp_df_tab)) {
    .fp_df_tab <- tibble::rownames_to_column(.fp_df_tab, "SampleID")
  }
  fp_val_col <- setdiff(colnames(.fp_df_tab), "SampleID")[1]
  if (!is.null(fp_val_col) && fp_val_col != "Faiths.PD") {
    names(.fp_df_tab)[names(.fp_df_tab) == fp_val_col] <- "Faiths.PD"
  }
  .fp_df_tab <- dplyr::left_join(.fp_df_tab, metadata[, intersect(c("SampleID", "Group"),
                                                                 colnames(metadata)), drop = FALSE],
                                 by = "SampleID")
  .fp_df_tab <- .fp_df_tab[!is.na(.fp_df_tab$Faiths.PD) & !is.na(.fp_df_tab$Group), , drop = FALSE]

  if (nrow(.fp_df_tab) > 0 && length(unique(.fp_df_tab$Group)) >= 2) {
    write.csv(.fp_df_tab, file.path(.report_dir, "alpha_diversity_faith_pd.csv"),
              row.names = FALSE)
    lev_fp <- levels(as.factor(.fp_df_tab$Group))
    L.pairs.fp <- combn(seq_along(lev_fp), 2, simplify = FALSE,
                        FUN = function(i) lev_fp[i])
    p_fp <- ggviolin(.fp_df_tab, x = "Group", y = "Faiths.PD", fill = "Group",
                     add = "boxplot",
                     add.params = list(fill = "white", outlier.shape = NA)) +
      theme_bw() + ggtitle("Faith's PD by Group") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      ylab("Faith's PD")
    if (length(L.pairs.fp) > 0) {
      p_fp <- p_fp + stat_compare_means(comparisons = L.pairs.fp,
                                        method = "wilcox.test",
                                        label = "p.format")
    }
    print(p_fp)

    kw_fp <- kruskal.test(Faiths.PD ~ as.factor(Group), data = .fp_df_tab)
    cat("\nKruskal-Wallis (Faith's PD) chi-squared =", round(kw_fp$statistic, 3), "\n")
    cat("Kruskal-Wallis (Faith's PD) p-value     =", signif(kw_fp$p.value, 4), "\n")

    pw_fp <- tryCatch(pairwise_wilcox_test(data = .fp_df_tab,
                                           formula = Faiths.PD ~ Group,
                                           p.adjust.method = "BH"),
                      error = function(e) NULL)
    if (!is.null(pw_fp)) {
      write.csv(pw_fp,
                file.path(.report_dir, "alpha_diversity_faith_pd_pairwise_wilcoxon.csv"),
                row.names = FALSE)
      fgcz_dt(pw_fp[, intersect(c("group1", "group2", "n1", "n2", "statistic",
                                  "p", "p.adj", "p.adj.signif"), colnames(pw_fp))],
              "Pairwise Wilcoxon (Faith's PD)")
    }
  } else {
    cat("Faith's PD vector available but not enough groups -- skipping comparison.\n")
  }
} else {
  cat("Faith's PD vector not provided -- skipping (older runs without faith_pd_vector.qza).\n")
}
```

# Beta diversity {.tabset}

PERMANOVA R^2 = fraction of community variation explained by Group. Betadisper checks
homogeneity of group dispersions; significant betadisper means PERMANOVA differences may
partly reflect dispersion rather than location.

## Jaccard {.tabset}

```{r beta_jaccard_prep}
.jaccard_data <- NULL
.jaccard_mat  <- NULL
if (exists("jaccard") && !is.null(jaccard)) {
  .jaccard_data <- if (is.list(jaccard) && !is.null(jaccard$data)) jaccard$data else jaccard
}
if (exists("jaccardmatrix") && !is.null(jaccardmatrix)) {
  .jaccard_mat <- if (is.list(jaccardmatrix) && !is.null(jaccardmatrix$data)) as.matrix(jaccardmatrix$data) else as.matrix(jaccardmatrix)
}
if (is.null(.jaccard_mat)) {
  .jaccard_mat <- tryCatch(as.matrix(phyloseq::distance(physeq_filtered, method = "jaccard")),
                           error = function(e) NULL)
}
```

### PCoA

```{r jaccard_pcoa, fig.width=7, fig.height=6}
if (!is.null(.jaccard_mat)) {
  ord_j <- ordinate(physeq_filtered, "PCoA", as.dist(.jaccard_mat))
  print(plot_ordination(physeq_filtered, ord_j, color = "Group") +
          geom_point(size = 3, alpha = 0.6) + stat_ellipse() + theme_bw() +
          ggtitle("Jaccard PCoA"))
} else {
  cat("No Jaccard matrix available.\n")
}
```

### PERMANOVA

```{r jaccard_permanova}
if (!is.null(.jaccard_mat)) {
  meta2 <- metadata[metadata$SampleID %in% rownames(.jaccard_mat), , drop = FALSE]
  .ng <- length(unique(na.omit(meta2$Group)))
  if (.ng < 2) {
    cat("PERMANOVA skipped: needs >=2 Group levels (found ", .ng, ").\n", sep = "")
  } else {
    perm  <- vegan::adonis2(.jaccard_mat ~ meta2$Group, permutations = 999, method = "jaccard")
    cat("Global PERMANOVA p-value =", signif(perm$`Pr(>F)`[1], 3), "\n")
    cat("Global PERMANOVA F        =", round(perm$F[1], 2), "\n")
    if (.ng >= 3) {
      pw_jac <- pairwise.adonis(.jaccard_mat, meta2$Group, perm = 999, sim.method = "jaccard")
      fgcz_dt(pw_jac, "Pairwise PERMANOVA -- Jaccard")
    }
  }
}
```

### Betadispersion

```{r jaccard_betadisper}
if (!is.null(.jaccard_mat)) {
  meta2 <- metadata[metadata$SampleID %in% rownames(.jaccard_mat), , drop = FALSE]
  .ng <- length(unique(na.omit(meta2$Group)))
  if (.ng < 2) {
    cat("Betadispersion skipped: needs >=2 Group levels (found ", .ng, ").\n", sep = "")
  } else {
    bd <- vegan::betadisper(as.dist(.jaccard_mat), meta2$Group)
    print(anova(bd))
    print(vegan::permutest(bd, pairwise = TRUE))
  }
}
```

## Bray-Curtis {.tabset}

```{r beta_bray_prep}
.bray_data <- NULL
.bray_mat  <- NULL
if (exists("bray") && !is.null(bray)) {
  .bray_data <- if (is.list(bray) && !is.null(bray$data)) bray$data else bray
}
if (exists("braymatrix") && !is.null(braymatrix)) {
  .bray_mat <- if (is.list(braymatrix) && !is.null(braymatrix$data)) as.matrix(braymatrix$data) else as.matrix(braymatrix)
}
if (is.null(.bray_mat)) {
  .bray_mat <- tryCatch(as.matrix(phyloseq::distance(physeq_filtered, method = "bray")),
                        error = function(e) NULL)
}
```

### PCoA

```{r bray_pcoa, fig.width=7, fig.height=6}
if (!is.null(.bray_mat)) {
  ord_b <- ordinate(physeq_filtered, "PCoA", as.dist(.bray_mat))
  print(plot_ordination(physeq_filtered, ord_b, color = "Group") +
          geom_point(size = 3, alpha = 0.6) + stat_ellipse() + theme_bw() +
          ggtitle("Bray-Curtis PCoA"))
}
```

### NMDS

```{r bray_nmds, fig.width=7, fig.height=6}
if (!is.null(.bray_mat)) {
  tryCatch({
    ps_rel <- microbiome::transform(physeq_filtered, "compositional")
    p <- plot_landscape(ps_rel, "NMDS", "bray", col = grouping) +
           labs(title = "NMDS / Bray")
    print(p)
  }, error = function(e) cat("NMDS failed: ", e$message, "\n"))
}
```

### PERMANOVA

```{r bray_permanova}
if (!is.null(.bray_mat)) {
  meta2 <- metadata[metadata$SampleID %in% rownames(.bray_mat), , drop = FALSE]
  .ng <- length(unique(na.omit(meta2$Group)))
  if (.ng < 2) {
    cat("PERMANOVA skipped: needs >=2 Group levels (found ", .ng, ").\n", sep = "")
  } else {
    perm  <- vegan::adonis2(.bray_mat ~ meta2$Group, permutations = 999, method = "bray")
    cat("Global PERMANOVA p-value =", signif(perm$`Pr(>F)`[1], 3), "\n")
    cat("Global PERMANOVA F        =", round(perm$F[1], 2), "\n")
    if (.ng >= 3) {
      pw_bc <- pairwise.adonis(.bray_mat, meta2$Group, perm = 999, sim.method = "bray")
      fgcz_dt(pw_bc, "Pairwise PERMANOVA -- Bray-Curtis")
    }
  }
}
```

### Betadispersion

```{r bray_betadisper}
if (!is.null(.bray_mat)) {
  meta2 <- metadata[metadata$SampleID %in% rownames(.bray_mat), , drop = FALSE]
  .ng <- length(unique(na.omit(meta2$Group)))
  if (.ng < 2) {
    cat("Betadispersion skipped: needs >=2 Group levels (found ", .ng, ").\n", sep = "")
  } else {
    bd <- vegan::betadisper(as.dist(.bray_mat), meta2$Group)
    print(anova(bd))
    print(vegan::permutest(bd, pairwise = TRUE))
  }
}
```

## Weighted UniFrac {.tabset}

```{r beta_unifrac_prep}
.uf_mat <- NULL
if (exists("unifracmatrix") && !is.null(unifracmatrix)) {
  .uf_mat <- if (is.list(unifracmatrix) && !is.null(unifracmatrix$data)) as.matrix(unifracmatrix$data) else as.matrix(unifracmatrix)
}
if (is.null(.uf_mat) && !is.null(phy_tree(physeq_filtered, errorIfNULL = FALSE))) {
  .uf_mat <- tryCatch(as.matrix(phyloseq::distance(physeq_filtered, method = "wunifrac")),
                      error = function(e) NULL)
}

## Prefer the QIIME2-computed PCoA vectors from `unifrac$data$Vectors` for the
## scatter plot (matches the Jaccard/Bray pcoa_results pattern). Falls back to
## the local ordinate() call if the full unifrac QZA was not supplied.
.uf_vectors <- NULL
if (exists("unifrac") && !is.null(unifrac)) {
  .uf_vectors <- tryCatch({
    v <- if (is.list(unifrac) && !is.null(unifrac$data) &&
             !is.null(unifrac$data$Vectors)) unifrac$data$Vectors else NULL
    if (!is.null(v)) as.data.frame(v, stringsAsFactors = FALSE) else NULL
  }, error = function(e) NULL)
}
```

### PCoA

```{r uf_pcoa, fig.width=7, fig.height=6}
if (!is.null(.uf_vectors) && nrow(.uf_vectors) > 0) {
  ## QIIME2 pcoa_results -> Vectors: SampleID + Axis.1 + Axis.2 + ...
  uf_v <- .uf_vectors
  id_col <- intersect(c("SampleID", "Sample.ID", "Sample-ID", "sample-id",
                        "SampleId", "sample_id"), colnames(uf_v))
  if (length(id_col) == 0) {
    ## Some qiime2R versions put the id in the first column; coerce to SampleID.
    uf_v$SampleID <- as.character(uf_v[[1]])
    id_col <- "SampleID"
  } else {
    names(uf_v)[names(uf_v) == id_col[1]] <- "SampleID"
    id_col <- "SampleID"
  }
  ax_cols <- grep("^(PC|Axis\\.?)\\s*\\d+$", colnames(uf_v), value = TRUE)
  if (length(ax_cols) < 2) ax_cols <- setdiff(colnames(uf_v), "SampleID")[1:2]
  pc1 <- ax_cols[1]; pc2 <- ax_cols[2]
  uf_v <- dplyr::left_join(uf_v, metadata[, intersect(c("SampleID", "Group"),
                                                     colnames(metadata)), drop = FALSE],
                           by = "SampleID")
  print(ggplot(uf_v, aes(x = .data[[pc1]], y = .data[[pc2]], color = Group)) +
          geom_point(size = 3, alpha = 0.6) +
          stat_ellipse() +
          theme_bw() +
          labs(x = pc1, y = pc2, title = "Weighted UniFrac PCoA (QIIME2)"))
} else if (!is.null(.uf_mat)) {
  ord_u <- ordinate(physeq_filtered, "PCoA", as.dist(.uf_mat))
  print(plot_ordination(physeq_filtered, ord_u, color = "Group") +
          geom_point(size = 3, alpha = 0.6) + stat_ellipse() + theme_bw() +
          ggtitle("Weighted UniFrac PCoA"))
}
```

### PERMANOVA

```{r uf_permanova}
if (!is.null(.uf_mat)) {
  meta2 <- metadata[metadata$SampleID %in% rownames(.uf_mat), , drop = FALSE]
  .ng <- length(unique(na.omit(meta2$Group)))
  if (.ng < 2) {
    cat("PERMANOVA skipped: needs >=2 Group levels (found ", .ng, ").\n", sep = "")
  } else {
    perm  <- vegan::adonis2(.uf_mat ~ meta2$Group, permutations = 999)
    cat("Global PERMANOVA p-value =", signif(perm$`Pr(>F)`[1], 3), "\n")
    cat("Global PERMANOVA F        =", round(perm$F[1], 2), "\n")
    if (.ng >= 3) {
      pw_uf <- pairwise.adonis(.uf_mat, meta2$Group, perm = 999)
      fgcz_dt(pw_uf, "Pairwise PERMANOVA -- Weighted UniFrac")
    }
  }
}
```

### Betadispersion

```{r uf_betadisper}
if (!is.null(.uf_mat)) {
  meta2 <- metadata[metadata$SampleID %in% rownames(.uf_mat), , drop = FALSE]
  .ng <- length(unique(na.omit(meta2$Group)))
  if (.ng < 2) {
    cat("Betadispersion skipped: needs >=2 Group levels (found ", .ng, ").\n", sep = "")
  } else {
    bd <- vegan::betadisper(as.dist(.uf_mat), meta2$Group)
    print(anova(bd))
    print(vegan::permutest(bd, pairwise = TRUE))
  }
}
```

# Differential abundance {.tabset}

## Properties of taxonomic profiling data

```{r da_props, fig.width=12, fig.height=4}
otu_mat_f <- as(otu_table(physeq_filtered), "matrix")
if (!taxa_are_rows(physeq_filtered)) otu_mat_f <- t(otu_mat_f)

## Overdispersion: variance > mean
df_disp <- reshape2::melt(otu_mat_f, varnames = c("Taxon", "Sample"), value.name = "Reads") %>%
  dplyr::group_by(Taxon) %>%
  dplyr::summarise(mean = mean(Reads), variance = var(Reads), .groups = "drop")
p_disp <- ggplot(df_disp, aes(x = mean, y = variance)) +
  geom_point(alpha = 0.5) + geom_abline(intercept = 0, slope = 1) +
  scale_x_log10() + scale_y_log10() +
  labs(title = "Overdispersion (variance > mean)") + theme_bw()

## Sparsity
log10ab <- reshape2::melt(log10(1 + otu_mat_f))$value
p_sparse <- ggplot(data.frame(v = log10ab), aes(x = v)) +
  geom_histogram(bins = 80) + theme_bw() +
  labs(title = "Sparsity (log10 abundance distribution)", x = "log10(1 + reads)")

## Rarity
df_rare <- reshape2::melt(otu_mat_f) %>% dplyr::arrange(dplyr::desc(value))
df_rare$Var1 <- factor(df_rare$Var1, levels = unique(df_rare$Var1))
p_rare <- ggplot(df_rare, aes(x = Var1, y = value)) +
  geom_boxplot(outlier.size = 0.5) + theme_bw() +
  theme(axis.text.x = element_blank()) +
  labs(title = "Rarity (per-taxon abundance)",
       x = "Taxonomic group", y = "Reads") +
  scale_y_log10()

print(p_disp + p_sparse + p_rare + plot_layout(ncol = 3))

cat("\n**Rarefy or not?**\n\n",
    "We do *not* rarefy the count table for differential abundance because rarefaction throws away data ",
    "and introduces artificial uncertainty. See ",
    "[Waste Not, Want Not](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1003531) ",
    "for the canonical argument.\n")
```

## ANCOM-BC {.tabset}

ANCOM-BC fits a log-linear model with bias correction and reports per-taxon log-fold-changes,
adjusted p-values, and a structural-zero flag for taxa absent in entire groups.

```{r ancombc_genus, fig.width=12, fig.height=8}
.ancombc_res_genus <- NULL
if (.have_ancombc && length(group_levels) >= 2 && ntaxa(physeq_Genus) > 0) {
  .ancombc_res_genus <- tryCatch(
    ANCOMBC::ancombc2(data = physeq_filtered,
                      fix_formula = "Group",
                      tax_level    = "Genus",
                      p_adj_method = "BH",
                      lib_cut      = 0,
                      prv_cut      = .cfg$prevalence_frac,
                      group        = "Group",
                      struc_zero   = TRUE,
                      neg_lb       = TRUE,
                      iter_control = list(tol = 1e-5, max_iter = 100, verbose = FALSE),
                      alpha        = 0.05,
                      global       = TRUE),
    error = function(e) { cat("ANCOM-BC Genus failed: ", e$message, "\n"); NULL }
  )
}
```

### Primary Results -- Genus

```{r ancombc_genus_primary}
if (!is.null(.ancombc_res_genus)) {
  res <- .ancombc_res_genus$res
  if (!is.null(res) && nrow(res) > 0) {
    write.csv(res, file.path(.report_dir, "ancombc2_genus_primary.csv"), row.names = FALSE)
    fgcz_dt(res, "ANCOM-BC2 -- primary results (Genus)")
  }
}
```

### Global test -- Genus

```{r ancombc_genus_global}
if (!is.null(.ancombc_res_genus) && length(group_levels) > 2 &&
    !is.null(.ancombc_res_genus$res_global)) {
  res_g <- .ancombc_res_genus$res_global
  write.csv(res_g, file.path(.report_dir, "ancombc2_genus_global.csv"), row.names = FALSE)
  fgcz_dt(res_g, "ANCOM-BC2 -- global test results (Genus)")
} else if (length(group_levels) <= 2) {
  cat("Only two groups present -- ANCOM-BC global test not applicable.\n")
}
```

```{r ancombc_family, fig.width=12, fig.height=8}
.ancombc_res_family <- NULL
if (.have_ancombc && length(group_levels) >= 2 && ntaxa(physeq_Family) > 0) {
  .ancombc_res_family <- tryCatch(
    ANCOMBC::ancombc2(data = physeq_filtered,
                      fix_formula = "Group",
                      tax_level    = "Family",
                      p_adj_method = "BH",
                      lib_cut      = 0,
                      prv_cut      = .cfg$prevalence_frac,
                      group        = "Group",
                      struc_zero   = TRUE,
                      neg_lb       = TRUE,
                      iter_control = list(tol = 1e-5, max_iter = 100, verbose = FALSE),
                      alpha        = 0.05,
                      global       = TRUE),
    error = function(e) { cat("ANCOM-BC Family failed: ", e$message, "\n"); NULL }
  )
}
```

### Primary Results -- Family

```{r ancombc_family_primary}
if (!is.null(.ancombc_res_family)) {
  res <- .ancombc_res_family$res
  if (!is.null(res) && nrow(res) > 0) {
    write.csv(res, file.path(.report_dir, "ancombc2_family_primary.csv"), row.names = FALSE)
    fgcz_dt(res, "ANCOM-BC2 -- primary results (Family)")
  }
}
```

### Global test -- Family

```{r ancombc_family_global}
if (!is.null(.ancombc_res_family) && length(group_levels) > 2 &&
    !is.null(.ancombc_res_family$res_global)) {
  res_g <- .ancombc_res_family$res_global
  write.csv(res_g, file.path(.report_dir, "ancombc2_family_global.csv"), row.names = FALSE)
  fgcz_dt(res_g, "ANCOM-BC2 -- global test results (Family)")
}
```

## DESeq2 + edgeR {.tabset}

### DESeq2 (per-contrast)

```{r deseq2_run}
.deseq_reses <- NULL
.sigtab      <- NULL
tryCatch({
  Group <- get_variable(physeq_filtered, grouping)
  ds2   <- phyloseq_to_deseq2(physeq_filtered, ~ 0 + Group)
  geoMeans <- apply(counts(ds2), 1, gm_mean)
  ds2 <- estimateSizeFactors(ds2, geoMeans = geoMeans)
  ds2 <- DESeq(ds2, test = "Wald", fitType = "local")

  if (length(group_levels) >= 2) {
    comps <- combn(group_levels, 2, simplify = FALSE)
    reses <- do.call(rbind, lapply(comps, function(i) {
      a <- as.character(i[1]); b <- as.character(i[2])
      r <- as.data.frame(results(ds2, contrast = c(grouping, a, b), cooksCutoff = FALSE))
      r$log10p   <- -log10(r$pvalue)
      r$contrast <- paste0(a, "-", b)
      cbind(r, as(tax_table(physeq_filtered)[rownames(r), ], "matrix"))
    }))
    .deseq_reses <- reses[!duplicated(reses), ]
    .sigtab      <- .deseq_reses[which(.deseq_reses$padj < 0.01), ]
    write.csv(.deseq_reses, file.path(.report_dir, "deseq2_all.csv"), row.names = FALSE)
  }
}, error = function(e) cat("DESeq2 failed: ", e$message, "\n"))
```

```{r deseq2_plots, fig.width=12, fig.height=6}
if (!is.null(.deseq_reses)) {
  p2 <- ggplot(.deseq_reses, aes(x = log2FoldChange)) + geom_histogram(bins = 60) +
          facet_wrap(~contrast, ncol = 3) + theme_bw()
  p3 <- ggplot(.deseq_reses, aes(x = pvalue)) + geom_histogram(bins = 60) +
          facet_wrap(~contrast, ncol = 3) + theme_bw()
  print(p2 / p3)
}
```

### edgeR (per-contrast)

```{r edger_run}
.edger_all <- NULL
.edger_sig <- NULL
tryCatch({
  Group <- get_variable(physeq_filtered, grouping)
  design <- model.matrix(~ 0 + Group)
  dge <- as(otu_table(physeq_filtered), "matrix") + 1L
  taxonomy <- data.frame(as.data.frame(tax_table(physeq_filtered)), "matrix")
  dge <- DGEList(counts = dge, group = Group, genes = taxonomy, remove.zeros = TRUE)
  dge <- calcNormFactors(dge, method = "RLE")
  dge <- estimateGLMCommonDisp(dge, design)
  dge <- estimateGLMTrendedDisp(dge, design)
  dge <- estimateGLMTagwiseDisp(dge, design)
  fit <- glmFit(dge, design)
  comps <- combn(group_levels, 2, simplify = FALSE)
  resAll <- do.call(rbind, lapply(comps, function(i) {
    myargs <- list(paste0("Group", i[1], "-Group", i[2]), levels = design)
    my.con <- do.call(makeContrasts, myargs)
    lrt <- glmLRT(fit, contrast = my.con)
    res <- topTags(lrt, n = nrow(dge), adjust.method = "BH", sort.by = "PValue")@.Data[[1]]
    res <- tibble::rownames_to_column(res, "ASV")
    res$log10p   <- -log10(res$PValue)
    res$contrast <- paste0(i[1], "-", i[2])
    res
  }))
  .edger_all <- resAll[!duplicated(resAll), ]
  .edger_sig <- .edger_all[.edger_all$FDR <= 0.01, ]
  write.csv(.edger_all, file.path(.report_dir, "edger_all.csv"), row.names = FALSE)
}, error = function(e) cat("edgeR failed: ", e$message, "\n"))
```

```{r edger_plots, fig.width=12, fig.height=6}
if (!is.null(.edger_all)) {
  p2 <- ggplot(.edger_all, aes(x = logFC)) + geom_histogram(bins = 60) +
          facet_wrap(~contrast, ncol = 3) + theme_bw()
  p3 <- ggplot(.edger_all, aes(x = PValue)) + geom_histogram(bins = 60) +
          facet_wrap(~contrast, ncol = 3) + theme_bw()
  print(p2 / p3)
}
```

### Combined DESeq2 + edgeR

```{r combined_venn, fig.width=12, fig.height=6}
if (!is.null(.sigtab) && !is.null(.edger_sig)) {
  contrasts <- intersect(unique(.sigtab$contrast), unique(.edger_sig$contrast))
  if (length(contrasts) > 0) {
    ggs <- lapply(contrasts, function(crb) {
      ggvenn(list(DESeq2 = rownames(.sigtab[.sigtab$contrast == crb, ]),
                  edgeR  = .edger_sig$ASV[.edger_sig$contrast == crb]),
             text_size = 4) + ggtitle(crb)
    })
    print(ggarrange(plotlist = ggs, ncol = min(length(ggs), 3)))
  }

  ## Overlap table
  res_overlap <- do.call(rbind, lapply(contrasts, function(crb) {
    a <- rownames(.sigtab[.sigtab$contrast == crb, ])
    b <- .edger_sig$ASV[.edger_sig$contrast == crb]
    ov <- intersect(a, b)
    if (length(ov) == 0) return(NULL)
    data.frame(contrast = crb, ASV = ov, stringsAsFactors = FALSE)
  }))
  if (!is.null(res_overlap)) {
    write.csv(res_overlap, file.path(.report_dir, "deseq2_edger_overlap.csv"), row.names = FALSE)
    fgcz_dt(res_overlap, "ASVs significant in both DESeq2 and edgeR")
  }
} else {
  cat("Not enough significant results to build the overlap.\n")
}
```

# PICRUSt2 {.tabset}

```{r picrust_gate}
.picrust_enabled <- isTRUE(.cfg$run_picrust2) &&
                    ((exists("ko_abundance") && !is.null(ko_abundance)) ||
                     (exists("ec_abundance") && !is.null(ec_abundance)) ||
                     (exists("pathway_abundance") && !is.null(pathway_abundance)) ||
                     (exists("picrust2_ko") && !is.null(picrust2_ko)) ||
                     (exists("picrust2_ec") && !is.null(picrust2_ec)) ||
                     (exists("picrust2_pathway") && !is.null(picrust2_pathway)))
if (!.picrust_enabled) {
  cat("PICRUSt2 not requested or input objects not present -- section skipped.\n")
}

## Normalise input names to picrust2_ko / picrust2_ec / picrust2_pathway.
## Parentheses required so R sees the whole if/else-if/else as one expression.
.pi_ko   <- (if (exists("picrust2_ko")      && !is.null(picrust2_ko))      picrust2_ko
             else if (exists("ko_abundance") && !is.null(ko_abundance))    ko_abundance else NULL)
.pi_ec   <- (if (exists("picrust2_ec")      && !is.null(picrust2_ec))      picrust2_ec
             else if (exists("ec_abundance") && !is.null(ec_abundance))    ec_abundance else NULL)
.pi_path <- (if (exists("picrust2_pathway") && !is.null(picrust2_pathway)) picrust2_pathway
             else if (exists("pathway_abundance") && !is.null(pathway_abundance)) pathway_abundance else NULL)
```

```{r picrust_helpers}
.daa_aldex2 <- function(mat, groups) {
  if (!.have_aldex2) return(NULL)
  tryCatch({
    aldex_res <- ALDEx2::aldex(round(as.matrix(mat)), conditions = groups,
                               mc.samples = 64, test = "kw", effect = TRUE,
                               include.sample.summary = FALSE, denom = "all", verbose = FALSE)
    aldex_res <- tibble::rownames_to_column(as.data.frame(aldex_res), "feature")
    aldex_res
  }, error = function(e) { cat("ALDEx2 failed: ", e$message, "\n"); NULL })
}

.daa_metagenomeSeq <- function(mat, groups) {
  if (!.have_metagenomeSeq) return(NULL)
  tryCatch({
    pairs <- combn(unique(groups), 2, simplify = FALSE)
    results_list <- lapply(pairs, function(p) {
      keep <- groups %in% p
      m2   <- mat[, keep, drop = FALSE]
      g2   <- groups[keep]
      m2   <- m2[rowSums(m2) > 0, , drop = FALSE]
      if (ncol(m2) < 4) return(NULL)
      pheno <- metagenomeSeq::AnnotatedDataFrame(data.frame(Group = g2,
                                                           row.names = colnames(m2)))
      mr <- metagenomeSeq::newMRexperiment(round(m2), phenoData = pheno)
      mr <- metagenomeSeq::cumNorm(mr, p = metagenomeSeq::cumNormStat(mr))
      design <- model.matrix(~ Group, data = pData(mr))
      fit    <- metagenomeSeq::fitFeatureModel(mr, design)
      res    <- metagenomeSeq::MRcoefs(fit, number = nrow(m2))
      res$feature  <- rownames(res)
      res$contrast <- paste(p[2], "vs", p[1])
      res
    })
    do.call(rbind, results_list[!sapply(results_list, is.null)])
  }, error = function(e) { cat("metagenomeSeq failed: ", e$message, "\n"); NULL })
}

.render_picrust_panel <- function(mat, label) {
  if (is.null(mat)) { cat("No ", label, " input.\n"); return(invisible(NULL)) }
  ## Defensive: if the input still has a non-numeric first column (e.g. the
  ## "function" / "pathway" ID column from a naive read.table), hoist it to
  ## rownames so as.matrix() doesn't force everything to character.
  if (is.data.frame(mat) && ncol(mat) >= 2 && !is.numeric(mat[[1]])) {
    .ids <- as.character(mat[[1]])
    mat  <- mat[, -1, drop = FALSE]
    rownames(mat) <- make.unique(.ids)
  }
  mat <- as.matrix(mat)
  storage.mode(mat) <- "numeric"
  if (ncol(mat) < 4 || nrow(mat) < 2) {
    cat("Insufficient ", label, " data (", nrow(mat), " features x ", ncol(mat), " samples).\n", sep = "")
    return(invisible(NULL))
  }
  samp_ids <- intersect(colnames(mat), metadata$SampleID)
  if (length(samp_ids) < 4) {
    cat("Not enough samples shared with metadata for ", label, ".\n")
    return(invisible(NULL))
  }
  mat <- mat[, samp_ids, drop = FALSE]
  groups <- metadata$Group[match(samp_ids, metadata$SampleID)]

  cat("#### ALDEx2 (global, all groups)\n\n")
  ald <- .daa_aldex2(mat, groups)
  if (!is.null(ald)) {
    write.csv(ald, file.path(.report_dir, paste0("picrust2_aldex2_", label, ".csv")), row.names = FALSE)
    print(fgcz_dt(head(ald[order(ald$kw.eBH), ], 200),
                  paste("ALDEx2 -- ", label, " (top 200 by KW eBH p)")))
  }

  cat("\n\n#### metagenomeSeq (pairwise)\n\n")
  mg <- .daa_metagenomeSeq(mat, groups)
  if (!is.null(mg) && nrow(mg) > 0) {
    write.csv(mg, file.path(.report_dir, paste0("picrust2_metagenomeSeq_", label, ".csv")), row.names = FALSE)
    print(fgcz_dt(head(mg[order(mg$adjPvalues), ], 200),
                  paste("metagenomeSeq -- ", label, " (top 200 by adjusted p)")))
  }
}
```

## MetaCyc Pathways

```{r picrust_pathway, eval=.picrust_enabled, results='asis'}
.render_picrust_panel(.pi_path, "pathway")
```

## KEGG Orthologs (KO)

```{r picrust_ko, eval=.picrust_enabled, results='asis'}
.render_picrust_panel(.pi_ko, "ko")
```

## Enzyme Commission (EC)

```{r picrust_ec, eval=.picrust_enabled, results='asis'}
.render_picrust_panel(.pi_ec, "ec")
```

## ASV contributors

```{r picrust_contributors, eval=.picrust_enabled, results='asis'}
.contrib_path <- NULL
if (exists("picrust2_contrib") && !is.null(picrust2_contrib)) {
  .contrib_path <- picrust2_contrib
}
if (!is.null(.contrib_path) && length(.contrib_path) > 0 && file.exists(.contrib_path[1])) {
  contrib <- tryCatch(read.delim(.contrib_path[1], sep = "\t", check.names = FALSE),
                      error = function(e) NULL)
  if (!is.null(contrib) && nrow(contrib) > 0) {
    cat("Top 200 (function, ASV) contributors by mean abundance:\n\n")
    top <- contrib %>%
      dplyr::group_by(`function`, taxon) %>%
      dplyr::summarise(mean_taxon_function_abun = mean(taxon_function_abun, na.rm = TRUE), .groups = "drop") %>%
      dplyr::arrange(dplyr::desc(mean_taxon_function_abun)) %>%
      head(200)
    write.csv(top, file.path(.report_dir, "picrust2_top_contributors.csv"), row.names = FALSE)
    print(fgcz_dt(top, "Top 200 ASV-to-function contributors"))
  }
} else {
  cat("No stratified PICRUSt2 contribution file provided -- skipping.\n")
}
```

# Taxon-level CLR analysis {.tabset}

CLR (centered log-ratio) transforms each sample's count vector into log-ratios relative to
the geometric mean -- the natural representation for compositional sequencing data.
Per-taxon Wilcoxon tests are run on the CLR values; violin plots show the top 5 by
mean(|CLR|).

```{r clr_helpers}
clr_wilcoxon_level <- function(ps, level) {
  if (is.null(ps) || ntaxa(ps) == 0) return(NULL)
  ps_lvl <- tryCatch(tax_glom(ps, taxrank = level, NArm = TRUE), error = function(e) NULL)
  if (is.null(ps_lvl) || ntaxa(ps_lvl) == 0) return(NULL)
  mat <- as(otu_table(ps_lvl), "matrix")
  if (!taxa_are_rows(ps_lvl)) mat <- t(mat)
  tax_df <- as.data.frame(tax_table(ps_lvl), stringsAsFactors = FALSE)
  taxon_label <- tax_df[[level]]
  if (any(is.na(taxon_label) | taxon_label == "")) {
    bad <- which(is.na(taxon_label) | taxon_label == "")
    taxon_label[bad] <- paste0("unknown_", bad)
  }
  taxon_label <- make.unique(taxon_label)
  rownames(mat) <- taxon_label
  clr_mat <- clr_transform_matrix(mat)
  samp_groups <- sample_data(ps_lvl)$Group
  samp_groups <- as.character(samp_groups)
  group_pairs <- combn(unique(samp_groups), 2, simplify = FALSE)
  results <- do.call(rbind, lapply(rownames(clr_mat), function(tx) {
    do.call(rbind, lapply(group_pairs, function(p) {
      x <- clr_mat[tx, samp_groups == p[1]]
      y <- clr_mat[tx, samp_groups == p[2]]
      if (length(x) < 2 || length(y) < 2) return(NULL)
      w <- suppressWarnings(wilcox.test(x, y))
      data.frame(taxon = tx, group1 = p[1], group2 = p[2],
                 mean_clr_g1 = mean(x, na.rm = TRUE),
                 mean_clr_g2 = mean(y, na.rm = TRUE),
                 W = w$statistic, p = w$p.value,
                 stringsAsFactors = FALSE)
    }))
  }))
  if (is.null(results) || nrow(results) == 0) return(NULL)
  results$p.adj <- p.adjust(results$p, method = "BH")
  results <- results[order(results$p.adj), ]
  attr(results, "clr_matrix") <- clr_mat
  results
}

clr_top5_violin <- function(clr_mat, samp_groups, level_name, top_n = 5) {
  if (is.null(clr_mat)) return(NULL)
  mean_abs <- rowMeans(abs(clr_mat), na.rm = TRUE)
  top <- names(sort(mean_abs, decreasing = TRUE)[seq_len(min(top_n, length(mean_abs)))])
  long <- reshape2::melt(clr_mat[top, , drop = FALSE], varnames = c("Taxon", "Sample"),
                         value.name = "CLR")
  long$Group <- samp_groups[long$Sample]
  ggplot(long, aes(x = Group, y = CLR, fill = Group)) +
    geom_violin(scale = "width") +
    geom_jitter(width = 0.1, height = 0, size = 0.7, alpha = 0.6) +
    facet_wrap(~Taxon, scales = "free_y") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
    labs(title = paste("Top", length(top), level_name, "by mean(|CLR|)"))
}

render_clr_section <- function(ps, ps_label) {
  if (is.null(ps)) return(invisible(NULL))
  samp_groups <- setNames(as.character(sample_data(ps)$Group), sample_names(ps))
  for (rank in c("Species", "Genus", "Family", "Phylum")) {
    if (rank == "Species" && !has_species) next
    cat(sprintf("\n## CLR Wilcoxon -- %s (%s) {.tabset}\n\n", rank, ps_label))
    res <- clr_wilcoxon_level(ps, rank)
    if (is.null(res) || nrow(res) == 0) {
      cat("_No CLR Wilcoxon results at this rank._\n")
      next
    }
    write.csv(res, file.path(.report_dir, paste0("CLR_Wilcoxon_", rank, "_", ps_label, ".csv")),
              row.names = FALSE)
    cat("\n### Pairwise Wilcoxon\n\n")
    print(fgcz_dt(head(res, 200), paste(rank, "-- CLR Wilcoxon (", ps_label, ", top 200 by BH p)")))
    cat("\n\n### Top 5 by mean(|CLR|)\n\n")
    p <- clr_top5_violin(attr(res, "clr_matrix"), samp_groups, rank, top_n = 5)
    if (!is.null(p)) print(p)
  }
}
```

## Pre-filter CLR

```{r clr_pre, fig.width=14, fig.height=8, results='asis'}
render_clr_section(physeq, "pre")
```

## Post-filter CLR

```{r clr_post, fig.width=14, fig.height=8, results='asis'}
render_clr_section(physeq_filtered, "post")
```

# CLR composition barplots {.tabset}

```{r clr_barplots, fig.width=20, fig.height=10, results='asis'}
clr_barplot <- function(ps, level, label) {
  ps_lvl <- tryCatch(tax_glom(ps, level, NArm = TRUE), error = function(e) NULL)
  if (is.null(ps_lvl) || ntaxa(ps_lvl) == 0) return(NULL)
  mat <- as(otu_table(ps_lvl), "matrix")
  if (!taxa_are_rows(ps_lvl)) mat <- t(mat)
  rownames(mat) <- make.unique(as.character(tax_table(ps_lvl)[, level]))
  clr_mat <- clr_transform_matrix(mat)
  long <- reshape2::melt(clr_mat, varnames = c("Taxon", "Sample"), value.name = "CLR")
  long$Group <- as.character(sample_data(ps_lvl)$Group[match(long$Sample, sample_names(ps_lvl))])
  ggplot(long, aes(x = Sample, y = CLR, fill = Taxon)) +
    geom_bar(stat = "identity") +
    facet_grid(~Group, scales = "free_x") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7), legend.position = "none") +
    labs(title = paste("CLR composition --", level, "(", label, ")"),
         y = "CLR contribution")
}

for (lvl in c("Phylum", "Family", "Genus")) {
  cat(sprintf("\n## %s {.tabset}\n\n", lvl))
  cat("### Pre-filter\n\n")
  p_pre <- clr_barplot(physeq, lvl, "pre")
  if (!is.null(p_pre)) print(p_pre)
  cat("\n\n### Post-filter\n\n")
  p_post <- clr_barplot(physeq_filtered, lvl, "post")
  if (!is.null(p_post)) print(p_post)
}
if (has_species) {
  cat("\n## Species {.tabset}\n\n### Pre-filter\n\n")
  p <- clr_barplot(physeq, "Species", "pre")
  if (!is.null(p)) print(p)
  cat("\n\n### Post-filter\n\n")
  p <- clr_barplot(physeq_filtered, "Species", "post")
  if (!is.null(p)) print(p)
}
```

# CLR abundance tables export

```{r clr_export}
export_clr_for_level <- function(ps, level, label) {
  if (is.null(ps)) return(invisible(NULL))
  ps_lvl <- tryCatch(tax_glom(ps, level, NArm = TRUE), error = function(e) NULL)
  if (is.null(ps_lvl) || ntaxa(ps_lvl) == 0) return(invisible(NULL))
  mat <- as(otu_table(ps_lvl), "matrix")
  if (!taxa_are_rows(ps_lvl)) mat <- t(mat)
  rownames(mat) <- make.unique(as.character(tax_table(ps_lvl)[, level]))
  clr_mat <- clr_transform_matrix(mat)
  write.csv(as.data.frame(mat),
            file.path(.report_dir, paste0("raw_abundance_table_", tolower(level), "_", label, ".csv")))
  write.csv(as.data.frame(clr_mat),
            file.path(.report_dir, paste0("CLR_abundance_table_", tolower(level), "_", label, ".csv")))
}

for (lvl in c("Phylum", "Family", "Genus")) {
  export_clr_for_level(physeq,         lvl, "pre")
  export_clr_for_level(physeq_filtered, lvl, "post")
}
if (has_species) {
  export_clr_for_level(physeq,         "Species", "pre")
  export_clr_for_level(physeq_filtered, "Species", "post")
}
cat("Exported raw + CLR tables for Phylum, Family, Genus",
    if (has_species) ", Species" else "",
    " (pre and post) to the report directory.\n", sep = "")
```

# General taxonomic composition barplots {.tabset}

Plots include reads whose label is missing at the current rank by relabelling them with
their parent rank (e.g. `unknown Family (Phylum=Bacteroidota)`). Useful for cross-checking
against the main barplot section, which drops `NA` rows.

```{r general_barplots, fig.width=20, fig.height=12, results='asis'}
relabel_for_general <- function(ps, rank) {
  tax <- as.data.frame(tax_table(ps), stringsAsFactors = FALSE)
  parent_ranks <- switch(rank,
    "Species" = c("Genus", "Family", "Order", "Class", "Phylum", "Kingdom"),
    "Genus"   = c("Family", "Order", "Class", "Phylum", "Kingdom"),
    "Family"  = c("Order", "Class", "Phylum", "Kingdom"),
    "Phylum"  = c("Kingdom"),
    character(0))
  lab <- tax[[rank]]
  bad <- is.na(lab) | lab == ""
  if (any(bad) && length(parent_ranks) > 0) {
    for (pr in parent_ranks) {
      if (pr %in% colnames(tax)) {
        lab[bad] <- paste0("unknown ", rank, " (", pr, "=", tax[bad, pr], ")")
        bad <- is.na(lab) | lab == "" | grepl("=NA", lab)
        if (!any(bad)) break
      }
    }
  }
  tax[[rank]] <- lab
  tax_table(ps) <- tax_table(as.matrix(tax))
  ps
}

for (rank in c("Phylum", "Family", "Genus")) {
  cat(sprintf("\n## %s (Top 20) {.tabset}\n\n### Plot\n\n", rank))
  ps_lab <- relabel_for_general(physeq_filtered, rank)
  print(plot_taxa_bar_topN(ps_lab, rank, 20, paste("General Top 20", rank)))
  cat(sprintf("\n\n## %s (Top 50) {.tabset}\n\n### Plot\n\n", rank))
  print(plot_taxa_bar_topN(ps_lab, rank, 50, paste("General Top 50", rank)))
}
if (has_species) {
  cat("\n## Species (Top 20) {.tabset}\n\n### Plot\n\n")
  ps_lab <- relabel_for_general(physeq_filtered, "Species")
  print(plot_taxa_bar_topN(ps_lab, "Species", 20, "General Top 20 Species"))
  cat("\n\n## Species (Top 50) {.tabset}\n\n### Plot\n\n")
  print(plot_taxa_bar_topN(ps_lab, "Species", 50, "General Top 50 Species"))
}
```

# Interactive widgets

The following widgets were saved next to this report; download or click to open in a browser:

```{r widgets_summary, results='asis'}
widget_files <- list.files(.report_dir, pattern = "^interactive_.*\\.html$", full.names = FALSE)
if (length(widget_files) > 0) {
  for (wf in sort(widget_files)) cat(sprintf("- [%s](%s)\n", wf, wf))
} else {
  cat("_No interactive widgets were generated._\n")
}
```

# Session info

<details>
<summary>Click to expand session info</summary>
```{r session_info}
sessionInfo()
```
</details>
