---
title: "`r if (exists('reportTitle')) reportTitle else 'QIIME2 SUSHI Report'`"
output:
  html_document:
    mathjax: https://fgcz-gstore.uzh.ch/reference/mathjax.js
    highlight: pygments
    theme: sand
    code_folding: hide
    toc: yes
    toc_depth: 4
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package="ezRun")
    css: !expr system.file("templates/fgcz.css", package="ezRun")
editor_options:
  chunk_output_type: inline
---

<style>
/* fgcz.css sets h1 to 8px (used across FGCZ reports where sections start at
   h2). This report's sections are h1, so restore readable header sizes
   without touching the shared CSS. */
h1 { font-size: 1.9em; font-weight: 700; margin-top: 1.2em; }
h2 { font-size: 1.5em; font-weight: 700; margin-top: 1.0em; }
h3 { font-size: 1.2em; font-weight: 700; }
h4 { font-size: 1.05em; font-weight: 700; }
details > summary { cursor: pointer; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment    = '',
  fig.width  = 12,
  fig.height = 8,
  warning    = FALSE,
  error      = FALSE,
  message    = FALSE,
  cache      = TRUE,
  echo       = FALSE
)
```

```{r load_pkgs, include=FALSE}
## Mandatory packages -- the SUSHI app already enforces these in its renv/module
suppressMessages(suppressWarnings({
  library(phyloseq)
  library(qiime2R)
  library(microbiome)
  library(vegan)
  library(pairwiseAdonis)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(tibble)
  library(reshape2)
  library(ggplot2)
  library(ggpubr)
  library(ggrepel)
  library(patchwork)
  library(cowplot)
  library(scales)
  library(RColorBrewer)
  library(rstatix)
  library(kableExtra)
  library(DT)
  library(plotly)
  library(htmlwidgets)
  library(htmltools)
  library(ggvenn)
  library(ComplexHeatmap)
  library(DESeq2)
  library(edgeR)
}))

## Optional packages -- gracefully degrade when not available
.have_ancombc        <- requireNamespace("ANCOMBC",        quietly = TRUE)
.have_aldex2         <- requireNamespace("ALDEx2",         quietly = TRUE)
.have_ggpicrust2     <- requireNamespace("ggpicrust2",     quietly = TRUE)
.have_microbiomeMrk  <- requireNamespace("microbiomeMarker", quietly = TRUE)
.have_dada2          <- requireNamespace("dada2",          quietly = TRUE)
.have_shortread      <- requireNamespace("ShortRead",      quietly = TRUE)
.have_biostrings     <- requireNamespace("Biostrings",     quietly = TRUE)
.have_biomformat     <- requireNamespace("biomformat",     quietly = TRUE)
.have_phangorn       <- requireNamespace("phangorn",       quietly = TRUE)
.have_picante        <- requireNamespace("picante",        quietly = TRUE)
.have_gunifrac       <- requireNamespace("GUniFrac",       quietly = TRUE)
.have_enhancedvolc   <- requireNamespace("EnhancedVolcano",quietly = TRUE)
.have_scatter3d      <- requireNamespace("scatterplot3d",  quietly = TRUE)
.have_microViz       <- requireNamespace("microViz",       quietly = TRUE)
.have_ggtree         <- requireNamespace("ggtree",         quietly = TRUE)
.have_ape            <- requireNamespace("ape",            quietly = TRUE)
.have_metagenomeSeq  <- requireNamespace("metagenomeSeq",  quietly = TRUE)
.have_car            <- requireNamespace("car",            quietly = TRUE)

if (.have_biostrings) suppressMessages(suppressWarnings(library(Biostrings)))
if (.have_shortread) suppressMessages(suppressWarnings(library(ShortRead)))
if (.have_dada2)     suppressMessages(suppressWarnings(library(dada2)))
if (.have_biomformat) suppressMessages(suppressWarnings(library(biomformat)))
if (.have_phangorn)  suppressMessages(suppressWarnings(library(phangorn)))
if (.have_picante)   suppressMessages(suppressWarnings(library(picante)))
if (.have_gunifrac)  suppressMessages(suppressWarnings(library(GUniFrac)))
if (.have_ape)       suppressMessages(suppressWarnings(library(ape)))
if (.have_ggtree)    suppressMessages(suppressWarnings(library(ggtree)))
if (.have_microViz)  suppressMessages(suppressWarnings(library(microViz)))

## shiny::validate masking guard from the project memory
## (jsonlite gets loaded transitively and masks validate)
.qiime2_validate <- function(...) if (requireNamespace("shiny", quietly = TRUE)) shiny::validate(...) else invisible(NULL)
```

```{r load_passed_objects, include=FALSE}
## makeRmdReport() in ezRun (util.R) serialises every named argument it
## receives to <name>.qs2 in the working directory and then renders the Rmd
## in a fresh env. The Rmd must read those objects back in -- nothing is
## auto-bound. Pull in every .qs2 sitting next to the report.
suppressMessages(suppressWarnings(library(qs2)))
.passed_files <- list.files(".", pattern = "\\.qs2$", full.names = FALSE)
for (.f in .passed_files) {
  .nm <- sub("\\.qs2$", "", .f)
  tryCatch(
    assign(.nm, qs_read(.f), envir = knitr::knit_global()),
    error = function(e) message("Failed to qs_read ", .f, ": ", conditionMessage(e))
  )
}
suppressWarnings(rm(list = intersect(c(".passed_files", ".f", ".nm"), ls())))
```

# Configuration

```{r config}
## Pull configuration from the SUSHI `param` object passed by makeRmdReport.
## All knobs are optional: sensible defaults kick in when missing.
if (!exists("param") || is.null(param)) param <- list()

.cfg <- list(
  grouping        = if (!is.null(param$grouping) && nzchar(param$grouping)) param$grouping else "Group",
  group_cols      = if (!is.null(param$grouping) && nzchar(param$grouping)) strsplit(param$grouping, ",")[[1]] else character(0),
  samples_to_prune= if (!is.null(param$samples_to_prune)) param$samples_to_prune else character(0),
  prevalence_frac = if (!is.null(param$prevalence_frac))  param$prevalence_frac  else 0.10,
  rel_abund_pct   = if (!is.null(param$rel_abund_pct))    param$rel_abund_pct    else 0.10,
  forward_primer  = if (!is.null(param$forward_primer))   param$forward_primer   else "CCTACGGGNGGCWGCAG",
  reverse_primer  = if (!is.null(param$reverse_primer))   param$reverse_primer   else "GACTACHVGGGTATCTAATCC",
  run_picrust2    = isTRUE(param$run_picrust2),
  run_fastp       = isTRUE(param$run_fastp)
)

## Resolve output directory. The R backend calls setwdNew(basename(ResultDir))
## and renders the Rmd from inside that working directory, so getwd() is the
## canonical place to drop CSV exports. Do NOT use output$getColumn("ResultDir")
## here -- that returns the gstore destination path that doesn't exist yet at
## render time (SUSHI copies results there only after the job finishes).
.report_dir <- getwd()

## Build a final `grouping` scalar (single column carrying the Group factor inside physeq).
grouping <- "Group"

## Verbose echo for traceability
cat("Configuration:\n")
cat("  Grouping column   :", .cfg$grouping, "\n")
cat("  Group sub-columns :", paste(.cfg$group_cols, collapse = ", "), "\n")
cat("  Prevalence filter :", .cfg$prevalence_frac * 100, "% of samples\n")
cat("  Abundance filter  :", .cfg$rel_abund_pct, "% of total reads\n")
cat("  Forward primer    :", .cfg$forward_primer, "\n")
cat("  Reverse primer    :", .cfg$reverse_primer, "\n")
cat("  PICRUSt2 enabled  :", .cfg$run_picrust2, "\n")
cat("  fastp enabled     :", .cfg$run_fastp, "\n")
```

```{r helpers}
## ---------------------------------------------------------------------------
## Helper functions used across the report (defined once, reused everywhere).
## ---------------------------------------------------------------------------

## Strip square brackets, fix common SILVA quirks (UCG-, A2, group, uncultured).
clean_taxonomy <- function(ps) {
  tax <- as.data.frame(tax_table(ps), stringsAsFactors = FALSE)
  if ("Genus" %in% colnames(tax)) {
    tax$Genus <- gsub("\\[|\\]", "", tax$Genus)
    tax$Genus[tax$Genus == "uncultured uncultured"] <- "unknown uncultured"
    idx_grp <- grepl("group", tax$Genus)
    tax$Genus[idx_grp] <- str_remove(tax$Genus[idx_grp], "_group")
    tax$Genus[tax$Genus == "UCG-010"] <- "Oscillospirales UCG-010"
    if ("Family" %in% colnames(tax)) {
      idx <- grepl("^UCG", tax$Genus) & !grepl("^UCG", tax$Family) & !is.na(tax$Family)
      tax$Genus[idx] <- paste0(tax$Family[idx], "_", tax$Genus[idx])
    }
    tax$Genus[tax$Genus == "A2"] <- "Lachnospiraceae_A2"
  }
  if ("Family" %in% colnames(tax)) tax$Family <- gsub("\\[|\\]", "", tax$Family)
  tax_table(ps) <- tax_table(as.matrix(tax))
  ps
}

## NULL if no taxa survive after pruning zero-sum taxa.
safe_prune_nonzero <- function(physeq) {
  if (is.null(physeq)) return(NULL)
  keep <- taxa_sums(physeq) > 0
  if (sum(keep) == 0) return(NULL)
  prune_taxa(keep, physeq)
}

## Standard DT widget matching the rest of the FGCZ reports.
fgcz_dt <- function(x, caption = NULL, escape = TRUE, page_length = 10) {
  datatable(x,
            caption    = caption,
            escape     = escape,
            extensions = "Buttons",
            rownames   = FALSE,
            options    = list(scrollX    = TRUE,
                              dom        = "lftBip",
                              buttons    = c("copy", "csv", "excel"),
                              pageLength = page_length,
                              lengthMenu = list(c(10, 20, 100, 500, -1),
                                                c("10", "20", "100", "500", "All"))))
}

## ---------------------------------------------------------------------------
## knit_child machinery for emitting htmlwidgets (DT/plotly) inside dynamically
## generated tabsets. A print(datatable(...)) inside a results='asis' chunk
## emits NOTHING (auto-print via knit_print.htmlwidget only fires for a normal
## chunk's last expression). The reliable way to interleave markdown headers
## with widgets is to build Rmd source text and knit_child() it: the child is
## re-parsed, so headers become sections AND widgets auto-print with their JS
## dependencies. Widgets/plots are stashed in .kc_registry so the generated
## child chunks can reference them by name (avoids recomputation / quoting).
.kc_registry <- new.env(parent = emptyenv())
.kc_counter  <- 0L
## Return Rmd source lines that, when knit, auto-print `obj` in a child chunk.
## Each block gets a UNIQUE label (kc<n>) so multiple knit_child() calls don't
## collide on auto-generated "unnamed-chunk-*" labels, and cache=FALSE so the
## global cache=TRUE doesn't stale these dynamically generated chunks.
kc_block <- function(obj, chunk_opts = "fig.width=8, fig.height=6") {
  .kc_counter <<- .kc_counter + 1L
  nm <- paste0("obj", .kc_counter)
  assign(nm, obj, envir = .kc_registry)
  c(sprintf("```{r kc%d, echo=FALSE, cache=FALSE, %s}", .kc_counter, chunk_opts),
    sprintf(".kc_registry$%s", nm),
    "```", "")
}
## Render a character vector of Rmd source (markdown + kc_block output) as a
## child document and write it straight into the report.
kc_render <- function(lines) {
  cat(knitr::knit_child(text = paste(lines, collapse = "\n"), quiet = TRUE))
  cat("\n\n")
}

## Save a ggplot/plotly widget as a self-contained HTML alongside the report.
save_interactive_widget <- function(widget, fname) {
  fpath <- file.path(.report_dir, fname)
  tryCatch({
    htmlwidgets::saveWidget(widget, fpath, selfcontained = TRUE)
    fname
  }, error = function(e) { message("saveWidget failed: ", e$message); "" })
}

## Aggregate to a rank and return a long df with per-sample rel. abundance.
taxa_aggregate_rank <- function(ps, rank) {
  df <- psmelt(ps)
  df <- df[!is.na(df[[rank]]), , drop = FALSE]
  df %>%
    dplyr::group_by(Sample, Group, !!rlang::sym(rank)) %>%
    dplyr::summarise(Abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
    dplyr::group_by(Sample) %>%
    dplyr::mutate(SampleTotal  = sum(Abundance, na.rm = TRUE),
                  RelAbundance = ifelse(SampleTotal > 0, Abundance / SampleTotal, 0)) %>%
    dplyr::select(-SampleTotal) %>%
    dplyr::ungroup()
}

plot_taxa_bar_all <- function(ps, rank, ylab) {
  agg <- taxa_aggregate_rank(ps, rank)
  if (nrow(agg) == 0) return(ggplot() + theme_void() + ggtitle(paste("No", rank, "data")))
  cols <- colorRampPalette(brewer.pal(12, "Set3"))(length(unique(agg[[rank]])))
  ggplot(agg, aes(x = Sample, y = Abundance, fill = .data[[rank]])) +
    facet_grid(~Group, scales = "free_x") +
    geom_bar(stat = "identity", position = "fill") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom") +
    labs(y = ylab) +
    scale_fill_manual(values = cols)
}

plot_taxa_bar_topN <- function(ps, rank, N, ylab) {
  agg <- taxa_aggregate_rank(ps, rank)
  if (nrow(agg) == 0) return(ggplot() + theme_void() + ggtitle(paste("No", rank, "data")))
  ranking <- agg %>%
    dplyr::group_by(.data[[rank]]) %>%
    dplyr::summarise(MeanRel = mean(RelAbundance, na.rm = TRUE), .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(MeanRel))
  top <- as.character(ranking[[rank]][seq_len(min(N, nrow(ranking)))])
  agg$TaxonDisp <- ifelse(as.character(agg[[rank]]) %in% top,
                          as.character(agg[[rank]]), "Other")
  agg <- agg %>%
    dplyr::group_by(Sample, Group, TaxonDisp) %>%
    dplyr::summarise(Abundance = sum(Abundance, na.rm = TRUE), .groups = "drop")
  lvls <- c(top, "Other")
  lvls <- lvls[lvls %in% unique(agg$TaxonDisp)]
  agg$TaxonDisp <- factor(agg$TaxonDisp, levels = lvls)
  cols <- colorRampPalette(brewer.pal(12, "Set3"))(length(lvls))
  if ("Other" %in% lvls) cols[length(lvls)] <- "grey70"
  ggplot(agg, aes(x = Sample, y = Abundance, fill = TaxonDisp)) +
    facet_grid(~Group, scales = "free_x") +
    geom_bar(stat = "identity", position = "fill") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom") +
    labs(y = ylab, fill = NULL) +
    scale_fill_manual(values = cols)
}

make_taxa_wide_table <- function(ps, rank) {
  agg <- taxa_aggregate_rank(ps, rank) %>%
    dplyr::mutate(RelAbundancePct = round(RelAbundance * 100, 3)) %>%
    dplyr::select(Sample, !!rlang::sym(rank), RelAbundancePct)
  wide <- tidyr::pivot_wider(agg, names_from = Sample,
                             values_from = RelAbundancePct, values_fill = 0)
  sample_cols <- setdiff(colnames(wide), rank)
  mat <- as.matrix(wide[, sample_cols, drop = FALSE]); mode(mat) <- "numeric"
  wide$`Mean%`       <- round(rowMeans(mat, na.rm = TRUE), 3)
  wide$`Prevalence%` <- round(rowSums(mat > 0, na.rm = TRUE) / length(sample_cols) * 100, 1)
  wide <- wide[, c(rank, "Mean%", "Prevalence%", sample_cols), drop = FALSE]
  wide[order(-wide$`Mean%`), , drop = FALSE]
}

## Per-sample relative abundance transform that doesn't divide by zero.
to_rel <- function(ps) {
  transform_sample_counts(ps, function(x) {
    s <- sum(x, na.rm = TRUE)
    if (s == 0) return(x)
    x / s * 100
  })
}

plot_level_heatmap <- function(ps, level = "Genus", title = "") {
  df <- psmelt(to_rel(ps))
  df <- df[!is.na(df[[level]]), , drop = FALSE]
  if (nrow(df) == 0) return(ggplot() + theme_void() + ggtitle(paste("No", level, "data")))
  df <- df %>%
    dplyr::group_by(Sample, Group, .data[[level]]) %>%
    dplyr::summarise(Abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
    dplyr::mutate(NormAbundance = log10(Abundance + 0.01))
  ggplot(df, aes(x = Sample, y = .data[[level]], fill = NormAbundance)) +
    geom_tile() +
    facet_grid(~Group, scales = "free_x") +
    scale_fill_viridis_c(name = "Log10 Abundance") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
    labs(y = level, x = "Sample ID", title = title)
}

plot_topN_heatmap <- function(ps, N = 30, title = "") {
  ps_rel <- to_rel(ps)
  taxa_means <- taxa_sums(ps_rel) / nsamples(ps_rel)
  N <- min(N, length(taxa_means))
  topN <- names(sort(taxa_means, decreasing = TRUE)[seq_len(N)])
  df <- psmelt(ps_rel) %>%
    dplyr::mutate(
      CleanGenus  = dplyr::case_when(
        !is.na(Genus)  & Genus  != "" ~ as.character(Genus),
        !is.na(Family) & Family != "" ~ paste0("f__", Family),
        TRUE ~ as.character(OTU)
      ),
      UniqueTaxa  = paste0(CleanGenus, "_", substr(OTU, 1, 6)),
      DisplayTaxa = dplyr::if_else(OTU %in% topN, UniqueTaxa, "Remainder")
    ) %>%
    dplyr::group_by(Sample, Group, DisplayTaxa) %>%
    dplyr::summarise(Abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
    dplyr::mutate(NormAbundance = log10(Abundance + 0.01))
  ggplot(df, aes(x = Sample, y = DisplayTaxa, fill = NormAbundance)) +
    geom_tile() +
    facet_grid(~Group, scales = "free_x") +
    scale_fill_viridis_c(name = "Log10 Abundance") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(y = paste("Top", N, "Taxa"), x = "Sample ID", title = title)
}

## CLR transform helper: works column-wise (per sample) on a feature×sample matrix.
clr_transform_matrix <- function(mat) {
  mat <- as.matrix(mat) + 0.5  # pseudocount
  apply(mat, 2, function(col) log(col) - mean(log(col)))
}

## Geometric mean for DESeq2 size factor estimation on sparse OTU tables.
gm_mean <- function(x, na.rm = TRUE) exp(sum(log(x[x > 0]), na.rm = na.rm) / length(x))

## Resolve an ANCOM-BC taxon row id to a readable taxonomy.
resolve_taxon_da <- function(ids, tax_tbl) {
  tax_df <- as.data.frame(tax_tbl)
  sapply(ids, function(id) {
    if (id %in% rownames(tax_df)) {
      tx <- tax_df[id, ]
      if (!is.null(tx$Species) && !is.na(tx$Species) && tx$Species != "")
        return(paste0(tx$Genus, " ", tx$Species))
      if (!is.null(tx$Genus)   && !is.na(tx$Genus)   && tx$Genus != "")   return(tx$Genus)
      if (!is.null(tx$Family)  && !is.na(tx$Family)  && tx$Family != "")  return(paste0("f__", tx$Family))
      if (!is.null(tx$Phylum)  && !is.na(tx$Phylum)  && tx$Phylum != "")  return(paste0("p__", tx$Phylum))
    }
    id
  })
}
```

# Data Loading

```{r load_data}
## physeq is supplied by the R backend via makeRmdReport.
## Apply prevalence + relative abundance filters and glom up the standard ranks.

if (!exists("physeq")) stop("`physeq` was not passed to the Rmd by makeRmdReport.")

## Inject combined Group factor if grouping columns are supplied.
if (length(.cfg$group_cols) > 0) {
  sd <- as(sample_data(physeq), "data.frame")
  present <- intersect(.cfg$group_cols, colnames(sd))
  if (length(present) > 0) {
    sd$Group <- do.call(paste, c(lapply(present, function(col) sd[[col]]), sep = "-"))
    sample_data(physeq) <- sample_data(sd)
  } else if (!"Group" %in% colnames(sd)) {
    ## Fall back: pretend everything is one group so the rest of the report still renders.
    sd$Group <- "all"
    sample_data(physeq) <- sample_data(sd)
  }
}

physeq <- clean_taxonomy(physeq)

## Filter 1: prevalence — keep ASVs present (count > 1) in at least N% of samples.
physeq_filter <- genefilter_sample(physeq,
                                   filterfun_sample(function(x) x > 1),
                                   A = .cfg$prevalence_frac * nsamples(physeq))
physeq_prev <- prune_taxa(physeq_filter, physeq)

## Filter 2: relative abundance — drop ASVs below .cfg$rel_abund_pct of total reads.
total_reads <- sum(sample_sums(physeq_prev))
if (total_reads > 0) {
  asv_rel_abund <- taxa_sums(physeq_prev) / total_reads * 100
  physeq_filtered <- prune_taxa(asv_rel_abund >= .cfg$rel_abund_pct, physeq_prev)
} else {
  physeq_filtered <- physeq_prev
}

## Optional manual sample drop list.
if (length(.cfg$samples_to_prune) > 0) {
  drop <- intersect(sample_names(physeq_filtered), .cfg$samples_to_prune)
  if (length(drop) > 0) {
    physeq_filtered <- prune_samples(!sample_names(physeq_filtered) %in% drop, physeq_filtered)
  }
}

## Relative abundance transform of the post-filter object.
physeq_transformed <- to_rel(physeq_filtered)

## Glom to standard ranks (Species optional).
has_species <- "Species" %in% colnames(tax_table(physeq_filtered)) &&
  any(!is.na(tax_table(physeq_filtered)[, "Species"]) &
        tax_table(physeq_filtered)[, "Species"] != "")
physeq_Species <- if (has_species) tax_glom(physeq_filtered, "Species", NArm = TRUE) else NULL
physeq_Genus   <- tax_glom(physeq_filtered, "Genus",   NArm = TRUE)
physeq_Family  <- tax_glom(physeq_filtered, "Family",  NArm = TRUE)
physeq_Phylum  <- tax_glom(physeq_filtered, "Phylum",  NArm = TRUE)

## Build a metadata data.frame to use across the report.
if (!exists("metadata") || is.null(metadata)) {
  metadata <- data.frame(sample_data(physeq_filtered))
  metadata$SampleID <- rownames(metadata)
} else {
  metadata <- as.data.frame(metadata)
  ## Find the sample-id column under any of the common aliases (QIIME2 uses
  ## "sample-id"; R's read.table with default check.names mangles that to
  ## "sample.id"; qiime2R/phyloseq pipelines sometimes pass "SampleID" or "#SampleID").
  if (!"SampleID" %in% colnames(metadata)) {
    .id_aliases <- c("sample-id", "sample.id", "sampleid", "Sample.ID",
                     "Sample-ID", "#SampleID", "sample_id")
    .id_match <- intersect(.id_aliases, colnames(metadata))
    if (length(.id_match) > 0) {
      metadata$SampleID <- as.character(metadata[[.id_match[1]]])
    } else if (!is.null(rownames(metadata)) &&
               !all(rownames(metadata) == as.character(seq_len(nrow(metadata))))) {
      ## Only use rownames if they are NOT the default 1..N integer index.
      metadata$SampleID <- rownames(metadata)
    } else {
      ## Last resort: assume the first column is the sample id.
      metadata$SampleID <- as.character(metadata[[1]])
    }
  }
  if (!"Group" %in% colnames(metadata)) {
    metadata$Group <- as.character(sample_data(physeq_filtered)$Group[match(metadata$SampleID, sample_names(physeq_filtered))])
  }
}

## Levels of the grouping factor — drives all per-subgroup tabsets.
group_levels <- levels(factor(sample_data(physeq_filtered)$Group))
```

# Filtering Summary

```{r filtering_summary}
filt_tbl <- data.frame(
  Stage  = c("Raw input",
             paste0("After ", .cfg$prevalence_frac * 100, "% prevalence filter"),
             paste0("After ", .cfg$rel_abund_pct, "% abundance filter"),
             "Species-level glom",
             "Genus-level glom",
             "Family-level glom",
             "Phylum-level glom"),
  ASVs   = c(ntaxa(physeq), ntaxa(physeq_prev), ntaxa(physeq_filtered),
             if (!is.null(physeq_Species)) ntaxa(physeq_Species) else NA_integer_,
             ntaxa(physeq_Genus), ntaxa(physeq_Family), ntaxa(physeq_Phylum)),
  Samples = c(nsamples(physeq), nsamples(physeq_prev), nsamples(physeq_filtered),
              if (!is.null(physeq_Species)) nsamples(physeq_Species) else NA_integer_,
              nsamples(physeq_Genus), nsamples(physeq_Family), nsamples(physeq_Phylum))
)
kable(filt_tbl, caption = "Filtering pipeline summary") %>% kable_styling()

if (!has_species) {
  cat("\nNote: no Species-level annotations survive filtering -- all Species sub-tabs are skipped.\n")
}
```

# QC {.tabset}

## Library sizes (pre-filter) {.tabset}

### Plot

```{r lib_sizes_pre, fig.width=12, fig.height=8}
## Prefer the pre-computed `library_sizes_pre` object from the R backend
## (single source of truth) and fall back to local recomputation if missing.
.asvs_detected_pre <- colSums(as(otu_table(physeq), "matrix") > 0)[sample_names(physeq)]
if (exists("library_sizes_pre") && !is.null(library_sizes_pre)) {
  .lp <- if (is.data.frame(library_sizes_pre)) {
    if ("SampleID" %in% colnames(library_sizes_pre)) {
      stats::setNames(library_sizes_pre[[setdiff(colnames(library_sizes_pre), "SampleID")[1]]],
                      as.character(library_sizes_pre$SampleID))
    } else stats::setNames(library_sizes_pre[[1]], rownames(library_sizes_pre))
  } else library_sizes_pre
  samp_ids_pre <- intersect(names(.lp), sample_names(physeq))
  if (length(samp_ids_pre) == 0) samp_ids_pre <- sample_names(physeq)
  lib_size_pre_df <- data.frame(
    Sample = samp_ids_pre,
    Reads  = as.numeric(.lp[samp_ids_pre]),
    ASVs_detected = as.integer(.asvs_detected_pre[samp_ids_pre])
  )
} else {
  lib_size_pre_df <- data.frame(
    Sample = sample_names(physeq),
    Reads  = sample_sums(physeq),
    ASVs_detected = .asvs_detected_pre
  )
}
lib_size_pre_df$Sample <- factor(lib_size_pre_df$Sample,
                                 levels = lib_size_pre_df$Sample[order(lib_size_pre_df$Reads)])
p_lib_pre <- ggplot(lib_size_pre_df,
                    aes(x = Reads, y = Sample,
                        text = paste("Sample:", Sample, "<br>Reads:", scales::comma(Reads)))) +
  geom_col(fill = "#00A6A6") +
  geom_text(aes(label = scales::comma(Reads)), hjust = -0.1, size = 3) +
  geom_vline(xintercept = 10000, color = "red", linetype = "dashed") +
  labs(x = "Read Count", y = "", title = "Library Sizes (before filtering)") +
  theme_bw() +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15)))
ggplotly(p_lib_pre, tooltip = "text")
```

### Table

```{r lib_sizes_pre_table}
lib_pre_out <- data.frame(
  SampleID = as.character(lib_size_pre_df$Sample),
  Reads    = lib_size_pre_df$Reads,
  ASVs_detected = lib_size_pre_df$ASVs_detected
)
write.csv(lib_pre_out, file.path(.report_dir, "library_sizes_pre.csv"), row.names = FALSE)
fgcz_dt(lib_pre_out, "Library sizes (pre-filtering)")
```

## Library sizes (post-filter) {.tabset}

### Plot

```{r lib_sizes_post, fig.width=12, fig.height=8}
## Prefer the pre-computed `library_sizes_post` object from the R backend
## (single source of truth) and fall back to local recomputation if missing.
.asvs_detected_post <- colSums(as(otu_table(physeq_filtered), "matrix") > 0)[sample_names(physeq_filtered)]
if (exists("library_sizes_post") && !is.null(library_sizes_post)) {
  .lpo <- if (is.data.frame(library_sizes_post)) {
    if ("SampleID" %in% colnames(library_sizes_post)) {
      stats::setNames(library_sizes_post[[setdiff(colnames(library_sizes_post), "SampleID")[1]]],
                      as.character(library_sizes_post$SampleID))
    } else stats::setNames(library_sizes_post[[1]], rownames(library_sizes_post))
  } else library_sizes_post
  samp_ids_post <- intersect(names(.lpo), sample_names(physeq_filtered))
  if (length(samp_ids_post) == 0) samp_ids_post <- sample_names(physeq_filtered)
  lib_size_df <- data.frame(
    Sample = samp_ids_post,
    Reads  = as.numeric(.lpo[samp_ids_post]),
    ASVs_detected = as.integer(.asvs_detected_post[samp_ids_post])
  )
} else {
  lib_size_df <- data.frame(
    Sample = sample_names(physeq_filtered),
    Reads  = sample_sums(physeq_filtered),
    ASVs_detected = .asvs_detected_post
  )
}
lib_size_df$Sample <- factor(lib_size_df$Sample,
                             levels = lib_size_df$Sample[order(lib_size_df$Reads)])
p_lib <- ggplot(lib_size_df,
                aes(x = Reads, y = Sample,
                    text = paste("Sample:", Sample, "<br>Reads:", scales::comma(Reads)))) +
  geom_col(fill = "#00A6A6") +
  geom_text(aes(label = scales::comma(Reads)), hjust = -0.1, size = 3) +
  geom_vline(xintercept = 10000, color = "red", linetype = "dashed") +
  labs(x = "Read Count", y = "", title = "Library Sizes (after filtering)") +
  theme_bw() +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15)))
ggplotly(p_lib, tooltip = "text")
```

### Table

```{r lib_sizes_post_table}
lib_post_out <- data.frame(
  SampleID = as.character(lib_size_df$Sample),
  Reads    = lib_size_df$Reads,
  ASVs_detected = lib_size_df$ASVs_detected
)
write.csv(lib_post_out, file.path(.report_dir, "library_sizes_post.csv"), row.names = FALSE)
fgcz_dt(lib_post_out, "Library sizes (after filtering)")
```

## Read quality profiles

### Forward reads

```{r qc_forward, fig.width=9, fig.height=4.5}
.fnFs <- character(0)
if (exists("raw_reads_path") && length(raw_reads_path) > 0 && nzchar(raw_reads_path[1]) && dir.exists(raw_reads_path[1])) {
  .fnFs <- sort(list.files(raw_reads_path[1],
                           pattern = "_R1\\.fastq\\.gz$|_R1_001\\.fastq\\.gz$",
                           full.names = TRUE))
}
if (length(.fnFs) > 0 && .have_dada2) {
  qc_F <- dada2::plotQualityProfile(.fnFs, aggregate = TRUE) +
    ggtitle("Forward Reads -- Aggregated Quality Profile")
  print(qc_F)
} else {
  cat("Forward-read FASTQs not available -- skipping quality profile.\n")
}
```

### Reverse reads

```{r qc_reverse, fig.width=9, fig.height=4.5}
.fnRs <- character(0)
if (exists("raw_reads_path") && length(raw_reads_path) > 0 && nzchar(raw_reads_path[1]) && dir.exists(raw_reads_path[1])) {
  .fnRs <- sort(list.files(raw_reads_path[1],
                           pattern = "_R2\\.fastq\\.gz$|_R2_001\\.fastq\\.gz$",
                           full.names = TRUE))
}
if (length(.fnRs) > 0 && .have_dada2) {
  qc_R <- dada2::plotQualityProfile(.fnRs, aggregate = TRUE) +
    ggtitle("Reverse Reads -- Aggregated Quality Profile")
  print(qc_R)
} else {
  cat("Reverse-read FASTQs not available -- skipping quality profile.\n")
}
```

## Primer presence check

```{r primer_check}
.primer_check_df <- NULL
if (exists("primer_check") && !is.null(primer_check) && is.data.frame(primer_check)) {
  .primer_check_df <- primer_check
} else if (exists("dada2_rep_seqs_fasta") &&
           length(dada2_rep_seqs_fasta) > 0 && nzchar(dada2_rep_seqs_fasta[1]) &&
           file.exists(dada2_rep_seqs_fasta[1]) && .have_shortread) {
  seqs <- ShortRead::sread(ShortRead::readFasta(dada2_rep_seqs_fasta[1]))
  n_seqs <- length(seqs)
  fwd_hits <- Biostrings::vcountPattern(.cfg$forward_primer, seqs, fixed = FALSE)
  rev_hits <- Biostrings::vcountPattern(.cfg$reverse_primer, seqs, fixed = FALSE)
  rev_comp_r <- as.character(Biostrings::reverseComplement(Biostrings::DNAString(.cfg$reverse_primer)))
  rev_comp_hits <- Biostrings::vcountPattern(rev_comp_r, seqs, fixed = FALSE)
  .primer_check_df <- data.frame(
    Primer        = c("Forward (5')", "Reverse (3')", "Reverse complement"),
    Sequence      = c(.cfg$forward_primer, .cfg$reverse_primer, rev_comp_r),
    Seqs_with_hit = c(sum(fwd_hits > 0), sum(rev_hits > 0), sum(rev_comp_hits > 0)),
    Percent       = c(round(sum(fwd_hits > 0) / n_seqs * 100, 2),
                      round(sum(rev_hits > 0) / n_seqs * 100, 2),
                      round(sum(rev_comp_hits > 0) / n_seqs * 100, 2))
  )
}

if (!is.null(.primer_check_df)) {
  kable(.primer_check_df, caption = "Primer presence in representative sequences") %>% kable_styling()
  if (any(.primer_check_df$Percent > 5, na.rm = TRUE)) {
    cat("\nWARNING: primers detected in >5% of sequences -- consider revising trim parameters.\n")
  } else if (all(.primer_check_df$Percent < 1, na.rm = TRUE)) {
    cat("\nPrimers successfully removed (<1% residual).\n")
  }
} else {
  cat("No representative-sequence FASTA available -- skipping primer check.\n")
}
```

## DADA2 denoise stats

```{r dada_denoise_table}
.denoise_df <- NULL
if (exists("denoise_stats_raw") && !is.null(denoise_stats_raw)) {
  .denoise_df <- as.data.frame(denoise_stats_raw)
} else if (exists("dada2") && !is.null(dada2)) {
  .denoise_df <- tryCatch({
    if (is.list(dada2) && !is.null(dada2$data)) as.data.frame(dada2$data) else as.data.frame(dada2)
  }, error = function(e) NULL)
}
if (!is.null(.denoise_df)) {
  if (!"SampleID" %in% colnames(.denoise_df)) {
    .denoise_df <- tibble::rownames_to_column(.denoise_df, "SampleID")
  }
  write.csv(.denoise_df, file.path(.report_dir, "denoise_stats.csv"), row.names = FALSE)
  fgcz_dt(.denoise_df, "DADA2 denoising stats per sample")
} else {
  cat("DADA2 denoise stats not available.\n")
}
```

```{r dada_denoise_plot, fig.width=12, fig.height=6}
if (!is.null(.denoise_df) && "non-chimeric" %in% colnames(.denoise_df) ||
    !is.null(.denoise_df) && "non.chimeric" %in% colnames(.denoise_df) ||
    !is.null(.denoise_df) && "non_chimeric" %in% colnames(.denoise_df)) {
  ncol_name <- intersect(c("non-chimeric", "non.chimeric", "non_chimeric"), colnames(.denoise_df))[1]
  d2 <- .denoise_df
  d2$.nonchim <- suppressWarnings(as.numeric(as.character(d2[[ncol_name]])))
  d2 <- d2[!is.na(d2$.nonchim), , drop = FALSE]
  if (nrow(d2) > 0) {
    p_reads <- ggplot(d2, aes(x = SampleID, y = .nonchim, fill = SampleID)) +
      geom_col() +
      theme_bw() +
      geom_text(aes(label = scales::comma(.nonchim)), vjust = -0.3, size = 3) +
      labs(title = "Reads retained after DADA2", x = NULL, y = "Non-chimeric reads") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
    print(p_reads)
  }
}
```

## fastp QC

```{r fastp_qc_build, results='asis'}
## Split into a text (asis) chunk + a widget chunk: DT widgets are only emitted
## into the report when auto-printed from a NORMAL chunk (see asv_table_blast
## note). Here we build the data frames and emit only prose; the next chunk
## auto-prints the tables via an htmltools::tagList.
## `fastp_reports` may be a character vector / list of paths or a data.frame.
.fastp_ok <- exists("fastp_reports") && !is.null(fastp_reports) && length(fastp_reports) > 0
.fastp_summary   <- NULL
.fastp_links_df  <- NULL
.fastp_json_df   <- NULL
if (.fastp_ok) {
  .fastp_paths <- if (is.data.frame(fastp_reports)) {
    path_col <- intersect(c("path", "Path", "report", "Report", "html", "HTML",
                            "json", "JSON", "file", "File"),
                          colnames(fastp_reports))
    if (length(path_col) > 0) as.character(fastp_reports[[path_col[1]]])
    else as.character(fastp_reports[[ncol(fastp_reports)]])
  } else {
    as.character(unlist(fastp_reports))
  }
  .fastp_paths <- .fastp_paths[nzchar(.fastp_paths)]

  ## Split HTML vs JSON reports.
  .fastp_html <- .fastp_paths[grepl("\\.html?$", .fastp_paths, ignore.case = TRUE)]
  .fastp_json <- .fastp_paths[grepl("\\.json$",  .fastp_paths, ignore.case = TRUE)]

  ## Best-effort sample naming -- strip the common fastp suffix patterns.
  derive_sample <- function(p) {
    bn <- basename(p)
    bn <- sub("\\.fastp\\.(html|json)$", "", bn, ignore.case = TRUE)
    bn <- sub("_fastp\\.(html|json)$",  "", bn, ignore.case = TRUE)
    bn <- sub("\\.(html|json)$",         "", bn, ignore.case = TRUE)
    bn
  }

  ## Try to summarise reads before/after from JSON, defensively. Paths that do
  ## not resolve (e.g. relative paths after a setwd) are skipped, so we resolve
  ## against the report dir as a fallback.
  .resolve_fastp <- function(p) {
    if (file.exists(p)) return(p)
    alt <- file.path(.report_dir, p)
    if (file.exists(alt)) return(alt)
    alt2 <- file.path(.report_dir, basename(dirname(p)), basename(p))
    if (file.exists(alt2)) return(alt2)
    p
  }
  if (length(.fastp_json) > 0 && requireNamespace("jsonlite", quietly = TRUE)) {
    rows <- lapply(.fastp_json, function(jp) {
      js <- tryCatch(jsonlite::fromJSON(.resolve_fastp(jp), simplifyVector = TRUE),
                     error = function(e) NULL)
      if (is.null(js)) return(NULL)
      rb <- tryCatch(js$summary$before_filtering$total_reads, error = function(e) NA)
      ra <- tryCatch(js$summary$after_filtering$total_reads,  error = function(e) NA)
      if (is.null(rb)) rb <- NA
      if (is.null(ra)) ra <- NA
      data.frame(Sample      = derive_sample(jp),
                 Reads_before = suppressWarnings(as.numeric(rb)),
                 Reads_after  = suppressWarnings(as.numeric(ra)),
                 stringsAsFactors = FALSE)
    })
    rows <- rows[!sapply(rows, is.null)]
    if (length(rows) > 0) {
      .fastp_summary <- do.call(rbind, rows)
      .fastp_summary$Pct_kept <- ifelse(
        is.na(.fastp_summary$Reads_before) | .fastp_summary$Reads_before == 0, NA,
        round(.fastp_summary$Reads_after / .fastp_summary$Reads_before * 100, 2)
      )
    }
  }

  if (!is.null(.fastp_summary) && nrow(.fastp_summary) > 0) {
    cat("Per-sample reads before and after fastp filtering:\n\n")
    write.csv(.fastp_summary,
              file.path(.report_dir, "fastp_summary.csv"), row.names = FALSE)
  } else {
    cat("Per-sample fastp JSON summaries unavailable (JSON shape unfamiliar or jsonlite missing). ",
        "Showing report links only.\n\n", sep = "")
  }

  if (length(.fastp_html) > 0) {
    .fastp_links_df <- data.frame(
      Sample = derive_sample(.fastp_html),
      Report = sprintf('<a href="%s" target="_blank" rel="noopener">view</a>', .fastp_html),
      Path   = .fastp_html,
      stringsAsFactors = FALSE
    )
  } else if (length(.fastp_json) > 0) {
    .fastp_json_df <- data.frame(Sample = derive_sample(.fastp_json),
                                 JSON   = .fastp_json,
                                 stringsAsFactors = FALSE)
  }
} else {
  cat("fastp not run (run_fastp = false) or no reports available -- skipping fastp QC.\n")
}
```

```{r fastp_qc_tables}
## Auto-printed htmlwidgets (normal chunk). tagList emits every widget it holds
## together with their JS/CSS dependencies.
.fastp_widgets <- list()
if (!is.null(.fastp_summary) && nrow(.fastp_summary) > 0) {
  .fastp_widgets[[length(.fastp_widgets) + 1]] <- fgcz_dt(.fastp_summary, "fastp summary (per sample)")
}
if (!is.null(.fastp_links_df)) {
  .fastp_widgets[[length(.fastp_widgets) + 1]] <- DT::datatable(
    .fastp_links_df, caption = "fastp HTML reports", escape = FALSE, rownames = FALSE,
    extensions = "Buttons",
    options = list(scrollX = TRUE, dom = "lftBip", buttons = c("copy", "csv", "excel"),
                   pageLength = 10, lengthMenu = c(10, 20, 100, 200)))
} else if (!is.null(.fastp_json_df)) {
  .fastp_widgets[[length(.fastp_widgets) + 1]] <- fgcz_dt(.fastp_json_df, "fastp JSON reports")
}
if (length(.fastp_widgets) > 0) htmltools::tagList(.fastp_widgets) else invisible(NULL)
```

# ASV table with BLAST links

```{r asv_table_blast}
## NOTE: this is a NORMAL chunk (no results='asis'). An htmlwidget (DT) is only
## emitted into the report when knitr auto-prints it via knit_print.htmlwidget,
## which happens for the LAST top-level expression of a normal chunk. Under
## results='asis' a print(datatable(...)) silently produces nothing. So we
## build `asv_tax_df` here and leave the datatable as the final expression.
## Fast-path: if the R backend already passed a fully-built `asv_taxonomy_df`
## (with ASV_ID + per-sample counts + BLAST link), use it directly.
.use_passed_asv_df <- exists("asv_taxonomy_df") && !is.null(asv_taxonomy_df) &&
  is.data.frame(asv_taxonomy_df) &&
  any(c("ASV_ID", "Feature.ID", "Feature ID", "FeatureID") %in% colnames(asv_taxonomy_df))

## physeq_all is needed for downstream chunks too -- compute once here.
physeq_all <- if (exists("physeq_unfiltered") && !is.null(physeq_unfiltered)) {
  clean_taxonomy(physeq_unfiltered)
} else if (exists("table_unfiltered") && !is.null(table_unfiltered)) {
  clean_taxonomy(table_unfiltered)
} else {
  physeq
}

if (.use_passed_asv_df &&
    "ASV_ID" %in% colnames(asv_taxonomy_df) &&
    "BLAST"  %in% colnames(asv_taxonomy_df) &&
    any(sample_names(physeq) %in% colnames(asv_taxonomy_df))) {
  ## Backend provided a complete, render-ready table.
  asv_tax_df <- as.data.frame(asv_taxonomy_df, stringsAsFactors = FALSE)
  write.csv(asv_tax_df[, setdiff(colnames(asv_tax_df), "BLAST"), drop = FALSE],
            file.path(.report_dir, "asv_ids_taxonomy.csv"), row.names = FALSE)
  .asv_caption <- paste0("All ASVs (", nrow(asv_tax_df), " total).")
} else {
  asv_tax_df <- data.frame(
    ASV_ID = taxa_names(physeq_all),
    as.data.frame(tax_table(physeq_all)),
    stringsAsFactors = FALSE
  )

  ## Merge in any extra columns from the backend-provided asv_taxonomy_df
  ## (e.g. the QIIME2 raw Taxon string / Confidence) so users can still see
  ## the source taxonomy alongside the cleaned phyloseq tax_table.
  if (.use_passed_asv_df) {
    extra <- as.data.frame(asv_taxonomy_df, stringsAsFactors = FALSE)
    id_col <- intersect(c("ASV_ID", "Feature.ID", "Feature ID", "FeatureID"),
                        colnames(extra))[1]
    if (!is.null(id_col) && id_col != "ASV_ID") {
      extra$ASV_ID <- as.character(extra[[id_col]])
    }
    keep_cols <- setdiff(colnames(extra),
                         c(colnames(asv_tax_df), id_col))
    if (length(keep_cols) > 0) {
      extra <- extra[, c("ASV_ID", keep_cols), drop = FALSE]
      asv_tax_df <- dplyr::left_join(asv_tax_df, extra, by = "ASV_ID")
    }
  }

  otu_pre_mat <- as(otu_table(physeq_all), "matrix")
  if (!taxa_are_rows(physeq_all)) otu_pre_mat <- t(otu_pre_mat)
  otu_pre_mat <- otu_pre_mat[asv_tax_df$ASV_ID, , drop = FALSE]
  sample_cols <- colnames(otu_pre_mat)
  asv_tax_df$Total_reads <- as.integer(rowSums(otu_pre_mat))
  asv_tax_df$N_samples   <- as.integer(rowSums(otu_pre_mat > 0))
  asv_tax_df$In_filtered <- asv_tax_df$ASV_ID %in% taxa_names(physeq_filtered)

  ## Attach representative sequences + BLAST link if available.
  .rep_seqs <- NULL
  if (exists("rep_seqs") && !is.null(rep_seqs)) {
    .rep_seqs <- if (inherits(rep_seqs, "DNAStringSet")) rep_seqs
                 else if (is.list(rep_seqs) && !is.null(rep_seqs$data)) rep_seqs$data
                 else NULL
  } else if (exists("dada2_rep_seqs_fasta") &&
             length(dada2_rep_seqs_fasta) > 0 && file.exists(dada2_rep_seqs_fasta[1]) &&
             .have_biostrings) {
    .rep_seqs <- tryCatch(Biostrings::readDNAStringSet(dada2_rep_seqs_fasta[1]),
                          error = function(e) NULL)
  }
  if (!is.null(.rep_seqs) && length(.rep_seqs) > 0) {
    seq_lookup <- setNames(as.character(.rep_seqs), names(.rep_seqs))
    asv_tax_df$Sequence <- seq_lookup[asv_tax_df$ASV_ID]
    asv_tax_df$Length   <- nchar(asv_tax_df$Sequence)
    asv_tax_df$BLAST    <- ifelse(
      is.na(asv_tax_df$Sequence), "",
      sprintf('<a href="https://blast.ncbi.nlm.nih.gov/Blast.cgi?PROGRAM=blastn&PAGE_TYPE=BlastSearch&DATABASE=nt&QUERY=%s" target="_blank" rel="noopener">BLAST</a>',
              asv_tax_df$Sequence)
    )
  }

  ## Per-sample read counts at the end so BLAST stays visible.
  asv_tax_df[, sample_cols] <- as.data.frame(otu_pre_mat, stringsAsFactors = FALSE)

  ## CSV without the HTML BLAST column.
  write.csv(asv_tax_df[, setdiff(colnames(asv_tax_df), "BLAST")],
            file.path(.report_dir, "asv_ids_taxonomy.csv"), row.names = FALSE)

  .asv_caption <- paste0("All ASVs (", nrow(asv_tax_df),
                         " total). In_filtered=TRUE marks the ",
                         sum(asv_tax_df$In_filtered),
                         " ASVs that survived the prevalence + abundance filter.")
}

## Final expression: auto-printed as an htmlwidget (do NOT wrap in print()).
DT::datatable(asv_tax_df,
  caption = .asv_caption,
  escape  = FALSE,
  extensions = "Buttons", rownames = FALSE,
  options = list(scrollX = TRUE, dom = "lftBip",
                 buttons = c("copy", "csv", "excel"),
                 pageLength = 10, lengthMenu = c(10, 20, 100, 200)))
```

# Taxa ASVs abundance per sample {.tabset}

## Pre-filtering

```{r prevalence_pre, fig.width=15, fig.height=7}
otu_full <- as(otu_table(physeq_all), "matrix")
if (!taxa_are_rows(physeq_all)) otu_full <- t(otu_full)
asv_per_sample <- colSums(otu_full)
asv_df <- data.frame(
  SampleID      = names(asv_per_sample),
  Total_reads   = asv_per_sample,
  ASVs_detected = colSums(otu_full > 0)
)
p_pre <- ggplot(asv_df, aes(x = SampleID, y = Total_reads, fill = SampleID)) +
  geom_col() +
  theme_bw() +
  geom_text(aes(label = scales::comma(Total_reads)), vjust = -0.3, size = 3) +
  labs(title = "Total ASV abundance per sample (pre-filter)",
       x = NULL, y = "Total ASV abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
print(p_pre)
write.csv(asv_df, file.path(.report_dir, "asv_per_sample_pre.csv"), row.names = FALSE)
fgcz_dt(asv_df, "Per-sample read totals and ASV richness (pre-filter)")
```

## After filtering

```{r prevalence_post, fig.width=15, fig.height=7}
otu_filt <- as(otu_table(physeq_filtered), "matrix")
if (!taxa_are_rows(physeq_filtered)) otu_filt <- t(otu_filt)
asv_df_filt <- data.frame(
  SampleID      = colnames(otu_filt),
  Total_reads   = colSums(otu_filt),
  ASVs_detected = colSums(otu_filt > 0)
)
p_post <- ggplot(asv_df_filt, aes(x = SampleID, y = Total_reads, fill = SampleID)) +
  geom_col() +
  theme_bw() +
  geom_text(aes(label = scales::comma(Total_reads)), vjust = -0.3, size = 3) +
  labs(title = "Total ASV abundance per sample (after filtering)",
       x = NULL, y = "Total ASV abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
print(p_post)
write.csv(asv_df_filt, file.path(.report_dir, "asv_per_sample_post.csv"), row.names = FALSE)
fgcz_dt(asv_df_filt, "Per-sample read totals and ASV richness (after filter)")
```

# Taxa prevalence {.tabset}

Taxa prevalence is how widespread a taxon is across samples (independent of abundance).
The All samples tabset shows the global view; per-group sub-tabsets are generated
automatically for every level of the grouping factor.

```{r prevalence_helpers}
## Drop rows where the target rank is missing/blank/prefix-only.
## SILVA-cleaned tax_tables often carry `""`, `"NA"`, or bare `"g__"` strings,
## none of which is.na() would catch — we filter all of those.
remove_na_rank <- function(ps, rank) {
  tax <- as.data.frame(tax_table(ps), stringsAsFactors = FALSE)
  vals <- as.character(tax[[rank]])
  keep <- !is.na(vals) & nzchar(vals) & vals != "NA" &
          !grepl("^[a-z]__$", vals)
  if (sum(keep) == 0) return(NULL)
  prune_taxa(rownames(tax)[keep], ps)
}

## Custom prevalence plot: agglomerate to `rank` (dropping taxa unclassified at
## that rank), then plot prevalence (fraction of samples in which the taxon
## occurs) vs total abundance, coloured by Phylum. Returns NULL only when no
## taxon is classified at the rank (e.g. Species with empty s__ everywhere) --
## replaces microbiome::plot_taxa_prevalence, which drew sparse/empty panels.
custom_prevalence_plot <- function(ps, rank, title) {
  ps_r <- tryCatch(tax_glom(ps, rank, NArm = TRUE), error = function(e) NULL)
  if (is.null(ps_r) || ntaxa(ps_r) == 0) return(NULL)
  otu <- as(otu_table(ps_r), "matrix")
  if (!taxa_are_rows(ps_r)) otu <- t(otu)
  nsamp <- ncol(otu)
  if (nsamp == 0) return(NULL)
  tax <- as.data.frame(tax_table(ps_r), stringsAsFactors = FALSE)
  df <- data.frame(
    Taxon          = tax[[rank]],
    Phylum         = if ("Phylum" %in% colnames(tax)) tax[["Phylum"]] else NA_character_,
    Prevalence     = rowSums(otu > 0) / nsamp,
    TotalAbundance = rowSums(otu),
    stringsAsFactors = FALSE
  )
  df <- df[df$TotalAbundance > 0, , drop = FALSE]
  if (nrow(df) == 0) return(NULL)
  ## `text` is not a real ggplot aesthetic, but plotly::ggplotly(tooltip="text")
  ## reads it to build the hover label -- taxon name, prevalence, abundance.
  df$text <- paste0(rank, ": ", df$Taxon,
                    "<br>Prevalence: ", round(df$Prevalence, 3),
                    " (", round(df$Prevalence * nsamp), "/", nsamp, " samples)",
                    "<br>Total abundance: ", df$TotalAbundance,
                    "<br>Phylum: ", df$Phylum)
  ggplot(df, aes(x = TotalAbundance, y = Prevalence, color = Phylum, text = text)) +
    geom_point(size = 2.5, alpha = 0.75) +
    scale_x_log10() +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Total abundance (reads, log10)",
         y = "Prevalence (fraction of samples)", title = title) +
    theme_bw() +
    theme(legend.position = "none")
}

## Build Rmd source lines for one prevalence panel: an interactive plotly widget
## (hover shows taxon/prevalence/abundance) wrapped in kc_block so it renders
## inside a dynamically generated tabset, or an italic note when there is no
## data / no taxa classified at the rank. Returns a character vector to splice
## into a knit_child() document.
prev_lines <- function(ps, rank, title) {
  if (is.null(ps)) return(paste0("_", title, " -- no data_\n"))
  p <- tryCatch(custom_prevalence_plot(ps, rank, title), error = function(e) NULL)
  if (is.null(p)) return(paste0("_", title, " -- no taxa classified at ", rank, "._\n"))
  w <- tryCatch(
    plotly::ggplotly(p, tooltip = "text") %>% plotly::layout(showlegend = FALSE),
    error = function(e) NULL)
  ## Fall back to the static ggplot if ggplotly ever fails, so a panel is never blank.
  kc_block(if (is.null(w)) p else w, chunk_opts = "fig.width=12, fig.height=6")
}
```

```{r prev_all_dynamic, results='asis'}
## Every prevalence panel is an interactive plotly widget (hover a point to read
## its taxon, prevalence and total abundance). plotly/DT widgets do NOT auto-print
## under results='asis', so we build the whole section as Rmd source (markdown
## headers + kc_block widget chunks) and knit_child it -- the same pattern used by
## the CLR and PICRUSt2 sections. Ranks ordered Species -> Genus -> Family -> Phylum.
.prev_ranks <- c("Species", "Genus", "Family", "Phylum")

.prev_all_lines <- function(ps, stage) {
  unlist(lapply(.prev_ranks, function(rk) {
    if (rk == "Species" && !has_species) return(character(0))
    c(sprintf("### %s", rk), "",
      prev_lines(ps, rk, sprintf("%s -- All samples (%s)", rk, stage)), "")
  }))
}

kc_render(c("## Pre-filtering (all samples) {.tabset}", "",
            .prev_all_lines(physeq, "pre")))
kc_render(c("## After filtering (all samples) {.tabset}", "",
            .prev_all_lines(physeq_filtered, "post")))

## Per-group tabsets (post-filter)
.prev_grp_lines <- c("## Per-group tabsets (post-filter) {.tabset}", "")
for (grp in group_levels) {
  .keep <- as.character(sample_data(physeq_filtered)$Group) == grp
  ps_grp <- safe_prune_nonzero(prune_samples(.keep, physeq_filtered))
  if (is.null(ps_grp)) {
    .prev_grp_lines <- c(.prev_grp_lines, sprintf("### %s", grp), "", "_No data in this group._", "")
    next
  }
  .prev_grp_lines <- c(.prev_grp_lines, sprintf("### %s {.tabset}", grp), "")
  for (rk in .prev_ranks) {
    if (rk == "Species" && !has_species) next
    .prev_grp_lines <- c(.prev_grp_lines, sprintf("#### %s", rk), "",
                         prev_lines(ps_grp, rk, sprintf("%s -- %s", rk, grp)), "")
  }
}
kc_render(.prev_grp_lines)
```

# Heatmaps {.tabset}

Log10-transformed relative abundance heatmaps. Each value is `log10(rel% + 0.01)`; absent taxa render at `-2`.

## Species level (post-filter)
```{r heatmap_species_post, fig.width=20, fig.height=12, eval=has_species}
plot_level_heatmap(physeq_filtered, level = "Species", title = "Species -- Post-filter")
```

## Genus level (pre-filter)
```{r heatmap_genus_pre, fig.width=20, fig.height=12}
plot_level_heatmap(physeq, level = "Genus", title = "Genus -- Pre-filter")
```

## Genus level (post-filter)
```{r heatmap_genus_post, fig.width=20, fig.height=12}
plot_level_heatmap(physeq_filtered, level = "Genus", title = "Genus -- Post-filter")
```

## Family level (post-filter)
```{r heatmap_family_post, fig.width=20, fig.height=12}
plot_level_heatmap(physeq_filtered, level = "Family", title = "Family -- Post-filter")
```

## Phylum level (post-filter)
```{r heatmap_phylum_post, fig.width=20, fig.height=12}
plot_level_heatmap(physeq_filtered, level = "Phylum", title = "Phylum -- Post-filter")
```

## Top 30 ASVs (post-filter)
```{r heatmap_top30_post, fig.width=20, fig.height=14}
plot_topN_heatmap(physeq_filtered, N = 30, title = "Top 30 ASVs -- Post-filter")
```

## Interactive heatmaps (HTML files)

```{r heatmap_interactive, results='asis'}
items <- list(
  list("Genus -- Pre",  function() plot_level_heatmap(physeq, "Genus", "Genus -- Pre"),  "interactive_heatmap_genus_pre.html",  TRUE),
  list("Genus -- Post", function() plot_level_heatmap(physeq_filtered, "Genus", "Genus -- Post"), "interactive_heatmap_genus_post.html", TRUE),
  list("Family -- Post",function() plot_level_heatmap(physeq_filtered, "Family", "Family -- Post"), "interactive_heatmap_family_post.html", TRUE),
  list("Phylum -- Post",function() plot_level_heatmap(physeq_filtered, "Phylum", "Phylum -- Post"), "interactive_heatmap_phylum_post.html", TRUE),
  list("Top 30 -- Post",function() plot_topN_heatmap(physeq_filtered, 30, "Top 30 ASVs -- Post"), "interactive_heatmap_top30_post.html", TRUE),
  list("Species -- Post",function() plot_level_heatmap(physeq_filtered, "Species", "Species -- Post"), "interactive_heatmap_species_post.html", has_species)
)
for (it in items) {
  if (!it[[4]]) next
  tryCatch({
    widget <- ggplotly(it[[2]]())
    fn <- save_interactive_widget(widget, it[[3]])
    if (nzchar(fn)) cat(sprintf("- [%s](%s)\n", it[[1]], fn))
  }, error = function(e) cat(sprintf("- %s skipped: %s\n", it[[1]], e$message)))
}
```

# Composition barplots {.tabset}

Relative abundance per sample, faceted by Group. The All tab shows every taxon at the rank;
Top 20 / Top 50 lump the rest into "Other".

```{r barplots_setup}
.do_bar_rank <- function(rank) {
  cat(sprintf("\n### %s {.tabset}\n\n", rank))
  cat("#### All\n\n")
  print(plot_taxa_bar_all(physeq, rank, paste("Relative Abundance", rank, "(pre)")))
  print(plot_taxa_bar_all(physeq_filtered, rank, paste("Relative Abundance", rank, "(post)")))
  cat("\n\n#### Top 20\n\n")
  print(plot_taxa_bar_topN(physeq, rank, 20, paste("Top 20", rank, "(pre)")))
  print(plot_taxa_bar_topN(physeq_filtered, rank, 20, paste("Top 20", rank, "(post)")))
  cat("\n\n#### Top 50\n\n")
  print(plot_taxa_bar_topN(physeq, rank, 50, paste("Top 50", rank, "(pre)")))
  print(plot_taxa_bar_topN(physeq_filtered, rank, 50, paste("Top 50", rank, "(post)")))
}
```

## Species
```{r barplot_species, fig.width=20, fig.height=12, eval=has_species}
print(plot_taxa_bar_all(physeq_filtered, "Species", "Relative Abundance Species (post)"))
print(plot_taxa_bar_topN(physeq_filtered, "Species", 20, "Top 20 Species (post)"))
print(plot_taxa_bar_topN(physeq_filtered, "Species", 50, "Top 50 Species (post)"))
species_tbl_post <- make_taxa_wide_table(physeq_filtered, "Species")
write.csv(species_tbl_post, file.path(.report_dir, "taxa_abundance_species_post.csv"), row.names = FALSE)
fgcz_dt(species_tbl_post, "Species -- Relative abundance (%) per sample, post-filter")
```

## Genus
```{r barplot_genus, fig.width=20, fig.height=12}
print(plot_taxa_bar_all(physeq_filtered, "Genus", "Relative Abundance Genus (post)"))
print(plot_taxa_bar_topN(physeq_filtered, "Genus", 20, "Top 20 Genus (post)"))
print(plot_taxa_bar_topN(physeq_filtered, "Genus", 50, "Top 50 Genus (post)"))
genus_tbl_post <- make_taxa_wide_table(physeq_filtered, "Genus")
write.csv(genus_tbl_post, file.path(.report_dir, "taxa_abundance_genus_post.csv"), row.names = FALSE)
fgcz_dt(genus_tbl_post, "Genus -- Relative abundance (%) per sample, post-filter")
```

## Family
```{r barplot_family, fig.width=20, fig.height=12}
print(plot_taxa_bar_all(physeq_filtered, "Family", "Relative Abundance Family (post)"))
print(plot_taxa_bar_topN(physeq_filtered, "Family", 20, "Top 20 Family (post)"))
print(plot_taxa_bar_topN(physeq_filtered, "Family", 50, "Top 50 Family (post)"))
family_tbl_post <- make_taxa_wide_table(physeq_filtered, "Family")
write.csv(family_tbl_post, file.path(.report_dir, "taxa_abundance_family_post.csv"), row.names = FALSE)
fgcz_dt(family_tbl_post, "Family -- Relative abundance (%) per sample, post-filter")
```

## Phylum
```{r barplot_phylum, fig.width=20, fig.height=12}
print(plot_taxa_bar_all(physeq_filtered, "Phylum", "Relative Abundance Phylum (post)"))
print(plot_taxa_bar_topN(physeq_filtered, "Phylum", 20, "Top 20 Phylum (post)"))
phylum_tbl_post <- make_taxa_wide_table(physeq_filtered, "Phylum")
write.csv(phylum_tbl_post, file.path(.report_dir, "taxa_abundance_phylum_post.csv"), row.names = FALSE)
fgcz_dt(phylum_tbl_post, "Phylum -- Relative abundance (%) per sample, post-filter")
```

## Interactive (HTML files)

```{r barplot_interactive, results='asis'}
ranks_iv <- c("Phylum", "Family", "Genus")
if (has_species) ranks_iv <- c("Species", ranks_iv)
for (rank in ranks_iv) {
  for (state in c("pre", "post")) {
    ps_use <- if (state == "pre") physeq else physeq_filtered
    tryCatch({
      p <- plot_taxa_bar_all(ps_use, rank, paste("Relative Abundance", rank, "(", state, ")"))
      widget <- ggplotly(p)
      fn <- save_interactive_widget(widget,
                                    paste0("interactive_barplot_", tolower(rank), "_", state, ".html"))
      if (nzchar(fn)) cat(sprintf("- [%s -- %s-filter](%s)\n", rank, state, fn))
    }, error = function(e) cat(sprintf("- %s %s skipped: %s\n", rank, state, e$message)))
  }
}
```

# Phylogenetic tree

```{r phylo_tree, fig.width=15, fig.height=10}
.tree_obj <- NULL
if (exists("tree") && !is.null(tree)) {
  .tree_obj <- tree
} else if (!is.null(phy_tree(physeq_filtered, errorIfNULL = FALSE))) {
  .tree_obj <- phy_tree(physeq_filtered)
}

if (!is.null(.tree_obj)) {
  if (.have_ggtree) {
    tryCatch({
      melt_simple <- psmelt(physeq_filtered) %>%
        dplyr::filter(Abundance < 50) %>%
        dplyr::select(OTU, val = Abundance)
      p <- ggtree::ggtree(physeq_filtered, layout = "fan", open.angle = 10) +
        ggtree::geom_tippoint(mapping = aes(color = Genus), size = 1.5, show.legend = TRUE)
      print(p + ggtitle("Phylogenetic tree (ggtree)"))
    }, error = function(e) {
      cat("ggtree plot failed: ", e$message, "\n")
      if (.have_ape) ape::plot.phylo(.tree_obj, type = "fan", show.tip.label = FALSE,
                                     main = "Phylogenetic tree (ape fallback)")
    })
  } else if (.have_ape) {
    ape::plot.phylo(.tree_obj, type = "fan", show.tip.label = FALSE,
                    main = "Phylogenetic tree")
  } else {
    cat("Neither ggtree nor ape available -- skipping phylogenetic tree.\n")
  }
} else {
  cat("No tree object available -- skipping phylogenetic tree.\n")
}
```

# Venn diagram across grouping levels

```{r venn, fig.width=20, fig.height=9}
otu_pre  <- as(otu_table(physeq), "matrix")
if (!taxa_are_rows(physeq)) otu_pre <- t(otu_pre)
otu_post <- as(otu_table(physeq_filtered), "matrix")
if (!taxa_are_rows(physeq_filtered)) otu_post <- t(otu_post)

groups_post <- as.character(sample_data(physeq_filtered)$Group)
names(groups_post) <- sample_names(physeq_filtered)

make_group_sets <- function(otu_mat, samp_groups) {
  setNames(lapply(unique(samp_groups), function(g) {
    samp_in <- names(samp_groups)[samp_groups == g]
    samp_in <- intersect(samp_in, colnames(otu_mat))
    if (length(samp_in) == 0) return(character(0))
    asvs <- rowSums(otu_mat[, samp_in, drop = FALSE] >= 1) > 0
    rownames(otu_mat)[asvs]
  }), unique(samp_groups))
}

sets_post <- make_group_sets(otu_post, groups_post)

if (length(sets_post) >= 2 && length(sets_post) <= 5) {
  print(ggvenn(sets_post, set_name_size = 4, text_size = 4) +
          ggtitle(paste("Shared ASVs across", length(sets_post), "groups")))
} else if (length(sets_post) > 5) {
  m <- make_comb_mat(rapply(sets_post, na.omit, how = "replace"), mode = "distinct")
  print(UpSet(m,
              comb_order = order(comb_size(m), decreasing = TRUE),
              top_annotation = upset_top_annotation(m, add_numbers = TRUE),
              right_annotation = upset_right_annotation(m, add_numbers = TRUE)))
}

## Membership table
membership <- as.data.frame(
  do.call(cbind, lapply(sets_post, function(asvs) as.integer(rownames(otu_post) %in% asvs))),
  stringsAsFactors = FALSE
)
membership$ASV_ID    <- rownames(otu_post)
membership$N_groups  <- rowSums(membership[, names(sets_post), drop = FALSE])
write.csv(membership, file.path(.report_dir, "venn_membership_post.csv"), row.names = FALSE)
fgcz_dt(head(membership[order(-membership$N_groups), ], 200),
        "Top 200 ASVs by group-membership count (post-filter)")
```

# Alpha diversity {.tabset}

Shannon -- richness + evenness combined. Observed -- raw ASV count. Faith's PD -- phylogenetic
breadth (summed branch lengths). Rarefaction equalises sequencing depth across samples.

## Rarefaction curve

```{r rarecurve, fig.width=6, fig.height=6}
otu_tab <- t(as(otu_table(physeq_filtered), "matrix"))
if (taxa_are_rows(physeq_filtered)) otu_tab <- as(otu_table(physeq_filtered), "matrix") %>% t()
tryCatch(rarecurve(otu_tab, step = 50, cex = 0.5),
         error = function(e) cat("rarecurve failed: ", e$message, "\n"))
```

## All measures (rarefied)

```{r alpha_div_rarefied, fig.width=14, fig.height=10}
set.seed(123)
ps_clean <- prune_samples(sample_sums(physeq_filtered) > 0, physeq_filtered)
ps_clean <- prune_taxa(taxa_sums(ps_clean) > 0, ps_clean)
min_lib  <- min(sample_sums(ps_clean))

ps_rarefied <- tryCatch(rarefy_even_depth(ps_clean, sample.size = min_lib, rngseed = 123,
                                          replace = FALSE, verbose = FALSE),
                        error = function(e) NULL)

richness_rar <- NULL
if (!is.null(ps_rarefied)) {
  richness_rar <- estimate_richness(ps_rarefied,
                                    measures = c("Observed", "Chao1", "Shannon", "Simpson")) %>%
    tibble::rownames_to_column("SampleID")
  richness_rar$SampleID <- str_remove(richness_rar$SampleID, "^X")

  ## Prefer QIIME2-supplied evenness vector (Pielou's) from `evenness$data`
  ## over re-computing with microbiome::evenness.
  pielou_rar <- NULL
  if (exists("evenness") && !is.null(evenness)) {
    .ev_df <- if (is.list(evenness) && !is.null(evenness$data)) evenness$data else evenness
    .ev_df <- as.data.frame(.ev_df, stringsAsFactors = FALSE)
    if (!"SampleID" %in% colnames(.ev_df)) {
      .ev_df <- tibble::rownames_to_column(.ev_df, "SampleID")
    }
    ## QIIME2 column is usually `pielou_evenness`; rename to `pielou` to match
    ## downstream alpha_metrics intersect().
    ev_val_col <- setdiff(colnames(.ev_df), "SampleID")[1]
    if (!is.null(ev_val_col) && ev_val_col != "pielou") {
      names(.ev_df)[names(.ev_df) == ev_val_col] <- "pielou"
    }
    pielou_rar <- .ev_df[, c("SampleID", "pielou"), drop = FALSE]
  } else {
    pielou_rar <- microbiome::evenness(ps_rarefied, index = "pielou") %>%
      tibble::rownames_to_column("SampleID")
  }

  ## Prefer QIIME2-supplied observed-features vector from `richness$data` over
  ## the locally-estimated Observed column (overwrite when available).
  if (exists("richness") && !is.null(richness)) {
    .rich_df <- if (is.list(richness) && !is.null(richness$data)) richness$data else richness
    .rich_df <- as.data.frame(.rich_df, stringsAsFactors = FALSE)
    if (!"SampleID" %in% colnames(.rich_df)) {
      .rich_df <- tibble::rownames_to_column(.rich_df, "SampleID")
    }
    rich_val_col <- setdiff(colnames(.rich_df), "SampleID")[1]
    if (!is.null(rich_val_col)) {
      names(.rich_df)[names(.rich_df) == rich_val_col] <- "Observed.qiime2"
      richness_rar <- dplyr::left_join(richness_rar, .rich_df[, c("SampleID", "Observed.qiime2")],
                                       by = "SampleID")
      ## Replace Observed with QIIME2's value where available.
      ok <- !is.na(richness_rar$Observed.qiime2)
      richness_rar$Observed[ok] <- richness_rar$Observed.qiime2[ok]
      richness_rar$Observed.qiime2 <- NULL
    }
  }

  meta_rar   <- as(sample_data(ps_rarefied), "data.frame") %>%
    tibble::rownames_to_column("SampleID")
  richness_rar <- richness_rar %>%
    dplyr::left_join(pielou_rar, by = "SampleID") %>%
    dplyr::left_join(meta_rar,   by = "SampleID")
  richness_rar$Shannon.Effective <- round(exp(richness_rar$Shannon), 2)

  ## Prefer QIIME2-supplied Faith's PD vector over the picante recomputation.
  if (exists("faith_pd") && !is.null(faith_pd)) {
    .fp_df <- if (is.list(faith_pd) && !is.null(faith_pd$data)) faith_pd$data else faith_pd
    .fp_df <- as.data.frame(.fp_df, stringsAsFactors = FALSE)
    if (!"SampleID" %in% colnames(.fp_df)) {
      .fp_df <- tibble::rownames_to_column(.fp_df, "SampleID")
    }
    fp_val_col <- setdiff(colnames(.fp_df), "SampleID")[1]
    if (!is.null(fp_val_col) && fp_val_col != "Faiths.PD") {
      names(.fp_df)[names(.fp_df) == fp_val_col] <- "Faiths.PD"
    }
    richness_rar <- dplyr::left_join(richness_rar,
                                     .fp_df[, c("SampleID", "Faiths.PD")],
                                     by = "SampleID")
  } else if (.have_phangorn && !is.null(phy_tree(ps_rarefied, errorIfNULL = FALSE)) && .have_picante) {
    my_tree <- phy_tree(ps_rarefied)
    if (!ape::is.rooted(my_tree)) my_tree <- phangorn::midpoint(my_tree)
    faith_rar <- tryCatch(picante::pd(t(as.data.frame(otu_table(ps_rarefied))),
                                      my_tree, include.root = TRUE),
                          error = function(e) NULL)
    if (!is.null(faith_rar)) {
      faith_rar$SampleID <- rownames(faith_rar)
      richness_rar <- richness_rar %>%
        dplyr::left_join(faith_rar[, c("SampleID", "PD")], by = "SampleID") %>%
        dplyr::rename(Faiths.PD = PD)
    }
  }

  lev <- levels(as.factor(richness_rar$Group))
  L.pairs <- if (length(lev) >= 2) combn(seq_along(lev), 2, simplify = FALSE, FUN = function(i) lev[i]) else list()
  alpha_metrics <- intersect(c("Observed", "Chao1", "Shannon", "Simpson", "pielou", "Shannon.Effective", "Faiths.PD"),
                             colnames(richness_rar))
  alpha_plots <- lapply(alpha_metrics, function(metric) {
    p <- ggviolin(richness_rar, x = "Group", y = metric, fill = "Group",
                  add = "boxplot", add.params = list(fill = "white", outlier.shape = NA)) +
      theme_bw() + ggtitle(paste(metric, "-- Rarefied")) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    if (length(L.pairs) > 0) {
      p <- p + stat_compare_means(comparisons = L.pairs, method = "wilcox.test", label = "p.format")
    }
    p
  })
  if (length(alpha_plots) > 0) print(wrap_plots(alpha_plots, ncol = 3))
} else {
  cat("Rarefaction failed -- skipping rarefied alpha diversity.\n")
}
```

## Shannon Kruskal-Wallis

```{r kw_test}
.use_alpha <- NULL
if (!is.null(richness_rar) && "Shannon" %in% colnames(richness_rar)) {
  .use_alpha <- richness_rar
} else if (exists("shannon") && !is.null(shannon)) {
  s <- if (is.list(shannon) && !is.null(shannon$data)) shannon$data else as.data.frame(shannon)
  .use_alpha <- s %>% tibble::rownames_to_column("SampleID")
  if (!"Shannon" %in% colnames(.use_alpha) && "shannon_entropy" %in% colnames(.use_alpha)) {
    .use_alpha$Shannon <- .use_alpha$shannon_entropy
  }
  .use_alpha <- dplyr::left_join(.use_alpha, metadata, by = "SampleID")
}

if (!is.null(.use_alpha) && "Shannon" %in% colnames(.use_alpha) && "Group" %in% colnames(.use_alpha)) {
  if (length(unique(.use_alpha$Group)) >= 2) {
    kw <- kruskal.test(Shannon ~ as.factor(Group), data = .use_alpha)
    cat("Global Kruskal-Wallis chi-squared = ", round(kw$statistic, 3), "\n")
    cat("Global Kruskal-Wallis p-value     = ", signif(kw$p.value, 4), "\n\n")

    pw <- tryCatch(pairwise_wilcox_test(data = .use_alpha,
                                        formula = Shannon ~ Group,
                                        p.adjust.method = "BH"),
                   error = function(e) NULL)
    if (!is.null(pw)) {
      write.csv(pw, file.path(.report_dir, "alpha_diversity_pairwise_wilcoxon.csv"), row.names = FALSE)
      fgcz_dt(pw[, intersect(c("group1", "group2", "n1", "n2", "statistic", "p", "p.adj", "p.adj.signif"),
                              colnames(pw))],
              "Pairwise Wilcoxon (Shannon)")
    }
  } else {
    cat("Only one group level present -- skipping KW test.\n")
  }
} else {
  cat("Shannon values not available -- skipping KW test.\n")
}
```

## Faith's PD

```{r faith_pd_tab, fig.width=8, fig.height=6}
## Render a Faith's PD violin/boxplot by Group with Kruskal-Wallis, using the
## QIIME2-supplied `faith_pd` vector. Guarded so the section disappears when
## the input is missing (e.g. older runs without faith_pd_vector.qza).
if (exists("faith_pd") && !is.null(faith_pd)) {
  .fp_df_tab <- if (is.list(faith_pd) && !is.null(faith_pd$data)) faith_pd$data else faith_pd
  .fp_df_tab <- as.data.frame(.fp_df_tab, stringsAsFactors = FALSE)
  if (!"SampleID" %in% colnames(.fp_df_tab)) {
    .fp_df_tab <- tibble::rownames_to_column(.fp_df_tab, "SampleID")
  }
  fp_val_col <- setdiff(colnames(.fp_df_tab), "SampleID")[1]
  if (!is.null(fp_val_col) && fp_val_col != "Faiths.PD") {
    names(.fp_df_tab)[names(.fp_df_tab) == fp_val_col] <- "Faiths.PD"
  }
  .fp_df_tab <- dplyr::left_join(.fp_df_tab, metadata[, intersect(c("SampleID", "Group"),
                                                                 colnames(metadata)), drop = FALSE],
                                 by = "SampleID")
  .fp_df_tab <- .fp_df_tab[!is.na(.fp_df_tab$Faiths.PD) & !is.na(.fp_df_tab$Group), , drop = FALSE]

  if (nrow(.fp_df_tab) > 0 && length(unique(.fp_df_tab$Group)) >= 2) {
    write.csv(.fp_df_tab, file.path(.report_dir, "alpha_diversity_faith_pd.csv"),
              row.names = FALSE)
    lev_fp <- levels(as.factor(.fp_df_tab$Group))
    L.pairs.fp <- combn(seq_along(lev_fp), 2, simplify = FALSE,
                        FUN = function(i) lev_fp[i])
    p_fp <- ggviolin(.fp_df_tab, x = "Group", y = "Faiths.PD", fill = "Group",
                     add = "boxplot",
                     add.params = list(fill = "white", outlier.shape = NA)) +
      theme_bw() + ggtitle("Faith's PD by Group") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      ylab("Faith's PD")
    if (length(L.pairs.fp) > 0) {
      p_fp <- p_fp + stat_compare_means(comparisons = L.pairs.fp,
                                        method = "wilcox.test",
                                        label = "p.format")
    }
    print(p_fp)

    kw_fp <- kruskal.test(Faiths.PD ~ as.factor(Group), data = .fp_df_tab)
    cat("\nKruskal-Wallis (Faith's PD) chi-squared =", round(kw_fp$statistic, 3), "\n")
    cat("Kruskal-Wallis (Faith's PD) p-value     =", signif(kw_fp$p.value, 4), "\n")

    pw_fp <- tryCatch(pairwise_wilcox_test(data = .fp_df_tab,
                                           formula = Faiths.PD ~ Group,
                                           p.adjust.method = "BH"),
                      error = function(e) NULL)
    if (!is.null(pw_fp)) {
      write.csv(pw_fp,
                file.path(.report_dir, "alpha_diversity_faith_pd_pairwise_wilcoxon.csv"),
                row.names = FALSE)
      fgcz_dt(pw_fp[, intersect(c("group1", "group2", "n1", "n2", "statistic",
                                  "p", "p.adj", "p.adj.signif"), colnames(pw_fp))],
              "Pairwise Wilcoxon (Faith's PD)")
    }
  } else {
    cat("Faith's PD vector available but not enough groups -- skipping comparison.\n")
  }
} else {
  cat("Faith's PD vector not provided -- skipping (older runs without faith_pd_vector.qza).\n")
}
```

# Beta diversity {.tabset}

PERMANOVA R^2 = fraction of community variation explained by Group. Betadisper checks
homogeneity of group dispersions; significant betadisper means PERMANOVA differences may
partly reflect dispersion rather than location.

## Jaccard {.tabset}

```{r beta_jaccard_prep}
.jaccard_data <- NULL
.jaccard_mat  <- NULL
if (exists("jaccard") && !is.null(jaccard)) {
  .jaccard_data <- if (is.list(jaccard) && !is.null(jaccard$data)) jaccard$data else jaccard
}
if (exists("jaccardmatrix") && !is.null(jaccardmatrix)) {
  .jaccard_mat <- if (is.list(jaccardmatrix) && !is.null(jaccardmatrix$data)) as.matrix(jaccardmatrix$data) else as.matrix(jaccardmatrix)
}
if (is.null(.jaccard_mat)) {
  .jaccard_mat <- tryCatch(as.matrix(phyloseq::distance(physeq_filtered, method = "jaccard")),
                           error = function(e) NULL)
}
```

### PCoA

```{r jaccard_pcoa, fig.width=7, fig.height=6}
if (!is.null(.jaccard_mat)) {
  ord_j <- ordinate(physeq_filtered, "PCoA", as.dist(.jaccard_mat))
  print(plot_ordination(physeq_filtered, ord_j, color = "Group") +
          geom_point(size = 3, alpha = 0.6) + stat_ellipse() + theme_bw() +
          ggtitle("Jaccard PCoA"))
} else {
  cat("No Jaccard matrix available.\n")
}
```

### PERMANOVA

```{r jaccard_permanova}
if (!is.null(.jaccard_mat)) {
  meta2 <- metadata[metadata$SampleID %in% rownames(.jaccard_mat), , drop = FALSE]
  .ng <- length(unique(na.omit(meta2$Group)))
  if (.ng < 2) {
    cat("PERMANOVA skipped: needs >=2 Group levels (found ", .ng, ").\n", sep = "")
  } else {
    perm  <- vegan::adonis2(.jaccard_mat ~ meta2$Group, permutations = 999, method = "jaccard")
    cat("Global PERMANOVA p-value =", signif(perm$`Pr(>F)`[1], 3), "\n")
    cat("Global PERMANOVA F        =", round(perm$F[1], 2), "\n")
    if (.ng >= 3) {
      pw_jac <- pairwise.adonis(.jaccard_mat, meta2$Group, perm = 999, sim.method = "jaccard")
      fgcz_dt(pw_jac, "Pairwise PERMANOVA -- Jaccard")
    }
  }
}
```

### Betadispersion

```{r jaccard_betadisper}
if (!is.null(.jaccard_mat)) {
  meta2 <- metadata[metadata$SampleID %in% rownames(.jaccard_mat), , drop = FALSE]
  .ng <- length(unique(na.omit(meta2$Group)))
  if (.ng < 2) {
    cat("Betadispersion skipped: needs >=2 Group levels (found ", .ng, ").\n", sep = "")
  } else {
    bd <- vegan::betadisper(as.dist(.jaccard_mat), meta2$Group)
    print(anova(bd))
    print(vegan::permutest(bd, pairwise = TRUE))
  }
}
```

## Bray-Curtis {.tabset}

```{r beta_bray_prep}
.bray_data <- NULL
.bray_mat  <- NULL
if (exists("bray") && !is.null(bray)) {
  .bray_data <- if (is.list(bray) && !is.null(bray$data)) bray$data else bray
}
if (exists("braymatrix") && !is.null(braymatrix)) {
  .bray_mat <- if (is.list(braymatrix) && !is.null(braymatrix$data)) as.matrix(braymatrix$data) else as.matrix(braymatrix)
}
if (is.null(.bray_mat)) {
  .bray_mat <- tryCatch(as.matrix(phyloseq::distance(physeq_filtered, method = "bray")),
                        error = function(e) NULL)
}
```

### PCoA

```{r bray_pcoa, fig.width=7, fig.height=6}
if (!is.null(.bray_mat)) {
  ord_b <- ordinate(physeq_filtered, "PCoA", as.dist(.bray_mat))
  print(plot_ordination(physeq_filtered, ord_b, color = "Group") +
          geom_point(size = 3, alpha = 0.6) + stat_ellipse() + theme_bw() +
          ggtitle("Bray-Curtis PCoA"))
}
```

### NMDS

```{r bray_nmds, fig.width=7, fig.height=6}
if (!is.null(.bray_mat)) {
  tryCatch({
    ps_rel <- microbiome::transform(physeq_filtered, "compositional")
    p <- plot_landscape(ps_rel, "NMDS", "bray", col = grouping) +
           labs(title = "NMDS / Bray")
    print(p)
  }, error = function(e) cat("NMDS failed: ", e$message, "\n"))
}
```

### PERMANOVA

```{r bray_permanova}
if (!is.null(.bray_mat)) {
  meta2 <- metadata[metadata$SampleID %in% rownames(.bray_mat), , drop = FALSE]
  .ng <- length(unique(na.omit(meta2$Group)))
  if (.ng < 2) {
    cat("PERMANOVA skipped: needs >=2 Group levels (found ", .ng, ").\n", sep = "")
  } else {
    perm  <- vegan::adonis2(.bray_mat ~ meta2$Group, permutations = 999, method = "bray")
    cat("Global PERMANOVA p-value =", signif(perm$`Pr(>F)`[1], 3), "\n")
    cat("Global PERMANOVA F        =", round(perm$F[1], 2), "\n")
    if (.ng >= 3) {
      pw_bc <- pairwise.adonis(.bray_mat, meta2$Group, perm = 999, sim.method = "bray")
      fgcz_dt(pw_bc, "Pairwise PERMANOVA -- Bray-Curtis")
    }
  }
}
```

### Betadispersion

```{r bray_betadisper}
if (!is.null(.bray_mat)) {
  meta2 <- metadata[metadata$SampleID %in% rownames(.bray_mat), , drop = FALSE]
  .ng <- length(unique(na.omit(meta2$Group)))
  if (.ng < 2) {
    cat("Betadispersion skipped: needs >=2 Group levels (found ", .ng, ").\n", sep = "")
  } else {
    bd <- vegan::betadisper(as.dist(.bray_mat), meta2$Group)
    print(anova(bd))
    print(vegan::permutest(bd, pairwise = TRUE))
  }
}
```

## Weighted UniFrac {.tabset}

```{r beta_unifrac_prep}
.uf_mat <- NULL
if (exists("unifracmatrix") && !is.null(unifracmatrix)) {
  .uf_mat <- if (is.list(unifracmatrix) && !is.null(unifracmatrix$data)) as.matrix(unifracmatrix$data) else as.matrix(unifracmatrix)
}
if (is.null(.uf_mat) && !is.null(phy_tree(physeq_filtered, errorIfNULL = FALSE))) {
  .uf_mat <- tryCatch(as.matrix(phyloseq::distance(physeq_filtered, method = "wunifrac")),
                      error = function(e) NULL)
}

## Prefer the QIIME2-computed PCoA vectors from `unifrac$data$Vectors` for the
## scatter plot (matches the Jaccard/Bray pcoa_results pattern). Falls back to
## the local ordinate() call if the full unifrac QZA was not supplied.
.uf_vectors <- NULL
if (exists("unifrac") && !is.null(unifrac)) {
  .uf_vectors <- tryCatch({
    v <- if (is.list(unifrac) && !is.null(unifrac$data) &&
             !is.null(unifrac$data$Vectors)) unifrac$data$Vectors else NULL
    if (!is.null(v)) as.data.frame(v, stringsAsFactors = FALSE) else NULL
  }, error = function(e) NULL)
}
```

### PCoA

```{r uf_pcoa, fig.width=7, fig.height=6}
if (!is.null(.uf_vectors) && nrow(.uf_vectors) > 0) {
  ## QIIME2 pcoa_results -> Vectors: SampleID + Axis.1 + Axis.2 + ...
  uf_v <- .uf_vectors
  id_col <- intersect(c("SampleID", "Sample.ID", "Sample-ID", "sample-id",
                        "SampleId", "sample_id"), colnames(uf_v))
  if (length(id_col) == 0) {
    ## Some qiime2R versions put the id in the first column; coerce to SampleID.
    uf_v$SampleID <- as.character(uf_v[[1]])
    id_col <- "SampleID"
  } else {
    names(uf_v)[names(uf_v) == id_col[1]] <- "SampleID"
    id_col <- "SampleID"
  }
  ax_cols <- grep("^(PC|Axis\\.?)\\s*\\d+$", colnames(uf_v), value = TRUE)
  if (length(ax_cols) < 2) ax_cols <- setdiff(colnames(uf_v), "SampleID")[1:2]
  pc1 <- ax_cols[1]; pc2 <- ax_cols[2]
  uf_v <- dplyr::left_join(uf_v, metadata[, intersect(c("SampleID", "Group"),
                                                     colnames(metadata)), drop = FALSE],
                           by = "SampleID")
  print(ggplot(uf_v, aes(x = .data[[pc1]], y = .data[[pc2]], color = Group)) +
          geom_point(size = 3, alpha = 0.6) +
          stat_ellipse() +
          theme_bw() +
          labs(x = pc1, y = pc2, title = "Weighted UniFrac PCoA (QIIME2)"))
} else if (!is.null(.uf_mat)) {
  ord_u <- ordinate(physeq_filtered, "PCoA", as.dist(.uf_mat))
  print(plot_ordination(physeq_filtered, ord_u, color = "Group") +
          geom_point(size = 3, alpha = 0.6) + stat_ellipse() + theme_bw() +
          ggtitle("Weighted UniFrac PCoA"))
}
```

### PERMANOVA

```{r uf_permanova}
if (!is.null(.uf_mat)) {
  meta2 <- metadata[metadata$SampleID %in% rownames(.uf_mat), , drop = FALSE]
  .ng <- length(unique(na.omit(meta2$Group)))
  if (.ng < 2) {
    cat("PERMANOVA skipped: needs >=2 Group levels (found ", .ng, ").\n", sep = "")
  } else {
    perm  <- vegan::adonis2(.uf_mat ~ meta2$Group, permutations = 999)
    cat("Global PERMANOVA p-value =", signif(perm$`Pr(>F)`[1], 3), "\n")
    cat("Global PERMANOVA F        =", round(perm$F[1], 2), "\n")
    if (.ng >= 3) {
      pw_uf <- pairwise.adonis(.uf_mat, meta2$Group, perm = 999)
      fgcz_dt(pw_uf, "Pairwise PERMANOVA -- Weighted UniFrac")
    }
  }
}
```

### Betadispersion

```{r uf_betadisper}
if (!is.null(.uf_mat)) {
  meta2 <- metadata[metadata$SampleID %in% rownames(.uf_mat), , drop = FALSE]
  .ng <- length(unique(na.omit(meta2$Group)))
  if (.ng < 2) {
    cat("Betadispersion skipped: needs >=2 Group levels (found ", .ng, ").\n", sep = "")
  } else {
    bd <- vegan::betadisper(as.dist(.uf_mat), meta2$Group)
    print(anova(bd))
    print(vegan::permutest(bd, pairwise = TRUE))
  }
}
```

# Differential abundance {.tabset}

## Properties of taxonomic profiling data

```{r da_props, fig.width=12, fig.height=4}
otu_mat_f <- as(otu_table(physeq_filtered), "matrix")
if (!taxa_are_rows(physeq_filtered)) otu_mat_f <- t(otu_mat_f)

## Overdispersion: variance > mean
df_disp <- reshape2::melt(otu_mat_f, varnames = c("Taxon", "Sample"), value.name = "Reads") %>%
  dplyr::group_by(Taxon) %>%
  dplyr::summarise(mean = mean(Reads), variance = var(Reads), .groups = "drop")
p_disp <- ggplot(df_disp, aes(x = mean, y = variance)) +
  geom_point(alpha = 0.5) + geom_abline(intercept = 0, slope = 1) +
  scale_x_log10() + scale_y_log10() +
  labs(title = "Overdispersion (variance > mean)") + theme_bw()

## Sparsity
log10ab <- reshape2::melt(log10(1 + otu_mat_f))$value
p_sparse <- ggplot(data.frame(v = log10ab), aes(x = v)) +
  geom_histogram(bins = 80) + theme_bw() +
  labs(title = "Sparsity (log10 abundance distribution)", x = "log10(1 + reads)")

## Rarity
df_rare <- reshape2::melt(otu_mat_f) %>% dplyr::arrange(dplyr::desc(value))
df_rare$Var1 <- factor(df_rare$Var1, levels = unique(df_rare$Var1))
p_rare <- ggplot(df_rare, aes(x = Var1, y = value)) +
  geom_boxplot(outlier.size = 0.5) + theme_bw() +
  theme(axis.text.x = element_blank()) +
  labs(title = "Rarity (per-taxon abundance)",
       x = "Taxonomic group", y = "Reads") +
  scale_y_log10()

print(p_disp + p_sparse + p_rare + plot_layout(ncol = 3))

cat("\n**Rarefy or not?**\n\n",
    "We do *not* rarefy the count table for differential abundance because rarefaction throws away data ",
    "and introduces artificial uncertainty. See ",
    "[Waste Not, Want Not](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1003531) ",
    "for the canonical argument.\n")
```

## ANCOM-BC2 {.tabset}

ANCOM-BC2 (`ANCOMBC::ancombc2`) fits a log-linear model with bias correction and reports
per-taxon log-fold-changes, adjusted p-values, and a structural-zero flag for taxa absent
in entire groups. It is run at Species (when available), Genus and Family. With more than two
groups it also reports the Global test, all Pairwise contrasts, and Dunnett-type contrasts
against the reference level.

```{r ancombc2_run, message=FALSE, warning=FALSE}
## Robust ANCOM-BC2 runner. On sparse multi-group data (e.g. distinct body sites)
## every taxon can be flagged as a structural zero, which makes ancombc2 abort with
## "All taxa contain structural zeros". We use the conservative structural-zero
## definition (neg_lb = FALSE) and, if the call still fails, transparently retry with
## struc_zero = FALSE so the section is never empty. lib_cut is kept at 0 because this
## workflow supports shallow libraries -- a non-zero lib_cut could drop every sample.
.ancombc2_msg <- list()
.ancombc2_run <- function(tax_level) {
  if (!.have_ancombc || length(group_levels) < 2) return(NULL)
  ps_ok <- switch(tax_level,
                  Species = !is.null(physeq_Species) && ntaxa(physeq_Species) > 0,
                  Genus   = ntaxa(physeq_Genus)  > 0,
                  Family  = ntaxa(physeq_Family) > 0,
                  FALSE)
  if (!isTRUE(ps_ok)) return(NULL)
  multi <- length(group_levels) > 2
  base_args <- list(
    data = physeq_filtered, fix_formula = "Group", tax_level = tax_level,
    p_adj_method = "BH", lib_cut = 0, prv_cut = .cfg$prevalence_frac,
    group = "Group", neg_lb = FALSE, alpha = 0.05, pseudo_sens = FALSE,
    global = multi, pairwise = multi, dunnet = multi,
    iter_control = list(tol = 1e-2, max_iter = 20, verbose = FALSE),
    em_control   = list(tol = 1e-5, max_iter = 100))
  res <- tryCatch(do.call(ANCOMBC::ancombc2, c(base_args, list(struc_zero = TRUE))),
                  error = function(e) { .ancombc2_msg[[tax_level]] <<- e$message; NULL })
  if (is.null(res)) {
    res <- tryCatch(do.call(ANCOMBC::ancombc2, c(base_args, list(struc_zero = FALSE))),
                    error = function(e) { .ancombc2_msg[[tax_level]] <<- e$message; NULL })
    if (!is.null(res)) attr(res, "sz_fallback") <- TRUE
  }
  res
}
.ancombc2_res <- list(
  Species = if (has_species) .ancombc2_run("Species") else NULL,
  Genus   = .ancombc2_run("Genus"),
  Family  = .ancombc2_run("Family"))
```

```{r ancombc2_render, results='asis'}
## DT widgets do not auto-print under results='asis', so build the whole ANCOM-BC2
## tabset as Rmd source (headers + kc_block tables) and knit_child it.
.ancombc2_dt <- function(df, cap, csv) {
  if (is.null(df) || !is.data.frame(df) || nrow(df) == 0)
    return(paste0("_", cap, " -- no results at this rank._\n"))
  write.csv(df, file.path(.report_dir, csv), row.names = FALSE)
  kc_block(fgcz_dt(df, cap))
}
.ancombc2_lines <- character(0)
for (rk in c("Species", "Genus", "Family")) {
  if (rk == "Species" && !has_species) next
  res <- .ancombc2_res[[rk]]
  .ancombc2_lines <- c(.ancombc2_lines, sprintf("### %s {.tabset}", rk), "")
  if (is.null(res)) {
    msg <- .ancombc2_msg[[rk]]
    .ancombc2_lines <- c(.ancombc2_lines,
      paste0("_ANCOM-BC2 could not be computed at ", rk,
             if (!is.null(msg)) paste0(": ", msg) else " (insufficient data or ANCOMBC unavailable)", "._"), "")
    next
  }
  if (isTRUE(attr(res, "sz_fallback")))
    .ancombc2_lines <- c(.ancombc2_lines,
      paste0("_Note: all taxa were flagged as structural zeros at ", rk,
             ", so results below were recomputed with `struc_zero = FALSE`._"), "")
  lrk <- tolower(rk)
  .ancombc2_lines <- c(.ancombc2_lines, "#### Primary", "",
    .ancombc2_dt(res$res, paste0("ANCOM-BC2 -- primary results (", rk, ")"),
                 paste0("ancombc2_", lrk, "_primary.csv")))
  if (length(group_levels) > 2) {
    .ancombc2_lines <- c(.ancombc2_lines, "#### Global", "",
      .ancombc2_dt(res$res_global, paste0("ANCOM-BC2 -- global test (", rk, ")"),
                   paste0("ancombc2_", lrk, "_global.csv")))
    .ancombc2_lines <- c(.ancombc2_lines, "#### Pairwise", "",
      .ancombc2_dt(res$res_pair, paste0("ANCOM-BC2 -- pairwise contrasts (", rk, ")"),
                   paste0("ancombc2_", lrk, "_pairwise.csv")))
    .ancombc2_lines <- c(.ancombc2_lines, "#### Dunnett", "",
      .ancombc2_dt(res$res_dunn, paste0("ANCOM-BC2 -- Dunnett vs reference (", rk, ")"),
                   paste0("ancombc2_", lrk, "_dunnett.csv")))
  } else {
    .ancombc2_lines <- c(.ancombc2_lines, "#### Contrast note", "",
      "_Only two groups present: the primary contrast already covers the single pairwise comparison; the Global test and Dunnett contrasts are not applicable._", "")
  }
}
kc_render(.ancombc2_lines)
```

## DESeq2 + edgeR {.tabset}

### DESeq2 (per-contrast)

```{r deseq2_run}
.deseq_reses <- NULL
.sigtab      <- NULL
tryCatch({
  Group <- get_variable(physeq_filtered, grouping)
  ds2   <- phyloseq_to_deseq2(physeq_filtered, ~ 0 + Group)
  geoMeans <- apply(counts(ds2), 1, gm_mean)
  ds2 <- estimateSizeFactors(ds2, geoMeans = geoMeans)
  ds2 <- DESeq(ds2, test = "Wald", fitType = "local")

  if (length(group_levels) >= 2) {
    comps <- combn(group_levels, 2, simplify = FALSE)
    reses <- do.call(rbind, lapply(comps, function(i) {
      a <- as.character(i[1]); b <- as.character(i[2])
      r <- as.data.frame(results(ds2, contrast = c(grouping, a, b), cooksCutoff = FALSE))
      r$log10p   <- -log10(r$pvalue)
      r$contrast <- paste0(a, "-", b)
      cbind(r, as(tax_table(physeq_filtered)[rownames(r), ], "matrix"))
    }))
    .deseq_reses <- reses[!duplicated(reses), ]
    .sigtab      <- .deseq_reses[which(.deseq_reses$padj < 0.01), ]
    write.csv(.deseq_reses, file.path(.report_dir, "deseq2_all.csv"), row.names = FALSE)
  }
}, error = function(e) cat("DESeq2 failed: ", e$message, "\n"))
```

```{r deseq2_plots, fig.width=12, fig.height=6}
if (!is.null(.deseq_reses)) {
  p2 <- ggplot(.deseq_reses, aes(x = log2FoldChange)) + geom_histogram(bins = 60) +
          facet_wrap(~contrast, ncol = 3) + theme_bw()
  p3 <- ggplot(.deseq_reses, aes(x = pvalue)) + geom_histogram(bins = 60) +
          facet_wrap(~contrast, ncol = 3) + theme_bw()
  print(p2 / p3)
}
```

### edgeR (per-contrast)

```{r edger_run}
.edger_all <- NULL
.edger_sig <- NULL
tryCatch({
  Group <- get_variable(physeq_filtered, grouping)
  design <- model.matrix(~ 0 + Group)
  dge <- as(otu_table(physeq_filtered), "matrix") + 1L
  taxonomy <- data.frame(as.data.frame(tax_table(physeq_filtered)), "matrix")
  dge <- DGEList(counts = dge, group = Group, genes = taxonomy, remove.zeros = TRUE)
  dge <- calcNormFactors(dge, method = "RLE")
  dge <- estimateGLMCommonDisp(dge, design)
  dge <- estimateGLMTrendedDisp(dge, design)
  dge <- estimateGLMTagwiseDisp(dge, design)
  fit <- glmFit(dge, design)
  comps <- combn(group_levels, 2, simplify = FALSE)
  resAll <- do.call(rbind, lapply(comps, function(i) {
    myargs <- list(paste0("Group", i[1], "-Group", i[2]), levels = design)
    my.con <- do.call(makeContrasts, myargs)
    lrt <- glmLRT(fit, contrast = my.con)
    res <- topTags(lrt, n = nrow(dge), adjust.method = "BH", sort.by = "PValue")@.Data[[1]]
    res <- tibble::rownames_to_column(res, "ASV")
    res$log10p   <- -log10(res$PValue)
    res$contrast <- paste0(i[1], "-", i[2])
    res
  }))
  .edger_all <- resAll[!duplicated(resAll), ]
  .edger_sig <- .edger_all[.edger_all$FDR <= 0.01, ]
  write.csv(.edger_all, file.path(.report_dir, "edger_all.csv"), row.names = FALSE)
}, error = function(e) cat("edgeR failed: ", e$message, "\n"))
```

```{r edger_plots, fig.width=12, fig.height=6}
if (!is.null(.edger_all)) {
  p2 <- ggplot(.edger_all, aes(x = logFC)) + geom_histogram(bins = 60) +
          facet_wrap(~contrast, ncol = 3) + theme_bw()
  p3 <- ggplot(.edger_all, aes(x = PValue)) + geom_histogram(bins = 60) +
          facet_wrap(~contrast, ncol = 3) + theme_bw()
  print(p2 / p3)
}
```

### Combined DESeq2 + edgeR

```{r combined_venn, fig.width=12, fig.height=6}
if (!is.null(.sigtab) && !is.null(.edger_sig)) {
  contrasts <- intersect(unique(.sigtab$contrast), unique(.edger_sig$contrast))
  if (length(contrasts) > 0) {
    ggs <- lapply(contrasts, function(crb) {
      ggvenn(list(DESeq2 = rownames(.sigtab[.sigtab$contrast == crb, ]),
                  edgeR  = .edger_sig$ASV[.edger_sig$contrast == crb]),
             text_size = 4) + ggtitle(crb)
    })
    print(ggarrange(plotlist = ggs, ncol = min(length(ggs), 3)))
  }

  ## Overlap table
  res_overlap <- do.call(rbind, lapply(contrasts, function(crb) {
    a <- rownames(.sigtab[.sigtab$contrast == crb, ])
    b <- .edger_sig$ASV[.edger_sig$contrast == crb]
    ov <- intersect(a, b)
    if (length(ov) == 0) return(NULL)
    data.frame(contrast = crb, ASV = ov, stringsAsFactors = FALSE)
  }))
  if (!is.null(res_overlap)) {
    write.csv(res_overlap, file.path(.report_dir, "deseq2_edger_overlap.csv"), row.names = FALSE)
    fgcz_dt(res_overlap, "ASVs significant in both DESeq2 and edgeR")
  }
} else {
  cat("Not enough significant results to build the overlap.\n")
}
```

## LEfSe {.tabset}

LEfSe (Linear discriminant analysis Effect Size, via `microbiomeMarker::run_lefse`) combines a
Kruskal-Wallis test across groups with an LDA effect-size estimate to find taxa that best
characterise each group. Run at Species (when available), Genus and Family.

```{r lefse_helpers}
.lefse_msg <- list()
.run_lefse_rank <- function(rank) {
  if (!.have_microbiomeMrk || is.null(physeq_filtered) || length(group_levels) < 2) return(NULL)
  if (rank == "Species" && !has_species) return(NULL)
  tryCatch(
    microbiomeMarker::run_lefse(physeq_filtered, group = "Group", taxa_rank = rank,
                                norm = "CPM", kw_cutoff = 0.05,
                                wilcoxon_cutoff = 0.05, lda_cutoff = 2),
    error = function(e) { .lefse_msg[[rank]] <<- e$message; NULL })
}
```

```{r lefse_render, results='asis'}
.lefse_lines <- character(0)
for (rk in c("Species", "Genus", "Family")) {
  if (rk == "Species" && !has_species) next
  mm  <- .run_lefse_rank(rk)
  ## marker_table is an S4 class that *contains* data.frame -- coerce via as(.,"data.frame").
  mt  <- if (!is.null(mm)) tryCatch(as(microbiomeMarker::marker_table(mm), "data.frame"),
                                    error = function(e) NULL) else NULL
  if (is.null(mt) || nrow(mt) == 0) {
    msg <- .lefse_msg[[rk]]
    .lefse_lines <- c(.lefse_lines, sprintf("### %s", rk), "",
      paste0("_No significant LEfSe markers at ", rk,
             if (!is.null(msg)) paste0(" (", msg, ")") else "", "._"), "")
    next
  }
  write.csv(mt, file.path(.report_dir, paste0("lefse_", tolower(rk), ".csv")), row.names = FALSE)
  p <- tryCatch(microbiomeMarker::plot_ef_bar(mm), error = function(e) NULL)
  .lefse_lines <- c(.lefse_lines, sprintf("### %s {.tabset}", rk), "")
  if (!is.null(p))
    .lefse_lines <- c(.lefse_lines, "#### Effect size (LDA)", "",
                      kc_block(p, chunk_opts = "fig.width=10, fig.height=7"))
  .lefse_lines <- c(.lefse_lines, "#### Marker table", "",
                    kc_block(fgcz_dt(mt, paste0("LEfSe markers (", rk, ")"))))
}
if (!.have_microbiomeMrk)
  .lefse_lines <- "_microbiomeMarker package not available -- LEfSe skipped._"
kc_render(.lefse_lines)
```

## ALDEx2 (taxonomic) {.tabset}

ALDEx2 on the taxonomic count tables (in addition to its use on the PICRUSt2 functional
profiles). For more than two groups it uses the Kruskal-Wallis / GLM test; for two groups the
Welch/Wilcoxon test with effect sizes. Run at Species (when available), Genus and Family.

```{r aldex_taxa_helpers}
.da_rank_ps      <- list(Species = physeq_Species, Genus = physeq_Genus, Family = physeq_Family)
.aldex_taxa_msg  <- list()

## Readable per-taxon labels for an agglomerated phyloseq at `rank`.
.rank_labels <- function(ps, rank) {
  tax <- as.data.frame(tax_table(ps), stringsAsFactors = FALSE)
  lab <- as.character(tax[[rank]])
  blank <- is.na(lab) | !nzchar(lab)
  lab[blank] <- rownames(tax)[blank]
  make.unique(lab)
}

.aldex_taxa_df <- function(rank) {
  ps <- .da_rank_ps[[rank]]
  if (!.have_aldex2 || is.null(ps) || ntaxa(ps) < 2 || length(group_levels) < 2) return(NULL)
  otu <- as(otu_table(ps), "matrix"); if (!taxa_are_rows(ps)) otu <- t(otu)
  groups <- as.character(sample_data(ps)$Group)
  test <- if (length(unique(groups)) > 2) "kw" else "t"
  res <- tryCatch(
    ALDEx2::aldex(round(otu), conditions = groups, mc.samples = 128,
                  test = test, effect = (test == "t"), denom = "all", verbose = FALSE),
    error = function(e) { .aldex_taxa_msg[[rank]] <<- e$message; NULL })
  if (is.null(res)) return(NULL)
  res <- as.data.frame(res)
  lab <- .rank_labels(ps, rank)
  res$Taxon <- lab[match(rownames(res), rownames(otu))]
  res[, c("Taxon", setdiff(colnames(res), "Taxon")), drop = FALSE]
}

## Bar chart of the top features by ALDEx2 significance (works for kw and t tests).
.aldex_taxa_plot <- function(df) {
  pcol <- intersect(c("kw.eBH", "we.eBH", "wi.eBH", "glm.eBH"), colnames(df))
  if (length(pcol) == 0) return(NULL)
  pcol <- pcol[1]
  d <- df[order(df[[pcol]]), , drop = FALSE]
  d <- head(d, 20)
  d$score <- -log10(pmax(d[[pcol]], .Machine$double.xmin))
  d$Taxon <- factor(d$Taxon, levels = rev(unique(d$Taxon)))
  ggplot(d, aes(x = Taxon, y = score)) +
    geom_col(fill = "steelblue") + coord_flip() +
    labs(x = NULL, y = paste0("-log10(", pcol, ")"),
         title = "Top 20 features by ALDEx2 significance") + theme_bw()
}
```

```{r aldex_taxa_render, results='asis'}
.aldex_taxa_lines <- character(0)
for (rk in c("Species", "Genus", "Family")) {
  if (rk == "Species" && !has_species) next
  df <- .aldex_taxa_df(rk)
  if (is.null(df) || nrow(df) == 0) {
    msg <- .aldex_taxa_msg[[rk]]
    .aldex_taxa_lines <- c(.aldex_taxa_lines, sprintf("### %s", rk), "",
      paste0("_ALDEx2 produced no result at ", rk,
             if (!is.null(msg)) paste0(" (", msg, ")") else "", "._"), "")
    next
  }
  write.csv(df, file.path(.report_dir, paste0("aldex2_taxa_", tolower(rk), ".csv")), row.names = FALSE)
  .aldex_taxa_lines <- c(.aldex_taxa_lines, sprintf("### %s {.tabset}", rk), "")
  p <- .aldex_taxa_plot(df)
  if (!is.null(p))
    .aldex_taxa_lines <- c(.aldex_taxa_lines, "#### Significance plot", "",
                           kc_block(p, chunk_opts = "fig.width=10, fig.height=7"))
  .aldex_taxa_lines <- c(.aldex_taxa_lines, "#### Table", "",
                         kc_block(fgcz_dt(df, paste0("ALDEx2 (", rk, ")"))))
}
if (!.have_aldex2)
  .aldex_taxa_lines <- "_ALDEx2 package not available -- taxonomic ALDEx2 skipped._"
kc_render(.aldex_taxa_lines)
```

# PICRUSt2 {.tabset}

```{r picrust_gate}
.picrust_enabled <- isTRUE(.cfg$run_picrust2) &&
                    ((exists("ko_abundance") && !is.null(ko_abundance)) ||
                     (exists("ec_abundance") && !is.null(ec_abundance)) ||
                     (exists("pathway_abundance") && !is.null(pathway_abundance)) ||
                     (exists("picrust2_ko") && !is.null(picrust2_ko)) ||
                     (exists("picrust2_ec") && !is.null(picrust2_ec)) ||
                     (exists("picrust2_pathway") && !is.null(picrust2_pathway)))
if (!.picrust_enabled) {
  cat("PICRUSt2 not requested or input objects not present -- section skipped.\n")
}

## Normalise input names to picrust2_ko / picrust2_ec / picrust2_pathway.
## Parentheses required so R sees the whole if/else-if/else as one expression.
.pi_ko   <- (if (exists("picrust2_ko")      && !is.null(picrust2_ko))      picrust2_ko
             else if (exists("ko_abundance") && !is.null(ko_abundance))    ko_abundance else NULL)
.pi_ec   <- (if (exists("picrust2_ec")      && !is.null(picrust2_ec))      picrust2_ec
             else if (exists("ec_abundance") && !is.null(ec_abundance))    ec_abundance else NULL)
.pi_path <- (if (exists("picrust2_pathway") && !is.null(picrust2_pathway)) picrust2_pathway
             else if (exists("pathway_abundance") && !is.null(pathway_abundance)) pathway_abundance else NULL)
```

```{r picrust_helpers}
.daa_aldex2 <- function(mat, groups) {
  if (!.have_aldex2) return(NULL)
  tryCatch({
    aldex_res <- ALDEx2::aldex(round(as.matrix(mat)), conditions = groups,
                               mc.samples = 64, test = "kw", effect = TRUE,
                               include.sample.summary = FALSE, denom = "all", verbose = FALSE)
    aldex_res <- tibble::rownames_to_column(as.data.frame(aldex_res), "feature")
    aldex_res
  }, error = function(e) { cat("ALDEx2 failed: ", e$message, "\n"); NULL })
}

.daa_metagenomeSeq <- function(mat, groups) {
  if (!.have_metagenomeSeq) return(NULL)
  tryCatch({
    pairs <- combn(unique(groups), 2, simplify = FALSE)
    results_list <- lapply(pairs, function(p) {
      keep <- groups %in% p
      m2   <- mat[, keep, drop = FALSE]
      g2   <- groups[keep]
      m2   <- m2[rowSums(m2) > 0, , drop = FALSE]
      if (ncol(m2) < 4) return(NULL)
      ## AnnotatedDataFrame / pData are exported by Biobase, not metagenomeSeq.
      pheno <- Biobase::AnnotatedDataFrame(data.frame(Group = g2,
                                                      row.names = colnames(m2)))
      mr <- metagenomeSeq::newMRexperiment(round(m2), phenoData = pheno)
      mr <- metagenomeSeq::cumNorm(mr, p = metagenomeSeq::cumNormStat(mr))
      design <- model.matrix(~ Group, data = Biobase::pData(mr))
      fit    <- metagenomeSeq::fitFeatureModel(mr, design)
      res    <- metagenomeSeq::MRcoefs(fit, number = nrow(m2))
      res$feature  <- rownames(res)
      res$contrast <- paste(p[2], "vs", p[1])
      res
    })
    do.call(rbind, results_list[!sapply(results_list, is.null)])
  }, error = function(e) { cat("metagenomeSeq failed: ", e$message, "\n"); NULL })
}

## ---- Functional abundance heatmap (top-N features) ----
## Per-sample CLR (features x samples): log(x+1) minus the per-sample mean-log.
.picrust_clr <- function(m) {
  m <- m + 1
  sweep(log(m), 2, colMeans(log(m)), "-")
}
## Clustered-ish abundance heatmap: pick the top-N features by total abundance,
## then colour by per-feature Z-score (RAW) or per-sample CLR, faceted by group.
.picrust_heatmap <- function(mat, groups, topN, mode = c("raw", "clr"), label = "") {
  mode <- match.arg(mode)
  if (nrow(mat) < 2 || ncol(mat) < 2) return(NULL)
  keep <- head(order(rowSums(mat, na.rm = TRUE), decreasing = TRUE), min(topN, nrow(mat)))
  m <- mat[keep, , drop = FALSE]
  if (mode == "clr") {
    m <- .picrust_clr(m); fillname <- "CLR"
  } else {
    m <- t(scale(t(m))); m[is.na(m)] <- 0; fillname <- "Z-score"
  }
  ord <- rownames(m)                              # keep abundance ordering
  df  <- reshape2::melt(as.matrix(m), varnames = c("feature", "SampleID"), value.name = "val")
  df$feature  <- factor(df$feature, levels = rev(ord))
  df$Group    <- groups[match(as.character(df$SampleID), colnames(mat))]
  ggplot(df, aes(x = SampleID, y = feature, fill = val)) +
    geom_tile() +
    facet_grid(. ~ Group, scales = "free_x", space = "free_x") +
    scale_fill_gradient2(low = "navy", mid = "white", high = "firebrick",
                         midpoint = 0, name = fillname) +
    labs(x = NULL, y = NULL,
         title = paste0(label, " -- top ", nrow(m), " features (", toupper(mode), ")")) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
          axis.text.y = element_text(size = 6),
          panel.grid  = element_blank())
}

## Top-feature significance bar chart from an ALDEx2 result frame.
.picrust_aldex_bar <- function(ald, label) {
  pcol <- intersect(c("kw.eBH", "we.eBH", "wi.eBH", "glm.eBH"), colnames(ald))
  if (length(pcol) == 0) return(NULL)
  pcol <- pcol[1]
  d <- head(ald[order(ald[[pcol]]), , drop = FALSE], 20)
  if (nrow(d) == 0) return(NULL)
  d$score <- -log10(pmax(d[[pcol]], .Machine$double.xmin))
  d$feature <- factor(d$feature, levels = rev(unique(d$feature)))
  ggplot(d, aes(x = feature, y = score)) +
    geom_col(fill = "steelblue") + coord_flip() +
    labs(x = NULL, y = paste0("-log10(", pcol, ")"),
         title = paste("ALDEx2 top features --", label)) + theme_bw()
}
## ALDEx2 volcano (effect vs significance); only defined for the 2-group case.
.picrust_aldex_volcano <- function(ald, label) {
  if (!("effect" %in% colnames(ald))) return(NULL)
  pcol <- intersect(c("we.eBH", "wi.eBH"), colnames(ald))
  if (length(pcol) == 0) return(NULL)
  pcol <- pcol[1]
  d <- ald
  d$sig <- d[[pcol]] < 0.05
  ggplot(d, aes(x = effect, y = -log10(pmax(d[[pcol]], .Machine$double.xmin)), color = sig)) +
    geom_point(alpha = 0.7) +
    scale_color_manual(values = c(`FALSE` = "grey60", `TRUE` = "firebrick"), guide = "none") +
    labs(x = "ALDEx2 effect size", y = paste0("-log10(", pcol, ")"),
         title = paste("ALDEx2 volcano --", label)) + theme_bw()
}
## metagenomeSeq per-contrast top-feature bar chart.
.picrust_mgseq_bar <- function(mg, label) {
  if (is.null(mg) || nrow(mg) == 0 || !("adjPvalues" %in% colnames(mg))) return(NULL)
  d <- mg %>%
    dplyr::group_by(contrast) %>%
    dplyr::arrange(adjPvalues, .by_group = TRUE) %>%
    dplyr::slice_head(n = 15) %>%
    dplyr::ungroup()
  if (nrow(d) == 0) return(NULL)
  d$score <- -log10(pmax(d$adjPvalues, .Machine$double.xmin))
  d$feature <- factor(d$feature, levels = rev(unique(d$feature)))
  ggplot(d, aes(x = feature, y = score)) +
    geom_col(fill = "darkorange") + coord_flip() +
    facet_wrap(~contrast, scales = "free_y", ncol = 2) +
    labs(x = NULL, y = "-log10(adjusted p)",
         title = paste("metagenomeSeq top features --", label)) + theme_bw()
}

.render_picrust_panel <- function(mat, label) {
  if (is.null(mat)) { cat("No ", label, " input.\n"); return(invisible(NULL)) }
  ## Defensive: if the input still has a non-numeric first column (e.g. the
  ## "function" / "pathway" ID column from a naive read.table), hoist it to
  ## rownames so as.matrix() doesn't force everything to character.
  if (is.data.frame(mat) && ncol(mat) >= 2 && !is.numeric(mat[[1]])) {
    .ids <- as.character(mat[[1]])
    mat  <- mat[, -1, drop = FALSE]
    rownames(mat) <- make.unique(.ids)
  }
  mat <- as.matrix(mat)
  storage.mode(mat) <- "numeric"
  if (ncol(mat) < 4 || nrow(mat) < 2) {
    cat("Insufficient ", label, " data (", nrow(mat), " features x ", ncol(mat), " samples).\n", sep = "")
    return(invisible(NULL))
  }
  samp_ids <- intersect(colnames(mat), metadata$SampleID)
  if (length(samp_ids) < 4) {
    cat("Not enough samples shared with metadata for ", label, ".\n")
    return(invisible(NULL))
  }
  mat <- mat[, samp_ids, drop = FALSE]
  groups <- metadata$Group[match(samp_ids, metadata$SampleID)]

  ## Compute DAA first (these may cat a specific failure message), then emit all
  ## sub-sections (headers + tables + plots) through knit_child so the DT/plot
  ## widgets actually render. Tables are shown in full (DT length-menu handles rows).
  ald <- .daa_aldex2(mat, groups)
  mg  <- .daa_metagenomeSeq(mat, groups)

  ## ---- Abundance heatmaps ----
  lines <- c("### Abundance heatmaps {.tabset}", "")
  hm <- list(
    c("#### Top 15 (RAW)",  "raw", 15L), c("#### Top 30 (RAW)",  "raw", 30L),
    c("#### Top 15 (CLR)",  "clr", 15L), c("#### Top 30 (CLR)",  "clr", 30L))
  for (h in hm) {
    p <- tryCatch(.picrust_heatmap(mat, groups, as.integer(h[3]), h[2], label),
                  error = function(e) NULL)
    lines <- c(lines, h[1], "",
               if (is.null(p)) "_Heatmap unavailable for this level._\n"
               else kc_block(p, chunk_opts = "fig.width=12, fig.height=8"))
  }

  ## ---- ALDEx2 ----
  lines <- c(lines, "### ALDEx2 (global, all groups) {.tabset}", "")
  if (!is.null(ald)) {
    write.csv(ald, file.path(.report_dir, paste0("picrust2_aldex2_", label, ".csv")), row.names = FALSE)
    lines <- c(lines, "#### Table", "",
               kc_block(fgcz_dt(ald[order(ald$kw.eBH), ],
                                paste("ALDEx2 --", label, "(sorted by KW eBH p)"))))
    pbar <- .picrust_aldex_bar(ald, label)
    lines <- c(lines, "#### Top features", "",
               if (is.null(pbar)) "_No significance column for a bar chart._\n"
               else kc_block(pbar, chunk_opts = "fig.width=10, fig.height=7"))
    pvol <- .picrust_aldex_volcano(ald, label)
    lines <- c(lines, "#### Volcano", "",
               if (is.null(pvol))
                 "_Volcano (effect size) is only available for two-group comparisons; with >2 groups ALDEx2 uses the Kruskal-Wallis/GLM test, which has no single effect size._\n"
               else kc_block(pvol, chunk_opts = "fig.width=9, fig.height=7"))
  } else {
    lines <- c(lines, "_No ALDEx2 result for this level._\n")
  }

  ## ---- metagenomeSeq ----
  lines <- c(lines, "### metagenomeSeq (pairwise) {.tabset}", "")
  if (!is.null(mg) && nrow(mg) > 0) {
    write.csv(mg, file.path(.report_dir, paste0("picrust2_metagenomeSeq_", label, ".csv")), row.names = FALSE)
    lines <- c(lines, "#### Table", "",
               kc_block(fgcz_dt(mg[order(mg$adjPvalues), ],
                                paste("metagenomeSeq --", label, "(sorted by adjusted p)"))))
    pmg <- .picrust_mgseq_bar(mg, label)
    lines <- c(lines, "#### Top features", "",
               if (is.null(pmg)) "_No significance column for a bar chart._\n"
               else kc_block(pmg, chunk_opts = "fig.width=12, fig.height=8"))
  } else {
    lines <- c(lines, "_No metagenomeSeq result for this level._\n")
  }
  kc_render(lines)
}
```

## MetaCyc Pathways {.tabset}

```{r picrust_pathway, eval=.picrust_enabled, results='asis'}
.render_picrust_panel(.pi_path, "pathway")
```

## KEGG Orthologs (KO) {.tabset}

```{r picrust_ko, eval=.picrust_enabled, results='asis'}
.render_picrust_panel(.pi_ko, "ko")
```

## Enzyme Commission (EC) {.tabset}

```{r picrust_ec, eval=.picrust_enabled, results='asis'}
.render_picrust_panel(.pi_ec, "ec")
```

## ASV contributors

```{r picrust_contributors, eval=.picrust_enabled, results='asis'}
.contrib_path <- if (exists("picrust2_contrib") && !is.null(picrust2_contrib)) picrust2_contrib else NULL
if (!is.null(.contrib_path) && length(.contrib_path) > 0 && file.exists(.contrib_path[1])) {
  contrib <- tryCatch(read.delim(.contrib_path[1], sep = "\t", check.names = FALSE),
                      error = function(e) NULL)
  if (is.null(contrib) || nrow(contrib) == 0) {
    cat("Stratified PICRUSt2 file was present but empty or unreadable -- skipping.\n\n")
  } else {
    cn <- colnames(contrib)
    top <- NULL
    if (all(c("function", "taxon", "taxon_function_abun") %in% cn)) {
      ## Long "contrib" format (default --stratified output).
      top <- contrib %>%
        dplyr::group_by(`function`, taxon) %>%
        dplyr::summarise(mean_taxon_function_abun = mean(taxon_function_abun, na.rm = TRUE),
                         .groups = "drop") %>%
        dplyr::arrange(dplyr::desc(mean_taxon_function_abun))
    } else {
      ## Wide "strat" format (--stratified --wide_table): function + sequence
      ## id columns, then one numeric column per sample. Sum across samples.
      id_cols   <- intersect(c("function", "sequence", "taxon", "pathway", "OTU_ID"), cn)
      samp_cols <- setdiff(cn, id_cols)
      samp_cols <- samp_cols[sapply(contrib[samp_cols], is.numeric)]
      if (length(id_cols) >= 1 && length(samp_cols) >= 1) {
        contrib$total_abun <- rowSums(contrib[, samp_cols, drop = FALSE], na.rm = TRUE)
        top <- contrib[, c(id_cols, "total_abun"), drop = FALSE]
        top <- top[order(-top$total_abun), , drop = FALSE]
      }
    }
    if (is.null(top) || nrow(top) == 0) {
      cat("Stratified PICRUSt2 file present but its columns were not recognised ",
          "(saw: ", paste(head(cn, 10), collapse = ", "), "). Skipping.\n\n", sep = "")
    } else {
      write.csv(top, file.path(.report_dir, "picrust2_top_contributors.csv"), row.names = FALSE)
      cat("All (function, ASV) contributors by abundance (use the length menu to page through):\n\n")
      kc_render(kc_block(fgcz_dt(top, "ASV-to-function contributors")))
    }
  }
} else {
  cat("No stratified PICRUSt2 contribution file provided -- skipping.\n\n")
}
```

# Taxon-level CLR analysis {.tabset}

CLR (centered log-ratio) transforms each sample's count vector into log-ratios relative to
the geometric mean -- the natural representation for compositional sequencing data.
Per-taxon Wilcoxon tests are run on the CLR values; violin plots show the top 5 by
mean(|CLR|).

```{r clr_helpers}
clr_wilcoxon_level <- function(ps, level) {
  if (is.null(ps) || ntaxa(ps) == 0) return(NULL)
  ps_lvl <- tryCatch(tax_glom(ps, taxrank = level, NArm = TRUE), error = function(e) NULL)
  if (is.null(ps_lvl) || ntaxa(ps_lvl) == 0) return(NULL)
  mat <- as(otu_table(ps_lvl), "matrix")
  if (!taxa_are_rows(ps_lvl)) mat <- t(mat)
  tax_df <- as.data.frame(tax_table(ps_lvl), stringsAsFactors = FALSE)
  taxon_label <- tax_df[[level]]
  if (any(is.na(taxon_label) | taxon_label == "")) {
    bad <- which(is.na(taxon_label) | taxon_label == "")
    taxon_label[bad] <- paste0("unknown_", bad)
  }
  taxon_label <- make.unique(taxon_label)
  rownames(mat) <- taxon_label
  clr_mat <- clr_transform_matrix(mat)
  samp_groups <- sample_data(ps_lvl)$Group
  samp_groups <- as.character(samp_groups)
  group_pairs <- combn(unique(samp_groups), 2, simplify = FALSE)
  results <- do.call(rbind, lapply(rownames(clr_mat), function(tx) {
    do.call(rbind, lapply(group_pairs, function(p) {
      x <- clr_mat[tx, samp_groups == p[1]]
      y <- clr_mat[tx, samp_groups == p[2]]
      if (length(x) < 2 || length(y) < 2) return(NULL)
      w <- suppressWarnings(wilcox.test(x, y))
      data.frame(taxon = tx, group1 = p[1], group2 = p[2],
                 mean_clr_g1 = mean(x, na.rm = TRUE),
                 mean_clr_g2 = mean(y, na.rm = TRUE),
                 W = w$statistic, p = w$p.value,
                 stringsAsFactors = FALSE)
    }))
  }))
  if (is.null(results) || nrow(results) == 0) return(NULL)
  results$p.adj <- p.adjust(results$p, method = "BH")
  results <- results[order(results$p.adj), ]
  attr(results, "clr_matrix") <- clr_mat
  results
}

clr_top5_violin <- function(clr_mat, samp_groups, level_name, top_n = 5) {
  if (is.null(clr_mat)) return(NULL)
  mean_abs <- rowMeans(abs(clr_mat), na.rm = TRUE)
  top <- names(sort(mean_abs, decreasing = TRUE)[seq_len(min(top_n, length(mean_abs)))])
  long <- reshape2::melt(clr_mat[top, , drop = FALSE], varnames = c("Taxon", "Sample"),
                         value.name = "CLR")
  long$Group <- samp_groups[long$Sample]
  ggplot(long, aes(x = Group, y = CLR, fill = Group)) +
    geom_violin(scale = "width") +
    geom_jitter(width = 0.1, height = 0, size = 0.7, alpha = 0.6) +
    facet_wrap(~Taxon, scales = "free_y") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
    labs(title = paste("Top", length(top), level_name, "by mean(|CLR|)"))
}

render_clr_section <- function(ps, ps_label) {
  if (is.null(ps)) return(invisible(NULL))
  samp_groups <- setNames(as.character(sample_data(ps)$Group), sample_names(ps))
  ## Build the whole section as Rmd source and knit_child it, so the Pairwise
  ## Wilcoxon DT (an htmlwidget) actually renders. A print(fgcz_dt(...)) inside
  ## a results='asis' chunk emits nothing.
  lines <- character(0)
  for (rank in c("Species", "Genus", "Family", "Phylum")) {
    if (rank == "Species" && !has_species) next
    lines <- c(lines, sprintf("\n\n## CLR Wilcoxon -- %s (%s) {.tabset}\n", rank, ps_label))
    res <- clr_wilcoxon_level(ps, rank)
    if (is.null(res) || nrow(res) == 0) {
      lines <- c(lines, "_No CLR Wilcoxon results at this rank._\n")
      next
    }
    write.csv(res, file.path(.report_dir, paste0("CLR_Wilcoxon_", rank, "_", ps_label, ".csv")),
              row.names = FALSE)
    lines <- c(lines, "\n### Pairwise Wilcoxon\n",
               kc_block(fgcz_dt(head(res, 200),
                                paste(rank, "-- CLR Wilcoxon (", ps_label, ", top 200 by BH p)"))))
    p <- clr_top5_violin(attr(res, "clr_matrix"), samp_groups, rank, top_n = 5)
    lines <- c(lines, "\n### Top 5 by mean(|CLR|)\n")
    if (!is.null(p)) {
      lines <- c(lines, kc_block(p, chunk_opts = "fig.width=14, fig.height=8"))
    } else {
      lines <- c(lines, "_Not enough taxa for a violin plot._\n")
    }
  }
  kc_render(lines)
}
```

```{r clr_pre, fig.width=14, fig.height=8, results='asis'}
## No `## Pre-filter CLR` wrapper here: the outer `# Taxon-level CLR analysis
## {.tabset}` treats every level-2 header as a tab, and render_clr_section
## emits `## CLR Wilcoxon -- Rank (pre/post) {.tabset}` directly — those are
## the real tabs. A `## Pre-filter CLR` wrapper would appear as an empty
## sibling tab because content moves under its own emitted h2s.
render_clr_section(physeq, "pre")
```

```{r clr_post, fig.width=14, fig.height=8, results='asis'}
render_clr_section(physeq_filtered, "post")
```

# CLR composition barplots {.tabset}

```{r clr_barplots, fig.width=20, fig.height=10, results='asis'}
clr_barplot <- function(ps, level, label) {
  ps_lvl <- tryCatch(tax_glom(ps, level, NArm = TRUE), error = function(e) NULL)
  if (is.null(ps_lvl) || ntaxa(ps_lvl) == 0) return(NULL)
  mat <- as(otu_table(ps_lvl), "matrix")
  if (!taxa_are_rows(ps_lvl)) mat <- t(mat)
  rownames(mat) <- make.unique(as.character(tax_table(ps_lvl)[, level]))
  clr_mat <- clr_transform_matrix(mat)
  long <- reshape2::melt(clr_mat, varnames = c("Taxon", "Sample"), value.name = "CLR")
  long$Group <- as.character(sample_data(ps_lvl)$Group[match(long$Sample, sample_names(ps_lvl))])
  ggplot(long, aes(x = Sample, y = CLR, fill = Taxon)) +
    geom_bar(stat = "identity") +
    facet_grid(~Group, scales = "free_x") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7), legend.position = "none") +
    labs(title = paste("CLR composition --", level, "(", label, ")"),
         y = "CLR contribution")
}

ranks_clr_bar <- c(if (has_species) "Species" else NULL, "Genus", "Family", "Phylum")
for (lvl in ranks_clr_bar) {
  cat(sprintf("\n\n## %s {.tabset}\n\n", lvl))
  cat("\n### Pre-filter\n\n")
  p_pre <- clr_barplot(physeq, lvl, "pre")
  if (!is.null(p_pre)) print(p_pre)
  cat("\n\n### Post-filter\n\n")
  p_post <- clr_barplot(physeq_filtered, lvl, "post")
  if (!is.null(p_post)) print(p_post)
  cat("\n\n")
}
```

# CLR abundance tables export {.tabset}

```{r clr_export, results='asis'}
## Returns the DT widget (or NULL); emission is done via knit_child below so
## the widget actually renders inside the tabset.
render_clr_table <- function(ps, level, label) {
  if (is.null(ps)) return(NULL)
  ps_lvl <- tryCatch(tax_glom(ps, level, NArm = TRUE), error = function(e) NULL)
  if (is.null(ps_lvl) || ntaxa(ps_lvl) == 0) return(NULL)
  mat <- as(otu_table(ps_lvl), "matrix")
  if (!taxa_are_rows(ps_lvl)) mat <- t(mat)
  rownames(mat) <- make.unique(as.character(tax_table(ps_lvl)[, level]))
  clr_mat <- clr_transform_matrix(mat)
  raw_path <- file.path(.report_dir, paste0("raw_abundance_table_", tolower(level), "_", label, ".csv"))
  clr_path <- file.path(.report_dir, paste0("CLR_abundance_table_", tolower(level), "_", label, ".csv"))
  write.csv(as.data.frame(mat), raw_path)
  write.csv(as.data.frame(clr_mat), clr_path)
  clr_df <- data.frame(Taxon = rownames(clr_mat),
                       round(clr_mat, 3),
                       check.names = FALSE,
                       stringsAsFactors = FALSE)
  fgcz_dt(clr_df, sprintf("CLR abundance -- %s (%s)", level, label))
}

ranks_clr <- c(if (has_species) "Species" else NULL, "Genus", "Family", "Phylum")
.clr_export_lines <- character(0)
for (lvl in ranks_clr) {
  .clr_export_lines <- c(.clr_export_lines, sprintf("\n\n## %s {.tabset}\n", lvl))
  dt_pre <- render_clr_table(physeq, lvl, "pre")
  .clr_export_lines <- c(.clr_export_lines, "\n### Pre-filter\n",
                         if (!is.null(dt_pre)) kc_block(dt_pre) else "_No taxa at this rank._\n")
  dt_post <- render_clr_table(physeq_filtered, lvl, "post")
  .clr_export_lines <- c(.clr_export_lines, "\n### Post-filter\n",
                         if (!is.null(dt_post)) kc_block(dt_post) else "_No taxa at this rank._\n")
}
kc_render(.clr_export_lines)
```

# General taxonomic composition barplots {.tabset}

Plots include reads whose label is missing at the current rank by relabelling them with
their parent rank (e.g. `unknown Family (Phylum=Bacteroidota)`). Useful for cross-checking
against the main barplot section, which drops `NA` rows.

```{r general_barplots, fig.width=20, fig.height=12, results='asis'}
relabel_for_general <- function(ps, rank) {
  tax <- as.data.frame(tax_table(ps), stringsAsFactors = FALSE)
  parent_ranks <- switch(rank,
    "Species" = c("Genus", "Family", "Order", "Class", "Phylum", "Kingdom"),
    "Genus"   = c("Family", "Order", "Class", "Phylum", "Kingdom"),
    "Family"  = c("Order", "Class", "Phylum", "Kingdom"),
    "Phylum"  = c("Kingdom"),
    character(0))
  lab <- tax[[rank]]
  bad <- is.na(lab) | lab == ""
  if (any(bad) && length(parent_ranks) > 0) {
    for (pr in parent_ranks) {
      if (pr %in% colnames(tax)) {
        lab[bad] <- paste0("unknown ", rank, " (", pr, "=", tax[bad, pr], ")")
        bad <- is.na(lab) | lab == "" | grepl("=NA", lab)
        if (!any(bad)) break
      }
    }
  }
  tax[[rank]] <- lab
  tax_table(ps) <- tax_table(as.matrix(tax))
  ps
}

## One tabset per rank with Top 20 / Top 50 as sub-tabs. Emitting two
## sibling `## Rank (Top 20)` / `## Rank (Top 50)` tabsets in a row without
## proper blank-line separation caused the next iteration's `### Plot` to
## land as a second sub-tab of the previous rank's Top 50 tabset — hence
## "Genus top 20 showing up under Family (Top 50)".
ranks_gen <- c(if (has_species) "Species" else NULL, "Genus", "Family", "Phylum")
for (rank in ranks_gen) {
  ps_lab <- relabel_for_general(physeq_filtered, rank)
  cat(sprintf("\n\n## %s {.tabset}\n\n", rank))
  cat("\n### Top 20\n\n")
  print(plot_taxa_bar_topN(ps_lab, rank, 20, paste("General Top 20", rank)))
  cat("\n\n### Top 50\n\n")
  print(plot_taxa_bar_topN(ps_lab, rank, 50, paste("General Top 50", rank)))
  cat("\n\n")
}
```

# Interactive widgets

The following widgets were saved next to this report; download or click to open in a browser:

```{r widgets_summary, results='asis'}
widget_files <- list.files(.report_dir, pattern = "^interactive_.*\\.html$", full.names = FALSE)
if (length(widget_files) > 0) {
  for (wf in sort(widget_files)) cat(sprintf("- [%s](%s)\n", wf, wf))
} else {
  cat("_No interactive widgets were generated._\n")
}
```

# Session info

<details>
<summary>Click to expand session info</summary>
```{r session_info}
sessionInfo()
```
</details>
