---
params:
  reportTitle: "rMATS-turbo Alternative Splicing"
title: "`r params$reportTitle`"
---

<!-- Styling comes from the fgczQuartoTemplate assets staged next to this file
     (_metadata.yml is applied by Quarto automatically). ezRun stages them via
     makeQuartoReport(), which calls fgczQuartoTemplate::fgcz_render(). The report
     reads the objects saved by makeQuartoReport() from the working directory. -->

```{r setup}
#| include: false
library(ezRun)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(DT)

## ── Data loading (qs2 with rds fallback) ─────────────────────────────────
.fgcz_load <- function(name, nthreads = 4L) {
  paths <- c(
    paste0(name, ".qs2"), file.path("data", paste0(name, ".qs2")),
    paste0(name, ".rds"), file.path("data", paste0(name, ".rds"))
  )
  for (p in paths) {
    if (file.exists(p)) {
      if (grepl("\\.qs2$", p)) return(qs2::qs_read(p, nthreads = nthreads))
      return(readRDS(p))
    }
  }
  stop("No .qs2 or .rds file found for: ", name)
}

param   <- .fgcz_load("param")
ase     <- .fgcz_load("aseResult")
coldata <- .fgcz_load("coldata")

sampleGroup <- param$sampleGroup
refGroup    <- param$refGroup
fdrCut      <- as.numeric(param$FDR)
dpsiCut     <- as.numeric(param$deltaPSI)
topN        <- as.integer(param$topN)
statModel   <- param$statModel

## Colour-blind-safe qualitative palette (Okabe-Ito) — the user has red-green CVD,
## so event types are distinguished by these colours *and* by direct labels.
cbPalette <- c(
  "#0072B2", "#E69F00", "#009E73", "#CC79A7",
  "#56B4E9", "#D55E00", "#F0E442", "#999999"
)

## The 5 rMATS event types, in a fixed display order. Keep any unexpected types
## as their own levels rather than dropping them to NA.
eventLevels <- c("SE", "MXE", "A5SS", "A3SS", "RI")
knownLevels <- intersect(eventLevels, unique(ase$EventType))
otherLevels <- setdiff(unique(ase$EventType), eventLevels)
ase$EventType <- factor(ase$EventType, levels = c(knownLevels, otherLevels))

if (is.null(ase$gene_name)) ase$gene_name <- NA_character_

## Significance call: adjusted p below cutoff AND effect size above |dPSI| cutoff.
ase$significant <- !is.na(ase$FDR) & ase$FDR < fdrCut &
  !is.na(ase$abs_deltaPSI) & ase$abs_deltaPSI > dpsiCut
ase$direction <- ifelse(ase$deltaPSI > 0,
  paste0("up in ", sampleGroup), paste0("up in ", refGroup))

nSample <- as.integer(table(coldata$condition)[c(sampleGroup, refGroup)])
names(nSample) <- c(sampleGroup, refGroup)

## ── Tiny schematic SVGs + definitions for the event types ────────────────
## Blue = constitutive exon, orange = the alternative element; dashed line = the
## spliced-out (excluded) path. Included isoform = the rMATS "inclusion" form.
## NB: pandoc lowercases the SVG viewBox attribute to an invalid 'viewbox', so
## width/height are set equal to the viewBox dimensions to render 1:1 regardless.
eventInfo <- list(
  SE = list(
    name = "Skipped Exon",
    def  = "A cassette exon is either included or skipped. Inclusion form contains the target exon.",
    svg  = '<svg viewBox="0 0 170 48" width="170" height="48"><line x1="34" y1="24" x2="136" y2="24" stroke="#999" stroke-width="2"/><path d="M42 19 q43 -17 86 0" fill="none" stroke="#555" stroke-width="1.5" stroke-dasharray="3,2"/><rect x="4" y="16" width="30" height="16" fill="#0072B2"/><rect x="71" y="16" width="28" height="16" fill="#E69F00"/><rect x="136" y="16" width="30" height="16" fill="#0072B2"/></svg>'
  ),
  MXE = list(
    name = "Mutually Exclusive Exons",
    def  = "Exactly one of two adjacent exons is used (never both, never neither).",
    svg  = '<svg viewBox="0 0 176 52" width="176" height="52"><line x1="34" y1="26" x2="142" y2="26" stroke="#999" stroke-width="2"/><rect x="4" y="18" width="30" height="16" fill="#0072B2"/><rect x="62" y="4"  width="26" height="14" fill="#E69F00"/><rect x="96" y="34" width="26" height="14" fill="#D55E00"/><rect x="142" y="18" width="30" height="16" fill="#0072B2"/></svg>'
  ),
  A5SS = list(
    name = "Alternative 5′ Splice Site",
    def  = "Two competing donor sites at an exon's 3′ end give a longer or shorter exon.",
    svg  = '<svg viewBox="0 0 170 48" width="170" height="48"><rect x="4" y="16" width="44" height="16" fill="#0072B2"/><rect x="48" y="16" width="16" height="16" fill="#E69F00"/><line x1="64" y1="24" x2="136" y2="24" stroke="#999" stroke-width="2"/><rect x="136" y="16" width="30" height="16" fill="#0072B2"/></svg>'
  ),
  A3SS = list(
    name = "Alternative 3′ Splice Site",
    def  = "Two competing acceptor sites at an exon's 5′ end give a longer or shorter exon.",
    svg  = '<svg viewBox="0 0 170 48" width="170" height="48"><rect x="4" y="16" width="30" height="16" fill="#0072B2"/><line x1="34" y1="24" x2="106" y2="24" stroke="#999" stroke-width="2"/><rect x="106" y="16" width="16" height="16" fill="#E69F00"/><rect x="122" y="16" width="44" height="16" fill="#0072B2"/></svg>'
  ),
  RI = list(
    name = "Retained Intron",
    def  = "An intron is retained in the mature transcript instead of being spliced out.",
    svg  = '<svg viewBox="0 0 170 48" width="170" height="48"><rect x="4" y="16" width="40" height="16" fill="#0072B2"/><rect x="44" y="16" width="82" height="16" fill="#E69F00" opacity="0.5"/><path d="M44 24 q41 14 82 0" fill="none" stroke="#555" stroke-width="1.5" stroke-dasharray="3,2"/><rect x="126" y="16" width="40" height="16" fill="#0072B2"/></svg>'
  )
)
```

Rendered `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`.

::: {.panel-tabset}

# Overview

Differential alternative splicing for **`r sampleGroup`** vs **`r refGroup`**,
computed with **rMATS-turbo** (`r statModel` statistical model, junction counts / JC).

- Reference build: **`r param$refBuild`**
- Transcript biotypes: **`r if (ezIsSpecified(param$transcriptTypes)) gsub(",", ", ", paste(param$transcriptTypes, collapse=", ")) else "all"`**
- Samples analysed: **`r nrow(coldata)`**  (`r nSample[[sampleGroup]]` `r sampleGroup`, `r nSample[[refGroup]]` `r refGroup`)
- Events tested: **`r nrow(ase)`**
- Significance: adjusted p < **`r fdrCut`** and |ΔPSI| > **`r dpsiCut`**
- Significant events: **`r sum(ase$significant)`**

ΔPSI is the rMATS **IncLevelDifference** = mean(PSI, `r sampleGroup`) −
mean(PSI, `r refGroup`) (positive = higher inclusion in **`r sampleGroup`**). See
the **Data** tab for the full JC/JCEC result tables and the explorer bundle.

# Event-type breakdown

::: {.callout-note collapse="true"}
## What do the event types mean?

```{r eventLegend}
#| echo: false
#| results: asis
cat('<div style="display:grid;grid-template-columns:1fr 1fr;gap:12px 24px;align-items:center;">\n')
for (et in names(eventInfo)) {
  ei <- eventInfo[[et]]
  cat(sprintf(
    '<div style="display:flex;gap:10px;align-items:center;">%s<div><b>%s</b> &mdash; %s<br><span style="opacity:0.8;">%s</span></div></div>\n',
    ei$svg, et, ei$name, ei$def
  ))
}
cat('</div>\n\n')
cat('<p style="opacity:0.8;margin-top:8px;">Blue = constitutive exon, orange = the alternative element; dashed line = the spliced-out (excluded) path. PSI (Percent Spliced In) is the fraction of transcripts using the "included" isoform.</p>\n')
```
:::

```{r breakdown}
#| echo: false
brk <- ase |>
  group_by(EventType) |>
  summarise(
    Tested = n(),
    Significant = sum(significant),
    `Up in sampleGroup` = sum(significant & deltaPSI > 0),
    `Up in refGroup` = sum(significant & deltaPSI < 0),
    .groups = "drop"
  )
DT::datatable(
  brk,
  caption = paste0("Events per type. 'Significant' = adj.p < ", fdrCut,
                   " and |ΔPSI| > ", dpsiCut, "."),
  options = list(dom = "t", pageLength = 20), rownames = FALSE
)
```

```{r breakdownPlot}
#| echo: false
#| fig-width: 8
#| fig-height: 4.5
brkLong <- brk |>
  select(EventType, Significant) |>
  filter(!is.na(EventType))
if (nrow(brkLong) > 0 && sum(brkLong$Significant) > 0) {
  ggplot(brkLong, aes(x = EventType, y = Significant, fill = EventType)) +
    geom_col() +
    geom_text(aes(label = Significant), vjust = -0.3, size = 3.5) +
    scale_fill_manual(values = colorRampPalette(cbPalette)(nrow(brkLong)), guide = "none") +
    labs(x = "Event type", y = "Significant events",
         title = "Significant differential splicing events by type") +
    theme_bw()
} else {
  cat("No significant events to plot.\n")
}
```

# Volcano

```{r volcano}
#| echo: false
#| fig-width: 8
#| fig-height: 6
vol <- ase |> filter(!is.na(FDR), !is.na(deltaPSI))
if (nrow(vol) > 0) {
  vol$negLog10FDR <- -log10(pmax(vol$FDR, .Machine$double.xmin))
  vol$colourKey <- ifelse(vol$significant, as.character(vol$EventType), "n.s.")
  ## build a colour for every event type actually present (interpolate if there
  ## are more than the base palette), so scale_colour_manual never errors on an
  ## unmapped value.
  presentTypes <- setdiff(unique(vol$colourKey), "n.s.")
  typeColours <- setNames(
    grDevices::colorRampPalette(cbPalette)(max(length(presentTypes), 1))[seq_along(presentTypes)],
    presentTypes
  )
  colourValues <- c(typeColours, "n.s." = "grey75")
  ## label the top few significant events by gene (fall back to event name)
  vol$lab <- ifelse(!is.na(vol$gene_name) & vol$gene_name != "", vol$gene_name, vol$EventName)
  lab <- vol |> filter(significant) |> arrange(desc(abs_deltaPSI)) |> head(15)
  ggplot(vol, aes(x = deltaPSI, y = negLog10FDR)) +
    geom_point(aes(colour = colourKey), alpha = 0.6, size = 1.4) +
    geom_hline(yintercept = -log10(fdrCut), linetype = "dashed", colour = "grey40") +
    geom_vline(xintercept = c(-dpsiCut, dpsiCut), linetype = "dashed", colour = "grey40") +
    ggrepel::geom_text_repel(data = lab, aes(label = lab), size = 2.6, max.overlaps = 20) +
    scale_colour_manual(values = colourValues, name = "Event type") +
    labs(x = paste0("ΔPSI  (", sampleGroup, " − ", refGroup, ")"),
         y = "−log10 adjusted p") +
    theme_bw()
} else {
  cat("No testable events for the volcano plot.\n")
}
```

# Top events

```{r topTable}
#| echo: false
keepCols <- intersect(
  c("gene_name", "EventName", "EventType", "EventRegion", "deltaPSI",
    "abs_deltaPSI", "PValue", "FDR"),
  colnames(ase)
)
tbl <- ase |>
  filter(significant) |>
  arrange(desc(abs_deltaPSI)) |>
  select(all_of(keepCols))
if (nrow(tbl) == 0) {
  cat("No significant events at the current thresholds.\n")
} else {
  ## render gene_name as an Ensembl search link
  if ("gene_name" %in% colnames(tbl)) {
    g <- tbl$gene_name
    tbl$gene_name <- ifelse(
      is.na(g) | g == "", g,
      paste0('<a href="https://www.ensembl.org/Multi/Search/Results?q=', g,
             '" target="_blank">', g, "</a>")
    )
  }
  numCols <- intersect(c("deltaPSI", "abs_deltaPSI", "PValue", "FDR"), colnames(tbl))
  DT::datatable(
    tbl, rownames = FALSE, filter = "top", escape = FALSE,
    extensions = "Buttons",
    caption = "Significant events, ranked by |ΔPSI|.",
    options = list(dom = "Bfrtip", buttons = c("excel"),
                   pageLength = 25, scrollX = TRUE)
  ) |> DT::formatSignif(columns = numCols, digits = 3)
}
```

# Top-event PSI

Per-replicate PSI (Percent Spliced In) by condition for the top **`r topN`**
significant events (by |ΔPSI|), from the rMATS `IncLevel1` / `IncLevel2` values.

```{r psiBoxplots}
#| echo: false
#| layout-ncol: 2
#| fig-width: 5
#| fig-height: 4
topEvents <- ase |>
  filter(significant) |>
  arrange(desc(abs_deltaPSI)) |>
  head(topN)
if (nrow(topEvents) == 0) {
  cat("No significant events to draw PSI boxplots for.\n")
} else {
  for (i in seq_len(nrow(topEvents))) {
    ev <- topEvents$EventName[i]
    gn <- topEvents$gene_name[i]
    et <- as.character(topEvents$EventType[i])
    ttl <- sprintf("%s  %s  (ΔPSI=%.2f)",
                   ifelse(is.na(gn) | gn == "", ev, gn), et, topEvents$deltaPSI[i])
    tryCatch(
      {
        s1 <- suppressWarnings(as.numeric(strsplit(topEvents$IncLevel1[i], ",")[[1]]))
        s2 <- suppressWarnings(as.numeric(strsplit(topEvents$IncLevel2[i], ",")[[1]]))
        df <- data.frame(
          PSI = c(s1, s2),
          group = factor(
            c(rep(sampleGroup, length(s1)), rep(refGroup, length(s2))),
            levels = c(refGroup, sampleGroup)
          )
        )
        df <- df[!is.na(df$PSI), , drop = FALSE]
        if (nrow(df) == 0) stop("no PSI values")
        print(
          ggplot(df, aes(x = group, y = PSI, fill = group)) +
            geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5) +
            geom_jitter(width = 0.12, size = 1.6, alpha = 0.9) +
            scale_fill_manual(values = c(cbPalette[1], cbPalette[2]), guide = "none") +
            coord_cartesian(ylim = c(0, 1)) +
            labs(x = NULL, y = "PSI", title = ttl) +
            theme_bw() +
            theme(plot.title = element_text(size = 9))
        )
      },
      error = function(e) {
        print(
          ggplot() +
            annotate("text", 0, 0, size = 3,
                     label = paste0("PSI unavailable:\n", ev)) +
            theme_void()
        )
      }
    )
  }
}
```

# Data

```{r explorerLink}
#| echo: false
#| output: asis
if (ezIsSpecified(param$exploreURL)) {
  cat("### Interactive explorer\n\n")
  cat(sprintf(
    "[Open the rMATS-turbo results explorer](%s) to filter the result table, browse the volcano, and view per-event PSI boxplots by condition.\n\n",
    param$exploreURL
  ))
}
```

```{r downloads}
#| echo: false
#| output: asis
cat("### Download files\n\n")
files <- c(
  "rMATS_JC_all.txt",
  sort(list.files(".", pattern = "^rMATS_(SE|MXE|A5SS|A3SS|RI)\\.MATS\\.(JC|JCEC)\\.txt$")),
  "rMATS_summary.txt",
  "rmatsTurbo_explore.qs2"
)
files <- unique(files[file.exists(files)])
labels <- c(
  "rMATS_JC_all.txt"        = "All tested events, combined JC table (TSV)",
  "rMATS_summary.txt"       = "rMATS run summary (events + significant counts per type)",
  "rmatsTurbo_explore.qs2"  = "Explorer bundle (qs2; result table + per-sample PSI + conditions)"
)
cat("Result files in this report folder (JC = junction counts, JCEC = junction + exon-body counts):\n\n")
for (f in files) {
  lab <- if (f %in% names(labels)) labels[[f]] else {
    m <- sub("^rMATS_(.*)\\.MATS\\.(JC|JCEC)\\.txt$", "\\1 events, \\2 (TSV)", f)
    m
  }
  cat(sprintf("- [%s](%s) — %s\n", f, f, lab))
}
cat("\n")
```

# SessionInfo

```{r sessionInfo}
#| echo: false
ezSessionInfo()
```

:::
