---
title:  "test"
author: "Functional Genomics Center Zurich"
output: 
  html_document:
    self_contained: false
    includes:
      in_header: fgcz_header.html
    css: fgcz.css
---

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

```{r setup, include=FALSE}
require(ezRun)
require(Seurat)
require(dplyr)
## ----------------------------------------------
## debug
# title:  "`r metadata(sce)$param$name`"
sce <- readRDS("/scratch/gtan/dev/SCReports-p2284/AVM_17_08_2018_SCCountQC/sce.rds")
## end of debug
debug <- FALSE

scData <- metadata(sce)$scData
param <- metadata(sce)$param
```

```{r pos markers, echo=FALSE}
markers <- FindAllMarkers(object=scData, only.pos=TRUE,
                              return.thresh = 0.05)
posMarkersFn <- "pos_markers.tsv"
ezWrite.table(markers, posMarkersFn)
```

## Clustering Results {.tabset}

### Cell clustering tSNE

This analysis was done with the R package Seurat. The distance metric which drives the clustering analysis is based on Principal Component Analysis.  In the plot below each dot is a cell.
We eliminated the genes that are expressed in < 5 cells. We eliminated the cells that have < `r param$min_genes` genes detected, or > `r param$max_genes` genes.

```{r tSNE, echo=FALSE}
TSNEPlot(object=scData)
```

Number of cells in each cluster (`r names(summary(scData@ident))`, respectively):

`r summary(scData@ident)`

### Control of the cell clustering

```{r FeaturePlot, echo=FALSE}
FeaturePlot(object = scData, features.plot = "nGene", 
            reduction.use = "tsne", no.legend = FALSE)
FeaturePlot(object = scData, features.plot = "nUMI", 
            reduction.use = "tsne", no.legend = FALSE)
```

### Table with top cluster markers

List of all positive cluster markers can be accessed [here](`r posMarkersFn`).

```{r markers table, echo=FALSE}
cm <- markers[which(markers$p_val_adj<0.01 & abs(markers$avg_logFC)>0.5),
              c("gene","cluster","avg_logFC","p_val_adj")]
rownames(cm) <- NULL
caption ="Expression differences of cluster marker genes."
ezInteractiveTableRmd(cm, digits=4, title=caption)
```

### Visualizations of top cluster markers

```{r top cluster markers, fig.width=4, fig.height=3, echo=FALSE}
some_markers <- cm$gene[1:100]
for(i in 1:length(some_markers)){
  plot(VlnPlot(object = scData, some_markers[i], do.return=TRUE))
  FeaturePlot(object = scData, features.plot = some_markers[i],
              cols.use = c("gray", "red"),
              reduction.use = "tsne", no.legend = FALSE)
}
```

### Heatmap of cluster marker genes
Max 10 marker genes are shown for each cluster. 

```{r heatmap, fig.width=10, fig.height=20, echo=FALSE}
top10 <- markers %>% group_by(cluster) %>% top_n(10, avg_logFC)
DoHeatmap(object = scData, genes.use = top10$gene, slim.col.label = TRUE,
          remove.key=TRUE)
```

### Expression of selected genes

Each dot is a cell. On the right-hand plot, the color corresponds to the gene expression levels. 

```{r selected markers, fig.width=4, fig.height=3, echo=FALSE, results='asis'}
if(length(param$markersToCheck) >0){
  all_m = list()
  for(i in 1:length(param$markersToCheck)){
    cat(paste("####", names(param$markersToCheck)[i]))
    check_markers <- param$markersToCheck[[i]]
    check_markers <- rownames(scData@data)[toupper(rownames(scData@data)) %in%
                                             toupper(check_markers)]
    all_m[[i]] <- check_markers
    if(length(check_markers) == 0)
      next
    for(j in 1:length(check_markers)){
      plot(VlnPlot(object = scData, check_markers[j], do.return=TRUE))
      FeaturePlot(object = scData, features.plot = check_markers[j],
                  cols.use = c("gray", "red"),
                  reduction.use = "tsne", no.legend = FALSE)
    }
    cat("\n")
  }
}

```

### Seurat parameters
```{r, echo=FALSE}
PrintFindClustersParams(object = scData)
```

### SessionInfo
```{r, echo=FALSE}
sessionInfo()
```
