---
title:  "test"
author: "Functional Genomics Center Zurich"
output: 
  html_document:
    self_contained: false
    includes:
      in_header: fgcz_header.html
    css: fgcz.css
---

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

```{r setup, include=FALSE}
require(ezRun)
require(Seurat)
## ----------------------------------------------
## debug
# title:  "`r metadata(sce)$param$name`"
sce <- readRDS("/scratch/gtan/dev/SCReports-p2284/AVM_17_08_2018_SCCountQC/sce.rds")
## end of debug
debug <- FALSE

scData <- metadata(sce)$scData
param <- metadata(sce)$param
```

```{r pos markers, echo=FALSE}
markers <- FindAllMarkers(object=scData, only.pos=TRUE,
                              return.thresh = 0.05)
posMarkersFn <- "pos_markers.tsv"
ezWrite.table(markers, posMarkersFn)
```

## Clustering Results {.tabset}

### Cell clustering tSNE

This analysis was done with the R package Seurat. The distance metric which drives the clustering analysis is based on Principal Component Analysis.  In the plot below each dot is a cell.
We eliminated the genes that are expressed in < 5 cells. We eliminated the cells that have < `r param$min_genes` genes detected, or > `r param$max_genes` genes.

```{r tSNE, echo=FALSE}
TSNEPlot(object=scData)
```

Number of cells in each cluster (`r names(summary(scData@ident))`, respectively):

`r summary(scData@ident)`

### Control of the cell clustering

```{r FeaturePlot, echo=FALSE}
FeaturePlot(object = scData, features.plot = "nGene", 
            reduction.use = "tsne", no.legend = FALSE)
FeaturePlot(object = scData, features.plot = "nUMI", 
            reduction.use = "tsne", no.legend = FALSE)
```

### Table with top cluster markers

List of all positive cluster markers can be accessed [here](`r posMarkersFn`).

```{r markers table, echo=FALSE}
cm <- markers[ ,c("gene","cluster","avg_logFC","p_val_adj")]
cm <- cm[which(cm$p_val_adj<0.01 & abs(cm$avg_logFC)>0.5), ]
rownames(cm) <- NULL
caption ="Expression differences of cluster marker genes."
# datatable(cm, options = list(pageLength = 10), 
#                   rownames = FALSE,
#                   escape = FALSE,
#                   caption = htmltools::tags$caption(
#                     style = 'caption-side: bottom; text-align: center;',
#                     'Table: ', htmltools::em(caption)
#                     )
#             )
ezInteractiveTableRmd(cm, digits=3, title=caption)
```

### Seurat parameters
```{r, echo=FALSE}
PrintFindClustersParams(object = scData)
```

### SessionInfo
```{r, echo=FALSE}
sessionInfo()
```
