---
title:  "`r metadata(sce)$param$name`"
author: "Functional Genomics Center Zurich"
output: 
  html_document:
    self_contained: false
    includes:
      in_header: fgcz_header.html
    css: fgcz.css
editor_options: 
  chunk_output_type: console
---

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

```{r setup, include=FALSE}
# input for this report: sce, output
require(ezRun)
require(Seurat)
require(dplyr)
require(tibble)
require(kableExtra)
require(fastICA)
require(Matrix)
require(readr)
require(slingshot)
require(RColorBrewer)
require(SummarizedExperiment)
## ----------------------------------------------
## debug
# title:  "`r metadata(sce)$param$name`"
# sce <- readRDS("/scratch/gtan/dev/SCReports-p2284/AVM_17_08_2018_SCReport/sce.rds")
# sce <- readRDS("/srv/gstore/projects/p2838/SCReport_32064_2019-01-28--16-12-45/NCC_Trpv5_Tomato_sc_A01_SCReport/sce.rds")
# sce <- readRDS("/scratch/gtan/debug/quickDebug/sce.rds")
## end of debug
debug <- FALSE

scData <- metadata(sce)$scData
param <- metadata(sce)$param
output <- metadata(sce)$output

pvalue_allMarkers <- 0.05
pvalue_all2allMarkers <- 0.01
```

```{r javascript, echo=FALSE, results='asis'}
jsFile = system.file("extdata/enrichr.js", package="ezRun", mustWork=TRUE)
invisible(file.copy(from=jsFile, to=basename(jsFile), overwrite=TRUE))
cat(paste0("<SCRIPT language=\"JavaScript\" SRC=\"", basename(jsFile), "\"></SCRIPT>"))
```

```{r pos markers, echo=FALSE}
markers <- FindAllMarkers(object=scData, only.pos=TRUE,
                          return.thresh = pvalue_allMarkers)
posMarkersFn <- "pos_markers.tsv"
ezWrite.table(markers, posMarkersFn)

## Significant markers
cm <- markers[ ,c("gene","cluster","avg_logFC","p_val_adj")]
rownames(cm) <- NULL
```

```{r pos markers all2all, echo=FALSE}
if(param$all2allMarkers){
  clusterCombs <- combn(levels(scData@ident), m=2)
  all2allMarkers <- mcmapply(FindMarkers, as.integer(clusterCombs[1, ]), 
                             as.integer(clusterCombs[2, ]),
                             MoreArgs = list(object=scData,
                                             only.pos=FALSE),
                             mc.preschedule=FALSE,
                             mc.cores=min(4L, param$cores),
                             SIMPLIFY=FALSE)
  all2allMarkers <- lapply(all2allMarkers, function(x){
    x[x$p_val <= pvalue_all2allMarkers, ]
  })
  names(all2allMarkers) <- apply(clusterCombs, 2, paste, collapse="vs")
}
```

```{r fastICA, echo=FALSE}
tSNE_data <- as_tibble(scData@dr$tsne@cell.embeddings,
                       rownames="cells")
tSNE_data <- rename(tSNE_data, X=`tSNE_1`, Y=`tSNE_2`)
tSNE_data$cluster <- scData@ident

if(param$runPseudoTime){
  cd <- as.matrix(scData@data)
  set.seed(10)
  a <- fastICA(cd, 2, alg.typ = "deflation", fun = "logcosh",
               alpha = 1, method = "C", row.norm = FALSE, maxit = 200,
               tol = 0.0001, verbose = FALSE)
  tSNE_data$pseudotime <- a$A[1,]
}
tSNEFn <- "tSNE_data.tsv"
write_tsv(tSNE_data, path=tSNEFn)
```

## Clustering Results {.tabset}

### Cell clustering tSNE

This analysis was done with the R package Seurat. The distance metric which drives the clustering analysis is based on Principal Component Analysis.  In the plot below each dot is a cell.
We eliminated the genes that are expressed in < `r param$minCellsPerGene` cells. We eliminated the cells that have < `r param$minGenesPerCell` genes detected, or > `r param$maxGenesPerCell` genes. We have `r sum(summary(scData@ident))` cells left in total.

```{r tSNE, echo=FALSE, results='asis'}
# Coloured by plate/run if available
if(length(unique(scData@meta.data$Plate)) >= 2){
  TSNEPlot(object=scData, group.by="Plate")
}
# Coloured by cluster
TSNEPlot(object=scData, do.label=TRUE)

kable(tibble(Cluster=names(summary(scData@ident)),
             "# of cells"=summary(scData@ident)),
      row.names=FALSE, format="html",
      caption="Number of cells in each cluster") %>%
  kable_styling(bootstrap_options = "striped", full_width = F,
                position = "float_right")
```

```{r pc.genes, echo=FALSE}
if(ezIsSpecified(param$pcGenes)){
  cat("Selected genes for clustering:\n")
  metadata(sce)[["pc.genes"]]
  cat("\n")
}
```

```{r gene mean per cluster, echo=FALSE}
tr_cnts <- expm1(scData@data)
geneMeans <- rowsum(t(as.matrix(tr_cnts)), group=scData@ident)
geneMeans <- sweep(geneMeans, 1,
                   STATS=table(scData@ident)[rownames(geneMeans)], FUN="/")
geneMeans <- log1p(t(geneMeans))
colnames(geneMeans) <- paste("cluster", colnames(geneMeans), sep="_")
geneMeanPerClusterFn <- "gene_means_per_cluster.txt"
ezWrite.table(geneMeans, geneMeanPerClusterFn)

geneMeans <- Matrix::rowMeans(tr_cnts)
geneMeans <- log1p(geneMeans)
geneMeansFn <- "gene_means.txt"
ezWrite.table(geneMeans, geneMeansFn)
```

[`r geneMeanPerClusterFn`](`r geneMeanPerClusterFn`)

[`r geneMeansFn`](`r geneMeansFn`)

[`r tSNEFn`](`r tSNEFn`)

### Control of the cell clustering

```{r FeaturePlot, echo=FALSE, eval=!debug}
FeaturePlot(object = scData, features.plot = "nGene", 
            reduction.use = "tsne", no.legend = FALSE)
FeaturePlot(object = scData, features.plot = "nUMI",
            reduction.use = "tsne", no.legend = FALSE)
FeaturePlot(object = scData, features.plot = "perc_mito",
            reduction.use = "tsne", no.legend = FALSE)
```

### Top cluster markers
List of all positive cluster markers can be accessed [here](`r posMarkersFn`).

```{r enrichr markers, echo=FALSE, results='asis'}
if(doEnrichr(param)){
  genesPerCluster <- split(markers$gene, markers$cluster)
  jsCall = paste0('enrich({list: "', sapply(genesPerCluster, paste, collapse="\\n"), '", popup: true});')
  enrichrCalls <- paste0("<a href='javascript:void(0)' onClick='", jsCall, 
                         "'>Analyse at Enrichr website</a>")
  enrichrTable <- tibble(Cluster=names(genesPerCluster),
                         "# of markers"=lengths(genesPerCluster),
                         "Enrichr link"=enrichrCalls)
  kable(enrichrTable, format="html", escape=FALSE,
        caption=paste0("GeneSet enrichment: genes with pvalue ", pvalue_allMarkers)) %>%
    kable_styling("striped", full_width = F, position = "left")
}
```

```{r markers table, echo=FALSE}
caption ="Expression differences of cluster marker genes"
ezInteractiveTableRmd(cm, digits=4, title=caption)
```

```{r section all2all, echo=FALSE, results='asis'}
if(doEnrichr(param) && param$all2allMarkers){
  cat("#### All clusters against all clusters comparison")
  cat("\n")
}
```

```{r enrichr all2all markers, echo=FALSE, results='asis'}
if(doEnrichr(param) && param$all2allMarkers){
  for(comparison in names(all2allMarkers)){
    write_tsv(as_tibble(all2allMarkers[[comparison]], rownames="Gene"),
              path=paste0(comparison, ".tsv"))
  }
  genesAllPerCluster <- lapply(all2allMarkers, rownames)
  genesUpPerCluster <- lapply(all2allMarkers, function(x){rownames(x)[x$avg_logFC > 0]})
  genesDownPerCluster <- lapply(all2allMarkers, function(x){rownames(x)[x$avg_logFC < 0]})
  
  jsCall_all = paste0('enrich({list: "', sapply(genesAllPerCluster, paste, collapse="\\n"), '", popup: true});')
  jsCall_up = paste0('enrich({list: "', sapply(genesUpPerCluster, paste, collapse="\\n"), '", popup: true});')
  jsCall_down = paste0('enrich({list: "', sapply(genesDownPerCluster, paste, collapse="\\n"), '", popup: true});')
  
  enrichrCalls_all <- paste0("<a href='javascript:void(0)' onClick='", jsCall_all,
                         "'>Analyse at Enrichr website</a>")
  enrichrCalls_up <- paste0("<a href='javascript:void(0)' onClick='", jsCall_up,
                         "'>Analyse at Enrichr website</a>")
  enrichrCalls_down <- paste0("<a href='javascript:void(0)' onClick='", jsCall_down,
                         "'>Analyse at Enrichr website</a>")
  enrichrTable <- tibble(Comparison=names(all2allMarkers),
                         "# of differentially expressed genes"=lengths(genesAllPerCluster),
                         "Enrichr link: all significant genes"=enrichrCalls_all,
                         "Enrichr link: up-regulated genes"=enrichrCalls_up,
                         "Enrichr link: down-regulated genes"=enrichrCalls_down,
                         "List of differentially expressed genes"=text_spec(paste0(names(all2allMarkers), ".tsv"), link=paste0(names(all2allMarkers), ".tsv")))
  kable(enrichrTable, format="html", escape=FALSE,
        caption=paste0("GeneSet enrichment: genes with pvalue ", pvalue_all2allMarkers)) %>%
    kable_styling("striped", full_width = F, position = "left")
}
```

### Visualizations of top cluster markers

```{r top cluster markers, fig.width=4, fig.height=3, echo=FALSE, eval=!debug}
some_markers <- head(arrange(cm, p_val_adj) %>% select(gene) %>% pull(), 100)
for(i in 1:length(some_markers)){
  plot(VlnPlot(object = scData, some_markers[i], do.return=TRUE))
  FeaturePlot(object = scData, features.plot = some_markers[i],
              cols.use = c("gray", "red"),
              reduction.use = "tsne", no.legend = FALSE)
}
```

### Heatmap of cluster marker genes
Max `r param$markersToShow` marker genes are shown for each cluster. 

```{r heatmap, fig.width=10, fig.height=param$markersToShow*1.5, echo=FALSE, eval=!debug}
top10 <- markers %>% group_by(cluster) %>% top_n(param$markersToShow, avg_logFC)
DoHeatmap(object = scData, genes.use = top10$gene, slim.col.label = TRUE,
          remove.key=TRUE)
```

### Expression of selected genes

Each dot is a cell. On the right-hand plot, the color corresponds to the gene expression levels. 

```{r selected markers, fig.width=4, fig.height=3, echo=FALSE, results='asis', eval=!debug}
if(length(param$markersToCheck) >0){
  all_m = list()
  for(i in 1:length(param$markersToCheck)){
    cat("\n")
    cat(paste("####", names(param$markersToCheck)[i]))
    cat("\n")
    check_markers <- param$markersToCheck[[i]]
    check_markers <- rownames(scData@data)[toupper(rownames(scData@data)) %in%
                                             toupper(check_markers)]
    all_m[[i]] <- check_markers
    if(length(check_markers) > 0){
      for(j in 1:length(check_markers)){
        plot(VlnPlot(object = scData, check_markers[j], do.return=TRUE))
        FeaturePlot(object = scData, features.plot = check_markers[j],
                    cols.use = c("gray", "red"),
                    reduction.use = "tsne", no.legend = FALSE)
      }
    }
    cat("\n")
  }
}

```

### Trajectory reconstruction and pseudotime inference

#### Slingshot
Slingshot is a cluster based ineage reconstruction and pseudotime inference approach.
Here, we use reduced-dimensional space of tSNE derived above and cell clustering produced by gaussian mixture modeling, instead of cell clustering done by Seurat.
More details is available at [here](https://www.bioconductor.org/packages/devel/bioc/vignettes/slingshot/inst/doc/slingshot.html).

```{r slingshot compute, echo=FALSE, message=FALSE, eval=param$runPseudoTime & !debug}
sceSlingshot <- Convert(from = scData, to = "sce")

sceSlingshot <- slingshot(sceSlingshot, clusterLabels = 'ident',
                          reducedDim = 'TSNE')
colors <- colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)
nrPseudoT <- grep("slingPseudotime_\\d+", colnames(colData(sceSlingshot)),
                  value=TRUE)
for(i in nrPseudoT){
  plot(reducedDim(sceSlingshot, type="TSNE"),
       col=colors[cut(sceSlingshot[[i]], breaks=100)],
       pch=16, asp = 1, main=i)
  lines(SlingshotDataSet(sceSlingshot), lwd=2)
}
## Trajectory
if(length(levels(sceSlingshot$ident)) >= 10){
  colors <- colorRampPalette(brewer.pal(9,'Set1'))(length(levels(sceSlingshot$ident)))
}else{
  colors <- brewer.pal(9,'Set1')
}
plot(reducedDim(sceSlingshot, type="TSNE"), 
     col = colors[sceSlingshot$ident], 
     pch=16, asp = 1, main="Trajectory reconstruction")
lines(SlingshotDataSet(sceSlingshot), lwd=2, type = 'lineages')

##  smooth curves
crv1 <- getCurves(SlingshotDataSet(sceSlingshot))
plot(reducedDim(sceSlingshot, type="TSNE"),
     col = colors[sceSlingshot$ident], asp = 1, pch = 16,
     main="Smooth curve")
lines(crv1, lwd = 3)
```

#### fastICA
```{r fastICA plot, echo=FALSE, fig.width=8, fig.height=5, eval=param$runPseudoTime && !debug}
p <- ggplot(tSNE_data, aes(X, Y)) + 
  geom_point(shape=21, size=5, colour = "black",
             aes(fill=pseudotime), alpha=0.7) +
  scale_fill_gradient(low="yellow", high="blue") +
  xlab("") + ylab("")
print(p)
```

### Interactive explorer
```{r make shiny data, echo=FALSE}
scData@scale.data <- NULL
saveRDS(list(scData=scData, tSNE_data=tSNE_data),
        file=basename(output$getColumn("Live Report")))
```
[Shiny explorer](`r output$getColumn("Live Report")`)

### Parameters
```{r report parameters, echo=FALSE}
param[c("minCellsPerGene", "minGenesPerCell", "maxGenesPerCell", "minReadsPerCell", 
        "maxMitoFraction", "pcs",
        "pcGenes", "x.low.cutoff", "x.high.cutoff", "y.cutoff", "vars.to.regress",
        "resolution", "markersToCheck")]
```

```{r seurat parameters, echo=FALSE}
PrintFindClustersParams(object = scData)
```

### SessionInfo
```{r, echo=FALSE}
sessionInfo()
```
