---
title:  "SCReport of merged samples"
author: "Functional Genomics Center Zurich"
output: 
  html_document:
    self_contained: false
    includes:
      in_header: fgcz_header.html
    css: fgcz.css
editor_options: 
  chunk_output_type: console
---

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

```{r setup, include=FALSE}
require(SummarizedExperiment)
require(ezRun)
require(cowplot)
require(tibble)
require(dplyr)
require(readr)
require(Seurat)
require(kableExtra)
# input for this report: sce1URL, sce2URL, param
# sce1URL <- "https://fgcz-sushi.uzh.ch/projects/p2284/SCReport_31302_2019-01-25--20-51-21/Recurrent_GBM_04092018_SCReport/00index.html"
# sce2URL <- "https://fgcz-sushi.uzh.ch/projects/p2979/SCReport_32645_2018-12-21--21-25-59/Epilepsy_26112018_SCReport/00index.html"

sce1 <- readRDS(file.path("/srv/gstore/projects",
                          sub("https://fgcz-(gstore|sushi).uzh.ch/projects", "",
                              dirname(sce1URL)), "sce.rds"))
sce2 <- readRDS(file.path("/srv/gstore/projects",
                          sub("https://fgcz-(gstore|sushi).uzh.ch/projects", "",
                              dirname(sce2URL)), "sce.rds"))
# to test
# param <- metadata(sce1)$param
# param$chosenClusters1 <- ""
# param$chosenClusters2 <- ""
# param$resolution <- 0.5
# param$cc <- 20
# param$batchCorrection <- "CCA"

pvalue_allMarkers <- 0.05
pvalue_all2allMarkers <- 0.01

if(ezIsSpecified(param$chosenClusters1)){
  chosenCells1 <- names(metadata(sce1)$scData@ident)[metadata(sce1)$scData@ident %in% 
                                                       param$chosenClusters1]
  sce1 <- sce1[, chosenCells1]
  metadata(sce1)$scData <- SubsetData(metadata(sce1)$scData,
                                      ident.use=param$chosenClusters1)
}
if(ezIsSpecified(param$chosenClusters2)){
  chosenCells2 <- names(metadata(sce2)$scData@ident)[metadata(sce2)$scData@ident %in% 
                                                       param$chosenClusters2]
  sce2 <- sce2[, chosenCells2]
  metadata(sce2)$scData <- SubsetData(metadata(sce2)$scData,
                                      ident.use=param$chosenClusters2)
}
```

```{r javascript, echo=FALSE, results='asis'}
jsFile = system.file("extdata/enrichr.js", package="ezRun", mustWork=TRUE)
invisible(file.copy(from=jsFile, to=basename(jsFile), overwrite=TRUE))
cat(paste0("<SCRIPT language=\"JavaScript\" SRC=\"", basename(jsFile), "\"></SCRIPT>"))
```

## Clustering results of merged samples {.tabset}

### Cell clustering without correction

```{r tSNE no correction, include=FALSE}
sce_cbind <- SingleCellExperiment::cbind(sce1, sce2)
metadata(sce_cbind)$param$resolution <- param$resolution
sce_cbind <- seuratPreProcess(sce_cbind)
scData.None <- metadata(sce_cbind)$scData
```

```{r tSNE no correction plot, echo=FALSE}
TSNEPlot(object=scData.None, group.by="Plate",
         do.return=TRUE)
TSNEPlot(object=scData.None,
         do.return=TRUE, do.label=TRUE)
toTable <- tibble(Cluster=names(summary(scData.None@ident)),
                  "# of cells"=summary(scData.None@ident))
cellCountsByPlate <- t(sapply(split(scData.None@meta.data$Plate, 
                                    scData.None@ident), table))
cellPercByPlate <- sweep(cellCountsByPlate, 2, colSums(cellCountsByPlate), "/")
colnames(cellPercByPlate) <- paste0(colnames(cellPercByPlate), "_fraction")
toTable <- left_join(toTable, 
                     as_tibble(cellCountsByPlate, rownames="Cluster"),
                     by="Cluster") %>%
  left_join(as_tibble(cellPercByPlate, rownames="Cluster"), by="Cluster")
## TODO: add fisher test?

kable(toTable,
      row.names=FALSE, format="html",
      caption="Number of cells in each cluster") %>%
  kable_styling(bootstrap_options = "striped", full_width = F,
                position = "float_right")
```

### Cell clustering with CCA correction

```{r run CCA, echo=FALSE, message=FALSE}
g.1 <- head(rownames(metadata(sce1)$scData@hvg.info), 1000)
g.2 <- head(rownames(metadata(sce2)$scData@hvg.info), 1000)
  
genes.use <- unique(c(g.1, g.2))
genes.use <- intersect(genes.use, rownames(metadata(sce1)$scData@scale.data))
genes.use <- intersect(genes.use, rownames(metadata(sce2)$scData@scale.data))
scData.CCA <- RunCCA(metadata(sce1)$scData, 
                     metadata(sce2)$scData,
                     genes.use=genes.use, num.cc=30,
                     scale.factor=getSeuratScalingFactor(param$scProtocol))
```

```{r, include=FALSE}
rm(sce1, sce2)
gc()
```

Visualize results of CCA plot CC1 versus CC2 and look at a violin plot.

```{r CCA plot, echo=FALSE, message=FALSE}
p1 <- DimPlot(object=scData.CCA, reduction.use="cca", group.by="Plate",
              pt.size=0.5, do.return=TRUE)
p1
p2 <- VlnPlot(object=scData.CCA, features.plot="CC1", group.by="Plate",
              do.return=TRUE)
p2
```

MetageneBicorPlot examines a measure of correlation strength for each CC and
find that this statistic generally saturates after a reasonable number of CCs.

```{r CCA MetageneBicorPlot, echo=FALSE, message=FALSE}
p3 <- MetageneBicorPlot(scData.CCA, grouping.var="Plate", dims.eval=1:30,
                        display.progress=FALSE)
```

```{r align CCA subspace, echo=FALSE, message=FALSE}
scData.CCA <- AlignSubspace(scData.CCA, reduction.type="cca",
                            grouping.var="Plate", dims.align=1:param$cc,
                            do.par=TRUE, num.cores=param$cores)
```

We can visualize the aligned CCA and perform an integrated analysis.

```{r aligned CCA plot, echo=FALSE}
p1 <- VlnPlot(object=scData.CCA, features.plot = "ACC1",
              group.by="Plate", do.return=TRUE)
p1
p2 <- VlnPlot(object=scData.CCA, features.plot="ACC2", group.by="Plate", 
              do.return=TRUE)
p2
```

```{r integrated analysis on CCA, echo=FALSE}
scData.CCA <- RunTSNE(object=scData.CCA, reduction.use="cca.aligned",
                      dims.use=1:param$cc, tsne.method="Rtsne",
                      perplexity=ifelse(length(scData.CCA@ident) > 384, 30, 10),
                      num_threads=param$cores)
scData.CCA <- FindClusters(object=scData.CCA, reduction.type="cca.aligned",
                           dims.use=1:param$cc, save.SNN=TRUE,
                           force.recalc=FALSE, print.output=FALSE,
                           resolution=param$resolution)
```

Visualization on t-SNE plot

```{r tSNE CCA correction plot, echo=FALSE}
p1 <- TSNEPlot(object=scData.CCA, group.by="Plate",
               do.return=TRUE)
p1
p2 <- TSNEPlot(object=scData.CCA,
               do.return=TRUE, do.label=TRUE)
p2
toTable <- tibble(Cluster=names(summary(scData.CCA@ident)),
                  "# of cells"=summary(scData.CCA@ident))
cellCountsByPlate <- t(sapply(split(scData.CCA@meta.data$Plate, 
                                    scData.CCA@ident), table))
cellPercByPlate <- sweep(cellCountsByPlate, 2, colSums(cellCountsByPlate), "/")
colnames(cellPercByPlate) <- paste0(colnames(cellPercByPlate), "_fraction")
toTable <- left_join(toTable, 
                     as_tibble(cellCountsByPlate, rownames="Cluster"),
                     by="Cluster") %>%
  left_join(as_tibble(cellPercByPlate, rownames="Cluster"), by="Cluster")
## TODO: add fisher test?

kable(toTable,
      row.names=FALSE, format="html",
      caption="Number of cells in each cluster") %>%
  kable_styling(bootstrap_options = "striped", full_width = F,
                position = "float_right")
```

```{r select correction, echo=FALSE}
if(param$batchCorrection == "None"){
  scData <- scData.None
}else if(param$batchCorrection == "CCA"){
  scData <- scData.CCA
}
# This is to avoid error in SplitDotPlotGG
# It uses _ to separate the ident
scData@meta.data$Plate <- gsub("_", ".", scData@meta.data$Plate)
# scData@meta.data$Plate <- factor(scData@meta.data$Plate,
#                                  levels=unique(scData@meta.data$Plate))
```

```{r, include=FALSE}
rm(scData.CCA, scData.None)
gc()
```
### Control of the cell clustering

```{r control cell clustering, echo=FALSE}
FeaturePlot(object = scData, features.plot = "nGene", 
            reduction.use = "tsne", no.legend = FALSE)
FeaturePlot(object = scData, features.plot = "nUMI", 
            reduction.use = "tsne", no.legend = FALSE)
```


### Top cluster markers

```{r pos markers, echo=FALSE}
markers <- FindAllMarkers(object=scData, only.pos=TRUE,
                          return.thresh = pvalue_allMarkers)
posMarkersFn <- "pos_markers.tsv"
ezWrite.table(markers, posMarkersFn)

## Significant markers
cm <- markers[ ,c("gene","cluster","avg_logFC","p_val_adj")]
rownames(cm) <- NULL
```

List of all positive cluster markers can be accessed [here](`r posMarkersFn`).

```{r enrichr markers all, echo=FALSE, results='asis'}
if(doEnrichr(param)){
  genesPerCluster <- split(markers$gene, markers$cluster)
  jsCall = paste0('enrich({list: "', sapply(genesPerCluster, paste, collapse="\\n"), '", popup: true});')
  enrichrCalls <- paste0("<a href='javascript:void(0)' onClick='", jsCall, 
                         "'>Analyse at Enrichr website</a>")
  enrichrTable <- tibble(Cluster=names(genesPerCluster),
                         "# of markers"=lengths(genesPerCluster),
                         "Enrichr link"=enrichrCalls)
  kable(enrichrTable, format="html", escape=FALSE,
        caption=paste0("GeneSet enrichment: genes with pvalue ", pvalue_allMarkers)) %>%
    kable_styling("striped", full_width = F, position = "left")
}
```

```{r markers table all, echo=FALSE}
caption ="Expression differences of cluster marker genes"
ezInteractiveTableRmd(cm, digits=4, title=caption)
```

```{r section all2all, echo=FALSE, results='asis'}
if(doEnrichr(param) && param$all2allMarkers){
  cat("#### All clusters against all clusters comparison")
  cat("\n")
}
```
```{r pos markers all2all, echo=FALSE}
if(param$all2allMarkers){
  clusterCombs <- combn(levels(scData@ident), m=2)
  all2allMarkers <- mcmapply(FindMarkers, as.integer(clusterCombs[1, ]), 
                             as.integer(clusterCombs[2, ]),
                             MoreArgs = list(object=scData,
                                             only.pos=FALSE),
                             mc.preschedule=FALSE,
                             mc.cores=min(4L, param$cores),
                             SIMPLIFY=FALSE)
  all2allMarkers <- lapply(all2allMarkers, function(x){
    x[x$p_val <= pvalue_all2allMarkers, ]
  })
  names(all2allMarkers) <- apply(clusterCombs, 2, paste, collapse="vs")
}
```

```{r enrichr all2all markers, echo=FALSE, results='asis'}
if(doEnrichr(param) && param$all2allMarkers){
  for(comparison in names(all2allMarkers)){
    write_tsv(as_tibble(all2allMarkers[[comparison]], rownames="Gene"),
              path=paste0(comparison, ".tsv"))
  }
  genesAllPerCluster <- lapply(all2allMarkers, rownames)
  genesUpPerCluster <- lapply(all2allMarkers, function(x){rownames(x)[x$avg_logFC > 0]})
  genesDownPerCluster <- lapply(all2allMarkers, function(x){rownames(x)[x$avg_logFC < 0]})
  
  jsCall_all = paste0('enrich({list: "', sapply(genesAllPerCluster, paste, collapse="\\n"), '", popup: true});')
  jsCall_up = paste0('enrich({list: "', sapply(genesUpPerCluster, paste, collapse="\\n"), '", popup: true});')
  jsCall_down = paste0('enrich({list: "', sapply(genesDownPerCluster, paste, collapse="\\n"), '", popup: true});')
  
  enrichrCalls_all <- paste0("<a href='javascript:void(0)' onClick='", jsCall_all,
                         "'>Analyse at Enrichr website</a>")
  enrichrCalls_up <- paste0("<a href='javascript:void(0)' onClick='", jsCall_up,
                         "'>Analyse at Enrichr website</a>")
  enrichrCalls_down <- paste0("<a href='javascript:void(0)' onClick='", jsCall_down,
                         "'>Analyse at Enrichr website</a>")
  enrichrTable <- tibble(Comparison=names(all2allMarkers),
                         "# of differentially expressed genes"=lengths(genesAllPerCluster),
                         "Enrichr link: all significant genes"=enrichrCalls_all,
                         "Enrichr link: up-regulated genes"=enrichrCalls_up,
                         "Enrichr link: down-regulated genes"=enrichrCalls_down,
                         "List of differentially expressed genes"=text_spec(paste0(names(all2allMarkers), ".tsv"), link=paste0(names(all2allMarkers), ".tsv")))
  kable(enrichrTable, format="html", escape=FALSE,
        caption=paste0("GeneSet enrichment: genes with pvalue ", pvalue_all2allMarkers)) %>%
    kable_styling("striped", full_width = F, position = "left")
}
```

### Visualizations of top cluster markers

```{r top cluster markers, fig.width=4, fig.height=3, echo=FALSE}
some_markers <- head(dplyr::arrange(cm, p_val_adj) %>% select(gene) %>% pull(), 
                     100)
for(i in 1:length(some_markers)){
  plot(VlnPlot(object = scData, some_markers[i], do.return=TRUE))
  FeaturePlot(object = scData, features.plot = some_markers[i],
              cols.use = c("gray", "red"),
              reduction.use = "tsne", no.legend = FALSE)
}
```

### Heatmap of cluster marker genes
Max `r param$markersToShow` marker genes are shown for each cluster. 

```{r heatmap, fig.width=10, fig.height=param$markersToShow*1.5, echo=FALSE}
top10 <- markers %>% group_by(cluster) %>% top_n(param$markersToShow, avg_logFC)
DoHeatmap(object = scData, genes.use = top10$gene, slim.col.label = TRUE,
          remove.key=TRUE)
```


### Conserved cell type markers

```{r conserved markers, echo=FALSE, warning=FALSE}
markers <- list()
for(eachCluster in levels(scData@ident)){
   markersEach <- try(FindConservedMarkers(scData, ident.1=eachCluster, 
                                       grouping.var="Plate",
                                       print.bar=FALSE, only.pos=TRUE),
                      silent=TRUE)
   ## to skip some groups with few cells
   if(class(markersEach) != "try-error"){
     markers[[eachCluster]] <- as_tibble(markersEach, rownames="gene")
   }
}
## some of the cluster have no significant conserved markers
markers <- markers[sapply(markers, nrow) != 0L] 
markers <- bind_rows(markers, .id="cluster")

posConservedMarkersFn <- "pos_conserved_markers.tsv"
write_tsv(markers, path=posConservedMarkersFn)
```

List of all positive conserved cell type markers can be accessed [here](`r posConservedMarkersFn`).

```{r enrichr markers conserved, echo=FALSE, results='asis'}
if(doEnrichr(param)){
  genesPerCluster <- split(markers$gene, markers$cluster)
  jsCall = paste0('enrich({list: "', sapply(genesPerCluster, paste, collapse="\\n"), '", popup: true});')
  enrichrCalls <- paste0("<a href='javascript:void(0)' onClick='", jsCall, 
                         "'>Analyse at Enrichr website</a>")
  enrichrTable <- tibble(Cluster=names(genesPerCluster),
                         "# of markers"=lengths(genesPerCluster),
                         "Enrichr link"=enrichrCalls)
  kable(enrichrTable, format="html", escape=FALSE,
        caption=paste0("GeneSet enrichment analysis: conserved marker genes")) %>%
    kable_styling("striped", full_width = F, position = "left")
}
```


```{r markers table conserved, echo=FALSE}
caption ="Conserved cell type markers"
ezInteractiveTableRmd(markers, title=caption)
```

### Visualisation of conserved markers
Top 30 conserved cell type markers are shown.

```{r conserved markers FeaturePlot, echo=FALSE, fig.width=4, fig.height=3}
markers.to.plot <- head(dplyr::arrange(markers, minimump_p_val) %>% 
                          dplyr::select(gene) %>% pull(), 30)
for(i in 1:length(markers.to.plot)){
  plot(VlnPlot(object = scData, markers.to.plot[i], do.return=TRUE))
  FeaturePlot(object = scData, features.plot = markers.to.plot[i],
              cols.use = c("gray", "red"),
              reduction.use = "tsne", no.legend = FALSE)
}
```

```{r visulisation conserved markers SplitDotPlotGG, echo=FALSE, fig.width=10, fig.height=0.5+length(unique(scData@ident)) * length(unique(scData@meta.data$Plate))*0.5}
sdp <- SplitDotPlotGG(scData, genes.plot = unique(markers.to.plot),
                      # It gives error when there are duplicated markers
                      cols.use = c("blue", "red"), x.lab.rot = T,
                      plot.legend = T, dot.scale = 8, do.return = T,
                      grouping.var = "Plate")
```

### Differential expressed genes
Identify differential expressed genes across conditions.

```{r diff expressed genes, echo=FALSE, results='asis', warning=FALSE, fig.width=7, fig.height=10}
scData@meta.data$cluster.plate <- paste0(scData@ident, "_",
                                         scData@meta.data$Plate)
scData <- StashIdent(scData, save.name = "cluster")
scData <- SetAllIdent(scData, id = "cluster.plate")
uniquePlates <- unique(scData@meta.data$Plate)

diffGenes <- list()
for(eachCluster in gtools::mixedsort(unique(scData@meta.data$cluster))){
  markersEach <- try(FindMarkers(scData, ident.1=paste0(eachCluster, "_", uniquePlates[1]),
                                 ident.2=paste0(eachCluster, "_", uniquePlates[2]),
                                 print.bar=FALSE), silent=TRUE)
  ## to skip some groups with few cells
  if(class(markersEach) != "try-error"){
    diffGenes[[eachCluster]] <- as_tibble(markersEach, rownames="gene")
    
    cat("\n")
    cat("#### Top 10 differential expressed genes from cluster ", eachCluster)
    cat("\n")
  
    FeatureHeatmap(scData, features.plot=head(rownames(markersEach), 10),
                   group.by="Plate", pt.size=0.25, key.position="top",
                   max.exp=3)
    cat("\n")
  }
}
diffGenes <- bind_rows(diffGenes, .id="cluster")
diffGenesFn <- "diffExpGenesPerCluster.tsv"
write_tsv(diffGenes, path=diffGenesFn)
```

```{r diff genes table, echo=FALSE}
caption ="Differential expressed genes per cluster"
ezInteractiveTableRmd(diffGenes, title=caption)
```

List of differential expressed genes per cluster can be accessed [here](`r diffGenesFn`).

The comparison is `r paste(uniquePlates, collapse=" vs ")`.

### Data availability
```{r save, echo=FALSE}
metadata(sce_cbind)$scData <- scData

sceFn <- "sce.rds"
saveRDS(sce_cbind, file=sceFn)
```

```{r gene mean per cluster, echo=FALSE}
tr_cnts <- expm1(scData@data)
geneMeans <- rowsum(t(as.matrix(tr_cnts)), group=scData@ident)
geneMeans <- sweep(geneMeans, 1,
                   STATS=table(scData@ident)[rownames(geneMeans)], FUN="/")
geneMeans <- log1p(t(geneMeans))
colnames(geneMeans) <- paste("cluster", colnames(geneMeans), sep="_")
geneMeanPerClusterFn <- "gene_means_per_cluster.txt"
ezWrite.table(geneMeans, geneMeanPerClusterFn)

geneMeans <- Matrix::rowMeans(tr_cnts)
geneMeans <- log1p(geneMeans)
geneMeansFn <- "gene_means.txt"
ezWrite.table(geneMeans, geneMeansFn)

tSNE_data <- as_tibble(scData@dr$tsne@cell.embeddings,
                       rownames="cells")
tSNE_data <- dplyr::rename(tSNE_data, X=`tSNE_1`, Y=`tSNE_2`)
tSNE_data$cluster <- scData@ident
tSNEFn <- "tSNE_data.tsv"
write_tsv(tSNE_data, path=tSNEFn)
```

[`r geneMeanPerClusterFn`](`r geneMeanPerClusterFn`)

[`r geneMeansFn`](`r geneMeansFn`)

[`r tSNEFn`](`r tSNEFn`)

The SingleCellExperiment object with final Seurat object after merging is [here](`r sceFn`).

### Parameters
```{r report parameters, echo=FALSE}
param[c("batchCorrection", "cc", "resolution",
        "chosenClusters1", "chosenClusters2")]
```

```{r seurat parameters, echo=FALSE}
PrintFindClustersParams(object = scData)
```

### SessionInfo
```{r, echo=FALSE}
sessionInfo()
```
