---
title: "RNAVelocity"
author: "Functional Genomics Center Zurich"
output: 
  html_document:
    self_contained: false
    includes:
      in_header: fgcz_header.html
    css: fgcz.css
editor_options: 
  chunk_output_type: console
---

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

```{r setup, include=FALSE}
## input for this report: ldat, emb, param
knitr::opts_chunk$set(echo = TRUE)
require(velocyto.R)

ldat <- ans$ldat
emb <- ans$emb
fit.quantile <- 0.05
```

## RNA Velocity Results {.tabset}

### Gene filtering
Spliced expression magnitude distribution across genes:

```{r gene filtering, echo=FALSE, results='markup'}
# exonic read (spliced) expression matrix
emat <- ldat$spliced;
# intronic read (unspliced) expression matrix
nmat <- ldat$unspliced
# spanning read (intron+exon) expression matrix
smat <- ldat$spanning;
# filter expression matrices based on some minimum max-cluster averages
emat <- filter.genes.by.cluster.expression(emat,cell.colors,
                                           min.max.cluster.average = 5)
nmat <- filter.genes.by.cluster.expression(nmat,cell.colors,
                                           min.max.cluster.average = 1)
smat <- filter.genes.by.cluster.expression(smat,cell.colors,
                                           min.max.cluster.average = 0.5)
# look at the resulting gene set
length(intersect(rownames(emat),rownames(nmat)))
# and if we use spanning reads (smat)
length(intersect(intersect(rownames(emat),rownames(nmat)),rownames(smat)))
```

```{r gene filtering plot, echo=FALSE}
hist(log10(rowSums(ldat$spliced)+1), col='wheat',
     xlab='log10[ number of reads + 1]',
     main='number of reads per gene')
```

### Velocity estimates variant 1
Velocity estimates using gene-relative model.

We’ll start with what is perhaps the most robust estimate, that combines cell kNN pooling with the gamma fit based on an extreme quantiles:

Using min/max quantile fit, in which case gene-specific offsets do not require spanning read (smat) fit. Here the fit is based on the top/bottom 5% of cells (by spliced expression magnitude).

```{r velocity 1 estimate, echo=FALSE}
rvel.qf <- gene.relative.velocity.estimates(emat, nmat, deltaT=1,
                                            kCells=5,
                                            fit.quantile=fit.quantile)
```

We visualize the velocities by projecting observed and extrapolated cells onto the first 5 PCs:

```{r velocity 1 PCA, echo=FALSE}
pca.velocity.plot(rvel.qf, nPcs=5, plot.cols=2,
                  cell.colors=ac(cell.colors,alpha=0.7),
                  cex=1.2, pcount=0.1, pc.multipliers=c(1,-1,-1,-1,-1))
```

#### Fitting of individual genes
Show the fitting of individual genes if given.

```{r fitting gene 1, echo=FALSE, fig.width=14}
# define custom pallet for expression magnitude
if(ezIsSpecified(param$markersToCheck)){
  for(theGene in param$markersToCheck){
    ## Some of the genes may not be available.
    try(gene.relative.velocity.estimates(emat,nmat, kCells = 5,
                                         fit.quantile=fit.quantile,
                                         old.fit=rvel.qf,
                                         show.gene=theGene,
                                         cell.emb=emb,
                                         cell.colors=cell.colors))
  }
}
```

#### Visualization on an existing embedding
Here we use t-SNE embedding from the upstream clustering.

```{r velocity 1 tsne, echo=FALSE}
show.velocity.on.embedding.cor(emb, rvel.qf, n=100, scale='sqrt',
                               cell.colors=ac(cell.colors,
                                              alpha=cell.alpha),
                               cex=cell.cex, arrow.scale=arrow.scale,
                               arrow.lwd=1)
legend("topleft", legend=names(cell.colorsPalette),
       pch=19, col=cell.colorsPalette)
```

#### Cell trajectory modeling
A similar function can be used to model central trajectories by directed diffusion on embedding.

```{r velocity 1 tragectory, echo=FALSE}
x <- show.velocity.on.embedding.eu(emb, rvel.qf, n=40, scale='sqrt',
                                   cell.colors=ac(cell.colors,
                                                  alpha=cell.alpha),
                                   cex=cell.cex, nPcs=30, sigma=2.5,
                                   show.trajectories=TRUE,
                                   diffusion.steps=400,
                                   n.trajectory.clusters=15,
                                   ntop.trajectories=1, embedding.knn=T,
                                   control.for.neighborhood.density=TRUE,
                                   n.cores=param$cores)
legend("topleft",legend=names(cell.colorsPalette),
       pch=19, col=cell.colorsPalette)
```

### Velocity estimates variant 2
Use spanning reads (smat) to fit the gene offsets. This will result in more accurate offset estimates, but for much fewer genes (spanning reads are rare).

```{r velocity 2 estimate, echo=FALSE}
rvel <- gene.relative.velocity.estimates(emat,nmat,smat=smat,
                                         kCells=5,
                                         fit.quantile=fit.quantile,
                                         diagonal.quantiles=TRUE)
```

We visualize the velocities by projecting observed and extrapolated cells onto the first 5 PCs:

```{r velocity 2 PCA, echo=FALSE}
pca.velocity.plot(rvel, nPcs=5, plot.cols=2, 
                  cell.colors=ac(cell.colors,alpha=0.7),
                  cex=1.2, pcount=0.1, pc.multipliers=c(1,-1,1,1,1))
```

#### Fitting of individual genes
Show the fitting of individual genes if given.

```{r fitting gene 2, echo=FALSE, fig.width=14}
# define custom pallet for expression magnitude
if(ezIsSpecified(param$markersToCheck)){
  for(theGene in param$markersToCheck){
    ## Some of the genes may not be available.
    try(gene.relative.velocity.estimates(emat,nmat, kCells = 5,
                                         fit.quantile=fit.quantile,
                                         old.fit=rvel,
                                         show.gene=theGene,
                                         cell.emb=emb,
                                         cell.colors=cell.colors))
  }
}
```

#### Visualization on an existing embedding
Here we use t-SNE embedding from the upstream clustering.

```{r velocity 2 tsne, echo=FALSE}
show.velocity.on.embedding.cor(emb, rvel, n=100, scale='sqrt',
                               cell.colors=ac(cell.colors,
                                              alpha=cell.alpha),
                               cex=cell.cex, arrow.scale=arrow.scale,
                               arrow.lwd=1)
legend("topleft", legend=names(cell.colorsPalette),
       pch=19, col=cell.colorsPalette)
```

#### Cell trajectory modeling
A similar function can be used to model central trajectories by directed diffusion on embedding.

```{r velocity 2 tragectory, echo=FALSE}
x <- show.velocity.on.embedding.eu(emb, rvel, n=40, scale='sqrt',
                                   cell.colors=ac(cell.colors,
                                                  alpha=cell.alpha),
                                   cex=cell.cex, nPcs=30, sigma=2.5,
                                   show.trajectories=TRUE,
                                   diffusion.steps=400,
                                   n.trajectory.clusters=15,
                                   ntop.trajectories=1, embedding.knn=T,
                                   control.for.neighborhood.density=TRUE,
                                   n.cores=param$cores)
legend("topleft",legend=names(cell.colorsPalette),
       pch=19, col=cell.colorsPalette)
```


### Velocity estimates variant 3
Here we calculate the most basic version of velocity estimates, using relative gamma fit, without cell kNN smoothing (i.e. actual single-cell velocity):

```{r velocity 3 estimate, echo=FALSE}
rvel1 <- gene.relative.velocity.estimates(emat, nmat, deltaT=1,
                                          deltaT2=1, kCells=1,
                                          fit.quantile=fit.quantile)

```

```{r velocity 3 PCA, echo=FALSE}
pca.velocity.plot(rvel1, nPcs=5, plot.cols=2, 
                  cell.colors=ac(cell.colors,alpha=0.7),
                  cex=1.2, pcount=0.1, pc.multipliers=c(1,-1,1,1,1))
```

#### Fitting of individual genes
Show the fitting of individual genes if given.

```{r fitting gene 3, echo=FALSE, fig.width=14}
# define custom pallet for expression magnitude
if(ezIsSpecified(param$markersToCheck)){
  for(theGene in param$markersToCheck){
    ## Some of the genes may not be available.
    try(gene.relative.velocity.estimates(emat,nmat, kCells = 5,
                                         fit.quantile=fit.quantile,
                                         old.fit=rvel1,
                                         show.gene=theGene,
                                         cell.emb=emb,
                                         cell.colors=cell.colors))
  }
}
```

#### Visualization on an existing embedding
Here we use t-SNE embedding from the upstream clustering.

```{r velocity 3 tsne, echo=FALSE}
show.velocity.on.embedding.cor(emb, rvel1, n=100, scale='sqrt',
                               cell.colors=ac(cell.colors,
                                              alpha=cell.alpha),
                               cex=cell.cex, arrow.scale=arrow.scale,
                               arrow.lwd=1)
legend("topleft", legend=names(cell.colorsPalette),
       pch=19, col=cell.colorsPalette)
```

#### Cell trajectory modeling
A similar function can be used to model central trajectories by directed diffusion on embedding.

```{r velocity 3 tragectory, echo=FALSE}
x <- show.velocity.on.embedding.eu(emb, rvel1, n=40, scale='sqrt',
                                   cell.colors=ac(cell.colors,
                                                  alpha=cell.alpha),
                                   cex=cell.cex, nPcs=30, sigma=2.5,
                                   show.trajectories=TRUE,
                                   diffusion.steps=400,
                                   n.trajectory.clusters=15,
                                   ntop.trajectories=1, embedding.knn=T,
                                   control.for.neighborhood.density=TRUE,
                                   n.cores=param$cores)
legend("topleft",legend=names(cell.colorsPalette),
       pch=19, col=cell.colorsPalette)
```