---
title: "`r ans$param$name`"
author: "Functional Genomics Center Zurich"
output: 
  html_document:
    self_contained: false
    includes:
      in_header: fgcz_header.html
    css: fgcz.css
editor_options: 
  chunk_output_type: console
---

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

```{r setup, include=FALSE}
## input for this report: ldat, tSNE_data, param
knitr::opts_chunk$set(echo = TRUE)

require(ezRun)
require(velocyto.R)
require(Matrix)

# debug
# title: "`r ans$param$name`"
# 10X
# ans <- readRDS("/export/local/scratch/gtan/dev/RNAVelocity-p2284/SCReport_SCRNAVelocity/ans.rds")
# ans <- readRDS("/scratch/gtan/dev/quickdev/ans.rds")
# end of debug

ldat <- ans$ldat
tSNE_data <- ans$tSNE_data
cell.dist <- ans$cell.dist
param <- ans$param

cell.colorsPalette <- setNames(gg_color_hue(length(levels(tSNE_data$cluster))),
                               levels(tSNE_data$cluster))
cell.colors <- cell.colorsPalette[tSNE_data$cluster]
names(cell.colors) <- tSNE_data$cells
emb <- data.frame(row.names=tSNE_data$cells,
                  tSNE_data[ ,c("X", "Y")])
emb <- as.matrix(emb)

if(param$scProtocol == "smart-Seq2"){
  emat_cutoff <- 5
  nmat_cutoff <- 1
  smat_cutoff <- 0.5
  fit.quantile <- 0.05
  kCells <- 5
  cell.cex <- 1
  arrow.scale <- 3
  show.grid.flow <- FALSE
  grid.n <- 20
}else if(param$scProtocol == "10X"){
  emat_cutoff <- 0.5
  nmat_cutoff <- 0.05
  smat_cutoff <- 0.005
  fit.quantile <- 0.02
  kCells <- 20
  cell.cex <- 0.8
  arrow.scale <- 5
  show.grid.flow <- TRUE
  grid.n <- 40
}else{
  stop("Unsupported single cell protocol!")
}
cell.alpha <- 0.7
```

```{r gene filtering processing, echo=FALSE}
# exonic read (spliced) expression matrix
emat <- ldat$spliced
# intronic read (unspliced) expression matrix
nmat <- ldat$unspliced
# spanning read (intron+exon) expression matrix
smat <- ldat$spanning

# Fix the barcode for 10X data
if(param$scProtocol == "10X"){
  colnames(emat) <- sub("x$", "-1", colnames(emat))
  colnames(nmat) <- sub("x$", "-1", colnames(nmat))
  if(!is.null(smat)){
    colnames(smat) <- sub("x$", "-1", colnames(smat))
  }
}

# restrict to cells that from filtering upstream
emat <- emat[ , rownames(emb)]
nmat <- nmat[ , rownames(emb)]
if(!is.null(smat)){
  smat <- smat[ , rownames(emb)]
}

# filter expression matrices based on some minimum max-cluster averages
emat <- filter.genes.by.cluster.expression(emat, cell.colors,
                                           min.max.cluster.average = emat_cutoff)
nmat <- filter.genes.by.cluster.expression(nmat, cell.colors,
                                           min.max.cluster.average = nmat_cutoff)
if(!is.null(smat)){
  smat <- filter.genes.by.cluster.expression(smat, cell.colors,
                                             min.max.cluster.average = smat_cutoff)
}
```

## RNA Velocity Results {.tabset}

### Gene filtering
Spliced expression magnitude distribution across genes:

```{r gene filtering plot, echo=FALSE}
hist(log10(Matrix::rowSums(ldat$spliced)+1), col='wheat',
     xlab='log10[ number of reads + 1]',
     main='number of reads per gene')
```

```{r gene filtering output, echo=FALSE, results='markup'}
cat("The numner of genes after filtering by required minimum average expression count for exonic (spliced) and intronic (unspliced) reads: \n")
length(intersect(rownames(emat), rownames(nmat)))
if(!is.null(smat)){
  cat("The numner of genes after filtering by required minimum average expression count for exonic (spliced), intronic (unspliced) reads and spanning (intron+exon) reads: \n")
  length(intersect(intersect(rownames(emat), rownames(nmat)), rownames(smat)))
}
```

### Velocity estimates variant 1
Velocity estimates using gene-relative model.

We’ll start with what is perhaps the most robust estimate, that combines cell kNN pooling with the gamma fit based on an extreme quantiles:

Using min/max quantile fit, in which case gene-specific offsets do not require spanning read (smat) fit. Here the fit is based on the top/bottom 5% of cells (by spliced expression magnitude).

```{r velocity 1 estimate, echo=FALSE, message=FALSE, results='hide'}
rvel.qf <- gene.relative.velocity.estimates(emat, nmat, deltaT=1,
                                            kCells=kCells,
                                            cell.dist=cell.dist,
                                            fit.quantile=fit.quantile)
```

We visualize the velocities by projecting observed and extrapolated cells onto the first 5 PCs:

```{r velocity 1 PCA, echo=FALSE, message=FALSE, results='hide', fig.width=14, fig.height=14}
pca.velocity.plot(rvel.qf, nPcs=5, plot.cols=2,
                  cell.colors=ac(cell.colors, alpha=0.7),
                  cex=1.2, pcount=0.1, pc.multipliers=c(1,-1,-1,-1,-1),
                  show.grid.flow=TRUE, min.grid.cell.mass=0.5,
                  grid.n=grid.n, arrow.lwd=2)
```

#### Fitting of individual genes
Show the fitting of individual genes if given.

```{r fitting gene 1, echo=FALSE, message=FALSE, results='hide', fig.width=14}
# define custom pallet for expression magnitude
if(ezIsSpecified(param$markersToCheck)){
  for(theGene in param$markersToCheck){
    ## Some of the genes may not be available.
    try(gene.relative.velocity.estimates(emat,nmat, kCells = kCells,
                                         cell.dist=cell.dist,
                                         fit.quantile=fit.quantile,
                                         old.fit=rvel.qf,
                                         show.gene=theGene,
                                         cell.emb=emb,
                                         cell.colors=cell.colors))
  }
}
```

#### Visualization on an existing embedding
Here we use t-SNE embedding from the upstream clustering.

```{r velocity 1 tsne, echo=FALSE, results='hide', message=FALSE, fig.width=10, fig.height=8}
show.velocity.on.embedding.cor(emb, rvel.qf, n=100, scale='sqrt',
                               cell.colors=ac(cell.colors,
                                              alpha=cell.alpha),
                               cex=cell.cex, arrow.scale=arrow.scale,
                               arrow.lwd=1)
legend("topleft", legend=names(cell.colorsPalette),
       pch=19, col=cell.colorsPalette)

show.velocity.on.embedding.cor(emb, rvel.qf, n=100, scale='sqrt',
                               cell.colors=ac(cell.colors,
                                              alpha=cell.alpha),
                               cex=cell.cex, arrow.scale=arrow.scale,
                               show.grid.flow=TRUE,
                               min.grid.cell.mass=0.5, grid.n=grid.n,
                               arrow.lwd=2)
legend("topleft", legend=names(cell.colorsPalette),
       pch=19, col=cell.colorsPalette)
```

#### Cell trajectory modeling
A similar function can be used to model central trajectories by directed diffusion on embedding.
This is disable on 10X data.

```{r velocity 1 tragectory, echo=FALSE, message=FALSE, results='hide', eval=TRUE, fig.width=10, fig.height=8}
if(param$scProtocol != "10X"){
x <- show.velocity.on.embedding.eu(emb, rvel.qf, n=40, scale='sqrt',
                                   cell.colors=ac(cell.colors,
                                                  alpha=cell.alpha),
                                   cex=cell.cex, nPcs=30, sigma=2.5,
                                   show.trajectories=TRUE,
                                   diffusion.steps=400,
                                   n.trajectory.clusters=15,
                                   ntop.trajectories=1, embedding.knn=T,
                                   control.for.neighborhood.density=TRUE,
                                   n.cores=param$cores)
legend("topleft", legend=names(cell.colorsPalette),
       pch=19, col=cell.colorsPalette)
}
```

### Velocity estimates variant 2
Use spanning reads (smat) to fit the gene offsets. This will result in more accurate offset estimates, but for much fewer genes (spanning reads are rare).
This variant is not available for 10X data.

```{r velocity 2 estimate, echo=FALSE, message=FALSE, results='hide'}
if(!is.null(smat)){
  rvel <- gene.relative.velocity.estimates(emat,nmat,smat=smat,
                                           kCells=kCells,
                                           cell.dist=cell.dist,
                                           fit.quantile=fit.quantile,
                                           diagonal.quantiles=TRUE)
}
```

We visualize the velocities by projecting observed and extrapolated cells onto the first 5 PCs:

```{r velocity 2 PCA, echo=FALSE, message=FALSE, results='hide', fig.width=14, fig.height=14}
if(!is.null(smat)){
pca.velocity.plot(rvel, nPcs=5, plot.cols=2, 
                  cell.colors=ac(cell.colors,alpha=0.7),
                  cex=1.2, pcount=0.1, pc.multipliers=c(1,-1,1,1,1),
                  show.grid.flow=TRUE, min.grid.cell.mass=0.5,
                  grid.n=grid.n, arrow.lwd=2)
}
```

#### Fitting of individual genes
Show the fitting of individual genes if given.

```{r fitting gene 2, echo=FALSE, message=FALSE, results='hide', fig.width=14}
# define custom pallet for expression magnitude
if(!is.null(smat)){
if(ezIsSpecified(param$markersToCheck)){
  for(theGene in param$markersToCheck){
    ## Some of the genes may not be available.
    try(gene.relative.velocity.estimates(emat,nmat, kCells = kCells,
                                         cell.dist=cell.dist,
                                         fit.quantile=fit.quantile,
                                         old.fit=rvel,
                                         show.gene=theGene,
                                         cell.emb=emb,
                                         cell.colors=cell.colors))
  }
}
}
```

#### Visualization on an existing embedding
Here we use t-SNE embedding from the upstream clustering.

```{r velocity 2 tsne, echo=FALSE, message=FALSE, results='hide', fig.width=10, fig.height=8}
if(!is.null(smat)){
show.velocity.on.embedding.cor(emb, rvel, n=100, scale='sqrt',
                               cell.colors=ac(cell.colors,
                                              alpha=cell.alpha),
                               cex=cell.cex, arrow.scale=arrow.scale,
                               arrow.lwd=1)
legend("topleft", legend=names(cell.colorsPalette),
       pch=19, col=cell.colorsPalette)

show.velocity.on.embedding.cor(emb, rvel, n=100, scale='sqrt',
                               cell.colors=ac(cell.colors,
                                              alpha=cell.alpha),
                               cex=cell.cex, arrow.scale=arrow.scale,
                               show.grid.flow=TRUE, min.grid.cell.mass=0.5,
                               grid.n=grid.n, arrow.lwd=2)
legend("topleft", legend=names(cell.colorsPalette),
       pch=19, col=cell.colorsPalette)
}
```

#### Cell trajectory modeling
A similar function can be used to model central trajectories by directed diffusion on embedding.
This is disable on 10X data.

```{r velocity 2 tragectory, echo=FALSE, message=FALSE, results='hide', eval=TRUE, fig.width=10, fig.height=8}
if(!is.null(smat)){
  if(param$scProtocol != "10X"){
x <- show.velocity.on.embedding.eu(emb, rvel, n=40, scale='sqrt',
                                   cell.colors=ac(cell.colors,
                                                  alpha=cell.alpha),
                                   cex=cell.cex, nPcs=30, sigma=2.5,
                                   show.trajectories=TRUE,
                                   diffusion.steps=400,
                                   n.trajectory.clusters=15,
                                   ntop.trajectories=1, embedding.knn=T,
                                   control.for.neighborhood.density=TRUE,
                                   n.cores=param$cores)
legend("topleft",legend=names(cell.colorsPalette),
       pch=19, col=cell.colorsPalette)
}
}
```

### Velocity estimates variant 3
Here we calculate the most basic version of velocity estimates, using relative gamma fit, without cell kNN smoothing (i.e. actual single-cell velocity).
This is not suitable for 10X data.

```{r velocity 3 estimate, echo=FALSE, results='hide', message=FALSE}
rvel1 <- gene.relative.velocity.estimates(emat, nmat, deltaT=1,
                                          deltaT2=1, kCells=1,
                                          cell.dist=cell.dist,
                                          fit.quantile=fit.quantile)

```

```{r velocity 3 PCA, echo=FALSE, results='hide', message=FALSE, fig.width=14, fig.height=14}
pca.velocity.plot(rvel1, nPcs=5, plot.cols=2, 
                  cell.colors=ac(cell.colors,alpha=0.7),
                  cex=1.2, pcount=0.1, pc.multipliers=c(1,-1,1,1,1),
                  show.grid.flow=TRUE, min.grid.cell.mass=0.5,
                  grid.n=grid.n, arrow.lwd=2)
```

#### Fitting of individual genes
Show the fitting of individual genes if given.

```{r fitting gene 3, echo=FALSE, message=FALSE, results='hide', fig.width=14}
# define custom pallet for expression magnitude
if(ezIsSpecified(param$markersToCheck)){
  for(theGene in param$markersToCheck){
    ## Some of the genes may not be available.
    try(gene.relative.velocity.estimates(emat,nmat, kCells = 1,
                                         cell.dist=cell.dist,
                                         fit.quantile=fit.quantile,
                                         old.fit=rvel1,
                                         show.gene=theGene,
                                         cell.emb=emb,
                                         cell.colors=cell.colors))
  }
}
```

#### Visualization on an existing embedding
Here we use t-SNE embedding from the upstream clustering.

```{r velocity 3 tsne, echo=FALSE, message=FALSE, results='hide', fig.width=10, fig.height=10}
show.velocity.on.embedding.cor(emb, rvel1, n=100, scale='sqrt',
                               cell.colors=ac(cell.colors,
                                              alpha=cell.alpha),
                               cex=cell.cex, arrow.scale=arrow.scale,
                               arrow.lwd=1)
legend("topleft", legend=names(cell.colorsPalette),
       pch=19, col=cell.colorsPalette)

show.velocity.on.embedding.cor(emb, rvel1, n=100, scale='sqrt',
                               cell.colors=ac(cell.colors,
                                              alpha=cell.alpha),
                               cex=cell.cex, arrow.scale=arrow.scale,
                               show.grid.flow=TRUE,
                               min.grid.cell.mass=0.5, grid.n=grid.n,
                               arrow.lwd=2)
legend("topleft", legend=names(cell.colorsPalette),
       pch=19, col=cell.colorsPalette)
```

#### Cell trajectory modeling
A similar function can be used to model central trajectories by directed diffusion on embedding.
This is disable on 10X data.

```{r velocity 3 tragectory, echo=FALSE, message=FALSE, results='hide', eval=TRUE, fig.width=10, fig.height=8}
if(param$scProtocol != "10X"){
x <- show.velocity.on.embedding.eu(emb, rvel1, n=40, scale='sqrt',
                                   cell.colors=ac(cell.colors,
                                                  alpha=cell.alpha),
                                   cex=cell.cex, nPcs=30, sigma=2.5,
                                   show.trajectories=TRUE,
                                   diffusion.steps=400,
                                   n.trajectory.clusters=15,
                                   ntop.trajectories=1, embedding.knn=T,
                                   control.for.neighborhood.density=TRUE,
                                   n.cores=param$cores)
legend("topleft",legend=names(cell.colorsPalette),
       pch=19, col=cell.colorsPalette)
}
```

### SessionInfo
```{r, echo=FALSE}
sessionInfo()
```
