---
params:
  reportTitle: "SpliceWiz Alternative Splicing"
title: "`r params$reportTitle`"
---

<!-- Styling comes from the fgczQuartoTemplate assets staged next to this file
     (_metadata.yml is applied by Quarto automatically). ezRun stages them via
     makeQuartoReport(), which calls fgczQuartoTemplate::fgcz_render(). The report
     reads the objects saved by makeQuartoReport() from the working directory. -->

```{r setup}
#| include: false
library(ezRun)
library(SpliceWiz)
library(SummarizedExperiment)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(DT)

## ── Data loading (qs2 with rds fallback) ─────────────────────────────────
.fgcz_load <- function(name, nthreads = 4L) {
  paths <- c(
    paste0(name, ".qs2"), file.path("data", paste0(name, ".qs2")),
    paste0(name, ".rds"), file.path("data", paste0(name, ".rds"))
  )
  for (p in paths) {
    if (file.exists(p)) {
      if (grepl("\\.qs2$", p)) return(qs2::qs_read(p, nthreads = nthreads))
      return(readRDS(p))
    }
  }
  stop("No .qs2 or .rds file found for: ", name)
}

param     <- .fgcz_load("param")
ase       <- .fgcz_load("aseResult")
se        <- .fgcz_load("se")

sampleGroup <- param$sampleGroup
refGroup    <- param$refGroup
fdrCut      <- as.numeric(param$FDR)
dpsiCut     <- as.numeric(param$deltaPSI)
topN        <- as.integer(param$topN)
aseMethod   <- param$aseMethod

## Colour-blind-safe qualitative palette (Okabe-Ito) — the user has red-green CVD,
## so event types are distinguished by these colours *and* by direct labels.
cbPalette <- c(
  "#0072B2", "#E69F00", "#009E73", "#CC79A7",
  "#56B4E9", "#D55E00", "#F0E442", "#999999"
)

## The 7 canonical SpliceWiz event types, in a fixed display order.
## SpliceWiz labels intron retention "IR" (not "RI"). Keep any unexpected types
## as their own levels rather than dropping them to NA.
eventLevels <- c("SE", "MXE", "A5SS", "A3SS", "IR", "AFE", "ALE")
knownLevels <- intersect(eventLevels, unique(ase$EventType))
otherLevels <- setdiff(unique(ase$EventType), eventLevels)
ase$EventType <- factor(ase$EventType, levels = c(knownLevels, otherLevels))

## Significance call: adjusted p below cutoff AND effect size above |dPSI| cutoff.
ase$significant <- !is.na(ase$FDR) & ase$FDR < fdrCut &
  !is.na(ase$abs_deltaPSI) & ase$abs_deltaPSI > dpsiCut
ase$direction <- ifelse(ase$deltaPSI > 0,
  paste0("up in ", sampleGroup), paste0("up in ", refGroup))
```

Rendered `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`.

::: {.panel-tabset}

# Overview

Differential alternative splicing for **`r sampleGroup`** vs **`r refGroup`**,
computed with **SpliceWiz** (`ASE_`\ `r aseMethod`).

- Reference build: **`r param$refBuild`**
- Samples analysed: **`r ncol(se)`**  (`r sum(colData(se)$condition == sampleGroup)` `r sampleGroup`, `r sum(colData(se)$condition == refGroup)` `r refGroup`)
- Events tested (after filtering): **`r nrow(ase)`**
- Significance: adjusted p < **`r fdrCut`** and |ΔPSI| > **`r dpsiCut`**
- Significant events: **`r sum(ase$significant)`**

ΔPSI is the difference in Percent-Spliced-In between the two groups
(positive = higher inclusion in **`r sampleGroup`**). The full result table is
written to `SpliceWiz_ASE_all.txt`, with one file per event type
(`SpliceWiz_ASE_<EventType>.txt`); the filtered `NxtSE` is in
`spliceWiz_filtered_se.rds`.

# Event-type breakdown

```{r breakdown}
#| echo: false
brk <- ase |>
  group_by(EventType) |>
  summarise(
    Tested = n(),
    Significant = sum(significant),
    `Up in sampleGroup` = sum(significant & deltaPSI > 0),
    `Up in refGroup` = sum(significant & deltaPSI < 0),
    .groups = "drop"
  )
DT::datatable(
  brk,
  caption = paste0("Events per type. 'Significant' = adj.p < ", fdrCut,
                   " and |ΔPSI| > ", dpsiCut, "."),
  options = list(dom = "t", pageLength = 20), rownames = FALSE
)
```

```{r breakdownPlot}
#| echo: false
#| fig-width: 8
#| fig-height: 4.5
brkLong <- brk |>
  select(EventType, Significant) |>
  filter(!is.na(EventType))
if (nrow(brkLong) > 0 && sum(brkLong$Significant) > 0) {
  ggplot(brkLong, aes(x = EventType, y = Significant, fill = EventType)) +
    geom_col() +
    geom_text(aes(label = Significant), vjust = -0.3, size = 3.5) +
    scale_fill_manual(values = cbPalette, guide = "none") +
    labs(x = "Event type", y = "Significant events",
         title = "Significant differential splicing events by type") +
    theme_bw()
} else {
  cat("No significant events to plot.\n")
}
```

# Volcano

```{r volcano}
#| echo: false
#| fig-width: 8
#| fig-height: 6
vol <- ase |> filter(!is.na(FDR), !is.na(deltaPSI))
if (nrow(vol) > 0) {
  vol$negLog10FDR <- -log10(pmax(vol$FDR, .Machine$double.xmin))
  vol$colourKey <- ifelse(vol$significant, as.character(vol$EventType), "n.s.")
  ## build a colour for every event type actually present (interpolate if there
  ## are more than the base palette), so scale_colour_manual never errors on an
  ## unmapped value.
  presentTypes <- setdiff(unique(vol$colourKey), "n.s.")
  typeColours <- setNames(
    grDevices::colorRampPalette(cbPalette)(max(length(presentTypes), 1))[seq_along(presentTypes)],
    presentTypes
  )
  colourValues <- c(typeColours, "n.s." = "grey75")
  ## label the top few significant events by effect size
  lab <- vol |> filter(significant) |> arrange(desc(abs_deltaPSI)) |> head(15)
  ggplot(vol, aes(x = deltaPSI, y = negLog10FDR)) +
    geom_point(aes(colour = colourKey), alpha = 0.6, size = 1.4) +
    geom_hline(yintercept = -log10(fdrCut), linetype = "dashed", colour = "grey40") +
    geom_vline(xintercept = c(-dpsiCut, dpsiCut), linetype = "dashed", colour = "grey40") +
    ggrepel::geom_text_repel(data = lab, aes(label = EventName),
                             size = 2.6, max.overlaps = 20) +
    scale_colour_manual(values = colourValues, name = "Event type") +
    labs(x = paste0("ΔPSI  (", sampleGroup, " − ", refGroup, ")"),
         y = "−log10 adjusted p") +
    theme_bw()
} else {
  cat("No testable events for the volcano plot.\n")
}
```

# Top events

```{r topTable}
#| echo: false
keepCols <- intersect(
  c("EventName", "EventType", "EventRegion", "deltaPSI", "abs_deltaPSI",
    "PValue", "FDR", "flags"),
  colnames(ase)
)
tbl <- ase |>
  filter(significant) |>
  arrange(desc(abs_deltaPSI)) |>
  select(all_of(keepCols))
if (nrow(tbl) == 0) {
  cat("No significant events at the current thresholds.\n")
} else {
  numCols <- intersect(c("deltaPSI", "abs_deltaPSI", "PValue", "FDR"), colnames(tbl))
  DT::datatable(
    tbl, rownames = FALSE, filter = "top",
    caption = "Significant events, ranked by |ΔPSI|.",
    options = list(pageLength = 25, scrollX = TRUE)
  ) |> DT::formatSignif(columns = numCols, digits = 3)
}
```

# Coverage plots

Normalized-coverage tracks for the top **`r topN`** significant events (by |ΔPSI|),
averaged per group.

```{r coverage}
#| echo: false
#| results: asis
#| fig-width: 10
#| fig-height: 5
topEvents <- ase |>
  filter(significant) |>
  arrange(desc(abs_deltaPSI)) |>
  head(topN)
if (nrow(topEvents) == 0) {
  cat("\nNo significant events to draw coverage for.\n")
} else {
  cat("\n::: {.panel-tabset}\n")
  for (i in seq_len(nrow(topEvents))) {
    ev <- topEvents$EventName[i]
    cat("\n\n## ", ev, "\n\n", sep = "")
    ## Build AND render inside the tryCatch: plotCoverage can succeed yet the
    ## patchwork fail to print (e.g. a ggplot build/scale error), and an
    ## uncaught error here aborts the whole quarto render.
    ok <- tryCatch(
      {
        p <- plotCoverage(
          se, Event = ev,
          tracks = c(refGroup, sampleGroup),
          condition = "condition",
          plotJunctions = TRUE
        )
        print(p)
        TRUE
      },
      error = function(e) {
        cat("_Coverage plot unavailable for this event: ",
            conditionMessage(e), "_\n\n", sep = "")
        FALSE
      }
    )
    if (ok) cat("\n\n")
  }
  cat("\n:::\n\n")
}
```

# SessionInfo

```{r sessionInfo}
#| echo: false
ezSessionInfo()
```

:::
