---
params:
  reportTitle: "Parse split-pipe"
title: "`r params$reportTitle`"
---

<!-- Styling comes from the fgczQuartoTemplate assets staged next to this file by
     makeQuartoReport() -> fgczQuartoTemplate::fgcz_render(); nothing to declare
     here. The report reads the split-pipe outputs from the working directory. -->

```{r setup}
#| include: false
library(ezRun)
library(DT)
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)

param <- ezLoadRobj("param")
wddir <- "."

## Colour-blind-safe qualitative palette (Okabe-Ito) - avoid red/green reliance.
cbPalette <- c(
  "#0072B2", "#E69F00", "#009E73", "#CC79A7",
  "#56B4E9", "#D55E00", "#F0E442", "#999999"
)

## Per-biological-sample output directories (those with a 10x/DGE matrix).
sampleDirs <- list.dirs(wddir, recursive = FALSE, full.names = FALSE)
sampleDirs <- sampleDirs[vapply(
  sampleDirs,
  function(d) dir.exists(file.path(wddir, d, "DGE_filtered")),
  logical(1)
)]

## Fetch a column from a split-pipe summary data.frame by exact name or, when the
## metric is genome-prefixed (e.g. "GRCh38-p14_median_genes_per_cell"), by suffix.
getMetric <- function(df, name) {
  if (name %in% names(df)) {
    return(df[[name]])
  }
  hit <- grep(paste0("_", name, "$"), names(df), value = TRUE)
  if (length(hit) >= 1) {
    return(df[[hit[1]]])
  }
  rep(NA, nrow(df))
}

## Parse a STAR Log.final.out into a named list of the fields we surface.
readStarFinalLog <- function(fn) {
  ll <- readLines(fn, warn = FALSE)
  grab <- function(key) {
    hit <- grep(key, ll, fixed = TRUE, value = TRUE)
    if (length(hit) == 0) {
      return(NA_character_)
    }
    trimws(sub(".*\\|", "", hit[1]))
  }
  list(
    input = grab("Number of input reads"),
    uniquePct = grab("Uniquely mapped reads %"),
    multiPct = grab("% of reads mapped to multiple loci")
  )
}
```

Rendered `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`.

::: {.panel-tabset}

# Overview

This report summarizes the **Parse Biosciences split-pipe** run **`r param$name`**.

- Kit: **`r param$kit`**, chemistry: **`r param$chemistry`**
- Reference: **`r param$refBuild`**
- Biological samples: **`r paste(sampleDirs, collapse = ", ")`**

# Summary metrics (10x-style)

```{r metricsTable}
#| echo: false
aggFn <- file.path(wddir, "agg_sample_summary.csv")
if (file.exists(aggFn)) {
  agg <- read_csv(aggFn, show_col_types = FALSE)

  ## Total genes detected per sample (from each sample's report dir).
  totalGenes <- vapply(agg$sample, function(s) {
    mFn <- file.path(wddir, s, "report", "sample_mrtg_counts.csv")
    if (file.exists(mFn)) {
      mm <- read_csv(mFn, show_col_types = FALSE)
      v <- mm$count[mm$statistic == "total_gene_count"]
      if (length(v)) {
        return(as.numeric(v[1]))
      }
    }
    NA_real_
  }, numeric(1))

  metrics <- tibble(
    Metric = c(
      "Estimated Number of Cells",
      "Mean Reads per Cell",
      "Number of Reads",
      "Sequencing Saturation",
      "Valid Barcode Fraction",
      "Reads Mapped to Transcriptome",
      "Median Genes per Cell",
      "Median UMIs (Transcripts) per Cell",
      "Fraction Reads in Cells",
      "Fraction Exonic",
      "Total Genes Detected",
      "cDNA Q30"
    )
  )
  for (i in seq_len(nrow(agg))) {
    metrics[[agg$sample[i]]] <- c(
      getMetric(agg, "number_of_cells")[i],
      round(getMetric(agg, "mean_reads_per_cell")[i], 1),
      getMetric(agg, "number_of_reads")[i],
      getMetric(agg, "sequencing_saturation")[i],
      getMetric(agg, "valid_barcode_fraction")[i],
      getMetric(agg, "transcriptome_map_fraction")[i],
      getMetric(agg, "median_genes_per_cell")[i],
      getMetric(agg, "median_tscp_per_cell")[i],
      getMetric(agg, "fraction_reads_in_cells")[i],
      getMetric(agg, "fraction_exonic")[i],
      totalGenes[i],
      getMetric(agg, "cDNA_Q30")[i]
    )
  }
  DT::datatable(
    metrics,
    caption = "Per-sample summary. 'Median UMIs (Transcripts) per Cell' is the Parse transcript count (= UMIs); mapping fractions are transcriptome-level.",
    options = list(dom = "t", pageLength = 20, scrollX = TRUE),
    rownames = FALSE
  )
} else {
  cat("No agg_sample_summary.csv found.\n")
}
```

# Alignment (per sublibrary)

Genome mapping rate comes from the STAR alignment log of each **sublibrary**
(these are run-level, not per biological sample).

```{r alignTable}
#| echo: false
starLogs <- list.files(
  wddir,
  pattern = "Log\\.final\\.out$",
  recursive = TRUE,
  full.names = TRUE
)
if (length(starLogs) > 0) {
  aln <- lapply(starLogs, function(fn) {
    s <- readStarFinalLog(fn)
    uniq <- suppressWarnings(as.numeric(sub("%", "", s$uniquePct)))
    multi <- suppressWarnings(as.numeric(sub("%", "", s$multiPct)))
    tibble(
      Sublibrary = basename(dirname(dirname(fn))),
      `Input Reads` = s$input,
      `Uniquely Mapped %` = s$uniquePct,
      `Multi-loci %` = s$multiPct,
      `Reads Mapped to Genome %` = ifelse(
        is.na(uniq) | is.na(multi),
        NA_character_,
        sprintf("%.2f%%", uniq + multi)
      )
    )
  })
  DT::datatable(
    bind_rows(aln),
    caption = "STAR alignment per sublibrary. 'Reads Mapped to Genome' = uniquely mapped + multi-loci.",
    options = list(dom = "t", scrollX = TRUE),
    rownames = FALSE
  )
} else {
  cat("No STAR Log.final.out found.\n")
}
```

# Per-cell distributions

Genes, UMIs (transcripts) and reads per cell, labeled by **biological sample**
(computed from each sample's `cell_metadata.csv` - these use the real sample
names, unlike split-pipe's native genome-keyed knee plots below).

```{r perCellPrep}
#| echo: false
meta <- lapply(sampleDirs, function(s) {
  fn <- file.path(wddir, s, "DGE_filtered", "cell_metadata.csv")
  if (file.exists(fn)) {
    d <- read_csv(fn, show_col_types = FALSE)
    d$sampleDir <- s
    d
  } else {
    NULL
  }
})
meta <- bind_rows(meta)
hasMeta <- !is.null(meta) && nrow(meta) > 0
plotPerCell <- function(var, ylab) {
  if (!hasMeta) {
    cat("No cell_metadata.csv found.\n")
    return(invisible(NULL))
  }
  ## cbPalette holds 8 colours and the fill is mapped to one level per biological
  ## sample, so a manual scale errors outright ("Insufficient values in manual
  ## scale. 92 needed but only 8 provided") as soon as a run has more samples than
  ## colours - a WT_mega plate has up to 96. Interpolate instead. The fill is
  ## decorative (guide = "none"), so a ramp loses nothing.
  n <- length(unique(meta$sampleDir))
  fills <- if (n <= length(cbPalette)) {
    cbPalette[seq_len(n)]
  } else {
    colorRampPalette(cbPalette)(n)
  }
  print(
    ggplot(meta, aes(x = sampleDir, y = .data[[var]], fill = sampleDir)) +
      geom_violin(scale = "width", trim = TRUE) +
      geom_boxplot(width = 0.1, outlier.size = 0.3, fill = "white") +
      scale_y_log10() +
      scale_fill_manual(values = fills, guide = "none") +
      labs(x = "Biological sample", y = ylab) +
      theme_bw() +
      ## 96 violins do not fit horizontal labels
      theme(axis.text.x = element_text(
        angle = 90, vjust = 0.5, hjust = 1,
        size = if (n > 48) 5 else if (n > 24) 7 else 9
      ))
  )
}
```

::: {.panel-tabset}

## Genes per cell

```{r pcGenes}
#| echo: false
#| fig-width: 14
#| fig-height: 5
plotPerCell("gene_count", "Genes per cell (log10)")
```

## UMIs per cell

```{r pcUmis}
#| echo: false
#| fig-width: 14
#| fig-height: 5
plotPerCell("tscp_count", "UMIs / transcripts per cell (log10)")
```

## Reads per cell

```{r pcReads}
#| echo: false
#| fig-width: 14
#| fig-height: 5
plotPerCell("mread_count", "Reads per cell (log10)")
```

:::

# split-pipe native figures

split-pipe's own QC figures. Note the knee-plot legends are keyed by **genome
name** (e.g. `GRCh38-p14`), suffixed by sublibrary index in combined runs - they
are per-genome/per-sublibrary QC, not per biological sample.

```{r nativeFigs}
#| echo: false
#| results: asis
## Every figure below also lives in that sample's own split-pipe HTML, which is
## linked per tab. Inlining them all is fine for a handful of samples and ruinous
## for a plate: 92 samples means ~1080 PNGs, and because the report is
## self-contained that produced a 256 MB 00index.html no browser will open. So
## embed only for small runs and rely on the links otherwise.
embedFigures <- length(sampleDirs) <= 8
if (!embedFigures) {
  cat(
    "\nThis run has ", length(sampleDirs), " samples. Their split-pipe figures are",
    " not inlined here, because doing so makes this page too large to open;",
    " follow the per-sample link in each tab instead.\n\n",
    sep = ""
  )
}
cat("\n::: {.panel-tabset}\n\n")
for (s in sampleDirs) {
  cat("\n\n## ", s, "\n\n", sep = "")
  nativeHtml <- paste0(s, "_analysis_summary.html")
  if (file.exists(file.path(wddir, nativeHtml))) {
    cat("[Full split-pipe analysis summary](", nativeHtml, ")\n\n", sep = "")
  } else {
    cat("_No split-pipe summary for this sample (no cells were recovered)._\n\n")
  }
  figDir <- file.path(wddir, s, "figures")
  if (embedFigures && dir.exists(figDir)) {
    for (fig in sort(list.files(figDir, pattern = "\\.png$", full.names = TRUE))) {
      cat("\n\n![", basename(fig), "](", fig, ")\n\n", sep = "")
    }
  }
}
cat("\n:::\n\n")
zipFn <- file.path(wddir, "all_summaries.zip")
if (file.exists(zipFn)) {
  cat("\n\n[Download all_summaries.zip](all_summaries.zip)\n")
}
```

# SessionInfo

```{r sessionInfo}
#| echo: false
ezSessionInfo()
```

:::
