---
title: "`r if (exists('reportTitle')) reportTitle else 'SUSHI Report'`"
output:
  html_document:
    mathjax: https://fgcz-gstore.uzh.ch/reference/mathjax.js
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package="ezRun")
    css: !expr system.file("templates/fgcz.css", package="ezRun")
editor_options:
  chunk_output_type: inline
---

```{r setup TargetedSeqQC, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(DT)
```

```{r load data TargetedSeqQC, include=FALSE}
if (!exists("param")) {
  param <- ezLoadRobj("param")
}
if (!exists("dataset")) {
  dataset <- ezLoadRobj("dataset")
}
if (!exists("qcSummary")) {
  qcSummary <- ezLoadRobj("qcSummary")
}
if (!exists("multiqcRelPath")) {
  multiqcRelPath <- ezLoadRobj("multiqcRelPath")
}
```

# Targeted / Exome Sequencing QC {.tabset}

## Overview

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`.

This report gives per-sample capture QC for a cohort of aligned BAM files.
The metrics are produced by **Picard CollectHsMetrics**, **mosdepth**,
**samtools** (stats/flagstat/idxstats) and optionally
**Picard CollectInsertSizeMetrics** / **VerifyBamID2**, and aggregated into a
single **MultiQC** report.

```{r overview info, results='asis'}
cat("- **Samples:** ", nrow(qcSummary), "\n", sep = "")
if (!is.null(param$refBuild) && nzchar(param$refBuild)) {
  cat("- **Reference:** ", param$refBuild, "\n", sep = "")
}
if (!is.null(param$designName) && nzchar(param$designName)) {
  cat("- **Capture design:** ", param$designName, " (vendor: ", param$vendor, ")\n", sep = "")
}
if (isTRUE(as.logical(param$restrictToMainChromosomes))) {
  cat("- **Scope:** main chromosomes only (contig name <= ",
      param$mainChromMaxNchar, " chars; scaffolds excluded)\n", sep = "")
}
cat("- **Min MAPQ / BASEQ:** ", param$mapq, " / ", param$baseq, "\n", sep = "")
```

### Full MultiQC report

The complete interactive MultiQC report (all modules and plots) opens in a new
tab. It is also exposed as its own `MultiQC Report [Link]` column in the output
dataset.

```{r multiqc link, results='asis'}
## A plain hyperlink (not an iframe/img) so the self-contained report does not
## try to embed the multi-MB MultiQC HTML as a resource.
cat('<p style="font-size:1.1em;"><a href="', multiqcRelPath,
    '" target="_blank"><b>Open the full MultiQC report &rarr;</b></a></p>\n',
    sep = "")
```

## Key metrics {.tabset}

### Summary table

Headline WES capture metrics per sample. `PCT_*` values are fractions
(0-1); `FOLD_80_BASE_PENALTY` near 1 means uniform coverage; high
`ZERO_CVG_TARGETS_PCT` or low `PCT_TARGET_BASES_30X` flags weak captures.

```{r summary table}
numCols <- setdiff(colnames(qcSummary), "Sample")
DT::datatable(
  qcSummary,
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 25, scrollX = TRUE, autoWidth = TRUE)
) |>
  DT::formatRound(columns = intersect(numCols, colnames(qcSummary)), digits = 3)
```

## Input dataset

```{r input dataset}
DT::datatable(
  as.data.frame(dataset),
  options = list(pageLength = 25, scrollX = TRUE),
  caption = "Input dataset (from SUSHI)"
)
```

## SessionInfo

```{r sessionInfo}
ezSessionInfo()
```
