---
title: "`r paste('Two groups analysis:', param$comparison)`"
author: "Functional Genomics Center Zurich"
output: 
  html_document:
    self_contained: false
    includes:
      in_header: fgcz_header.html
    css: fgcz.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## This report requires objects: se, param, output, deResult
require(knitr)
require(kableExtra)
require(SummarizedExperiment)
require(webshot)
require(htmlwidgets)
debug <- FALSE
```
Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

## TwoGroupsAnalysis_Result {.tabset}

### Settings
```{r setting, echo=FALSE}
kable(makeCountResultSummary(param, se), row.names=TRUE, 
      col.names="Setting", format="html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F,
                position = "left")
```

### Result Summary
```{r summary, echo=FALSE}
settings = character()
settings["Number of features:"] = nrow(se)
if (!is.null(rowData(se)$isPresentProbe)){
  settings["Number of features with counts above threshold:"] = 
    sum(rowData(se)$isPresentProbe)
}
knitr::kable(as.data.frame(settings), format="html", 
             col.names="Number", row.names=TRUE) %>%
    kable_styling(bootstrap_options = "striped", full_width = F,
                  position = "left")
```

#### Number of significants by p-value and fold-change
```{r gene table, echo=FALSE}
knitr::kable(makeSignificantCounts(se), row.names=TRUE,  format="html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, 
                position = "left")
```

Full result table for opening with a spreadsheet program 
(e.g. Excel: when opening with Excel, make sure that the Gene symbols are loaded into a column formatted as 'text' that prevents conversion of the symbols to dates).

```{r result live report, echo=FALSE}
resultFile <- makeResultFile(param, se)

ezWrite.table(colData(se)[, "sf", drop=FALSE],
              file="scalingfactors.txt", head="Name", digits=4)

liveReportLink = output$getColumn("Live Report")
deResult$saveToFile(basename(output$getColumn("Live Report")))
```

[`r resultFile$resultZip`](`r resultFile$resultZip`)

[Live Report and Visualizations](`r liveReportLink`)

### Inspection of significant genes

#### Between-group Comparison

```{r make scatter data, echo=FALSE}
scatterData <- makeTestScatterData(param, se)
logSignal <- scatterData$logSignal
groupMeans <- scatterData$groupMeans
types <- scatterData$types
isTwoGroup <- ncol(groupMeans) == 2 & !is.null(param$sampleGroup) &
  !is.null(param$refGroup)
```

```{r, echo=FALSE}
settings = character()
if (!is.null(param$pValueHighlightThresh)){
  settings["P-value threshold:"] = paste("p <=", param$pValueHighlightThresh)
}
if (!is.null(param$log2RatioHighlightThresh)){
    settings["Log ratio threshold:"] = paste("log ratio >=", 
                                             param$log2RatioHighlightThresh)
}
settings["Number of significant genes:"] = sum(types$Significants)
knitr::kable(as.data.frame(settings), format="html", 
             col.names="Number", row.names=TRUE) %>%
    kable_styling(bootstrap_options = "striped", full_width = F,
                  position = "left")
```

Subsequent plots highlight significant genes in <span style="color:red">red</span>.

[Interactive table of significant genes](`r resultFile$resultHtml`)

```{r scatter comparison, echo=FALSE, eval=!debug}
if (isTRUE(isTwoGroup)){
  ## scatter plot
  sampleValues = 2^groupMeans[ , param$sampleGroup]
  refValues = 2^groupMeans[ , param$refGroup]
  p_scatter <- ezXYScatterPlotly(xVec=refValues, yVec=sampleValues,
                                 isPresent=rowData(se)$usedInTest, 
                                 types=types, names=rowData(se)$gene_name,
                                 xlab=param$refGroup, ylab=param$sampleGroup,
                                 main="Comparison of average expression")
  scatterPng <- paste0(param$comparison, "-scatter.png")
  export(p_scatter, file=scatterPng)
  scatterPdf <- sub("png$", "pdf", scatterPng)
  export(p_scatter, file=scatterPdf)
  scatterHtml <- paste0(param$comparison, "-scatter.html")
  saveWidget(as_widget(p_scatter), scatterHtml)
  ### To make legend more visible on non-retinal display
  include_graphics(scatterPng, dpi=72*1.5)
  }
```

```{r scatter comparison link, echo=FALSE, results='asis'}
if (isTRUE(isTwoGroup)){
  cat(paste0("[Plot in pdf](", scatterPdf, ")"), "\n")
  cat("\n")
  cat(paste0("[Interactive comparison plot (With thousands of genes, it can be slow and requires a modern web brwoser for best javascript performance!)](", scatterHtml, ")"), "\n")
}
```

```{r volcano p-value, echo=FALSE, eval=!debug}
if (isTRUE(isTwoGroup)){
    ## volcano plot with p-value
    p_volcano <- ezVolcanoPlotly(log2Ratio=rowData(se)$log2Ratio, 
                                 pValue=rowData(se)$pValue, yType="p-value",
                                 isPresent=rowData(se)$usedInTest, types=types,
                                 names=rowData(se)$gene_name,
                                 main=param$comparison)
    volcanoPng <- paste0(param$comparison, "-volcano.png")
    export(p_volcano, file=volcanoPng)
    volcanoPdf <- sub("png$", "pdf", volcanoPng)
    export(p_volcano, file=volcanoPdf)
    volcanoHtml <- paste0(param$comparison, "-volcano.html")
    saveWidget(as_widget(p_volcano), volcanoHtml)
    ### To make legend more visible on non-retinal display
    include_graphics(volcanoPng, dpi=72*1.5)
}
```

```{r volcano p-value link , echo=FALSE, results='asis'}
if (isTRUE(isTwoGroup)){
  cat(paste0("[Plot in pdf](", volcanoPdf, ")"), "\n")
  cat("\n")
  cat(paste0("[Interactive comparison plot (With thousands of genes, it can be slow and requires a modern web brwoser for best javascript performance!)](", volcanoHtml, ")"), "\n")
}
```

```{r scatter all pairs, echo=FALSE, results='asis'}
## All pairs
if (!isTRUE(isTwoGroup)){
  plotCmd = expression({
      ezAllPairScatter(x=2^groupMeans, isPresent=rowData(se)$usedInTest, 
                       types=types)
  })
  includeHtml <- ezImageFileLink(plotCmd, file=paste0(param$comparison, "-scatter.png"),
                                 width=min(max(ncol(groupMeans) * 200, 400), 2000),
                                 height=min(max(ncol(groupMeans) * 200, 400), 2000)) 
  # dynamic png with possibly many plots
  cat(includeHtml)
}
```

### Inspection of significant genes (Advanced plots)
```{r volcano FDR, echo=FALSE, warning=FALSE, eval=!debug}
if(isTRUE(isTwoGroup)){
  ## warning=FALSE because FDR contains NA. Shall we make them always 1?
  ## volcano plot with FDR
  p_volcano <- ezVolcanoPlotly(log2Ratio=rowData(se)$log2Ratio, 
                               pValue=rowData(se)$fdr, yType="FDR",
                               isPresent=rowData(se)$usedInTest, types=types,
                               names=rowData(se)$gene_name,
                               main=param$comparison)
  volcanoFDRPng <- paste0(param$comparison, "-FDR-volcano.png")
  export(p_volcano, file=volcanoFDRPng)
  volcanoFDRPdf <- sub("png$", "pdf", volcanoFDRPng)
  export(p_volcano, file=volcanoFDRPdf)
  volcanoFDRHtml <- paste0(param$comparison, "-FDR-volcano.html")
  saveWidget(as_widget(p_volcano), volcanoFDRHtml)
  ## To make legend more visible on non-retinal display
  include_graphics(volcanoFDRPng, dpi=72*1.5)
}
```

```{r volcano FDR link, echo=FALSE, results='asis'}
if(isTRUE(isTwoGroup)){
  cat(paste0("[Plot in pdf](", volcanoFDRPdf, ")"), "\n")
  cat("\n")
  cat(paste0("[Interactive comparison plot (With thousands of genes, it can be slow and requires a modern web brwoser for best javascript performance!)](", volcanoFDRHtml, ")"), "\n")
}
```

```{r pvalue hist, echo=FALSE, results='asis', eval=!debug}
plotCmd = expression({
    myBreaks = seq(0, 1, by=0.002)
    histUsed = hist(rowData(se)$pValue[rowData(se)$usedInTest], breaks=myBreaks, plot=FALSE)
    histAbs = hist(rowData(se)$pValue[!rowData(se)$usedInTest], breaks=myBreaks, plot=FALSE)
    xx = rbind(used=histUsed$counts, absent=histAbs$counts)
    xx = shrinkToRange(xx, c(0, max(xx["used", ])))
    barplot(xx, space=0, border=NA, col=c("blue", "darkorange"), 
            xlab="p-value", ylab="counts", ylim=c(0, max(xx["used", ])),
            main="p-value histogram")
    abline(h=sum(rowData(se)$usedInTest)/ncol(xx))
    at = c(0.01, 0.1, 0.25, 0.5, 0.75, 1)
    axis(1, at=at*ncol(xx), labels = at)
    legend("top", c("used", "not expressed"), col=c("blue", "darkorange"), pch=20, cex=1)
})
includeHtml <- ezImageFileLink(plotCmd, 
                               file=paste0(param$comparison, "-pValueHist.png"), 
                               width=800)
cat(includeHtml)
```

```{r intra group, echo=FALSE, results='asis', eval=!debug}
theRange = 2^(range(logSignal, na.rm=TRUE))
x <- logSignal
if (!ezIsSpecified(param$grouping2)){ ## TODO: we no longer use pairing, we now use batch which is more general; however these plots only work if batch is a real pairing
  for (group in unique(c(param$refGroup, colnames(groupMeans)))){
    idx = which(group == param$grouping)
    if (length(idx) > 1){
      cat("\n")
      cat(paste("#### Intra-group Comparison:", group), "\n")
      
      pngName = paste0(group, "-scatter.png")
      xlab = paste("Avg of", group)
      refValues = groupMeans[ , group]
      plotCmd = expression({
        ezScatter(x=2^refValues, y=2^x[, idx, drop=FALSE], 
                  isPresent=assays(se)$isPresent[, idx, drop=FALSE], 
                  types=types, lim=theRange, xlab=xlab)
      })
      includeHtml <- ezImageFileLink(plotCmd, file=pngName, 
                                     width=min(ncol(as.matrix(x[, idx, drop=FALSE])), 6) * 480,
                                     height=ceiling(ncol(as.matrix(x[, idx, drop=FALSE]))/6) * 480)
      # dynamic png with possibly many plots
      cat(includeHtml)
      
      if (ncol(groupMeans) == 2){
          otherGroup = setdiff(colnames(groupMeans), group)
          pngName = paste0(group, "-over-", otherGroup, "-scatter.png")
          xlab = paste("Avg of", otherGroup)
          refValues = groupMeans[ , otherGroup]
          plotCmd = expression({
            ezScatter(x=2^refValues, y=2^x[, idx, drop=FALSE], 
                      isPresent=assays(se)$isPresent[, idx, drop=FALSE], 
                      types=types, lim=theRange, xlab=xlab)
          })
          includeHtml <- ezImageFileLink(plotCmd, file=pngName,
                                         width=min(ncol(as.matrix(x[, idx, drop=FALSE])), 6) * 480,
                                         height=ceiling(ncol(as.matrix(x[, idx, drop=FALSE]))/6) * 480)
          # dynamic png with possibly many plots
          cat(includeHtml)
      }
    }
    cat("\n")
  }
}else{
  cat(paste("Pairs:", param$sampleGroup, "over", param$refGroup), "\n")
  use = param$grouping %in% c(param$sampleGroup, param$refGroup)
  if (all(table(param$grouping2[use], param$grouping[use]) == 1)){
      groups = paste(param$grouping, param$grouping2, sep="--")
      sampleGroups = sort(unique(groups[param$grouping == param$sampleGroup]))
      refGroups = sort(unique(groups[param$grouping == param$refGroup]))
      avgValues = averageColumns(x[ ,use], groups[use], mean)
      avgPresent= averageColumns(x[ ,use], groups[use], function(x){mean(x) > 0.5})
      sampleValues = avgValues[ , sampleGroups, drop=FALSE]
      refValues = avgValues[ , refGroups, drop=FALSE]
      samplePresent = avgPresent[ ,sampleGroups, drop=FALSE]
      refPresent = avgPresent[ , refGroups, drop=FALSE]
      pngName = paste0(param$sampleGroup, "-over-", param$refGroup, "-pairs.png")
      plotCmd = expression({
        ezScatter(x=2^refValues, y=2^sampleValues, isPresent=samplePresent | refPresent, types=types, lim=theRange, xlab=colnames(refValues))
      })
      includeHtml <- ezImageFileLink(plotCmd, file=pngName,
                                     width=min(ncol(as.matrix(sampleValues)), 6) * 400,
                                     height=ceiling(ncol(as.matrix(sampleValues))/6) * 400)
      cat(includeHtml)
  }
}
```

### Clustering of Significant Features
```{r cluster setup, echo=FALSE}
use = rowData(se)$pValue < param$pValueHighlightThresh & 
    abs(rowData(se)$log2Ratio) > param$log2RatioHighlightThresh & rowData(se)$usedInTest
use[is.na(use)] = FALSE
if (sum(use) > param$maxGenesForClustering){
    use[use] = rank(rowData(se)$pValue[use], ties.method="max") <= param$maxGenesForClustering
}
settings = character()
settings["Significance threshold:"] = param$pValueHighlightThresh
if (param$log2RatioHighlightThresh > 0){
  settings["log2 Ratio threshold:"] = param$log2RatioHighlightThresh
}
settings["Number of significant features:"] = sum(use)
knitr::kable(as.data.frame(settings), format="html", 
             col.names="Number", row.names=TRUE) %>%
    kable_styling(bootstrap_options = "striped", full_width = F,
                  position = "left")
```

#### Cluster Plot
```{r javascript, echo=FALSE, results='asis'}
jsFile = system.file("extdata/enrichr.js", package="ezRun", mustWork=TRUE)
invisible(file.copy(from=jsFile, to=basename(jsFile), overwrite=TRUE))
cat(paste0("<SCRIPT language=\"JavaScript\" SRC=\"", basename(jsFile), "\"></SCRIPT>"))
```

```{r cluster, echo=FALSE, eval=!debug, message=FALSE, fig.width=max(8,0.1*ncol(se)), fig.height=10, fig.retina=2}
## for clustering we use a moderated logSignal
logSignal = log2(assays(se)$xNorm + param$backgroundExpression)
if (sum(use) > param$minGenesForClustering){
  xCentered = logSignal[use , ]
  if (!is.null(param$useRefGroupAsBaseline) && param$useRefGroupAsBaseline){
    xCentered = xCentered - rowMeans(xCentered[ , param$grouping == param$refGroup])
  } else {
    xCentered = xCentered - rowMeans(xCentered)
  }
  xCentered = xCentered[, order(param$grouping)]
  sampleColors = getSampleColors(param$grouping)[order(param$grouping)]
  clusterPng = "cluster-heatmap.png"
  clusterColors = c("red", "yellow", "orange", "green", "blue", "cyan")
  clusterResult = clusterResults(xCentered, nClusters=6, clusterColors=clusterColors)
  #plotCmd = expression({
    clusterHeatmap(xCentered, param, clusterResult, file=NULL,
                   colColors=sampleColors, 
                   lim=c(-param$logColorRange, param$logColorRange))
  #})
  #eval(plotCmd)
  #clusterLink = ezImageFileLink(plotCmd, file=clusterPng, 
  #                              width=max(800, 400 + 10 * ncol(xCentered)), 
  #                              height=1000) # HEATMAP
  #cat(clusterLink)
  
  seqAnno <- data.frame(rowData(se), row.names=rownames(se),
                        check.names = FALSE, stringsAsFactors=FALSE)
  if (doGo(param, seqAnno)){
    clusterResult = goClusterResults(xCentered, param, clusterResult, seqAnno=seqAnno, universeProbeIds=rownames(seqAnno)[rowData(se)$isPresentProbe])
  }
  ## append the result file with the cluster colors
  resultLoaded = ezRead.table(resultFile$resultFile)
  resultLoaded$Cluster = clusterResult$clusterColors[clusterResult$clusterNumbers[rownames(resultLoaded)]]
  ezWrite.table(resultLoaded, file=resultFile$resultFile)
  if (param$doZip){
    zipFile(resultFile$resultFile)
  }
  if (!is.null(clusterResult$GO)){
    goTables = goClusterTableRmd(param, clusterResult, seqAnno)
    if (doEnrichr(param)){
        goAndEnrichr = cbind(goTables$linkTable, goTables$enrichrTable)
    } else {
      goAndEnrichr = goTables$linkTable
    }
    bgColors = gsub("FF$", "", clusterResult$clusterColors)
    ## kable solution
    #rownames(goAndEnrichr) <- paste0('<td bgcolor="', bgColors, 
    #                                 '">',rownames(goAndEnrichr),'</td>')
    rownames(goAndEnrichr) <- paste0('<font color="', bgColors,
                                     '">Cluster ', rownames(goAndEnrichr),
                                     '</font>')
    kable(goAndEnrichr, escape=FALSE, row.names=TRUE, format = "html",
          caption="GO categories of feature clusters") %>%
      kable_styling(bootstrap_options = "striped",
                    full_width = F, position = "float_right") %>%
      add_footnote(c("Cluster font color corresponds to the row colors in the heatmap plot."), 
                   notation = "alphabet")
    
    ## htmlTable solution
    #htmlTable(goAndEnrichr, col.rgroup = clusterColors)
  }
}
```

```{r render go cluster, echo=FALSE, results='hide', message=FALSE, warning=FALSE, eval=!debug}
if (sum(use) > param$minGenesForClustering){
  if (!is.null(clusterResult$GO)){
    ## GO Cluster tables
    file <- file.path(system.file("templates", package="ezRun"), 
                      "twoGroups_goClusterTable.Rmd")
    file.copy(from=file, to=basename(file), overwrite=TRUE)
    rmarkdown::render(input=basename(file),
                      output_dir=".", output_file="goClusterTable.html")
  }
}
```

```{r go cluster table link, echo=FALSE, results='asis'}
if (file.exists("goClusterTable.html")){
  cat(paste0("[GO cluster tables](", "goClusterTable.html", ")"), "\n")
  cat("\n")
}
```

### Enrichr
```{r enrichr, echo=FALSE, message=FALSE, eval=!debug}
if (doGo(param, seqAnno)){
    if (doEnrichr(param)){
      isSig = rowData(se)$pValue < param$pValThreshGO & rowData(se)$usedInTest
      isUp = rowData(se)$log2Ratio > param$log2RatioThreshGO & isSig
      isDown = rowData(se)$log2Ratio < -param$log2RatioThreshGO & isSig
      regulatedGenes = list()
      regulatedGenes$upGenes = na.omit(unique(seqAnno[isUp, "gene_name"]))
      regulatedGenes$downGenes = na.omit(unique(seqAnno[isDown, "gene_name"]))
      regulatedGenes$bothGenes = union(regulatedGenes$upGenes, regulatedGenes$downGenes)
      
      enrichrLinks = ezMatrix("", rows=names(regulatedGenes), cols=c("External", "Precomputed"))
      maxResultsPerLibrary = 5
      for (row in rownames(enrichrLinks)){
        genesToUse = regulatedGenes[[row]]
        jsCall = paste0('enrich({list: "', paste(genesToUse, collapse="\\n"), '", popup: true});')
        enrichrLinks[row, "External"] = as.html(pot(paste0("<a href='javascript:void(0)' onClick='", jsCall, "'>Analyse at Enrichr website</a>")))
        resMerged <- NA
        if (!is.null(genesToUse) && length(genesToUse) > 3 && param$doPrecomputeEnrichr) {
          resList = runEnrichr(genesToUse, maxResult = maxResultsPerLibrary)
          resList = lapply(names(resList), function(nm){return(cbind("Gene-set library"=nm, resList[[nm]][, c(2:5, 7:10)]))}) ## add the name as a first column
          if (length(resList) > 0) {
            resMerged = do.call("rbind", resList)
            resMerged <- resMerged[order(-resMerged[,5]), ]
            resMerged[, c(3,6:8)] <- apply(resMerged[, c(3,6:8)], 2, sprintf, fmt = "%0.2e")
            resMerged[, c(4,5)] <- apply(resMerged[, c(4,5)], 2, sprintf, fmt = "%0.3f")
            enrichrTablePath <- paste0("enrichrTable_", row, ".html")
            ezInteractiveTable(resMerged, tableLink=enrichrTablePath, title=paste("Enrichr report for ", row))
            enrichrLinks[row, "Precomputed"] = as.html(ezLink(link = enrichrTablePath, label = "Report", target = "_blank"))
          } else {
            enrichrLinks[row, "Precomputed"] = "No significant results"
          }
        } else {
          enrichrLinks[row, "Precomputed"] = "Not run (too few genes)"
        }
      }
      kable(enrichrLinks, escape=FALSE, row.names=TRUE, format = "html") %>%
        kable_styling(bootstrap_options = "striped",
                    full_width = F, position = "float_left")
    }
}
```

### GO Enrichment Analysis (goseq)
Maximum number of terms displayed: `r param$maxNumberGroupsDisplayed`.

```{r goseq compute, echo=FALSE, message=FALSE, warning=FALSE}
goResult <- twoGroupsGOSE(param, se, method=param$goseqMethod)
```

```{r goseq interactive table, echo=FALSE, eval=TRUE, message=FALSE}
goTables <- goUpDownTables(param, goResult)
kable(goTables$linkTable, escape=FALSE, row.names=TRUE, format = "html") %>%
        kable_styling(bootstrap_options = "striped",
                    full_width = F, position = "float_left")
ktables <- goTables$ktables
```

```{r enrichUp, echo=FALSE, eval=TRUE}
## The code in a for loop doesn't work.
kable(ktables[["enrichUp"]], escape=TRUE, row.names=FALSE, format="html",
      caption="GO categories that are overrepresented among significantly upregulated genes.") %>%
  kable_styling(bootstrap_options = "striped",
                full_width = F, position = "float_left") %>%
  add_header_above(c("Biological Proc. (BP)" = 4, "Molecular Func. (MF)" = 4, 
                     "Cellular Comp. (CC)" = 4))
```

```{r enrichDown, echo=FALSE, eval=TRUE}
kable(ktables[["enrichDown"]], escape=TRUE, row.names=FALSE, format="html",
      caption="GO categories that are overrepresented among significantly downregulated genes.") %>%
  kable_styling(bootstrap_options = "striped",
                full_width = F, position = "float_left") %>%
  add_header_above(c("Biological Proc. (BP)" = 4, "Molecular Func. (MF)" = 4, 
                     "Cellular Comp. (CC)" = 4))
```

```{r enrichBoth, echo=FALSE, eval=TRUE}
kable(ktables[["enrichBoth"]], escape=TRUE, row.names=FALSE, format="html",
      caption="GO categories that are overrepresented among all significantly regulated genes.") %>%
  kable_styling(bootstrap_options = "striped",
                full_width = F, position = "float_left") %>%
  add_header_above(c("Biological Proc. (BP)" = 4, "Molecular Func. (MF)" = 4, 
                     "Cellular Comp. (CC)" = 4))
```

### ReViGO
```{r revigo, echo=FALSE, eval=TRUE}
kable(revigoUpDownTables(param, goResult), escape = FALSE, row.names = TRUE,
      format="html") %>%
  kable_styling(bootstrap_options = "striped",
                full_width = F, position = "float_left")
```
### Input Dataset
```{r input, echo=FALSE}
dataset <- data.frame(colData(se), check.names=FALSE, stringsAsFactors=FALSE)
ezInteractiveTableRmd(values=dataset, digits=4)
```

### SessionInfo
```{r, echo=FALSE}
sessionInfo()
```
