---
  title: "`r paste('Two groups analysis:', param$comparison)`"
author: "Functional Genomics Center Zurich"
output: 
  html_document:
  self_contained: false
lib_dir: rmarkdownLib
includes:
  in_header: fgcz_header.html
css: fgcz.css
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
output = readRDS("output.rds")
deResult = readRDS("deResult.rds")
param = readRDS("param.rds")

debug = FALSE

require(ezRun)
require(knitr)
require(kableExtra)
require(SummarizedExperiment)
require(plotly)
require(webshot)
require(htmlwidgets)
require(enrichplot)
```

```{r prepare data, include=FALSE}
# here, we prepare the data for this report

# general
se = deResult$se
param = metadata(se)$param

dataset = data.frame(colData(se), check.names=FALSE, stringsAsFactors=FALSE)
design = ezDesignFromDataset(dataset, param)

# scatter plots
scatterData = makeTestScatterData(param, se)
logSignal = scatterData$logSignal
groupMeans = scatterData$groupMeans
types = scatterData$types
isTwoGroup = ncol(groupMeans) == 2 & !is.null(param$sampleGroup) &
 !is.null(param$refGroup)

# clustering
use = rowData(se)$pValue < param$pValueHighlightThresh & 
  abs(rowData(se)$log2Ratio) > param$log2RatioHighlightThresh & rowData(se)$usedInTest
use[is.na(use)] = FALSE

# enrichment analyses
enrichInput = compileEnrichmentInput(param, se)

## In some cases, such as p2710, there are millions of features, then we have to adjust some analyses below.
hasMassiveFeats = nrow(enrichInput$seqAnno) >= 200e3
```

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

## TwoGroupsAnalysis_Result {.tabset}

### Settings
```{r settings table, echo=FALSE}
kable(makeCountResultSummary(param, se), row.names=TRUE, 
      col.names="Setting", format="html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F,
                position = "left")
```

### Result summary
```{r features table, echo=FALSE}
settings = character()
settings["Number of features:"] = nrow(se)
if (!is.null(rowData(se)$isPresentProbe)){
  settings["Number of features with counts above threshold:"] = 
    sum(rowData(se)$isPresentProbe)
}
knitr::kable(as.data.frame(settings), format="html", 
             col.names="Number", row.names=TRUE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F,
                position = "left")
```

#### Number of significants by p-value and fold-change
```{r significant genes table, echo=FALSE, message=FALSE}
knitr::kable(makeSignificantCounts(se), row.names=TRUE,  format="html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, 
                position = "left")
```

Full result table in tabular text for opening with a spreadsheet program (e.g. Excel).

When opening with Excel, do make sure that the Gene symbols are loaded into a column formatted as 'text' that prevents
conversion of the symbols to dates). See

(https://www.genenames.org/help/importing-gene-symbol-data-into-excel-correctly)

```{r store result live report, echo=FALSE, message=FALSE, eval=TRUE, warning=FALSE, cache=FALSE}
resultFile = makeResultFile(param, se)
ezWrite.table(colData(se)[, "sf", drop=FALSE],
              file="scalingfactors.txt", head="Name", digits=4)
liveReportLink = output$getColumn("Live Report")
deResult$saveToFile(basename(output$getColumn("Live Report")))
```

[`r resultFile$resultZip`](`r resultFile$resultZip`)

[Live Report and Visualizations](`r liveReportLink`)

### Inspection of significant genes
#### Between-group comparison
```{r thresholds table, echo=FALSE}
settings = character()
if (!is.null(param$pValueHighlightThresh)){
  settings["P-value threshold:"] = paste("p <=", param$pValueHighlightThresh)
}
if (!is.null(param$log2RatioHighlightThresh)){
  settings["Log ratio threshold:"] = paste("log ratio >=", 
                                           param$log2RatioHighlightThresh)
}
settings["Number of significant genes:"] = sum(types$Significants)
knitr::kable(as.data.frame(settings), format="html", 
             col.names="Number", row.names=TRUE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F,
                position = "left")
```

Subsequent plots highlight significant genes in <span style="color:#A6CEE3FF">blue</span>.

[Interactive table of significant genes](`r resultFile$resultHtml`)

```{r scatter comparison, echo=FALSE, message=FALSE, fig.width=9, fig.width=7, eval=!debug}
if(isTRUE(isTwoGroup)){
  ## scatter plot
  sampleValues = 2^groupMeans[ , param$sampleGroup]
  refValues = 2^groupMeans[ , param$refGroup]
  scatterPng = paste0(param$comparison, "-scatter.png")
  scatterPdf = sub("png$", "pdf", scatterPng)
  p_scatter = ezXYScatter.2(xVec=refValues, yVec=sampleValues,
                             isPresent=rowData(se)$usedInTest, 
                             types=types, names=rowData(se)$gene_name,
                             xlab=param$refGroup, ylab=param$sampleGroup,
                             main="Comparison of average expression",
                             mode="ggplot2")
  
  print(p_scatter)
  
  ggsave(filename=scatterPdf, plot=p_scatter, device="pdf",
         width=9, height=7)
  
  if(!isTRUE(hasMassiveFeats)){
    p_scatter = ezXYScatter.2(xVec=refValues, yVec=sampleValues,
                               isPresent=rowData(se)$usedInTest, 
                               types=types, names=rowData(se)$gene_name,
                               xlab=param$refGroup, ylab=param$sampleGroup,
                               main="Comparison of average expression",
                               mode="plotly")
    scatterHtml = paste0(param$comparison, "-scatter.html")
    saveWidget(as_widget(p_scatter), scatterHtml)
  }
}
```

```{r volcano p-value, echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.width=7, eval=!debug}
if (isTRUE(isTwoGroup)){
  ## volcano plot with p-value
  volcanoPng = paste0(param$comparison, "-volcano.png")
  volcanoPdf = sub("png$", "pdf", volcanoPng)
  p_volcano = ezVolcano(log2Ratio=rowData(se)$log2Ratio,
                         pValue=rowData(se)$pValue, yType="p-value",
                         isPresent=rowData(se)$usedInTest, types=types,
                         names=rowData(se)$gene_name,
                         main=param$comparison,
                         mode="ggplot2")
  
  print(p_volcano)
  
  ggsave(filename=volcanoPdf, plot=p_volcano, device="pdf",
         width=9, height=7)
  if(!isTRUE(hasMassiveFeats)){
    p_volcano = ezVolcano(log2Ratio=rowData(se)$log2Ratio, 
                           pValue=rowData(se)$pValue, yType="p-value",
                           isPresent=rowData(se)$usedInTest, types=types,
                           names=rowData(se)$gene_name,
                           main=param$comparison,
                           mode="plotly")
    volcanoHtml = paste0(param$comparison, "-volcano.html")
    saveWidget(as_widget(p_volcano), volcanoHtml)
  }
}
```

```{r scatter all pairs, echo=FALSE, results='asis', message=FALSE, eval=!debug}
## All pairs
if (!isTRUE(isTwoGroup)){
  plotCmd = expression({
    ezAllPairScatter(x=2^groupMeans, isPresent=rowData(se)$usedInTest, 
                     types=types)
  })
  includeHtml = ezImageFileLink(plotCmd, file=paste0(param$comparison, "-scatter.png"),
                                 width=min(max(ncol(groupMeans) * 200, 400), 2000),
                                 height=min(max(ncol(groupMeans) * 200, 400), 2000)) 
  # dynamic png with possibly many plots
  cat(includeHtml)
}
```

### Inspection of significant genes (Advanced plots)
```{r volcano FDR, echo=FALSE, message=FALSE, fig.width=9, fig.width=7, eval=!debug, warning=FALSE}
if(isTRUE(isTwoGroup)){
  ## warning=FALSE because FDR contains NA. Shall we make them always 1?
  ## volcano plot with FDR
  volcanoFDRPng = paste0(param$comparison, "-FDR-volcano.png")
  volcanoFDRPdf = sub("png$", "pdf", volcanoFDRPng)
  p_volcano = ezVolcano(log2Ratio=rowData(se)$log2Ratio,
                         pValue=rowData(se)$fdr, yType="FDR",
                         isPresent=rowData(se)$usedInTest, types=types,
                         names=rowData(se)$gene_name,
                         main=param$comparison,
                         mode="ggplot2")
  print(p_volcano)
  
  ggsave(filename=volcanoFDRPdf, plot=p_volcano, device="pdf",
         width=9, height=7)
  
  if(!isTRUE(hasMassiveFeats)){
    p_volcano = ezVolcano(log2Ratio=rowData(se)$log2Ratio, 
                           pValue=rowData(se)$fdr, yType="FDR",
                           isPresent=rowData(se)$usedInTest, types=types,
                           names=rowData(se)$gene_name,
                           main=param$comparison,
                           mode="plotly")
    volcanoFDRHtml = paste0(param$comparison, "-FDR-volcano.html")
    saveWidget(as_widget(p_volcano), volcanoFDRHtml)
  }
}
```

```{r pvalue hist, echo=FALSE, results='asis', eval=!debug}
plotCmd = expression({
  myBreaks = seq(0, 1, by=0.002)
  histUsed = hist(rowData(se)$pValue[rowData(se)$usedInTest], breaks=myBreaks, plot=FALSE)
  histAbs = hist(rowData(se)$pValue[!rowData(se)$usedInTest], breaks=myBreaks, plot=FALSE)
  xx = rbind(used=histUsed$counts, absent=histAbs$counts)
  xx = shrinkToRange(xx, c(0, max(xx["used", ])))
  barplot(xx, space=0, border=NA, col=c("blue", "darkorange"), 
          xlab="p-value", ylab="counts", ylim=c(0, max(xx["used", ])),
          main="p-value histogram")
  abline(h=sum(rowData(se)$usedInTest)/ncol(xx))
  at = c(0.01, 0.1, 0.25, 0.5, 0.75, 1)
  axis(1, at=at*ncol(xx), labels = at)
  legend("top", c("used", "not expressed"), col=c("blue", "darkorange"), pch=20, cex=1)
})
eval(plotCmd)
```

```{r intra group, echo=FALSE, results='asis', eval=!debug, message=FALSE}
theRange = 2^(range(logSignal, na.rm=TRUE))
x = logSignal
if(ncol(x) <= 100){
  if (!ezIsSpecified(param$grouping2)){ ## TODO: we no longer use pairing, we now use batch which is more general; however these plots only work if batch is a real pairing
    for (group in unique(c(param$refGroup, colnames(groupMeans)))){
      idx = which(group == param$grouping)
      if (length(idx) > 1){
        cat("\n")
        cat(paste("#### Intra-group Comparison:", group), "\n")
        
        pngName = paste0(group, "-scatter.png")
        xlab = paste("Avg of", group)
        refValues = groupMeans[ , group]
        plotCmd = expression({
          ezScatter(x=2^refValues, y=2^x[, idx, drop=FALSE], 
                    isPresent=assays(se)$isPresent[, idx, drop=FALSE], 
                    types=types, lim=theRange, xlab=xlab)
        })
        includeHtml = ezImageFileLink(plotCmd, file=pngName, 
                                       width=min(ncol(as.matrix(x[, idx, drop=FALSE])), 6) * 480,
                                       height=ceiling(ncol(as.matrix(x[, idx, drop=FALSE]))/6) * 480)
        # dynamic png with possibly many plots
        cat(includeHtml)
        
        if (ncol(groupMeans) == 2){
          otherGroup = setdiff(colnames(groupMeans), group)
          pngName = paste0(group, "-over-", otherGroup, "-scatter.png")
          xlab = paste("Avg of", otherGroup)
          refValues = groupMeans[ , otherGroup]
          plotCmd = expression({
            ezScatter(x=2^refValues, y=2^x[, idx, drop=FALSE], 
                      isPresent=assays(se)$isPresent[, idx, drop=FALSE], 
                      types=types, lim=theRange, xlab=xlab)
          })
          includeHtml = ezImageFileLink(plotCmd, file=pngName,
                                         width=min(ncol(as.matrix(x[, idx, drop=FALSE])), 6) * 480,
                                         height=ceiling(ncol(as.matrix(x[, idx, drop=FALSE]))/6) * 480)
          # dynamic png with possibly many plots
          cat(includeHtml)
        }
      }
      cat("\n")
    }
  }else{
    cat(paste("Pairs:", param$sampleGroup, "over", param$refGroup), "\n")
    use = param$grouping %in% c(param$sampleGroup, param$refGroup)
    if (all(table(param$grouping2[use], param$grouping[use]) == 1)){
      groups = paste(param$grouping, param$grouping2, sep="--")
      sampleGroups = sort(unique(groups[param$grouping == param$sampleGroup]))
      refGroups = sort(unique(groups[param$grouping == param$refGroup]))
      avgValues = averageColumns(x[ ,use], groups[use], mean)
      avgPresent= averageColumns(x[ ,use], groups[use], function(x){mean(x) > 0.5})
      sampleValues = avgValues[ , sampleGroups, drop=FALSE]
      refValues = avgValues[ , refGroups, drop=FALSE]
      samplePresent = avgPresent[ ,sampleGroups, drop=FALSE]
      refPresent = avgPresent[ , refGroups, drop=FALSE]
      pngName = paste0(param$sampleGroup, "-over-", param$refGroup, "-pairs.png")
      plotCmd = expression({
        ezScatter(x=2^refValues, y=2^sampleValues, isPresent=samplePresent | refPresent, types=types, lim=theRange, xlab=colnames(refValues))
      })
      includeHtml = ezImageFileLink(plotCmd, file=pngName,
                                     width=min(ncol(as.matrix(sampleValues)), 6) * 400,
                                     height=ceiling(ncol(as.matrix(sampleValues))/6) * 400)
      cat(includeHtml)
    }
  }
}
```

### Clustering of significant features
```{r cluster setup, echo=FALSE, eval=!debug}
if (sum(use) > param$maxGenesForClustering){
  use[use] = rank(rowData(se)$pValue[use], ties.method="max") <= param$maxGenesForClustering
}
settings = character()
settings["Significance threshold:"] = param$pValueHighlightThresh
if (param$log2RatioHighlightThresh > 0){
  settings["log2 Ratio threshold:"] = param$log2RatioHighlightThresh
}
settings["Number of significant features:"] = sum(use)
knitr::kable(as.data.frame(settings), format="html", 
             col.names="Number", row.names=TRUE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F,
                position = "left")
```

#### Cluster plot
```{r javascript, echo=FALSE, results='asis', eval=!debug}
jsFile = system.file("extdata/enrichr.js", package="ezRun", mustWork=TRUE)
invisible(file.copy(from=jsFile, to=basename(jsFile), overwrite=TRUE))
cat(paste0("<SCRIPT language=\"JavaScript\" SRC=\"", basename(jsFile), "\"></SCRIPT>"))
```

```{r cluster, echo=FALSE, message=FALSE, fig.width=max(8,4+0.15*ncol(se)), fig.height=10, fig.retina=2, warning=FALSE, eval=!debug && !hasMassiveFeats}

## for clustering we use a moderated logSignal
logSignal = log2(assays(se)$xNorm + param$backgroundExpression)
if (sum(use) > param$minGenesForClustering){
  xCentered = logSignal[use , ]
  if(param$onlyCompGroupsHeatmap){
    columnsToSubset = param$grouping %in% c(param$refGroup, param$sampleGroup)
    param$grouping = param$grouping[columnsToSubset]
    xCentered = xCentered[ , columnsToSubset]
    design = design[columnsToSubset,,drop=FALSE]
  }
  if (!is.null(param$useRefGroupAsBaseline) && param$useRefGroupAsBaseline){
    xCentered = xCentered - rowMeans(xCentered[ , param$grouping == param$refGroup])
    xCentered = sweep(xCentered, MARGIN=1, 
                      rowMeans(xCentered[ , param$grouping == param$refGroup]),
                      FUN="-")
  } else {
    xCentered = sweep(xCentered, MARGIN=1, rowMeans(xCentered), FUN="-")
  }
  xCentered = xCentered[, order(param$grouping)]
  sampleColors = getSampleColors(param$grouping)[order(param$grouping)]
  
  clusterPng = "cluster-heatmap.png"
  clusterPdf = "cluster-heatmap.pdf"
  
  nClusters=6
  clusterColors = hcl.colors(nClusters, palette = "Spectral")
  clusterResult = clusterResults(xCentered, nClusters=nClusters, clusterColors=clusterColors)
  
  design = design[,c(param$groupingName,setdiff(colnames(design),param$groupingName)), drop=FALSE]
  
  plotCmd = expression({
    clusterPheatmap(xCentered, design, param, clusterColors,
                    lim=c(-param$logColorRange, param$logColorRange),
                    doClusterColumns=FALSE, sampleColors=sampleColors)
  })
  eval(plotCmd)
  
  if (doGo(param, enrichInput$seqAnno)){
    clusterResult = goClusterResults(x = xCentered, param, result = clusterResult, enrichInput)
  }
  ## append the result file with the cluster colors
  resultLoaded = ezRead.table(resultFile$resultFile)
  resultLoaded$Cluster = clusterResult$clusterColors[clusterResult$clusterNumbers[rownames(resultLoaded)]]
  ezWrite.table(resultLoaded, file=resultFile$resultFile)
  if (param$doZip){
    zipFile(resultFile$resultFile)
  }
  if (!is.null(clusterResult$GO)){
    goTables = goClusterTableRmd(param, clusterResult, enrichInput$seqAnno)
    if (doEnrichr(param)){
      goAndEnrichr = cbind(goTables$linkTable, goTables$enrichrTable)
    } else {
      goAndEnrichr = goTables$linkTable
    }
    bgColors = gsub("FF$", "", clusterResult$clusterColors)
    rownames(goAndEnrichr) = paste0('<font color="', bgColors,
                                     '">Cluster ', rownames(goAndEnrichr),
                                     '</font>')
    kable(goAndEnrichr, escape=FALSE, row.names=TRUE, format = "html",
          caption="GO categories of feature clusters") %>%
      kable_styling(bootstrap_options = "striped",
                    full_width = F, position = "float_right") %>%
      footnote(general="Cluster font color corresponds to the row colors in the heatmap plot.")
  }
}
```

```{r render go cluster, echo=FALSE, results='hide', message=FALSE, warning=FALSE, eval=!debug && !hasMassiveFeats}
if (sum(use) > param$minGenesForClustering && !is.null(clusterResult$GO)){
    ## GO Cluster tables
    file = file.path(system.file("templates", package="ezRun"), 
                      "twoGroups_goClusterTable.Rmd")
    file.copy(from=file, to=basename(file), overwrite=TRUE)
    rmarkdown::render(input=basename(file), envir = new.env(),
                      output_dir=".", output_file="goClusterTable.html",
                      quiet=TRUE)
}
```

```{r go cluster table link, echo=FALSE, results='asis', eval=!debug && !hasMassiveFeats}
if (file.exists("goClusterTable.html")){
  cat(paste0("[GO cluster tables](", "goClusterTable.html", ")"), "\n")
  cat("\n")
}
```

### Enrichr
```{r enrichr, echo=FALSE, message=FALSE, eval=!debug}
if (doGo(param, enrichInput$seqAnno)){
  if (doEnrichr(param)){
    isSig = rowData(se)$pValue < param$pValThreshGO & rowData(se)$usedInTest
    isUp = rowData(se)$log2Ratio > param$log2RatioThreshGO & isSig
    isDown = rowData(se)$log2Ratio < -param$log2RatioThreshGO & isSig
    regulatedGenes = list()
    regulatedGenes$upGenes = na.omit(unique(enrichInput$seqAnno[isUp, "gene_name"]))
    regulatedGenes$downGenes = na.omit(unique(enrichInput$seqAnno[isDown, "gene_name"]))
    regulatedGenes$bothGenes = union(regulatedGenes$upGenes, regulatedGenes$downGenes)
    
    enrichrLinks = ezMatrix("", rows=names(regulatedGenes), cols=c("Number of Genes", "External", "Precomputed"))
    maxResultsPerLibrary = 5
    for (row in rownames(enrichrLinks)){
      genesToUse = regulatedGenes[[row]]
      enrichrLinks[row, "Number of Genes"] = length(genesToUse)
      jsCall = paste0('enrich({list: "', paste(genesToUse, collapse="\\n"), '", popup: true});')
      enrichrLinks[row, "External"] = paste0("<a href='javascript:void(0)' onClick='", jsCall, "'>Analyse at Enrichr website</a>")
      resMerged = NA
      if (!is.null(genesToUse) && length(genesToUse) > 3 && param$doPrecomputeEnrichr) {
        resList = runEnrichr(genesToUse, maxResult = maxResultsPerLibrary)
        resList = lapply(names(resList), 
                         function(nm){return(cbind("Gene-set library"=nm, 
                                                   resList[[nm]][, c(2:5, 7:10)]))}) ## add the name as a first column
        if (length(resList) > 0) {
          resMerged = do.call("rbind", resList)
          resMerged = resMerged[order(-resMerged[,5]), ]
          resMerged[, c(3,6:8)] = apply(resMerged[, c(3,6:8)], 2, sprintf,
                                         fmt = "%0.2e")
          resMerged[, c(4,5)] = apply(resMerged[, c(4,5)], 2, sprintf, 
                                       fmt="%0.3f")
          enrichrTablePath = paste0("enrichrTable_", row, ".html")
          ezInteractiveTable(resMerged, tableLink=enrichrTablePath,
                             title=paste("Enrichr report for ", row))
          enrichrLinks[row, "Precomputed"] = ezLink(link=enrichrTablePath,
                                                    label="Report",
                                                    target="_blank")
        } else {
          enrichrLinks[row, "Precomputed"] = "No significant results"
        }
      } else {
        enrichrLinks[row, "Precomputed"] = "Not run (too few genes)"
      }
    }
    settings = character()
    if (!is.null(param$pValueHighlightThresh)){
      settings["P-value threshold:"] = paste("p <=", param$pValThreshGO)
    }
    if (!is.null(param$log2RatioHighlightThresh)){
      settings["Log ratio threshold:"] = paste("log ratio >=", param$log2RatioThreshGO)
    }
    kable(as.data.frame(settings), format="html", 
          col.names="Number", row.names=TRUE) %>%
      kable_styling(bootstrap_options = "striped", full_width = F,
                    position = "left")
    kable(enrichrLinks, escape=FALSE, row.names=TRUE, format = "html") %>%
      kable_styling(bootstrap_options = "striped",
                    full_width = F, position = "left")
  }
}
if(!doEnrichr(param)){
  cat("\n")
  cat(getOrganism(param$ezRef), "is not supported by Enrichr.")
  cat("\n")
}
```

### GO enrichment analysis (goseq)
Maximum number of terms displayed: `r param$maxNumberGroupsDisplayed`.

```{r goseq compute, echo=FALSE, message=FALSE, warning=FALSE, eval=!debug}
if (doGo(param, enrichInput$seqAnno)){
  goResult = twoGroupsGOSE(param, enrichInput, se)
  goTables = goUpDownTables(param, goResult)
  resFns <- c() ## Provide download links for these results
}
```

```{r goseq interactive table, echo=FALSE, eval=!debug}
if (exists("goTables")){
  kable(goTables$linkTable, escape=FALSE, row.names=TRUE, format = "html") %>%
    kable_styling(bootstrap_options = "striped",
                  full_width = F, position = "left")
}
```

```{r enrichUp, echo=FALSE, eval=!debug}
## The code in a for loop doesn't work.
## TODO: maybe explicit print or tagList method can work.
if (exists("goTables")){
  ktables = goTables$ktables
  kable(ktables[["enrichUp"]], escape=TRUE, row.names=FALSE, format="html",
        caption="GO categories that are overrepresented among significantly upregulated genes.") %>%
    kable_styling(bootstrap_options = c("striped", "bordered"),
                  full_width = F, position = "left") %>%
    add_header_above(c("Biological Proc. (BP)" = 4, "Molecular Func. (MF)" = 4, 
                       "Cellular Comp. (CC)" = 4))
}
```

```{r enrichDown, echo=FALSE, eval=!debug}
if (exists("ktables")){
  kable(ktables[["enrichDown"]], escape=TRUE, row.names=FALSE, format="html",
        caption="GO categories that are overrepresented among significantly downregulated genes.") %>%
    kable_styling(bootstrap_options = c("striped", "bordered"),
                  full_width = F, position = "left") %>%
    add_header_above(c("Biological Proc. (BP)" = 4, "Molecular Func. (MF)" = 4, 
                       "Cellular Comp. (CC)" = 4))
}
```

```{r enrichBoth, echo=FALSE, eval=!debug}
if (exists("ktables")){
  kable(ktables[["enrichBoth"]], escape=TRUE, row.names=FALSE, format="html",
        caption="GO categories that are overrepresented among all significantly regulated genes.") %>%
    kable_styling(bootstrap_options = c("striped", "bordered"),
                  full_width = F, position = "left") %>%
    add_header_above(c("Biological Proc. (BP)" = 4, "Molecular Func. (MF)" = 4, 
                       "Cellular Comp. (CC)" = 4))
}
```

#### Result files
```{r goseq result files, echo=FALSE, results='asis'}
if (exists("goTables")){
  for(resFn in goTables$txtFiles){
    cat("\n")
    cat(paste0("[", resFn, "](", resFn, ")"))
    cat("\n")
  }
}
```

#### dotplot
```{r dotplot, echo=FALSE, message=FALSE, eval=!debug}
if(exists("goResult")){
  for(onto in names(goResult)){
    if(is.null(goResult[[onto]]) || nrow(goResult[[onto]]$enrichBoth) == 0L){
      cat("No significant gene set.\n")
      next
    }else{
      for (enrichType in names(goResult[[onto]])){
        enrichResult = prepareEnrichResult(goTables$ktablesComplete[[enrichType]][[onto]])
        p = dotplot(enrichResult, 
                    showCategory=20, title=paste(onto,enrichType))
        print(p)
        resFn = paste(onto, enrichType, "GSEA.txt", sep="_")
        write.table(goResult[[onto]], file=resFn,
                    quote=FALSE, col.names = TRUE, row.names = FALSE)
        resFns = append(resFns, resFn)
      }
    }
  }
}
```

#### enrichMap
```{r enrichMap, echo=FALSE, fig.width=14, fig.height=14, eval=FALSE}
if(exists("goResult")){
  for(onto in names(goResult)){
    if(is.null(goResult[[onto]]) || nrow(goResult[[onto]]$enrichBoth) == 0L){
      cat("No significant gene set.\n")
      next
    }else{
      for (enrichType in names(goResult[[onto]])){
        enrichResult = prepareEnrichResult(goTables$ktablesComplete[[enrichType]][[onto]])
        p = suppressWarnings(emapplot(enrichResult, showCategory=50,main=onto))
        print(p)
      }
    }
  }
}
```

#### cnetplot
```{r cnetplot, echo=FALSE, fig.width=14, fig.height=14, eval=FALSE}
if(exists("goResult")){
  for(onto in names(goResult)){
    if(is.null(goResult[[onto]]) || nrow(goResult[[onto]]$enrichBoth) == 0L){
      cat("No significant gene set.\n")
      next
    }else{
      for (enrichType in names(goResult[[onto]])){
        # use gene names
        enrichResult = goTables$ktablesComplete[[enrichType]][[onto]]
        enrichResult$geneID = enrichResult$geneNames
        enrichResult = prepareEnrichResult(enrichResult)
        
        p = cnetplot(enrichResult,
                     foldChange=setNames(rowData(se)$log2Ratio, rowData(se)$gene_name),
                     main=onto)
        print(p)
      }
    }
  }
}
```

### Webgestalt
[InputFiles](./Webgestalt) for [Webgestalt](http://www.webgestalt.org/)

```{r webgestalt, include=FALSE}
makeWebgestaltFiles(param, resultFile$resultFile)
if(param$runGSEA){
  gseaFiles = list.files('Webgestalt', pattern = '^GSEA_Input', full.names = TRUE)
  for (i in 1:length(gseaFiles)){
    res = runWebgestaltGSEA(param, gseaFiles[i])
  }
  ##Keep only interactive html report & zip archive:
  ezSystem('mv Webgestalt/GSEA_Results/*/*.html Webgestalt/GSEA_Results')
  ezSystem('mv Webgestalt/GSEA_Results/*/*.zip Webgestalt/GSEA_Results')
  ezSystem('find Webgestalt/GSEA_Results -maxdepth 1 -mindepth 1 -type d -exec rm -Rf {} \\;')
}
ezSystem('zip Webgestalt/InputFiles_GSEA.zip Webgestalt/*.rnk')
ezSystem('rm Webgestalt/*.rnk')
ezSystem('zip Webgestalt/InputFiles_ORA.zip Webgestalt/*.txt')
ezSystem('rm Webgestalt/*.txt')
```

### Technical bias
We define 4 gene sets 

* high GC: the 5% of the genes with the highest GC content
* low GC: the 5% of the genes with the lowest GC content
* long genes: the 5% of the genes with the biggest length
* short genes: the 5% of the genes with the smalles length

And we test if the up- or down-regulated genes are associated with one of those gene sets. If there is a significant association, some of the significant genes are potentially false positives 
due to a technical bias.

Tests where the association p-value is below 0.001 are highlighted in red.
The column "overlapping/total genes" shows the number of overlapping genes
and the total number of genes in that category.


```{r bias, echo=FALSE, eval=TRUE}
x = rowData(se)[rowData(se)$usedInTest, ]

if(all(c("gc", "featWidth") %in% colnames(x))) {
  if(!all(is.na(x$gc)) & !all(is.na(x$featWidth))){
    gcThresh = quantile(x$gc, c(0.05, 0.95))
    widthThresh = quantile(x$featWidth, c(0.05, 0.95))
    biasTable = data.frame("low GC"=x$gc < gcThresh[1],
                            "high GC"=x$gc > gcThresh[2],
                            "short genes"= x$featWidth < widthThresh[1],
                            "long genes"=x$featWidth > widthThresh[2],
                            check.names=FALSE)
    isUp = x$pValue < param$pValueHighlightThresh & 
      x$log2Ratio > param$log2RatioHighlightThresh
    isDown = x$pValue < param$pValueHighlightThresh & 
      x$log2Ratio < -param$log2RatioHighlightThresh
    sigTable = data.frame("Up-regulation"=isUp,
                           "Down-regulation"=isDown,
                           check.names=FALSE)
    
    tests = expand.grid(sig=names(sigTable), bias=names(biasTable))
    testTable = ezFrame("overlapping/total genes"=character(), "odds ratio"=numeric(), "p-value"=numeric())
    for (i in 1:nrow(tests)){
      testName = paste(tests$bias[i], " -- ", tests$sig[i])
      myBias = biasTable[[tests$bias[i]]]
      mySig = sigTable[[tests$sig[i]]]
      if(any(mySig)){
        res.fisher = fisher.test(myBias, mySig, alternative="greater")
      } else {
        res.fisher = list(estimate=NA, p.value=NA)
      }
      testTable[testName, "overlapping/total genes"] = paste(sum(myBias & mySig), sum(myBias), sep="/")
      testTable[testName, "odds ratio"] = res.fisher$estimate
      testTable[testName, "p-value"] = res.fisher$p.value
    }
    t = kable(format(testTable, digits=3, scientific=-2), "html", 
               caption=paste("Association test:",
                             sum(isUp), "up-regulated and",
                             sum(isDown), "down-regulated genes")) %>%
      kable_styling(bootstrap_options = "striped", full_width = F, 
                    position = "left")
    isSig = testTable$`p-value` < 1e-3
    if(any(isSig, na.rm = TRUE)){
      t = t %>% 
        row_spec(which(isSig), bold = T, color = "white",
                 background = "#D7261E")
    }
    t
  }
}
```

### Input dataset
```{r input, echo=FALSE}
ezInteractiveTableRmd(values=dataset, digits=4)
```

### SessionInfo
```{r, echo=FALSE}
sessionInfo()
```
