---
title: "`r if (exists('reportTitle')) reportTitle else 'SUSHI Report'`"
output:
  html_document:
    mathjax: https://fgcz-gstore.uzh.ch/reference/mathjax.js
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package="ezRun")
    css: !expr system.file("templates/fgcz.css", package="ezRun")
editor_options:
  chunk_output_type: inline
---

`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(ggplot2)
library(DT)
wddir <- "."
```

# `r params$sample` {.tabset}

## Overview

### Parameters

```{r params-table}
kable(
  data.frame(
    Parameter = c(
      "Sample",
      "Paired end",
      "Host genome (hostBuild)",
      "Viral database (virBuild)",
      "Minimum read count for reporting"
    ),
    Value = c(
      params$sample,
      params$paired,
      params$hostBuild,
      params$virBuild,
      params$minReadCount
    )
  ),
  col.names = c("Parameter", "Value"),
  caption = "Analysis parameters"
)
```

### Contamination and virome mapping rates

Raw reads were trimmed and filtered based on quality scores, further filtered
by contaminated genomes (human and non-human host), before being mapped to
the reference virome database.

```{r counts, fig.width=7, fig.height=5, fig.cap="Read/pair counts retained at each pipeline stage"}
counts.file <- file.path(wddir, "read_count_summary.tsv")
if (file.exists(counts.file)) {
  counts <- read.table(
    counts.file,
    header = TRUE,
    sep = "\t",
    stringsAsFactors = FALSE
  )
  counts$Stage <- factor(counts$Stage, levels = counts$Stage)
  kable(counts, row.names = FALSE, caption = "Read/pair counts through the pipeline")
  ggplot(counts, aes(x = Stage, y = Count)) +
    geom_col(fill = "steelblue") +
    scale_y_log10() +
    labs(x = NULL, y = "Read pairs (log scale)", title = "Reads retained at each pipeline stage") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
} else {
  cat("Read count summary not available.")
}
```

## Detected viral genomes

Viral genomes detected in the reference virome database, after removing host
contamination.

```{r summary-table}
list.file <- file.path(wddir, "summary_table.tsv")
if (file.exists(list.file)) {
  full <- read.table(
    list.file,
    header = TRUE,
    stringsAsFactors = FALSE,
    sep = "\t"
  )
  sub <- full[full$mappedReads > params$minReadCount, ]
  sub <- na.omit(sub)
} else {
  sub <- data.frame()
}
if (nrow(sub) > 0) {
  DT::datatable(
    sub,
    rownames = FALSE,
    caption = paste(
      "Viral genomes detected with more than",
      params$minReadCount,
      "mapped reads"
    )
  )
} else {
  cat("No virus detected with more than", params$minReadCount, "reads.")
}
```

### Bubble plot of all detected viral families

```{r family-bubble, eval=nrow(sub)>0, fig.width=8, fig.height=6, fig.cap="Mean genome coverage and depth per detected viral family; bubble size is mean mapped bases"}
agg <- aggregate(
  cbind(mappedBases, genomeCov_pect, aveDepth) ~ Family,
  data = sub,
  mean
)
ggplot(agg, aes(x = genomeCov_pect, y = aveDepth, size = mappedBases, color = Family)) +
  geom_point(alpha = 0.7) +
  ggrepel::geom_text_repel(aes(label = Family), size = 3, show.legend = FALSE) +
  scale_size_area(max_size = 15) +
  labs(x = "% genome coverage", y = "Mean depth", title = "Detected viral families") +
  theme_bw()
```

### Bubble plot of viral species, grouped by viral family

```{r species-bubble, eval=nrow(sub)>0, fig.width=10, fig.height=7, fig.cap="Genome coverage and depth per detected viral species, faceted by family; bubble size is mapped bases"}
ggplot(sub, aes(x = genomeCov_pect, y = aveDepth, size = mappedBases, color = Family)) +
  geom_point(alpha = 0.7) +
  ggrepel::geom_text_repel(aes(label = CommonName), size = 3, show.legend = FALSE) +
  scale_size_area(max_size = 15) +
  facet_wrap(~Family, scales = "free") +
  labs(
    x = "% genome coverage",
    y = "Mean depth",
    title = "Detected viral species, grouped by family"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

## Coverage plots {.tabset}

Per-base coverage of each detected viral genome, sorted by abundance.

```{r coverage-tabs, results='asis', eval=nrow(sub)>0}
for (i in seq_len(nrow(sub))) {
  chr <- sub$ID[i]
  common_name <- sub$CommonName[i]
  csv.file <- file.path(wddir, paste0(chr, ".csv"))
  if (!file.exists(csv.file)) {
    next
  }
  cov <- read.table(
    csv.file,
    header = FALSE,
    sep = "\t",
    quote = "",
    stringsAsFactors = FALSE
  )
  cat("\n\n### ", common_name, "\n\n")
  p <- ggplot(cov, aes(x = V5, y = V6)) +
    geom_line(color = "steelblue") +
    labs(x = "Position", y = "Depth", title = common_name) +
    theme_bw()
  print(p)
  cat("\n\n")
}
```

```{r coverage-tabs-empty, eval=nrow(sub)==0, results='asis'}
cat("No virus detected with more than", params$minReadCount, "reads.")
```

## Session Info

```{r session-info}
ezSessionInfo()
```
