---
title: "`r if (exists('reportTitle')) reportTitle else 'SUSHI Report'`"
output:
  html_document:
    mathjax: https://fgcz-gstore.uzh.ch/reference/mathjax.js
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package="ezRun")
    css: !expr system.file("templates/fgcz.css", package="ezRun")
editor_options:
  chunk_output_type: console
---

```{css custom-tabs, echo=FALSE}
/* Level 1 tabs: larger and bold */
.nav-tabs > li > a {
  font-size: 1.15em;
  font-weight: bold;
  padding: 10px 18px;
}

/* Level 2 tabs (nested): normal size */
.tab-content .nav-tabs > li > a {
  font-size: 0.95em;
  font-weight: normal;
  padding: 8px 14px;
}
```

# {.tabset}

## Cell filtering {.tabset}

```{r setup, include=FALSE}
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
library(ezRun)
library(qs2)
library(DT)
library(pals)
library(scales)
library(stringr)
library(kableExtra)
library(readxl)
library(viridis)
library(UpSetR)
library(scater)
library(SingleCellExperiment)
library(SpatialExperiment)
library(grid)
library(ComplexHeatmap)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      knitr.table.format = "html")
```

```{r load_data}
# Load pre-computed data
scData <- qs_read("scData.qs2")
param <- qs_read("param.qs2")

# Load unfiltered data if available
if (file.exists("scData.unfiltered.qs2")) {
  scData_unfiltered <- qs_read("scData.unfiltered.qs2")
  has_unfiltered <- TRUE
} else {
  has_unfiltered <- FALSE
}

# Load markers
posMarkers <- readxl::read_xlsx("posMarkers.xlsx")
if (file.exists("posMarkersBanksy.xlsx")) {
  posMarkersBanksy <- readxl::read_xlsx("posMarkersBanksy.xlsx")
} else {
  posMarkersBanksy <- data.frame()
}

# Process markers
if (nrow(posMarkers) > 0) {
  posMarkers$gene <- as.factor(posMarkers$gene)
  clusterSet <- posMarkers$cluster |> as.character() |> as.integer() |>
    unique() |> sort(na.last = TRUE)
  if (any(is.na(clusterSet))) {
    posMarkers$cluster <- as.factor(posMarkers$cluster)
  } else {
    posMarkers$cluster <- factor(posMarkers$cluster, levels = clusterSet)
  }
  posMarkers$p_val_adj[posMarkers$p_val_adj == 0] <-
    min(posMarkers$p_val_adj[posMarkers$p_val_adj > 0])
}

# Get assay column names
nCountCol <- grep("^nCount_Xenium", colnames(scData@meta.data), value = TRUE) |>
  head(1)
nFeatureCol <- grep("^nFeature_Xenium", colnames(scData@meta.data), value = TRUE) |>
  head(1)

# Polychrome palette
cluster_colors <- unname(pals::polychrome())
```

```{r seurat_to_spe, eval=has_unfiltered}
# Convert unfiltered Seurat to SPE for plotColData (matching VisiumHD)
imageName <- names(scData_unfiltered@images)[1]
coords <- scData_unfiltered@images[[imageName]]@boundaries$centroids@coords
spe <- Seurat_to_SPE(scData_unfiltered, coords)
```

### Cell filtering

::: {.row}
::: {.col-md-6}

```{r upset_plot, echo=FALSE, message=FALSE, warning=FALSE, eval=has_unfiltered}

qcFrame <- colData(spe) |> data.frame() |> dplyr::select(starts_with("qc."))
qcFrame[is.na(qcFrame)] <- FALSE
## select only columns that have at least one flag
useColumn <- apply(qcFrame, 2, any)
qcFrame <- qcFrame[ , useColumn, drop=FALSE]
colnames(qcFrame) <- c("Transcript_Counts", "nGenes")


if (ncol(qcFrame) > 1) {
  print(UpSetR::upset(
    qcFrame + 0, ## convert from logical to integer
    nsets = min(5, ncol(qcFrame)),
    mainbar.y.label = "cells"
  ))
  grid.text("cells removed", x = 0.65, y = 0.95, gp = gpar(fontsize = 20))
}

```
:::

::: {.col-md-6}

```{r qc_stats_table, echo=FALSE, message=FALSE, warning=FALSE, eval=has_unfiltered}
meta <- colData(spe)

qcStats <- ezFrame(
  "Number of cells" = numeric(0),
  "Min Pass Value"  = numeric(0),
  "Max Pass Value"  = numeric(0),
  "Fraction"        = numeric(0)
)

qcStats["Transcript Counts", ] <- c(
  c(sum(meta$qc.lib),
    range(meta[[nCountCol]][meta$useCell])) %>% as.integer(),
  round(sum(meta$qc.lib) / nrow(meta), 2)
)

qcStats["# Genes", ] <- c(
  c(sum(meta$qc.nexprs),
    range(meta[[nFeatureCol]][meta$useCell])) %>% as.integer(),
  round(sum(meta$qc.nexprs) / nrow(meta), 2)
)

ezKable(qcStats)
```

:::

:::



<details>
  <summary>Metrics to identify cells of low quality</summary>

```{r qc_metrics, echo=FALSE, message=FALSE, warning=FALSE}
qcMetrics <- tibble(
  metric = character(),
  interpretation = character()
) |>
  add_row(
    metric = "Library size – total sum of counts across all relevant features for each cell",
    interpretation = "Small library sizes indicate low quality (RNA may have been lost during library preparation)"
  ) |>
  add_row(
    metric = "Number of expressed genes with non-zero counts in each cell",
    interpretation = "Very few expressed genes means poor capture of the transcript population"
  )

ezKable(qcMetrics)
```

Evaluation

```{r diagnostics_plots, echo=FALSE, message=FALSE, warning=FALSE}
qcDiagnostics <- tibble(
  check = character(),
  interpretation = character()
) |>
  add_row(
    check = "Independence of QC metrics",
    interpretation = "QC metrics should reflect technical quality, not biological state (e.g., low library size should not correspond to specific cell types)."
  ) |>
  add_row(
    check = "Distribution of values",
    interpretation = "Ideally, metrics follow a normal distribution, supporting threshold-based outlier detection."
  ) |>
  add_row(
    check = "Secondary modes",
    interpretation = "A large secondary peak (another mode) suggests correlation with biology, risking loss of distinct cell types."
  ) |>
  add_row(
    check = "Violin plots",
    interpretation = "Show cells kept (`FALSE`) vs discarded (`TRUE`) after QC filtering, highlighting potential biases."
  ) |>
  add_row(
    check = "Spatial plots",
    interpretation = "Visualize QC metrics across the tissue to detect spatial patterns of low quality."
  )

ezKable(qcDiagnostics)
```
</details>




```{r nCount_plots, fig.width=12, fig.height=6, eval=has_unfiltered}
plot1 <- plotColData(
  spe, x = "Sample",
  y = nCountCol,
  colour_by = "discard") + 
  scale_y_log10() +
  labs(title = "Transcript counts", y = "Transcript counts per cell") +
  theme(legend.position = "none")

plot2 <- plotColData(
  spe,
  x = nCountCol,
  y = nFeatureCol,
  colour_by = "discard") +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "", x = "Transcript counts", y = "#genes")

plot3 <- ImageFeaturePlot(
  scData_unfiltered, features = nCountCol,
  max.cutoff = "q95", size = 0.5, axes = TRUE
) +
  labs(title = "Transcripts per cell", fill = "Transcript counts") +
  scale_fill_viridis_c(option = "turbo", na.value = "grey95") +
  theme(legend.position = "bottom")

wrap_plots(
  (plot1 | plot2) + plot_layout(guides = "collect") & theme(legend.position = "bottom"),
  plot3,
  widths = c(2, 2))

```


```{r nGenes_plots, fig.width=12, fig.height=4, eval=has_unfiltered}
plot1 <- plotColData(
  spe, x = "Sample",
  y = nFeatureCol,
  colour_by = "discard"
) +
  labs(title = "Genes", y = "#genes per cell")+
  theme(legend.position = "bottom",
        legend.direction = "vertical")

plot2 <- ImageFeaturePlot(
  scData_unfiltered, features = nFeatureCol,
  max.cutoff = "q95", size = 0.5, axes = TRUE
) +
  labs(title = "Genes per cell", fill = "#genes") +
  scale_fill_viridis_c(option = "turbo", na.value = "grey95")+
  theme(legend.position = "bottom")

plot3 <- ImageDimPlot(
  scData_unfiltered, group.by = "Sample",
  size = 0.1, axes = TRUE, cols = "grey90"
) +
  labs(title = "Tissue overview") +
  guides(color = "none")+
  theme(legend.position = "bottom")

plot1 + plot2 + plot3 + plot_layout(widths = c(1, 2, 2))
```

## QC {.tabset}

### Overview

**Sample:** `r paste(unique(scData$Sample), collapse = ", ")`

**Total cells:** `r format(ncol(scData), big.mark = ",")`

**Total genes:** `r nrow(scData)`

**Assays:** `r base::paste(names(scData@assays), collapse = ", ")`

### Spatial QC {.tabset}

Spatial distribution of key quality control metrics across the tissue.

#### Transcript counts

```{r spatial_qc_counts, fig.width=14, fig.height=6}
p1 <- ImageFeaturePlot(scData, features = nCountCol, max.cutoff = "q95", size = 0.5) +
 ggtitle("Transcript counts - Spatial") +
 theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p2 <- FeaturePlot(scData, features = nCountCol, reduction = "umap",
                 max.cutoff = "q95", raster = FALSE, order = TRUE) +
 coord_fixed() +
 ggtitle("Transcript counts - UMAP") +
 theme_void() +
 theme(plot.title = element_text(hjust = 0.5, face = "bold"))

(p1 + p2) + plot_layout(widths = c(1.2, 1))
```

#### Genes detected

```{r spatial_qc_features, fig.width=14, fig.height=6}
p1 <- ImageFeaturePlot(scData, features = nFeatureCol, max.cutoff = "q95", size = 0.5) +
 ggtitle("Genes detected - Spatial") +
 theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p2 <- FeaturePlot(scData, features = nFeatureCol, reduction = "umap",
                 max.cutoff = "q95", raster = FALSE, order = TRUE) +
 coord_fixed() +
 ggtitle("Genes detected - UMAP") +
 theme_void() +
 theme(plot.title = element_text(hjust = 0.5, face = "bold"))

(p1 + p2) + plot_layout(widths = c(1.2, 1))
```

#### Blank codeword counts

```{r spatial_qc_blank, fig.width=14, fig.height=6}
if ("nCount_BlankCodeword" %in% colnames(scData@meta.data)) {
 p1 <- ImageFeaturePlot(scData, features = "nCount_BlankCodeword",
                        max.cutoff = "q95", size = 0.5) +
   ggtitle("Blank codeword - Spatial") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold"))

 p2 <- FeaturePlot(scData, features = "nCount_BlankCodeword", reduction = "umap",
                   max.cutoff = "q95", raster = FALSE, order = TRUE) +
   coord_fixed() +
   ggtitle("Blank codeword - UMAP") +
   theme_void() +
   theme(plot.title = element_text(hjust = 0.5, face = "bold"))

 print((p1 + p2) + plot_layout(widths = c(1.2, 1)))
} else {
 cat("Blank codeword information not available in metadata.")
}
```

#### Control codeword counts

```{r spatial_qc_control_codeword, fig.width=14, fig.height=6}
if ("nCount_ControlCodeword" %in% colnames(scData@meta.data)) {
 p1 <- ImageFeaturePlot(scData, features = "nCount_ControlCodeword",
                        max.cutoff = "q95", size = 0.5) +
   ggtitle("Control codeword - Spatial") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold"))

 p2 <- FeaturePlot(scData, features = "nCount_ControlCodeword", reduction = "umap",
                   max.cutoff = "q95", raster = FALSE, order = TRUE) +
   coord_fixed() +
   ggtitle("Control codeword - UMAP") +
   theme_void() +
   theme(plot.title = element_text(hjust = 0.5, face = "bold"))

 print((p1 + p2) + plot_layout(widths = c(1.2, 1)))
} else {
 cat("Control codeword information not available in metadata.")
}
```

#### Control probe counts

```{r spatial_qc_control_probe, fig.width=14, fig.height=6}
if ("nCount_ControlProbe" %in% colnames(scData@meta.data)) {
 p1 <- ImageFeaturePlot(scData, features = "nCount_ControlProbe",
                        max.cutoff = "q95", size = 0.5) +
   ggtitle("Control probe - Spatial") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold"))

 p2 <- FeaturePlot(scData, features = "nCount_ControlProbe", reduction = "umap",
                   max.cutoff = "q95", raster = FALSE, order = TRUE) +
   coord_fixed() +
   ggtitle("Control probe - UMAP") +
   theme_void() +
   theme(plot.title = element_text(hjust = 0.5, face = "bold"))

 print((p1 + p2) + plot_layout(widths = c(1.2, 1)))
} else {
 cat("Control probe information not available in metadata.")
}
```

### QC histograms {.tabset}

Distribution of transcript counts and genes detected across all cells.

#### Transcript distribution

```{r hist_transcripts, fig.width=10, fig.height=5}
ggplot(scData@meta.data, aes_string(x = nCountCol)) +
 geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
 scale_x_log10(labels = label_number()) +
 theme_minimal(base_size = 12) +
 labs(title = "Distribution of transcript counts per cell",
      x = "Transcript counts (log10 scale)", y = "Number of cells") +
 theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

#### Gene distribution

```{r hist_genes, fig.width=10, fig.height=5}
ggplot(scData@meta.data, aes_string(x = nFeatureCol)) +
 geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
 scale_x_log10(labels = label_number()) +
 theme_minimal(base_size = 12) +
 labs(title = "Distribution of genes detected per cell",
      x = "Genes detected (log10 scale)", y = "Number of cells") +
 theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

### Cell morphology {.tabset}

Cell and nucleus area distributions (if available from segmentation).

#### Cell area

```{r cell_area, fig.width=14, fig.height=5}
# Check for cell area column with valid numeric data
cell_area_col <- grep("cell_area|CellArea", colnames(scData@meta.data),
                     value = TRUE, ignore.case = TRUE) |> head(1)

has_valid_cell_area <- length(cell_area_col) > 0 && cell_area_col != "" &&
  is.numeric(scData@meta.data[[cell_area_col]]) &&
  sum(!is.na(scData@meta.data[[cell_area_col]])) > 0

if (has_valid_cell_area) {
 # Reference cell sizes (area = pi * (diameter/2)^2)
 ref_sizes <- data.frame(
   cell_type = c("Lymphocyte\n(~8um)", "Monocyte\n(~15um)",
                 "Macrophage\n(~25um)", "Large cell\n(~40um)"),
   area = c(50, 177, 491, 1257)
 )

 p1 <- ggplot(scData@meta.data, aes_string(x = cell_area_col)) +
   geom_histogram(bins = 50, fill = "#9b59b6", color = "white", alpha = 0.8) +
   geom_vline(data = ref_sizes, aes(xintercept = area),
              linetype = "dashed", color = "red", alpha = 0.7) +
   geom_text(data = ref_sizes, aes(x = area, y = Inf, label = cell_type),
             vjust = 1.5, hjust = -0.1, size = 2.5, color = "red", angle = 0) +
   theme_minimal(base_size = 12) +
   labs(title = "Cell area distribution",
        subtitle = "Red lines: typical cell type sizes",
        x = expression(paste("Cell area (", mu, m^2, ")")),
        y = "Number of cells") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold"))

 p2 <- ImageFeaturePlot(scData, features = cell_area_col, max.cutoff = "q95",
                        size = 0.5) +
   ggtitle("Cell area - Spatial") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold"))

 print(p1 + p2)
} else {
 cat("Cell area information not available in metadata.")
}
```

#### Nucleus area

```{r nucleus_area, fig.width=14, fig.height=5, results='asis'}
# Check for nucleus area column with valid numeric data
nucleus_area_col <- grep("nucleus_area|NucleusArea", colnames(scData@meta.data),
                        value = TRUE, ignore.case = TRUE) |> head(1)

has_valid_nucleus_area <- length(nucleus_area_col) > 0 && nucleus_area_col != "" &&
  is.numeric(scData@meta.data[[nucleus_area_col]])

if (has_valid_nucleus_area) {
 # Handle NAs in nucleus_area (cells without detected nuclei)
 na_count <- sum(is.na(scData@meta.data[[nucleus_area_col]]))
 if (na_count > 0) {
   scData@meta.data[[nucleus_area_col]][is.na(scData@meta.data[[nucleus_area_col]])] <- 0
   cat("\n\n*Note:* ", na_count, " cells had NA nucleus_area (no nucleus detected), set to 0.\n\n", sep = "")
 }

 p1 <- ggplot(scData@meta.data, aes_string(x = nucleus_area_col)) +
   geom_histogram(bins = 50, fill = "#e74c3c", color = "white", alpha = 0.8) +
   theme_minimal(base_size = 12) +
   labs(title = "Nucleus area distribution",
        x = expression(paste("Nucleus area (", mu, m^2, ")")),
        y = "Number of cells") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold"))

 p2 <- ImageFeaturePlot(scData, features = nucleus_area_col, max.cutoff = "q95",
                        size = 0.5) +
   ggtitle("Nucleus area - Spatial") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold"))

 print(p1 + p2)
} else {
 cat("Nucleus area information not available in metadata.")
}
```

#### Segmentation method

```{r segmentation_method, fig.width=10, fig.height=8}
# Check for segmentation_method column (available in newer Xenium outputs)
seg_col <- grep("segmentation_method", colnames(scData@meta.data),
               value = TRUE, ignore.case = TRUE) |> head(1)

has_valid_seg <- length(seg_col) > 0 && seg_col != "" &&
  !all(is.na(scData@meta.data[[seg_col]]))

if (has_valid_seg) {
 seg_vals <- unique(scData@meta.data[[seg_col]])

 # Build color map based on segmentation type patterns
 color_map <- sapply(seg_vals, function(v) {
   if (grepl("boundary", v, ignore.case = TRUE)) return("#FF9999")
   if (grepl("interior", v, ignore.case = TRUE)) return("#DAA520")
   if (grepl("nucle", v, ignore.case = TRUE)) return("#000080")
   return("grey50")
 })
 names(color_map) <- seg_vals

 ImageDimPlot(scData, fov = names(scData@images)[1], group.by = seg_col,
              cols = color_map, size = 0.5) +
   ggtitle("Segmentation method") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
   labs(fill = "Method")
} else {
 cat("Segmentation method information not available in metadata.")
}
```

### Dimensionality reduction

<details>
 <summary>PCA and UMAP methodology</summary>

```{r dimred_table}
dimred_tbl <- tibble(
 topic = character(),
 summary = character()
) |>
 add_row(
   topic = "Goal",
   summary = "Reduce dimensions by compressing information from many genes into fewer components."
 ) |>
 add_row(
   topic = "Benefits",
   summary = "Computational efficiency, noise reduction, and visualization."
 ) |>
 add_row(
   topic = "Method",
   summary = "PCA for linear reduction, UMAP for non-linear visualization."
 )

ezKable(dimred_tbl)
```
</details>

```{r pca_variance, fig.width=10, fig.height=5}
if ("pca" %in% names(scData@reductions)) {
 stdev <- scData[["pca"]]@stdev
 n_use <- min(30, length(stdev))

 var_pct <- (stdev^2) / sum(stdev^2) * 100
 cum_pct <- cumsum(var_pct)

 df <- data.frame(
   PC = seq_along(stdev),
   perc = var_pct,
   cum = cum_pct,
   used = seq_along(stdev) <= n_use
 )

 ggplot(df, aes(x = PC)) +
   geom_col(aes(y = perc, fill = used), width = 0.7, show.legend = FALSE) +
   scale_fill_manual(values = c(`TRUE` = "grey20", `FALSE` = "grey75")) +
   geom_line(aes(y = cum / max(cum) * max(perc), group = 1),
             linewidth = 0.5, color = "orange") +
   geom_point(aes(y = cum / max(cum) * max(perc)), size = 1, color = "orange") +
   geom_vline(xintercept = n_use, linetype = "dashed", color = "grey40") +
   scale_y_continuous(
     name = "Variance explained (%)",
     sec.axis = sec_axis(~ . / max(df$perc) * 100, name = "Cumulative variance (%)")
   ) +
   labs(title = "PCA variance explained",
        subtitle = paste0("PCs used: ", n_use, " | Cumulative: ",
                          sprintf("%.2f%%", df$cum[n_use])),
        x = NULL) +
   theme_minimal(base_size = 9) +
   theme(
     text = element_text(face = "bold"),
     axis.title.y.left = element_text(color = "grey20"),
     axis.title.y.right = element_text(color = "orange"),
     plot.title = element_text(hjust = 0.5),
     panel.grid.minor = element_blank()
   )
}
```

## Clusters {.tabset}

### Spatial + UMAP

```{r clustering, fig.width=14, fig.height=7}
n_clusters <- nlevels(Idents(scData))
show_leg <- n_clusters <= 20

Idents(scData) <- "seurat_clusters"

# Spatial plot first
p1 <- ImageDimPlot(scData, fov = names(scData@images)[1], group.by = "ident",
                  axes = TRUE, size = 0.5) +
 ggtitle("Spatial clusters") +
 labs(fill = "Clusters") +
 guides(fill = if (show_leg) guide_legend(title = "Clusters", override.aes = list(size = 3)) else "none")

# UMAP second with coord_fixed
p2 <- DimPlot(scData, reduction = "umap", label = FALSE, pt.size = 0.5, raster = FALSE) +
 coord_fixed() +
 labs(color = "Clusters") +
 theme_void() +
 ggtitle("UMAP") +
 theme(legend.position = "none")

p2 <- Seurat::LabelClusters(
 plot = p2,
 id = "ident",
 fontface = "bold",
 box = TRUE,
 repel = TRUE,
 color = "black",
 size = 4
)

print((p1 + p2) + plot_layout(nrow = 1, widths = c(1.2, 1), guides = if (show_leg) "collect" else "keep"))
```

`r paste("Resolution:", ifelse(!is.null(param$clusterResolution), param$clusterResolution, 0.5))`

<details>
 <summary>Clustering methodology</summary>

```{r cluster_method}
cluster_tbl <- tibble(
 Step = character(),
 Details = character()
) |>
 add_row(
   Step = "Build KNN graph",
   Details = "Each node = cell. Connect each cell to its k nearest neighbors in PCA space."
 ) |>
 add_row(
   Step = "Community detection",
   Details = "Apply Leiden/Louvain algorithm to find groups more connected internally."
 ) |>
 add_row(
   Step = "Visualization",
   Details = "Plot clusters on UMAP and in spatial context."
 )

ezKable(cluster_tbl)
```
</details>

::: {.row}
::: {.col-md-8}

```{r cells_per_cluster, fig.width=10, fig.height=5}
cellIdents <- scData@meta.data[, c("seurat_clusters", "Sample")]

ggplot(cellIdents, aes(x = seurat_clusters, fill = Sample)) +
 geom_bar(fill = "#93AECA") +
 labs(title = "Cells per cluster",
      x = NULL, y = "Cells") +
 theme_minimal() +
 scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
 theme(legend.position = "top",
       legend.justification = "left",
       panel.grid.minor = element_blank(),
       panel.grid.major = element_blank(),
       axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

:::
::: {.col-md-4}

```{r cells_table}
df <- cellsProportion(scData, groupVar1 = "seurat_clusters", groupVar2 = "Sample")
df <- df |> as.data.frame(check.names = FALSE, stringsAsFactors = FALSE) |>
 dplyr::rename("Cluster" = "seurat_clusters")

is_pct_col <- grepl("_fraction$", colnames(df))
colnames(df) <- colnames(df) |> str_replace_all("_fraction$", " (%)")
is_count_col <- !is_pct_col & colnames(df) != "Cluster"
colnames(df)[is_count_col] <- colnames(df)[is_count_col] |> paste("(n)")

ezInteractiveTableRmd(df, rowNames = FALSE, title = "Cells per cluster",
                     filter = "none") |>
 DT::formatPercentage(columns = is_pct_col, digits = 2)
```

:::
:::

### Cluster assessment

<details>
 <summary>Technical vs. biological clustering drivers</summary>

```{r driver_qc}
driver_qc_tbl <- tibble(
 Item = character(),
 Details = character()
) |>
 add_row(
   Item = "Purpose",
   Details = "Check whether clusters are driven by technical artifacts."
 ) |>
 add_row(
   Item = "Potential confounders",
   Details = "Library size (UMIs), genes detected."
 ) |>
 add_row(
   Item = "Diagnostics",
   Details = "Per-cluster box/violin plots for UMIs and genes."
 )

ezKable(driver_qc_tbl)
```
</details>

```{r cluster_assessment, fig.height=10, fig.width=12}
plot1 <- VlnPlot(scData, nCountCol, group.by = "seurat_clusters",
                pt.size = 0, raster = TRUE) +
 ggtitle("Number of UMIs vs cluster") +
 ylab(nCountCol) +
 theme_minimal(base_size = 12) +
 theme(legend.position = "none", axis.title.x = element_blank()) +
 scale_y_log10(breaks = log_breaks(base = 10), labels = label_number())

plot2 <- VlnPlot(scData, nFeatureCol, group.by = "seurat_clusters",
                pt.size = 0, raster = TRUE) +
 ggtitle("Number of genes vs cluster") +
 ylab(nFeatureCol) +
 theme_minimal(base_size = 12) +
 theme(legend.position = "none", axis.title.x = element_blank()) +
 scale_y_log10(breaks = log_breaks(base = 10), labels = label_number())

plot1 + plot2 + plot_layout(nrow = 2, ncol = 1)
```

### Cluster markers

<details>
 <summary>Positive marker detection</summary>

```{r marker_method}
marker_tbl <- tibble(
 Step = character(),
 Details = character()
) |>
 add_row(
   Step = "Goal",
   Details = "Identify genes significantly higher in a given cluster vs. all other cells."
 ) |>
 add_row(
   Step = "Test",
   Details = "Wilcoxon Rank Sum test (non-parametric, robust)."
 ) |>
 add_row(
   Step = "Effect size filter",
   Details = "Average log-fold change >= 0.25."
 ) |>
 add_row(
   Step = "Significance",
   Details = "Adjusted p-value < 0.05."
 )

ezKable(marker_tbl)
```
</details>

```{r markers_table}
if (nrow(posMarkers) > 0) {
 ezInteractiveTableRmd(posMarkers, digits = 3, rowNames = FALSE)
} else {
 cat("No significant markers found.")
}
```

### Marker plots

```{r dotplot_setup}
if (nrow(posMarkers) > 0) {
 # Sort by diff_pct if available, otherwise by avg_log2FC
 if ("diff_pct" %in% colnames(posMarkers)) {
   top5 <- posMarkers |>
     group_by(cluster) |>
     arrange(desc(diff_pct), desc(avg_log2FC), .by_group = TRUE) |>
     slice_head(n = 5)
 } else {
   top5 <- posMarkers |>
     group_by(cluster) |>
     arrange(desc(avg_log2FC), .by_group = TRUE) |>
     slice_head(n = 5)
 }

 genesToPlot <- unique(as.character(top5$gene))
 genesToPlot <- gsub("_", "-", genesToPlot)
 genesToPlot <- intersect(genesToPlot, rownames(scData))
}
```

#### DotPlot

Visualization of the top 5 markers in each cluster.

```{r dotplot, fig.width=15, fig.height=20, eval=nrow(posMarkers) > 0}
DotPlot(scData, features = rev(genesToPlot)) +
 coord_flip() +
 scale_y_discrete(limits = as.character(sort(as.integer(levels(Idents(scData)))))) +
 theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

### EnrichR

<details>
 <summary>Cell type scoring (Enrichr)</summary>

```{r enrichr_method}
enrichr_tbl <- tibble(
 Item = character(),
 Details = character()
) |>
 add_row(
   Item = "Goal",
   Details = "Score clusters by cell-type/pathway activity via gene set enrichment."
 ) |>
 add_row(
   Item = "Method",
   Details = "Use Enrichr to perform over-representation analysis on marker genes."
 )

ezKable(enrichr_tbl)
```
</details>

```{r enrichr_links, results='asis', eval=nrow(posMarkers) > 0}
markersPerClusterTable <- c()
for (eachCluster in levels(posMarkers$cluster)) {
 markersPerCluster <- dplyr::filter(posMarkers, cluster == eachCluster) |>
   dplyr::arrange(desc(avg_log2FC))
 markersPerCluster <- head(markersPerCluster, min(nrow(markersPerCluster), 500))
 markersPerClusterTable <- rbind(markersPerClusterTable, markersPerCluster)
}

genesPerCluster <- split(markersPerClusterTable$gene, markersPerClusterTable$cluster)

jsCall <- paste0('enrich({list: "',
                sapply(genesPerCluster, paste, collapse = "\\n"),
                '", popup: true});')

enrichrCalls <- paste0("<a href='javascript:void(0)' onClick='",
                      jsCall, "'>Open</a>")

enrichrTable <- tibble(
 Cluster = names(genesPerCluster),
 "N posMarkers" = lengths(genesPerCluster),
 "Enrichr link" = enrichrCalls
)

ezKable(enrichrTable, escape = FALSE, format = "html", format.args = list())
```

## Niches {.tabset}

```{r check_banksy}
has_banksy <- "banksy_cluster" %in% colnames(scData@meta.data)
has_niche_genes <- FALSE
```

### Spatial + UMAP

```{r banksy_spatial, fig.width=14, fig.height=7, eval=has_banksy}
Idents(scData) <- "banksy_cluster"

n_clusters_banksy <- nlevels(Idents(scData))
show_leg_banksy <- n_clusters_banksy <= 20

# Spatial plot first
p1 <- ImageDimPlot(scData, fov = names(scData@images)[1], group.by = "banksy_cluster",
                  axes = TRUE, size = 0.5) +
 ggtitle("Spatial niches") +
 labs(fill = "Niches") +
 guides(fill = if (show_leg_banksy) guide_legend(title = "Niches", override.aes = list(size = 3)) else "none")

# UMAP second with coord_fixed
p2 <- DimPlot(scData, reduction = "umap", label = FALSE, pt.size = 0.5, raster = FALSE) +
 coord_fixed() +
 labs(color = "Niches") +
 theme_void() +
 ggtitle("UMAP") +
 theme(legend.position = "none")

p2 <- Seurat::LabelClusters(
 plot = p2,
 id = "ident",
 fontface = "bold",
 box = TRUE,
 repel = TRUE,
 color = "black",
 size = 4
)

print((p1 + p2) + plot_layout(nrow = 1, widths = c(1.2, 1), guides = if (show_leg_banksy) "collect" else "keep"))
```

`r if(has_banksy) paste("Lambda:", ifelse(!is.null(param$lambda), param$lambda, 0.8))`

`r if(has_banksy) paste("Resolution:", ifelse(!is.null(param$nicheResolution), param$nicheResolution, 0.5))`

<details>
 <summary>BANKSY niche methodology</summary>

```{r banksy_method, eval=has_banksy}
banksy_tbl <- tibble(
 Concept = character(),
 Definition = character()
) |>
 add_row(
   Concept = "Niche (BANKSY cluster)",
   Definition = "Community of spatially adjacent cells with neighborhood-smoothed expression."
 ) |>
 add_row(
   Concept = "k_geom",
   Definition = "Local neighborhood size. Larger values yield larger domains."
 ) |>
 add_row(
   Concept = "lambda",
   Definition = "How strongly neighbors affect each cell's expression (0-1)."
 )

ezKable(banksy_tbl)
```
</details>

```{r no_banksy, eval=!has_banksy}
cat("BANKSY analysis was not performed or failed.")
```

::: {.row}
::: {.col-md-8}

```{r cells_per_niche, fig.width=10, fig.height=5, eval=has_banksy}
nicheIdents <- scData@meta.data[, c("banksy_cluster", "Sample")]

ggplot(nicheIdents, aes(x = banksy_cluster, fill = Sample)) +
 geom_bar(fill = "#93AECA") +
 labs(title = "Cells per niche",
      x = NULL, y = "Cells") +
 theme_minimal() +
 scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
 theme(legend.position = "top",
       legend.justification = "left",
       panel.grid.minor = element_blank(),
       panel.grid.major = element_blank(),
       axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

:::
::: {.col-md-4}

```{r niche_table, eval=has_banksy}
df_niche <- cellsProportion(scData, groupVar1 = "banksy_cluster", groupVar2 = "Sample")
df_niche <- df_niche |> as.data.frame(check.names = FALSE, stringsAsFactors = FALSE) |>
 dplyr::rename("Niche" = "banksy_cluster")

is_pct_col_niche <- grepl("_fraction$", colnames(df_niche))
colnames(df_niche) <- colnames(df_niche) |> str_replace_all("_fraction$", " (%)")
is_count_col_niche <- !is_pct_col_niche & colnames(df_niche) != "Niche"
colnames(df_niche)[is_count_col_niche] <- colnames(df_niche)[is_count_col_niche] |> paste("(n)")

ezInteractiveTableRmd(df_niche, rowNames = FALSE, title = "Cells per niche",
                     filter = "none") |>
 DT::formatPercentage(columns = is_pct_col_niche, digits = 2)
```

:::
:::

### Niche assessment

```{r niche_assessment, fig.height=10, fig.width=12, eval=has_banksy}
plot1 <- VlnPlot(scData, nCountCol, group.by = "banksy_cluster",
                pt.size = 0, raster = TRUE) +
 ggtitle("Number of UMIs vs niche") +
 ylab(nCountCol) +
 theme_minimal(base_size = 12) +
 theme(legend.position = "none", axis.title.x = element_blank()) +
 scale_y_log10(breaks = log_breaks(base = 10), labels = label_number())

plot2 <- VlnPlot(scData, nFeatureCol, group.by = "banksy_cluster",
                pt.size = 0, raster = TRUE) +
 ggtitle("Number of genes vs niche") +
 ylab(nFeatureCol) +
 theme_minimal(base_size = 12) +
 theme(legend.position = "none", axis.title.x = element_blank()) +
 scale_y_log10(breaks = log_breaks(base = 10), labels = label_number())

plot1 + plot2 + plot_layout(nrow = 2, ncol = 1)
```

### Niche markers

```{r niche_markers, eval=has_banksy && nrow(posMarkersBanksy) > 0}
ezInteractiveTableRmd(posMarkersBanksy, digits = 3, rowNames = FALSE)
```

```{r no_niche_markers, eval=!has_banksy || nrow(posMarkersBanksy) == 0}
cat("No BANKSY niche markers available.")
```

### Marker plots

```{r niche_dotplot_setup, eval=has_banksy && nrow(posMarkersBanksy) > 0}
top5_banksy <- posMarkersBanksy |>
 group_by(cluster) |>
 arrange(desc(diff_pct), desc(avg_log2FC), .by_group = TRUE) |>
 slice_head(n = 5)

genesToPlot_banksy <- unique(as.character(top5_banksy$gene))
genesToPlot_banksy <- gsub("\\.m[0-9]+$", "", genesToPlot_banksy)
genesToPlot_banksy <- gsub("_", "-", genesToPlot_banksy)
genesToPlot_banksy <- intersect(genesToPlot_banksy, rownames(scData[["Xenium"]]))
genesToPlot_banksy <- unique(genesToPlot_banksy)

has_niche_genes <- length(genesToPlot_banksy) > 0
```

```{r niche_dotplot, fig.width=15, fig.height=20, eval=has_banksy && nrow(posMarkersBanksy) > 0 && has_niche_genes}
Idents(scData) <- "banksy_cluster"
DotPlot(scData, features = rev(genesToPlot_banksy), assay = "Xenium") +
 coord_flip() +
 scale_y_discrete(limits = as.character(sort(as.integer(levels(Idents(scData)))))) +
 theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

```{r no_niche_genes, eval=has_banksy && nrow(posMarkersBanksy) > 0 && !has_niche_genes}
cat("No matching genes found in the Seurat object for the niche marker dotplot.")
```

### EnrichR

```{r enrichr_niches, results='asis', eval=has_banksy && nrow(posMarkersBanksy) > 0}
markersPerNicheTable <- posMarkersBanksy |>
 dplyr::group_by(cluster) |>
 dplyr::arrange(desc(avg_log2FC)) |>
 dplyr::slice_head(n = 500) |>
 dplyr::ungroup()

genesPerNiche <- split(markersPerNicheTable$gene, markersPerNicheTable$cluster)

jsCallNiche <- paste0('enrich({list: "',
                     sapply(genesPerNiche, paste, collapse = "\\n"),
                     '", popup: true});')

enrichrCallsNiche <- paste0("<a href='javascript:void(0)' onClick='",
                           jsCallNiche, "'>Open</a>")

enrichrTableNiche <- tibble(
 Niche = names(genesPerNiche),
 "N posMarkers" = lengths(genesPerNiche),
 "Enrichr link" = enrichrCallsNiche
)

ezKable(enrichrTableNiche, escape = FALSE, format = "html", format.args = list())
```

```{r no_enrichr_niches, eval=!has_banksy || nrow(posMarkersBanksy) == 0}
cat("EnrichR analysis not available for niches (no BANKSY results or no markers).")
```

### Niche x cluster

```{r niche_cluster_heatmap, fig.width=10, fig.height=8, eval=has_banksy}
# Calculate proportion of each cluster within each niche (row-normalized)
niche_cluster_df <- scData@meta.data |>
 dplyr::filter(!is.na(banksy_cluster), !is.na(seurat_clusters)) |>
 dplyr::group_by(banksy_cluster, seurat_clusters) |>
 dplyr::summarise(count = n(), .groups = "drop") |>
 dplyr::group_by(banksy_cluster) |>
 dplyr::mutate(proportion = count / sum(count)) |>
 dplyr::ungroup()

# Create proportion matrix
prop_matrix <- niche_cluster_df |>
 dplyr::select(banksy_cluster, seurat_clusters, proportion) |>
 tidyr::pivot_wider(names_from = seurat_clusters, values_from = proportion, values_fill = 0) |>
 tibble::column_to_rownames("banksy_cluster") |>
 as.matrix()

# Sort rows and columns numerically if possible
tryCatch({
 row_order <- order(as.numeric(rownames(prop_matrix)))
 col_order <- order(as.numeric(colnames(prop_matrix)))
 prop_matrix <- prop_matrix[row_order, col_order, drop = FALSE]
}, error = function(e) NULL)

# Convert to ggplot for better axis labels
prop_long <- prop_matrix |>
 as.data.frame() |>
 tibble::rownames_to_column("Niche") |>
 tidyr::pivot_longer(cols = -Niche, names_to = "Cluster", values_to = "Proportion")

# Order factors
prop_long$Niche <- factor(prop_long$Niche, levels = rownames(prop_matrix))
prop_long$Cluster <- factor(prop_long$Cluster, levels = colnames(prop_matrix))

ggplot(prop_long, aes(x = Cluster, y = Niche, fill = Proportion)) +
 geom_tile(color = "grey90") +
 scale_fill_gradient(low = "white", high = "#2166AC") +
 labs(x = "Cluster", y = "Niche", title = "Cluster composition per niche",
      fill = "Proportion") +
 theme_minimal(base_size = 12) +
 theme(axis.text.x = element_text(angle = 45, hjust = 1),
       plot.title = element_text(hjust = 0.5, face = "bold"))
```

```{r no_niche_cluster_heatmap, eval=!has_banksy}
cat("Niche x Cluster heatmap not available (no BANKSY results).")
```

## Cell type annotation {.tabset}

```{r check_annotation}
# Check for RCTD annotation columns
has_rctd <- "RCTD_Main" %in% colnames(scData@meta.data)
has_spot_class <- "spot_class" %in% colnames(scData@meta.data)

# Use RCTD_Main or predicted.celltype
if (has_rctd) {
 celltype_col <- "RCTD_Main"
} else if ("predicted.celltype" %in% colnames(scData@meta.data)) {
 celltype_col <- "predicted.celltype"
} else {
 celltype_col <- NULL
}

has_annotation <- !is.null(celltype_col)

# Get RCTD reference information from param
rctd_ref_name <- if (!is.null(param$rctdReference) && param$rctdReference != "" && param$rctdReference != "None") {
 param$rctdReference
} else {
 NULL
}

rctd_ref_file <- if (!is.null(param$rctdFile) && param$rctdFile != "") {
 param$rctdFile
} else {
 NULL
}
```

### Reference info

```{r rctd_reference_info, eval=has_annotation && !is.null(rctd_ref_name), results='asis'}
cat("**RCTD Reference:** ", rctd_ref_name, "\n\n", sep = "")
if (!is.null(rctd_ref_file)) {
 cat("**Reference file:** ", rctd_ref_file, "\n\n", sep = "")
}
umi_min_used <- ifelse(!is.null(param$rctdUMImin), param$rctdUMImin, 100)
cat("**UMI_min threshold:** ", umi_min_used, " (cells below this were not classified)\n\n", sep = "")
cat("**Number of cell types:** ", length(unique(na.omit(scData@meta.data[[celltype_col]]))), "\n\n", sep = "")
```

```{r no_rctd_ref, eval=!has_annotation || is.null(rctd_ref_name)}
cat("RCTD annotation was not performed or reference information not available.")
```

### Cell types

```{r celltype_spatial, fig.width=18, fig.height=10, eval=has_annotation}
# Get unique cell types and create consistent color mapping
celltypes_unique <- sort(unique(na.omit(scData@meta.data[[celltype_col]])))
n_celltypes <- length(celltypes_unique)
extended_colors <- c(
 cluster_colors,
 RColorBrewer::brewer.pal(9, "Set1"),
 RColorBrewer::brewer.pal(8, "Set2"),
 RColorBrewer::brewer.pal(12, "Set3")
)
extended_colors <- unique(extended_colors)
if (n_celltypes > length(extended_colors)) {
 extended_colors <- rep(extended_colors, length.out = n_celltypes)
}
celltype_colors <- setNames(extended_colors[1:n_celltypes], celltypes_unique)

# Create truncated labels for legend
truncate_label <- function(x, max_chars = 35) {
 ifelse(nchar(x) > max_chars, paste0(substr(x, 1, max_chars - 3), "..."), x)
}
celltype_labels <- setNames(truncate_label(celltypes_unique), celltypes_unique)

# Spatial plot first
p1 <- ImageDimPlot(scData, fov = names(scData@images)[1],
                  group.by = celltype_col,
                  axes = TRUE, size = 0.5, cols = celltype_colors) +
 scale_fill_manual(values = celltype_colors, labels = celltype_labels) +
 ggtitle("Cell types - Spatial") +
 labs(fill = "Cell type") +
 theme(legend.position = "bottom",
       legend.text = element_text(size = 7, color = "white")) +
 guides(fill = guide_legend(ncol = 4, override.aes = list(size = 3)))

# UMAP second with coord_fixed
p2 <- DimPlot(scData, group.by = celltype_col, cols = celltype_colors,
             label = TRUE, repel = TRUE, pt.size = 0.5, label.size = 3, raster = FALSE) +
 coord_fixed() +
 ggtitle("Cell types - UMAP") +
 theme_void() +
 theme(legend.position = "none")

(p1 + p2) + plot_layout(nrow = 1, widths = c(1.2, 1))
```

```{r no_annotation, eval=!has_annotation}
cat("Cell type annotation was not performed. Select an RCTD reference to enable annotation.")
```

### Cell type proportions

```{r celltype_barplot, fig.width=10, fig.height=8, eval=has_annotation}
if (!exists("celltype_colors")) {
 celltypes_unique <- sort(unique(na.omit(scData@meta.data[[celltype_col]])))
 n_celltypes <- length(celltypes_unique)
 extended_colors <- c(
   cluster_colors,
   RColorBrewer::brewer.pal(9, "Set1"),
   RColorBrewer::brewer.pal(8, "Set2"),
   RColorBrewer::brewer.pal(12, "Set3")
 )
 extended_colors <- unique(extended_colors)
 if (n_celltypes > length(extended_colors)) {
   extended_colors <- rep(extended_colors, length.out = n_celltypes)
 }
 celltype_colors <- setNames(extended_colors[1:n_celltypes], celltypes_unique)
}

if (!exists("truncate_label")) {
 truncate_label <- function(x, max_chars = 35) {
   ifelse(nchar(x) > max_chars, paste0(substr(x, 1, max_chars - 3), "..."), x)
 }
}

celltype_counts <- as.data.frame(table(scData@meta.data[[celltype_col]]))
colnames(celltype_counts) <- c("CellType", "Count")
celltype_counts <- celltype_counts[order(celltype_counts$Count, decreasing = FALSE), ]
celltype_counts$CellType_display <- truncate_label(as.character(celltype_counts$CellType))
celltype_counts$CellType <- factor(celltype_counts$CellType,
                                   levels = celltype_counts$CellType)

ggplot(celltype_counts, aes(x = CellType, y = Count, fill = CellType)) +
 geom_bar(stat = "identity") +
 scale_fill_manual(values = celltype_colors) +
 scale_x_discrete(labels = setNames(celltype_counts$CellType_display, celltype_counts$CellType)) +
 coord_flip() +
 theme_minimal(base_size = 12) +
 theme(legend.position = "none",
       axis.title.y = element_blank(),
       axis.text.y = element_text(size = 9)) +
 labs(title = "Cell type proportions", y = "Number of cells") +
 scale_y_continuous(expand = expansion(mult = c(0, 0.05)))
```

```{r celltype_table, eval=has_annotation}
celltype_prop <- as.data.frame(table(scData@meta.data[[celltype_col]]))
colnames(celltype_prop) <- c("CellType", "Count")
celltype_prop <- celltype_prop[order(celltype_prop$Count, decreasing = TRUE), ]
celltype_prop$Proportion <- celltype_prop$Count / sum(celltype_prop$Count)

ezInteractiveTableRmd(celltype_prop, rowNames = FALSE,
                     title = "Cell type proportions") |>
 DT::formatPercentage(columns = "Proportion", digits = 2)
```

### Niche x cell type

```{r niche_celltype_heatmap_check}
has_rctd_niche <- "RCTD_Main" %in% colnames(scData@meta.data)
if (has_rctd_niche) {
 celltype_col_niche <- "RCTD_Main"
} else if ("predicted.celltype" %in% colnames(scData@meta.data)) {
 celltype_col_niche <- "predicted.celltype"
} else {
 celltype_col_niche <- NULL
}
has_annotation_niche <- !is.null(celltype_col_niche)
```

```{r niche_celltype_heatmap prep, eval=has_banksy && has_annotation_niche}
# Calculate proportion of each cell type within each niche (row-normalized)
niche_celltype_df <- scData@meta.data |>
 dplyr::filter(!is.na(banksy_cluster), !is.na(.data[[celltype_col_niche]])) |>
 dplyr::group_by(banksy_cluster, .data[[celltype_col_niche]]) |>
 dplyr::summarise(count = n(), .groups = "drop") |>
 dplyr::group_by(banksy_cluster) |>
 dplyr::mutate(proportion = count / sum(count)) |>
 dplyr::ungroup()

# Create proportion matrix
prop_matrix_ct <- niche_celltype_df |>
 dplyr::select(banksy_cluster, all_of(celltype_col_niche), proportion) |>
 tidyr::pivot_wider(names_from = all_of(celltype_col_niche), values_from = proportion, values_fill = 0) |>
 tibble::column_to_rownames("banksy_cluster") |>
 as.matrix()

# Sort rows numerically if possible
tryCatch({
 row_order <- order(as.numeric(rownames(prop_matrix_ct)))
 prop_matrix_ct <- prop_matrix_ct[row_order, , drop = FALSE]
}, error = function(e) NULL)

# Truncate long cell type names for display
if (!exists("truncate_label")) {
 truncate_label <- function(x, max_chars = 35) {
   ifelse(nchar(x) > max_chars, paste0(substr(x, 1, max_chars - 3), "..."), x)
 }
}
colnames(prop_matrix_ct) <- truncate_label(colnames(prop_matrix_ct), max_chars = 40)

```

```{r niche_celltype_heatmap plot, eval=has_banksy && has_annotation_niche}

# Heatmap with reduced padding

ht <- ComplexHeatmap::Heatmap(
  prop_matrix_ct,
  cluster_rows = FALSE,
  cluster_columns = TRUE,
  col = colorRampPalette(c("white", "#B2182B"))(100),
  column_title = "Cell type composition per niche",
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8),
  column_names_rot = 45,
  column_names_max_height = max_text_width(colnames(prop_matrix_ct), gp = gpar(fontsize = 8)),
  rect_gp = gpar(col = "grey80", lwd = 0.5),
  heatmap_legend_param = list(title = "Proportion")
)

# Draw with explicit padding
draw(ht, padding = unit(c(2, 2, 2, 2), "cm"))  # bottom, left, top, right
```


```{r no_niche_celltype_heatmap, eval=!has_banksy || !has_annotation_niche}
cat("Niche x Cell type heatmap not available (no BANKSY results or no cell type annotation).")
```

### Cell type markers

```{r celltype_markers_compute, eval=has_annotation}
Idents(scData) <- celltype_col
celltype_markers <- FindAllMarkers(scData, only.pos = TRUE,
                                   min.pct = 0.25, logfc.threshold = 0.25,
                                   verbose = FALSE)
if (nrow(celltype_markers) > 0) {
 celltype_markers$diff_pct <- abs(celltype_markers$pct.1 - celltype_markers$pct.2)
}

has_celltype_markers <- exists("celltype_markers") && nrow(celltype_markers) > 0
```

```{r celltype_markers_table, eval=has_annotation && has_celltype_markers}
celltype_markers_sorted <- celltype_markers |>
 dplyr::arrange(cluster, desc(avg_log2FC)) |>
 dplyr::select(-diff_pct)

ezInteractiveTableRmd(celltype_markers_sorted,
                     rowNames = FALSE, title = "Cell type marker genes")
```

```{r no_celltype_markers, eval=has_annotation && !has_celltype_markers}
cat("No significant cell type markers found.")
```

### Cell type marker dotplot

```{r celltype_dotplot_setup, eval=has_annotation && has_celltype_markers}
top5_celltype <- celltype_markers |>
 dplyr::group_by(cluster) |>
 dplyr::arrange(desc(diff_pct)) |>
 dplyr::slice_head(n = 5) |>
 dplyr::ungroup()

sorted_clusters <- sort(unique(as.character(top5_celltype$cluster)))
genesToPlot_celltype <- character()
for (ct in sorted_clusters) {
 ct_genes <- top5_celltype$gene[top5_celltype$cluster == ct]
 genesToPlot_celltype <- c(genesToPlot_celltype, setdiff(ct_genes, genesToPlot_celltype))
}
genesToPlot_celltype <- as.character(genesToPlot_celltype)
genesToPlot_celltype <- gsub("_", "-", genesToPlot_celltype)
genesToPlot_celltype <- intersect(genesToPlot_celltype, rownames(scData))
genesToPlot_celltype <- make.unique(genesToPlot_celltype)

has_genes_to_plot <- length(genesToPlot_celltype) > 0
```

```{r celltype_dotplot, fig.width=15, fig.height=20, eval=has_annotation && has_celltype_markers && has_genes_to_plot, results='asis'}
original_types <- as.character(scData@meta.data[[celltype_col]])
cells_with_celltype <- which(!is.na(original_types))

if (length(cells_with_celltype) < ncol(scData)) {
 n_na <- ncol(scData) - length(cells_with_celltype)
 cat("\n\n*Note:* Excluding ", n_na, " cells with unassigned cell types from dotplot.\n\n", sep = "")
}

scData_subset <- subset(scData, cells = colnames(scData)[cells_with_celltype])

cleaned_types <- gsub("[/\\\\]", "_", original_types[cells_with_celltype])
cleaned_types <- gsub("\\s+", "_", cleaned_types)
cleaned_types <- make.names(cleaned_types)
scData_subset$celltype_clean <- factor(cleaned_types)

Idents(scData_subset) <- "celltype_clean"

tryCatch({
 p <- DotPlot(scData_subset, features = rev(genesToPlot_celltype), group.by = "celltype_clean") +
   coord_flip() +
   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
   labs(x = "Gene", y = "Cell type")
 print(p)
}, error = function(e) {
 tryCatch({
   cat("\n\n*Note:* Using fewer markers due to plotting constraints.\n\n")
   top_genes <- unique(head(genesToPlot_celltype, 20))
   p <- DotPlot(scData_subset, features = rev(top_genes), group.by = "celltype_clean") +
     coord_flip() +
     theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
     labs(x = "Gene", y = "Cell type")
   print(p)
 }, error = function(e2) {
   cat("Cell type marker dotplot could not be generated:", e2$message, "\n")
 })
})

rm(scData_subset)
```

```{r no_celltype_dotplot, eval=has_annotation && has_celltype_markers && !has_genes_to_plot}
cat("No valid marker genes found for dotplot.")
```

### RCTD classification

<details>
 <summary>RCTD classification definitions</summary>

```{r rctd_class_definitions}
rctd_def_tbl <- tibble::tibble(
 Class = c("singlet", "doublet_certain", "doublet_uncertain", "reject", "unassigned"),
 Definition = c(
   "Cell confidently assigned to a single cell type. The best-fit cell type has a significantly higher likelihood than the second-best type.",
   "Cell identified as containing two distinct cell types with high confidence. Both cell types have significant likelihood scores.",
   "Cell potentially contains two cell types, but assignment confidence is lower. The second cell type likelihood is moderately elevated.",
   "Cell could not be reliably classified. This occurs when: (1) no cell type fits well, (2) too few transcripts detected, or (3) the expression profile doesn't match any reference type.",
   "Cell was excluded by RCTD before analysis due to the UMI_min threshold. These cells have too few transcripts for reliable cell type assignment."
 ),
 Color = c("#2ecc71 (green)", "#e74c3c (red)", "#f39c12 (orange)", "#95a5a6 (gray)", "#3498db (blue)")
)

ezKable(rctd_def_tbl)
```

**Note:** High proportions of singlets indicate good data quality and reference match. Doublets may indicate overlapping cells or spots spanning cell boundaries. Rejects warrant investigation - they may represent novel cell types not in the reference, dying cells, or technical artifacts.

</details>

```{r rctd_class_combined, fig.width=16, fig.height=7, eval=has_spot_class}
# Define RCTD class colors including unassigned
rctd_class_colors <- c("singlet" = "#2ecc71", "doublet_certain" = "#e74c3c",
                      "doublet_uncertain" = "#f39c12", "reject" = "#95a5a6",
                      "unassigned" = "#3498db")

# Spatial plot
p1 <- ImageDimPlot(scData, fov = names(scData@images)[1], group.by = "spot_class",
            axes = TRUE, cols = rctd_class_colors, size = 0.5) +
 ggtitle("RCTD classification - Spatial") +
 labs(fill = "RCTD class")

# Barplot with unassigned
class_vector <- as.character(scData$spot_class)
class_vector[is.na(class_vector)] <- "unassigned"
class_counts <- as.data.frame(table(class_vector, useNA = "no"))
colnames(class_counts) <- c("Class", "Count")
class_counts <- class_counts[order(class_counts$Count, decreasing = TRUE), ]
class_counts$Class <- factor(class_counts$Class, levels = class_counts$Class)
class_counts$Proportion <- class_counts$Count / sum(class_counts$Count)

p2 <- ggplot(class_counts, aes(x = Class, y = Count, fill = Class)) +
 geom_bar(stat = "identity") +
 scale_fill_manual(values = rctd_class_colors) +
 geom_text(aes(label = paste0(round(Proportion * 100, 1), "%")),
           vjust = -0.5, size = 4) +
 theme_minimal(base_size = 12) +
 theme(legend.position = "none",
       axis.text.x = element_text(angle = 45, hjust = 1)) +
 labs(title = "RCTD classification distribution",
      x = "Classification", y = "Number of cells") +
 scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

(p1 + p2) + plot_layout(widths = c(1.3, 1))
```

```{r rctd_class_table, eval=has_spot_class}
class_vector <- as.character(scData$spot_class)
class_vector[is.na(class_vector)] <- "unassigned"
class_table <- as.data.frame(table(class_vector))
colnames(class_table) <- c("Class", "Count")
class_table$Proportion <- class_table$Count / sum(class_table$Count)
class_table <- class_table[order(class_table$Count, decreasing = TRUE), ]

ezInteractiveTableRmd(class_table, rowNames = FALSE,
                     title = "RCTD classification summary") |>
 DT::formatPercentage(columns = "Proportion", digits = 2)
```

```{r no_rctd_class, eval=!has_spot_class}
cat("RCTD classification (spot_class) not available. Run RCTD annotation to generate.")
```

### Prediction scores

```{r prediction_scores_violin, fig.width=14, fig.height=10, eval=has_annotation}
if (!exists("celltype_colors")) {
 celltypes_unique <- sort(unique(na.omit(scData@meta.data[[celltype_col]])))
 n_celltypes <- length(celltypes_unique)
 extended_colors <- c(
   cluster_colors,
   RColorBrewer::brewer.pal(9, "Set1"),
   RColorBrewer::brewer.pal(8, "Set2"),
   RColorBrewer::brewer.pal(12, "Set3")
 )
 extended_colors <- unique(extended_colors)
 if (n_celltypes > length(extended_colors)) {
   extended_colors <- rep(extended_colors, length.out = n_celltypes)
 }
 celltype_colors <- setNames(extended_colors[1:n_celltypes], celltypes_unique)
}

weight_cols <- grep("^rctd\\.weight\\.", colnames(scData@meta.data), value = TRUE)

if (length(weight_cols) > 0) {
 weight_matrix <- scData@meta.data[, weight_cols, drop = FALSE]
 max_weights <- apply(weight_matrix, 1, max, na.rm = TRUE)
 max_weights[is.infinite(max_weights)] <- NA
 scData$rctd_max_weight <- max_weights

 valid_weights <- sum(!is.na(scData$rctd_max_weight))

 if (valid_weights > 0) {
   if (!exists("truncate_label")) {
     truncate_label <- function(x, max_chars = 35) {
       ifelse(nchar(x) > max_chars, paste0(substr(x, 1, max_chars - 3), "..."), x)
     }
   }
   x_labels <- setNames(truncate_label(celltypes_unique), celltypes_unique)

   p1 <- VlnPlot(scData, features = "rctd_max_weight",
                 group.by = celltype_col, pt.size = 0, cols = celltype_colors) +
     scale_x_discrete(labels = x_labels) +
     stat_summary(fun = median, geom = "crossbar", width = 0.5,
                  color = "white", linewidth = 0.8) +
     ggtitle("RCTD max weight by cell type") +
     ylab("Max prediction weight") +
     theme_minimal(base_size = 12) +
     theme(legend.position = "none",
           axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
           axis.title.x = element_blank())
   print(p1)
 } else {
   cat("No valid RCTD prediction weights available for violin plot.")
 }

} else {
 score_cols <- grep("prediction.score", colnames(scData@meta.data), value = TRUE)

 if (length(score_cols) > 0 && "prediction.score.max" %in% score_cols) {
   if (!exists("truncate_label")) {
     truncate_label <- function(x, max_chars = 35) {
       ifelse(nchar(x) > max_chars, paste0(substr(x, 1, max_chars - 3), "..."), x)
     }
   }
   x_labels <- setNames(truncate_label(celltypes_unique), celltypes_unique)

   p1 <- VlnPlot(scData, features = "prediction.score.max",
                 group.by = celltype_col, pt.size = 0, cols = celltype_colors) +
     scale_x_discrete(labels = x_labels) +
     stat_summary(fun = median, geom = "crossbar", width = 0.5,
                  color = "white", linewidth = 0.8) +
     ggtitle("Prediction score by cell type") +
     theme_minimal(base_size = 12) +
     theme(legend.position = "none",
           axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
   print(p1)
 } else {
   cat("Prediction scores/weights not available in metadata.")
 }
}
```

```{r prediction_scores_spatial, fig.width=14, fig.height=6, eval=has_annotation, results='asis'}
weight_cols_sp <- grep("^rctd\\.weight\\.", colnames(scData@meta.data), value = TRUE)

if (length(weight_cols_sp) > 0 && "rctd_max_weight" %in% colnames(scData@meta.data)) {
 valid_weights_sp <- sum(!is.na(scData$rctd_max_weight))

 if (valid_weights_sp > 0) {
   tryCatch({
     scData$rctd_max_weight_plot <- scData$rctd_max_weight
     scData$rctd_max_weight_plot[is.na(scData$rctd_max_weight_plot)] <- 0

     p1 <- ImageFeaturePlot(scData, features = "rctd_max_weight_plot",
                            max.cutoff = "q95", size = 0.5) +
       scale_fill_gradientn(colors = viridis::plasma(100),
                            limits = c(0, 1),
                            name = "Max Weight") +
       ggtitle("Prediction confidence - Spatial") +
       theme(plot.title = element_text(hjust = 0.5, face = "bold"))

     p2 <- FeaturePlot(scData, features = "rctd_max_weight_plot",
                       reduction = "umap", max.cutoff = "q95", raster = FALSE, order = TRUE) +
       scale_color_gradientn(colors = viridis::plasma(100),
                             limits = c(0, 1),
                             name = "Max Weight") +
       coord_fixed() +
       ggtitle("Prediction confidence - UMAP") +
       theme_void() +
       theme(plot.title = element_text(hjust = 0.5, face = "bold"))

     print((p1 + p2) + plot_layout(widths = c(1.2, 1)))

     na_count <- sum(is.na(scData$rctd_max_weight))
     if (na_count > 0) {
       cat("\n\n*Note:* ", na_count, " cells without RCTD weights shown as 0 (dark color).\n\n", sep = "")
     }
   }, error = function(e) {
     cat("Spatial prediction confidence plot not available:", e$message, "\n")
   })
 } else {
   cat("No valid RCTD prediction weights available for spatial plot.")
 }
} else {
 cat("RCTD prediction weights not found in metadata.")
}
```

```{r prediction_scores_summary, eval=has_annotation}
if ("rctd_max_weight" %in% colnames(scData@meta.data)) {
 valid_count <- sum(!is.na(scData$rctd_max_weight))
 if (valid_count > 0) {
   cat("Cells with valid prediction weights:", valid_count,
       "of", ncol(scData), "total cells\n")
   cat("Mean max weight:", round(mean(scData$rctd_max_weight, na.rm = TRUE), 3), "\n")
   cat("Median max weight:", round(median(scData$rctd_max_weight, na.rm = TRUE), 3), "\n")
 }
}
```

## Links

### Interactive Vitessce explorer

The link below will take you to an interactive Vitessce viewer with linked spatial, UMAP,
and gene expression views. You can explore cell types, clusters, and marker genes interactively.

[Vitessce Explorer](`r {sample_name <- unique(scData$Sample)[1]; relative_path <- file.path(param$resultDir, sample_name); paste0("https://fgcz-shiny.uzh.ch/app/exploreVitessceXenium/?data=", relative_path, "/scData.qs2")}`){target="_blank"}

### Data files

##### Positive markers of each cluster

[posMarkers](posMarkers.xlsx)

##### Positive markers of each BANKSY niche

[posMarkersBanksy](posMarkersBanksy.xlsx)

##### The final Seurat object is [here](scData.qs2)

### Xenium Explorer import files

Download these CSV files to import annotations into Xenium Explorer:

```{r xenium_explorer_links, results='asis'}
cat("- [clusters_for_explorer.csv](clusters_for_explorer.csv) - Seurat clusters\n")

if (file.exists("niches_for_explorer.csv")) {
 cat("- [niches_for_explorer.csv](niches_for_explorer.csv) - BANKSY spatial niches\n")
}

if (file.exists("celltypes_for_explorer.csv")) {
 cat("- [celltypes_for_explorer.csv](celltypes_for_explorer.csv) - RCTD cell type annotations\n")
}
```

<details>
 <summary>How to import into Xenium Explorer</summary>

1. Download the CSV file(s) above
2. Open your Xenium dataset in Xenium Explorer
3. Go to: **File -> Import -> Cell Categories**
4. Select the downloaded CSV file
5. Cell annotations will appear as a new grouping option in the dropdown menu

**Tip:** You can import multiple CSV files to compare different annotation methods!
</details>

## Parameters

### Parameters used

```{r parameters}
param_list <- list(
 "Min counts" = ifelse(!is.null(param$minCounts), param$minCounts, 10),
 "Min features" = ifelse(!is.null(param$minFeatures), param$minFeatures, 5),
 "Clustering resolution" = ifelse(!is.null(param$clusterResolution), param$clusterResolution, 0.5),
 "BANKSY lambda" = ifelse(!is.null(param$lambda), param$lambda, 0.8),
 "Niche resolution" = ifelse(!is.null(param$nicheResolution),
                              param$nicheResolution, 0.5)
)

ezListAsKable(param_list, col.names = c("Parameter", "Value"),
             caption = "Parameters used")
```

## Session information

```{r session_info}
format(Sys.time(), "%Y-%m-%d %H:%M:%S")
ezSessionInfo()
```
